### 2020-06-06 src/bbData2020.Rmd

---
title: "descriptive anaysis"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

```{r settings, include = FALSE}
library(dplyr)
library(reshape2)
library(lubridate)
library(ggplot2)
library(openair)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(tmap)
library(classInt)
library(RColorBrewer)
library(knitr)
library(rprojroot)

# knitr::opts_chunk$set(echo = FALSE
#                       , comment = NA
#                       , warning = FALSE
#                       , error = FALSE
#                       , message = FALSE
#                       , tidy = TRUE)

# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# This is the project path
path <- find_rstudio_root_file()

```

```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
#upload data file
path.data <- paste(path, "./data/raw/speedtest.csv", sep = "")

#bb2020 <- read.csv("./data/raw/speedtest.csv")
bb2020 <- read.csv(path.data)

#name columns
names(bb2020)[1:6] <- c("datetext","speeddown","speedup","provider","lat", "lon")

str(bb2020)
summary(bb2020) #2030205 obs
#conversions needed to numeric? df$col <- as.numeric(df$col) 
#remove empty rows or coerced to NA? df[complete.cases(df),] or df <- df[!is.na(df$col),]

#remove outliers based on Riddlesden and Singleton 2014 for slower end
#check fastest upload and download speeds commercially available from Virgin Media (Gigaclear offers even faster!)
bb2020 <- bb2020[bb2020$speeddown>512,] 
bb2020 <- bb2020[bb2020$speeddown<362000,]
bb2020 <- bb2020[bb2020$speedup<21000,] #1839333

#remove below seconds, create columns for date and hour
bb2020$datetext <- strtrim(x = bb2020$datetext, width = 19) 
bb2020$datetext2 <- bb2020$datetext
bb2020$datetext2 <- strtrim(x = bb2020$datetext2, width = 13)
bb2020$datetext2 <- colsplit(bb2020$datetext2, pattern = " ", c("date", "hour"))
bb2020$hour <- bb2020$datetext2$hour
bb2020$date <- bb2020$datetext2$date
#delete datetext2 column(s)
bb2020 <- bb2020[ ,c(1:6,8:9)]
#convert to date form datetext and date columns
bb2020$datetext <- as.POSIXct(strptime(bb2020$datetext, format = "%Y-%m-%d %H:%M:%S", "GMT")) 
bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
```

# Temporal analysis

```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
#add column for day of the week
bb2020$weekday <- wday(bb2020$date, label = TRUE)
#create 2 dataframes for 2 matching periods
Wks2019 <- selectByDate(bb2020, start = "2019-01-21", end = "2019-06-02")
names(Wks2019)[1:2] <- c("dateOnly", "date")
Wks2020 <- selectByDate(bb2020, start = "2020-01-20", end = "2020-05-31")
names(Wks2020)[1:2] <- c("dateOnly", "date")
#timeplots
TimeVar2019 <- timeVariation(Wks2019, pollutant = "speeddown", statistic = "mean")
TimeVar2019up <- timeVariation(Wks2019, pollutant = "speedup", statistic = "mean")
TimeVar2020 <- timeVariation(Wks2020, pollutant = "speeddown", statistic = "mean")
TimeVar2020up <- timeVariation(Wks2020, pollutant = "speedup", statistic = "mean")
timePlot(Wks2019, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
timePlot(Wks2020, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
timePlot(Wks2019, pollutant = "speedup", avg.time = "day", smooth = TRUE)
timePlot(Wks2020, pollutant = "speedup", avg.time = "day", smooth = TRUE)
```

# Frequency per date

```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
DateCount19 <- count(Wks2019, dateOnly)
names(DateCount19) <- c("Date19", "tests19")
HrCount19 <- count(Wks2019, date, hour)
HrSpeed19 <- summarise(group_by(Wks2019, date, hour), mean(speeddown))
HrStats19 <- left_join(HrCount19, HrSpeed19)
names(HrStats19) <- c("Date19", "Hour19", "test19", "meanSp19")
DateCount20 <- count(Wks2020, dateOnly)
names(DateCount20) <- c("Date20", "tests20")
HrCount20 <- count(Wks2020, date, hour)
HrSpeed20 <- summarise(group_by(Wks2020, date, hour), mean(speeddown))
HrStats20 <- left_join(HrCount20, HrSpeed20)
names(HrStats20) <- c("Date20", "Hour20", "test20", "meanSp20")
DateCount <- cbind(DateCount19, DateCount20)
HrStats <- cbind(HrStats19, HrStats20)  # I GET AN ERROR HERE
write.csv(DateCount, "DailyTestFreq.csv")
write.csv(HrStats, "HrTestFreqMean.csv")
Plot2019 <- ggplot(DateCount19, aes(Date19, tests19))
Plot2019 + geom_line()
Plot2020 <- ggplot(DateCount20, aes(Date20, tests20))
Plot2020 + geom_line()
```

# Spatial analysis

```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
#convert to spatial object
coords_bb <- cbind(bb2020$lon, bb2020$lat)
bbNUTS3 <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020))
proj4string(bbNUTS3) <- CRS("+init=epsg:4326") #define projection

#get NUTS3 from up a level and read in (145 NUTS3)
NUTS3 <- readOGR("../NUTS_Level_3_January_2015_Super_Generalised_Clipped_Boundaries_in_England_and_Wales.shp") 
NUTS3 <- spTransform(NUTS3, CRS("+init=epsg:4326"))                                
NUTS3@data$nuts315nm <- as.character(NUTS3@data$nuts315nm)

#spatial join to NUTS3
bbNUTS3sp <- over(bbNUTS3, NUTS3[, "nuts315nm"]) #not a spatial object
bbNUTS3$nuts315nm <- bbNUTS3sp$nuts315nm

#create dataframe object to analyse (remove lat and lon?)
bb2020sp <- bbNUTS3@data
#% obs NA for NUTS3?

#summarise by date and geography
NUTS3datecount <- count(bb2020sp, nuts315nm, date)
NUTS3dateAvg <- summarise(group_by(NUTS3datecount, nuts315nm), mean(n))

#map NUTS3 mean test frequency per date
NUTS3@data <- left_join(NUTS3@data, NUTS3dateAvg, by = "nuts315nm")
var <- NUTS3@data[ ,'mean(n)']
breaks <- classIntervals(var, n = 5, style = "quantile")
my_colours <- brewer.pal(5, "Blues")
plot(NUTS3, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE,
     border = rgb(0.8, 0.8, 0.8, 0))
legend(x = "top", legend = leglabs(breaks$brks), fill = my_colours, bty = "n", cex = 0.5)
```

### 2020-06-06 src/bbData2020.Rmd
further updates to explore temporal features

@@
 Wks2019 <- selectByDate(bb2020, start = "2019-01-21", end = "2019-06-02")
-names(Wks2019)[1:2] <- c("dateOnly", "date")
+names(Wks2019)[1:2] <- c("dateOnly", "date") #413788 obs
 Wks2020 <- selectByDate(bb2020, start = "2020-01-20", end = "2020-05-31")
-names(Wks2020)[1:2] <- c("dateOnly", "date")
+names(Wks2020)[1:2] <- c("dateOnly", "date") #521773 obs
 #timeplots
-TimeVar2019 <- timeVariation(Wks2019, pollutant = "speeddown", statistic = "mean")
-TimeVar2019up <- timeVariation(Wks2019, pollutant = "speedup", statistic = "mean")
-TimeVar2020 <- timeVariation(Wks2020, pollutant = "speeddown", statistic = "mean")
-TimeVar2020up <- timeVariation(Wks2020, pollutant = "speedup", statistic = "mean")
-timePlot(Wks2019, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
-timePlot(Wks2020, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
-timePlot(Wks2019, pollutant = "speedup", avg.time = "day", smooth = TRUE)
-timePlot(Wks2020, pollutant = "speedup", avg.time = "day", smooth = TRUE)
+TimeVar2019 <- timeVariation(Wks2019, pollutant = "speeddown", statistic = "mean", name.pol = "speedown 2019")
+TimeVar2019up <- timeVariation(Wks2019, pollutant = "speedup", statistic = "mean", name.pol = "speedup 2019")
+TimeVar2020 <- timeVariation(Wks2020, pollutant = "speeddown", statistic = "mean", name.pol = "speeddown 2020")
+TimeVar2020up <- timeVariation(Wks2020, pollutant = "speedup", statistic = "mean", name.pol = "speedup 2020")
+timePlot(Wks2019, pollutant = "speeddown", avg.time = "day", smooth = TRUE, name.pol = "speeddown 2019")
+timePlot(Wks2020, pollutant = "speeddown", avg.time = "day", smooth = TRUE, name.pol = "speeddown 2020")
+timePlot(Wks2019, pollutant = "speedup", avg.time = "day", smooth = TRUE, name.pol = "speedup 2019")
+timePlot(Wks2020, pollutant = "speedup", avg.time = "day", smooth = TRUE, name.pol = "speedup 2020")
 ```
@@
 names(DateCount19) <- c("Date19", "tests19")
-HrCount19 <- count(Wks2019, date, hour)
-HrSpeed19 <- summarise(group_by(Wks2019, date, hour), mean(speeddown))
+HrCount19 <- count(Wks2019, dateOnly, hour)
+HrSpeed19 <- summarise(group_by(Wks2019, dateOnly, hour), mean(speeddown))
+HrSpeed19up <- summarise(group_by(Wks2019, dateOnly, hour), mean(speedup))
 HrStats19 <- left_join(HrCount19, HrSpeed19)
-names(HrStats19) <- c("Date19", "Hour19", "test19", "meanSp19")
+HrStats19 <- left_join(HrStats19, HrSpeed19up)
+names(HrStats19) <- c("dateOnly", "hour", "test19", "meanSp19", "meanSp19up")
 DateCount20 <- count(Wks2020, dateOnly)
 names(DateCount20) <- c("Date20", "tests20")
-HrCount20 <- count(Wks2020, date, hour)
-HrSpeed20 <- summarise(group_by(Wks2020, date, hour), mean(speeddown))
+HrCount20 <- count(Wks2020, dateOnly, hour)
+HrSpeed20 <- summarise(group_by(Wks2020, dateOnly, hour), mean(speeddown))
+HrSpeed20up <- summarise(group_by(Wks2020, dateOnly, hour), mean(speedup))
 HrStats20 <- left_join(HrCount20, HrSpeed20)
-names(HrStats20) <- c("Date20", "Hour20", "test20", "meanSp20")
+HrStats20 <- left_join(HrStats20, HrSpeed20up)
+names(HrStats20) <- c("dateOnly", "hour", "test20", "meanSp20", "meanSp20up")
 DateCount <- cbind(DateCount19, DateCount20)
-HrStats <- cbind(HrStats19, HrStats20)  # I GET AN ERROR HERE
-write.csv(DateCount, "DailyTestFreq.csv")
-write.csv(HrStats, "HrTestFreqMean.csv")
+HrStats <- cbind(HrStats19, HrStats20)
+#add columns tests per hour, mean speeddown per hr, mean speed up per hr
+Wks2019 <- left_join(Wks2019, HrStats19, by = c("dateOnly", "hour"))
+Wks2020 <- left_join(Wks2020, HrStats20, by = c("dateOnly", "hour"))
+#frequency plot
+timePlot(Wks2019, pollutant = "test19", avg.time = "hour", smooth = TRUE, name.pol = "hourlyTests 2019")
+timePlot(Wks2020, pollutant = "tests20", avg.time = "hour", smooth = TRUE, name.pol = "hourlyTests 2020")
+TimeVar2019n <- timeVariation(Wks2019, pollutant = "test19", statistic = "mean", name.pol = "hourlytests 2019")
+TimeVar2020n <- timeVariation(Wks2020, pollutant = "test20", statistic = "mean", name.pol = "hourlytests 2020")
 Plot2019 <- ggplot(DateCount19, aes(Date19, tests19))
@@
 Plot2020 + geom_line()
+
 ```

### 2020-06-06 src/bbData2020.R

@@ -43,36 +43,64 @@ names(Wks2019)[1:2] <- c("dateOnly", "date")
 Wks2020 <- selectByDate(bb2020, start = "2020-01-20", end = "2020-05-31")
 names(Wks2020)[1:2] <- c("dateOnly", "date")
 #timeplots
-TimeVar2019 <- timeVariation(Wks2019, pollutant = "speeddown", statistic = "mean")
-TimeVar2019up <- timeVariation(Wks2019, pollutant = "speedup", statistic = "mean")
-TimeVar2020 <- timeVariation(Wks2020, pollutant = "speeddown", statistic = "mean")
-TimeVar2020up <- timeVariation(Wks2020, pollutant = "speedup", statistic = "mean")
-timePlot(Wks2019, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
-timePlot(Wks2020, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
-timePlot(Wks2019, pollutant = "speedup", avg.time = "day", smooth = TRUE)
-timePlot(Wks2020, pollutant = "speedup", avg.time = "day", smooth = TRUE)
+TimeVar2019 <- timeVariation(Wks2019, pollutant = "speeddown", 
+                        statistic = "mean", name.pol = "speedown 2019")
+TimeVar2019up <- timeVariation(Wks2019, pollutant = "speedup",
+                        statistic = "mean", name.pol = "speedup 2019")
+TimeVar2020 <- timeVariation(Wks2020, pollutant = "speeddown", 
+                        statistic = "mean", name.pol = "speeddown 2020")
+TimeVar2020up <- timeVariation(Wks2020, pollutant = "speedup",
+                        statistic = "mean", name.pol = "speedup 2020")
+timePlot(Wks2019, pollutant = "speeddown", avg.time = "day", 
+         smooth = TRUE, name.pol = "speeddown 2019")
+timePlot(Wks2020, pollutant = "speeddown", avg.time = "day", 
+         smooth = TRUE, name.pol = "speeddown 2020")
+timePlot(Wks2019, pollutant = "speedup", avg.time = "day", 
+         smooth = TRUE, name.pol = "speedup 2019")
+timePlot(Wks2020, pollutant = "speedup", avg.time = "day", 
[…]

### 2020-06-08 src/bbData2020.Rmd
fix images in .Rmd

--- 2020-06-06 src/bbData2020.Rmd
@@
 
-# Temporal analysis
+## Temporal analysis
 
-```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
+```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE, fig.height=15, fig.width=15}
 #add column for day of the week
@@
 Wks2019 <- selectByDate(bb2020, start = "2019-01-21", end = "2019-06-02")
-names(Wks2019)[1:2] <- c("dateOnly", "date")
+names(Wks2019)[1:2] <- c("dateOnly", "date") #413788 obs
 Wks2020 <- selectByDate(bb2020, start = "2020-01-20", end = "2020-05-31")
-names(Wks2020)[1:2] <- c("dateOnly", "date")
+names(Wks2020)[1:2] <- c("dateOnly", "date") #521773 obs
 #timeplots
-TimeVar2019 <- timeVariation(Wks2019, pollutant = "speeddown", statistic = "mean")
-TimeVar2019up <- timeVariation(Wks2019, pollutant = "speedup", statistic = "mean")
-TimeVar2020 <- timeVariation(Wks2020, pollutant = "speeddown", statistic = "mean")
-TimeVar2020up <- timeVariation(Wks2020, pollutant = "speedup", statistic = "mean")
-timePlot(Wks2019, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
-timePlot(Wks2020, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
-timePlot(Wks2019, pollutant = "speedup", avg.time = "day", smooth = TRUE)
-timePlot(Wks2020, pollutant = "speedup", avg.time = "day", smooth = TRUE)
+TimeVar2019 <- timeVariation(Wks2019, pollutant = "speeddown", statistic = "mean", name.pol = "speedown 2019")
+TimeVar2019up <- timeVariation(Wks2019, pollutant = "speedup", statistic = "mean", name.pol = "speedup 2019")
+TimeVar2020 <- timeVariation(Wks2020, pollutant = "speeddown", statistic = "mean", name.pol = "speeddown 2020")
+TimeVar2020up <- timeVariation(Wks2020, pollutant = "speedup", statistic = "mean", name.pol = "speedup 2020")
+timePlot(Wks2019, pollutant = "speeddown", avg.time = "day", smooth = TRUE, name.pol = "speeddown 2019")
+timePlot(Wks2020, pollutant = "speeddown", avg.time = "day", smooth = TRUE, name.pol = "speeddown 2020")
+timePlot(Wks2019, pollutant = "speedup", avg.time = "day", smooth = TRUE, name.pol = "speedup 2019")
+timePlot(Wks2020, pollutant = "speedup", avg.time = "day", smooth = TRUE, name.pol = "speedup 2020")
 ```
 
-# Frequency per date
+## Frequency per date
 
-```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
+```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE, fig.height=15, fig.width=15}
 DateCount19 <- count(Wks2019, dateOnly)
 names(DateCount19) <- c("Date19", "tests19")
-HrCount19 <- count(Wks2019, date, hour)
-HrSpeed19 <- summarise(group_by(Wks2019, date, hour), mean(speeddown))
+HrCount19 <- count(Wks2019, dateOnly, hour)
+HrSpeed19 <- summarise(group_by(Wks2019, dateOnly, hour), mean(speeddown))
+HrSpeed19up <- summarise(group_by(Wks2019, dateOnly, hour), mean(speedup))
 HrStats19 <- left_join(HrCount19, HrSpeed19)
-names(HrStats19) <- c("Date19", "Hour19", "test19", "meanSp19")
+HrStats19 <- left_join(HrStats19, HrSpeed19up)
+names(HrStats19) <- c("dateOnly", "hour", "test19", "meanSp19", "meanSp19up")
 DateCount20 <- count(Wks2020, dateOnly)
 names(DateCount20) <- c("Date20", "tests20")
-HrCount20 <- count(Wks2020, date, hour)
-HrSpeed20 <- summarise(group_by(Wks2020, date, hour), mean(speeddown))
+HrCount20 <- count(Wks2020, dateOnly, hour)
+HrSpeed20 <- summarise(group_by(Wks2020, dateOnly, hour), mean(speeddown))
+HrSpeed20up <- summarise(group_by(Wks2020, dateOnly, hour), mean(speedup))
 HrStats20 <- left_join(HrCount20, HrSpeed20)
-names(HrStats20) <- c("Date20", "Hour20", "test20", "meanSp20")
+HrStats20 <- left_join(HrStats20, HrSpeed20up)
+names(HrStats20) <- c("dateOnly", "hour", "test20", "meanSp20", "meanSp20up")
 DateCount <- cbind(DateCount19, DateCount20)
-HrStats <- cbind(HrStats19, HrStats20)  # I GET AN ERROR HERE
-write.csv(DateCount, "DailyTestFreq.csv")
-write.csv(HrStats, "HrTestFreqMean.csv")
+HrStats <- cbind(HrStats19, HrStats20)
+#add columns tests per hour, mean speeddown per hr, mean speed up per hr
+Wks2019 <- left_join(Wks2019, HrStats19, by = c("dateOnly", "hour"))
+Wks2020 <- left_join(Wks2020, HrStats20, by = c("dateOnly", "hour"))
+#frequency plot
+timePlot(Wks2019, pollutant = "test19", avg.time = "hour", smooth = TRUE, name.pol = "hourlyTests 2019")
+timePlot(Wks2020, pollutant = "test20", avg.time = "hour", smooth = TRUE, name.pol = "hourlyTests 2020")
+TimeVar2019n <- timeVariation(Wks2019, pollutant = "test19", statistic = "mean", name.pol = "hourlytests 2019")
+TimeVar2020n <- timeVariation(Wks2020, pollutant = "test20", statistic = "mean", name.pol = "hourlytests 2020")
 Plot2019 <- ggplot(DateCount19, aes(Date19, tests19))
@@
 Plot2020 + geom_line()
+
 ```
 
-# Spatial analysis
+## Spatial analysis
 
-```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
-#convert to spatial object
-coords_bb <- cbind(bb2020$lon, bb2020$lat)
-bbNUTS3 <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020))
-proj4string(bbNUTS3) <- CRS("+init=epsg:4326") #define projection
+<!-- # Ctrl + Shift + c to remove all the below comments -->
+<!-- ```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE, fig.height=15, fig.width=10} -->
+<!-- #convert to spatial object -->
+<!-- coords_bb <- cbind(bb2020$lon, bb2020$lat) -->
+<!-- bbNUTS3 <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020)) -->
+<!-- proj4string(bbNUTS3) <- CRS("+init=epsg:4326") #define projection -->
 
-#get NUTS3 from up a level and read in (145 NUTS3)
-NUTS3 <- readOGR("../NUTS_Level_3_January_2015_Super_Generalised_Clipped_Boundaries_in_England_and_Wales.shp") 
-NUTS3 <- spTransform(NUTS3, CRS("+init=epsg:4326"))                                
-NUTS3@data$nuts315nm <- as.character(NUTS3@data$nuts315nm)
+<!-- #get NUTS3 from up a level and read in (145 NUTS3) -->
+<!-- NUTS3 <- readOGR("../NUTS_Level_3_January_2015_Super_Generalised_Clipped_Boundaries_in_England_and_Wales.shp") -->
+<!-- NUTS3 <- spTransform(NUTS3, CRS("+init=epsg:4326")) -->
+<!-- NUTS3@data$nuts315nm <- as.character(NUTS3@data$nuts315nm) -->
 
-#spatial join to NUTS3
-bbNUTS3sp <- over(bbNUTS3, NUTS3[, "nuts315nm"]) #not a spatial object
-bbNUTS3$nuts315nm <- bbNUTS3sp$nuts315nm
+<!-- #spatial join to NUTS3 -->
+<!-- bbNUTS3sp <- over(bbNUTS3, NUTS3[, "nuts315nm"]) #not a spatial object -->
+<!-- bbNUTS3$nuts315nm <- bbNUTS3sp$nuts315nm -->
 
-#create dataframe object to analyse (remove lat and lon?)
-bb2020sp <- bbNUTS3@data
-#% obs NA for NUTS3?
+<!-- #create dataframe object to analyse (remove lat and lon?) -->
+<!-- bb2020sp <- bbNUTS3@data -->
+<!-- #% obs NA for NUTS3? -->
 
-#summarise by date and geography
-NUTS3datecount <- count(bb2020sp, nuts315nm, date)
-NUTS3dateAvg <- summarise(group_by(NUTS3datecount, nuts315nm), mean(n))
+<!-- #summarise by date and geography -->
+<!-- NUTS3datecount <- count(bb2020sp, nuts315nm, date) -->
+<!-- NUTS3dateAvg <- summarise(group_by(NUTS3datecount, nuts315nm), mean(n)) -->
 
-#map NUTS3 mean test frequency per date
-NUTS3@data <- left_join(NUTS3@data, NUTS3dateAvg, by = "nuts315nm")
-var <- NUTS3@data[ ,'mean(n)']
-breaks <- classIntervals(var, n = 5, style = "quantile")
-my_colours <- brewer.pal(5, "Blues")
-plot(NUTS3, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE,
-     border = rgb(0.8, 0.8, 0.8, 0))
-legend(x = "top", legend = leglabs(breaks$brks), fill = my_colours, bty = "n", cex = 0.5)
-```
+<!-- #map NUTS3 mean test frequency per date -->
+<!-- NUTS3@data <- left_join(NUTS3@data, NUTS3dateAvg, by = "nuts315nm") -->
+<!-- var <- NUTS3@data[ ,'mean(n)'] -->
+<!-- breaks <- classIntervals(var, n = 5, style = "quantile") -->
+<!-- my_colours <- brewer.pal(5, "Blues") -->
+<!-- plot(NUTS3, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE, -->
+<!--      border = rgb(0.8, 0.8, 0.8, 0)) -->
+<!-- legend(x = "top", legend = leglabs(breaks$brks), fill = my_colours, bty = "n", cex = 0.5) -->
+<!-- ``` -->

### 2020-06-09 src/bbData2020.Rmd
Create temporal variables frequency and speed up and speed down mean and standard deviation by month, weekday, hour, date with and without night and weekend exclusions.

--- 2020-06-06 src/bbData2020.Rmd
@@
 bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
+#create dataframe with number of tests, mean and standard deviation by month for adding to dataframe later
+MonthFreq <- timeAverage(bb2020, avg.time = "month", statistic = "frequency")
+MonthMean <- timeAverage(bb2020, avg.time = "month", statistic = "mean")
+MonthSD <- timeAverage(bb2020, avg.time = "month", statistic = "sd")
+MonthStats <- cbind(MonthFreq[,1:2], MonthMean[,c(3:4)], 
+                    MonthSD[,c(3:4)])
+names(MonthStats) <- c("month", "monthlyTests", "monthlyMeanSpDown", 
+                       "monthlyMeanSpUp", "monthlySDSpDown", 
+                       "monthlySDSpUp")
+MonthStats$month <- month(MonthStats$month, label = TRUE, abbr = TRUE)
+#add column for month and day of the week to create stats below
+bb2020$month <- month(bb2020$date, label = TRUE, abbr = TRUE)
+bb2020$weekday <- wday(bb2020$date, label = TRUE)
+#divide MonthStats for future join with separate 2019 and 2020 dataframes
+MonthStats19 <- MonthStats[1:6,]
+MonthStats20 <- MonthStats[13:17,]
 ```
 
-# Temporal analysis
+# Temporal analysis - explore graphically
 
 ```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
-#add column for day of the week
-bb2020$weekday <- wday(bb2020$date, label = TRUE)
 #create 2 dataframes for 2 matching periods
 Wks2019 <- selectByDate(bb2020, start = "2019-01-21", end = "2019-06-02")
-names(Wks2019)[1:2] <- c("dateOnly", "date")
+names(Wks2019)[1:2] <- c("dateOnly", "date") #413788 obs
 Wks2020 <- selectByDate(bb2020, start = "2020-01-20", end = "2020-05-31")
-names(Wks2020)[1:2] <- c("dateOnly", "date")
+names(Wks2020)[1:2] <- c("dateOnly", "date") #521773 obs
 #timeplots
-TimeVar2019 <- timeVariation(Wks2019, pollutant = "speeddown", statistic = "mean")
-TimeVar2019up <- timeVariation(Wks2019, pollutant = "speedup", statistic = "mean")
-TimeVar2020 <- timeVariation(Wks2020, pollutant = "speeddown", statistic = "mean")
-TimeVar2020up <- timeVariation(Wks2020, pollutant = "speedup", statistic = "mean")
-timePlot(Wks2019, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
-timePlot(Wks2020, pollutant = "speeddown", avg.time = "day", smooth = TRUE)
-timePlot(Wks2019, pollutant = "speedup", avg.time = "day", smooth = TRUE)
-timePlot(Wks2020, pollutant = "speedup", avg.time = "day", smooth = TRUE)
+TimeVar2019 <- timeVariation(Wks2019, pollutant = "speeddown", statistic = "mean", name.pol = "speedown 2019")
+TimeVar2019up <- timeVariation(Wks2019, pollutant = "speedup", statistic = "mean", name.pol = "speedup 2019")
+TimeVar2020 <- timeVariation(Wks2020, pollutant = "speeddown", statistic = "mean", name.pol = "speeddown 2020")
+TimeVar2020up <- timeVariation(Wks2020, pollutant = "speedup", statistic = "mean", name.pol = "speedup 2020")
+timePlot(Wks2019, pollutant = "speeddown", avg.time = "day", smooth = TRUE, name.pol = "speeddown 2019")
+timePlot(Wks2020, pollutant = "speeddown", avg.time = "day", smooth = TRUE, name.pol = "speeddown 2020")
+timePlot(Wks2019, pollutant = "speedup", avg.time = "day", smooth = TRUE, name.pol = "speedup 2019")
+timePlot(Wks2020, pollutant = "speedup", avg.time = "day", smooth = TRUE, name.pol = "speedup 2020")
 ```
 
-# Frequency per date
+# Add temporal stats as column variables
 
 ```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
+#add stats for each month: test frequency, mean and standard deviation of speeddown and speedup
+#note that Jan 2019, Jun 2019 and Jan 2020 inc WHOLE month in stats and weekends and nights included
+Wks2019 <- left_join(Wks2019, MonthStats19)
+Wks2020 <- left_join(Wks2020, MonthStats20)
+#add stats for days of week, one set of stats with all hours and one set exc 00.00-06.00
+WkdyCount19 <- count(Wks2019, weekday)
+WkdyMeanSp19down <- summarise(group_by(Wks2019, weekday), mean(speeddown))
+WkdyMeanSp19up <- summarise(group_by(Wks2019, weekday), mean(speedup))
+WkdySDSp19down <- summarise(group_by(Wks2019, weekday), sd(speeddown))
+WkdySDSp19up <- summarise(group_by(Wks2019, weekday), sd(speedup))
+WkdyCount19ex <- count(Wks2019[which(Wks2019$hour != 0:5),], weekday)
+WkdyMeanSp19downex <- summarise(group_by(Wks2019[which(Wks2019$hour 
+                                != 0:5),], weekday), mean(speeddown))
+WkdyMeanSp19upex <- summarise(group_by(Wks2019[which(Wks2019$hour
+                                != 0:5),], weekday), mean(speedup))
+WkdySDSp19downex <- summarise(group_by(Wks2019[which(Wks2019$hour 
+                                != 0:5),], weekday), sd(speeddown))
+WkdySDSp19upex <- summarise(group_by(Wks2019[which(Wks2019$hour 
+                                != 0:5),], weekday), sd(speedup))
+WkdyStats19 <- cbind(WkdyCount19ex, WkdyCount19[,2], 
+                     WkdyMeanSp19downex[,2], WkdyMeanSp19down[,2],
+                     WkdyMeanSp19upex[,2], WkdyMeanSp19up[,2],
+                     WkdySDSp19downex[,2], WkdySDSp19down[,2],
+                     WkdySDSp19upex[,2],WkdySDSp19up[,2])
+names(WkdyStats19) <- c("weekday", "testsWkdyEx", "testsWkdy",
+                        "MeanSpDownWkdyEx", "MeanSpDownWkdy", 
+                        "MeanSpUpWkdyEx", "MeanSpUpWkdy", 
+                        "SDSpDownEx", "SDSpDown", "SDSpUpEx", "SDSpUp")
+Wks2019 <- left_join(Wks2019, WkdyStats19)
+WkdyCount20 <- count(Wks2020, weekday)
+WkdyMeanSp20down <- summarise(group_by(Wks2020, weekday), mean(speeddown))
+WkdyMeanSp20up <- summarise(group_by(Wks2020, weekday), mean(speedup))
+WkdySDSp20down <- summarise(group_by(Wks2020, weekday), sd(speeddown))
+WkdySDSp20up <- summarise(group_by(Wks2020, weekday), sd(speedup))
+WkdyCount20ex <- count(Wks2020[which(Wks2020$hour != 0:5),], weekday)
+WkdyMeanSp20downex <- summarise(group_by(Wks2020[which(Wks2020$hour 
+                                != 0:5),], weekday), mean(speeddown))
+WkdyMeanSp20upex <- summarise(group_by(Wks2020[which(Wks2020$hour
+                                != 0:5),], weekday), mean(speedup))
+WkdySDSp20downex <- summarise(group_by(Wks2020[which(Wks2020$hour 
+                                != 0:5),], weekday), sd(speeddown))
+WkdySDSp20upex <- summarise(group_by(Wks2020[which(Wks2020$hour 
+                                != 0:5),], weekday), sd(speedup))
+WkdyStats20 <- cbind(WkdyCount20ex, WkdyCount20[,2], 
+                     WkdyMeanSp20downex[,2], WkdyMeanSp20down[,2],
+                     WkdyMeanSp20upex[,2], WkdyMeanSp20up[,2],
+                     WkdySDSp20downex[,2], WkdySDSp20down[,2],
+                     WkdySDSp20upex[,2],WkdySDSp20up[,2])
+names(WkdyStats20) <- c("weekday", "testsWkdyEx", "testsWkdy",
+                        "MeanSpDownWkdyEx", "MeanSpDownWkdy", 
+                        "MeanSpUpWkdyEx", "MeanSpUpWkdy", 
+                        "SDSpDownEx", "SDSpDown", "SDSpUpEx", "SDSpUp")
+Wks2020 <- left_join(Wks2020, WkdyStats20)
+#add stats for hours of the day, one set with those hours from all days included and one set exc those hours on Sat-Sun
+HrCount19 <- count(Wks2019, hour)
+HrMeanSp19down <- summarise(group_by(Wks2019, hour), mean(speeddown))
+HrMeanSp19up <- summarise(group_by(Wks2019, hour), mean(speedup))
+HrSDSp19down <- summarise(group_by(Wks2019, hour), sd(speeddown))
+HrSDSp19up <- summarise(group_by(Wks2019, hour), sd(speedup))
+HrCount19ex <- count(Wks2019[which(Wks2019$weekday != "Sat" &
+                                   Wks2019$weekday != "Sun"),], hour)
+HrMeanSp19downex <- summarise(group_by(Wks2019[which(Wks2019$weekday 
+                              != "Sat" & Wks2019$weekday != "Sun"),], 
+                              hour), mean(speeddown))
+HrMeanSp19upex <- summarise(group_by(Wks2019[which(Wks2019$weekday 
+                              != "Sat" & Wks2019$weekday != "Sun"),], 
+                              hour), mean(speedup))
+HrSDSp19downex <- summarise(group_by(Wks2019[which(Wks2019$weekday 
+                              != "Sat" & Wks2019$weekday != "Sun"),], 
+                              hour), sd(speeddown))
+HrSDSp19upex <- summarise(group_by(Wks2019[which(Wks2019$weekday 
+                              != "Sat" & Wks2019$weekday != "Sun"),], 
+                              hour), sd(speedup))
+HrStats19 <- cbind(HrCount19ex, HrCount19[,2], 
+                     HrMeanSp19downex[,2], HrMeanSp19down[,2],
+                     HrMeanSp19upex[,2], HrMeanSp19up[,2],
+                     HrSDSp19downex[,2], HrSDSp19down[,2],
+                     HrSDSp19upex[,2],HrSDSp19up[,2])
+names(HrStats19) <- c("hour", "testsHrEx", "testsHr",
+                        "MeanSpDownHrEx", "MeanSpDownHr", 
+                        "MeanSpUpHrEx", "MeanSpUpHr", 
+                        "SDSpDownHrEx", "SDSpDownHr", "SDSpUpHrEx", 
+                        "SDSpUpHr")
+Wks2019 <- left_join(Wks2019, HrStats19)
+HrCount20 <- count(Wks2020, hour)
+HrMeanSp20down <- summarise(group_by(Wks2020, hour), mean(speeddown))
+HrMeanSp20up <- summarise(group_by(Wks2020, hour), mean(speedup))
+HrSDSp20down <- summarise(group_by(Wks2020, hour), sd(speeddown))
+HrSDSp20up <- summarise(group_by(Wks2020, hour), sd(speedup))
+HrCount20ex <- count(Wks2020[which(Wks2020$weekday != "Sat" &
+                               Wks2020$weekday != "Sun"),], hour)
+HrMeanSp20downex <- summarise(group_by(Wks2020[which(Wks2020$weekday 
+                               != "Sat" & Wks2020$weekday != "Sun"),], 
+                                       hour), mean(speeddown))
+HrMeanSp20upex <- summarise(group_by(Wks2020[which(Wks2020$weekday 
+                               != "Sat" & Wks2020$weekday != "Sun"),], 
+                                     hour), mean(speedup))
+HrSDSp20downex <- summarise(group_by(Wks2020[which(Wks2020$weekday 
+                               != "Sat" & Wks2020$weekday != "Sun"),], 
+                                     hour), sd(speeddown))
+HrSDSp20upex <- summarise(group_by(Wks2020[which(Wks2020$weekday 
+                                != "Sat" & Wks2020$weekday != "Sun"),],
+                                   hour), sd(speedup))
+HrStats20 <- cbind(HrCount20ex, HrCount20[,2], 
+                   HrMeanSp20downex[,2], HrMeanSp20down[,2],
+                   HrMeanSp20upex[,2], HrMeanSp20up[,2],
+                   HrSDSp20downex[,2], HrSDSp20down[,2],
+                   HrSDSp20upex[,2],HrSDSp20up[,2])
+names(HrStats20) <- c("hour", "testsHrEx", "testsHr",
+                      "MeanSpDownHrEx", "MeanSpDownHr", 
+                      "MeanSpUpHrEx", "MeanSpUpHr", 
+                      "SDSpDownHrEx", "SDSpDownHr", "SDSpUpHrEx", 
+                      "SDSpUpHr")
+Wks2020 <- left_join(Wks2020, HrStats20)
+#add daily stats - with and without 0:00-0:600
 DateCount19 <- count(Wks2019, dateOnly)
-names(DateCount19) <- c("Date19", "tests19")
-HrCount19 <- count(Wks2019, date, hour)
-HrSpeed19 <- summarise(group_by(Wks2019, date, hour), mean(speeddown))
-HrStats19 <- left_join(HrCount19, HrSpeed19)
-names(HrStats19) <- c("Date19", "Hour19", "test19", "meanSp19")
+DateSp19down <- summarise(group_by(Wks2019, dateOnly), mean(speeddown))
+DateSp19up <- summarise(group_by(Wks2019, dateOnly), mean(speedup))
+DateSDSp19down <- summarise(group_by(Wks2019, dateOnly), sd(speeddown))
+DateSDSp19up <- summarise(group_by(Wks2019, dateOnly), sd(speedup))
+DateCount19ex <- count(Wks2019[which(Wks2019$hour
+                      != 0:5),], dateOnly)
+DateSp19downEx <- summarise(group_by(Wks2019[which(Wks2019$hour
+                      != 0:5),], dateOnly), mean(speeddown))
+DateSp19upEx <- summarise(group_by(Wks2019[which(Wks2019$hour
+                      != 0:5),], dateOnly), mean(speedup))
+DateSDSp19downEx <- summarise(group_by(Wks2019[which(Wks2019$hour
+                      != 0:5),], dateOnly), sd(speeddown))
+DateSDSp19upEx <- summarise(group_by(Wks2019[which(Wks2019$hour
+                      != 0:5),], dateOnly), sd(speedup))
+DateStats19 <- cbind(DateCount19ex, DateCount19[,2], 
+                   DateSp19downEx[,2], DateSp19down[,2],
+                   DateSp19upEx[,2], DateSp19up[,2],
+                   DateSDSp19downEx[,2], DateSDSp19down[,2],
+                   DateSDSp19upEx[,2], DateSDSp19up[,2])
+names(DateStats19) <- c("dateOnly", "test19ex", "tests19",
+                      "MeanSpDownDateEx", "MeanSpDownDate", 
+                      "MeanSpUpDateEx", "MeanSpUpDate", 
+                      "SDSpDownDateEx", "SDSpDownDate", "SDSpUpDateEx", 
+                      "SDSpUpDate")
+Wks2019 <- left_join(Wks2019, DateStats19)
 DateCount20 <- count(Wks2020, dateOnly)
-names(DateCount20) <- c("Date20", "tests20")
-HrCount20 <- count(Wks2020, date, hour)
-HrSpeed20 <- summarise(group_by(Wks2020, date, hour), mean(speeddown))
-HrStats20 <- left_join(HrCount20, HrSpeed20)
-names(HrStats20) <- c("Date20", "Hour20", "test20", "meanSp20")
-DateCount <- cbind(DateCount19, DateCount20)
-HrStats <- cbind(HrStats19, HrStats20)  # I GET AN ERROR HERE
-write.csv(DateCount, "DailyTestFreq.csv")
-write.csv(HrStats, "HrTestFreqMean.csv")
-Plot2019 <- ggplot(DateCount19, aes(Date19, tests19))
+DateSp20down <- summarise(group_by(Wks2020, dateOnly), mean(speeddown))
+DateSp20up <- summarise(group_by(Wks2020, dateOnly), mean(speedup))
+DateSDSp20down <- summarise(group_by(Wks2020, dateOnly), sd(speeddown))
+DateSDSp20up <- summarise(group_by(Wks2020, dateOnly), sd(speedup))
+DateCount20ex <- count(Wks2020[which(Wks2020$hour != 0:5),], dateOnly)
+DateSp20downEx <- summarise(group_by(Wks2020[which(Wks2020$hour
+                              != 0:5),], dateOnly), mean(speeddown))
+DateSp20upEx <- summarise(group_by(Wks2020[which(Wks2020$hour
+                                   != 0:5),], dateOnly), mean(speedup))
+DateSDSp20downEx <- summarise(group_by(Wks2020[which(Wks2020$hour
+                              != 0:5),], dateOnly), sd(speeddown))
+DateSDSp20upEx <- summarise(group_by(Wks2020[which(Wks2020$hour
+                            != 0:5),], dateOnly), sd(speedup))
+DateStats20 <- cbind(DateCount20ex, DateCount20[,2], 
+                     DateSp20downEx[,2], DateSp20down[,2],
+                     DateSp20upEx[,2], DateSp20up[,2],
+                     DateSDSp20downEx[,2], DateSDSp20down[,2],
+                     DateSDSp20upEx[,2], DateSDSp20up[,2])
+names(DateStats20) <- c("dateOnly", "test20ex", "tests20",
+                        "MeanSpDownDateEx", "MeanSpDownDate", 
+                        "MeanSpUpDateEx", "MeanSpUpDate", 
+                        "SDSpDownDateEx", "SDSpDownDate", "SDSpUpDateEx", 
+                        "SDSpUpDate")
+Wks2020 <- left_join(Wks2020, DateStats20)
+#frequency plot
+Plot2019 <- ggplot(Wks2019, aes(dateOnly, tests19))
 Plot2019 + geom_line()
-Plot2020 <- ggplot(DateCount20, aes(Date20, tests20))
+Plot2020 <- ggplot(Wks2020, aes(dateOnly, tests20))
 Plot2020 + geom_line()
+HrDateCount19 <- count(Wks2019, dateOnly, hour)
+HrDateCount20 <- count(Wks2020, dateOnly, hour)
+names(HrDateCount19)[3] <- "testFreq19"
+names(HrDateCount20)[3] <- "testFreq20"
+Wks2019 <- left_join(Wks2019, HrDateCount19)
+Wks2020 <- left_join(Wks2020, HrDateCount20)
+TimeVar2019n <- timeVariation(Wks2019, pollutant = "testFreq19", 
+                    statistic = "mean")
+TimeVar2020n <- timeVariation(Wks2020, pollutant = "testFreq20", 
+                              statistic = "mean")
+#save files
+write.csv(Wks2019, "Wks2019.csv")
+write.csv(Wks2020, "Wks2020.csv")
+
 ```
@@
 ```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
-#convert to spatial object
-coords_bb <- cbind(bb2020$lon, bb2020$lat)
-bbNUTS3 <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020))
-proj4string(bbNUTS3) <- CRS("+init=epsg:4326") #define projection
 
-#get NUTS3 from up a level and read in (145 NUTS3)
-NUTS3 <- readOGR("../NUTS_Level_3_January_2015_Super_Generalised_Clipped_Boundaries_in_England_and_Wales.shp") 
-NUTS3 <- spTransform(NUTS3, CRS("+init=epsg:4326"))                                
-NUTS3@data$nuts315nm <- as.character(NUTS3@data$nuts315nm)
-
-#spatial join to NUTS3
-bbNUTS3sp <- over(bbNUTS3, NUTS3[, "nuts315nm"]) #not a spatial object
-bbNUTS3$nuts315nm <- bbNUTS3sp$nuts315nm
-
-#create dataframe object to analyse (remove lat and lon?)
-bb2020sp <- bbNUTS3@data
-#% obs NA for NUTS3?
-
-#summarise by date and geography
-NUTS3datecount <- count(bb2020sp, nuts315nm, date)
-NUTS3dateAvg <- summarise(group_by(NUTS3datecount, nuts315nm), mean(n))
-
-#map NUTS3 mean test frequency per date
-NUTS3@data <- left_join(NUTS3@data, NUTS3dateAvg, by = "nuts315nm")
-var <- NUTS3@data[ ,'mean(n)']
-breaks <- classIntervals(var, n = 5, style = "quantile")
-my_colours <- brewer.pal(5, "Blues")
-plot(NUTS3, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE,
-     border = rgb(0.8, 0.8, 0.8, 0))
-legend(x = "top", legend = leglabs(breaks$brks), fill = my_colours, bty = "n", cex = 0.5)
 ```

### 2020-06-09 src/bbData2020.R

@@ -18,7 +18,7 @@ summary(bb2020) #2030205 obs
 #check fastest upload and download speeds commercially available from Virgin Media (Gigaclear offers even faster!)
 bb2020 <- bb2020[bb2020$speeddown>512,] 
 bb2020 <- bb2020[bb2020$speeddown<362000,]
-bb2020 <- bb2020[bb2020$speedup<21000,] #1839333
+bb2020 <- bb2020[bb2020$speedup<21000,] #1839332 obs
 
 #remove below seconds, create columns for date and hour
 bb2020$datetext <- strtrim(x = bb2020$datetext, width = 19) 
@@ -35,8 +35,23 @@ bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
 
 #add library to explore data temporally
 library(openair)
-#add column for day of the week
+#create dataframe with number of tests, mean and standard deviation by month for adding to dataframe later
+MonthFreq <- timeAverage(bb2020, avg.time = "month", statistic = "frequency")
+MonthMean <- timeAverage(bb2020, avg.time = "month", statistic = "mean")
+MonthSD <- timeAverage(bb2020, avg.time = "month", statistic = "sd")
+MonthStats <- cbind(MonthFreq[,1:2], MonthMean[,c(3:4)], 
+                    MonthSD[,c(3:4)])
+names(MonthStats) <- c("month", "monthlyTests", "monthlyMeanSpDown", 
+                       "monthlyMeanSpUp", "monthlySDSpDown", 
+                       "monthlySDSpUp")
+MonthStats$month <- month(MonthStats$month, label = TRUE, abbr = TRUE)
+#add column for month and day of the week to create stats below
+bb2020$month <- month(bb2020$date, label = TRUE, abbr = TRUE)
 bb2020$weekday <- wday(bb2020$date, label = TRUE)
+#divide MonthStats for future join with separate 2019 and 2020 dataframes
+MonthStats19 <- MonthStats[1:6,]
+MonthStats20 <- MonthStats[13:17,]
+
 #create 2 dataframes for 2 matching periods
 Wks2019 <- selectByDate(bb2020, start = "2019-01-21", end = "2019-06-02")
 names(Wks2019)[1:2] <- c("dateOnly", "date")
@@ -60,87 +75,191 @@ timePlot(Wks2019, pollutant = "speedup", avg.time = "day",
[…]

### 2020-06-10 src/help_timeseries.R
lit + new help script

@@ -1,0 +1,100 @@
+library(workflows)
+library(parsnip)
+library(recipes)
+library(yardstick)
+library(glmnet)
+library(tidyverse)
+library(tidyquant)
+library(timetk) # Use >= 0.1.3, remotes::install_github("business-science/timetk")
+
+
+# Add time series signature
+bb2020test <- bb2020 %>% 
+  select(datetext,speedup)
+    
+# Add time series signature
+recipe_spec_timeseries <- recipe(speedup ~ ., data = bb2020test) %>%
+  step_timeseries_signature(datetext) 
+
+bb2020test_ <- bake(prep(recipe_spec_timeseries), new_data = bb2020test)
+options(tibble.width = Inf) 
+head(bb2020test_)
+
+bb2020test_final <- recipe_spec_timeseries %>%
+  step_rm(date) %>%
+  step_rm(contains("iso"), 
+          contains("second"), contains("minute"), contains("hour"),
+          contains("am.pm"), contains("xts")) %>%
+  step_normalize(contains("index.num"), date_year) %>%
+  step_interact(~ date_month.lbl * date_day) %>%
+  step_interact(~ date_month.lbl * date_mweek) %>%
+  step_interact(~ date_month.lbl * date_wday.lbl * date_yday) %>%
+  step_dummy(contains("lbl"), one_hot = TRUE) 
+
+bake(prep(bb2020test_final), new_data = bb2020test_)
+
+bake(prep(recipe_spec_final), new_data = train_tbl)
+
+##########
+# install.packages("devtools")
+devtools::install_github("robjhyndman/anomalous-acm")
+library(anomalousACM)
+
+z <- ts(matrix(rnorm(3000),ncol=100),freq=4)
+y <- tsmeasures(z)
+biplot.features(y)
+anomaly(y)
+
+y_ <- tsmeasures(bb2020test$speedup)
+
+
+########
+# tsfeatures is the cran and advanced version of anomalousACM
+library(tsfeatures)
+ft <- tsfeatures(bb2020test$speedup)
+head(ft)
+dim(ft)
+entropy(bb2020test$speedup)
+stability(bb2020test$speedup)
+lumpiness(bb2020test$speedup)
+crossing_points(bb2020test$speedup)
+length(bb2020test$speedup)
+flat_spots(bb2020test$speedup)
+hurst(bb2020test$speedup)
+acf_features(bb2020test$speedup)
+
+# reproducing Hyndman, Wang & Laptev (ICDM 2015)
[…]

### 2020-06-10 src/bbData2020.Rmd

@@
 
-```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
+```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE, fig.height=15, fig.width=15}
+#OLD# #add column for day of the week
+#OLD# bb2020$weekday <- wday(bb2020$date, label = TRUE)
+
 #create 2 dataframes for 2 matching periods
@@
 
-```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
+```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE, fig.height=15, fig.width=15}
 #add stats for each month: test frequency, mean and standard deviation of speeddown and speedup
@@
 Plot2019 <- ggplot(Wks2019, aes(dateOnly, tests19))
+
+#OLD# timePlot(Wks2019, pollutant = "test19", avg.time = "hour", smooth = TRUE, name.pol = "hourlyTests 2019")
+#OLD# timePlot(Wks2020, pollutant = "test20", avg.time = "hour", smooth = TRUE, name.pol = "hourlyTests 2020")
+#OLD# TimeVar2019n <- timeVariation(Wks2019, pollutant = "test19", statistic = "mean", name.pol = "hourlytests 2019")
+#OLD# TimeVar2020n <- timeVariation(Wks2020, pollutant = "test20", statistic = "mean", name.pol = "hourlytests 2020")
+#OLD# Plot2019 <- ggplot(DateCount19, aes(Date19, tests19))
+
 Plot2019 + geom_line()
@@
 
-# Spatial analysis
+## Spatial analysis
 
-```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
+<!-- # Ctrl + Shift + c to remove all the below comments -->
+<!-- ```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE, fig.height=15, fig.width=10} -->
+<!-- #convert to spatial object -->
+<!-- coords_bb <- cbind(bb2020$lon, bb2020$lat) -->
+<!-- bbNUTS3 <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020)) -->
+<!-- proj4string(bbNUTS3) <- CRS("+init=epsg:4326") #define projection -->
 
-```
+<!-- #get NUTS3 from up a level and read in (145 NUTS3) -->
+<!-- NUTS3 <- readOGR("../NUTS_Level_3_January_2015_Super_Generalised_Clipped_Boundaries_in_England_and_Wales.shp") -->
+<!-- NUTS3 <- spTransform(NUTS3, CRS("+init=epsg:4326")) -->
+<!-- NUTS3@data$nuts315nm <- as.character(NUTS3@data$nuts315nm) -->
+
+<!-- #spatial join to NUTS3 -->
+<!-- bbNUTS3sp <- over(bbNUTS3, NUTS3[, "nuts315nm"]) #not a spatial object -->
+<!-- bbNUTS3$nuts315nm <- bbNUTS3sp$nuts315nm -->
+
+<!-- #create dataframe object to analyse (remove lat and lon?) -->
+<!-- bb2020sp <- bbNUTS3@data -->
+<!-- #% obs NA for NUTS3? -->
+
+<!-- #summarise by date and geography -->
+<!-- NUTS3datecount <- count(bb2020sp, nuts315nm, date) -->
+<!-- NUTS3dateAvg <- summarise(group_by(NUTS3datecount, nuts315nm), mean(n)) -->
+
+<!-- #map NUTS3 mean test frequency per date -->
+<!-- NUTS3@data <- left_join(NUTS3@data, NUTS3dateAvg, by = "nuts315nm") -->
+<!-- var <- NUTS3@data[ ,'mean(n)'] -->
+<!-- breaks <- classIntervals(var, n = 5, style = "quantile") -->
+<!-- my_colours <- brewer.pal(5, "Blues") -->
+<!-- plot(NUTS3, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE, -->
+<!--      border = rgb(0.8, 0.8, 0.8, 0)) -->
+<!-- legend(x = "top", legend = leglabs(breaks$brks), fill = my_colours, bty = "n", cex = 0.5) -->
+<!-- ``` -->

### 2020-06-12 src/bbData2020.Rmd
Revert "Revert "fix Rmd conflicts, fix paths to data temp outputs, knit html, remove .gitignore from temp data folder.""
This reverts commit ceca99313e95ed5b36d6eef6837fbcc9debc08f2.

@@
 #save files
-write.csv(Wks2019, "Wks2019.csv")
-write.csv(Wks2020, "Wks2020.csv")
-
+out.Wks2019 <- paste(path, "./data/temp/Wks2019.csv", sep = "")
+out.Wks2020 <- paste(path, "./data/temp/Wks2020.csv", sep = "")
+write.csv(Wks2019, out.Wks2019)
+write.csv(Wks2020, out.Wks2020)
 ```

### 2020-06-12 src/timeseries_plots.Rmd
1st attempt for ts

---
title: "Time series plots"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

```{r settings, include = FALSE}
library(tsbox)
library(ggplot2)
library(tidyverse)
library(geojsonio)
library(tsfeatures)
library(rprojroot)
library(openair)
library(reshape2)
library(lubridate)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(classInt)
library(RColorBrewer)
library(knitr)
library(rprojroot)
library(viridis)
library(hrbrthemes)

# This is the project path
path <- find_rstudio_root_file()
```

```{r include = FALSE}

# load and clean the data.
# copied from bbData2020.Rmd

#upload data file
path.data <- paste(path, "./data/raw/speedtest.csv", sep = "")

#bb2020 <- read.csv("./data/raw/speedtest.csv")
bb2020 <- read.csv(path.data)

#name columns
names(bb2020)[1:6] <- c("datetext","speeddown","speedup","provider","lat", "lon")

str(bb2020)
summary(bb2020) #2030205 obs
#conversions needed to numeric? df$col <- as.numeric(df$col) 
#remove empty rows or coerced to NA? df[complete.cases(df),] or df <- df[!is.na(df$col),]

#remove outliers based on Riddlesden and Singleton 2014 for slower end
#check fastest upload and download speeds commercially available from Virgin Media (Gigaclear offers even faster!)
bb2020 <- bb2020[bb2020$speeddown>512,] 
bb2020 <- bb2020[bb2020$speeddown<362000,]
bb2020 <- bb2020[bb2020$speedup<21000,] #1839333

#remove below seconds, create columns for date and hour
bb2020$datetext <- strtrim(x = bb2020$datetext, width = 19) 
bb2020$datetext2 <- bb2020$datetext
bb2020$datetext2 <- strtrim(x = bb2020$datetext2, width = 13)
bb2020$datetext2 <- colsplit(bb2020$datetext2, pattern = " ", c("date", "hour"))
bb2020$hour <- bb2020$datetext2$hour
bb2020$date <- bb2020$datetext2$date
#delete datetext2 column(s)
bb2020 <- bb2020[ ,c(1:6,8:9)]
#convert to date form datetext and date columns
bb2020$datetext <- as.POSIXct(strptime(bb2020$datetext, format = "%Y-%m-%d %H:%M:%S", "GMT")) 
bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
#create dataframe with number of tests, mean and standard deviation by month for adding to dataframe later
MonthFreq <- timeAverage(bb2020, avg.time = "month", statistic = "frequency")
MonthMean <- timeAverage(bb2020, avg.time = "month", statistic = "mean")
MonthSD <- timeAverage(bb2020, avg.time = "month", statistic = "sd")
MonthStats <- cbind(MonthFreq[,1:2], MonthMean[,c(3:4)], 
                    MonthSD[,c(3:4)])
names(MonthStats) <- c("month", "monthlyTests", "monthlyMeanSpDown", 
                       "monthlyMeanSpUp", "monthlySDSpDown", 
                       "monthlySDSpUp")
MonthStats$month <- month(MonthStats$month, label = TRUE, abbr = TRUE)
#add column for month and day of the week to create stats below
bb2020$month <- month(bb2020$date, label = TRUE, abbr = TRUE)
bb2020$weekday <- wday(bb2020$date, label = TRUE)
#divide MonthStats for future join with separate 2019 and 2020 dataframes
MonthStats19 <- MonthStats[1:6,]
MonthStats20 <- MonthStats[13:17,]
```

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}

# Structured TS

# convert to spatial object
coords_bb <- cbind(bb2020$lon, bb2020$lat)
bb.la <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020))
proj4string(bb.la) <- CRS("+init=epsg:4326") #define projection

# get LA 
# (i) directly from the web
#la <- readOGR("http://geoportal1-ons.opendata.arcgis.com/datasets/b6d2e15801de45328b760a4f55d74318_0.geojson?outSR={%22latestWkid%22:3857,%22wkid%22:102100}")#, layer="OGRGeoJSON")
# or, (ii) from the json I saved locally
path.json <- paste(path, "./data/raw/Local_Authority_Districts_(April_2019)_Boundaries_UK_BFE.geojson", sep = "")
la <- readOGR(path.json)#, layer="OGRGeoJSON")
# UK BGC la <- readOGR("https://opendata.arcgis.com/datasets/0e07a8196454415eab18c40a54dfbbef_0.geojson")
# UK BFC la <- readOGR("https://opendata.arcgis.com/datasets/1d78d47c87df4212b79fe2323aae8e08_0.geojson")

# source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe

# spatial transformations
la <- spTransform(la, CRS("+init=epsg:4326"))
la@data$LAD19NM <- as.character(la@data$LAD19NM)

# spatial join to LAD
bb.la.sp <- over(bb.la, la[, "LAD19NM"]) #not a spatial object
bb.la$LAD19NM <- bb.la.sp$LAD19NM

# create dataframe object to analyse (remove lat and lon?)
bb2020sp <- bb.la@data

# drop NAs
bb2020sp <- bb2020sp[!is.na(bb2020sp$LAD19NM),]
#13431/1839332 = .0073 NA LAD19NM, coord outside UK

# summarise by date and geography
la.date.count <- count(bb2020sp, LAD19NM, date)
la.dateAvg <- summarise(group_by(la.date.count, LAD19NM), mean(n))
range(la.dateAvg$`mean(n)`)

# map LA mean test frequency per date
la@data <- left_join(la@data, la.dateAvg, by = "LAD19NM")
var <- la@data[ ,'mean(n)']
breaks <- classIntervals(var, n = 5, style = "quantile")
my_colours <- brewer.pal(5, "Blues")
plot(la, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE,
                 border = rgb(0.8, 0.8, 0.8, 0))
# drop LA
rm(la)
```

## Sub-daily timeseries

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=500, fig.width=12}

# create sub daily blocks
bb2020sp$block[bb2020sp$hour>=0 & bb2020sp$hour<8] <- 1
bb2020sp$block[bb2020sp$hour>=8 & bb2020sp$hour<16] <- 2
bb2020sp$block[bb2020sp$hour>=16 & bb2020sp$hour<24] <- 3

#bb2020sp$date.block <- paste(bb2020sp$date, bb2020sp$date.block, sep = "_")
# I use block as hour for the ts_wide()
bb2020sp$date.block = ymd_h(paste(bb2020sp$date, bb2020sp$block))
head(bb2020sp)

ts <- bb2020sp %>%
  group_by(date.block,LAD19NM) %>%
  summarise(mean.down = mean(speeddown))#, mean.down = mean(speeddown), n.tests = n())
sapply(ts, function(x) sum(is.na(x)))

#hist(ts$n.tests)
#summary(ts$n.tests)
# I am keeping all the obs. despite the n. of tests for now

# spagetti plot
ggplot(ts[ts$LAD19NM=="City of London",], 
       aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = LAD19NM)) +
  geom_line() + guides(colour=FALSE) + ggtitle("City of London") + #xlab("") +
  ylab("download speed") +
  #geom_text_repel(aes(label=outlier.io), cex = 4) + #this line is from the previous version
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5))

for (i in ts[,2]){
  spaghetti <- ggplot(ts[ts$LAD19NM==i,], 
       aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = LAD19NM)) +
    geom_line() + guides(colour=FALSE) + ggtitle("City of London") + #xlab("") +
    ylab("download speed") +
    #geom_text_repel(aes(label=outlier.io), cex = 4) + #this line is from the previous version
    scale_y_continuous(labels = scales::comma) +
    scale_x_datetime(date_breaks = "1 month") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
    plot.title = element_text(hjust = 0.5))
  print(spaghetti)
}

# https://www.data-to-viz.com/caveat/spaghetti.html
tmp <- ts %>%
  mutate(LAD19NM2=LAD19NM)

tmp %>%
  ggplot( aes(x=as.POSIXct(date.block), y=mean.down)) +
    geom_line( data=tmp %>% dplyr::select(-LAD19NM), aes(group=LAD19NM2), color="grey", size=0.5, alpha=0.5) +
    geom_line( aes(color=LAD19NM), color="#69b3a2", size=1.2 )+
    scale_color_viridis(discrete = TRUE) +
    theme_ipsum() +
    theme(
      legend.position="none",
      plot.title = element_text(size=14),
      panel.grid = element_blank()
    ) +
    ggtitle("A spaghetti chart of baby names popularity") +
    facet_wrap(~LAD19NM)

ggplot(subset(ts,date.block>"2019-12-31"), aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = "green")) +
  geom_line() + guides(colour=FALSE) + ggtitle("Download speeds") + #xlab("") +
  stat_summary(fun.y=mean, geom="line", colour="grey", aes(group=1)) + 
  ylab("download speed") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  #geom_vline(xintercept = as.numeric(as.Date("2020-03-01"))) +#, linetype="dashed", color = "blue", size=1.5) +
  facet_wrap(~LAD19NM, ncol = 2)

```

### 2020-06-12 src/help_timeseries.R

@@ -6,6 +6,8 @@ library(glmnet)
 library(tidyverse)
 library(tidyquant)
 library(timetk) # Use >= 0.1.3, remotes::install_github("business-science/timetk")
+library(geojsonio)
+
 
 
 # Add time series signature
@@ -83,18 +85,131 @@ hwl <- bind_cols(
          crossing_points, max_kl_shift, time_kl_shift)
 head(hwl)
 
+
+# Structured TS ----
+
+# convert to spatial object
+coords_bb <- cbind(bb2020$lon, bb2020$lat)
+bb.la <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020))
+proj4string(bb.la) <- CRS("+init=epsg:4326") #define projection
+
+# get LA 
+la <- readOGR("http://geoportal1-ons.opendata.arcgis.com/datasets/b6d2e15801de45328b760a4f55d74318_0.geojson?outSR={%22latestWkid%22:3857,%22wkid%22:102100}")#, layer="OGRGeoJSON")
+# from https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe
+la <- spTransform(la, CRS("+init=epsg:4326"))
+la@data$LAD19NM <- as.character(la@data$LAD19NM)
+
+# spatial join to LAD
+bb.la.sp <- over(bb.la, la[, "LAD19NM"]) #not a spatial object
+bb.la$LAD19NM <- bb.la.sp$LAD19NM
+
+# create dataframe object to analyse (remove lat and lon?)
+bb2020sp <- bb.la@data
+
+# drop NAs
+bb2020sp <- bb2020sp[!is.na(bb2020sp$LAD19NM),]
+#13431/1839332 = .0073 NA LAD19NM, coord outside UK
+
+# summarise by date and geography
+la.date.count <- count(bb2020sp, LAD19NM, date)
+la.dateAvg <- summarise(group_by(la.date.count, LAD19NM), mean(n))
+range(la.dateAvg$`mean(n)`)
+
+# map LA mean test frequency per date
+la@data <- left_join(la@data, la.dateAvg, by = "LAD19NM")
+var <- la@data[ ,'mean(n)']
+breaks <- classIntervals(var, n = 5, style = "quantile")
+my_colours <- brewer.pal(5, "Blues")
[…]

### 2020-06-18 src/bbData2020.Rmd
time series clustering time series plots literature YAMLchange for output dir

@@
     toc_float: true
+knit: (function(inputFile, encoding) {
+    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../output")
+  })
 ---

### 2020-06-18 src/help_timeseries.R

@@ -203,6 +203,57 @@ hwl_pca$x %>%
   geom_point()
 
 
+
+# on our data using tsfeatures
+hwl.bb <- bind_cols(
+  tsfeatures(ts.wide[,-1],
+             c("acf_features","entropy","lumpiness",
+               "flat_spots","crossing_points")),
+  tsfeatures(ts.wide[,-1],"stl_features", s.window='periodic', robust=TRUE),
+  tsfeatures(ts.wide[,-1], "max_kl_shift", width=48),
+  tsfeatures(ts.wide[,-1],
+             c("mean","var"), scale=FALSE, na.rm=TRUE),
+  tsfeatures(ts.wide[,-1],
+             c("max_level_shift","max_var_shift"), trim=TRUE)) %>%
+  select(mean, var, x_acf1, trend, linearity, curvature,
+         #     seasonal_strength, peak, trough,
+         entropy, lumpiness, spike, max_level_shift, max_var_shift, flat_spots,
+         crossing_points, max_kl_shift, time_kl_shift)
+head(hwl.bb)
+glimpse(hwl.bb)
+
+# remove nperiods and seasonal_period
+#hwl.bb <- hwl.bb[,-(11:12)]
+
+hwl.bb <- hwl.bb[,-(14:15)]
+
+library(ggplot2)
+hwl_pca <- hwl.bb %>%
+  na.omit() %>%
+  prcomp(scale=T)
+hwl_pca$x %>%
+  as_tibble() %>%
+  ggplot(aes(x=PC1, y=PC2)) +
+  geom_point()
+
+features <- tsfeatures(ts.wide[,-1], na.action = na.omit) # -1 to drop ts.wide$date
+# bring LADs
+features$lads <- names(ts.wide[-1])
+head(features)
+dim(features)
+glimpse(features)
+
+range(features$entropy, na.rm = T)
+entropy1 <- features[features$entropy==1,]
+dim(entropy1)
+unique(entropy1$lads)
+unique(features$lads)
+
+entropy(ts.wide[,-1])
+
+
+
+
 library(babynames)
 # Load dataset from github
 data <- babynames %>% 
@@ -213,3 +264,5 @@ head(data)
 tmp <- data %>%
   mutate(name2=name)
 head(tmp)
+
+

### 2020-06-18 src/test.R

@@ -1,21 +1,0 @@
-# This is to test the path.
-# The wd() is the project path,
-# so we can easily read from the ./data 
-# and save to ./outputs.
-# I renamed the data file to speedtest.csv
-# Paste the file on this location localy and rename it.
-# The .gitignore files prevent the sync of the folder contents.
-# So, the below .csv will not sync from you laptop to git.
-# We don't want that because (i) it's too big for gut and
-# (ii) we don't want to have these data public.
-# I have similar .gitignore files to other folders
-# because git cannot sync empty folders.
-# Once we have outputs we can delete these files in
-# order to sync their content with git
-
-library(data.table)
-
-getwd()
-
-test <- fread("./data/raw/speedtest.csv")
-head(test)

### 2020-06-19 src/bbData2020.Rmd
allow merge

--- 2020-06-10 src/bbData2020.Rmd
@@
 
-```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE, fig.height=15, fig.width=15}
-#OLD# #add column for day of the week
-#OLD# bb2020$weekday <- wday(bb2020$date, label = TRUE)
-
+```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
 #create 2 dataframes for 2 matching periods
@@
 
-```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE, fig.height=15, fig.width=15}
+```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
 #add stats for each month: test frequency, mean and standard deviation of speeddown and speedup
@@
                       "MeanSpUpDateEx", "MeanSpUpDate", 
-                      "SDSpDownDateEx", "SDSpDownDate", "SDSpUpDateEx", 
-                      "SDSpUpDate")
+                      "SDSpDownDateEx", "SDSpDownDate", "SDSpUpDateEx",                       "SDSpUpDate")
 Wks2019 <- left_join(Wks2019, DateStats19)
@@
 Plot2019 <- ggplot(Wks2019, aes(dateOnly, tests19))
-
-#OLD# timePlot(Wks2019, pollutant = "test19", avg.time = "hour", smooth = TRUE, name.pol = "hourlyTests 2019")
-#OLD# timePlot(Wks2020, pollutant = "test20", avg.time = "hour", smooth = TRUE, name.pol = "hourlyTests 2020")
-#OLD# TimeVar2019n <- timeVariation(Wks2019, pollutant = "test19", statistic = "mean", name.pol = "hourlytests 2019")
-#OLD# TimeVar2020n <- timeVariation(Wks2020, pollutant = "test20", statistic = "mean", name.pol = "hourlytests 2020")
-#OLD# Plot2019 <- ggplot(DateCount19, aes(Date19, tests19))
-
 Plot2019 + geom_line()
@@
 
-## Spatial analysis
+# Spatial analysis
 
-<!-- # Ctrl + Shift + c to remove all the below comments -->
-<!-- ```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE, fig.height=15, fig.width=10} -->
-<!-- #convert to spatial object -->
-<!-- coords_bb <- cbind(bb2020$lon, bb2020$lat) -->
-<!-- bbNUTS3 <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020)) -->
-<!-- proj4string(bbNUTS3) <- CRS("+init=epsg:4326") #define projection -->
+```{r echo=FALSE, results= 'asis', message=FALSE, warning=TRUE}
 
-<!-- #get NUTS3 from up a level and read in (145 NUTS3) -->
-<!-- NUTS3 <- readOGR("../NUTS_Level_3_January_2015_Super_Generalised_Clipped_Boundaries_in_England_and_Wales.shp") -->
-<!-- NUTS3 <- spTransform(NUTS3, CRS("+init=epsg:4326")) -->
-<!-- NUTS3@data$nuts315nm <- as.character(NUTS3@data$nuts315nm) -->
-
-<!-- #spatial join to NUTS3 -->
-<!-- bbNUTS3sp <- over(bbNUTS3, NUTS3[, "nuts315nm"]) #not a spatial object -->
-<!-- bbNUTS3$nuts315nm <- bbNUTS3sp$nuts315nm -->
-
-<!-- #create dataframe object to analyse (remove lat and lon?) -->
-<!-- bb2020sp <- bbNUTS3@data -->
-<!-- #% obs NA for NUTS3? -->
-
-<!-- #summarise by date and geography -->
-<!-- NUTS3datecount <- count(bb2020sp, nuts315nm, date) -->
-<!-- NUTS3dateAvg <- summarise(group_by(NUTS3datecount, nuts315nm), mean(n)) -->
-
-<!-- #map NUTS3 mean test frequency per date -->
-<!-- NUTS3@data <- left_join(NUTS3@data, NUTS3dateAvg, by = "nuts315nm") -->
-<!-- var <- NUTS3@data[ ,'mean(n)'] -->
-<!-- breaks <- classIntervals(var, n = 5, style = "quantile") -->
-<!-- my_colours <- brewer.pal(5, "Blues") -->
-<!-- plot(NUTS3, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE, -->
-<!--      border = rgb(0.8, 0.8, 0.8, 0)) -->
-<!-- legend(x = "top", legend = leglabs(breaks$brks), fill = my_colours, bty = "n", cex = 0.5) -->
-<!-- ``` -->
+```

### 2020-06-19 src/bbData2020.R

@@ -74,11 +74,14 @@ timePlot(Wks2019, pollutant = "speedup", avg.time = "day",
          smooth = TRUE, name.pol = "speedup 2019")
 timePlot(Wks2020, pollutant = "speedup", avg.time = "day", 
          smooth = TRUE, name.pol = "speedup 2020")
-
-#add temporal stats, first all months
+#add stats for each month: 
+#test frequency, mean and standard deviation of speeddown and speedup
+#note that Jan 2019, Jun 2019 and Jan 2020 inc WHOLE month in stats 
+#weekends and nights included
 Wks2019 <- left_join(Wks2019, MonthStats19)
 Wks2020 <- left_join(Wks2020, MonthStats20)
-#add temporal stats for days of week, all hours and exc 00.00-06.00
+#add stats for days of week, one set of stats with all hours 
+#one set exc 00.00-06.00
 WkdyCount19 <- count(Wks2019, weekday)
 WkdyMeanSp19down <- summarise(group_by(Wks2019, weekday), mean(speeddown))
 WkdyMeanSp19up <- summarise(group_by(Wks2019, weekday), mean(speedup))
@@ -129,7 +132,9 @@ names(WkdyStats20) <- c("weekday", "testsWkdyEx", "testsWkdy",
                         "SDSpDownWkdyEx", "SDSpDownWkdy", "SDSpUpWkdyEx",
                         "SDSpUpWkdy")
 Wks2020 <- left_join(Wks2020, WkdyStats20)
-#add temporal stats for hours of the day, all days and exc Sat-Sun
+#add stats for hours of the day, 
+#one set with those hours from all days included
+#one set exc those hours on Sat-Sun
 HrCount19 <- count(Wks2019, hour)
 HrMeanSp19down <- summarise(group_by(Wks2019, hour), mean(speeddown))
 HrMeanSp19up <- summarise(group_by(Wks2019, hour), mean(speedup))

### 2020-06-19 src/bbData2020.Rmd

--- 2020-06-12 src/bbData2020.Rmd
@@
 #save files
-out.Wks2019 <- paste(path, "./data/temp/Wks2019.csv", sep = "")
-out.Wks2020 <- paste(path, "./data/temp/Wks2020.csv", sep = "")
-write.csv(Wks2019, out.Wks2019)
-write.csv(Wks2020, out.Wks2020)
+write.csv(Wks2019, "Wks2019.csv")
+write.csv(Wks2020, "Wks2020.csv")
+
 ```

### 2020-06-19 src/ts_clusters.Rmd

--- 2020-06-12 src/timeseries_plots.Rmd
@@
 ---
-title: "Time series plots"
+title: "Time series clusters"
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
@@
 library(hrbrthemes)
+library(imputeTS)
+library(tsbox)
+library(dtwclust)
 
@@
 
-```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+## Structured time series
 
-# Structured TS
+```{r eval=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
 
@@
                  border = rgb(0.8, 0.8, 0.8, 0))
-# drop LA
-rm(la)
+
 ```
 
-## Sub-daily timeseries
+## Keep working hours and days
 
-```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=500, fig.width=12}
+The time series are based on mean upload and download speeds as well as the total number of tests for working days and during working hours
 
-# create sub daily blocks
-bb2020sp$block[bb2020sp$hour>=0 & bb2020sp$hour<8] <- 1
-bb2020sp$block[bb2020sp$hour>=8 & bb2020sp$hour<16] <- 2
-bb2020sp$block[bb2020sp$hour>=16 & bb2020sp$hour<24] <- 3
+```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
 
-#bb2020sp$date.block <- paste(bb2020sp$date, bb2020sp$date.block, sep = "_")
-# I use block as hour for the ts_wide()
-bb2020sp$date.block = ymd_h(paste(bb2020sp$date, bb2020sp$block))
+bb2020sp <- bb2020sp %>% arrange(datetext)
+
+# keep working hours
+bb2020sp <- bb2020sp[bb2020sp$hour>=9 & bb2020sp$hour<17,]
+
+# drop w/e
+bb2020sp <- bb2020sp %>%
+  filter(bb2020sp$weekday!="Sat" & bb2020sp$weekday!="Sun")
+
 head(bb2020sp)
+```
 
-ts <- bb2020sp %>%
-  group_by(date.block,LAD19NM) %>%
+## Time series clusters
+
+I create clusters based on upload, download and number of tests time series for tests working hours and days from 16/2/2020 onwards.
+
+More work needs to be done to select the optimal *k* as well as the clustering methods.
+The `dtwclust` package comes with quite some readings, saved in the `.\literature\`.
+
+Also, we need to think about upload / download / differnces / period + **imputation**.
+
+```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# download
+ts.down <- bb2020sp %>%
+  group_by(date,LAD19NM) %>%
   summarise(mean.down = mean(speeddown))#, mean.down = mean(speeddown), n.tests = n())
-sapply(ts, function(x) sum(is.na(x)))
 
-#hist(ts$n.tests)
-#summary(ts$n.tests)
-# I am keeping all the obs. despite the n. of tests for now
+# upload
+ts.up <- bb2020sp %>%
+  group_by(date,LAD19NM) %>%
+  summarise(mean.up = mean(speedup))
 
-# spagetti plot
-ggplot(ts[ts$LAD19NM=="City of London",], 
-       aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = LAD19NM)) +
-  geom_line() + guides(colour=FALSE) + ggtitle("City of London") + #xlab("") +
-  ylab("download speed") +
-  #geom_text_repel(aes(label=outlier.io), cex = 4) + #this line is from the previous version
-  scale_y_continuous(labels = scales::comma) +
-  scale_x_datetime(date_breaks = "1 month") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
-        plot.title = element_text(hjust = 0.5))
+# tests
+ts.tests <- bb2020sp %>%
+  group_by(date,LAD19NM) %>%
+  summarise(n.tests = n())
+```
 
-for (i in ts[,2]){
-  spaghetti <- ggplot(ts[ts$LAD19NM==i,], 
-       aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = LAD19NM)) +
-    geom_line() + guides(colour=FALSE) + ggtitle("City of London") + #xlab("") +
-    ylab("download speed") +
-    #geom_text_repel(aes(label=outlier.io), cex = 4) + #this line is from the previous version
-    scale_y_continuous(labels = scales::comma) +
-    scale_x_datetime(date_breaks = "1 month") +
-    theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
-    plot.title = element_text(hjust = 0.5))
-  print(spaghetti)
-}
+### Upload speed
 
-# https://www.data-to-viz.com/caveat/spaghetti.html
-tmp <- ts %>%
-  mutate(LAD19NM2=LAD19NM)
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
 
-tmp %>%
-  ggplot( aes(x=as.POSIXct(date.block), y=mean.down)) +
-    geom_line( data=tmp %>% dplyr::select(-LAD19NM), aes(group=LAD19NM2), color="grey", size=0.5, alpha=0.5) +
-    geom_line( aes(color=LAD19NM), color="#69b3a2", size=1.2 )+
-    scale_color_viridis(discrete = TRUE) +
-    theme_ipsum() +
-    theme(
-      legend.position="none",
-      plot.title = element_text(size=14),
-      panel.grid = element_blank()
-    ) +
-    ggtitle("A spaghetti chart of baby names popularity") +
-    facet_wrap(~LAD19NM)
+# upload working hours and days
 
-ggplot(subset(ts,date.block>"2019-12-31"), aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = "green")) +
-  geom_line() + guides(colour=FALSE) + ggtitle("Download speeds") + #xlab("") +
-  stat_summary(fun.y=mean, geom="line", colour="grey", aes(group=1)) + 
-  ylab("download speed") +
-  scale_y_continuous(labels = scales::comma) +
-  scale_x_datetime(date_breaks = "1 month") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
-        plot.title = element_text(hjust = 0.5)) +
-  #geom_vline(xintercept = as.numeric(as.Date("2020-03-01"))) +#, linetype="dashed", color = "blue", size=1.5) +
-  facet_wrap(~LAD19NM, ncol = 2)
+ts.wide <- ts.up %>%
+  select(date, LAD19NM, mean.up) %>%
+  filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide <- ts_wide(ts.wide)
+
+# impute with means
+library(imputeTS)
+ts.wide <- na_mean(ts.wide)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# extract names
+ts.wide.names <- names(ts.wide[,-1])
+ts.wide <- ts.wide[,-1]
+
+# standardise
+ts.wide <- zscore(ts.wide)
+
+# convert to list for tsclust
+ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
+
+# 5 - 20 cluster sollutions
+pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
+                type="partitional",
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+# decide k, see RJ-2019-023 for max/min
+names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+sapply(pc_k, cvi, type = "internal")
+
+# I need to work on the selection
+
+pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+plot(pc_k)
+
+cluster_up <- data.frame(LAD=ts.wide.names,
+                           cluster=pc_k@cluster)
+```
+
+### Download speed
+
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# downlad working hours and days
+
+ts.wide <- ts.down %>%
+  select(date, LAD19NM, mean.down) %>%
+  filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide <- ts_wide(ts.wide)
+
+# impute with means
+library(imputeTS)
+ts.wide <- na_mean(ts.wide)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# extract names
+ts.wide.names <- names(ts.wide[,-1])
+ts.wide <- ts.wide[,-1]
+
+# standardise
+ts.wide <- zscore(ts.wide)
+
+# convert to list for tsclust
+ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
+
+# 5 - 20 cluster sollutions
+pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
+                type="partitional",
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+# decide k, see RJ-2019-023 for max/min
+names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+sapply(pc_k, cvi, type = "internal")
+
+# I need to work on the selection
+
+pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+plot(pc_k)
+
+cluster_down <- data.frame(LAD=ts.wide.names,
+                         cluster=pc_k@cluster)
+```
+
+### N. of speeds
+
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# tests working hours and days
+
+ts.wide <- ts.tests %>%
+  select(date, LAD19NM, n.tests) %>%
+  filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide <- ts_wide(ts.wide)
+
+# impute with means
+library(imputeTS)
+ts.wide <- na_mean(ts.wide)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# extract names
+ts.wide.names <- names(ts.wide[,-1])
+ts.wide <- ts.wide[,-1]
+
+# standardise
+ts.wide <- zscore(ts.wide)
+
+# convert to list for tsclust
+ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
+
+# 5 - 20 cluster sollutions
+pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
+                type="partitional",
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+# decide k, see RJ-2019-023 for max/min
+names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+sapply(pc_k, cvi, type = "internal")
+
+# I need to work on the selection
+
+pc_k <- tsclust(ts.wide, k = 15, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+plot(pc_k)
+
+cluster_tests <- data.frame(LAD=ts.wide.names,
+                           cluster=pc_k@cluster)
+
+clusters <- merge(cluster_up, cluster_down, by = "LAD")
+names(clusters)[2] <- "cluster.up"
+names(clusters)[3] <- "cluster.down"
+clusters <- merge(clusters, cluster_tests, by = "LAD")
+names(clusters)[4] <- "cluster.tests"
+#sapply(clusters, function(x) sum(is.na(x)))
+
+clusters  %>% arrange(cluster.up) %>% kable() 
+```
+
+## Mapping the clusters
+
+```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+la@data <- merge(la@data, clusters, by.x = "LAD19NM", by.y = "LAD")
+
+var <- la@data[ ,'cluster.up']
+plot(la, col = var, main = "Upload clusters",
+     border = rgb(0.8, 0.8, 0.8, 0))
+
+var <- la@data[ ,'cluster.down']
+plot(la, col = var, main = "Download clusters",
+     border = rgb(0.8, 0.8, 0.8, 0))
+
+var <- la@data[ ,'cluster.tests']
+plot(la, col = var, main = "Tests clusters",
+     border = rgb(0.8, 0.8, 0.8, 0))
 

### 2020-06-26 Data_Spatial.Rmd
New .Rmd for subset data by date, day, hour and then point in polygon

---
title: "Data_Spatial"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output: html_document
df_print: paged
    toc: true
    toc_float: true
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../output")
  })
---

```{r settings, include = FALSE}
library(tsbox)
library(ggplot2)
library(tidyverse)
library(geojsonio)
library(tsfeatures)
library(rprojroot)
library(openair)
library(reshape2)
library(lubridate)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(classInt)
library(RColorBrewer)
library(knitr)
library(rprojroot)
library(viridis)
library(hrbrthemes)

# This is the project path
path <- find_rstudio_root_file()
```

```{r include = FALSE}

# load and clean the data.
# copied from bbData2020.Rmd

#upload data file
path.data <- paste(path, "./data/raw/speedtest.csv", sep = "")

#bb2020 <- read.csv("./data/raw/speedtest.csv")
bb2020 <- read.csv(path.data)

#name columns
names(bb2020)[1:6] <- c("datetext","speeddown","speedup","provider","lat", "lon")

str(bb2020)
summary(bb2020) #2030205 obs
#conversions needed to numeric? df$col <- as.numeric(df$col) 
#remove empty rows or coerced to NA? df[complete.cases(df),] or df <- df[!is.na(df$col),]

#remove outliers based on Riddlesden and Singleton 2014 for slower end
#check fastest upload and download speeds commercially available from Virgin Media (Gigaclear offers even faster!)
bb2020 <- bb2020[bb2020$speeddown>512,] 
bb2020 <- bb2020[bb2020$speeddown<362000,]
bb2020 <- bb2020[bb2020$speedup<21000,] #1839333

#remove below seconds, create columns for date and hour
bb2020$datetext <- strtrim(x = bb2020$datetext, width = 19) 
bb2020$datetext2 <- bb2020$datetext
bb2020$datetext2 <- strtrim(x = bb2020$datetext2, width = 13)
bb2020$datetext2 <- colsplit(bb2020$datetext2, pattern = " ", c("date", "hour"))
bb2020$hour <- bb2020$datetext2$hour
bb2020$date <- bb2020$datetext2$date
#delete datetext2 column(s)
bb2020 <- bb2020[ ,c(1:6,8:9)]
#convert to date form datetext and date columns
bb2020$datetext <- as.POSIXct(strptime(bb2020$datetext, format = "%Y-%m-%d %H:%M:%S", "GMT")) 
bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
#add column for month and day of the week to create stats below
bb2020$month <- month(bb2020$date, label = TRUE, abbr = TRUE)
bb2020$weekday <- wday(bb2020$date, label = TRUE)
#create 2 dataframes for 2 matching periods
Wks2019 <- selectByDate(bb2020, start = "2019-01-21", end = "2019-06-02")
names(Wks2019)[1:2] <- c("dateOnly", "date") #413788 obs
Wks2019$Year <- rep(2019)
Wks2020 <- selectByDate(bb2020, start = "2020-01-20", end = "2020-05-31")
names(Wks2020)[1:2] <- c("dateOnly", "date") #521773 obs
Wks2020$Year <- rep(2020)
TSbb19_20 <- rbind(Wks2019, Wks2020) #935561 obs
#exclude obs from Saturdays, Sundays, bank holidays and midnight-6am
TSbb19_20 <- selectByDate(TSbb19_20, day = "weekday", hour = 6:23) #641830 obs
TSbb19_20$dateOnly <- as.character(TSbb19_20$dateOnly)
TSbb19_20 <- TSbb19_20[which(TSbb19_20$dateOnly != "2019-04-19" &
                               TSbb19_20$dateOnly != "2019-04-22" &
                               TSbb19_20$dateOnly != "2019-05-06" &
                               TSbb19_20$dateOnly != "2019-05-27" &
                               TSbb19_20$dateOnly != "2020-04-10" &
                               TSbb19_20$dateOnly != "2020-04-13" &
                               TSbb19_20$dateOnly != "2020-05-08" &
                     TSbb19_20$dateOnly != "2020-05-25"),] #619821 obs
rm(bb2020)
rm(Wks2019)
rm(Wks2020)
```

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}

# Structured TS

# convert to spatial object
coords_bb <- cbind(TSbb19_20$lon, TSbb19_20$lat)
bb.la <- SpatialPointsDataFrame(coords_bb, data = data.frame(TSbb19_20))
proj4string(bb.la) <- CRS("+init=epsg:4326") #define projection

# get LA 
# (i) directly from the web
#la <- readOGR("http://geoportal1-ons.opendata.arcgis.com/datasets/b6d2e15801de45328b760a4f55d74318_0.geojson?outSR={%22latestWkid%22:3857,%22wkid%22:102100}")#, layer="OGRGeoJSON")
# or, (ii) from the json I saved locally
path.json <- paste(path, "./data/raw/Local_Authority_Districts_(April_2019)_Boundaries_UK_BFE.geojson", sep = "")
la <- readOGR(path.json)#, layer="OGRGeoJSON")
# UK BGC la <- readOGR("https://opendata.arcgis.com/datasets/0e07a8196454415eab18c40a54dfbbef_0.geojson")
# UK BFC la <- readOGR("https://opendata.arcgis.com/datasets/1d78d47c87df4212b79fe2323aae8e08_0.geojson")

# source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe

# spatial transformations
la <- spTransform(la, CRS("+init=epsg:4326"))
la@data$LAD19NM <- as.character(la@data$LAD19NM)

# spatial join to LAD
bb.la.sp <- over(bb.la, la[, "LAD19NM"]) #not a spatial object
bb.la$LAD19NM <- bb.la.sp$LAD19NM

# create dataframe object to analyse (remove lat and lon?)
TSbb19_20sp <- bb.la@data

# drop NAs
TSbb19_20sp <- TSbb19_20sp[!is.na(TSbb19_20sp$LAD19NM),]
#NA LAD19NM, coord outside UK =  615516 obs (99.3% of obs retained)
```

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
#split by year before create stats by each hour of each day of the week
TS2019 <- selectByDate(TSbb19_20sp, year = 2019) #272248 obs
TS2020 <- selectByDate(TSbb19_20sp, year = 2020) #343268 obs

#5 weekdays x 18 hours x 382 LA = 34,380 obs
ts.tests <- TS2020 %>%
  group_by(weekday, hour, LAD19NM) %>%
  summarise(n.tests = n())
sapply(ts.tests, function(x) sum(is.na(x)))

# summarise by frequency of tests and geography
la.dateAvg <- summarise(group_by(ts.tests, LAD19NM), mean(n.tests))
range(la.dateAvg$`mean(n.tests)`)
#exclude LAs with average of less than 3 tests per hour per weekday
la.exc <- la.dateAvg[which(la.dateAvg$`mean(n.tests)` <3),]
la.dateAvg <- la.dateAvg[which(la.dateAvg$`mean(n.tests)` >=3),] #374 or 98% of LADs, so sufficient to keep 90 data points in ts
la.exc
#remove 8 LAD from TS2019 and TS2020?

# map LA mean test frequency per date
la@data <- left_join(la@data, la.dateAvg, by = "LAD19NM")
var <- la@data[ ,"mean(n.tests)"]
breaks <- classIntervals(var, n = 5, style = "quantile")
my_colours <- brewer.pal(5, "Blues")
plot(la, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE,
                 border = rgb(0.8, 0.8, 0.8, 0))

#look at 2019?
ts.tests <- TS2019 %>%
  group_by(weekday, hour, LAD19NM) %>%
  summarise(n.tests = n())
sapply(ts.tests, function(x) sum(is.na(x)))

```

### 2020-06-29 src/Data_Spatial.Rmd
I just moved the .Rmd to the .\src and added some spaces to makeit more readable.

--- 2020-06-26 Data_Spatial.Rmd
@@
 bb2020$date <- bb2020$datetext2$date
+
 #delete datetext2 column(s)
 bb2020 <- bb2020[ ,c(1:6,8:9)]
+
 #convert to date form datetext and date columns
@@
 bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
+
 #add column for month and day of the week to create stats below
@@
 bb2020$weekday <- wday(bb2020$date, label = TRUE)
+
 #create 2 dataframes for 2 matching periods
@@
 TSbb19_20 <- rbind(Wks2019, Wks2020) #935561 obs
+
 #exclude obs from Saturdays, Sundays, bank holidays and midnight-6am
@@
 la <- readOGR(path.json)#, layer="OGRGeoJSON")
-# UK BGC la <- readOGR("https://opendata.arcgis.com/datasets/0e07a8196454415eab18c40a54dfbbef_0.geojson")
-# UK BFC la <- readOGR("https://opendata.arcgis.com/datasets/1d78d47c87df4212b79fe2323aae8e08_0.geojson")
-
 # source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe
@@
 ```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
+
 #split by year before create stats by each hour of each day of the week

### 2020-07-06 src/Data_Spatial.Rmd
ts clusters

@@
 #NA LAD19NM, coord outside UK =  615516 obs (99.3% of obs retained)
+
+#save files
+out.TSbb19_20sp <- paste(path, "./data/temp/TSbb19_20sp.csv", sep = "")
+write_csv(TSbb19_20sp, out.TSbb19_20sp)
 ```

### 2020-07-06 src/ts_clusters.Rmd
ts clusters

@@
 library(dtwclust)
+library(leaflet)
 
@@
 
-# load and clean the data.
-# copied from bbData2020.Rmd
+# load the data created by Data_Spatial.Rmd
 
-#upload data file
-path.data <- paste(path, "./data/raw/speedtest.csv", sep = "")
+path.data <- paste(path, "/data/temp/TSbb19_20sp.csv", sep = "")
 
-#bb2020 <- read.csv("./data/raw/speedtest.csv")
-bb2020 <- read.csv(path.data)
+TSbb19_20sp <- read_csv(path.data)
 
-#name columns
-names(bb2020)[1:6] <- c("datetext","speeddown","speedup","provider","lat", "lon")
-
-str(bb2020)
-summary(bb2020) #2030205 obs
-#conversions needed to numeric? df$col <- as.numeric(df$col) 
-#remove empty rows or coerced to NA? df[complete.cases(df),] or df <- df[!is.na(df$col),]
-
-#remove outliers based on Riddlesden and Singleton 2014 for slower end
-#check fastest upload and download speeds commercially available from Virgin Media (Gigaclear offers even faster!)
-bb2020 <- bb2020[bb2020$speeddown>512,] 
-bb2020 <- bb2020[bb2020$speeddown<362000,]
-bb2020 <- bb2020[bb2020$speedup<21000,] #1839333
-
-#remove below seconds, create columns for date and hour
-bb2020$datetext <- strtrim(x = bb2020$datetext, width = 19) 
-bb2020$datetext2 <- bb2020$datetext
-bb2020$datetext2 <- strtrim(x = bb2020$datetext2, width = 13)
-bb2020$datetext2 <- colsplit(bb2020$datetext2, pattern = " ", c("date", "hour"))
-bb2020$hour <- bb2020$datetext2$hour
-bb2020$date <- bb2020$datetext2$date
-#delete datetext2 column(s)
-bb2020 <- bb2020[ ,c(1:6,8:9)]
-#convert to date form datetext and date columns
-bb2020$datetext <- as.POSIXct(strptime(bb2020$datetext, format = "%Y-%m-%d %H:%M:%S", "GMT")) 
-bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
-#create dataframe with number of tests, mean and standard deviation by month for adding to dataframe later
-MonthFreq <- timeAverage(bb2020, avg.time = "month", statistic = "frequency")
-MonthMean <- timeAverage(bb2020, avg.time = "month", statistic = "mean")
-MonthSD <- timeAverage(bb2020, avg.time = "month", statistic = "sd")
-MonthStats <- cbind(MonthFreq[,1:2], MonthMean[,c(3:4)], 
-                    MonthSD[,c(3:4)])
-names(MonthStats) <- c("month", "monthlyTests", "monthlyMeanSpDown", 
-                       "monthlyMeanSpUp", "monthlySDSpDown", 
-                       "monthlySDSpUp")
-MonthStats$month <- month(MonthStats$month, label = TRUE, abbr = TRUE)
-#add column for month and day of the week to create stats below
-bb2020$month <- month(bb2020$date, label = TRUE, abbr = TRUE)
-bb2020$weekday <- wday(bb2020$date, label = TRUE)
-#divide MonthStats for future join with separate 2019 and 2020 dataframes
-MonthStats19 <- MonthStats[1:6,]
-MonthStats20 <- MonthStats[13:17,]
-```
-
-## Structured time series
-
-```{r eval=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
-
-# convert to spatial object
-coords_bb <- cbind(bb2020$lon, bb2020$lat)
-bb.la <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020))
-proj4string(bb.la) <- CRS("+init=epsg:4326") #define projection
-
-# get LA 
-# (i) directly from the web
-#la <- readOGR("http://geoportal1-ons.opendata.arcgis.com/datasets/b6d2e15801de45328b760a4f55d74318_0.geojson?outSR={%22latestWkid%22:3857,%22wkid%22:102100}")#, layer="OGRGeoJSON")
-# or, (ii) from the json I saved locally
-path.json <- paste(path, "./data/raw/Local_Authority_Districts_(April_2019)_Boundaries_UK_BFE.geojson", sep = "")
-la <- readOGR(path.json)#, layer="OGRGeoJSON")
-# UK BGC la <- readOGR("https://opendata.arcgis.com/datasets/0e07a8196454415eab18c40a54dfbbef_0.geojson")
-# UK BFC la <- readOGR("https://opendata.arcgis.com/datasets/1d78d47c87df4212b79fe2323aae8e08_0.geojson")
-
-# source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe
-
-# spatial transformations
-la <- spTransform(la, CRS("+init=epsg:4326"))
-la@data$LAD19NM <- as.character(la@data$LAD19NM)
-
-# spatial join to LAD
-bb.la.sp <- over(bb.la, la[, "LAD19NM"]) #not a spatial object
-bb.la$LAD19NM <- bb.la.sp$LAD19NM
-
-# create dataframe object to analyse (remove lat and lon?)
-bb2020sp <- bb.la@data
-
-# drop NAs
-bb2020sp <- bb2020sp[!is.na(bb2020sp$LAD19NM),]
-#13431/1839332 = .0073 NA LAD19NM, coord outside UK
-
-# summarise by date and geography
-la.date.count <- count(bb2020sp, LAD19NM, date)
-la.dateAvg <- summarise(group_by(la.date.count, LAD19NM), mean(n))
-range(la.dateAvg$`mean(n)`)
-
-# map LA mean test frequency per date
-la@data <- left_join(la@data, la.dateAvg, by = "LAD19NM")
-var <- la@data[ ,'mean(n)']
-breaks <- classIntervals(var, n = 5, style = "quantile")
-my_colours <- brewer.pal(5, "Blues")
-plot(la, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE,
-                 border = rgb(0.8, 0.8, 0.8, 0))
-
-```
-
-## Keep working hours and days
-
-The time series are based on mean upload and download speeds as well as the total number of tests for working days and during working hours
-
-```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
-
-bb2020sp <- bb2020sp %>% arrange(datetext)
-
-# keep working hours
-bb2020sp <- bb2020sp[bb2020sp$hour>=9 & bb2020sp$hour<17,]
-
-# drop w/e
-bb2020sp <- bb2020sp %>%
-  filter(bb2020sp$weekday!="Sat" & bb2020sp$weekday!="Sun")
-
-head(bb2020sp)
+#split by year before create stats by each hour of each day of the week
+TS2019 <- selectByDate(TSbb19_20sp, year = 2019) #272248 obs
+TS2020 <- selectByDate(TSbb19_20sp, year = 2020) #343268 obs
 ```
@@
 
-# download
-ts.down <- bb2020sp %>%
-  group_by(date,LAD19NM) %>%
+# download 2020
+ts.down2020 <- TS2020 %>%
+  group_by(weekday, hour, LAD19NM) %>%
   summarise(mean.down = mean(speeddown))#, mean.down = mean(speeddown), n.tests = n())
 
-# upload
-ts.up <- bb2020sp %>%
-  group_by(date,LAD19NM) %>%
+# upload 2020
+ts.up2020 <- TS2020 %>%
+  group_by(weekday, hour, LAD19NM) %>%
   summarise(mean.up = mean(speedup))
 
-# tests
-ts.tests <- bb2020sp %>%
-  group_by(date,LAD19NM) %>%
+# tests 2020
+ts.tests2020 <- TS2020 %>%
+  group_by(weekday, hour, LAD19NM) %>%
   summarise(n.tests = n())
+
+# download 2019
+ts.down2019 <- TS2019 %>%
+  group_by(weekday, hour, LAD19NM) %>%
+  summarise(mean.down = mean(speeddown))#, mean.down = mean(speeddown), n.tests = n())
+
+# upload 2019
+ts.up2019 <- TS2019 %>%
+  group_by(weekday, hour, LAD19NM) %>%
+  summarise(mean.up = mean(speedup))
+
+# tests 2019
+ts.tests2019 <- TS2019 %>%
+  group_by(weekday, hour, LAD19NM) %>%
+  summarise(n.tests = n())
+
 ```
@@
 
-# upload working hours and days
+# create a 'fake' week for a ts id
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2020, month = 2, day = 3),
+                make_datetime(year = 2020, month = 2, day = 4),
+                make_datetime(year = 2020, month = 2, day = 5),
+                make_datetime(year = 2020, month = 2, day = 6),
+                make_datetime(year = 2020, month = 2, day = 7)),
+                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+ts.wide <- merge(ts.up2020, help.dates, by = "weekday")
+ts.wide$hour <- paste(ts.wide$hour, ":00", sep="")
+ts.wide$help.date.time <- with(ts.wide, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
 
-ts.wide <- ts.up %>%
-  select(date, LAD19NM, mean.up) %>%
-  filter(date > "2020-2-15")
+ts.wide <- ts.wide %>%
+  select(help.date.time, LAD19NM, mean.up) #%>%
+  #filter(date > "2020-2-15")
 
@@
 
-# impute with means
+# impute with means of each TS
+# TODO check sensitivity
 library(imputeTS)
@@
 names(pc_k) <- paste0("k_", c("5","10","15", "20"))
-sapply(pc_k, cvi, type = "internal")
+cvis <- sapply(pc_k, cvi, type = "internal")
+cvis
+# max Sil
+# max SF
+# max CH
+# min DB
+# min DBstar
+# max D
+# min COP
 
-# I need to work on the selection
+# loop for k decision
+pref.cvi <- NULL
 
-pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
+max.cvis <- c(1:3,6)
+for (i in max.cvis){
+  a <- which.max(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+min.cvis <- c(4:5,7)
+for (i in min.cvis){
+  a <- which.min(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+print(pref.cvi) # cvis[2,] is always 0
+
+pc_k <- tsclust(ts.wide, k = 10, seed = 94L, 
                 type="partitional", 
@@
                 args = tsclust_args(dist = list(window.size = 20L)))
-
 plot(pc_k)
@@
 
-# downlad working hours and days
+ts.wide <- merge(ts.down2020, help.dates, by = "weekday")
+ts.wide$hour <- paste(ts.wide$hour, ":00", sep="")
+ts.wide$help.date.time <- with(ts.wide, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
 
-ts.wide <- ts.down %>%
-  select(date, LAD19NM, mean.down) %>%
-  filter(date > "2020-2-15")
+ts.wide <- ts.wide %>%
+  select(help.date.time, LAD19NM, mean.down) #%>%
+  #filter(date > "2020-2-15")
 
@@
 names(pc_k) <- paste0("k_", c("5","10","15", "20"))
-sapply(pc_k, cvi, type = "internal")
+cvis <- sapply(pc_k, cvi, type = "internal")
+cvis
+# max Sil
+# max SF
+# max CH
+# min DB
+# min DBstar
+# max D
+# min COP
 
-# I need to work on the selection
+# loop for k decision
+pref.cvi <- NULL
+
+max.cvis <- c(1:3,6)
+for (i in max.cvis){
+  a <- which.max(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+min.cvis <- c(4:5,7)
+for (i in min.cvis){
+  a <- which.min(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+print(pref.cvi) # cvis[2,] is always 0
 
@@
 
-### N. of speeds
+### N. of speed tests
 
@@
 
-# tests working hours and days
+# tests 
+ts.wide <- merge(ts.tests2020, help.dates, by = "weekday")
+ts.wide$hour <- paste(ts.wide$hour, ":00", sep="")
+ts.wide$help.date.time <- with(ts.wide, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
 
-ts.wide <- ts.tests %>%
-  select(date, LAD19NM, n.tests) %>%
-  filter(date > "2020-2-15")
+ts.wide <- ts.wide %>%
+  select(help.date.time, LAD19NM, n.tests) #%>%
+  #filter(date > "2020-2-15")
 
@@
 names(pc_k) <- paste0("k_", c("5","10","15", "20"))
-sapply(pc_k, cvi, type = "internal")
+cvis <- sapply(pc_k, cvi, type = "internal")
+cvis
+# max Sil
+# max SF
+# max CH
+# min DB
+# min DBstar
+# max D
+# min COP
 
-# I need to work on the selection
+# loop for k decision
+pref.cvi <- NULL
 
-pc_k <- tsclust(ts.wide, k = 15, seed = 94L, 
+max.cvis <- c(1:3,6)
+for (i in max.cvis){
+  a <- which.max(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+min.cvis <- c(4:5,7)
+for (i in min.cvis){
+  a <- which.min(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+print(pref.cvi) # cvis[2,] is always 0
+
+pc_k <- tsclust(ts.wide, k = 5, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+plot(pc_k)
+
+cluster_tests <- data.frame(LAD=ts.wide.names,
+                           cluster=pc_k@cluster)
+```
+
+### Upload speed diff 2019-2020
+
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# 2020 copied from above
+
+# create a 'fake' week for a ts id
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2020, month = 2, day = 3),
+                make_datetime(year = 2020, month = 2, day = 4),
+                make_datetime(year = 2020, month = 2, day = 5),
+                make_datetime(year = 2020, month = 2, day = 6),
+                make_datetime(year = 2020, month = 2, day = 7)),
+                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+ts.wide <- merge(ts.up2020, help.dates, by = "weekday")
+ts.wide$hour <- paste(ts.wide$hour, ":00", sep="")
+ts.wide$help.date.time <- with(ts.wide, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+
+ts.wide <- ts.wide %>%
+  select(help.date.time, LAD19NM, mean.up) #%>%
+  #filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide <- ts_wide(ts.wide)
+
+# impute with means of each TS
+# TODO check sensitivity
+library(imputeTS)
+ts.wide <- na_mean(ts.wide)
+
+# 2019
+
+# create a 'fake' week for a ts id
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2019, month = 2, day = 4),
+                make_datetime(year = 2019, month = 2, day = 5),
+                make_datetime(year = 2019, month = 2, day = 6),
+                make_datetime(year = 2019, month = 2, day = 7),
+                make_datetime(year = 2019, month = 2, day = 8)),
+                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+ts.wide2019 <- merge(ts.up2019, help.dates, by = "weekday")
+ts.wide2019$hour <- paste(ts.wide2019$hour, ":00", sep="")
+ts.wide2019$help.date.time <- with(ts.wide2019, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+
+ts.wide2019 <- ts.wide2019 %>%
+  select(help.date.time, LAD19NM, mean.up) #%>%
+  #filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide2019 <- ts_wide(ts.wide2019)
+
+# impute with means of each TS
+# TODO check sensitivity
+library(imputeTS)
+ts.wide2019 <- na_mean(ts.wide2019)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# diff 2020-2019
+ts.wide <- mapply('-', ts.wide, ts.wide2019, SIMPLIFY = FALSE)
+ts.wide <- as.data.frame(ts.wide)
+
+# extract names
+ts.wide.names <- names(ts.wide[,-1])
+ts.wide <- ts.wide[,-1]
+
+# standardise
+ts.wide <- zscore(ts.wide)
+
+# convert to list for tsclust
+ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
+
+# 5 - 20 cluster sollutions
+pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
                 type="partitional",
@@
 
+# decide k, see RJ-2019-023 for max/min
+names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+cvis <- sapply(pc_k, cvi, type = "internal")
+cvis
+# max Sil
+# max SF
+# max CH
+# min DB
+# min DBstar
+# max D
+# min COP
+
+# loop for k decision
+pref.cvi <- NULL
+
+max.cvis <- c(1:3,6)
+for (i in max.cvis){
+  a <- which.max(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+min.cvis <- c(4:5,7)
+for (i in min.cvis){
+  a <- which.min(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+print(pref.cvi) # cvis[2,] is always 0
+
+pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
 plot(pc_k)
 
-cluster_tests <- data.frame(LAD=ts.wide.names,
+cluster_up.diff <- data.frame(LAD=ts.wide.names,
                            cluster=pc_k@cluster)
 
-clusters <- merge(cluster_up, cluster_down, by = "LAD")
-names(clusters)[2] <- "cluster.up"
-names(clusters)[3] <- "cluster.down"
-clusters <- merge(clusters, cluster_tests, by = "LAD")
-names(clusters)[4] <- "cluster.tests"
-#sapply(clusters, function(x) sum(is.na(x)))
+```
 
-clusters  %>% arrange(cluster.up) %>% kable() 
+### download speed diff 2019-2020
+
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# 2020 copied from above
+
+# create a 'fake' week for a ts id
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2020, month = 2, day = 3),
+                make_datetime(year = 2020, month = 2, day = 4),
+                make_datetime(year = 2020, month = 2, day = 5),
+                make_datetime(year = 2020, month = 2, day = 6),
+                make_datetime(year = 2020, month = 2, day = 7)),
+                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+ts.wide <- merge(ts.down2020, help.dates, by = "weekday")
+ts.wide$hour <- paste(ts.wide$hour, ":00", sep="")
+ts.wide$help.date.time <- with(ts.wide, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+
+ts.wide <- ts.wide %>%
+  select(help.date.time, LAD19NM, mean.down) #%>%
+  #filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide <- ts_wide(ts.wide)
+
+# impute with means of each TS
+# TODO check sensitivity
+library(imputeTS)
+ts.wide <- na_mean(ts.wide)
+
+# 2019
+
+# create a 'fake' week for a ts id
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2019, month = 2, day = 4),
+                make_datetime(year = 2019, month = 2, day = 5),
+                make_datetime(year = 2019, month = 2, day = 6),
+                make_datetime(year = 2019, month = 2, day = 7),
+                make_datetime(year = 2019, month = 2, day = 8)),
+                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+ts.wide2019 <- merge(ts.down2019, help.dates, by = "weekday")
+ts.wide2019$hour <- paste(ts.wide2019$hour, ":00", sep="")
+ts.wide2019$help.date.time <- with(ts.wide2019, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+
+ts.wide2019 <- ts.wide2019 %>%
+  select(help.date.time, LAD19NM, mean.down) #%>%
+  #filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide2019 <- ts_wide(ts.wide2019)
+
+# impute with means of each TS
+# TODO check sensitivity
+library(imputeTS)
+ts.wide2019 <- na_mean(ts.wide2019)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# diff 2020-2019
+ts.wide <- mapply('-', ts.wide, ts.wide2019, SIMPLIFY = FALSE)
+ts.wide <- as.data.frame(ts.wide)
+
+# extract names
+ts.wide.names <- names(ts.wide[,-1])
+ts.wide <- ts.wide[,-1]
+
+# standardise
+ts.wide <- zscore(ts.wide)
+
+# convert to list for tsclust
+ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
+
+# 5 - 20 cluster sollutions
+pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
+                type="partitional",
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+# decide k, see RJ-2019-023 for max/min
+names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+cvis <- sapply(pc_k, cvi, type = "internal")
+cvis
+# max Sil
+# max SF
+# max CH
+# min DB
+# min DBstar
+# max D
+# min COP
+
+# loop for k decision
+pref.cvi <- NULL
+
+max.cvis <- c(1:3,6)
+for (i in max.cvis){
+  a <- which.max(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+min.cvis <- c(4:5,7)
+for (i in min.cvis){
+  a <- which.min(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+print(pref.cvi) # cvis[2,] is always 0
+
+pc_k <- tsclust(ts.wide, k = 10, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 15L)))
+plot(pc_k)
+
+cluster_down.diff <- data.frame(LAD=ts.wide.names,
+                           cluster=pc_k@cluster)
 ```
@@
 
-la@data <- merge(la@data, clusters, by.x = "LAD19NM", by.y = "LAD")
+# clusters data frame
+clusters <- merge(cluster_up, cluster_down, by = "LAD")
+names(clusters)[2] <- "cluster.up"
+names(clusters)[3] <- "cluster.down"
 
-var <- la@data[ ,'cluster.up']
-plot(la, col = var, main = "Upload clusters",
-     border = rgb(0.8, 0.8, 0.8, 0))
+clusters <- merge(clusters, cluster_up.diff, by = "LAD")
+names(clusters)[4] <- "cluster.up.diff"
 
-var <- la@data[ ,'cluster.down']
-plot(la, col = var, main = "Download clusters",
-     border = rgb(0.8, 0.8, 0.8, 0))
+clusters <- merge(clusters, cluster_down.diff, by = "LAD")
+names(clusters)[5] <- "cluster.down.diff"
 
-var <- la@data[ ,'cluster.tests']
-plot(la, col = var, main = "Tests clusters",
-     border = rgb(0.8, 0.8, 0.8, 0))
+clusters <- merge(clusters, cluster_tests, by = "LAD")
+names(clusters)[6] <- "cluster.tests"
+#sapply(clusters, function(x) sum(is.na(x)))
+
+clusters  %>% arrange(cluster.up) %>% kable()
+
+# save clusters file
+out.clusters <- paste(path, "./data/temp/clusters.csv", sep = "")
+write_csv(clusters, out.clusters)
+
+# get LA 
+# this is the heaviest version:
+# (i) directly from the web
+#la <- readOGR("http://geoportal1-ons.opendata.arcgis.com/datasets/b6d2e15801de45328b760a4f55d74318_0.geojson?outSR={%22latestWkid%22:3857,%22wkid%22:102100}")#, layer="OGRGeoJSON")
+# or, (ii) from the json I saved locally
+# path.json <- paste(path, "./data/raw/Local_Authority_Districts_(April_2019)_Boundaries_UK_BFE.geojson", sep = "")
+# la <- readOGR(path.json)#, layer="OGRGeoJSON")
+# UK BGC: 
+la <- readOGR("https://opendata.arcgis.com/datasets/0e07a8196454415eab18c40a54dfbbef_0.geojson")
+# UK BFC: la <- readOGR("https://opendata.arcgis.com/datasets/1d78d47c87df4212b79fe2323aae8e08_0.geojson")
+
+# source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe
+
+# spatial transformations
+la <- spTransform(la, CRS("+init=epsg:4326"))
+#la@data$LAD19NM <- as.character(la@data$LAD19NM)
+la@data$lad19nm <- as.character(la@data$lad19nm)
+
+#la@data <- merge(la@data, clusters, by.x = "LAD19NM", by.y = "LAD")
+la@data <- merge(la@data, clusters, by.x = "lad19nm", by.y = "LAD")
+
+# plot maps
+# var <- la@data[ ,'cluster.up']
+# plot(la, col = var, main = "Upload clusters",
+#      border = rgb(0.8, 0.8, 0.8, 0))
+# 
+# var <- la@data[ ,'cluster.up.diff']
+# plot(la, col = var, main = "Upload difference clusters",
+#      border = rgb(0.8, 0.8, 0.8, 0))
+# 
+# 
+# var <- la@data[ ,'cluster.down']
+# plot(la, col = var, main = "Download clusters",
+#      border = rgb(0.8, 0.8, 0.8, 0))
+# 
+# var <- la@data[ ,'cluster.tests']
+# plot(la, col = var, main = "Tests clusters",
+#      border = rgb(0.8, 0.8, 0.8, 0))
+
+# leaflet maps
+
+# factors
+la@data$cluster.up <- as.factor(la@data$cluster.up)
+la@data$cluster.down <- as.factor(la@data$cluster.down)
+la@data$cluster.up.diff <- as.factor(la@data$cluster.up.diff)
+la@data$cluster.down.diff <- as.factor(la@data$cluster.down.diff)
+la@data$cluster.tests <- as.factor(la@data$cluster.tests)
+
+# palettes
+factpal.up <- colorFactor("RdYlBu", la@data$cluster.up)
+factpal.down <- colorFactor("RdYlBu", la@data$cluster.down)
+factpal.up.diff <- colorFactor("RdYlBu", la@data$cluster.up.diff)
+factpal.down.diff <- colorFactor("RdYlBu", la@data$cluster.down.diff)
+factpal.tests <- colorFactor("RdYlBu", la@data$cluster.tests)
+
+leaflet(la) %>%
+  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
+    color = ~factpal.up(cluster.up), group = "cluster.up") %>%
+  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
+    color = ~factpal.down(cluster.down), group = "cluster.down") %>%
+  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
+    color = ~factpal.up.diff(cluster.up.diff), group = "cluster.up.diff") %>%
+  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
+    color = ~factpal.down.diff(cluster.down.diff), group = "cluster.down.diff") %>%
+  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
+    color = ~factpal.tests(cluster.tests), group = "cluster.tests") %>%
+  hideGroup("cluster.up") %>%
+  hideGroup("cluster.down") %>%
+  hideGroup("cluster.up.diff") %>%
+  hideGroup("cluster.down.diff") %>%
+  #addLegend("topright",
+  #            pal = factpal,
+  #            values = la@data$cluster.up,
+  #            title = "Clusters") %>%
+  addLayersControl(
+    #baseGroups = c("Upload)", "Download"),
+    overlayGroups = c("cluster.up","cluster.down", "cluster.down", "cluster.up.diff", "cluster.down.diff", "cluster.tests"),
+    options = layersControlOptions(collapsed = FALSE)
+  )
 

### 2020-07-13 src/Data_Spatial.Rmd

@@
 #create 2 dataframes for 2 matching periods
-Wks2019 <- selectByDate(bb2020, start = "2019-01-21", end = "2019-06-02")
+Wks2019 <- selectByDate(bb2020, start = "2019-03-04", end = "2019-05-31")
 names(Wks2019)[1:2] <- c("dateOnly", "date") #413788 obs
 Wks2019$Year <- rep(2019)
-Wks2020 <- selectByDate(bb2020, start = "2020-01-20", end = "2020-05-31")
+Wks2020 <- selectByDate(bb2020, start = "2020-03-02", end = "2020-05-29")
 names(Wks2020)[1:2] <- c("dateOnly", "date") #521773 obs

### 2020-07-13 src/Data_Spatial.Rmd
I applied the new dates and estimated the clusters. leaflet map is the same. No more time to work on this this morning. I will try again tonight for the legends.

@@
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
-output: html_document
+output: 
+  html_document:
     df_print: paged

### 2020-07-13 src/ts_clusters.Rmd
I applied the new dates and estimated the clusters. leaflet map is the same. No more time to work on this this morning. I will try again tonight for the legends.

@@
 # create a 'fake' week for a ts id
-help.dates <- tibble(help.date.time = c(make_datetime(year = 2020, month = 2, day = 3),
-                make_datetime(year = 2020, month = 2, day = 4),
-                make_datetime(year = 2020, month = 2, day = 5),
-                make_datetime(year = 2020, month = 2, day = 6),
-                make_datetime(year = 2020, month = 2, day = 7)),
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2020, month = 3, day = 2),
+                make_datetime(year = 2020, month = 3, day = 3),
+                make_datetime(year = 2020, month = 3, day = 4),
+                make_datetime(year = 2020, month = 3, day = 5),
+                make_datetime(year = 2020, month = 3, day = 6)),
                 weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
@@
 # TODO check sensitivity
-library(imputeTS)
 ts.wide <- na_mean(ts.wide)
@@
 # create a 'fake' week for a ts id
-help.dates <- tibble(help.date.time = c(make_datetime(year = 2019, month = 2, day = 4),
-                make_datetime(year = 2019, month = 2, day = 5),
-                make_datetime(year = 2019, month = 2, day = 6),
-                make_datetime(year = 2019, month = 2, day = 7),
-                make_datetime(year = 2019, month = 2, day = 8)),
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2019, month = 3, day = 4),
+                make_datetime(year = 2019, month = 3, day = 5),
+                make_datetime(year = 2019, month = 3, day = 6),
+                make_datetime(year = 2019, month = 3, day = 7),
+                make_datetime(year = 2019, month = 3, day = 8)),
                 weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
@@
 # TODO check sensitivity
-library(imputeTS)
 ts.wide2019 <- na_mean(ts.wide2019)

### 2020-07-13 src/ts_clusters.Rmd
minor tweak to the previous

@@
     overlayGroups = c("cluster.up","cluster.down", "cluster.down", "cluster.up.diff", "cluster.down.diff", "cluster.tests"),
-    options = layersControlOptions(collapsed = FALSE)
-  )
+    options = layersControlOptions(collapsed = FALSE)) #%>%
+  #addLegend(values = 1, group = "cluster.up", position = "bottomleft", colors= ~factpal.up(cluster.up)) #%>% 
+  #addLegend(values = 2, group = "cluster.down", position = "bottomleft", labels = "2" ,colors= "blue")
 

### 2020-07-13 src/ts_clusters.Rmd
Leaflet map update. Legends don't work v. good. Click and unclick all boxes and you will be able to see the correct legend.

--- 2020-07-13 src/ts_clusters.Rmd
@@
 leaflet(la) %>%
+  # add polygons as groups
   addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
@@
     color = ~factpal.tests(cluster.tests), group = "cluster.tests") %>%
+  # hide groups, so they don't appear all on in the begining 
   hideGroup("cluster.up") %>%
@@
   hideGroup("cluster.down.diff") %>%
-  #addLegend("topright",
-  #            pal = factpal,
-  #            values = la@data$cluster.up,
-  #            title = "Clusters") %>%
+  hideGroup("cluster.tests") %>%
+  # layer control
   addLayersControl(
@@
     overlayGroups = c("cluster.up","cluster.down", "cluster.down", "cluster.up.diff", "cluster.down.diff", "cluster.tests"),
-    options = layersControlOptions(collapsed = FALSE)
-  )
+    options = layersControlOptions(collapsed = FALSE)) %>%
+  # add different legends
+  addLegend(pal = factpal.up, 
+            values = la@data$cluster.up,
+            opacity = 1,
+            group = "cluster.up",
+            title = "Upload clusters") %>%
+  addLegend(pal = factpal.down, 
+            values = la@data$cluster.down,
+            opacity = 1,
+            group = "cluster.down",
+            title = "Download clusters") %>%
+  addLegend(pal = factpal.up.diff, 
+            values = la@data$cluster.up.diff,
+            opacity = 1,
+            group = "cluster.up.diff",
+            title = "Upload diff. clusters") %>%
+  addLegend(pal = factpal.down.diff, 
+            values = la@data$cluster.down.diff,
+            opacity = 1,
+            group = "cluster.down.diff",
+            title = "Download diff. clusters") %>%
+  addLegend(pal = factpal.tests, 
+            values = la@data$cluster.tests,
+            opacity = 1,
+            group = "cluster.tests",
+            title = "Test clusters")
 

### 2020-07-14 descriptive_Clusters.Rmd

---
title: "descriptive_Clusters"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output: 
  html_document:
    df_print: paged
    toc: true
    toc_float: true
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../output")
  })
---

```{r settings, include = FALSE}
library(tsbox)
library(ggplot2)
library(tidyverse)
library(geojsonio)
library(tsfeatures)
library(rprojroot)
library(openair)
library(reshape2)
library(lubridate)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(classInt)
library(RColorBrewer)
library(knitr)
library(rprojroot)
library(viridis)
library(hrbrthemes)

# This is the project path
path <- find_rstudio_root_file()
```

```{r include = FALSE}

# load the data created by Data_Spatial.Rmd and ts_clusters.Rmd

LAD.cluster <- paste(path, "./data/temp/clusters.csv", sep = "")

clusters <- read_csv(LAD.cluster)

path.data <- paste(path, "/data/temp/TSbb19_20sp.csv", sep = "")

TSbb19_20sp <- read_csv(path.data)

#split out 2020 before join to clusters to create descriptives
TS2020 <- selectByDate(TSbb19_20sp, year = 2020)

#join clusters to TS2020
names(clusters) [1] <- "LAD19NM"
TS2020 <- left_join(TS2020, clusters)
#remove NAs? - otherwise cannot create unified dataframes for below stats

#summarise by LADs per cluster and Total tests / Mean speed per cluster
LAD.cluster.tests <- clusters %>%
  group_by(cluster.tests) %>%
  summarise(n.LADs = n())
LAD.cluster.tests$Total <- TS2020 %>%
  group_by(cluster.tests) %>%
  summarise(n.cluster = n())
LAD.cluster.up <- clusters %>%
  group_by(cluster.up) %>%
  summarise(up.LADs = n())
LAD.cluster.up$Mean <- summarise(group_by(TS2020, cluster.up),
  mean(speedup))
LAD.cluster.up$SD <- summarise(group_by(TS2020, cluster.up),
  sd(speedup))
LAD.cluster.down <- clusters %>%
  group_by(cluster.down) %>%
  summarise(down.LADs = n())
LAD.cluster.down$Mean <- summarise(group_by(TS2020, cluster.down),
  mean(speeddown))
LAD.cluster.down$SD <- summarise(group_by(TS2020, cluster.down),
  sd(speeddown))
LAD.cluster.diffUp <- clusters %>%
  group_by(cluster.up.diff) %>%
  summarise(diffUp.LADs = n())
LAD.cluster.diffDown <- clusters %>%
  group_by(cluster.down.diff) %>%
  summarise(diffDown.LADs = n())

```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r }
#summarise test frequency clusters by hour of each weekday and plot
cluster.tests <- TS2020 %>%
  group_by(cluster.tests, weekday, hour) %>%
  summarise(n.tests = n())
names(cluster.tests)[1] <- "clusterID"
cluster.tests$clusterID <- as.factor(cluster.tests$clusterID)
#add fake dates for plotting
help.dates <- tibble(help.date.time = c(
  make_datetime(year = 2020, month = 3, day = 2),
  make_datetime(year = 2020, month = 3, day = 3),
  make_datetime(year = 2020, month = 3, day = 4),
  make_datetime(year = 2020, month = 3, day = 5),
  make_datetime(year = 2020, month = 3, day = 6)),
                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
cluster.tests <- merge(cluster.tests, help.dates, by = "weekday")
cluster.tests$hour <- paste(cluster.tests$hour, ":00", sep="")
cluster.tests$help.date.time <- with(cluster.tests, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
#plot cluster statistic for number tests
TestCluster <- ggplot(cluster.tests, aes(help.date.time, n.tests, 
                                         colour = clusterID))
TestCluster + geom_line() + labs(x = "Weekdays March-May") + 
  scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
                    "2020-03-03 15:00:00", "2020-03-04 15:00:00",
                    "2020-03-05 15:00:00", "2020-03-06 15:00:00"), 
                    labels = "%a"), labels = 
                     c("Mon", "Tues", "Weds", "Thurs", "Fri"))

#summarise upload speed clusters by hour of each weekday and plot
cluster.up <- TS2020 %>%
  group_by(cluster.up, weekday, hour) %>%
  summarise(mean.up = mean(speedup))
names(cluster.up)[1] <- "clusterID"
names(LAD.cluster.up)[1] <- "clusterID"
cluster.up <- left_join(cluster.up, LAD.cluster.up)
cluster.up <- cluster.up[which(cluster.up$up.LADs > 5),]
cluster.up$clusterID <- as.factor(cluster.up$clusterID)
#add fake dates for plotting
help.dates <- tibble(help.date.time = c(
  make_datetime(year = 2020, month = 3, day = 2),
  make_datetime(year = 2020, month = 3, day = 3),
  make_datetime(year = 2020, month = 3, day = 4),
  make_datetime(year = 2020, month = 3, day = 5),
  make_datetime(year = 2020, month = 3, day = 6)),
                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
cluster.up <- merge(cluster.up, help.dates, by = "weekday")
cluster.up$hour <- paste(cluster.up$hour, ":00", sep="")
cluster.up$help.date.time <- with(cluster.up, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
#plot cluster statistic for mean upload speed
UpCluster <- ggplot(cluster.up, aes(help.date.time, mean.up, 
                                         colour = clusterID))
UpCluster + geom_line() + labs(x = "Weekdays March-May") + 
  scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
                    "2020-03-03 15:00:00", "2020-03-04 15:00:00",
                    "2020-03-05 15:00:00", "2020-03-06 15:00:00"), 
                    labels = "%a"), labels = 
                     c("Mon", "Tues", "Weds", "Thurs", "Fri"))

#summarise download speed clusters by hour of each weekday and plot
cluster.down <- TS2020 %>%
  group_by(cluster.down, weekday, hour) %>%
  summarise(mean.down = mean(speeddown))
names(cluster.down)[1] <- "clusterID"
names(LAD.cluster.down)[1] <- "clusterID"
cluster.down <- left_join(cluster.down, LAD.cluster.down)
cluster.down <- cluster.down[which(cluster.down$down.LADs > 5),]
cluster.down$clusterID <- as.factor(cluster.down$clusterID)
#add fake dates for plotting
help.dates <- tibble(help.date.time = c(
  make_datetime(year = 2020, month = 3, day = 2),
  make_datetime(year = 2020, month = 3, day = 3),
  make_datetime(year = 2020, month = 3, day = 4),
  make_datetime(year = 2020, month = 3, day = 5),
  make_datetime(year = 2020, month = 3, day = 6)),
                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
cluster.down <- merge(cluster.down, help.dates, by = "weekday")
cluster.down$hour <- paste(cluster.down$hour, ":00", sep="")
cluster.down$help.date.time <- with(cluster.down, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
#plot cluster statistic for mean download speeds
TestCluster <- ggplot(cluster.down, aes(help.date.time, mean.down, 
                                         colour = clusterID))
TestCluster + geom_line() + labs(x = "Weekdays March-May") + 
  scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
                    "2020-03-03 15:00:00", "2020-03-04 15:00:00",
                    "2020-03-05 15:00:00", "2020-03-06 15:00:00"), 
                    labels = "%a"), labels = 
                     c("Mon", "Tues", "Weds", "Thurs", "Fri"))

```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2020-07-14 descriptive_Clusters.Rmd

@@
   summarise(n.LADs = n())
-LAD.cluster.tests$Total <- TS2020 %>%
+LAD.cluster.testsTotal <- TS2020 %>%
   group_by(cluster.tests) %>%
   summarise(n.cluster = n())
+LAD.cluster.tests <- left_join(LAD.cluster.tests, LAD.cluster.testsTotal)
 LAD.cluster.up <- clusters %>%
@@
   summarise(up.LADs = n())
-LAD.cluster.up$Mean <- summarise(group_by(TS2020, cluster.up),
+LAD.cluster.upMean <- summarise(group_by(TS2020, cluster.up),
   mean(speedup))
-LAD.cluster.up$SD <- summarise(group_by(TS2020, cluster.up),
+LAD.cluster.upSD <- summarise(group_by(TS2020, cluster.up),
   sd(speedup))
+LAD.cluster.up <- cbind(LAD.cluster.up, LAD.cluster.upMean[,2], LAD.cluster.upSD[,2])
 LAD.cluster.down <- clusters %>%
@@
   summarise(down.LADs = n())
-LAD.cluster.down$Mean <- summarise(group_by(TS2020, cluster.down),
+LAD.cluster.downMean <- summarise(group_by(TS2020, cluster.down),
   mean(speeddown))
-LAD.cluster.down$SD <- summarise(group_by(TS2020, cluster.down),
+LAD.cluster.downSD <- summarise(group_by(TS2020, cluster.down),
   sd(speeddown))
+LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2])
 LAD.cluster.diffUp <- clusters %>%
@@
   summarise(diffDown.LADs = n())
+
+#write csvs to outputs
+out.clustersLAD <- paste(path, "./data/temp/LAD.cluster.tests.csv", sep = "")
+write_csv(LAD.cluster.tests, out.clustersLAD)
+out.clustersLADup <- paste(path, "./data/temp/LAD.cluster.up.csv", sep = "")
+write_csv(LAD.cluster.up, out.clustersLADup)
+out.clustersLADdown <- paste(path, "./data/temp/LAD.cluster.down.csv", sep = "")
+write_csv(LAD.cluster.down, out.clustersLADdown)
 

### 2020-07-15 src/descriptive_Clusters.Rmd
descriptive_Clusters.Rmd now runs

--- 2020-07-14 descriptive_Clusters.Rmd
@@
 
-LAD.cluster <- paste(path, "./data/temp/clusters.csv", sep = "")
+LAD.cluster <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
 
@@
 LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2])
-LAD.cluster.diffUp <- clusters %>%
-  group_by(cluster.up.diff) %>%
-  summarise(diffUp.LADs = n())
-LAD.cluster.diffDown <- clusters %>%
-  group_by(cluster.down.diff) %>%
-  summarise(diffDown.LADs = n())
+
+# commented to run the script
+# LAD.cluster.diffUp <- clusters %>%
+#   group_by(cluster.up.diff) %>%
+#   summarise(diffUp.LADs = n())
+# LAD.cluster.diffDown <- clusters %>%
+#   group_by(cluster.down.diff) %>%
+#   summarise(diffDown.LADs = n())
 
@@
 ```
-
-## R Markdown
-
-This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
-
-When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
-You can embed an R code chunk like this:
 
@@
 ```
-
-## Including Plots
-
-You can also embed plots, for example:
-
-```{r pressure, echo=FALSE}
-plot(pressure)
-```
-
-Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

### 2020-07-15 src/ts_clusters.Rmd
descriptive_Clusters.Rmd now runs

@@
 
-# clusters data frame
+# all clusters data frame
 clusters <- merge(cluster_up, cluster_down, by = "LAD")
@@
 # save clusters file
-out.clusters <- paste(path, "./data/temp/clusters.csv", sep = "")
+out.clusters <- paste(path, "/data/temp/clusters_all.csv", sep = "")
 write_csv(clusters, out.clusters)
+
+# clusters data frame WITHOUT the diff clusters
+clusters.nodiff <- merge(cluster_up, cluster_down, by = "LAD")
+names(clusters.nodiff)[2] <- "cluster.up"
+names(clusters.nodiff)[3] <- "cluster.down"
+
+clusters.nodiff <- merge(clusters.nodiff, cluster_tests, by = "LAD")
+names(clusters.nodiff)[4] <- "cluster.tests"
+#sapply(clusters, function(x) sum(is.na(x)))
+
+clusters.nodiff  %>% arrange(cluster.up) %>% kable()
+
+# save clusters file
+out.clusters <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
+write_csv(clusters.nodiff, out.clusters)
 

### 2020-07-29 src/descriptive_Clusters.Rmd

@@
 TS2020 <- selectByDate(TSbb19_20sp, year = 2020) 
+TS2019 <- selectByDate(TSbb19_20sp, year = 2019)
 
@@
 TS2020 <- left_join(TS2020, clusters)
+TS2019 <- left_join(TS2019, clusters)
 #remove NAs? - otherwise cannot create unified dataframes for below stats
@@
   summarise(n.cluster = n()) 
-LAD.cluster.tests <- left_join(LAD.cluster.tests, LAD.cluster.testsTotal)
+LAD.cluster.tests19 <- TS2019 %>%
+  group_by(cluster.tests) %>%
+  summarise(n.cluster19 = n()) 
+LAD.cluster.tests <- cbind(LAD.cluster.tests, LAD.cluster.testsTotal[,2], LAD.cluster.tests19[,2])
+LAD.cluster.tests$n.20_19.diff <- LAD.cluster.tests$n.cluster - LAD.cluster.tests$n.cluster19
 LAD.cluster.up <- clusters %>%
@@
   sd(speedup))
-LAD.cluster.up <- cbind(LAD.cluster.up, LAD.cluster.upMean[,2], LAD.cluster.upSD[,2])
+LAD.cluster.upAM <- summarise(group_by(TS2020[which(TS2020$hour == 10 | TS2020$hour == 11),], cluster.up), mean(speedup))
+LAD.cluster.upPM <- summarise(group_by(TS2020[which(TS2020$hour == 19 | TS2020$hour == 20),], cluster.up), mean(speedup))
+LAD.cluster.up <- cbind(LAD.cluster.up, LAD.cluster.upMean[,2], LAD.cluster.upSD[,2], LAD.cluster.upAM[,2], LAD.cluster.upPM[,2])
 LAD.cluster.down <- clusters %>%
@@
   sd(speeddown))
-LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2])
+LAD.cluster.downAM <- summarise(group_by(TS2020[which(TS2020$hour == 10 | TS2020$hour == 11),], cluster.down), mean(speeddown))
+LAD.cluster.downPM <- summarise(group_by(TS2020[which(TS2020$hour == 19 | TS2020$hour == 20),], cluster.down), mean(speeddown))
+LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2], LAD.cluster.downAM[,2], LAD.cluster.downAM[,2])
 
@@
                     labels = "%a"), labels = 
-                     c("Mon", "Tues", "Weds", "Thurs", "Fri"))
+                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + geom_point()
 

### 2020-07-31 src/descriptive_Clusters.Rmd

@@
 #summarise by LADs per cluster and Total tests / Mean speed per cluster
+#Test Frequency cluster
 LAD.cluster.tests <- clusters %>%
@@
 LAD.cluster.tests$n.20_19.diff <- LAD.cluster.tests$n.cluster - LAD.cluster.tests$n.cluster19
+#Upload Speed Clusters
 LAD.cluster.up <- clusters %>%
@@
 LAD.cluster.up <- cbind(LAD.cluster.up, LAD.cluster.upMean[,2], LAD.cluster.upSD[,2], LAD.cluster.upAM[,2], LAD.cluster.upPM[,2])
+names(LAD.cluster.up) [3:6] <- c("upMean", "upSD", "upAMmean",
+                                 "upPMmean")
+#Download speed clusters  
 LAD.cluster.down <- clusters %>%
@@
 LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2], LAD.cluster.downAM[,2], LAD.cluster.downAM[,2])
-
+names(LAD.cluster.down)[3:6] <- c("downMean", "downSD", "downAMmean",
+                                  "downPMmean")
 # commented to run the script

### 2020-07-31 src/descriptive_Clusters.Rmd
AM typo

--- 2020-07-29 src/descriptive_Clusters.Rmd
@@
 LAD.cluster.downPM <- summarise(group_by(TS2020[which(TS2020$hour == 19 | TS2020$hour == 20),], cluster.down), mean(speeddown))
-LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2], LAD.cluster.downAM[,2], LAD.cluster.downAM[,2])
-
+LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2], LAD.cluster.downAM[,2], LAD.cluster.downPM[,2])
+names(LAD.cluster.down)[3:6] <- c("downMean", "downSD", "downAMmean",
+                                  "downPMmean")
 # commented to run the script

### 2020-08-05 src/descriptive_Clusters.Rmd
add 2019 tests graph, change AM hours for LAD descriptives.

--- 2020-07-29 src/descriptive_Clusters.Rmd
@@
 LAD.cluster.tests$n.20_19.diff <- LAD.cluster.tests$n.cluster - LAD.cluster.tests$n.cluster19
+LAD.cluster.tests$perCent.change <- LAD.cluster.tests$n.20_19.diff/
+  LAD.cluster.tests$n.cluster19
+LAD.cluster.tests$perCent.change <- paste(round(100*LAD.cluster.tests$perCent.change, 2), "%", sep="")
+#Upload Speed Clusters
 LAD.cluster.up <- clusters %>%
@@
   sd(speedup))
-LAD.cluster.upAM <- summarise(group_by(TS2020[which(TS2020$hour == 10 | TS2020$hour == 11),], cluster.up), mean(speedup))
+LAD.cluster.upAM <- summarise(group_by(TS2020[which(TS2020$hour == 9 | TS2020$hour == 10),], cluster.up), mean(speedup))
 LAD.cluster.upPM <- summarise(group_by(TS2020[which(TS2020$hour == 19 | TS2020$hour == 20),], cluster.up), mean(speedup))
 LAD.cluster.up <- cbind(LAD.cluster.up, LAD.cluster.upMean[,2], LAD.cluster.upSD[,2], LAD.cluster.upAM[,2], LAD.cluster.upPM[,2])
+names(LAD.cluster.up) [3:6] <- c("upMean", "upSD", "upAMmean",
+                                 "upPMmean")
+#Download speed clusters  
 LAD.cluster.down <- clusters %>%
@@
   sd(speeddown))
-LAD.cluster.downAM <- summarise(group_by(TS2020[which(TS2020$hour == 10 | TS2020$hour == 11),], cluster.down), mean(speeddown))
+LAD.cluster.downAM <- summarise(group_by(TS2020[which(TS2020$hour == 9 | TS2020$hour == 10),], cluster.down), mean(speeddown))
 LAD.cluster.downPM <- summarise(group_by(TS2020[which(TS2020$hour == 19 | TS2020$hour == 20),], cluster.down), mean(speeddown))
-LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2], LAD.cluster.downAM[,2], LAD.cluster.downAM[,2])
-
+LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2], LAD.cluster.downAM[,2], LAD.cluster.downPM[,2])
+names(LAD.cluster.down)[3:6] <- c("downMean", "downSD", "downAMmean",
+                                  "downPMmean")
 # commented to run the script
@@
 ```
+
+As people came to rely more on their broadband services, it is unsurprising that the rate of testing increased in 2020 compared to 2019.
+This increase was greatest in cluster 2, where 56% more tests were run compared to the same period in 2019.
+Cluster 2 also has the most local authorities, 148, so the sum of tests run in all these authorities is much higher than in all the other clusters, both in 2019 and 2020.
+Cluster 1 has the second most local authorities, with 93, but the frequency of testing in these authorities is much lower in 2019 and 2020 and there was the lowest rate of increase between the two time periods at just 33%.
+In comparison, cluster 5 has the fewest local authorities, at 28, but the second highest total number of tests and the second largest increase in testing between 2019 and 2020, at 52%.
+A possible explanation is that cluster 5 contains some of the most populous local authorities in the country, including the cities of Birmingham, Manchester, Leeds, Sheffield, Glasgow and Edinburgh.
+However, this cluster also includes large rural counties, such as Cornwall and Northumberland.
+Yet the temporal profile in 2020 is very different to 2019, as can be seen in graphs [.] and [.].
+The number of broadband speed tests run in local authorities in cluster 2 peaked each Monday to Thursday in the hour 19.00-19.59 in March through May 2019.
+On Fridays, the peak was slightly earlier.
+A similar, although less stark pattern could be seen in the other clusters, along with much smaller peaks around the beginning of the working day, at either 9.00-9.59 or 10.00-10.59.
 
@@
                      c("Mon", "Tues", "Weds", "Thurs", "Fri")) + geom_point()
-
+#summarise test frequency clusters by hour of each weekday 2019 and plot
+cluster.tests19 <- TS2019 %>%
+  group_by(cluster.tests, weekday, hour) %>%
+  summarise(n.tests = n())
+names(cluster.tests19)[1] <- "clusterID"
+cluster.tests19$clusterID <- as.factor(cluster.tests19$clusterID)
+#add fake dates for plotting
+help.dates <- tibble(help.date.time = c(
+  make_datetime(year = 2019, month = 3, day = 4),
+  make_datetime(year = 2019, month = 3, day = 5),
+  make_datetime(year = 2019, month = 3, day = 6),
+  make_datetime(year = 2019, month = 3, day = 7),
+  make_datetime(year = 2019, month = 3, day = 8)),
+                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+cluster.tests19 <- merge(cluster.tests19, help.dates, by = "weekday")
+cluster.tests19$hour <- paste(cluster.tests19$hour, ":00", sep="")
+cluster.tests19$help.date.time <- with(cluster.tests19, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+#plot cluster statistic for number tests
+TestCluster19 <- ggplot(cluster.tests19, aes(help.date.time, n.tests, 
+                                         colour = clusterID))
+TestCluster19 + geom_line() + labs(x = "Weekdays March-May 2019") + 
+  scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
+                    "2020-03-03 15:00:00", "2020-03-04 15:00:00",
+                    "2020-03-05 15:00:00", "2020-03-06 15:00:00"), 
+                    labels = "%a"), labels = 
+                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + geom_point()
 #summarise upload speed clusters by hour of each weekday and plot

### 2020-08-10 src/Data_Spatial.Rmd
minor change on comments
knitted descriptive_clusters.Rmd

@@
 TSbb19_20sp <- TSbb19_20sp[!is.na(TSbb19_20sp$LAD19NM),]
-#NA LAD19NM, coord outside UK =  615516 obs (99.3% of obs retained)
+#NA LAD19NM, coord outside UK =  2618 / 405568 obs (99.4% of obs retained)
 
 #save files
-out.TSbb19_20sp <- paste(path, "./data/temp/TSbb19_20sp.csv", sep = "")
+out.TSbb19_20sp <- paste(path, "/data/temp/TSbb19_20sp.csv", sep = "")
 write_csv(TSbb19_20sp, out.TSbb19_20sp)

### 2020-08-10 src/ts_clusters.Rmd
knitted descriptive_clusters.Rmd

@@
 library(leaflet)
+library(ggmap)
 
@@
 
-clusters  %>% arrange(cluster.up) %>% kable()
+# clusters  %>% arrange(cluster.up) %>% kable()
 
@@
 
+# ggplot2 maps
+la.f <- fortify(la, region = "lad19nm")
+la.f <- merge(la.f, clusters.nodiff, by.x = "id", by.y = "LAD")
+la.f <- la.f[order(la.f$order),] # if i don't order merge.nuts.f loses order and the map has gaps
+
+ggplot(la.f, aes(x = long, y = lat)) +
+  geom_polygon(aes( group = group, fill = as.factor(cluster.tests))) +
+  theme_nothing(legend = TRUE) +
+  labs(title = "Test clusters") +
+  scale_fill_viridis_d()+
+  theme_void()
+
+ggplot(la.f, aes(x = long, y = lat)) +
+  geom_polygon(aes( group = group, fill = as.factor(cluster.up))) +
+  theme_nothing(legend = TRUE) +
+  labs(title = "Upload clusters") +
+  scale_fill_viridis_d()+
+  theme_void()
+
+ggplot(la.f, aes(x = long, y = lat)) +
+  geom_polygon(aes( group = group, fill = as.factor(cluster.down))) +
+  theme_nothing(legend = TRUE) +
+  labs(title = "Download clusters") +
+  scale_fill_viridis_d()+
+  theme_void()
+
+```
+
+```{r, eval=FALSE}
+# leaflet maps
+
+# !!!WRONG CATEGORY VISUALISATION!!!
+
 #la@data <- merge(la@data, clusters, by.x = "LAD19NM", by.y = "LAD")
-la@data <- merge(la@data, clusters, by.x = "lad19nm", by.y = "LAD")
+la@data <- merge(la@data, clusters.nodiff, by.x = "lad19nm", by.y = "LAD")
 
-# plot maps
-# var <- la@data[ ,'cluster.up']
-# plot(la, col = var, main = "Upload clusters",
-#      border = rgb(0.8, 0.8, 0.8, 0))
-# 
-# var <- la@data[ ,'cluster.up.diff']
-# plot(la, col = var, main = "Upload difference clusters",
-#      border = rgb(0.8, 0.8, 0.8, 0))
-# 
-# 
-# var <- la@data[ ,'cluster.down']
-# plot(la, col = var, main = "Download clusters",
-#      border = rgb(0.8, 0.8, 0.8, 0))
-# 
-# var <- la@data[ ,'cluster.tests']
-# plot(la, col = var, main = "Tests clusters",
-#      border = rgb(0.8, 0.8, 0.8, 0))
+# from handbook
+la@data$cluster.tests <- as.factor(la@data$cluster.tests)
+factpal <- colorFactor(topo.colors(5), la@data$cluster.tests)
 
-# leaflet maps
+leaflet(la) %>%
+  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
+    color = ~factpal(la@data$cluster.tests)) %>%
+  addMarkers(lng=la$long, lat=la$lat, popup=la@data$lad19nm) %>%
+  addLegend(pal = factpal, 
+            values = ~la@data$cluster.tests,
+            opacity = 1,
+            #group = "cluster.tests",
+            title = "Test clusters")
+
+leaflet(la) %>%
+  addPolygons(
+    fillColor = factpal(la$cluster.tests),
+    weight = 2,
+    opacity = 1,
+    color = "white",
+    dashArray = "3",
+    fillOpacity = 0.7) %>%
+  addMarkers(lng=la$long, lat=la$lat, popup=la@data$lad19nm)
+
+# circles
+leaflet(la@data) %>%
+  addTiles() %>%
+  addCircles(lng = ~long, lat = ~lat, weight = 1, 
+             radius = 500L, popup = ~lad19nm, color = ~factpal(cluster.tests))
 
@@
 la@data$cluster.down <- as.factor(la@data$cluster.down)
-la@data$cluster.up.diff <- as.factor(la@data$cluster.up.diff)
-la@data$cluster.down.diff <- as.factor(la@data$cluster.down.diff)
+#la@data$cluster.up.diff <- as.factor(la@data$cluster.up.diff)
+#la@data$cluster.down.diff <- as.factor(la@data$cluster.down.diff)
 la@data$cluster.tests <- as.factor(la@data$cluster.tests)
@@
 factpal.down <- colorFactor("RdYlBu", la@data$cluster.down)
-factpal.up.diff <- colorFactor("RdYlBu", la@data$cluster.up.diff)
-factpal.down.diff <- colorFactor("RdYlBu", la@data$cluster.down.diff)
+#factpal.up.diff <- colorFactor("RdYlBu", la@data$cluster.up.diff)
+#factpal.down.diff <- colorFactor("RdYlBu", la@data$cluster.down.diff)
 factpal.tests <- colorFactor("RdYlBu", la@data$cluster.tests)
@@
     color = ~factpal.down(cluster.down), group = "cluster.down") %>%
-  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
-    color = ~factpal.up.diff(cluster.up.diff), group = "cluster.up.diff") %>%
-  addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
-    color = ~factpal.down.diff(cluster.down.diff), group = "cluster.down.diff") %>%
+  #addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
+  #  color = ~factpal.up.diff(cluster.up.diff), group = "cluster.up.diff") %>%
+  #addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
+  #  color = ~factpal.down.diff(cluster.down.diff), group = "cluster.down.diff") %>%
   addPolygons(stroke = FALSE, smoothFactor = 0.2, fillOpacity = 1,
     color = ~factpal.tests(cluster.tests), group = "cluster.tests") %>%
+  addMarkers(lng=la$long, lat=la$lat, popup=la$lad19nm) %>%
   # hide groups, so they don't appear all on in the begining 
@@
   hideGroup("cluster.down") %>%
-  hideGroup("cluster.up.diff") %>%
-  hideGroup("cluster.down.diff") %>%
+  #hideGroup("cluster.up.diff") %>%
+  #hideGroup("cluster.down.diff") %>%
   hideGroup("cluster.tests") %>%
@@
     #baseGroups = c("Upload)", "Download"),
-    overlayGroups = c("cluster.up","cluster.down", "cluster.down", "cluster.up.diff", "cluster.down.diff", "cluster.tests"),
+    overlayGroups = c("cluster.up","cluster.down", "cluster.down", "cluster.tests"), #"cluster.up.diff", "cluster.down.diff", 
     options = layersControlOptions(collapsed = FALSE)) %>%
@@
             title = "Download clusters") %>%
-  addLegend(pal = factpal.up.diff, 
-            values = la@data$cluster.up.diff,
-            opacity = 1,
-            group = "cluster.up.diff",
-            title = "Upload diff. clusters") %>%
-  addLegend(pal = factpal.down.diff, 
-            values = la@data$cluster.down.diff,
-            opacity = 1,
-            group = "cluster.down.diff",
-            title = "Download diff. clusters") %>%
+  #addLegend(pal = factpal.up.diff, 
+  #          values = la@data$cluster.up.diff,
+  #          opacity = 1,
+  #          group = "cluster.up.diff",
+  #          title = "Upload diff. clusters") %>%
+  #addLegend(pal = factpal.down.diff, 
+  #          values = la@data$cluster.down.diff,
+  #          opacity = 1,
+  #          group = "cluster.down.diff",
+  #          title = "Download diff. clusters") %>%
   addLegend(pal = factpal.tests, 

### 2020-08-18 src/descriptive_Clusters.Rmd
Text on clusters.

@@
 A similar, although less stark pattern could be seen in the other clusters, along with much smaller peaks around the beginning of the working day, at either 9.00-9.59 or 10.00-10.59.
+In comparison, the daily profile of broadband speed testing in 2020 is flatter, with less variation hour to hour.
+Although there is still an identifiable evening peak for almost all days of the week and all clusters, it is much less pronounced throughout, there is less of a gap between the number of tests in the morning peak and the evening peak, and there is less of a dip in testing in the middle of the day.
+All this suggests that more of the additional testing generated in 2020 compared to 2019 is occurring outside the traditional peak in the evening at streaming primetime.
+Moreover, this is particularly noticeable for cluster 2, but the change is less pronounced for cluster 5, perhaps because of the diversity of urban and rural populations represented in cluster 5.
+The graphs also show that Cluster 1, with the lowest increase in testing, falls from the middle position in terms of total number of tests to the second from bottom.
+For upload speeds, 345 of 382 local authorities fall into cluster 6 or cluster 9.
+Graph [.] shows that both of these clusters have relatively similar temporal profiles, with relatively stable speeds.
+However, the upload speeds at all times for cluster 9 are substantially higher than for cluster 6.
+This difference is reflected not only in the mean speeds for these clusters for the whole sample, but also the mean upload speeds in evening peak period from 19:00-20:59 and the morning peak from 9:00-10:59.
+Which sort of areas are benefitting from these higher speeds?
+The 115 Local Authority areas are mostly in urban or suburban areas, including 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation.
+There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, as well as many suburban districts from the South East of England to South Tyneside.
+On the other hand, the 230 local authorities in cluster 6 still include urban areas, including Bristol, Liverpool and Leeds, and many suburban areas.
+However, it also includes some of the most rural areas in the country, and could be more seen as a typical level of service for upload speeds, rather than being particularly slow.
+The differences between morning and evening upload speeds for both clusters are minimal, at 1% slower in the morning in cluster 9 and 4% slower in cluster 6.
+This is an indication of the reliability of broadband services in places in both clusters, and suggests that upload speeds have generally been robust throughout the study period.
+The exceptions can be found in seven of the eight much smaller clusters of between 1 and 10 local authority districts, where the standard deviation from mean speeds is between 3% and 14% greater than in cluster 6.
+The temporal profiles on Graph [.] for the three of those clusters with more than 5 local authorities also shows that speeds are far from dependable in these clusters.
+Furthermore, AM peak upload speeds are between 6% and 18% slower than PM peak upload speeds in the 35 local authorities in these seven clusters, although the mean speeds for each cluster are higher than cluster 6.
+Included in this eclectic group are central London borough of Westminster and the remote Isles of Scilly, and even the largest of these clusters includes a Surrey district on the outskirts of London, the West Wales authority of Ceredigion, and the northern market town of Darlington.
+Download speeds are the most commonly used measure of the quality and reliability of broadband services.
+They also vary the most, with standard deviations in speed tending to be in a similar range to the mean speeds for each cluster, compared to upload speeds, where standard deviations are half to two thirds the level of the mean speeds.
+This temporal variation can be seen in Graph [.], which shows the 10 of 20 clusters with over five local authorities in them.
+The largest of these 10 clusters, number 19 with 123 Local Authorities, as well as cluster 1 with 16 Local Authorities have lower mean speeds than the other clusters, and this is consistent throughout the temporal profile on the graph.
+However, this means they also have lower standard deviations, without such extreme peaks and troughs as, for example, clusters 7, 9 and 11, with 36 Local Authorities.
+What is of greater interest, however, is where download speeds are slower in the morning peak when broadband is more likely to be used for work purposes than in the evening peak.
+This difference is most stark in some of the smaller clusters, although since there are only 11 Local Authorities in these clusters even added together, their inclusion in these clusters may be more a reflection of quite localised issues.
+The largest two clusters of Local Authorities have 1-3% slower speeds in the AM peak, and a few clusters have faster download speeds in the AM peak.
+However, there are still 58 Local Authorities, including the 11 mentioned previously with download speeds at least 6% lower in the morning peak than in the evening peak.
+Is the reduction in download speeds in the morning peak in more places than the reduction in upload speeds a result of ISP management to support reliable services for telecommuters?
+There are only eight Local Authorities where both upload and download speeds are more than 5% slower in the AM peak than the PM peak.
 
@@
                                          colour = clusterID))
-TestCluster + geom_line() + labs(x = "Weekdays March-May") + 
+TestCluster + geom_line() + labs(x = "Weekdays March-May 2020") + 
   scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 

### 2020-08-18 src/LAs_Clusters.Rmd
multinomial regressions

---
title: "LAs_clusters"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output: 
  html_document:
    df_print: paged
    toc: true
    toc_float: FALSE
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../output")
  })
---

```{r settings, include = FALSE}
library(ggplot2)
library(tidyverse)
library(rprojroot)
library(lubridate)
library(knitr)
library(rprojroot)
library(nnet)
library(stargazer)
library(openair)
library(rgdal)
library(corrplot)
library(DescTools)

# This is the project path
path <- find_rstudio_root_file()
```

## Load cluster data

```{r include = FALSE}

# load the data created by Data_Spatial.Rmd and ts_clusters.Rmd
path.cluster <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
clusters <- read_csv(path.cluster)
names(clusters) [1] <- "LAD19NM"

path.data <- paste(path, "/data/temp/TSbb19_20sp.csv", sep = "")
TSbb19_20sp <- read_csv(path.data)

#split out 2020 before join to clusters to create descriptives
TS2020 <- selectByDate(TSbb19_20sp, year = 2020) 
TS2019 <- selectByDate(TSbb19_20sp, year = 2019)

# LAD codes and names
# get LA 
# this is the heaviest version:
# (i) directly from the web
#la <- readOGR("http://geoportal1-ons.opendata.arcgis.com/datasets/b6d2e15801de45328b760a4f55d74318_0.geojson?outSR={%22latestWkid%22:3857,%22wkid%22:102100}")#, layer="OGRGeoJSON")
# or, (ii) from the json I saved locally
# path.json <- paste(path, "./data/raw/Local_Authority_Districts_(April_2019)_Boundaries_UK_BFE.geojson", sep = "")
# la <- readOGR(path.json)#, layer="OGRGeoJSON")
# UK BGC: 
la <- readOGR("https://opendata.arcgis.com/datasets/0e07a8196454415eab18c40a54dfbbef_0.geojson")
codes <- la@data %>%
  select(lad19cd, lad19nm) %>%
  rename(LAD19NM = lad19nm)

# LAD broadband characteristics AM
LAD.up.am19 <- TS2019 %>%
  filter(hour == 9 | hour == 10) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.up.am19 = mean(speedup))

LAD.up.am20 <- TS2020 %>%
  filter(hour == 9 | hour == 10) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.up.am20 = mean(speedup))

LAD.down.am19 <- TS2019 %>%
  filter(hour == 9 | hour == 10) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.down.am19 = mean(speeddown))

LAD.down.am20 <- TS2020 %>%
  filter(hour == 9 | hour == 10) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.down.am20 = mean(speeddown))

LAD.tests.am19 <- TS2019 %>%
  filter(hour == 9 | hour == 10) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.tests.am19 = n())

LAD.tests.am20 <- TS2020 %>%
  filter(hour == 9 | hour == 10) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.tests.am20 = n())

# LAD broadband characteristics PM
LAD.up.pm19 <- TS2019 %>%
  filter(hour == 19 | hour == 20) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.up.pm19 = mean(speedup))

LAD.up.pm20 <- TS2020 %>%
  filter(hour == 19 | hour == 20) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.up.pm20 = mean(speedup))

LAD.down.pm19 <- TS2019 %>%
  filter(hour == 19 | hour == 20) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.down.pm19 = mean(speeddown))

LAD.down.pm20 <- TS2020 %>%
  filter(hour == 19 | hour == 20) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.down.pm20 = mean(speeddown))

LAD.tests.pm19 <- TS2019 %>%
  filter(hour == 19 | hour == 20) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.tests.pm19 = n())

LAD.tests.pm20 <- TS2020 %>%
  filter(hour == 19 | hour == 20) %>%
  group_by(LAD19NM) %>%
  summarise(LAD.tests.pm20 = n())

# LAD characteristcs merge
lad <- list(codes,
            clusters,
            LAD.up.am19,
            LAD.up.am20,
            LAD.down.am19,
            LAD.down.am20,
            LAD.tests.am19,
            LAD.tests.am20,
            LAD.up.pm19,
            LAD.up.pm20,
            LAD.down.pm19,
            LAD.down.pm20,
            LAD.tests.pm19,
            LAD.tests.pm20)

sapply(lad, function(x) dim(x))
sapply(lad, function(x) sum(is.na(x)))

# merge with reduce
lad <- lad %>% 
  reduce(inner_join, by = "LAD19NM")
sapply(lad, function(x) sum(is.na(x)))
```

## Job density

```{r}
job.dens <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_57_1.data.csv?geography=1816133633...1816133848,1820327937...1820328318&date=latest&item=3&measures=20100")

job.dens <- job.dens %>%
  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, OBS_VALUE) %>%
  rename(job.dens2018 = OBS_VALUE) %>%
  rename(lad19cd = GEOGRAPHY_CODE) %>%
  distinct()

# test <- left_join(lad, job.dens, by = "lad19cd")
```

## Population

```{r}
pop <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_31_1.data.csv?geography=1807745025...1807745028,1807745030...1807745032,1807745034...1807745155,1807745157...1807745164,1807745166...1807745170,1807745172...1807745177,1807745179...1807745194,1807745196,1807745197,1807745199,1807745201...1807745218,1807745221,1807745222,1807745224,1807745226...1807745231,1807745233,1807745234,1807745236...1807745244,1807745271...1807745281,1811939329...1811939332,1811939334...1811939336,1811939338...1811939497,1811939499...1811939501,1811939503,1811939505...1811939507,1811939509...1811939517,1811939519,1811939520,1811939524...1811939570,1811939575...1811939599,1811939601...1811939628,1811939630...1811939634,1811939636...1811939647,1811939649,1811939655...1811939664,1811939667...1811939680,1811939682,1811939683,1811939685,1811939687...1811939704,1811939707,1811939708,1811939710,1811939712...1811939717,1811939719,1811939720,1811939722...1811939730,1811939757...1811939767&date=latestMINUS1-latest&sex=7&age=0,22&measures=20100")

pop <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_31_1.data.csv?geography=1820327937...1820328318,1816133633...1816133848&date=latestMINUS1-latest&sex=7&age=0,22&measures=20100")

pop <- pop %>%
  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, AGE_NAME, OBS_VALUE) %>%
  distinct(GEOGRAPHY_CODE, AGE_NAME, .keep_all = T) %>% 
  spread(AGE_NAME, OBS_VALUE) %>% 
  rename(pop16_64 = `Aged 16 - 64`) %>%
  rename(pop = `All ages`) %>%
  rename(lad19cd = GEOGRAPHY_CODE)

# test <- left_join(lad, pop, by = "lad19cd")
# sapply(test, function(x) sum(is.na(x)))
```

## Labour supply

```{r}
labour <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_17_5.data.csv?geography=1816133633...1816133837,1820327937...1820328307&date=latest&variable=18,1532...1540,290,720&measures=20599,21001,21002,21003")

labour <- labour %>%
  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, VARIABLE_NAME, OBS_VALUE) %>%
  distinct(GEOGRAPHY_CODE, VARIABLE_NAME, .keep_all = T) %>% 
  spread(VARIABLE_NAME, OBS_VALUE) %>% 
  rename(lad19cd = GEOGRAPHY_CODE) %>%
  rename(managers =  `% all in employment who are - 1: managers, directors and senior officials (SOC2010)`) %>%
  rename(prof = `% all in employment who are - 2: professional occupations (SOC2010)`) %>%
  rename(tech = `% all in employment who are - 3: associate prof & tech occupations (SOC2010)`) %>%
  rename(admin = `% all in employment who are - 4: administrative and secretarial occupations (SOC2010)`) %>%
  rename(skilled = `% all in employment who are - 5: skilled trades occupations (SOC2010)`) %>%
  rename(caring = `% all in employment who are - 6: caring, leisure and other service occupations (SOC2010)`) %>%
  rename(sales = `% all in employment who are - 7: sales and customer service occupations (SOC2010)`) %>%
  rename(plant = `% all in employment who are - 8: process, plant and machine operatives (SOC2010)`) %>%
  rename(elementary = `% all in employment who are - 9: elementary occupations (SOC2010)`) %>%
  rename(NVQ3 = `% with NVQ3+ - aged 16-64`) %>%
  rename(NVQ4 = `% with NVQ4+ - aged 16-64`) %>%
  rename(econ.act = `Economic activity rate - aged 16-64`)

# test <- left_join(lad, labour, by = "lad19cd")
# sapply(test, function(x) sum(is.na(x)))
```

## Earnings

```{r}
earnings <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_30_1.data.csv?geography=1946157348,2092957698,2013265929,1816133633...1816133837,1820327937...1820328307&date=latest&sex=8&item=2&pay=1&measures=20100,20701")

earnings <- earnings %>%
  filter(MEASURES_NAME == "Value") %>% # dropiing the confedence
  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, OBS_VALUE) %>%
  rename(lad19cd = GEOGRAPHY_CODE) %>%
  rename(earnings =  OBS_VALUE) %>%
  distinct()

# test <- left_join(lad, earnings, by = "lad19cd")
# sapply(test, function(x) sum(is.na(x)))
```

## Business counts

```{r}
# busi <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_141_1.data.csv?geography=1816133633...1816133848,1820327937...1820328318,1870659585...1870659791,1870659801,1870659792...1870659800,1879048193...1879048573,1879048583,1879048574...1879048582&date=latest&industry=163577857...163577874&employment_sizeband=0&legal_status=0&measures=20100")

busi <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_141_1.data.csv?geography=1816133633...1816133848,1820327937...1820328318,1870659585...1870659791,1870659801,1870659792...1870659800,1879048193...1879048573,1879048583,1879048574...1879048582&date=latest&industry=150994945...150994965&employment_sizeband=0&legal_status=0&measures=20100")

busi <- busi %>%
  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, INDUSTRY_NAME, OBS_VALUE) %>%
  distinct(GEOGRAPHY_CODE, INDUSTRY_NAME, .keep_all = T) %>% 
  spread(INDUSTRY_NAME, OBS_VALUE) %>% 
  rename(lad19cd = GEOGRAPHY_CODE) %>%
  rename(A = `A : Agriculture, forestry and fishing`) %>%
  rename(B = `B : Mining and quarrying`) %>%
  rename(C = `C : Manufacturing`) %>%
  rename(D = `D : Electricity, gas, steam and air conditioning supply`) %>%
  rename(E = `E : Water supply; sewerage, waste management and remediation activities`) %>%
  rename(F = `F : Construction`) %>%
  rename(G = `G : Wholesale and retail trade; repair of motor vehicles and motorcycles`) %>%
  rename(H = `H : Transportation and storage`) %>%
  rename(I = `I : Accommodation and food service activities`) %>%
  rename(J = `J : Information and communication`) %>%
  rename(K = `K : Financial and insurance activities`) %>%
  rename(L = `L : Real estate activities`) %>%
  rename(M = `M : Professional, scientific and technical activities`) %>%
  rename(N = `N : Administrative and support service activities`) %>%
  rename(O = `O : Public administration and defence; compulsory social security`) %>%
  rename(P = `P : Education`) %>%
  rename(Q = `Q : Human health and social work activities`) %>%
  rename(R = `R : Arts, entertainment and recreation`) %>%
  rename(S = `S : Other service activities`) %>%
  rename(T = `T : Activities of households as employers;undifferentiated goods-and services-producing activities of households for own use`) %>%
  rename(U = `U : Activities of extraterritorial organisations and bodies`) %>%
  mutate(total.busi = rowSums(.[3:23]))

# test <- left_join(lad, busi, by = "lad19cd")
# sapply(test, function(x) sum(is.na(x)))
```

## merge

```{r}
data <- list(lad, job.dens, pop, labour, earnings, busi)
sapply(data, function(x) dim(x))
sapply(data, function(x) names(x))

# merge with reduce
data <- data %>% 
  reduce(inner_join, by = "lad19cd") %>%
  #distinct(.keep_all = T)
  unique()
sapply(data, function(x) sum(is.na(x)))
```

## Descriptive statistics and correlations

There is surprisingly low correlation between upload and download, morning and night. 
Number of tests are the most correlated variables.

```{r correlations}
data.cor <- data %>%
  dplyr::select(LAD.up.am19, LAD.down.am19,  
                LAD.tests.am19, LAD.up.pm19, 
                LAD.down.pm19, LAD.tests.pm19, 
                LAD.up.am20, LAD.down.am20,
                LAD.tests.am20, LAD.up.pm20, 
                LAD.down.pm20, LAD.tests.pm20)

m <- cor(data.cor)
corrplot(m, type="upper",method = "number", number.cex = .5, tl.cex = .75)

data.cor <- data %>%
  dplyr::select(LAD.up.am20, LAD.up.pm20,
                LAD.tests.pm20,
                pop, pop16_64,
                managers, tech,
                skilled, earnings,
                total.busi)

m <- cor(data.cor, use = "pairwise.complete.obs")
corrplot(m, type="upper",method = "number", number.cex = .5, tl.cex = .75)
```

## Auxiliary Regression

```{r regressions, results= 'asis', message=F, error=F, warning=F}
# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/

aux <- multinom(cluster.up ~ LAD.up.am20 + LAD.up.pm20 + LAD.tests.pm20 + #  LAD.down.am20 + LAD.down.pm20
                   pop + pop16_64 + # how about young population share
                   managers + tech + skilled +
                   earnings + total.busi, data = data, 
                trace = F) # to avoid messages

# McFadden's R squared
aux.null <- multinom(cluster.up ~ 1, data = data)
r2 <- 1-logLik(aux)/logLik(aux.null)

stargazer(aux, type = "html",
          add.lines = list(c("McFadden's R squared",                    # use rownames as line name. in this case Weak insterumetns
                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
                              round(r2[1],3), round(r2[1],3), round(r2[1],3))))

```

## Auxiliary Regression without upload variables

```{r regressions no upload and test, results= 'asis', message=F, error=F, warning=F}
# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/

aux <- multinom(cluster.up ~ pop + pop16_64 + # how about young population share
                  managers + tech + skilled +
                  earnings + total.busi, data = data, 
                trace = F) # to avoid messages

# McFadden's R squared
aux.null <- multinom(cluster.up ~ 1, data = data)
r2 <- 1-logLik(aux)/logLik(aux.null)

stargazer(aux, type = "html",
          add.lines = list(c("McFadden's R squared",                    # use rownames as line name. in this case Weak insterumetns
                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
                              round(r2[1],3), round(r2[1],3), round(r2[1],3))))

```

### 2020-08-18 src/LAs_Clusters.Rmd
now with a download regression too

@@
 
-## Auxiliary Regression
+## Auxiliary regression, dep. var.: upload clusters
 
-```{r regressions, results= 'asis', message=F, error=F, warning=F}
+```{r upload regression, results= 'asis', message=F, error=F, warning=F}
 # helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
@@
 
-## Auxiliary Regression without upload variables
+## Auxiliary regression, dep. var.: upload clusters, without upload variables
 
-```{r regressions no upload and test, results= 'asis', message=F, error=F, warning=F}
+```{r upload regression no upload and test, results= 'asis', message=F, error=F, warning=F}
 # helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
@@
 ```
+
+## Auxiliary regression, dep. var.: download clusters
+
+```{r download regression, results= 'asis', message=F, error=F, warning=F}
+# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
+# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/
+
+aux <- multinom(cluster.down ~ LAD.down.am20 + LAD.down.pm20 + LAD.tests.pm20 + #  LAD.down.am20 + LAD.down.pm20
+                   pop + pop16_64 + # how about young population share
+                   managers + tech + skilled +
+                   earnings + total.busi, data = data, 
+                trace = F) # to avoid messages
+
+# McFadden's R squared
+aux.null <- multinom(cluster.down ~ 1, data = data)
+r2 <- 1-logLik(aux)/logLik(aux.null)
+
+stargazer(aux, type = "html",
+          add.lines = list(c("McFadden's R squared",                    # use rownames as line name. in this case Weak insterumetns
+                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                              round(r2[1],3))))
+
+```

### 2020-08-26 src/LAs_Clusters.Rmd

@@
 library(DescTools)
+library(geosphere)
 
@@
 sapply(lad, function(x) sum(is.na(x)))
+```
+
+## Distance to Metro Area
+
+```{r}
+# get LA 
+# (i) directly from the web
+#la <- readOGR("http://geoportal1-ons.opendata.arcgis.com/datasets/b6d2e15801de45328b760a4f55d74318_0.geojson?outSR={%22latestWkid%22:3857,%22wkid%22:102100}")#, layer="OGRGeoJSON")
+# or, (ii) from the json I saved locally
+path.json <- paste(path, "./data/raw/Local_Authority_Districts_(April_2019)_Boundaries_UK_BFE.geojson", sep = "")
+la <- readOGR(path.json)#, layer="OGRGeoJSON")
+# source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe
+
+# spatial transformations
+la <- spTransform(la, CRS("+init=epsg:4326"))
+la@data$LAD19NM <- as.character(la@data$LAD19NM)
+ladDist <- la@data[,c("LAD19CD", "LAD19NM")]
+names(ladDist)[1] <- "lad19cd"
+
+#distance from London and other 
+coords_LAD <- la@data[,c("LONG", "LAT")]
+London <- la[which(la@data$LAD19NM == "City of London"),]
+London <- gCentroid(London)
+distLondon <- distm(coords_LAD, London, fun=distCosine) 
+#result in meters - change to km
+distLondon <- round((distLondon/1000),2)
+Birmingham <- la[which(la@data$LAD19NM == "Birmingham"),]
+Birmingham <- gCentroid(Birmingham)
+distBham <- distm(coords_LAD, Birmingham, fun=distCosine)
+distBham <- round((distBham/1000),2)
+Manchester <- la[which(la@data$LAD19NM == "Manchester"),]
+Manchester <- gCentroid(Manchester)
+distManc <- distm(coords_LAD, Manchester, fun=distCosine)
+distManc <- round((distManc/1000),2)
+Leeds <- la[which(la@data$LAD19NM == "Leeds"),]
+Leeds <- gCentroid(Leeds)
+distLeeds <- distm(coords_LAD, Leeds, fun=distCosine)
+distLeeds <- round((distLeeds/1000),2)
+Liverpool <- la[which(la@data$LAD19NM == "Liverpool"),]
+Liverpool <- gCentroid(Liverpool)
+distLivp <- distm(coords_LAD, Liverpool, fun=distCosine)
+distLivp <- round((distLivp/1000),2)
+Newcastle <- la[which(la@data$LAD19NM == "Newcastle upon Tyne"),]
+Newcastle <- gCentroid(Newcastle)
+distNewc <- distm(coords_LAD, Newcastle, fun=distCosine)
+distNewc <- round((distNewc/1000),2)
+Sheffield <- la[which(la@data$LAD19NM == "Sheffield"),]
+Sheffield <- gCentroid(Sheffield)
+distShef <- distm(coords_LAD, Sheffield, fun=distCosine)
+distShef <- round((distShef/1000),2)
+SouthHants <- la[which(la@data$LAD19NM == "Fareham"),]
+SouthHants <- gCentroid(SouthHants)
+distSHam <- distm(coords_LAD, SouthHants, fun=distCosine)
+distSHam <- round((distSHam/1000),2)
+Nottingham <- la[which(la@data$LAD19NM == "Nottingham"),]
+Nottingham <- gCentroid(Nottingham)
+distNotts <- distm(coords_LAD, Nottingham, fun=distCosine)
+distNotts <- round((distNotts/1000),2)
+Bristol <- la[which(la@data$LAD19NM == "Bristol, City of"),]
+Bristol <- gCentroid(Bristol)
+distBris <- distm(coords_LAD, Bristol, fun=distCosine)
+distBris <- round((distBris/1000),2)
+Cardiff <- la[which(la@data$LAD19NM == "Cardiff"),]
+Cardiff <- gCentroid(Cardiff)
+distCard <- distm(coords_LAD, Cardiff, fun=distCosine)
+distCard <- round((distCard/1000),2)
+Glasgow <- la[which(la@data$LAD19NM == "Glasgow City"),]
+Glasgow <- gCentroid(Glasgow)
+distGlas <- distm(coords_LAD, Glasgow, fun=distCosine) 
+distGlas <- round((distGlas/1000),2)
+
+ladDist <- cbind(ladDist, distLondon, distBham, distManc, distLeeds,
+                 distLivp, distNewc, distShef, distSHam, distNotts, 
+                 distBris, distCard, distGlas)
+
+ladDist$distMet <- apply(ladDist[,3:14], 1, min)
+
+#test <- left_join(lad, ladDist)
 ```

### 2020-08-26 src/descriptive_Clusters.Rmd
knit descriptive_Clusters.Rmd

@@
 
+#add LAD population data
+pop <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_31_1.data.csv?geography=1820327937...1820328318,1816133633...1816133848&date=latestMINUS1-latest&sex=7&age=0,22&measures=20100")
+
+pop <- pop %>%
+  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, AGE_NAME, OBS_VALUE) %>%
+  distinct(GEOGRAPHY_NAME, AGE_NAME, .keep_all = T) %>% 
+  spread(AGE_NAME, OBS_VALUE) %>% 
+  rename(pop16_64 = `Aged 16 - 64`) %>%
+  rename(pop = `All ages`) %>%
+  rename(LAD19NM = GEOGRAPHY_NAME)
+
+#join to clusters
+pop$LAD19NM <- if_else(pop$LAD19NM == "Rhondda Cynon Taff", "Rhondda Cynon Taf", pop$LAD19NM)
+clusters <- left_join(clusters, pop)
+
 #summarise by LADs per cluster and Total tests / Mean speed per cluster
@@
   summarise(n.LADs = n())
+LAD.cluster.pop <- clusters %>%
+  group_by(cluster.tests) %>%
+  summarise(pop.LADs = sum(pop))
 LAD.cluster.testsTotal <- TS2020 %>%
@@
   summarise(n.cluster19 = n()) 
-LAD.cluster.tests <- cbind(LAD.cluster.tests, LAD.cluster.testsTotal[,2], LAD.cluster.tests19[,2])
+LAD.cluster.tests <- cbind(LAD.cluster.tests, LAD.cluster.pop [2], LAD.cluster.testsTotal[,2], LAD.cluster.tests19[,2])
+LAD.cluster.tests$TestsPop20 <- LAD.cluster.tests$n.cluster / LAD.cluster.tests$pop.LADs
+LAD.cluster.tests$TestsPop19 <- LAD.cluster.tests$n.cluster19 / LAD.cluster.tests$pop.LADs
 LAD.cluster.tests$n.20_19.diff <- LAD.cluster.tests$n.cluster - LAD.cluster.tests$n.cluster19
@@
   summarise(up.LADs = n())
+LAD.cluster.upPop <- clusters %>%
+  group_by(cluster.up) %>%
+  summarise(pop.LADs = sum(pop))
 LAD.cluster.upMean <- summarise(group_by(TS2020, cluster.up),
@@
 LAD.cluster.upPM <- summarise(group_by(TS2020[which(TS2020$hour == 19 | TS2020$hour == 20),], cluster.up), mean(speedup))
-LAD.cluster.up <- cbind(LAD.cluster.up, LAD.cluster.upMean[,2], LAD.cluster.upSD[,2], LAD.cluster.upAM[,2], LAD.cluster.upPM[,2])
-names(LAD.cluster.up) [3:6] <- c("upMean", "upSD", "upAMmean",
+LAD.cluster.up <- cbind(LAD.cluster.up, LAD.cluster.upPop[,2], LAD.cluster.upMean[,2], LAD.cluster.upSD[,2], LAD.cluster.upAM[,2], LAD.cluster.upPM[,2])
+names(LAD.cluster.up) [4:7] <- c("upMean", "upSD", "upAMmean",
                                  "upPMmean")
+LAD.cluster.up$AM_slower <- (LAD.cluster.up$upPMmean - LAD.cluster.up$upAMmean) / LAD.cluster.up$upPMmean
+LAD.cluster.up$SDtoMean <- LAD.cluster.up$upSD/LAD.cluster.up$upMean
+LAD.cluster.up$AM_slower <- paste(round(100*LAD.cluster.up$AM_slower, 2), "%", sep="")
+LAD.cluster.up$SDtoMean <- paste(round(100*LAD.cluster.up$SDtoMean, 2), "%", sep="")
 #Download speed clusters  
@@
   summarise(down.LADs = n())
+LAD.cluster.downPop <- clusters %>%
+  group_by(cluster.down) %>%
+  summarise(pop.LADs = sum(pop))
 LAD.cluster.downMean <- summarise(group_by(TS2020, cluster.down),
@@
 LAD.cluster.downPM <- summarise(group_by(TS2020[which(TS2020$hour == 19 | TS2020$hour == 20),], cluster.down), mean(speeddown))
-LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downMean[,2], LAD.cluster.downSD[,2], LAD.cluster.downAM[,2], LAD.cluster.downPM[,2])
-names(LAD.cluster.down)[3:6] <- c("downMean", "downSD", "downAMmean",
+LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downPop[,2], LAD.cluster.downMean[,2], LAD.cluster.downSD[,2], LAD.cluster.downAM[,2], LAD.cluster.downPM[,2])
+names(LAD.cluster.down)[4:7] <- c("downMean", "downSD", "downAMmean",
                                   "downPMmean")
+LAD.cluster.down$AM_slower <- (LAD.cluster.down$downPMmean - LAD.cluster.down$downAMmean) / LAD.cluster.down$downPMmean
+LAD.cluster.down$SDtoMean <- LAD.cluster.down$downSD / LAD.cluster.down$downMean
+LAD.cluster.down$AM_slower <- paste(round(100 * LAD.cluster.down$AM_slower, 2), "%", sep="")
+LAD.cluster.down$SDtoMean <- paste(round(100*LAD.cluster.down$SDtoMean, 2), "%", sep="")
 # commented to run the script
@@
 
-As people came to rely more on their broadband services, it is unsurprising that the rate of testing increased in 2020 compared to 2019.
+As people came to rely more on their broadband services, it is unsurprising that both the amount and rate of testing increased in 2020 during the lockdown compared to the same period in 2019.
 This increase was greatest in cluster 2, where 56% more tests were run compared to the same period in 2019.
-Cluster 2 also has the most local authorities, 148, so the sum of tests run in all these authorities is much higher than in all the other clusters, both in 2019 and 2020.
-Cluster 1 has the second most local authorities, with 93, but the frequency of testing in these authorities is much lower in 2019 and 2020 and there was the lowest rate of increase between the two time periods at just 33%.
-In comparison, cluster 5 has the fewest local authorities, at 28, but the second highest total number of tests and the second largest increase in testing between 2019 and 2020, at 52%.
-A possible explanation is that cluster 5 contains some of the most populous local authorities in the country, including the cities of Birmingham, Manchester, Leeds, Sheffield, Glasgow and Edinburgh.
+Cluster 2 also has the most local authorities, 148, which are home to more than 28.5 million people, so the sum of tests run in all these authorities is much higher than in all the other clusters, both in 2019 and 2020.
+Cluster 1 in comparison, not only has has the lowest  frequency of testing per capita in both 2019 and 2020, it also shows the lowest rate of increase in testing between the two time periods at just 33%.
+Meanwhile, cluster 5 has the fewest local authorities, at 28, but the second highest total number of tests in both periods and the second largest increase in testing between 2019 and 2020, at 52%.
+It also has the highest number of tests per capita in both periods.
+A possible explanation is that cluster 5, with a population of just under 12 million contains some of the most populous local authorities in the country, including the cities of Birmingham, Manchester, Leeds, Sheffield, Glasgow and Edinburgh.
 However, this cluster also includes large rural counties, such as Cornwall and Northumberland.
+Are urban dwellers or rural households more likely to test their broadband speeds?
+
 Yet the temporal profile in 2020 is very different to 2019, as can be seen in graphs [.] and [.].
@@
 The graphs also show that Cluster 1, with the lowest increase in testing, falls from the middle position in terms of total number of tests to the second from bottom.
+
 For upload speeds, 345 of 382 local authorities fall into cluster 6 or cluster 9.
@@
 There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, as well as many suburban districts from the South East of England to South Tyneside.
+
 On the other hand, the 230 local authorities in cluster 6 still include urban areas, including Bristol, Liverpool and Leeds, and many suburban areas.
@@
 This is an indication of the reliability of broadband services in places in both clusters, and suggests that upload speeds have generally been robust throughout the study period.
-The exceptions can be found in seven of the eight much smaller clusters of between 1 and 10 local authority districts, where the standard deviation from mean speeds is between 3% and 14% greater than in cluster 6.
+The exceptions can be found in seven of the eight much smaller clusters including 35 local authority districts, where AM peak upload speeds are between 6% and 18% slower than PM peak upload speeds, although the mean speeds for each cluster are higher than cluster 6.
 The temporal profiles on Graph [.] for the three of those clusters with more than 5 local authorities also shows that speeds are far from dependable in these clusters.
-Furthermore, AM peak upload speeds are between 6% and 18% slower than PM peak upload speeds in the 35 local authorities in these seven clusters, although the mean speeds for each cluster are higher than cluster 6.
-Included in this eclectic group are central London borough of Westminster and the remote Isles of Scilly, and even the largest of these clusters includes a Surrey district on the outskirts of London, the West Wales authority of Ceredigion, and the northern market town of Darlington.
+Indeed, in 25 local authorities with a combined population of almost 3 million, speeds are 13% or more slower in the morning than in the evening.
+Included in this latter group are central London borough of Westminster and the London Borough of Newham, rural authorities like Eden and West Devon, and small cities like Dundee and Carlisle.
+
 Download speeds are the most commonly used measure of the quality and reliability of broadband services.
@@
 This temporal variation can be seen in Graph [.], which shows the 10 of 20 clusters with over five local authorities in them.
-The largest of these 10 clusters, number 19 with 123 Local Authorities, as well as cluster 1 with 16 Local Authorities have lower mean speeds than the other clusters, and this is consistent throughout the temporal profile on the graph.
-However, this means they also have lower standard deviations, without such extreme peaks and troughs as, for example, clusters 7, 9 and 11, with 36 Local Authorities.
+The second largest of these 10 clusters, number 19 with 123 Local Authorities and a population of almost 18 million, along with cluster 1 with 16 local authorities and an additional 1.5million people have lower mean speeds than the other clusters.
+This is consistent throughout the temporal profile on the graph.
+However, cluster 19 also has the second lowest standard deviation at 88% of the mean speed, whilst the standard deviation of cluster 1 is 110% and thus greater than the mean speed.
+
 What is of greater interest, however, is where download speeds are slower in the morning peak when broadband is more likely to be used for work purposes than in the evening peak.
-This difference is most stark in some of the smaller clusters, although since there are only 11 Local Authorities in these clusters even added together, their inclusion in these clusters may be more a reflection of quite localised issues.
-The largest two clusters of Local Authorities have 1-3% slower speeds in the AM peak, and a few clusters have faster download speeds in the AM peak.
-However, there are still 58 Local Authorities, including the 11 mentioned previously with download speeds at least 6% lower in the morning peak than in the evening peak.
-Is the reduction in download speeds in the morning peak in more places than the reduction in upload speeds a result of ISP management to support reliable services for telecommuters?
+This difference is most stark in some of the smaller clusters, although since there are only 11 Local Authorities and a little less than 1.2 million people in these clusters altogether, the differences in speeds may be a reflection of quite localised issues.
+The largest two clusters of Local Authorities, which are home to over 39 million people, have speeds only 1-3% slower in the AM peak, and a few clusters have faster download speeds in the AM peak.
+However, there are 4 clusters with over 9 million people in 47 Local Authorities with download speeds 5-8% lower in the morning peak than in the evening peak, a noticeable reduction and one equivalent to the contention monitored in the evening peak by OfCom to measure service performance.
+Is the 5% or greater reduction in download speeds in the morning peak compared to the evening peak suffered in more places by more people than suffer a similar reduction in upload speeds a result of ISP management to support reliable services for telecommuters?
 There are only eight Local Authorities where both upload and download speeds are more than 5% slower in the AM peak than the PM peak.

### 2020-09-10 src/LAs_Clusters.Rmd
minor changes for population

--- 2020-08-18 src/LAs_Clusters.Rmd
@@
 ```{r}
-pop <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_31_1.data.csv?geography=1807745025...1807745028,1807745030...1807745032,1807745034...1807745155,1807745157...1807745164,1807745166...1807745170,1807745172...1807745177,1807745179...1807745194,1807745196,1807745197,1807745199,1807745201...1807745218,1807745221,1807745222,1807745224,1807745226...1807745231,1807745233,1807745234,1807745236...1807745244,1807745271...1807745281,1811939329...1811939332,1811939334...1811939336,1811939338...1811939497,1811939499...1811939501,1811939503,1811939505...1811939507,1811939509...1811939517,1811939519,1811939520,1811939524...1811939570,1811939575...1811939599,1811939601...1811939628,1811939630...1811939634,1811939636...1811939647,1811939649,1811939655...1811939664,1811939667...1811939680,1811939682,1811939683,1811939685,1811939687...1811939704,1811939707,1811939708,1811939710,1811939712...1811939717,1811939719,1811939720,1811939722...1811939730,1811939757...1811939767&date=latestMINUS1-latest&sex=7&age=0,22&measures=20100")
+# pop <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_31_1.data.csv?geography=1807745025...1807745028,1807745030...1807745032,1807745034...1807745155,1807745157...1807745164,1807745166...1807745170,1807745172...1807745177,1807745179...1807745194,1807745196,1807745197,1807745199,1807745201...1807745218,1807745221,1807745222,1807745224,1807745226...1807745231,1807745233,1807745234,1807745236...1807745244,1807745271...1807745281,1811939329...1811939332,1811939334...1811939336,1811939338...1811939497,1811939499...1811939501,1811939503,1811939505...1811939507,1811939509...1811939517,1811939519,1811939520,1811939524...1811939570,1811939575...1811939599,1811939601...1811939628,1811939630...1811939634,1811939636...1811939647,1811939649,1811939655...1811939664,1811939667...1811939680,1811939682,1811939683,1811939685,1811939687...1811939704,1811939707,1811939708,1811939710,1811939712...1811939717,1811939719,1811939720,1811939722...1811939730,1811939757...1811939767&date=latestMINUS1-latest&sex=7&age=0,22&measures=20100")
 

### 2020-09-11 src/LAs_Clusters.Rmd
the N in the regression table is not correct. it should be equal to the min N from the descriptive stats.

@@
 library(geosphere)
+library(rgeos)
+library(raster)
 
@@
 codes <- la@data %>%
-  select(lad19cd, lad19nm) %>%
+  dplyr::select(lad19cd, lad19nm) %>%
   rename(LAD19NM = lad19nm)
@@
 ladDist$distMet <- apply(ladDist[,3:14], 1, min)
+ladDist <- ladDist %>%
+  dplyr::select(lad19cd, distLondon, distMet)
 
@@
 
-## Job density
+## Job density 2018
 
 ```{r}
+
 job.dens <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_57_1.data.csv?geography=1816133633...1816133848,1820327937...1820328318&date=latest&item=3&measures=20100")
@@
 
-## Population
+## Population 2018
 
@@
 
-## Labour supply
+## Population density 2018
 
 ```{r}
+
+# this is just population
+pop.dens <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_31_1.data.csv?geography=1820327937...1820328318,1816133633...1816133848&date=latestMINUS1-latest&sex=7&age=0,22&measures=20100")
+
+pop.dens <- pop.dens %>%
+  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, AGE_NAME, OBS_VALUE) %>%
+  distinct(GEOGRAPHY_CODE, AGE_NAME, .keep_all = T) %>% 
+  spread(AGE_NAME, OBS_VALUE) %>% 
+  rename(pop16_64 = `Aged 16 - 64`) %>%
+  rename(pop = `All ages`) %>%
+  rename(lad19cd = GEOGRAPHY_CODE) %>%
+  dplyr::select(-pop16_64)
+
+crs(la)
+la$area_sqkm <- area(la) / 1000000
+
+pop.dens <- merge(pop.dens, la@data, by.x = "lad19cd", by.y = "LAD19CD", all.x = T)
+
+pop.dens <- pop.dens %>%
+  mutate(pop.dens = pop/area_sqkm) %>%
+  dplyr::select(lad19cd,pop.dens)
+
+# sapply(pop.dens, function(x) sum(is.na(x)))
+# test <- pop.dens[is.na(pop.dens$area_sqkm),]
+```
+
+## N/S divide
+
+```{r}
+
+# It only includes only the south
+
+ns <- read_csv("https://opendata.arcgis.com/datasets/3ba3daf9278f47daba0f561889c3521a_0.csv") %>%
+  mutate(south = ifelse(RGN19NM=="London" | 
+                        RGN19NM=="South West" |
+                        RGN19NM=="East of England" |
+                        RGN19NM=="South East", 1, 0)) %>%
+  dplyr::select(LAD19CD, south) %>% 
+  dplyr::filter(south==1) %>%
+  rename(lad19cd = LAD19CD)
+```
+
+## Labour supply March 2020
+
+```{r}
+
 labour <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_17_5.data.csv?geography=1816133633...1816133837,1820327937...1820328307&date=latest&variable=18,1532...1540,290,720&measures=20599,21001,21002,21003")
@@
 
-## Earnings
+## Earnings 2019
 
 ```{r}
+
 earnings <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_30_1.data.csv?geography=1946157348,2092957698,2013265929,1816133633...1816133837,1820327937...1820328307&date=latest&sex=8&item=2&pay=1&measures=20100,20701")
@@
 
-## Business counts
+## Business counts 2019
 
 ```{r}
+
 # busi <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_141_1.data.csv?geography=1816133633...1816133848,1820327937...1820328318,1870659585...1870659791,1870659801,1870659792...1870659800,1879048193...1879048573,1879048583,1879048574...1879048582&date=latest&industry=163577857...163577874&employment_sizeband=0&legal_status=0&measures=20100")
@@
 ```{r}
-data <- list(lad, job.dens, pop, labour, earnings, busi)
+data <- list(lad, ladDist, job.dens, pop, pop.dens, labour, earnings, busi) # ns 
 sapply(data, function(x) dim(x))
@@
 sapply(data, function(x) sum(is.na(x)))
+
+data <- merge(data, ns, all.x = T)
+data$south <- ifelse(is.na(data$south),0, data$south)
 ```
@@
 
-```{r correlations}
+```{r correlations, results = 'asis'}
 data.cor <- data %>%
@@
 data.cor <- data %>%
-  dplyr::select(LAD.up.am20, LAD.up.pm20,
-                LAD.tests.pm20,
+  dplyr::select(LAD.tests.am20,
                 pop, pop16_64,
@@
 
+data.cor <- data %>%
+  dplyr::select(pop16_64, 
+                pop.dens,
+                job.dens2018, 
+                distMet,
+                distLondon,
+                south,
+                managers, 
+                tech,
+                skilled,
+                prof,
+                admin, 
+                caring,
+                plant,
+                earnings,
+                total.busi,
+                LAD.tests.am20,
+                pop) %>%
+  mutate('LAD.tests.am20/pop' = LAD.tests.am20/pop) %>%
+    mutate('total.busi/pop' = total.busi/pop) %>%
+  dplyr::select(-LAD.tests.am20,-pop, -total.busi)
+
 m <- cor(data.cor, use = "pairwise.complete.obs")
 corrplot(m, type="upper",method = "number", number.cex = .5, tl.cex = .75)
-```
 
-## Auxiliary regression, dep. var.: upload clusters
-
-```{r upload regression, results= 'asis', message=F, error=F, warning=F}
-# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
-# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/
-
-aux <- multinom(cluster.up ~ LAD.up.am20 + LAD.up.pm20 + LAD.tests.pm20 + #  LAD.down.am20 + LAD.down.pm20
-                   pop + pop16_64 + # how about young population share
-                   managers + tech + skilled +
-                   earnings + total.busi, data = data, 
-                trace = F) # to avoid messages
-
-# McFadden's R squared
-aux.null <- multinom(cluster.up ~ 1, data = data)
-r2 <- 1-logLik(aux)/logLik(aux.null)
-
-stargazer(aux, type = "html",
-          add.lines = list(c("McFadden's R squared",                    # use rownames as line name. in this case Weak insterumetns
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3))))
-
+stargazer(data.cor,
+          type="html",
+          #type="text", 
+          summary = T,
+          nobs = T)
 ```
@@
 
-aux <- multinom(cluster.up ~ pop + pop16_64 + # how about young population share
-                  managers + tech + skilled +
-                  earnings + total.busi, data = data, 
+# convert dependent variable to factor
+data$cluster.up <- as.factor(data$cluster.up)
+data$cluster.up <- relevel(data$cluster.up, "5") # base category: cluster 5
+
+aux <- multinom(as.factor(cluster.up) ~ pop16_64 + # how about young population share, pop +
+                  pop.dens + job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
+                  distMet + distLondon + south +
+                  managers + tech + skilled + prof + admin + caring + plant +
+                  earnings + I(total.busi/pop) +
+                  I(LAD.tests.am20/pop), data = data, 
                 trace = F) # to avoid messages
@@
 
-stargazer(aux, type = "html",
-          add.lines = list(c("McFadden's R squared",                    # use rownames as line name. in this case Weak insterumetns
+stargazer(aux, 
+          type = "html",
+          #type = "text",
+          nobs = T,
+          covariate.labels=c("pop. 16-64, 2018",
+                             "pop. densuty, 2018",
+                             "job density, 2018",
+                             "distance to nearest met. area",
+                             "distance to London",
+                             "South of the UK",
+                             "% of managerial jobs, 2020",
+                             "% of tech jobs, 2020",
+                             "% of skilled trade jobs, 2020",
+                             "% of professional jobs, 2020",
+                             "% of dministrative jobs, 2020",
+                             "% of leisure jobs, 2020",
+                             "% of machine operation jobs, 2020",
+                             "earnings, 2019",
+                             "n. business est. per hab., 2019",
+                             "n. of AM broadband tests per hab., 2019"),
+          add.lines = list(c("McFadden's R squared",
                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3))))
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3)),
+                           (c("N",
+                            nrow(data), nrow(data), nrow(data),
+                            nrow(data), nrow(data), nrow(data),
+                            nrow(data), nrow(data), nrow(data)))))
 
 ```
-
-## Auxiliary regression, dep. var.: download clusters
-
-```{r download regression, results= 'asis', message=F, error=F, warning=F}
-# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
-# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/
-
-aux <- multinom(cluster.down ~ LAD.down.am20 + LAD.down.pm20 + LAD.tests.pm20 + #  LAD.down.am20 + LAD.down.pm20
-                   pop + pop16_64 + # how about young population share
-                   managers + tech + skilled +
-                   earnings + total.busi, data = data, 
-                trace = F) # to avoid messages
-
-# McFadden's R squared
-aux.null <- multinom(cluster.down ~ 1, data = data)
-r2 <- 1-logLik(aux)/logLik(aux.null)
-
-stargazer(aux, type = "html",
-          add.lines = list(c("McFadden's R squared",                    # use rownames as line name. in this case Weak insterumetns
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                              round(r2[1],3))))
-
-```

### 2020-09-11 src/LAs_Clusters.Rmd
minor changes

@@
 data$south <- ifelse(is.na(data$south),0, data$south)
+
+# NOT RUN
+# export file for future reference / backup
+# data.out.path <- paste0(path, "/data/temp/data_for_aux.csv")
+# write.csv(data, data.out.path)
 ```
@@
           nobs = T,
+          digits = 3,
           covariate.labels=c("pop. 16-64, 2018",

### 2020-09-11 src/LAs_Clusters.Rmd
more aux regressions

@@
 
-## Auxiliary regression, dep. var.: upload clusters, without upload variables
+## Auxiliary regression, dep. var.: upload clusters
 
-```{r upload regression no upload and test, results= 'asis', message=F, error=F, warning=F}
+```{r upload regression , results= 'asis', message=F, error=F, warning=F}
 # helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
@@
 
-aux <- multinom(as.factor(cluster.up) ~ pop16_64 + # how about young population share, pop +
-                  pop.dens + job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
+aux <- multinom(cluster.up ~ 
+                  #pop16_64 + # non sign, while pop is
+                  pop +
+                  #pop.dens + 
+                  job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
                   distMet + distLondon + south +
@@
                   earnings + I(total.busi/pop) +
-                  I(LAD.tests.am20/pop), data = data, 
+                  I(LAD.tests.am20/pop) # non sensitive to addition of other tests pm
+                , data = data, 
                 trace = F) # to avoid messages
@@
 
+n.obs <- dim(aux$fitted.values)[1]
+
 stargazer(aux, 
@@
           digits = 3,
-          covariate.labels=c("pop. 16-64, 2018",
-                             "pop. densuty, 2018",
+          covariate.labels=c(#"pop. 16-64, 2018",
+                             "pop, 2018",
+                             #"pop. densuty, 2018",
                              "job density, 2018",
@@
                              "% of professional jobs, 2020",
-                             "% of dministrative jobs, 2020",
+                             "% of administrative jobs, 2020",
                              "% of leisure jobs, 2020",
@@
                            (c("N",
-                            nrow(data), nrow(data), nrow(data),
-                            nrow(data), nrow(data), nrow(data),
-                            nrow(data), nrow(data), nrow(data)))))
+                            n.obs, n.obs, n.obs,
+                            n.obs, n.obs, n.obs,
+                            n.obs, n.obs, n.obs))))
 
 ```
+
+## Auxiliary regression, dep. var.: test clusters
+
+```{r test regression, results= 'asis', message=F, error=F, warning=F}
+# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
+# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/
+
+# convert dependent variable to factor
+data$cluster.tests <- as.factor(data$cluster.tests)
+data$cluster.tests <- relevel(data$cluster.tests, "5") # base category: cluster 5
+
+aux <- multinom(cluster.tests ~ 
+                  #pop16_64 + # non sign, while pop is
+                  pop +
+                  #pop.dens + 
+                  job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
+                  distMet + distLondon + south +
+                  managers + tech + skilled + prof + admin + caring + plant +
+                  earnings + I(total.busi/pop) +
+                  I(LAD.tests.am20/pop) # non sensitive to addition of other tests pm
+                , data = data, 
+                trace = F) # to avoid messages
+
+# McFadden's R squared
+aux.null <- multinom(cluster.tests ~ 1, data = data)
+r2 <- 1-logLik(aux)/logLik(aux.null)
+
+n.obs <- dim(aux$fitted.values)[1]
+
+stargazer(aux, 
+          type = "html",
+          #type = "text",
+          digits = 3,
+          covariate.labels=c(#"pop. 16-64, 2018",
+                             "pop, 2018",
+                             #"pop. densuty, 2018",
+                             "job density, 2018",
+                             "distance to nearest met. area",
+                             "distance to London",
+                             "South of the UK",
+                             "% of managerial jobs, 2020",
+                             "% of tech jobs, 2020",
+                             "% of skilled trade jobs, 2020",
+                             "% of professional jobs, 2020",
+                             "% of administrative jobs, 2020",
+                             "% of leisure jobs, 2020",
+                             "% of machine operation jobs, 2020",
+                             "earnings, 2019",
+                             "n. business est. per hab., 2019",
+                             "n. of AM broadband tests per hab., 2019"),
+          add.lines = list(c("McFadden's R squared",
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                             round(r2[1],3)),
+                           (c("N",
+                            n.obs, n.obs, n.obs,
+                            n.obs))))
+
+```
+
+## Auxiliary regression, dep. var.: down clusters
+
+```{r down regression, results= 'asis', message=F, error=F, warning=F}
+# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
+# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/
+
+# convert dependent variable to factor
+data$cluster.down <- as.factor(data$cluster.down)
+data$cluster.down <- relevel(data$cluster.down, "2") # base category: cluster 2
+
+aux <- multinom(cluster.down ~ 
+                  #pop16_64 + # non sign, while pop is
+                  pop +
+                  #pop.dens + 
+                  job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
+                  distMet + distLondon + south +
+                  managers + tech + skilled + prof + admin + caring + plant +
+                  earnings + I(total.busi/pop) +
+                  I(LAD.tests.am20/pop) # non sensitive to addition of other tests pm
+                , data = data, 
+                trace = F) # to avoid messages
+
+# McFadden's R squared
+aux.null <- multinom(cluster.down ~ 1, data = data)
+r2 <- 1-logLik(aux)/logLik(aux.null)
+
+n.obs <- dim(aux$fitted.values)[1]
+
+stargazer(aux, 
+          type = "html",
+          #type = "text",
+          digits = 3,
+          covariate.labels=c(#"pop. 16-64, 2018",
+                             "pop, 2018",
+                             #"pop. densuty, 2018",
+                             "job density, 2018",
+                             "distance to nearest met. area",
+                             "distance to London",
+                             "South of the UK",
+                             "% of managerial jobs, 2020",
+                             "% of tech jobs, 2020",
+                             "% of skilled trade jobs, 2020",
+                             "% of professional jobs, 2020",
+                             "% of administrative jobs, 2020",
+                             "% of leisure jobs, 2020",
+                             "% of machine operation jobs, 2020",
+                             "earnings, 2019",
+                             "n. business est. per hab., 2019",
+                             "n. of AM broadband tests per hab., 2019"),
+          add.lines = list(c("McFadden's R squared",
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
+                             round(r2[1],3)),
+                           (c("N",
+                            n.obs, n.obs, n.obs,
+                            n.obs, n.obs, n.obs,
+                            n.obs, n.obs, n.obs,
+                            n.obs, n.obs, n.obs,
+                            n.obs, n.obs, n.obs,
+                            n.obs, n.obs, n.obs,
+                            n.obs))))
+
+```

### 2020-10-07 src/LAs_Clusters.Rmd
share of virgin, rebase upload regression

@@
 
+# this is a share of Virgin provider variable
+virgin <- TS2020 %>%
+  filter(provider == "Virgin Media") %>%
+  group_by(LAD19NM) %>%
+  summarise(virgin = n())
+
+virgin.denom <- TS2020 %>%
+  #filter(provider == "Virgin Media") %>%
+  group_by(LAD19NM) %>%
+  summarise(virgin.denom = n())
+
+virgin <- merge(virgin.denom, virgin, by = "LAD19NM", all.x = T)
+virgin$virgin <- ifelse(is.na(virgin$virgin), 0, virgin$virgin)
+virgin$virgin <- virgin$virgin/virgin.denom$virgin.denom
+virgin$virgin.denom <- NULL
+
 # LAD characteristcs merge
@@
             LAD.tests.pm19,
-            LAD.tests.pm20)
+            LAD.tests.pm20,
+            virgin)
 
@@
                 LAD.tests.am20,
-                pop) %>%
+                pop,
+                virgin) %>%
   mutate('LAD.tests.am20/pop' = LAD.tests.am20/pop) %>%
@@
 data$cluster.up <- as.factor(data$cluster.up)
-data$cluster.up <- relevel(data$cluster.up, "5") # base category: cluster 5
+data$cluster.up <- relevel(data$cluster.up, "3") # base category: cluster 3
 
@@
                   earnings + I(total.busi/pop) +
-                  I(LAD.tests.am20/pop) # non sensitive to addition of other tests pm
+                  I(LAD.tests.am20/pop) + # non sensitive to addition of other tests pm
+                  virgin
                 , data = data, 
@@
                              "n. business est. per hab., 2019",
-                             "n. of AM broadband tests per hab., 2019"),
+                             "n. of AM broadband tests per hab., 2019",
+                             "% of Virgin Media connections"),
           add.lines = list(c("McFadden's R squared",

### 2020-10-07 src/descriptive_Clusters.Rmd

@@
 All this suggests that more of the additional testing generated in 2020 compared to 2019 is occurring outside the traditional peak in the evening at streaming primetime.
-Moreover, this is particularly noticeable for cluster 2, but the change is less pronounced for cluster 5, perhaps because of the diversity of urban and rural populations represented in cluster 5.
+
+The increase in daytime compared to evening testing is particularly noticeable for cluster 2, but the change is less pronounced for cluster 5.
+Indeed, in an auxiliary regression, the probability of additional tests per person run between 9.00 and 11.00 was significantly higher for those living in cluster 2 than in cluster 5, whilst the probability was significantly lower in cluster 1.
+??
+perhaps because of the diversity of urban and rural populations represented in cluster 5 ??
+#re-base auxiliary regression so reference case is either 1 or 3 - we wanted to talk most about 5?
 The graphs also show that Cluster 1, with the lowest increase in testing, falls from the middle position in terms of total number of tests to the second from bottom.
 
-For upload speeds, 345 of 382 local authorities fall into cluster 6 or cluster 9.
+For upload speeds, 345 of 382 local authorities, or over 62 million people, fall into cluster 6 or cluster 9.
 Graph [.] shows that both of these clusters have relatively similar temporal profiles, with relatively stable speeds.
@@
 This difference is reflected not only in the mean speeds for these clusters for the whole sample, but also the mean upload speeds in evening peak period from 19:00-20:59 and the morning peak from 9:00-10:59.
-Which sort of areas are benefitting from these higher speeds?
-The 115 Local Authority areas are mostly in urban or suburban areas, including 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation.
+The 115 Local Authority areas in cluster 9 are mostly in urban or suburban areas, including 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation.
 There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, as well as many suburban districts from the South East of England to South Tyneside.
 
-On the other hand, the 230 local authorities in cluster 6 still include urban areas, including Bristol, Liverpool and Leeds, and many suburban areas.
-However, it also includes some of the most rural areas in the country, and could be more seen as a typical level of service for upload speeds, rather than being particularly slow.
-The differences between morning and evening upload speeds for both clusters are minimal, at 1% slower in the morning in cluster 9 and 4% slower in cluster 6.
+Meanwhile, the 230 local authorities in cluster 6 still include major urban areas, such as Bristol, Liverpool and Leeds, and many suburban areas, but also include some of the most rural areas in the country.
+Slower speeds could be seen as a typical level of service in such areas, and indeed, upload speeds in cluster 6 are only 4% slower in the morning than in the evening, compared to 12-19% slower for the almost 3 million people living in five of the smaller clusters.
+And yet, auxiliary regressions show that those living in cluster 6 had the highest probability of testing their broadband between 9:00 and 11:00 of any of the 10 upload speed clusters.
+Cluster 6 also experienced the joint highest ratio of standard deviation to mean across the time period under assessment.
+This suggests that although the time profile is relatively flat in graph [.], the experience is one of speeds that fluctuate from a lower mean, and therefore might more often impact on the activities users are doing.
+
+Cluster 1, with 10 local authorities that are home to over 1 million people, is also interesting, as it has the second slowest speeds in the morning compared to the evening 'peaks' and the second highest ratio of standard deviation to mean.
+Indeed, the time profile in graph [.] shows upload speeds in cluster 1 are on average lower at certain times of day during the study period than any other cluster.
+Yet the auxiliary regression suggests that there are not many tests being run during the am peak in cluster 1.
+This may be because there are fewer people working at home, anxiously checking their broadband than in most other clusters.
+Indeed, the auxiliary regressions indicate that cluster 1 has the highest job density or proportion of jobs to working-age population, which is likely to due to the presence of Westminster, central London, cluster 1's most populous local authority.
+Westminster not only has more workplaces than residents, but it is reasonable to presume that many who would normally work in Westminster, but be able to work from home during lockdown are likely to live outside central London and not be subject to the fluctuating speeds there.
+Workplaces, meanwhile, some of which would still be open in a place like Westminster, could be running programmes that cause the slowdown and variation, but would be more likely to have their own in-house diagnostics, rather than using a service like Speedchecker Ltd.
+
+Furthermore, whilst cluster 1 has the second highest number of established businesses per capita, it is far behind the cluster with the most business per capita, cluster 6.
+This suggests that cluster 1 has more large employers than cluster 6, which fits with the likelihood that those working from home whose large employer is based in cluster 1 are much less likely than those in cluster 6 to also live in a local authority in cluster 1 (or at least Westminster, which probably skews the sample).
+Meanwhile, cluster 9 has the lowest number of established businesses per inhabitant.
+Considering how many central urban areas fall into cluster 9, this suggests an even greater dominance of large employers.
+Yet the higher upload speeds in cluster 9 may mean that people working from home in these local authorities noticed fewer problems.
+There is also the least difference between morning and evening upload speeds for cluster 9, at 1% slower in the morning.
+However, that is not to say that cluster 9 did not include many people telecommuting during lockdown.
+
+Indeed, the only cluster where upload speeds were slower in the evening than in the morning was cluster 5, made up of 2 local authorities with less than 200,000 people: Three Rivers, a suburban district north of London, and Fylde, a seaside suburb of Blackpool.
+The auxiliary regressions show that compared to these two places, all the other clusters had a lower percentage of working people in managerial, professional and administrative jobs.???
+
 This is an indication of the reliability of broadband services in places in both clusters, and suggests that upload speeds have generally been robust throughout the study period.
@@
 Download speeds are the most commonly used measure of the quality and reliability of broadband services.
-They also vary the most, with standard deviations in speed tending to be in a similar range to the mean speeds for each cluster, compared to upload speeds, where standard deviations are half to two thirds the level of the mean speeds.
+They are most sensitive, which does result in substantial noise.
+This is clear from the way that download speeds vary the most, with the largest range and standard deviations in speed tending to be in a similar range to the mean speeds for each cluster, compared to upload speeds, where standard deviations are half to two thirds the level of the mean speeds.
 This temporal variation can be seen in Graph [.], which shows the 10 of 20 clusters with over five local authorities in them.

### 2020-10-07 paper/v1/broadband.speed.covid.Rmd
paper structure

---
title: |
  WFH and broadband speed (title needs rework) 
type: ARTICLE TEMPLATE
author:
  - name: A. N. Author
    affil: a
    email: latex.helpdesk@tandf.co.uk
  - name: John Smith
    affil: b
    email: john.smith@uni-freiburg.de
affiliation:
  - num: a
    address: |
      Taylor & Francis, 4 Park Square, Milton Park, Abingdon, UK
  - num: b
    address: |
      Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany
bibliography: bibliography.bib
appendix: appendix.tex
abstract: |
  TBC
keywords: |
  covid; internet; working from home; broadband speed; time series clusters
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

# Introduction

[our Google doc](https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit#heading=h.i5om1o8wpcd9)

**PARA1**: working from home is a challenge and maybe an opportunity [use arguments from @BUDNITZ2020102713]
It has been altered dramatically because of covid

**PARA2**: covind and working from home, cities, urban structure
Potentially useful readings for covid and cities:

- https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas
- https://www.coronavirusandtheeconomy.com/question/will-coronavirus-cause-big-city-exodus
- [EPB commentaries](https://journals.sagepub.com/toc/EPB/current)
- https://journals.sagepub.com/doi/full/10.1177/2399808320926912?casa_token=gh3SgKFCZ44AAAAA:MjjMWJ61DqlmxzkWQyy_wxPnU20QdyQYkC4fVFpHLPfLk7McwmQGkJ7x2Q7LZOjTf6vcYaqNwug
- [WFM and covid, econ](https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home)
- [productivity and WFM](https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity)

It is important to become aware of working from home patterns because of:
- transport planning reasons (see para 1)
- WFM might increase coparing to pre-covid level

**PARA3**: How can we observe observe WFM? 
Directly through surveys (@felstead2020homeworking).
Add survey limitations (cost, reprentativeness, time lag between survey and data availability)
Data gap.
Passive data collection through internet speeds
Another option could be mobility data.
How is internet speed data a better source than mobility?

**PARA4**: Contention

**PARA5**: Data and methods

**PARA6**: Contribution:

# Literature review

## broadband studies, divides, broadband tech stuff

## from telecommuting to #WFH

Some new papers google recommended to me:

- https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf
- https://link.springer.com/article/10.1007/s11116-020-10136-6
- https://www.sciencedirect.com/science/article/pii/S0966692319311305
- check who cites the above and what they cite

# Time series clustering

Desription of the method

# Data and descriptive statistics

Data details and some figures, descriptive stats

# Results

Clusters, cluster description and aux regressions

# Conclusions

# Acknowledgement(s) {-}

An unnumbered section, e.g.\ `\section*{Acknowledgements}`, may be used for thanks, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.

# Funding {-}

An unnumbered section, e.g.\ `\section*{Funding}`, may be used for grant details, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.

# References

### 2020-10-07 paper/v1/broadband.speed.covid.tex
paper structure

% interactcadsample.tex
% v1.03 - April 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

% see https://stackoverflow.com/a/47122900

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}

\begin{document}

\articletype{ARTICLE TEMPLATE}

\title{WFH and broadband speed (title needs rework)}

\author{\name{A. N. Author$^{a}$, John Smith$^{b}$}
\affil{$^{a}$Taylor \& Francis, 4 Park Square, Milton Park, Abingdon, UK; $^{b}$Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany}
}

\thanks{CONTACT A. N. Author. Email: \href{mailto:latex.helpdesk@tandf.co.uk}{\nolinkurl{latex.helpdesk@tandf.co.uk}}, John Smith. Email: \href{mailto:john.smith@uni-freiburg.de}{\nolinkurl{john.smith@uni-freiburg.de}}}

\maketitle

\begin{abstract}
TBC
\end{abstract}

\begin{keywords}
covid; internet; working from home; broadband speed; time series
clusters
\end{keywords}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\href{https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit\#heading=h.i5om1o8wpcd9}{our
Google doc}

\textbf{PARA1}: working from home is a challenge and maybe an
opportunity \citep[use arguments from][]{BUDNITZ2020102713} It has been
altered dramatically because of covid

\textbf{PARA2}: covind and working from home, cities, urban structure
Potentially useful readings for covid and cities:

\begin{itemize}
\tightlist
\item
  \url{https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas}
\item
  \url{https://www.coronavirusandtheeconomy.com/question/will-coronavirus-cause-big-city-exodus}
\item
  \href{https://journals.sagepub.com/toc/EPB/current}{EPB commentaries}
\item
  \url{https://journals.sagepub.com/doi/full/10.1177/2399808320926912?casa_token=gh3SgKFCZ44AAAAA:MjjMWJ61DqlmxzkWQyy_wxPnU20QdyQYkC4fVFpHLPfLk7McwmQGkJ7x2Q7LZOjTf6vcYaqNwug}
\item
  \href{https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home}{WFM
  and covid, econ}
\item
  \href{https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity}{productivity
  and WFM}
\end{itemize}

It is important to become aware of working from home patterns because
of: - transport planning reasons (see para 1) - WFM might increase
coparing to pre-covid level

\textbf{PARA3}: How can we observe observe WFM? Directly through surveys
(\citet{felstead2020homeworking}). Add survey limitations (cost,
reprentativeness, time lag between survey and data availability) Data
gap. Passive data collection through internet speeds Another option
could be mobility data. How is internet speed data a better source than
mobility?

\textbf{PARA4}: Contention

\textbf{PARA5}: Data and methods

\textbf{PARA6}: Contribution:

\hypertarget{literature-review}{%
\section{Literature review}\label{literature-review}}

\hypertarget{broadband-studies-divides-broadband-tech-stuff}{%
\subsection{broadband studies, divides, broadband tech
stuff}\label{broadband-studies-divides-broadband-tech-stuff}}

\hypertarget{from-telecommuting-to-wfh}{%
\subsection{from telecommuting to
\#WFH}\label{from-telecommuting-to-wfh}}

Some new papers google recommended to me:

\begin{itemize}
\tightlist
\item
  \url{https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf}
\item
  \url{https://link.springer.com/article/10.1007/s11116-020-10136-6}
\item
  \url{https://www.sciencedirect.com/science/article/pii/S0966692319311305}
\item
  check who cites the above and what they cite
\end{itemize}

\hypertarget{time-series-clustering}{%
\section{Time series clustering}\label{time-series-clustering}}

Desription of the method

\hypertarget{data-and-descriptive-statistics}{%
\section{Data and descriptive
statistics}\label{data-and-descriptive-statistics}}

Data details and some figures, descriptive stats

\hypertarget{results}{%
\section{Results}\label{results}}

Clusters, cluster description and aux regressions

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

\hypertarget{acknowledgements}{%
\section*{Acknowledgement(s)}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgement(s)}

An unnumbered section,
e.g.~\texttt{\textbackslash{}section*\{Acknowledgements\}}, may be used
for thanks, etc.~if required and included \emph{in the non-anonymous
version} before any Notes or References.

\hypertarget{funding}{%
\section*{Funding}\label{funding}}
\addcontentsline{toc}{section}{Funding}

An unnumbered section,
e.g.~\texttt{\textbackslash{}section*\{Funding\}}, may be used for grant
details, etc.~if required and included \emph{in the non-anonymous
version} before any Notes or References.

\bibliographystyle{tfcad}
\bibliography{bibliography.bib}

\input{"appendix.tex"}

\end{document}

### 2020-10-21 paper/v1/broadband.speed.covid.Rmd
juut brought some of the google doc text

@@
 
-**PARA1**: working from home is a challenge and maybe an opportunity [use arguments from @BUDNITZ2020102713]
-It has been altered dramatically because of covid
+<!-- **PARA1**: working from home is a challenge and maybe an opportunity [use arguments from @BUDNITZ2020102713] -->
+<!-- It has been altered dramatically because of covid -->
+
+During the pandemic, working from home using Information and Communication
+Technologies (ICT), whether partially or exclusively, was transformed from a
+niche means of accessing work, albeit one that had been on a slow, upward trend,
+to a widespread way of life in many countries.
+The ability to work from home or telecommute meant millions retained their jobs
+and, to a varying extent, maintained productivity during periods of strict
+lockdown around the world.
+However, this ability was not evenly distributed socially or spatially, creating
+a new type of digital divide -- one of economic resilience and personal safety on
+one side and unemployment or essential front-line work on the other.
+The ability to telecommute is partly a product of whether the majority of work
+tasks in an occupation can be performed independently of being in a particular
+location or co-locating with colleagues.
+It is also a product of the quality of ICT services, particularly home internet
+connections, sufficient to successfully complete work tasks with a minimum of
+delay or interruption.
+Therefore, in this paper, we aim to look at patterns of broadband speed tests
+and upload speeds in the UK in order to gain insights into both the quality and
+reliability of broadband services across the country and the distribution of
+telecommuting at the time when the population were told to work from home if
+possible.
+From this analysis, we provide a greater understanding of how telecommuting and
+technology intersect at a time of extreme demand, and what lessons this time has
+for a future where telecommuting is likely to remain a common means of accessing
+work and broadband services, as well as infrastructure, must be fit for purpose.
 

### 2020-10-21 src/descriptive_Clusters.Rmd
More writing...

@@
 Cluster 1 in comparison, not only has has the lowest  frequency of testing per capita in both 2019 and 2020, it also shows the lowest rate of increase in testing between the two time periods at just 33%.
+
 Meanwhile, cluster 5 has the fewest local authorities, at 28, but the second highest total number of tests in both periods and the second largest increase in testing between 2019 and 2020, at 52%.
@@
 Are urban dwellers or rural households more likely to test their broadband speeds?
+Auxiliary regressions suggest that cluster 5 also had the highest earnings of the clusters, but was least likely to have a Virgin Media connection.
 
@@
 The increase in daytime compared to evening testing is particularly noticeable for cluster 2, but the change is less pronounced for cluster 5.
-Indeed, in an auxiliary regression, the probability of additional tests per person run between 9.00 and 11.00 was significantly higher for those living in cluster 2 than in cluster 5, whilst the probability was significantly lower in cluster 1.
+Indeed, in an auxiliary regression, the probability of additional tests per person run between 9.00 and 11.00 was significantly higher for those living in cluster 2 than in cluster 5, but both were higher than the other three clusters.
 ??
 perhaps because of the diversity of urban and rural populations represented in cluster 5 ??
+Cluster 5 mostly same as cluster 6 upload, few cluster 9, but same for cluster 2?
 #re-base auxiliary regression so reference case is either 1 or 3 - we wanted to talk most about 5?
@@
 However, the upload speeds at all times for cluster 9 are substantially higher than for cluster 6.
-This difference is reflected not only in the mean speeds for these clusters for the whole sample, but also the mean upload speeds in evening peak period from 19:00-20:59 and the morning peak from 9:00-10:59.
-The 115 Local Authority areas in cluster 9 are mostly in urban or suburban areas, including 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation.
-There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, as well as many suburban districts from the South East of England to South Tyneside.
+This difference is reflected not only in the mean speeds for these clusters for the whole sample, but also the mean upload speeds in the morning peak from 9:00-10:59, as well as the evening peak period from 19:00-20:59.
+The quality of broadband service is thus much better in the 115 Local Authority areas in cluster 9, which are mostly in urban or suburban areas.
+These include 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation.
+There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, and many suburban districts from the South East of England to South Tyneside.
+Auxiliary regressions indicate that the speed tests in these authorities are also more likely to have been run on services provided by Virgin Media, suggesting they are in the half of the country with the most lucrative ICT market, which originally attracted the cable TV provider (OfCom...).
 
 Meanwhile, the 230 local authorities in cluster 6 still include major urban areas, such as Bristol, Liverpool and Leeds, and many suburban areas, but also include some of the most rural areas in the country.
-Slower speeds could be seen as a typical level of service in such areas, and indeed, upload speeds in cluster 6 are only 4% slower in the morning than in the evening, compared to 12-19% slower for the almost 3 million people living in five of the smaller clusters.
-And yet, auxiliary regressions show that those living in cluster 6 had the highest probability of testing their broadband between 9:00 and 11:00 of any of the 10 upload speed clusters.
+For example, although auxiliary regressions show that Cluster 6 local authorities are more likely to be in the South of the UK than Cluster 9, the cluster notably includes Southern rural districts from Cornwall to North Norfolk.
+Slower speeds could reflect the lower quality of service in rural areas compared to urban and suburban areas.
+The reliability of service in these areas is also in question.
+For although upload speeds in cluster 6 are only 4% slower in the morning than in the evening, compared to 12-19% slower for the almost 3 million people living in five of the smaller clusters, auxiliary regressions show that those living in cluster 6 had the highest probability of testing their broadband between 9:00 and 11:00 of any of the 10 upload speed clusters.
 Cluster 6 also experienced the joint highest ratio of standard deviation to mean across the time period under assessment.
@@
 
-Cluster 1, with 10 local authorities that are home to over 1 million people, is also interesting, as it has the second slowest speeds in the morning compared to the evening 'peaks' and the second highest ratio of standard deviation to mean.
+In comparison, there is negligible difference between morning and evening upload speeds for cluster 9, at 1% slower in the morning.
+This suggests a high level of reliability of service, a conclusion supported by the finding in the auxiliary regressions that those in cluster 9 ran the fewest speed tests per person during the morning period of any cluster.
+Now this may be an indication of fewer people working from home, less contention, and less resultant frustration.
+Cluster 9 comprises many central urban areas and has the lowest number of established businesses per inhabitant, which could be interpreted as a dominance of large employers.
+However, the job density is lower than cluster 6, meaning there are not as many jobs per resident in these areas.
+As the cluster includes many suburban areas too, which may be largely residential, could the quality and reliability of internet service be reinforcing patterns of telecommuting by those in wealthier suburbs who can work from home?
+Earnings in cluster 9 are second highest of all the clusters, with only Cluster 2 (comprising just North East Lincolnshire and East Lothian, population 265k) earning more per person.
+
+Meanwhile, Cluster 1, with 10 local authorities that are home to over 1 million people has the second slowest speeds in the morning compared to the evening 'peaks' and the second highest ratio of standard deviation to mean.
 Indeed, the time profile in graph [.] shows upload speeds in cluster 1 are on average lower at certain times of day during the study period than any other cluster.
 Yet the auxiliary regression suggests that there are not many tests being run during the am peak in cluster 1.
-This may be because there are fewer people working at home, anxiously checking their broadband than in most other clusters.
+This may be because there are fewer people working at home checking their broadband than in most other clusters.
 Indeed, the auxiliary regressions indicate that cluster 1 has the highest job density or proportion of jobs to working-age population, which is likely to due to the presence of Westminster, central London, cluster 1's most populous local authority.
 Westminster not only has more workplaces than residents, but it is reasonable to presume that many who would normally work in Westminster, but be able to work from home during lockdown are likely to live outside central London and not be subject to the fluctuating speeds there.
-Workplaces, meanwhile, some of which would still be open in a place like Westminster, could be running programmes that cause the slowdown and variation, but would be more likely to have their own in-house diagnostics, rather than using a service like Speedchecker Ltd.
-
-Furthermore, whilst cluster 1 has the second highest number of established businesses per capita, it is far behind the cluster with the most business per capita, cluster 6.
-This suggests that cluster 1 has more large employers than cluster 6, which fits with the likelihood that those working from home whose large employer is based in cluster 1 are much less likely than those in cluster 6 to also live in a local authority in cluster 1 (or at least Westminster, which probably skews the sample).
-Meanwhile, cluster 9 has the lowest number of established businesses per inhabitant.
-Considering how many central urban areas fall into cluster 9, this suggests an even greater dominance of large employers.
-Yet the higher upload speeds in cluster 9 may mean that people working from home in these local authorities noticed fewer problems.
-There is also the least difference between morning and evening upload speeds for cluster 9, at 1% slower in the morning.
-However, that is not to say that cluster 9 did not include many people telecommuting during lockdown.
+Workplaces, meanwhile, some of which would still have been open, could be running programmes that cause the slowdown and variation, but would be more likely to have their own in-house diagnostics, rather than using a service like Speedchecker Ltd.
 
 Indeed, the only cluster where upload speeds were slower in the evening than in the morning was cluster 5, made up of 2 local authorities with less than 200,000 people: Three Rivers, a suburban district north of London, and Fylde, a seaside suburb of Blackpool.
-The auxiliary regressions show that compared to these two places, all the other clusters had a lower percentage of working people in managerial, professional and administrative jobs.???
+However, these are likely to be outliers and may not have many tests from which the clusters are calcualted.
+**Take out smaller clusters!?** The auxiliary regressions show that compared to these two places, all the other clusters had a lower percentage of working people in managerial, professional and administrative jobs.???
 
-This is an indication of the reliability of broadband services in places in both clusters, and suggests that upload speeds have generally been robust throughout the study period.
+This is an indication of the reliability **or maybe not - 5% slowdown from low base can be noticeable, and sd of cluster 6 bigger** of broadband services in places in both clusters, and suggests that upload speeds have generally been robust throughout the study period.
 The exceptions can be found in seven of the eight much smaller clusters including 35 local authority districts, where AM peak upload speeds are between 6% and 18% slower than PM peak upload speeds, although the mean speeds for each cluster are higher than cluster 6.

### 2020-10-23 paper/v1/broadband.speed.covid.Rmd
Intro drafted. Where says 'insert ref' I think I have some ideas, but if I've added a question mark, I'm not sure. Also, some bits copied across into data and results sections. Needs more work and discussion, but hopefully captures many of our ideas discussed yesterday. I may have more time before you do, in which case I will commit further changes.

--- 2020-10-07 paper/v1/broadband.speed.covid.Rmd
@@
 
-[our Google doc](https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit#heading=h.i5om1o8wpcd9)
+**PARA1**During the pandemic, working from home using Information and Communication Technologies (ICT), whether partially or exclusively, was transformed from a niche means of accessing work, albeit one that had been on a slow, upward trend, to a widespread way of life in many countries (Felstead and Reuschke, 2020; **insert further refs from elsewhere**).
+The ability to work from home or telecommute has enabled millions to retain their jobs and, to a varying extent, maintain productivity during periods of strict lockdown around the world.
+However, this ability has not been evenly distributed socially or spatially, creating a new type of digital divide.
+On one side are those who can work from home and have been able to enjoy both economic resilience and greater personal safety.
+On the other side, previously employed individuals have been forced to accept furlough or redundancy packages unless they are part of the cadre of essential workers, who are potentially at high risk of infection.
+Using the UK as a case study, this paper aims to improve our understanding of the spatial and social dimensions of this new digital divide by assessing the resilience of broadband speeds in terms of both quality and reliability of service, and whether this reinforces or redresses prior digital divisions.
 
-**PARA1**: working from home is a challenge and maybe an opportunity [use arguments from @BUDNITZ2020102713]
-It has been altered dramatically because of covid
+**PARA2**The capability to work from home has previously been studied from the perspective of whether work tasks in a given occupation both can be and are allowed to be performed independently of location or co-location with colleagues, including supervisors (**insert refs**).
+However, successful telecommuting also requires that the quality and reliability of ICT services, particularly home internet connection speeds, enable the completion of work tasks with a minimum of delay or interruption.
+In reality, the performance of broadband speeds with respect to telecommuters has never been tested before at scale, as working from home and connecting to colleagues and workplace resources via broadband has previously been the purview of a small minority of workers.
+In the UK, the most recent statistics indicate that **x%** of working age adults identified their home as their usual workplace, with an additional **x%** who said they work from home at least once a week (**insert stats from 2018 or 2019 NTS - also x% of these use ICT to complete their work at home**).
+Their demands on internet services have been minor compared to the demands of leisure users, such that the broadband performance offered by different Internet Service Providers (ISPs) has been benchmarked according to download speeds during the evening 'primetime', when video streaming services are at their peak, rather than during the working  day (**OfCom, 2017**).
+Yet the pandemic has fundamentally changed not only how many people work from home - 47% of working age adults at the peak in April, 2020 (**ref**) - but also their technical requirements when doing so.
+Whilst it may have been common in the past for telecommuters to complete solitary work tasks at home, but still visit their workplace for meetings (**ref?**), during the pandemic the replacement of face-to-face contact with bandwidth-intensive video conference calling was seen as essential to suppressing the spread of infection.
+This type of broadband use can particularly affect upload speeds **ref?**, so we seek to develop a new benchmark (**is it a benchmark?**)to assess the quality and reliability of upload speeds by time of day and day of the week across local authorities in the UK during the time when the population were told to work from home if at all possible.
 
-**PARA2**: covind and working from home, cities, urban structure
+**PARA3**The experience of upload speeds is not the same as the maximum speeds offered by an ISP or possible speeds over a particular type and length of connection.
+Therefore, we use volunteered geographic information in the form of speed-checks run by users to test their experienced broadband speeds, upload and download.
+Meanwhile, the quality and reliability of upload speeds, like ICT more generally, vary in time and space due to both supply and demand-side influences, and can be measured in a number of ways.
+These include: a) mean, experienced connection speed, b) standard deviation or the amount of fluctuation from the mean, and c) the variation in speeds at particular times of day when working from home is more likely to take place.
+We take account of all three measurements in order to describe upload speeds as fully as possible.
+We start by calculating the mean upload speed for every local government district in the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank holidays.
+We then cluster the local authorities by these temporal profiles, allowing us to identify patterns and describe the overall means, standard deviations, and other relevant statistics for each spatial cluster.
+Thus we aim to answer the first part of our research question: How resilient are broadband speeds as experienced in different parts of the UK during a time of extreme demand?
+
+**PARA4**The cause of these different experiences of broadband resilience may be different in different areas, as they may reflect similarities in patterns of demand or similar quality of infrastructure.
+Our approach is also limited by potential endogeneity, as for example, better quality connections with high mean speeds may enable more working from home, but greater demand may cause slower speeds, less reliability and greater variability of speed at different times of day or week.
+Therefore, we avoid attributing any cause to our analysis of the experienced level of quality and reliability of upload speeds.
+Instead, we run auxilliary regressions in order to understand how the spatial and temporal patterns of internet service relate to the economic geography of the UK.
+We discuss how the different patterns might support or undermine efforts to work from home and maintain safe productivity and whether they reinforce existing spatial and social inequalities.
+From this analysis, we hope to provide a greater understanding of how telecommuting and technology intersect at a time of extreme demand, and what lessons this time has for a future where telecommuting is likely to remain a common means of accessing work and broadband services, as well as infrastructure, must be fit for purpose.
+
+**PARA5*The structure of this paper is as follows.
+
+# Literature review
+
+## broadband studies, digital divides, broadband tech stuff - benchmarking?
+
+## from telecommuting to #WFH - Introduce space-time geography - circumstances where work can be carried out more flexibly in space and time- how relate to digital divides - capability theory of mobility?
+suppressed demand...
+
+##covid and working from home, cities, urban structure - economic geography of UK
 Potentially useful readings for covid and cities:
@@
 - [productivity and WFM](https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity)
-
-It is important to become aware of working from home patterns because of:
-- transport planning reasons (see para 1)
-- WFM might increase coparing to pre-covid level
-
-**PARA3**: How can we observe observe WFM? 
-Directly through surveys (@felstead2020homeworking).
-Add survey limitations (cost, reprentativeness, time lag between survey and data availability)
-Data gap.
-Passive data collection through internet speeds
-Another option could be mobility data.
-How is internet speed data a better source than mobility?
-
-**PARA4**: Contention
-
-**PARA5**: Data and methods
-
-**PARA6**: Contribution:
-
-# Literature review
-
-## broadband studies, divides, broadband tech stuff
-
-## from telecommuting to #WFH
 
@@
 
-Data details and some figures, descriptive stats
+The primary data analysed in this paper was provided by Speedchecker Ltd, a private company that allows internet users to check their own broadband upload / download speeds, and stores every speed-check with a timestamp and geolocation.
+
+Our approach involves aggregating all the speed-checks during the 13 weeks of March to May inclusive for weekdays in 2020 by each hour of the day and day of the week.
+As our research aims to identify the geography of internet service resilience for work purposes, bank holidays and the hours between midnight and 6am were also excluded.
+The composite week time series thus comprise 18 hours multiplied by 5 weekdays or 90 time points per series.
+These time series were calculated for each of the 382 Local Authority Districts (LADs) in the UK, standardised, and then a k-means clustering algorithm was applied, including dynamic time warping.
+The LADs were assigned to 10 clusters for upload speeds.
+The cluster id for each LAD is then reattached to the speed-check dataframe to identify the characteristics of each cluster, including number of LADs, and the descriptives statistics of upload speeds in that cluster, and the temporal profile by hour of the day and day of the week.
+
+**Data details and some figures, descriptive stats - include whole sample time profile for 2019 and 2020 frequency of tests run as part of why we chose to create the time profiles by hour of the day and day of the week rather than daily over the whole period.**
 
@@
 
-Clusters, cluster description and aux regressions
+##Upload Clusters / cluster description
+
+The temporal profiles used to cluster the local authorities have been summarised in graph [.], which shows a composite profile of mean upload speeds per hour per day for each cluster.
+For upload speeds, 345 of 382 local authorities, or over 62 million people, fall into cluster 6 or cluster 9.
+Graph [.] shows that both of these clusters have relatively similar temporal profiles, which are flatter than the other, smaller clusters, suggesting better reliability of service.
+However, the upload speeds at all times for cluster 9 are substantially higher than for cluster 6, which is an indication of better quality of service.
+This difference is reflected not only in the mean speeds for these clusters for the whole sample, but also the mean upload speeds in the morning peak from 9:00-10:59, as well as the evening peak period from 19:00-20:59.
+In comparison, the time profile in graph [.] shows upload speeds in cluster 1 are on average lower at certain times of day during the study period than any other cluster, whilst the profile for cluster 3 appears to show speeds fluctuating as much as cluster 1, but at levels usually higher even than cluster 9.
+
+The variability of the smaller clusters may be related to the fewer speed tests from fewer local authorities that have been averaged, whilst averaging greater numbers of speed tests could artificially flatten the profile.
+This appears not to be the case for cluster 9, as there is negligible difference between morning and evening upload speeds, at 1% slower in the morning, confirming a high level of reliability of service.
+In comparison, upload speeds in cluster 6 are 4% slower in the morning than in the evening and experienced the joint highest ratio of standard deviation to mean across the time period under assessment.
+This suggests that although the time profile is relatively flat in graph [.], the experience is one of speeds that fluctuate from a lower mean, and therefore might more often impact on online activities.
+Still, in most of the smaller clusters the reliability of service is worse than in cluster 9 in terms of the ratio of standard deviation to mean and worse than cluster 6 as well in terms of the ratio of upload speed in the morning peak compared to the evening peak.
+In five of the smaller clusters, which are home to almost 3 million people, speeds are 12-19% slower between 9:00-10:59 compared to 19:00-20:59.
+
+The quality and reliability of broadband service is thus much better in the 115 Local Authority areas in cluster 9, which are mostly in urban or suburban areas, compared to cluster 6.
+These include 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation.
+There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, and many suburban districts from the South East of England to South Tyneside.
+Meanwhile, the 230 local authorities in cluster 6, which have lower speeds on average and more variation in service still include major urban areas, such as Bristol, Liverpool and Leeds, and many suburban areas, but also include some of the most rural areas in the country.
+Meanwhile, Cluster 1, with 10 local authorities that are home to over 1 million people has the second slowest speeds in the morning compared to the evening 'peaks' and the second highest ratio of standard deviation to mean.
+This cluster's most populous area is Westminster in central London.
+
+**further description of temporal profiles of other clusters - include all in graph?** Indeed, the only cluster where upload speeds were slower in the evening than in the morning was cluster 5, made up of 2 local authorities with less than 200,000 people: Three Rivers, a suburban district north of London, and Fylde, a seaside suburb of Blackpool.
+However, these are likely to be outliers and may not have many tests from which the clusters are calcualted.
+**Take out smaller clusters?** The exceptions can be found in seven of the eight much smaller clusters including 35 local authority districts, where AM peak upload speeds are between 6% and 18% slower than PM peak upload speeds, although the mean speeds for each cluster are higher than cluster 6.
+Indeed, in 25 local authorities with a combined population of almost 3 million, speeds are 13% or more slower in the morning than in the evening.
+Included in this latter group are central London borough of Westminster and the London Borough of Newham, rural authorities like Eden and West Devon, and small cities like Dundee and Carlisle.
+
+##aux regressions
+
+Auxiliary regressions indicate that the speed tests in cluster 9 authorities are also more likely to have been run on services provided by Virgin Media, suggesting they are in the half of the country with the most lucrative ICT market, which originally attracted the cable TV provider (OfCom...).
+For example, although auxiliary regressions show that Cluster 6 local authorities are more likely to be in the South of the UK than Cluster 9, the cluster notably includes Southern rural districts from Cornwall to North Norfolk.
+Slower speeds could reflect the lower quality of service in rural areas compared to urban and suburban areas.
+There is frustration, however, as auxiliary regressions show that those living in cluster 6 had the highest probability of testing their broadband between 9:00 and 11:00 of any of the 10 upload speed clusters.
+
+the finding in the auxiliary regressions that those in cluster 9 ran the fewest speed tests per person during the morning period of any cluster.
+Now this may be an indication of fewer people working from home, less contention, and less resultant frustration.
+Cluster 9 comprises many central urban areas and has the lowest number of established businesses per inhabitant, which could be interpreted as a dominance of large employers.
+However, the job density is lower than cluster 6, meaning there are not as many jobs per resident in these areas.
+As the cluster includes many suburban areas too, which may be largely residential, could the quality and reliability of internet service be reinforcing patterns of telecommuting by those in wealthier suburbs who can work from home?
+Earnings in cluster 9 are second highest of all the clusters, with only Cluster 2 (comprising just North East Lincolnshire and East Lothian, population 265k) earning more per person.
+
+The auxiliary regressions show that compared to the two authorities in cluster 5, all the other clusters had a lower percentage of working people in managerial, professional and administrative jobs.???
+Yet the auxiliary regression suggests that there are not many tests being run during the am peak in cluster 1.
+This may be because there are fewer people working at home checking their broadband than in most other clusters.
+Indeed, the auxiliary regressions indicate that cluster 1 has the highest job density or proportion of jobs to working-age population, which is likely to due to the presence of Westminster, central London, cluster 1's most populous local authority.
+Westminster not only has more workplaces than residents, but it is reasonable to presume that many who would normally work in Westminster, but be able to work from home during lockdown are likely to live outside central London and not be subject to the fluctuating speeds there.
+Workplaces, meanwhile, some of which would still have been open, could be running programmes that cause the slowdown and variation, but would be more likely to have their own in-house diagnostics, rather than using a service like Speedchecker Ltd.
 

### 2020-11-12 paper/v1/broadband.speed.covid.Rmd
I wrote a page for the intro, but I am not 100% sure we nailed the RQ. Our aim makes sense (understand telecomuting as a 2 dimensional thing) but I think we need to further clarify the RQ. Maybe a quick call after you read it?

--- 2020-10-21 paper/v1/broadband.speed.covid.Rmd
@@
 
-<!-- **PARA1**: working from home is a challenge and maybe an opportunity [use arguments from @BUDNITZ2020102713] -->
-<!-- It has been altered dramatically because of covid -->
-
 During the pandemic, working from home using Information and Communication
@@
 delay or interruption.
-Therefore, in this paper, we aim to look at patterns of broadband speed tests
+Therefore, in this paper, we look at patterns of broadband speed tests
 and upload speeds in the UK in order to gain insights into both the quality and
@@
 possible.
-From this analysis, we provide a greater understanding of how telecommuting and
+To do so, we employ unique data regarding individual broadband speed tests and
+state of the art time series clustering methods.
+Our analysis enables us to better understand how telecommuting and
 technology intersect at a time of extreme demand, and what lessons this time has
@@
 work and broadband services, as well as infrastructure, must be fit for purpose.
+**I THINK WE NEED A MORE CLEAR RESEARCH QUESTION**
 
-**PARA2**: covind and working from home, cities, urban structure
-Potentially useful readings for covid and cities:
+<!-- How much do we work form home-->
+The shift towards telecommuting during various stages of lockdown around the world
+has been drastic and, importantly, speculations indicate that the post-Covid
+tendency to work from home will be much higher than the pre-Covid one.
+A back of the envelope calculation suggests that up to 40% of the working force
+could work from home in the UK [@batty2020editorial].
+Observational data pointed an even higher share of people in employment in the
+UK who worked from home in April $2020$ ($47$%), while the same figure only reached
+$5$% the year before [@ons2020; @ons2020lm2019].
+Similar figures have been reported for other countries around the world
+[@felstead2020homeworking].
+For instance, $37$% of the workforce worked from home in Europe in April $2020$ with
+countries like Finland reaching $60$% [@eurofound2020].
+In the US almost half of the working population worked from home during the same
+period because of the pandemic [@brynjolfsson2020covid].
+More broadly, a recent estimate indicated that $37$% of all jobs in the US can be
+permanently performed entirely from home [@NBERw26948].
 
-- https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas
-- https://www.coronavirusandtheeconomy.com/question/will-coronavirus-cause-big-city-exodus
-- [EPB commentaries](https://journals.sagepub.com/toc/EPB/current)
-- https://journals.sagepub.com/doi/full/10.1177/2399808320926912?casa_token=gh3SgKFCZ44AAAAA:MjjMWJ61DqlmxzkWQyy_wxPnU20QdyQYkC4fVFpHLPfLk7McwmQGkJ7x2Q7LZOjTf6vcYaqNwug
-- [WFM and covid, econ](https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home)
-- [productivity and WFM](https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity)
+<!-- Who works from home-->
+There is a consensus that opportunities for working from home especially during
+the current pandemic are not equally spread across the workforce.
+@NBERw26948 indicated that in the US managers, educators, as well as those
+working in computer-related occupations, finance, and law can easily work from
+home.
+On the contrary, workers in farming, construction, and manufacturing do not have
+such opportunities.
+Not surprisingly, occupations with opportunities to telecommute are associated
+with higher earnings.
+This is not the case for the workforce occupied in less footloose occupations as
+as they tend to be lower-income, non-white, without a university degree,
+live in rental accommodation and lack health insurance [@NBERw27085].
+Although these figures refer to the US, similar trends can be observed for other
+countries.
+For example, $75$% of workers with tertiary education worked from home in Europe
+during spring $2020$, while the same share for workers with secondary and
+primary education dropped to $34$% and $14$% respectively.
+Moreover, employees living in cities, women and younger employees were
+have worked from home [@eurofound2020].
 
-It is important to become aware of working from home patterns because of:
-- transport planning reasons (see para 1)
-- WFM might increase coparing to pre-covid level
+<!-- ICT infrastructure-->
+None of these changes could have happened in the absence of ICT
+infrastructure -- both in terms of software and hardware.
+But while innovations in software are easily diffused across space and
+society^[See for example the huge success of videoconferencing apps
+[@marks2020zoom].],
+the same does not apply for ICT hardware infrastructure such as internet
+broadband connectivity.
+The literature exemplifies digital divides in terms of internet access and its
+quality.
+For instance, @riddlesden2014broadband highlighted the broadband divides in the
+UK, while the systematic review from @SALEMINK2017360 reinforced our
+understanding for the infrastructure quality differences between urban and rural
+areas.
 
-**PARA3**: How can we observe observe WFM?
-Directly through surveys (@felstead2020homeworking).
-Add survey limitations (cost, reprentativeness, time lag between survey and data availability)
+<!-- WFH and digital divides-->
+Our framework to understand telecommuting as a function of occupations and
+quality of ICT infrastructure is aligned with current debates on digital divides.
+While the, so-called, first level digital divides are associated with access
+and quality of internet connectivity, the second level ones are linked
+to the necessary skills to effectively utilise ICT and the internet
+[@blank2014dimensions; @van2011internet].
+Importantly, the differentiated capacity to telecommute, which to a certain
+extend are related to the first and second level digital divides, leads to
+differentiated outcomes regarding the economic resilience of people and
+places against the current pandemic.
+And to the extend that the quality of the internet infrastructure and occupation
+variation are spatially dependent and clustered in space, the spatial footprint
+of telecommuting is of great interest. In a way similar to the third level
+digital divide, which
+focuses on the differentiated returns of internet use [@stern2009levels;
+@van2014digital; @van2015third] places with high rates of telecommuting during
+the Covid pandemic illustrate higher economic resilience against the current pandemic.
+<!-- Econ resilience-->
+This is aligned with the regional economic resilience literature, which
+underlines the differentiated capacity of cities and regions to escape or
+recover from economic crises [@martin2012regional; @kitsos2018economic].
+
+<!-- WFH broader consequences-->
+The long-term effects of such drastic changes in telecommuting and attitudes
+towards working from home are difficult to predict. Nevertheless,
+they might span through various aspects of economy and society: from changes to
+transportation planning due to changes in commuting patters to changes in land
+use and urban planning to accommodate people who work from home
+(@BUDNITZ2020102713); and from productivity and innovation changes to a change
+in agglomeration externalities and the attraction of large cities [@econobs]
+just to name a few. This paper is positioned to support such endeavours by
+providing an explanatory **[NOTE SURE ABOUT THIS]** framework  based on
+the quality of internet connectivity and the frequencies of occupations within
+Local Authorities in the UK.
+
+<!-- How can we observe observe WFM? -->
+<!-- removed for now -->
+<!-- Directly through surveys (@felstead2020homeworking).
+Add survey limitations (cost, representativeness, time lag between survey and data availability)
 Data gap.
@@
 Another option could be mobility data.
-How is internet speed data a better source than mobility?
+How is internet speed data a better source than mobility?-->
 
-**PARA4**: Contention
+**MORE SOURCES:**
+
+- https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas
+- [EPB commentaries](https://journals.sagepub.com/toc/EPB/current)
+- https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home
+- https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity
+- https://www.coronavirusandtheeconomy.com/question/how-will-economic-effects-coronavirus-vary-across-areas-uk
+- https://www.coronavirusandtheeconomy.com/which-parts-uk-have-been-hit-hardest-covid-19-crisis
+- https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas
 
@@
 - check who cites the above and what they cite
+
+In this analysis, the terms ‘teleworking’ and ‘working from home’ are used interchangeably. An appropriate definition of teleworking is ‘the remote
+provision of labour that would otherwise be carried out within company premises’ (European Commission, 2020b). In practice, during the COVID crisis,
+most such work was carried out in the homes of individual employees rather than any other location.
+
+**PARA about Contention**
 

### 2020-11-12 paper/v1/broadband.speed.covid.tex
I wrote a page for the intro, but I am not 100% sure we nailed the RQ. Our aim makes sense (understand telecomuting as a 2 dimensional thing) but I think we need to further clarify the RQ. Maybe a quick call after you read it?

@@
 
-\textbf{PARA1}: working from home is a challenge and maybe an
-opportunity \citep[use arguments from][]{BUDNITZ2020102713} It has been
-altered dramatically because of covid
+During the pandemic, working from home using Information and
+Communication Technologies (ICT), whether partially or exclusively, was
+transformed from a niche means of accessing work, albeit one that had
+been on a slow, upward trend, to a widespread way of life in many
+countries. The ability to work from home or telecommute meant millions
+retained their jobs and, to a varying extent, maintained productivity
+during periods of strict lockdown around the world. However, this
+ability was not evenly distributed socially or spatially, creating a new
+type of digital divide -- one of economic resilience and personal safety
+on one side and unemployment or essential front-line work on the other.
+The ability to telecommute is partly a product of whether the majority
+of work tasks in an occupation can be performed independently of being
+in a particular location or co-locating with colleagues. It is also a
+product of the quality of ICT services, particularly home internet
+connections, sufficient to successfully complete work tasks with a
+minimum of delay or interruption. Therefore, in this paper, we look at
+patterns of broadband speed tests and upload speeds in the UK in order
+to gain insights into both the quality and reliability of broadband
+services across the country and the distribution of telecommuting at the
+time when the population were told to work from home if possible. To do
+so, we employ unique data regarding individual broadband speed tests and
+state of the art time series clustering methods. Our analysis enables us
+to better understand how telecommuting and technology intersect at a
+time of extreme demand, and what lessons this time has for a future
+where telecommuting is likely to remain a common means of accessing work
+and broadband services, as well as infrastructure, must be fit for
+purpose. \textbf{I THINK WE NEED A MORE CLEAR RESEARCH QUESTION}
 
-\textbf{PARA2}: covind and working from home, cities, urban structure
-Potentially useful readings for covid and cities:
+The shift towards telecommuting during various stages of lockdown around
+the world has been drastic and, importantly, speculations indicate that
+the post-Covid tendency to work from home will be much higher than the
+pre-Covid one. A back of the envelope calculation suggests that up to
+40\% of the working force could work from home in the UK
+\citep{batty2020editorial}. Observational data pointed an even higher
+share of people in employment in the UK who worked from home in April
+\(2020\) (\(47\)\%), while the same figure only reached \(5\)\% the year
+before \citep{ons2020, ons2020lm2019}. Similar figures have been
+reported for other countries around the world
+\citep{felstead2020homeworking}. For instance, \(37\)\% of the workforce
+worked from home in Europe in April \(2020\) with countries like Finland
+reaching \(60\)\% \citep{eurofound2020}. In the US almost half of the
+working population worked from home during the same period because of
+the pandemic \citep{brynjolfsson2020covid}. More broadly, a recent
+estimate indicated that \(37\)\% of all jobs in the US can be
+permanently performed entirely from home \citep{NBERw26948}.
+
+There is a consensus that opportunities for working from home especially
+during the current pandemic are not equally spread across the workforce.
+\citet{NBERw26948} indicated that in the US managers, educators, as well
+as those working in computer-related occupations, finance, and law can
+easily work from home. On the contrary, workers in farming,
+construction, and manufacturing do not have such opportunities. Not
+surprisingly, occupations with opportunities to telecommute are
+associated with higher earnings. This is not the case for the workforce
+occupied in less footloose occupations as as they tend to be
+lower-income, non-white, without a university degree, live in rental
+accommodation and lack health insurance \citep{NBERw27085}. Although
+these figures refer to the US, similar trends can be observed for other
+countries. For example, \(75\)\% of workers with tertiary education
+worked from home in Europe during spring \(2020\), while the same share
+for workers with secondary and primary education dropped to \(34\)\% and
+\(14\)\% respectively. Moreover, employees living in cities, women and
+younger employees were have worked from home \citep{eurofound2020}.
+
+None of these changes could have happened in the absence of ICT
+infrastructure -- both in terms of software and hardware. But while
+innovations in software are easily diffused across space and
+society\footnote{See for example the huge success of videoconferencing
+  apps \citep{marks2020zoom}.}, the same does not apply for ICT hardware
+infrastructure such as internet broadband connectivity. The literature
+exemplifies digital divides in terms of internet access and its quality.
+For instance, \citet{riddlesden2014broadband} highlighted the broadband
+divides in the UK, while the systematic review from
+\citet{SALEMINK2017360} reinforced our understanding for the
+infrastructure quality differences between urban and rural areas.
+
+Our framework to understand telecommuting as a function of occupations
+and quality of ICT infrastructure is aligned with current debates on
+digital divides. While the, so-called, first level digital divides are
+associated with access and quality of internet connectivity, the second
+level ones are linked to the necessary skills to effectively utilise ICT
+and the internet \citep{blank2014dimensions, van2011internet}.
+Importantly, the differentiated capacity to telecommute, which to a
+certain extend are related to the first and second level digital
+divides, leads to differentiated outcomes regarding the economic
+resilience of people and places against the current pandemic. And to the
+extend that the quality of the internet infrastructure and occupation
+variation are spatially dependent and clustered in space, the spatial
+footprint of telecommuting is of great interest. In a way similar to the
+third level digital divide, which focuses on the differentiated returns
+of internet use \citep{stern2009levels, van2014digital, van2015third}
+places with high rates of telecommuting during the Covid pandemic
+illustrate higher economic resilience against the current pandemic. This
+is aligned with the regional economic resilience literature, which
+underlines the differentiated capacity of cities and regions to escape
+or recover from economic crises
+\citep{martin2012regional, kitsos2018economic}.
+
+The long-term effects of such drastic changes in telecommuting and
+attitudes towards working from home are difficult to predict.
+Nevertheless, they might span through various aspects of economy and
+society: from changes to transportation planning due to changes in
+commuting patters to changes in land use and urban planning to
+accommodate people who work from home (\citet{BUDNITZ2020102713}); and
+from productivity and innovation changes to a change in agglomeration
+externalities and the attraction of large cities \citep{econobs} just to
+name a few. This paper is positioned to support such endeavours by
+providing an explanatory \textbf{{[}NOTE SURE ABOUT THIS{]}} framework
+based on the quality of internet connectivity and the frequencies of
+occupations within Local Authorities in the UK.
+
+\textbf{MORE SOURCES:}
 
@@
 \item
-  \url{https://www.coronavirusandtheeconomy.com/question/will-coronavirus-cause-big-city-exodus}
-\item
   \href{https://journals.sagepub.com/toc/EPB/current}{EPB commentaries}
 \item
-  \url{https://journals.sagepub.com/doi/full/10.1177/2399808320926912?casa_token=gh3SgKFCZ44AAAAA:MjjMWJ61DqlmxzkWQyy_wxPnU20QdyQYkC4fVFpHLPfLk7McwmQGkJ7x2Q7LZOjTf6vcYaqNwug}
+  \url{https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home}
 \item
-  \href{https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home}{WFM
-  and covid, econ}
+  \url{https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity}
 \item
-  \href{https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity}{productivity
-  and WFM}
+  \url{https://www.coronavirusandtheeconomy.com/question/how-will-economic-effects-coronavirus-vary-across-areas-uk}
+\item
+  \url{https://www.coronavirusandtheeconomy.com/which-parts-uk-have-been-hit-hardest-covid-19-crisis}
+\item
+  \url{https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas}
 \end{itemize}
-
-It is important to become aware of working from home patterns because
-of: - transport planning reasons (see para 1) - WFM might increase
-coparing to pre-covid level
-
-\textbf{PARA3}: How can we observe observe WFM? Directly through surveys
-(\citet{felstead2020homeworking}). Add survey limitations (cost,
-reprentativeness, time lag between survey and data availability) Data
-gap. Passive data collection through internet speeds Another option
-could be mobility data. How is internet speed data a better source than
-mobility?
-
-\textbf{PARA4}: Contention
 
@@
 \end{itemize}
+
+In this analysis, the terms `teleworking' and `working from home' are
+used interchangeably. An appropriate definition of teleworking is `the
+remote provision of labour that would otherwise be carried out within
+company premises' (European Commission, 2020b). In practice, during the
+COVID crisis, most such work was carried out in the homes of individual
+employees rather than any other location.
+
+\textbf{PARA about Contention}
 

### 2020-11-16 paper/v1/broadband.speed.covid.Rmd

--- 2020-10-23 paper/v1/broadband.speed.covid.Rmd
@@
 
+HB
 **PARA2**The capability to work from home has previously been studied from the perspective of whether work tasks in a given occupation both can be and are allowed to be performed independently of location or co-location with colleagues, including supervisors (**insert refs**).
@@
 
+ET
+During the pandemic, working from home using Information and Communication
+Technologies (ICT), whether partially or exclusively, was transformed from a
+niche means of accessing work, albeit one that had been on a slow, upward trend,
+to a widespread way of life in many countries.
+The ability to work from home or telecommute meant millions retained their jobs
+and, to a varying extent, maintained productivity during periods of strict
+lockdown around the world.
+However, this ability was not evenly distributed socially or spatially, creating
+a new type of digital divide -- one of economic resilience and personal safety on
+one side and unemployment or essential front-line work on the other.
+The ability to telecommute is partly a product of whether the majority of work
+tasks in an occupation can be performed independently of being in a particular
+location or co-locating with colleagues.
+It is also a product of the quality of ICT services, particularly home internet
+connections, sufficient to successfully complete work tasks with a minimum of
+delay or interruption.
+Therefore, in this paper, we look at patterns of broadband speed tests
+and upload speeds in the UK in order to gain insights into both the quality and
+reliability of broadband services across the country and the distribution of
+telecommuting at the time when the population were told to work from home if
+possible.
+To do so, we employ unique data regarding individual broadband speed tests and
+state of the art time series clustering methods.
+Our analysis enables us to better understand how telecommuting and
+technology intersect at a time of extreme demand, and what lessons this time has
+for a future where telecommuting is likely to remain a common means of accessing
+work and broadband services, as well as infrastructure, must be fit for purpose.
+**I THINK WE NEED A MORE CLEAR RESEARCH QUESTION**
+
+<!-- How much do we work form home-->
+The shift towards telecommuting during various stages of lockdown around the world
+has been drastic and, importantly, speculations indicate that the post-Covid
+tendency to work from home will be much higher than the pre-Covid one.
+A back of the envelope calculation suggests that up to 40% of the working force
+could work from home in the UK [@batty2020editorial].
+Observational data pointed an even higher share of people in employment in the
+UK who worked from home in April $2020$ ($47$%), while the same figure only reached
+$5$% the year before [@ons2020; @ons2020lm2019].
+Similar figures have been reported for other countries around the world
+[@felstead2020homeworking].
+For instance, $37$% of the workforce worked from home in Europe in April $2020$ with
+countries like Finland reaching $60$% [@eurofound2020].
+In the US almost half of the working population worked from home during the same
+period because of the pandemic [@brynjolfsson2020covid].
+More broadly, a recent estimate indicated that $37$% of all jobs in the US can be
+permanently performed entirely from home [@NBERw26948].
+
+<!-- Who works from home-->
+There is a consensus that opportunities for working from home especially during
+the current pandemic are not equally spread across the workforce.
+@NBERw26948 indicated that in the US managers, educators, as well as those
+working in computer-related occupations, finance, and law can easily work from
+home.
+On the contrary, workers in farming, construction, and manufacturing do not have
+such opportunities.
+Not surprisingly, occupations with opportunities to telecommute are associated
+with higher earnings.
+This is not the case for the workforce occupied in less footloose occupations as
+as they tend to be lower-income, non-white, without a university degree,
+live in rental accommodation and lack health insurance [@NBERw27085].
+Although these figures refer to the US, similar trends can be observed for other
+countries.
+For example, $75$% of workers with tertiary education worked from home in Europe
+during spring $2020$, while the same share for workers with secondary and
+primary education dropped to $34$% and $14$% respectively.
+Moreover, employees living in cities, women and younger employees were
+have worked from home [@eurofound2020].
+
+<!-- ICT infrastructure-->
+None of these changes could have happened in the absence of ICT
+infrastructure -- both in terms of software and hardware.
+But while innovations in software are easily diffused across space and
+society^[See for example the huge success of videoconferencing apps
+[@marks2020zoom].],
+the same does not apply for ICT hardware infrastructure such as internet
+broadband connectivity.
+The literature exemplifies digital divides in terms of internet access and its
+quality.
+For instance, @riddlesden2014broadband highlighted the broadband divides in the
+UK, while the systematic review from @SALEMINK2017360 reinforced our
+understanding for the infrastructure quality differences between urban and rural
+areas.
+
+<!-- WFH and digital divides-->
+Our framework to understand telecommuting as a function of occupations and
+quality of ICT infrastructure is aligned with current debates on digital divides.
+While the, so-called, first level digital divides are associated with access
+and quality of internet connectivity, the second level ones are linked
+to the necessary skills to effectively utilise ICT and the internet
+[@blank2014dimensions; @van2011internet].
+Importantly, the differentiated capacity to telecommute, which to a certain
+extend are related to the first and second level digital divides, leads to
+differentiated outcomes regarding the economic resilience of people and
+places against the current pandemic.
+And to the extend that the quality of the internet infrastructure and occupation
+variation are spatially dependent and clustered in space, the spatial footprint
+of telecommuting is of great interest. In a way similar to the third level
+digital divide, which
+focuses on the differentiated returns of internet use [@stern2009levels;
+@van2014digital; @van2015third] places with high rates of telecommuting during
+the Covid pandemic illustrate higher economic resilience against the current pandemic.
+<!-- Econ resilience-->
+This is aligned with the regional economic resilience literature, which
+underlines the differentiated capacity of cities and regions to escape or
+recover from economic crises [@martin2012regional; @kitsos2018economic].
+
+<!-- WFH broader consequences-->
+The long-term effects of such drastic changes in telecommuting and attitudes
+towards working from home are difficult to predict. Nevertheless,
+they might span through various aspects of economy and society: from changes to
+transportation planning due to changes in commuting patters to changes in land
+use and urban planning to accommodate people who work from home
+(@BUDNITZ2020102713); and from productivity and innovation changes to a change
+in agglomeration externalities and the attraction of large cities [@econobs]
+just to name a few. This paper is positioned to support such endeavours by
+providing an explanatory **[NOTE SURE ABOUT THIS]** framework  based on
+the quality of internet connectivity and the frequencies of occupations within
+Local Authorities in the UK.
+
+<!-- How can we observe observe WFM? -->
+<!-- removed for now -->
+<!-- Directly through surveys (@felstead2020homeworking).
+Add survey limitations (cost, representativeness, time lag between survey and data availability)
+Data gap.
+Passive data collection through internet speeds
+Another option could be mobility data.
+How is internet speed data a better source than mobility?-->
+
+**MORE SOURCES:**
+
+- https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas
+- [EPB commentaries](https://journals.sagepub.com/toc/EPB/current)
+- https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home
+- https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity
+- https://www.coronavirusandtheeconomy.com/question/how-will-economic-effects-coronavirus-vary-across-areas-uk
+- https://www.coronavirusandtheeconomy.com/which-parts-uk-have-been-hit-hardest-covid-19-crisis
+- https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas
+
+**PARA5**: Data and methods
+
+**PARA6**: Contribution:
+
+# Literature review
+
+## broadband studies, divides, broadband tech stuff
+
+## from telecommuting to #WFH
+
 Some new papers google recommended to me:

### 2020-11-16 paper/v1/broadband.speed.covid.Rmd
merged intros

@@
 
-**PARA1**During the pandemic, working from home using Information and Communication Technologies (ICT), whether partially or exclusively, was transformed from a niche means of accessing work, albeit one that had been on a slow, upward trend, to a widespread way of life in many countries (Felstead and Reuschke, 2020; **insert further refs from elsewhere**).
-The ability to work from home or telecommute has enabled millions to retain their jobs and, to a varying extent, maintain productivity during periods of strict lockdown around the world.
-However, this ability has not been evenly distributed socially or spatially, creating a new type of digital divide.
-On one side are those who can work from home and have been able to enjoy both economic resilience and greater personal safety.
-On the other side, previously employed individuals have been forced to accept furlough or redundancy packages unless they are part of the cadre of essential workers, who are potentially at high risk of infection.
-Using the UK as a case study, this paper aims to improve our understanding of the spatial and social dimensions of this new digital divide by assessing the resilience of broadband speeds in terms of both quality and reliability of service, and whether this reinforces or redresses prior digital divisions.
+- [our google doc](https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit#heading=h.i5om1o8wpcd9)
 
-HB
-**PARA2**The capability to work from home has previously been studied from the perspective of whether work tasks in a given occupation both can be and are allowed to be performed independently of location or co-location with colleagues, including supervisors (**insert refs**).
-However, successful telecommuting also requires that the quality and reliability of ICT services, particularly home internet connection speeds, enable the completion of work tasks with a minimum of delay or interruption.
-In reality, the performance of broadband speeds with respect to telecommuters has never been tested before at scale, as working from home and connecting to colleagues and workplace resources via broadband has previously been the purview of a small minority of workers.
-In the UK, the most recent statistics indicate that **x%** of working age adults identified their home as their usual workplace, with an additional **x%** who said they work from home at least once a week (**insert stats from 2018 or 2019 NTS - also x% of these use ICT to complete their work at home**).
-Their demands on internet services have been minor compared to the demands of leisure users, such that the broadband performance offered by different Internet Service Providers (ISPs) has been benchmarked according to download speeds during the evening 'primetime', when video streaming services are at their peak, rather than during the working  day (**OfCom, 2017**).
-Yet the pandemic has fundamentally changed not only how many people work from home - 47% of working age adults at the peak in April, 2020 (**ref**) - but also their technical requirements when doing so.
-Whilst it may have been common in the past for telecommuters to complete solitary work tasks at home, but still visit their workplace for meetings (**ref?**), during the pandemic the replacement of face-to-face contact with bandwidth-intensive video conference calling was seen as essential to suppressing the spread of infection.
-This type of broadband use can particularly affect upload speeds **ref?**, so we seek to develop a new benchmark (**is it a benchmark?**)to assess the quality and reliability of upload speeds by time of day and day of the week across local authorities in the UK during the time when the population were told to work from home if at all possible.
-
-**PARA3**The experience of upload speeds is not the same as the maximum speeds offered by an ISP or possible speeds over a particular type and length of connection.
-Therefore, we use volunteered geographic information in the form of speed-checks run by users to test their experienced broadband speeds, upload and download.
-Meanwhile, the quality and reliability of upload speeds, like ICT more generally, vary in time and space due to both supply and demand-side influences, and can be measured in a number of ways.
-These include: a) mean, experienced connection speed, b) standard deviation or the amount of fluctuation from the mean, and c) the variation in speeds at particular times of day when working from home is more likely to take place.
-We take account of all three measurements in order to describe upload speeds as fully as possible.
-We start by calculating the mean upload speed for every local government district in the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank holidays.
-We then cluster the local authorities by these temporal profiles, allowing us to identify patterns and describe the overall means, standard deviations, and other relevant statistics for each spatial cluster.
-Thus we aim to answer the first part of our research question: How resilient are broadband speeds as experienced in different parts of the UK during a time of extreme demand?
-
-**PARA4**The cause of these different experiences of broadband resilience may be different in different areas, as they may reflect similarities in patterns of demand or similar quality of infrastructure.
-Our approach is also limited by potential endogeneity, as for example, better quality connections with high mean speeds may enable more working from home, but greater demand may cause slower speeds, less reliability and greater variability of speed at different times of day or week.
-Therefore, we avoid attributing any cause to our analysis of the experienced level of quality and reliability of upload speeds.
-Instead, we run auxilliary regressions in order to understand how the spatial and temporal patterns of internet service relate to the economic geography of the UK.
-We discuss how the different patterns might support or undermine efforts to work from home and maintain safe productivity and whether they reinforce existing spatial and social inequalities.
-From this analysis, we hope to provide a greater understanding of how telecommuting and technology intersect at a time of extreme demand, and what lessons this time has for a future where telecommuting is likely to remain a common means of accessing work and broadband services, as well as infrastructure, must be fit for purpose.
-
-**PARA5*The structure of this paper is as follows.
-
-# Literature review
-
-## broadband studies, digital divides, broadband tech stuff - benchmarking?
-
-## from telecommuting to #WFH - Introduce space-time geography - circumstances where work can be carried out more flexibly in space and time- how relate to digital divides - capability theory of mobility?
-suppressed demand...
-
-##covid and working from home, cities, urban structure - economic geography of UK
-Potentially useful readings for covid and cities:
-
-- https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas
-- https://www.coronavirusandtheeconomy.com/question/will-coronavirus-cause-big-city-exodus
-- [EPB commentaries](https://journals.sagepub.com/toc/EPB/current)
-- https://journals.sagepub.com/doi/full/10.1177/2399808320926912?casa_token=gh3SgKFCZ44AAAAA:MjjMWJ61DqlmxzkWQyy_wxPnU20QdyQYkC4fVFpHLPfLk7McwmQGkJ7x2Q7LZOjTf6vcYaqNwug
-- [WFM and covid, econ](https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home)
-- [productivity and WFM](https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity)
-
-ET
 During the pandemic, working from home using Information and Communication
@@
 niche means of accessing work, albeit one that had been on a slow, upward trend,
-to a widespread way of life in many countries.
+to a widespread way of life in many countries [@felstead2020homeworking].
+**HANNAH, DO WE REALLY NEED A REFERENCE IN THE PREVIOUS SENTENCE?**
 The ability to work from home or telecommute meant millions retained their jobs
@@
 lockdown around the world.
-However, this ability was not evenly distributed socially or spatially, creating
+<!--OLD TEXT: However, this ability was not evenly distributed socially or spatially, creating
 a new type of digital divide -- one of economic resilience and personal safety on
@@
 tasks in an occupation can be performed independently of being in a particular
-location or co-locating with colleagues.
-It is also a product of the quality of ICT services, particularly home internet
-connections, sufficient to successfully complete work tasks with a minimum of
-delay or interruption.
-Therefore, in this paper, we look at patterns of broadband speed tests
-and upload speeds in the UK in order to gain insights into both the quality and
+location or co-locating with colleagues.-->
+However, this ability has not been evenly distributed socially or spatially,
+creating a new type of digital divide.
+On one side are those who can work from home and have been able to enjoy both
+economic resilience and greater personal safety.
+On the other side, previously employed individuals have been forced to accept
+furlough or redundancy packages unless they are part of the cadre of essential
+workers, who are potentially at high risk of infection.
+Using the UK as a case study, this paper aims to improve our understanding of
+the spatial and social dimensions of this new digital divide by assessing the
+resilience of broadband speeds in terms of both quality and reliability of
+service, and whether this reinforces or redresses prior digital divisions.
+<!--and upload speeds in the UK in order to gain insights into both the quality and
 reliability of broadband services across the country and the distribution of
 telecommuting at the time when the population were told to work from home if
-possible.
-To do so, we employ unique data regarding individual broadband speed tests and
-state of the art time series clustering methods.
+possible.-->
+To do so, we employ unique volunteered geographic data on individual broadband
+speed tests and state-of-the-art time-series clustering methods, which enable us
+to create clusters of UK local authorities with similar temporal signatures of
+experienced internet speeds.
+We then associate these clusters of local authorities with their socioeconomic
+and geographic characteristics to explore whether they are linked with existing
+divides and the economic geography of the UK.
+**HANNAH, HOW DOES THIS SOUND?**
 Our analysis enables us to better understand how telecommuting and
@@
 work and broadband services, as well as infrastructure, must be fit for purpose.
-**I THINK WE NEED A MORE CLEAR RESEARCH QUESTION**
+**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**
+
+<!-- Our framework: capability to WFH; HB old PARA2 -->
+The capability to work from home has previously been studied from the
+perspective of whether work tasks in a given occupation both can be and are
+allowed to be performed independently of location or co-location with
+colleagues, including supervisors (**insert refs**).
+However, successful telecommuting also requires that the quality and
+reliability of ICT services, particularly home internet connection speeds,
+enable the completion of work tasks with a minimum of delay or interruption.
+In reality, the performance of broadband speeds with respect to telecommuters
+has never been tested before at scale, as working from home and connecting to
+colleagues and workplace resources via broadband has previously been the
+purview of a small minority of workers.
+This papers understands telecommuting as a function of the quality of ICT
+services, particularly home internet connections, and whether work
+tasks in an occupation can be performed independently of being in a particular
+location or co-locating with colleagues
 
@@
 This is not the case for the workforce occupied in less footloose occupations as
-as they tend to be lower-income, non-white, without a university degree,
+they tend to be lower-income, non-white, without a university degree,
 live in rental accommodation and lack health insurance [@NBERw27085].
@@
 <!-- ICT infrastructure-->
-None of these changes could have happened in the absence of ICT
+None of these changes could have happened in the absence of reliable ICT
 infrastructure -- both in terms of software and hardware.
-But while innovations in software are easily diffused across space and
+But while software innovations are easily diffused across space and
 society^[See for example the huge success of videoconferencing apps
@@
 [@blank2014dimensions; @van2011internet].
-Importantly, the differentiated capacity to telecommute, which to a certain
-extend are related to the first and second level digital divides, leads to
+Importantly, the capacity to telecommute, which to a certain
+extend is related to the first and second level digital divides, leads to
 differentiated outcomes regarding the economic resilience of people and
@@
 towards working from home are difficult to predict. Nevertheless,
-they might span through various aspects of economy and society: from changes to
-transportation planning due to changes in commuting patters to changes in land
+they span through various aspects of economy and society: from changes to
+transportation planning due to altered commuting patters, to changes in land
 use and urban planning to accommodate people who work from home
-(@BUDNITZ2020102713); and from productivity and innovation changes to a change
+[@BUDNITZ2020102713]; and from productivity and innovation changes, to changes
 in agglomeration externalities and the attraction of large cities [@econobs]
-just to name a few. This paper is positioned to support such endeavours by
-providing an explanatory **[NOTE SURE ABOUT THIS]** framework  based on
-the quality of internet connectivity and the frequencies of occupations within
-Local Authorities in the UK.
+just to name a few. This paper is positioned to support endeavours in
+understanding the effects of increased telecommuting by
+exposing the spatial and social dimensions of telecommuting including the
+resilience of broadband speeds in terms of both quality and reliability of
+service, and whether this reinforces or redresses prior digital divisions.
+**HANNAH, HOW DOES THIS SOUND**
+
+The structure of this paper is as follows.
+...
 
@@
 
-**PARA5**: Data and methods
-
-**PARA6**: Contribution:
-
 # Literature review
@@
 
+- PARA about Contention??
+
+<!-- WFH and infrastructure, HB old PARA2 -->
+The infrastructural demands of working from home on internet services have been
+minor compared to the demands of leisure users, such that the broadband
+performance offered by different Internet Service Providers (ISPs) has been
+benchmarked according to download speeds during the evening 'primetime',
+when video streaming services are at their peak, rather than during the working
+day (**OfCom, 2017**).
+Yet the pandemic has fundamentally changed not only how many people work from
+home, but also their technical requirements when doing so.
+Whilst it may have been common in the past for telecommuters to complete
+solitary work tasks at home, but still visit their workplace for meetings
+(**ref?**), during the pandemic the replacement of face-to-face contact with
+bandwidth-intensive video conference calling was seen as essential to
+suppressing the spread of infection.
+This type of broadband use can particularly affect upload
+speeds **ref?**.
+This paper offers a framework to assess the quality and reliability of upload
+speeds by time of day and day of the week across local authorities in the UK
+during the time when the population were told to work from home if at all
+possible.
+
 ## from telecommuting to #WFH
+
+- Introduce space-time geography
+- circumstances where work can be carried out more flexibly in space and time
+- how relate to digital divides
+- capability theory of mobility? suppressed demand...
 
@@
 most such work was carried out in the homes of individual employees rather than any other location.
+[@eurofound2020]
 
-**PARA about Contention**
-
-# Time series clustering
-
-Desription of the method
+## covid and working from home, cities, urban structure
+- economic geography of UK
 
 # Data and descriptive statistics
+
+<!-- HB old PARA3 -->
+The experience of upload speeds is not the same as the maximum speeds offered by an ISP or possible speeds over a particular type and length of connection.
+Therefore, we use volunteered geographic information in the form of speed-checks run by users to test their experienced broadband speeds, upload and download.
+Meanwhile, the quality and reliability of upload speeds, like ICT more generally, vary in time and space due to both supply and demand-side influences, and can be measured in a number of ways.
+These include: a) mean, experienced connection speed, b) standard deviation or the amount of fluctuation from the mean, and c) the variation in speeds at particular times of day when working from home is more likely to take place.
+We take account of all three measurements in order to describe upload speeds as fully as possible.
+We start by calculating the mean upload speed for every local government district in the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank holidays.
+We then cluster the local authorities by these temporal profiles, allowing us to identify patterns and describe the overall means, standard deviations, and other relevant statistics for each spatial cluster.
+Thus we aim to answer the first part of our research question: How resilient are broadband speeds as experienced in different parts of the UK during a time of extreme demand?
+
+<!-- HB old PARA4 -->
+The cause of these different experiences of broadband resilience may be different in different areas, as they may reflect similarities in patterns of demand or similar quality of infrastructure.
+Our approach is also limited by potential endogeneity, as for example, better quality connections with high mean speeds may enable more working from home, but greater demand may cause slower speeds, less reliability and greater variability of speed at different times of day or week.
+Therefore, we avoid attributing any cause to our analysis of the experienced level of quality and reliability of upload speeds.
+Instead, we run auxilliary regressions in order to understand how the spatial and temporal patterns of internet service relate to the economic geography of the UK.
+We discuss how the different patterns might support or undermine efforts to work from home and maintain safe productivity and whether they reinforce existing spatial and social inequalities.
+From this analysis, we hope to provide a greater understanding of how telecommuting and technology intersect at a time of extreme demand, and what lessons this time has for a future where telecommuting is likely to remain a common means of accessing work and broadband services, as well as infrastructure, must be fit for purpose.
 
@@
 
+# Time series clustering
+
+Description of the method
+
 # Results
 
-##Upload Clusters / cluster description
+## Upload Clusters / cluster description
 
@@
 
-##aux regressions
+## aux regressions
 

### 2020-11-16 paper/v1/broadband.speed.covid.tex
merged intros

@@
 
+\begin{itemize}
+\tightlist
+\item
   \href{https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit\#heading=h.i5om1o8wpcd9}{our
-Google doc}
+  google doc}
+\end{itemize}
 
@@
 been on a slow, upward trend, to a widespread way of life in many
-countries. The ability to work from home or telecommute meant millions
-retained their jobs and, to a varying extent, maintained productivity
-during periods of strict lockdown around the world. However, this
-ability was not evenly distributed socially or spatially, creating a new
-type of digital divide -- one of economic resilience and personal safety
-on one side and unemployment or essential front-line work on the other.
-The ability to telecommute is partly a product of whether the majority
-of work tasks in an occupation can be performed independently of being
-in a particular location or co-locating with colleagues. It is also a
-product of the quality of ICT services, particularly home internet
-connections, sufficient to successfully complete work tasks with a
-minimum of delay or interruption. Therefore, in this paper, we look at
-patterns of broadband speed tests and upload speeds in the UK in order
-to gain insights into both the quality and reliability of broadband
-services across the country and the distribution of telecommuting at the
-time when the population were told to work from home if possible. To do
-so, we employ unique data regarding individual broadband speed tests and
-state of the art time series clustering methods. Our analysis enables us
+countries \citep{felstead2020homeworking}. \textbf{HANNAH, DO WE REALLY
+NEED A REFERENCE IN THE PREVIOUS SENTENCE?} The ability to work from
+home or telecommute meant millions retained their jobs and, to a varying
+extent, maintained productivity during periods of strict lockdown around
+the world. However, this ability has not been evenly distributed
+socially or spatially, creating a new type of digital divide. On one
+side are those who can work from home and have been able to enjoy both
+economic resilience and greater personal safety. On the other side,
+previously employed individuals have been forced to accept furlough or
+redundancy packages unless they are part of the cadre of essential
+workers, who are potentially at high risk of infection. Using the UK as
+a case study, this paper aims to improve our understanding of the
+spatial and social dimensions of this new digital divide by assessing
+the resilience of broadband speeds in terms of both quality and
+reliability of service, and whether this reinforces or redresses prior
+digital divisions. To do so, we employ unique volunteered geographic
+data on individual broadband speed tests and state-of-the-art
+time-series clustering methods, which enable us to create clusters of UK
+local authorities with similar temporal signatures of experienced
+internet speeds. We then associate these clusters of local authorities
+with their socioeconomic and geographic characteristics to explore
+whether they are linked with existing divides and the economic geography
+of the UK. \textbf{HANNAH, HOW DOES THIS SOUND?} Our analysis enables us
 to better understand how telecommuting and technology intersect at a
@@
 and broadband services, as well as infrastructure, must be fit for
-purpose. \textbf{I THINK WE NEED A MORE CLEAR RESEARCH QUESTION}
+purpose. \textbf{LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE
+MORE THE RQ}
+
+The capability to work from home has previously been studied from the
+perspective of whether work tasks in a given occupation both can be and
+are allowed to be performed independently of location or co-location
+with colleagues, including supervisors (\textbf{insert refs}). However,
+successful telecommuting also requires that the quality and reliability
+of ICT services, particularly home internet connection speeds, enable
+the completion of work tasks with a minimum of delay or interruption. In
+reality, the performance of broadband speeds with respect to
+telecommuters has never been tested before at scale, as working from
+home and connecting to colleagues and workplace resources via broadband
+has previously been the purview of a small minority of workers. This
+papers understands telecommuting as a function of the quality of ICT
+services, particularly home internet connections, and whether work tasks
+in an occupation can be performed independently of being in a particular
+location or co-locating with colleagues
 
@@
 associated with higher earnings. This is not the case for the workforce
-occupied in less footloose occupations as as they tend to be
-lower-income, non-white, without a university degree, live in rental
-accommodation and lack health insurance \citep{NBERw27085}. Although
-these figures refer to the US, similar trends can be observed for other
-countries. For example, \(75\)\% of workers with tertiary education
-worked from home in Europe during spring \(2020\), while the same share
-for workers with secondary and primary education dropped to \(34\)\% and
-\(14\)\% respectively. Moreover, employees living in cities, women and
-younger employees were have worked from home \citep{eurofound2020}.
+occupied in less footloose occupations as they tend to be lower-income,
+non-white, without a university degree, live in rental accommodation and
+lack health insurance \citep{NBERw27085}. Although these figures refer
+to the US, similar trends can be observed for other countries. For
+example, \(75\)\% of workers with tertiary education worked from home in
+Europe during spring \(2020\), while the same share for workers with
+secondary and primary education dropped to \(34\)\% and \(14\)\%
+respectively. Moreover, employees living in cities, women and younger
+employees were have worked from home \citep{eurofound2020}.
 
-None of these changes could have happened in the absence of ICT
+None of these changes could have happened in the absence of reliable ICT
 infrastructure -- both in terms of software and hardware. But while
-innovations in software are easily diffused across space and
+software innovations are easily diffused across space and
 society\footnote{See for example the huge success of videoconferencing
@@
 and the internet \citep{blank2014dimensions, van2011internet}.
-Importantly, the differentiated capacity to telecommute, which to a
-certain extend are related to the first and second level digital
-divides, leads to differentiated outcomes regarding the economic
-resilience of people and places against the current pandemic. And to the
-extend that the quality of the internet infrastructure and occupation
-variation are spatially dependent and clustered in space, the spatial
-footprint of telecommuting is of great interest. In a way similar to the
-third level digital divide, which focuses on the differentiated returns
-of internet use \citep{stern2009levels, van2014digital, van2015third}
-places with high rates of telecommuting during the Covid pandemic
-illustrate higher economic resilience against the current pandemic. This
-is aligned with the regional economic resilience literature, which
-underlines the differentiated capacity of cities and regions to escape
-or recover from economic crises
-\citep{martin2012regional, kitsos2018economic}.
+Importantly, the capacity to telecommute, which to a certain extend is
+related to the first and second level digital divides, leads to
+differentiated outcomes regarding the economic resilience of people and
+places against the current pandemic. And to the extend that the quality
+of the internet infrastructure and occupation variation are spatially
+dependent and clustered in space, the spatial footprint of telecommuting
+is of great interest. In a way similar to the third level digital
+divide, which focuses on the differentiated returns of internet use
+\citep{stern2009levels, van2014digital, van2015third} places with high
+rates of telecommuting during the Covid pandemic illustrate higher
+economic resilience against the current pandemic. This is aligned with
+the regional economic resilience literature, which underlines the
+differentiated capacity of cities and regions to escape or recover from
+economic crises \citep{martin2012regional, kitsos2018economic}.
 
@@
 attitudes towards working from home are difficult to predict.
-Nevertheless, they might span through various aspects of economy and
-society: from changes to transportation planning due to changes in
-commuting patters to changes in land use and urban planning to
-accommodate people who work from home (\citet{BUDNITZ2020102713}); and
-from productivity and innovation changes to a change in agglomeration
-externalities and the attraction of large cities \citep{econobs} just to
-name a few. This paper is positioned to support such endeavours by
-providing an explanatory \textbf{{[}NOTE SURE ABOUT THIS{]}} framework
-based on the quality of internet connectivity and the frequencies of
-occupations within Local Authorities in the UK.
+Nevertheless, they span through various aspects of economy and society:
+from changes to transportation planning due to altered commuting
+patters, to changes in land use and urban planning to accommodate people
+who work from home \citep{BUDNITZ2020102713}; and from productivity and
+innovation changes, to changes in agglomeration externalities and the
+attraction of large cities \citep{econobs} just to name a few. This
+paper is positioned to support endeavours in understanding the effects
+of increased telecommuting by exposing the spatial and social dimensions
+of telecommuting including the resilience of broadband speeds in terms
+of both quality and reliability of service, and whether this reinforces
+or redresses prior digital divisions. \textbf{HANNAH, HOW DOES THIS
+SOUND}
+
+The structure of this paper is as follows. \ldots{}
 
@@
 
-\textbf{PARA5}: Data and methods
-
-\textbf{PARA6}: Contribution:
-
 \hypertarget{literature-review}{%
@@
 
+\begin{itemize}
+\tightlist
+\item
+  PARA about Contention??
+\end{itemize}
+
+The infrastructural demands of working from home on internet services
+have been minor compared to the demands of leisure users, such that the
+broadband performance offered by different Internet Service Providers
+(ISPs) has been benchmarked according to download speeds during the
+evening `primetime', when video streaming services are at their peak,
+rather than during the working day (\textbf{OfCom, 2017}). Yet the
+pandemic has fundamentally changed not only how many people work from
+home, but also their technical requirements when doing so. Whilst it may
+have been common in the past for telecommuters to complete solitary work
+tasks at home, but still visit their workplace for meetings
+(\textbf{ref?}), during the pandemic the replacement of face-to-face
+contact with bandwidth-intensive video conference calling was seen as
+essential to suppressing the spread of infection. This type of broadband
+use can particularly affect upload speeds \textbf{ref?}. This paper
+offers a framework to assess the quality and reliability of upload
+speeds by time of day and day of the week across local authorities in
+the UK during the time when the population were told to work from home
+if at all possible.
+
 \hypertarget{from-telecommuting-to-wfh}{%
@@
 \#WFH}\label{from-telecommuting-to-wfh}}
+
+\begin{itemize}
+\tightlist
+\item
+  Introduce space-time geography
+\item
+  circumstances where work can be carried out more flexibly in space and
+  time
+\item
+  how relate to digital divides
+\item
+  capability theory of mobility? suppressed demand\ldots{}
+\end{itemize}
 
@@
 COVID crisis, most such work was carried out in the homes of individual
-employees rather than any other location.
+employees rather than any other location. \citep{eurofound2020}
 
-\textbf{PARA about Contention}
+\hypertarget{covid-and-working-from-home-cities-urban-structure}{%
+\subsection{covid and working from home, cities, urban
+structure}\label{covid-and-working-from-home-cities-urban-structure}}
 
-\hypertarget{time-series-clustering}{%
-\section{Time series clustering}\label{time-series-clustering}}
-
-Desription of the method
+\begin{itemize}
+\tightlist
+\item
+  economic geography of UK
+\end{itemize}
 
@@
 
-Data details and some figures, descriptive stats
+The experience of upload speeds is not the same as the maximum speeds
+offered by an ISP or possible speeds over a particular type and length
+of connection. Therefore, we use volunteered geographic information in
+the form of speed-checks run by users to test their experienced
+broadband speeds, upload and download. Meanwhile, the quality and
+reliability of upload speeds, like ICT more generally, vary in time and
+space due to both supply and demand-side influences, and can be measured
+in a number of ways. These include: a) mean, experienced connection
+speed, b) standard deviation or the amount of fluctuation from the mean,
+and c) the variation in speeds at particular times of day when working
+from home is more likely to take place. We take account of all three
+measurements in order to describe upload speeds as fully as possible. We
+start by calculating the mean upload speed for every local government
+district in the UK for each hour of each weekday, excluding midnight to
+6:00, weekends and bank holidays. We then cluster the local authorities
+by these temporal profiles, allowing us to identify patterns and
+describe the overall means, standard deviations, and other relevant
+statistics for each spatial cluster. Thus we aim to answer the first
+part of our research question: How resilient are broadband speeds as
+experienced in different parts of the UK during a time of extreme
+demand?
+
+The cause of these different experiences of broadband resilience may be
+different in different areas, as they may reflect similarities in
+patterns of demand or similar quality of infrastructure. Our approach is
+also limited by potential endogeneity, as for example, better quality
+connections with high mean speeds may enable more working from home, but
+greater demand may cause slower speeds, less reliability and greater
+variability of speed at different times of day or week. Therefore, we
+avoid attributing any cause to our analysis of the experienced level of
+quality and reliability of upload speeds. Instead, we run auxilliary
+regressions in order to understand how the spatial and temporal patterns
+of internet service relate to the economic geography of the UK. We
+discuss how the different patterns might support or undermine efforts to
+work from home and maintain safe productivity and whether they reinforce
+existing spatial and social inequalities. From this analysis, we hope to
+provide a greater understanding of how telecommuting and technology
+intersect at a time of extreme demand, and what lessons this time has
+for a future where telecommuting is likely to remain a common means of
+accessing work and broadband services, as well as infrastructure, must
+be fit for purpose.
+
+The primary data analysed in this paper was provided by Speedchecker
+Ltd, a private company that allows internet users to check their own
+broadband upload / download speeds, and stores every speed-check with a
+timestamp and geolocation.
+
+Our approach involves aggregating all the speed-checks during the 13
+weeks of March to May inclusive for weekdays in 2020 by each hour of the
+day and day of the week. As our research aims to identify the geography
+of internet service resilience for work purposes, bank holidays and the
+hours between midnight and 6am were also excluded. The composite week
+time series thus comprise 18 hours multiplied by 5 weekdays or 90 time
+points per series. These time series were calculated for each of the 382
+Local Authority Districts (LADs) in the UK, standardised, and then a
+k-means clustering algorithm was applied, including dynamic time
+warping. The LADs were assigned to 10 clusters for upload speeds. The
+cluster id for each LAD is then reattached to the speed-check dataframe
+to identify the characteristics of each cluster, including number of
+LADs, and the descriptives statistics of upload speeds in that cluster,
+and the temporal profile by hour of the day and day of the week.
+
+\textbf{Data details and some figures, descriptive stats - include whole
+sample time profile for 2019 and 2020 frequency of tests run as part of
+why we chose to create the time profiles by hour of the day and day of
+the week rather than daily over the whole period.}
+
+\hypertarget{time-series-clustering}{%
+\section{Time series clustering}\label{time-series-clustering}}
+
+Description of the method
 
@@
 
-Clusters, cluster description and aux regressions
+\hypertarget{upload-clusters-cluster-description}{%
+\subsection{Upload Clusters / cluster
+description}\label{upload-clusters-cluster-description}}
+
+The temporal profiles used to cluster the local authorities have been
+summarised in graph {[}.{]}, which shows a composite profile of mean
+upload speeds per hour per day for each cluster. For upload speeds, 345
+of 382 local authorities, or over 62 million people, fall into cluster 6
+or cluster 9. Graph {[}.{]} shows that both of these clusters have
+relatively similar temporal profiles, which are flatter than the other,
+smaller clusters, suggesting better reliability of service. However, the
+upload speeds at all times for cluster 9 are substantially higher than
+for cluster 6, which is an indication of better quality of service. This
+difference is reflected not only in the mean speeds for these clusters
+for the whole sample, but also the mean upload speeds in the morning
+peak from 9:00-10:59, as well as the evening peak period from
+19:00-20:59. In comparison, the time profile in graph {[}.{]} shows
+upload speeds in cluster 1 are on average lower at certain times of day
+during the study period than any other cluster, whilst the profile for
+cluster 3 appears to show speeds fluctuating as much as cluster 1, but
+at levels usually higher even than cluster 9.
+
+The variability of the smaller clusters may be related to the fewer
+speed tests from fewer local authorities that have been averaged, whilst
+averaging greater numbers of speed tests could artificially flatten the
+profile. This appears not to be the case for cluster 9, as there is
+negligible difference between morning and evening upload speeds, at 1\%
+slower in the morning, confirming a high level of reliability of
+service. In comparison, upload speeds in cluster 6 are 4\% slower in the
+morning than in the evening and experienced the joint highest ratio of
+standard deviation to mean across the time period under assessment. This
+suggests that although the time profile is relatively flat in graph
+{[}.{]}, the experience is one of speeds that fluctuate from a lower
+mean, and therefore might more often impact on online activities. Still,
+in most of the smaller clusters the reliability of service is worse than
+in cluster 9 in terms of the ratio of standard deviation to mean and
+worse than cluster 6 as well in terms of the ratio of upload speed in
+the morning peak compared to the evening peak. In five of the smaller
+clusters, which are home to almost 3 million people, speeds are 12-19\%
+slower between 9:00-10:59 compared to 19:00-20:59.
+
+The quality and reliability of broadband service is thus much better in
+the 115 Local Authority areas in cluster 9, which are mostly in urban or
+suburban areas, compared to cluster 6. These include 13 London Boroughs
+(of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7
+constituent authorities of the West Midlands Combined Authority, as well
+as cities like Glasgow, Leicester, Nottingham, Sheffield, and the
+Portsmouth and Southampton conurbation. There are also some notable
+medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes,
+and York, and many suburban districts from the South East of England to
+South Tyneside. Meanwhile, the 230 local authorities in cluster 6, which
+have lower speeds on average and more variation in service still include
+major urban areas, such as Bristol, Liverpool and Leeds, and many
+suburban areas, but also include some of the most rural areas in the
+country. Meanwhile, Cluster 1, with 10 local authorities that are home
+to over 1 million people has the second slowest speeds in the morning
+compared to the evening `peaks' and the second highest ratio of standard
+deviation to mean. This cluster's most populous area is Westminster in
+central London.
+
+\textbf{further description of temporal profiles of other clusters -
+include all in graph?} Indeed, the only cluster where upload speeds were
+slower in the evening than in the morning was cluster 5, made up of 2
+local authorities with less than 200,000 people: Three Rivers, a
+suburban district north of London, and Fylde, a seaside suburb of
+Blackpool. However, these are likely to be outliers and may not have
+many tests from which the clusters are calcualted. \textbf{Take out
+smaller clusters?} The exceptions can be found in seven of the eight
+much smaller clusters including 35 local authority districts, where AM
+peak upload speeds are between 6\% and 18\% slower than PM peak upload
+speeds, although the mean speeds for each cluster are higher than
+cluster 6. Indeed, in 25 local authorities with a combined population of
+almost 3 million, speeds are 13\% or more slower in the morning than in
+the evening. Included in this latter group are central London borough of
+Westminster and the London Borough of Newham, rural authorities like
+Eden and West Devon, and small cities like Dundee and Carlisle.
+
+\hypertarget{aux-regressions}{%
+\subsection{aux regressions}\label{aux-regressions}}
+
+Auxiliary regressions indicate that the speed tests in cluster 9
+authorities are also more likely to have been run on services provided
+by Virgin Media, suggesting they are in the half of the country with the
+most lucrative ICT market, which originally attracted the cable TV
+provider (OfCom\ldots). For example, although auxiliary regressions show
+that Cluster 6 local authorities are more likely to be in the South of
+the UK than Cluster 9, the cluster notably includes Southern rural
+districts from Cornwall to North Norfolk. Slower speeds could reflect
+the lower quality of service in rural areas compared to urban and
+suburban areas. There is frustration, however, as auxiliary regressions
+show that those living in cluster 6 had the highest probability of
+testing their broadband between 9:00 and 11:00 of any of the 10 upload
+speed clusters.
+
+the finding in the auxiliary regressions that those in cluster 9 ran the
+fewest speed tests per person during the morning period of any cluster.
+Now this may be an indication of fewer people working from home, less
+contention, and less resultant frustration. Cluster 9 comprises many
+central urban areas and has the lowest number of established businesses
+per inhabitant, which could be interpreted as a dominance of large
+employers. However, the job density is lower than cluster 6, meaning
+there are not as many jobs per resident in these areas. As the cluster
+includes many suburban areas too, which may be largely residential,
+could the quality and reliability of internet service be reinforcing
+patterns of telecommuting by those in wealthier suburbs who can work
+from home? Earnings in cluster 9 are second highest of all the clusters,
+with only Cluster 2 (comprising just North East Lincolnshire and East
+Lothian, population 265k) earning more per person.
+
+The auxiliary regressions show that compared to the two authorities in
+cluster 5, all the other clusters had a lower percentage of working
+people in managerial, professional and administrative jobs.??? Yet the
+auxiliary regression suggests that there are not many tests being run
+during the am peak in cluster 1. This may be because there are fewer
+people working at home checking their broadband than in most other
+clusters. Indeed, the auxiliary regressions indicate that cluster 1 has
+the highest job density or proportion of jobs to working-age population,
+which is likely to due to the presence of Westminster, central London,
+cluster 1's most populous local authority. Westminster not only has more
+workplaces than residents, but it is reasonable to presume that many who
+would normally work in Westminster, but be able to work from home during
+lockdown are likely to live outside central London and not be subject to
+the fluctuating speeds there. Workplaces, meanwhile, some of which would
+still have been open, could be running programmes that cause the
+slowdown and variation, but would be more likely to have their own
+in-house diagnostics, rather than using a service like Speedchecker Ltd.
 

### 2020-11-19 paper/v1/broadband.speed.covid.Rmd

@@
 
-During the pandemic, working from home using Information and Communication
-Technologies (ICT), whether partially or exclusively, was transformed from a
-niche means of accessing work, albeit one that had been on a slow, upward trend,
-to a widespread way of life in many countries [@felstead2020homeworking].
-**HANNAH, DO WE REALLY NEED A REFERENCE IN THE PREVIOUS SENTENCE?**
+During the pandemic, working from home using digital technologies, whether partially 
+or exclusively, was transformed from a niche means of accessing work, albeit one that 
+had been on a slow, upward trend, to a widespread way of life in many countries.
 The ability to work from home or telecommute meant millions retained their jobs
@@
 lockdown around the world.
-<!--OLD TEXT: However, this ability was not evenly distributed socially or spatially, creating
-a new type of digital divide -- one of economic resilience and personal safety on
-one side and unemployment or essential front-line work on the other.
-The ability to telecommute is partly a product of whether the majority of work
-tasks in an occupation can be performed independently of being in a particular
-location or co-locating with colleagues.-->
 However, this ability has not been evenly distributed socially or spatially,
 creating a new type of digital divide.
-On one side are those who can work from home and have been able to enjoy both
-economic resilience and greater personal safety.
+On one side are those who can work from home, supported by digital technologies, 
+and have thus been able to enjoy both economic resilience and greater personal safety.
 On the other side, previously employed individuals have been forced to accept
@@
 workers, who are potentially at high risk of infection.
-Using the UK as a case study, this paper aims to improve our understanding of
-the spatial and social dimensions of this new digital divide by assessing the
-resilience of broadband speeds in terms of both quality and reliability of
-service, and whether this reinforces or redresses prior digital divisions.
-<!--and upload speeds in the UK in order to gain insights into both the quality and
-reliability of broadband services across the country and the distribution of
-telecommuting at the time when the population were told to work from home if
-possible.-->
-To do so, we employ unique volunteered geographic data on individual broadband
-speed tests and state-of-the-art time-series clustering methods, which enable us
-to create clusters of UK local authorities with similar temporal signatures of
-experienced internet speeds.
+Whilst the basis for this new digital divide can be viewed as mainly occupational, 
+we consider whether the divide is also technological.
+Using the UK as a case study, this paper aims to understand how the resilience
+of broadband speeds in terms of both quality and reliability of service may reinforce
+or redress the spatial and social dimensions of the digital division exposed by the
+pandemic.
+To do so, we employ volunteered geographic data on individual broadband speed tests
+and state-of-the-art time-series clustering methods to create clusters of UK local
+authorities with similar temporal signatures of experienced internet speeds.
 We then associate these clusters of local authorities with their socioeconomic
-and geographic characteristics to explore whether they are linked with existing
-divides and the economic geography of the UK.
-**HANNAH, HOW DOES THIS SOUND?**
-Our analysis enables us to better understand how telecommuting and
-technology intersect at a time of extreme demand, and what lessons this time has
-for a future where telecommuting is likely to remain a common means of accessing
-work and broadband services, as well as infrastructure, must be fit for purpose.
+and geographic characteristics to explore how they overlap with or diverge from
+the existing economic and digital geography of the UK. **need to add in urban-rural**
+Our analysis enables us to better understand how the spatial and social distribution 
+of both occupations and online accessibility intersect to enable or hinder the 
+practice of telecommuting at a time of extreme demand. 
+We will also consider what lessons can be learned from this time for a future where
+telecommuting is likely to remain a common means of accessing work, and broadband
+services and infrastructure must be fit for purpose.
 **LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**
@@
 perspective of whether work tasks in a given occupation both can be and are
-allowed to be performed independently of location or co-location with
-colleagues, including supervisors (**insert refs**).
+allowed to be performed using digital technologies independently of 
+location or co-location with colleagues, including supervisors 
+[@allen2015effective; @singh2013modeling].
 However, successful telecommuting also requires that the quality and
-reliability of ICT services, particularly home internet connection speeds,
+reliability of digital services, particularly home internet connection speeds,
 enable the completion of work tasks with a minimum of delay or interruption.
-In reality, the performance of broadband speeds with respect to telecommuters
-has never been tested before at scale, as working from home and connecting to
-colleagues and workplace resources via broadband has previously been the
-purview of a small minority of workers.
-This papers understands telecommuting as a function of the quality of ICT
-services, particularly home internet connections, and whether work
-tasks in an occupation can be performed independently of being in a particular
-location or co-locating with colleagues
-
-<!-- How much do we work form home-->
-The shift towards telecommuting during various stages of lockdown around the world
-has been drastic and, importantly, speculations indicate that the post-Covid
-tendency to work from home will be much higher than the pre-Covid one.
+Prior to the pandemic, the performance of broadband services with respect to
+telecommuters was never tested at scale, as working from home and connecting to
+colleagues and workplace resources via broadband was the purview of a small minority 
+of workers. 
+Instead, in the UK, for example, benchmarking broadband performance and service 
+delivery by different Internet Service Providers (ISPs) focused upon the quality of 
+service for evening leisure use, when video streaming services are at their peak, 
+rather than during the working day [@ofcom2017].
+Yet the shift towards telecommuting during various stages of lockdown around the world
+has been drastic and there are speculations that post-Covid, the tendency to work from
+home will be much higher than the pre-Covid, raising questions around whether 
+internet services can accommodate the increased demand.
+For example, $47$% of people in employment in the UK worked soley from home in April
+$2020$, whilst the same figure only reached $5$% the year before [@ons2020; 
+@ons2020lm2019]. 
 A back of the envelope calculation suggests that up to 40% of the working force
-could work from home in the UK [@batty2020editorial].
-Observational data pointed an even higher share of people in employment in the
-UK who worked from home in April $2020$ ($47$%), while the same figure only reached
-$5$% the year before [@ons2020; @ons2020lm2019].
+could work from home on an ongoing basis [@batty2020editorial].
 Similar figures have been reported for other countries around the world
 [@felstead2020homeworking].
-For instance, $37$% of the workforce worked from home in Europe in April $2020$ with
+For instance, $37$% of the European workforce worked from home in April $2020$ with
 countries like Finland reaching $60$% [@eurofound2020].
-In the US almost half of the working population worked from home during the same
+In the US, almost half of the working population worked from home during the same
 period because of the pandemic [@brynjolfsson2020covid].
-More broadly, a recent estimate indicated that $37$% of all jobs in the US can be
+Furthermore, a recent estimate indicated that $37$% of all jobs in the US can be
 permanently performed entirely from home [@NBERw26948].
 
-<!-- Who works from home-->
-There is a consensus that opportunities for working from home especially during
-the current pandemic are not equally spread across the workforce.
-@NBERw26948 indicated that in the US managers, educators, as well as those
-working in computer-related occupations, finance, and law can easily work from
-home.
-On the contrary, workers in farming, construction, and manufacturing do not have
-such opportunities.
-Not surprisingly, occupations with opportunities to telecommute are associated
-with higher earnings.
-This is not the case for the workforce occupied in less footloose occupations as
-they tend to be lower-income, non-white, without a university degree,
-live in rental accommodation and lack health insurance [@NBERw27085].
-Although these figures refer to the US, similar trends can be observed for other
-countries.
-For example, $75$% of workers with tertiary education worked from home in Europe
-during spring $2020$, while the same share for workers with secondary and
-primary education dropped to $34$% and $14$% respectively.
-Moreover, employees living in cities, women and younger employees were
-have worked from home [@eurofound2020].
-
-<!-- ICT infrastructure-->
+<!-- ICT infrastructure and digital divides-->
 None of these changes could have happened in the absence of reliable ICT
@@
 broadband connectivity.
-The literature exemplifies digital divides in terms of internet access and its
-quality.
-For instance, @riddlesden2014broadband highlighted the broadband divides in the
-UK, while the systematic review from @SALEMINK2017360 reinforced our
-understanding for the infrastructure quality differences between urban and rural
-areas.
-
-<!-- WFH and digital divides-->
-Our framework to understand telecommuting as a function of occupations and
-quality of ICT infrastructure is aligned with current debates on digital divides.
-While the, so-called, first level digital divides are associated with access
-and quality of internet connectivity, the second level ones are linked
-to the necessary skills to effectively utilise ICT and the internet
+The literature describes first level digital divides in terms of the availability and
+quality of internet connectivity, such as that manifest in different geographies 
+in the UK [@riddlesden2014broadband; @philip2017digital]. 
+Second level digital divides consider the presence or lack of the necessary skills 
+to effectively utilise digital technologies and the internet 
 [@blank2014dimensions; @van2011internet].
-Importantly, the capacity to telecommute, which to a certain
-extend is related to the first and second level digital divides, leads to
-differentiated outcomes regarding the economic resilience of people and
-places against the current pandemic.
-And to the extend that the quality of the internet infrastructure and occupation
-variation are spatially dependent and clustered in space, the spatial footprint
-of telecommuting is of great interest. In a way similar to the third level
-digital divide, which
-focuses on the differentiated returns of internet use [@stern2009levels;
-@van2014digital; @van2015third] places with high rates of telecommuting during
-the Covid pandemic illustrate higher economic resilience against the current pandemic.
-<!-- Econ resilience-->
-This is aligned with the regional economic resilience literature, which
-underlines the differentiated capacity of cities and regions to escape or
-recover from economic crises [@martin2012regional; @kitsos2018economic].
-
-<!-- WFH broader consequences-->
-The long-term effects of such drastic changes in telecommuting and attitudes
-towards working from home are difficult to predict. Nevertheless,
-they span through various aspects of economy and society: from changes to
-transportation planning due to altered commuting patters, to changes in land
-use and urban planning to accommodate people who work from home
-[@BUDNITZ2020102713]; and from productivity and innovation changes, to changes
-in agglomeration externalities and the attraction of large cities [@econobs]
-just to name a few. This paper is positioned to support endeavours in
-understanding the effects of increased telecommuting by
-exposing the spatial and social dimensions of telecommuting including the
-resilience of broadband speeds in terms of both quality and reliability of
-service, and whether this reinforces or redresses prior digital divisions.
-**HANNAH, HOW DOES THIS SOUND**
+The third level focuses on the differentiated returns of internet use **for...?**
+[@stern2009levels; @van2014digital; @van2015third].
+The capability to telecommute is related to all three levels of digital divides,
+but more importantly leads to differentiated outcomes regarding the economic 
+resilience of people and places during the current pandemic.
+We identify this as a new digital divide, one that fundamentally alters the 
+potential returns of internet use for the user and wider community, 
+assumes skills that are present in some occupations but not in others, 
+and relies upon access to high quality internet services.
+As the quality of internet infrastructure and services, as well as variation in 
+occupations are spatially dependent and clustered in space, our approach offers a 
+framework for understanding which types of places, socially and spatially, 
+are more likely to land on the right side of this new digital divide. 
+By asking how resilient broadband speeds are as experienced in different parts of the 
+UK during a time of extreme demand, we will better understand which places benefit 
+from greater economic resilience not only during the pandemic, but also into the 
+future.
 
 The structure of this paper is as follows.
+First we will review the literature on telecommuting and digital divides to better 
+understand the origins of the new digital divide revealed by the pandemic and its
+impact on the economic resilience of different places.
+We then describe our data and methodology.
+Our results section first describes the how internet services vary in the 
+different clusters and then assesses whether these clusters replicate or repudiate
+other socio-economic and geographic patterns of economic resilience.
+
 ...
@@
 
-## broadband studies, divides, broadband tech stuff
+## from telecommuting to #WFH
 
-- PARA about Contention??
+<!-- Who works from home-->
+In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used 
+interchangeably, as most remote labour during the Covid-19 crisis was carried out 
+in the homes of individual employees rather than any other location [@eurofound2020]. 
+However, it should be noted that previous research has explored how telecommuting 
+can occur in other places, including satellite offices or on public transport 
+[@felstead2012rapid; @siha2006telecommuting].
+Previous research has also used a variety of definitions to measure the level of 
+telecommuting within different workforces, distinguising, for example, between those 
+directly employed, indirectly employed, self-employed, full-time or part-time, and 
+who use digital technologies to work remotely full-days or part-days 
+[@allen2015effective; @bailey2002review; @haddad2009examination].
+<!--old text: An appropriate definition of teleworking is ‘the remote provision of labour that would otherwise be carried out within company premises’ (European Commission, 2020b).-->
+No matter the definition, the option and capability to telecommute or work from home 
+has never been equally distributed spatially or socio-economically any more than 
+different industries and employment opportunities have.
+For example, studies from the United States, the Netherlands, and the UK indicate 
+that telecommuters are most likely to hold professional, managerial, and technical 
+occupations, where the workforce is better educated and wealthier, and that there is
+suppressed demand among women and part-time workers
+[@headicar2016move; @peters2004employees; @singh2013modeling].
+Opportunities for working from home during the current pandemic have likewise not 
+been equally spread across the workforce.
+@NBERw26948 indicated that in the US, managers, educators, as well as those
+working in computer-related occupations, finance, and law can easily work from
+home, and that occupations with opportunities to telecommute are associated
+with higher earnings.
+This is not the case for the workforce occupied in more spatially fixed occupations,
+from farming, construction, and manufacturing to hospitality and care services.
+In the US, these occupations tend to be lower-income, non-white, 
+without a university degree, live in rental accommodation and lack health insurance
+[@NBERw27085].
+Similar trends can be observed for other countries.
+For example, $75$% of workers with tertiary education worked from home in Europe
+during spring $2020$, whilst only $34$% of workers with secondary education and
+$14$% of those primary education did so [@eurofound2020].
 
-<!-- WFH and infrastructure, HB old PARA2 -->
-The infrastructural demands of working from home on internet services have been
-minor compared to the demands of leisure users, such that the broadband
-performance offered by different Internet Service Providers (ISPs) has been
-benchmarked according to download speeds during the evening 'primetime',
-when video streaming services are at their peak, rather than during the working
-day (**OfCom, 2017**).
-Yet the pandemic has fundamentally changed not only how many people work from
-home, but also their technical requirements when doing so.
-Whilst it may have been common in the past for telecommuters to complete
-solitary work tasks at home, but still visit their workplace for meetings
-(**ref?**), during the pandemic the replacement of face-to-face contact with
-bandwidth-intensive video conference calling was seen as essential to
-suppressing the spread of infection.
-This type of broadband use can particularly affect upload
-speeds **ref?**.
-This paper offers a framework to assess the quality and reliability of upload
-speeds by time of day and day of the week across local authorities in the UK
-during the time when the population were told to work from home if at all
-possible.
+## digital divides and economic resilience
+<!-- WFH and digital divides-->
+Our understanding of telecommuting as a product of enabled occupations can be 
+described as a manifestation of the third level digital divide, as those who are able 
+to use digital technologies to work from home benefit from a high rate of return on 
+their use of the internet in terms of autonomy, flexibility, and time saved from 
+commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
+These returns have been even greater during the Covid-19 crisis, when those with the 
+ability to telecommute also have the ability to maintain their employment whilst 
+protecting their health.
+However, the success of these arrangements has also been dependent upon the first 
+level digital divide, which is associated with access and quality of internet 
+connectivity.
+For example, the systematic review from @SALEMINK2017360 highlights the 
+infrastructure quality differences between urban and rural areas.
+Whether this variation in infrastructure quality affects the spatial footprint of 
+telecommuting has not previously been measured, although there are indications that
+those who purchase high speed connections consume more data of all sorts and use their
+connections for a variety of purposes [@hauge2011consumer].
+Another study identifies a correlation between access to internet services and a 
+reduction in household transport spend across 33 coutries, although the authors are 
+uncertain whether this implies less travel, for example because of increased
+telecommuting, or whether internet access enables more efficient travel 
+[@bris2017ict].
+However, the extreme demand during the pandemic provides a new opportunity to 
+understand how infrastructure accessibility, quality, and reliability affects 
+telecommuting, particularly as working from home during the pandemic required
+high volumes of bandwidth-intensive video conferencing in order to avoid the 
+face-to-face contact that could increase the spread of infection.
+Thus by answering questions about internet service resilience, we can also refine our
+understanding of how this has contributed to or reduced the new digital divide where
+economic resilience has been dependent upon the capability to work from home.
 
-## from telecommuting to #WFH
+<!-- Econ resilience-->
+This is aligned with the regional economic resilience literature, which
+underlines the differentiated capacity of cities and regions to escape or
+recover from economic crises [@martin2012regional; @kitsos2018economic].
+
+## other subsection?
 
@@
 
-In this analysis, the terms ‘teleworking’ and ‘working from home’ are used interchangeably. An appropriate definition of teleworking is ‘the remote
-provision of labour that would otherwise be carried out within company premises’ (European Commission, 2020b). In practice, during the COVID crisis,
-most such work was carried out in the homes of individual employees rather than any other location.
-[@eurofound2020]
-
 ## covid and working from home, cities, urban structure
@@
 <!-- HB old PARA3 -->
-The experience of upload speeds is not the same as the maximum speeds offered by an ISP or possible speeds over a particular type and length of connection.
-Therefore, we use volunteered geographic information in the form of speed-checks run by users to test their experienced broadband speeds, upload and download.
-Meanwhile, the quality and reliability of upload speeds, like ICT more generally, vary in time and space due to both supply and demand-side influences, and can be measured in a number of ways.
-These include: a) mean, experienced connection speed, b) standard deviation or the amount of fluctuation from the mean, and c) the variation in speeds at particular times of day when working from home is more likely to take place.
-We take account of all three measurements in order to describe upload speeds as fully as possible.
-We start by calculating the mean upload speed for every local government district in the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank holidays.
-We then cluster the local authorities by these temporal profiles, allowing us to identify patterns and describe the overall means, standard deviations, and other relevant statistics for each spatial cluster.
-Thus we aim to answer the first part of our research question: How resilient are broadband speeds as experienced in different parts of the UK during a time of extreme demand?
+This paper assesses the quality and reliability of upload speeds by time of day and 
+day of the week across local authorities in the UK during the time when the population
+were told to work from home if at all possible.
+The experience of upload speeds is not the same as the maximum speeds offered by an 
+ISP or possible speeds over a particular type and length of connection. 
+We use volunteered geographic information in the form of speed-checks run by users to 
+test their experienced broadband speeds, upload and download. 
+Since the quality and reliability of internet services vary in time and space due to 
+both supply and demand-side influences, we use a number of different measures of 
+experienced upload speeds. 
+These include: a) mean, experienced connection speed, 
+b) standard deviation or the amount of fluctuation from the mean, and 
+c) the variation in speeds at particular times of day when working from home is more 
+likely to take place. 
+We take account of all three measurements in order to describe upload speeds as fully 
+as possible. 
+We start by calculating the mean upload speed for every local government district in 
+the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank 
+holidays. 
+We then cluster the local authorities by these temporal profiles, allowing us to 
+identify patterns and describe the overall means, standard deviations, and other 
+relevant statistics for each spatial cluster. 
+Thus we aim to answer the first part of our research question: 
+How resilient are broadband speeds as experienced in different parts of the UK during 
+a time of extreme demand?
 
 <!-- HB old PARA4 -->
-The cause of these different experiences of broadband resilience may be different in different areas, as they may reflect similarities in patterns of demand or similar quality of infrastructure.
-Our approach is also limited by potential endogeneity, as for example, better quality connections with high mean speeds may enable more working from home, but greater demand may cause slower speeds, less reliability and greater variability of speed at different times of day or week.
-Therefore, we avoid attributing any cause to our analysis of the experienced level of quality and reliability of upload speeds.
-Instead, we run auxilliary regressions in order to understand how the spatial and temporal patterns of internet service relate to the economic geography of the UK.
-We discuss how the different patterns might support or undermine efforts to work from home and maintain safe productivity and whether they reinforce existing spatial and social inequalities.
-From this analysis, we hope to provide a greater understanding of how telecommuting and technology intersect at a time of extreme demand, and what lessons this time has for a future where telecommuting is likely to remain a common means of accessing work and broadband services, as well as infrastructure, must be fit for purpose.
+The cause of these different experiences of broadband resilience may be different in 
+different areas, as they may reflect similarities in patterns of demand or similar 
+quality of infrastructure. 
+Our approach is also limited by potential endogeneity, as for example, better quality 
+connections with high mean speeds may enable more working from home, but greater 
+demand may cause slower speeds, less reliability and greater variability of speed at 
+different times of day or week. 
+Therefore, we avoid attributing any cause to our analysis of the experienced level of 
+quality and reliability of upload speeds. 
+Instead, we run auxilliary regressions in order to understand how the spatial and 
+temporal patterns of internet service relate to the economic geography of the UK. 
+We discuss how the different patterns might support or undermine efforts to work from 
+home and maintain safe productivity and whether they reinforce existing spatial and 
+social inequalities. 
+From this analysis, we hope to provide a greater understanding of how telecommuting 
+and technology intersect at a time of extreme demand, and what lessons this time has 
+for a future where telecommuting is likely to remain a common means of accessing work 
+and broadband services, as well as infrastructure, must be fit for purpose.
 
@@
 # Conclusions
+<!-- WFH broader consequences-->
+The long-term effects of such drastic changes in telecommuting and attitudes
+towards working from home are difficult to predict. Nevertheless,
+they span through various aspects of economy and society: from changes to
+transportation planning due to altered commuting patters, to changes in land
+use and urban planning to accommodate people who work from home
+[@BUDNITZ2020102713]; and from productivity and innovation changes, to changes
+in agglomeration externalities and the attraction of large cities [@econobs]
+just to name a few. This paper is positioned to support endeavours in
+understanding the effects of increased telecommuting by
+exposing the spatial and social dimensions of telecommuting including the
+resilience of broadband speeds in terms of both quality and reliability of
+service, and whether this reinforces or redresses prior digital divisions.
+**HANNAH, HOW DOES THIS SOUND - this last bit probably in the discussion at the end?**
 

### 2020-11-24 paper/v1/broadband.speed.covid.Rmd
started working on the methods sectrion. haven't finished yet :(

--- 2020-11-16 paper/v1/broadband.speed.covid.Rmd
@@
 
-Description of the method
+The core of the methodological framework of this paper lies upon clustering methods.
+Clustering is a widely used family of techniques in geography (**ADD REF**),
+which can be defined within the modern machine learning framework as an unsupervised
+learning task, which involves partitioning unlabeled observations into homogeneous
+groups known as clusters [@montero2014tsclust]. The key idea is that observations
+within clusters tend to be more similar than observations between clusters.
+Clustering is particularly useful for exploratory studies as it identifies
+structures within the data [@aghabozorgi2015time].
+
+Various applications of clustering can be found within geography: **ADD EXAMPLES**
+
+Key characteristic of these studies is the cross-sectional nature of the data they
+employed. Indeed, most clustering problems in geography deal with observations
+that are fixed in time. However, for this paper we are interested in creating
+clusters of local authorities in the UK with similar temporal signatures of
+experienced internet speeds over time. Hence, we deviate from the established
+geographical clustering tools and employ time series clustering methods.
+
+Time series clustering methods have been developed in order to deal with clustering
+problems linked to, for instance, stock or other financial data, economic,
+governmental or medical data as well as machine monitoring
+[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
+The main challenge -- and also the difference with cross-sectional clustering
+problems -- is data dimensionality given the multiplicity of data points for every
+individual included in the data set. Time series data are dynamic data as the
+value of the observations change as a function of time [@aghabozorgi2015time].
+This high dimensionality leads to (i) computational and algorithmic challenges
+regarding handling these data and building algorithms to perform clustering over
+long time series, and (ii) open questions regarding the choice of similarity
+measures in order to cluster similar times series together considering the whole
+length of the time series and overcoming issues around noise, outliers and shifts
+[@lin2004iterative; @aghabozorgi2015time].
+
+Time series clustering methods utilising the whole length of time series can be
+grouped in three categories. The first -- model-based approaches -- is based on
+recovering the underlying model for each time series and then apply clustering
+algorithms on the model parameters [@aghabozorgi2015time]. The main criticism
+is the cluster accuracy for near clusters [@mitsa2010temporal]. The second approach
+is a based the formation of vectors of features based on the original time series.
+These new data of reduced dimensionality is then clustered using conventional
+clustering algorithms.
+
+For this paper we utilise the third category of time series clustering methods
+known as shape-based approaches. The main idea is to match any two sperate time
+series based on the similarity of their shapes through the calculation of
+distances among these two shapes.
+
+@sardatime
+Ideally, all members of the same cluster are similar to each other, but are as
+dissimilar as possible from objects in a different cluster.
+
+Time-series is a common type of dynamic data that naturally arises in many different scenarios,
+such as stock data, medical data, and machine monitoring, just to name a few (Aghabozorgi et al.,
+2015; Aggarwal and Reddy, 2013). They pose some challenging issues due to the large size and high
+dimensionality commonly associated with time-series (Aghabozorgi et al., 2015). In this context,
+dimensionality of a series is related to time, and it can be understood as the length of the series.
+
+the data and how the underlying grouping is performed. One classification depends on whether the
+whole series, a subsequence, or individual time points are to be clustered. On the other hand, the
+clustering itself may be shape-based, feature-based, or model-based.
+
+In the context of shape-based time-series clustering, it is common to utilize the Dynamic Time
+Warping (DTW) distance as dissimilarity measure (Aghabozorgi et al., 2015). The calculation of the
+DTW distance involves a dynamic programming algorithm that tries to find the optimum warping
+path between two series under certain constraints
 

### 2020-11-24 paper/v1/broadband.speed.covid.tex
started working on the methods sectrion. haven't finished yet :(

@@
 
-Description of the method
+The core of the methodological framework of this paper lies upon
+clustering methods. Clustering is a widely used family of techniques in
+geography (\textbf{ADD REF}), which can be defined within the modern
+machine learning framework as an unsupervised learning task, which
+involves partitioning unlabeled observations into homogeneous groups
+known as clusters \citep{montero2014tsclust}. The key idea is that
+observations within clusters tend to be more similar than observations
+between clusters. Clustering is particularly useful for exploratory
+studies as it identifies structures within the data
+\citep{aghabozorgi2015time}.
+
+Various applications of clustering can be found within geography:
+\textbf{ADD EXAMPLES}
+
+Key characteristic of these studies is the cross-sectional nature of the
+data they employed. Indeed, most clustering problems in geography deal
+with observations that are fixed in time. However, for this paper we are
+interested in creating clusters of local authorities in the UK with
+similar temporal signatures of experienced internet speeds over time.
+Hence, we deviate from the established geographical clustering tools and
+employ time series clustering methods.
+
+Time series clustering methods have been developed in order to deal with
+clustering problems linked to, for instance, stock or other financial
+data, economic, governmental or medical data as well as machine
+monitoring
+\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
+The main challenge -- and also the difference with cross-sectional
+clustering problems -- is data dimensionality given the multiplicity of
+data points for every individual included in the data set. Time series
+data are dynamic data as the value of the observations change as a
+function of time \citep{aghabozorgi2015time}. This high dimensionality
+leads to (i) computational and algorithmic challenges regarding handling
+these data and building algorithms to perform clustering over long time
+series, and (ii) open questions regarding the choice of similarity
+measures in order to cluster similar times series together considering
+the whole length of the time series and overcoming issues around noise,
+outliers and shifts \citep{lin2004iterative, aghabozorgi2015time}.
+
+Time series clustering methods utilising the whole length of time series
+can be grouped in three categories. The first -- model-based approaches
+-- is based on recovering the underlying model for each time series and
+then apply clustering algorithms on the model parameters
+\citep{aghabozorgi2015time}. The main criticism is the cluster accuracy
+for near clusters \citep{mitsa2010temporal}. The second approach is a
+based the formation of vectors of features based on the original time
+series. These new data of reduced dimensionality is then clustered using
+conventional clustering algorithms.
+
+For this paper we utilise the third category of time series clustering
+methods known as shape-based approaches. The main idea is to match any
+two sperate time series based on the similarity of their shapes through
+the calculation of distances among these two shapes.
+
+\citet{sardatime} Ideally, all members of the same cluster are similar
+to each other, but are as dissimilar as possible from objects in a
+different cluster.
+
+Time-series is a common type of dynamic data that naturally arises in
+many different scenarios, such as stock data, medical data, and machine
+monitoring, just to name a few (Aghabozorgi et al., 2015; Aggarwal and
+Reddy, 2013). They pose some challenging issues due to the large size
+and high dimensionality commonly associated with time-series
+(Aghabozorgi et al., 2015). In this context, dimensionality of a series
+is related to time, and it can be understood as the length of the
+series.
+
+the data and how the underlying grouping is performed. One
+classification depends on whether the whole series, a subsequence, or
+individual time points are to be clustered. On the other hand, the
+clustering itself may be shape-based, feature-based, or model-based.
+
+In the context of shape-based time-series clustering, it is common to
+utilize the Dynamic Time Warping (DTW) distance as dissimilarity measure
+(Aghabozorgi et al., 2015). The calculation of the DTW distance involves
+a dynamic programming algorithm that tries to find the optimum warping
+path between two series under certain constraints
 

### 2020-11-24 paper/v1/broadband.speed.covid.Rmd

--- 2020-11-19 paper/v1/broadband.speed.covid.Rmd
@@
 had been on a slow, upward trend, to a widespread way of life in many countries.
-The ability to work from home or telecommute meant millions retained their jobs
-and, to a varying extent, maintained productivity during periods of strict
+The ability to work from home or telecommute meant millions retained their jobs and,
+to a varying extent, maintained productivity during periods of strict
 lockdown around the world.
@@
 we consider whether the divide is also technological.
+
+<!-- our approach -->
 Using the UK as a case study, this paper aims to understand how the resilience
@@
 and geographic characteristics to explore how they overlap with or diverge from
-the existing economic and digital geography of the UK. **need to add in urban-rural**
+the existing economic and digital geography of the UK. **need to add urban-rural to analysis**
 Our analysis enables us to better understand how the spatial and social distribution 
@@
 
-<!-- Our framework: capability to WFH; HB old PARA2 -->
+<!-- capability to WFH -->
 The capability to work from home has previously been studied from the
@@
 of workers. 
-Instead, in the UK, for example, benchmarking broadband performance and service 
-delivery by different Internet Service Providers (ISPs) focused upon the quality of 
-service for evening leisure use, when video streaming services are at their peak, 
-rather than during the working day [@ofcom2017].
+Instead, leisure use in the evening, when video streaming services are at their peak, 
+has been used to benchmark broadband performance and service delivery by different 
+Internet Service Providers (ISPs), at least in the UK [@ofcom2017].
 Yet the shift towards telecommuting during various stages of lockdown around the world
 has been drastic and there are speculations that post-Covid, the tendency to work from
-home will be much higher than the pre-Covid, raising questions around whether 
-internet services can accommodate the increased demand.
+home will be much higher than pre-Covid, raising questions around whether internet 
+services can accommodate the increased demand.
 For example, $47$% of people in employment in the UK worked soley from home in April
@@
 could work from home on an ongoing basis [@batty2020editorial].
-Similar figures have been reported for other countries around the world
-[@felstead2020homeworking].
+Similar figures have been reported for other countries [@felstead2020homeworking].
 For instance, $37$% of the European workforce worked from home in April $2020$ with
@@
 In the US, almost half of the working population worked from home during the same
-period because of the pandemic [@brynjolfsson2020covid].
-Furthermore, a recent estimate indicated that $37$% of all jobs in the US can be
-permanently performed entirely from home [@NBERw26948].
+period because of the pandemic [@brynjolfsson2020covid], 
+and a recent estimate indicated that $37$% of all jobs in the US can be permanently 
+performed entirely from home [@NBERw26948].
 
@@
 [@blank2014dimensions; @van2011internet].
-The third level focuses on the differentiated returns of internet use **for...?**
+The third level focuses on the differentiated returns of internet use **for...? add a phrase here introducing economic resilience?**
 [@stern2009levels; @van2014digital; @van2015third].
@@
 potential returns of internet use for the user and wider community, 
-assumes skills that are present in some occupations but not in others, 
+assumes skills or functions that are present in some occupations but not in others, 
 and relies upon access to high quality internet services.
@@
 occupations are spatially dependent and clustered in space, our approach offers a 
-framework for understanding which types of places, socially and spatially, 
+framework for understanding which types of places,  
 are more likely to land on the right side of this new digital divide. 
 By asking how resilient broadband speeds are as experienced in different parts of the 
-UK during a time of extreme demand, we will better understand which places benefit 
-from greater economic resilience not only during the pandemic, but also into the 
-future.
+UK during a time of extreme demand, 
+we interrogate which places benefit from greater the economic resilience 
+digital technologies can offer, not only during the pandemic, 
+but also into the future.
 
@@
 
-...
+# Literature review
+
+## from telecommuting to #WFH
+
+<!-- Who works from home-->
+In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used 
+interchangeably, as most remote labour during the Covid-19 crisis was carried out 
+in the homes of individual employees rather than any other location [@eurofound2020]. 
+However, it should be noted that previous research has explored how telecommuting 
+can occur in other places, including satellite offices or on public transport 
+[@felstead2012rapid; @siha2006telecommuting].
+Previous research has also used a variety of definitions to measure the level of 
+telecommuting within different workforces, distinguishing, for example, between those 
+directly employed, indirectly employed, self-employed, full-time or part-time, and 
+those who use digital technologies to work remotely full-days or part-days 
+[@allen2015effective; @bailey2002review; @haddad2009examination].
+No matter the definition, the option and capability to telecommute or work from home 
+has never been equally distributed spatially or socio-economically any more than 
+different industries and employment opportunities have.
+For example, studies from the United States, the Netherlands, and the UK indicate 
+that telecommuters are most likely to hold professional, managerial, and technical 
+occupations, where the workforce is better educated and wealthier, and that there is
+suppressed demand among women and part-time workers
+[@headicar2016move; @peters2004employees; @singh2013modeling].
+Opportunities for working from home during the current pandemic have likewise not 
+been equally spread across the workforce.
+@NBERw26948 indicated that in the US, managers, educators, as well as those
+working in computer-related occupations, finance, and law can easily work from home,
+and that occupations with opportunities to telecommute are associated
+with higher earnings.
+This is not the case for the workforce occupied in more spatially fixed occupations,
+from farming, construction and manufacturing to hospitality and care services.
+In the US, these occupations tend to be lower-income, non-white, 
+without a university degree, live in rental accommodation and lack health insurance
+[@NBERw27085].
+Similar trends can be observed for other countries.
+For example, $75$% of workers with tertiary education worked from home in Europe
+during spring $2020$, whilst only $34$% of workers with secondary education and
+$14$% of those primary education did so [@eurofound2020].
+
+## digital divides and economic resilience
+<!-- WFH and digital divides-->
+Our understanding of telecommuting as a product of enabled occupations can be 
+described as a manifestation of the third level digital divide, as those who are able 
+to use digital technologies to work from home benefit from a high rate of return on 
+their use of the internet in terms of autonomy, flexibility, and time saved from 
+commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
+These returns have been even greater during the Covid-19 crisis, 
+when those with the ability to telecommute also have the ability to maintain 
+their employment whilst protecting their health.
+However, the success of these arrangements has been dependent upon the first 
+level digital divide, which is associated with access and quality of internet 
+connectivity.
+For example, the systematic review from @SALEMINK2017360 highlights the 
+infrastructure quality differences between urban and rural areas 
+in various advanced economies.
+Whether this variation in infrastructure quality affects the spatial footprint of 
+telecommuting has not previously been measured, although there are indications that
+those who purchase high speed connections consume more data of all sorts and use their
+connections for a variety of purposes [@hauge2011consumer].
+Another study identifies a correlation between access to internet services and a 
+reduction in household transport spend across 33 coutries [@bris2017ict]. 
+Whether the implication is less travel because of increased telecommuting, 
+or whether internet access enables more efficient travel, this finding is 
+an indication of the potential household savings better internet services offer.
+Thus, the extreme demand during the pandemic provides a new opportunity to 
+understand how infrastructure accessibility, quality, and reliability 
+affects telecommuting, particularly as working from home during the pandemic required
+high volumes of bandwidth-intensive video conferencing in order to avoid the 
+face-to-face contact that could increase the spread of infection.
+By first answering questions about internet service resilience, we can also refine our
+understanding of how this has contributed to or reduced the new digital divide, 
+where economic resilience has been dependent upon the capability to work from home.
+
+## covid and working from home, cities, urban structure
+This is aligned with the regional economic resilience literature, which
+underlines the differentiated capacity of cities and regions to escape or
+recover from economic crises [@martin2012regional; @kitsos2018economic]. 
+Different cities have different industrial and occupational profiles, 
+and these will affect the potential for telecommuting... 
+**Fill in more here - relate to 2nd and 3rd level digital divide, economic geography of UK**
+
+## other subsection?
+- circumstances where work can be carried out more flexibly in space and time
+- how relate to digital divides
+- capability theory of mobility?
+- broadband tech stuff? broadband studies / divides / resilience
+
+Some new papers google recommended to me:
+
+- https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf
+- https://link.springer.com/article/10.1007/s11116-020-10136-6
+- https://www.sciencedirect.com/science/article/pii/S0966692319311305
+- check who cites the above and what they cite
 
@@
 
-# Literature review
-
-## from telecommuting to #WFH
-
-<!-- Who works from home-->
-In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used 
-interchangeably, as most remote labour during the Covid-19 crisis was carried out 
-in the homes of individual employees rather than any other location [@eurofound2020]. 
-However, it should be noted that previous research has explored how telecommuting 
-can occur in other places, including satellite offices or on public transport 
-[@felstead2012rapid; @siha2006telecommuting].
-Previous research has also used a variety of definitions to measure the level of 
-telecommuting within different workforces, distinguising, for example, between those 
-directly employed, indirectly employed, self-employed, full-time or part-time, and 
-who use digital technologies to work remotely full-days or part-days 
-[@allen2015effective; @bailey2002review; @haddad2009examination].
-<!--old text: An appropriate definition of teleworking is ‘the remote provision of labour that would otherwise be carried out within company premises’ (European Commission, 2020b).-->
-No matter the definition, the option and capability to telecommute or work from home 
-has never been equally distributed spatially or socio-economically any more than 
-different industries and employment opportunities have.
-For example, studies from the United States, the Netherlands, and the UK indicate 
-that telecommuters are most likely to hold professional, managerial, and technical 
-occupations, where the workforce is better educated and wealthier, and that there is
-suppressed demand among women and part-time workers
-[@headicar2016move; @peters2004employees; @singh2013modeling].
-Opportunities for working from home during the current pandemic have likewise not 
-been equally spread across the workforce.
-@NBERw26948 indicated that in the US, managers, educators, as well as those
-working in computer-related occupations, finance, and law can easily work from
-home, and that occupations with opportunities to telecommute are associated
-with higher earnings.
-This is not the case for the workforce occupied in more spatially fixed occupations,
-from farming, construction, and manufacturing to hospitality and care services.
-In the US, these occupations tend to be lower-income, non-white, 
-without a university degree, live in rental accommodation and lack health insurance
-[@NBERw27085].
-Similar trends can be observed for other countries.
-For example, $75$% of workers with tertiary education worked from home in Europe
-during spring $2020$, whilst only $34$% of workers with secondary education and
-$14$% of those primary education did so [@eurofound2020].
-
-## digital divides and economic resilience
-<!-- WFH and digital divides-->
-Our understanding of telecommuting as a product of enabled occupations can be 
-described as a manifestation of the third level digital divide, as those who are able 
-to use digital technologies to work from home benefit from a high rate of return on 
-their use of the internet in terms of autonomy, flexibility, and time saved from 
-commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
-These returns have been even greater during the Covid-19 crisis, when those with the 
-ability to telecommute also have the ability to maintain their employment whilst 
-protecting their health.
-However, the success of these arrangements has also been dependent upon the first 
-level digital divide, which is associated with access and quality of internet 
-connectivity.
-For example, the systematic review from @SALEMINK2017360 highlights the 
-infrastructure quality differences between urban and rural areas.
-Whether this variation in infrastructure quality affects the spatial footprint of 
-telecommuting has not previously been measured, although there are indications that
-those who purchase high speed connections consume more data of all sorts and use their
-connections for a variety of purposes [@hauge2011consumer].
-Another study identifies a correlation between access to internet services and a 
-reduction in household transport spend across 33 coutries, although the authors are 
-uncertain whether this implies less travel, for example because of increased
-telecommuting, or whether internet access enables more efficient travel 
-[@bris2017ict].
-However, the extreme demand during the pandemic provides a new opportunity to 
-understand how infrastructure accessibility, quality, and reliability affects 
-telecommuting, particularly as working from home during the pandemic required
-high volumes of bandwidth-intensive video conferencing in order to avoid the 
-face-to-face contact that could increase the spread of infection.
-Thus by answering questions about internet service resilience, we can also refine our
-understanding of how this has contributed to or reduced the new digital divide where
-economic resilience has been dependent upon the capability to work from home.
-
-<!-- Econ resilience-->
-This is aligned with the regional economic resilience literature, which
-underlines the differentiated capacity of cities and regions to escape or
-recover from economic crises [@martin2012regional; @kitsos2018economic].
-
-## other subsection?
-
-- Introduce space-time geography
-- circumstances where work can be carried out more flexibly in space and time
-- how relate to digital divides
-- capability theory of mobility? suppressed demand...
-
-Some new papers google recommended to me:
-
-- https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf
-- https://link.springer.com/article/10.1007/s11116-020-10136-6
-- https://www.sciencedirect.com/science/article/pii/S0966692319311305
-- check who cites the above and what they cite
-
-## covid and working from home, cities, urban structure
-- economic geography of UK
-
 # Data and descriptive statistics
 
-<!-- HB old PARA3 -->
+<!-- PARA introduce broadband speeds -->
 This paper assesses the quality and reliability of upload speeds by time of day and 
@@
 test their experienced broadband speeds, upload and download. 
+This primary data was provided by Speedchecker Ltd, 
+a private company that allows internet users to check their own broadband upload / 
+download speeds, and stores every speed-check with a timestamp and geolocation.
 Since the quality and reliability of internet services vary in time and space due to 
@@
 as possible.
-We start by calculating the mean upload speed for every local government district in 
-the UK for each hour of each weekday, excluding midnight to 6:00, weekends and bank 
-holidays. 
-We then cluster the local authorities by these temporal profiles, allowing us to 
-identify patterns and describe the overall means, standard deviations, and other 
-relevant statistics for each spatial cluster. 
+The following **graphs / tables** show these statistics for the sample as a whole,
+as well as similar statistics related to frequency of testing.
+
+**Data details and some figures, descriptive stats - include whole sample time profile for 2019 and 2020 frequency of tests run as part of why we chose to create the time profiles by hour of the day and day of the week rather than daily over the whole period.**
+
+**Just looking back, it was partly the frequency of tests, but that was also by geography, so do we need maps? And I feel there was something else I'm forgetting...**
+
+# Time series clustering
+
+Clustering methods are a widely used family of techniques in **economic?** geography (**ADD REF**).
+It can be defined within the modern machine learning framework as an unsupervised
+learning task, which involves partitioning unlabeled observations into homogeneous
+groups known as clusters [@montero2014tsclust]. 
+The key idea is that observations within clusters tend to be more similar than 
+observations between clusters.
+Clustering is particularly useful for exploratory studies as it identifies
+structures within the data [@aghabozorgi2015time].
+
+Various applications of clustering can be found within geography: **ADD EXAMPLES**
+
+Key characteristic of these studies is the cross-sectional nature of the data they
+employed. 
+Indeed, most clustering problems in geography deal with observations 
+that are fixed in time. 
+However, for this paper we are interested in creating clusters of local authorities 
+in the UK with similar temporal signatures of experienced internet speeds over time. 
+Hence, we deviate from the established geographical clustering tools and 
+employ time series clustering methods.
+
+Time-series is a common type of dynamic data that naturally arises in many 
+different scenarios, such as stock data, medical data, and machine monitoring, 
+just to name a few **(Aghabozorgi et al., 2015; Aggarwal and Reddy, 2013)**. 
+They pose some challenging issues due to the large size and high dimensionality 
+commonly associated with time-series **(Aghabozorgi et al., 2015)**. 
+In this context, dimensionality of a series is related to time, 
+and it can be understood as the length of the series.
+
+Time series clustering methods have been developed in order to deal with clustering
+problems linked to, for instance, stock or other financial data, economic,
+governmental or medical data as well as machine monitoring
+[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
+The main challenge -- and also the difference with cross-sectional clustering
+problems -- is data dimensionality given the multiplicity of data points for every
+individual included in the data set. 
+Time series are dynamic data as the value of the observations change as 
+a function of time [@aghabozorgi2015time].
+This high dimensionality leads to 
+(i) computational and algorithmic challenges regarding handling these data and 
+building algorithms to perform clustering over long time series, 
+and (ii) open questions regarding the choice of similarity measures in order to 
+cluster similar times series together considering the whole length of the time series 
+and overcoming issues around noise, outliers and shifts [@lin2004iterative; 
+@aghabozorgi2015time].
+
+Time series clustering methods utilising the whole length of time series can be
+grouped in three categories. 
+The first -- model-based approaches -- is based on recovering the underlying model 
+for each time series and then apply clustering algorithms on the model parameters 
+[@aghabozorgi2015time]. 
+The main criticism is the cluster accuracy for near clusters [@mitsa2010temporal]. 
+The second approach is a based the formation of vectors of features based on the 
+original time series.
+These new data of reduced dimensionality is then clustered using conventional
+clustering algorithms.
+
+For this paper we utilise the third category of time series clustering methods
+known as shape-based approaches. 
+The main idea is to match any two sperate time series based on the similarity 
+of their shapes through the calculation of distances among these two shapes.
+
+@sardatime
+Ideally, all members of the same cluster are similar to each other, but are as
+dissimilar as possible from objects in a different cluster.
+
+the data and how the underlying grouping is performed. 
+One classification depends on whether the whole series, a subsequence, 
+or individual time points are to be clustered. 
+On the other hand, the clustering itself may be shape-based, feature-based, or model-based.
+
+In the context of shape-based time-series clustering, it is common to utilize the Dynamic Time Warping (DTW) distance as dissimilarity measure **(Aghabozorgi et al., 2015)**. 
+The calculation of the DTW distance involves a dynamic programming algorithm 
+that tries to find the optimum warping path between two series under certain constraints
+
+**moved this section from data and descriptives as need to end methods of clustering with how it applies to what we did, but needs further edit after you're done**
+To apply these methods to our data, 
+we aggregate all the speed-checks during the 13 weeks of March 
+to May inclusive for weekdays in 2020 by each hour of the day and day of the week. 
+As our research aims to identify the geography of internet service resilience for 
+work purposes, bank holidays and the hours between midnight and 6am were also 
+excluded. 
+The composite week time series thus comprise 18 hours multiplied by 5 weekdays or 
+90 time points per series. 
+These time series were calculated for each of 
+the 382 Local Authority Districts (LADs) in the UK, standardised, 
+and then a k-means clustering algorithm was applied, including dynamic time warping. 
+The LADs were assigned to 10 clusters for upload speeds. 
+The cluster id for each LAD is then reattached to the speed-check dataframe to 
+identify the characteristics of each cluster, including number of LADs, and the 
+descriptives statistics of upload speeds in that cluster, 
+and the temporal profile by hour of the day and day of the week.
 Thus we aim to answer the first part of our research question: 
@@
 
-<!-- HB old PARA4 -->
+<!-- HB old PARA4 - link to auxillary regressions method, or put in results to introduce results of auxillary regressions? -->
 The cause of these different experiences of broadband resilience may be different in 
-different areas, as they may reflect similarities in patterns of demand or similar 
-quality of infrastructure. 
+different areas, as they may reflect either similarities in patterns of demand 
+or similar quality of infrastructure. 
 Our approach is also limited by potential endogeneity, as for example, better quality 
@@
 
-The primary data analysed in this paper was provided by Speedchecker Ltd, a private company that allows internet users to check their own broadband upload / download speeds, and stores every speed-check with a timestamp and geolocation.
-
-Our approach involves aggregating all the speed-checks during the 13 weeks of March to May inclusive for weekdays in 2020 by each hour of the day and day of the week.
-As our research aims to identify the geography of internet service resilience for work purposes, bank holidays and the hours between midnight and 6am were also excluded.
-The composite week time series thus comprise 18 hours multiplied by 5 weekdays or 90 time points per series.
-These time series were calculated for each of the 382 Local Authority Districts (LADs) in the UK, standardised, and then a k-means clustering algorithm was applied, including dynamic time warping.
-The LADs were assigned to 10 clusters for upload speeds.
-The cluster id for each LAD is then reattached to the speed-check dataframe to identify the characteristics of each cluster, including number of LADs, and the descriptives statistics of upload speeds in that cluster, and the temporal profile by hour of the day and day of the week.
-
-**Data details and some figures, descriptive stats - include whole sample time profile for 2019 and 2020 frequency of tests run as part of why we chose to create the time profiles by hour of the day and day of the week rather than daily over the whole period.**
-
-# Time series clustering
-
-Description of the method
-
 # Results
@@
 
-The temporal profiles used to cluster the local authorities have been summarised in graph [.], which shows a composite profile of mean upload speeds per hour per day for each cluster.
-For upload speeds, 345 of 382 local authorities, or over 62 million people, fall into cluster 6 or cluster 9.
-Graph [.] shows that both of these clusters have relatively similar temporal profiles, which are flatter than the other, smaller clusters, suggesting better reliability of service.
-However, the upload speeds at all times for cluster 9 are substantially higher than for cluster 6, which is an indication of better quality of service.
-This difference is reflected not only in the mean speeds for these clusters for the whole sample, but also the mean upload speeds in the morning peak from 9:00-10:59, as well as the evening peak period from 19:00-20:59.
-In comparison, the time profile in graph [.] shows upload speeds in cluster 1 are on average lower at certain times of day during the study period than any other cluster, whilst the profile for cluster 3 appears to show speeds fluctuating as much as cluster 1, but at levels usually higher even than cluster 9.
+The temporal profiles used to cluster the local authorities have been summarised in 
+graph [.], which 
+shows a composite profile of mean upload speeds per hour per day for each cluster. 
+For upload speeds, 345 of 382 local authorities, or over 62 million people, 
+fall into cluster 6 or cluster 9. 
+Graph [.] suggests that these two clusters have relatively similar temporal profiles, 
+which are flatter than the other, smaller clusters, 
+suggesting better reliability of service. 
+However, the upload speeds at all times for cluster 9 are substantially higher 
+than for cluster 6, which is an indication of better quality of service. 
+This difference is reflected not only in the mean speeds for these clusters for the 
+whole sample, but also the mean upload speeds in the morning peak from 9:00-10:59, 
+as well as the evening peak period from 19:00-20:59. 
+In comparison, the time profile in graph [.] shows upload speeds in cluster 1 are 
+on average lower at certain times of day during the study period than any other 
+cluster, whilst the profile for cluster 3 appears to show speeds fluctuating as much 
+as cluster 1, but at levels usually higher even than cluster 9.
 
-The variability of the smaller clusters may be related to the fewer speed tests from fewer local authorities that have been averaged, whilst averaging greater numbers of speed tests could artificially flatten the profile.
-This appears not to be the case for cluster 9, as there is negligible difference between morning and evening upload speeds, at 1% slower in the morning, confirming a high level of reliability of service.
-In comparison, upload speeds in cluster 6 are 4% slower in the morning than in the evening and experienced the joint highest ratio of standard deviation to mean across the time period under assessment.
-This suggests that although the time profile is relatively flat in graph [.], the experience is one of speeds that fluctuate from a lower mean, and therefore might more often impact on online activities.
-Still, in most of the smaller clusters the reliability of service is worse than in cluster 9 in terms of the ratio of standard deviation to mean and worse than cluster 6 as well in terms of the ratio of upload speed in the morning peak compared to the evening peak.
-In five of the smaller clusters, which are home to almost 3 million people, speeds are 12-19% slower between 9:00-10:59 compared to 19:00-20:59.
+The variability of the smaller clusters may be related to the fewer speed tests from 
+fewer local authorities that have been averaged, whilst averaging greater numbers of 
+speed tests could artificially flatten the profile. 
+This appears not to be the case for cluster 9, as there is negligible difference 
+between morning and evening upload speeds, at 1% slower in the morning, confirming a 
+high level of reliability of service. 
+In comparison, upload speeds in cluster 6 are 4% slower in the morning than in the 
+evening and experienced the joint highest ratio of standard deviation to mean across 
+the time period under assessment. 
+This suggests that although the time profile is relatively flat in graph [.], 
+the experience is one of speeds that fluctuate from a lower mean, 
+and therefore might more often impact on online activities. 
+Still, in most of the smaller clusters the reliability of service is worse than in 
+cluster 9 in terms of the ratio of standard deviation to mean. 
+The smaller clusters are mainly worse than clusters 9 and 6 in terms of 
+the ratio of upload speed in the morning peak compared to the evening peak. 
+In five of the smaller clusters, which are home to almost 3 million people, 
+speeds are 12-19% slower between 9:00-10:59 compared to 19:00-20:59.
 
-The quality and reliability of broadband service is thus much better in the 115 Local Authority areas in cluster 9, which are mostly in urban or suburban areas, compared to cluster 6.
+The quality and reliability of broadband service is much better in 
+the 115 Local Authority areas in cluster 9, 
+which are mostly in urban or suburban areas, compared to cluster 6. 
 These include 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation.
@@
 # Conclusions
+
 <!-- WFH broader consequences-->
@@
 use and urban planning to accommodate people who work from home
-[@BUDNITZ2020102713]; and from productivity and innovation changes, to changes
+[@BUDNITZ2020102713]**also 2020 Swedish article from JTG**; and from productivity and innovation changes, to changes
 in agglomeration externalities and the attraction of large cities [@econobs]
@@
 service, and whether this reinforces or redresses prior digital divisions.
-**HANNAH, HOW DOES THIS SOUND - this last bit probably in the discussion at the end?**
+**took this bit you wrote to put in the discussion at the end?**
 

### 2020-11-24 paper/v1/broadband.speed.covid.Rmd
changes after we spoke on 24/11

--- 2020-11-24 paper/v1/broadband.speed.covid.Rmd
@@
 series based on the similarity of their shapes through the calculation of
-distances among these two shapes.
+distances among these two shapes. Such approached are better equipped to capture
+similarities between short length time series [@aghabozorgi2015time].
+This approach serves best this paper because (i) we aim to identify clusters of
+UK local authorities with similar temporal signatures -- i.e. shapes -- of
+experienced internet speeds and (ii) the length of our time series is short
+(**see also discussion in Section XX**).
+
+Other important dimensions regarding time series clustering are related with the
+clustering algorithm. Similar to clustering of cross-sectional data, we can employ
+partitioning algorithms, which lead to non-overlapping clusters, hierarchical
+clustering, which result to a hierarchy of clusters and fuzzy algorithms, which
+create overlapping clusters [@sardatime]. Because of the simplicity of the
+implementation and the interpretability of the results, we utilise here
+partitioning clustering based on the widely *k-Means* algorithm. This is an iterative
+algorithm, which begins with defining the desired number of clusters *k*. Then,
+each observation is randomly assigned to a cluster from the $[1,k]$ space. This
+is the initial cluster assignment, which is followed by iterations in order to
+minimise the distance between the centroids of the clusters and the observations
+assigned to these clusters [@james2013introduction].
+
+There are a number of differences between the above described application of
+*k-Means* for cross-sectional data and its application for times series data.
+
+k-Meidoids
+distance between points in time series
+dt
+
+comparison with density-based clustering DBSCAN [187]
+where a cluster is expanded if its neighbours are dense.
+
+- evaluation
+- standardised
 
 @sardatime
-Ideally, all members of the same cluster are similar to each other, but are as
-dissimilar as possible from objects in a different cluster.
-
-Time-series is a common type of dynamic data that naturally arises in many different scenarios,
-such as stock data, medical data, and machine monitoring, just to name a few (Aghabozorgi et al.,
-2015; Aggarwal and Reddy, 2013). They pose some challenging issues due to the large size and high
-dimensionality commonly associated with time-series (Aghabozorgi et al., 2015). In this context,
-dimensionality of a series is related to time, and it can be understood as the length of the series.
-
-the data and how the underlying grouping is performed. One classification depends on whether the
-whole series, a subsequence, or individual time points are to be clustered. On the other hand, the
-clustering itself may be shape-based, feature-based, or model-based.
-
 In the context of shape-based time-series clustering, it is common to utilize the Dynamic Time

### 2020-11-25 paper/v1/broadband.speed.covid.Rmd
pasting from HB's emailed .Rmd

--- 2020-11-24 paper/v1/broadband.speed.covid.Rmd
@@
 For this paper we utilise the third category of time series clustering methods
-known as shape-based approaches. 
-The main idea is to match any two sperate time series based on the similarity 
-of their shapes through the calculation of distances among these two shapes.
+known as shape-based approaches. The main idea is to match any two sperate time
+series based on the similarity of their shapes through the calculation of
+distances among these two shapes. Such approached are better equipped to capture
+similarities between short length time series [@aghabozorgi2015time].
+This approach serves best this paper because (i) we aim to identify clusters of
+UK local authorities with similar temporal signatures -- i.e. shapes -- of
+experienced internet speeds and (ii) the length of our time series is short
+(**see also discussion in Section XX**).
+
+Other important dimensions regarding time series clustering are related with the
+clustering algorithm. Similar to clustering of cross-sectional data, we can employ
+partitioning algorithms, which lead to non-overlapping clusters, hierarchical
+clustering, which result to a hierarchy of clusters and fuzzy algorithms, which
+create overlapping clusters [@sardatime]. Because of the simplicity of the
+implementation and the interpretability of the results, we utilise here
+partitioning clustering based on the widely *k-Means* algorithm. This is an iterative
+algorithm, which begins with defining the desired number of clusters *k*. Then,
+each observation is randomly assigned to a cluster from the $[1,k]$ space. This
+is the initial cluster assignment, which is followed by iterations in order to
+minimise the distance between the centroids of the clusters and the observations
+assigned to these clusters [@james2013introduction].
+
+There are a number of differences between the above described application of
+*k-Means* for cross-sectional data and its application for times series data.
+
+k-Meidoids
+distance between points in time series
+dt
+
+comparison with density-based clustering DBSCAN [187]
+where a cluster is expanded if its neighbours are dense.
+
+- evaluation
+- standardised
 
 @sardatime
-Ideally, all members of the same cluster are similar to each other, but are as
-dissimilar as possible from objects in a different cluster.
-
-the data and how the underlying grouping is performed. 
-One classification depends on whether the whole series, a subsequence, 
-or individual time points are to be clustered. 
-On the other hand, the clustering itself may be shape-based, feature-based, or model-based.
-
-In the context of shape-based time-series clustering, it is common to utilize the Dynamic Time Warping (DTW) distance as dissimilarity measure **(Aghabozorgi et al., 2015)**. 
-The calculation of the DTW distance involves a dynamic programming algorithm 
-that tries to find the optimum warping path between two series under certain constraints
+In the context of shape-based time-series clustering, it is common to utilize the Dynamic Time
+Warping (DTW) distance as dissimilarity measure (Aghabozorgi et al., 2015). The calculation of the
+DTW distance involves a dynamic programming algorithm that tries to find the optimum warping
+path between two series under certain constraints
 

### 2020-11-27 paper/v1/broadband.speed.covid.Rmd
after conflicts resolved, addition in intro, lit rev, methods and data.

@@
 keywords: |
-  covid; internet; working from home; broadband speed; time series clusters
+  covid; internet; working from home; broadband speed; time-series clusters
 header-includes: |
@@
 
+```{r settings, include = FALSE}
+library(ggplot2)
+library(tidyverse)
+library(geojsonio)
+library(tsfeatures)
+library(rprojroot)
+library(openair)
+library(reshape2)
+library(lubridate)
+library(rgdal)
+library(maptools)
+library(sp)
+library(rgeos)
+library(classInt)
+library(RColorBrewer)
+library(knitr)
+library(rprojroot)
+library(viridis)
+library(hrbrthemes)
+
+# This is the project path
+path <- find_rstudio_root_file()
+```
+
 # Introduction
 
-- [our google doc](https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit#heading=h.i5om1o8wpcd9)
+<!-- [our google doc](https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit#heading=h.i5om1o8wpcd9) -->
 
@@
 workers, who are potentially at high risk of infection.
-Whilst the basis for this new digital divide can be viewed as mainly occupational, 
-we consider whether the divide is also technological.
+Whilst the basis for this new digital divide has been viewed as mainly occupational,
+here we consider whether the divide is also technological.
 
 <!-- our approach -->
-Using the UK as a case study, this paper aims to understand how the resilience
-of broadband speeds in terms of both quality and reliability of service may reinforce
-or redress the spatial and social dimensions of the digital division exposed by the
-pandemic.
+Using the UK as a case study, this paper aims to understand how the quality and
+reliability of internet service, as reflected in *experienced* internet speeds,
+may reinforce or redress the spatial and social dimensions of the digital division
+exposed by the pandemic.
 To do so, we employ volunteered geographic data on individual broadband speed tests
@@
 telecommuters was never tested at scale, as working from home and connecting to
-colleagues and workplace resources via broadband was the purview of a small minority 
-of workers. 
+colleagues and workplace resources via the internet was the purview of a small minority of workers.
 Instead, leisure use in the evening, when video streaming services are at their peak,
@@
 <!-- ICT infrastructure and digital divides-->
-None of these changes could have happened in the absence of reliable ICT
-infrastructure -- both in terms of software and hardware.
+None of these changes could have happened in the absence of reliable information
+and communication technology (ICT) infrastructure -- both in terms of software
+and hardware.
 But while software innovations are easily diffused across space and
-society^[See for example the huge success of videoconferencing apps
+society^[See for example the huge success of videoconferencing apps such as Zoom
 [@marks2020zoom].],
@@
 broadband connectivity.
-The literature describes first level digital divides in terms of the availability and
-quality of internet connectivity, such as that manifest in different geographies 
-in the UK [@riddlesden2014broadband; @philip2017digital]. 
+The literature describes first level digital divides in terms of the
+availability and quality of internet connectivity, such as that manifest in
+different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
 Second level digital divides consider the presence or lack of the necessary skills
@@
 [@blank2014dimensions; @van2011internet].
-The third level focuses on the differentiated returns of internet use **for...? add a phrase here introducing economic resilience?**
+The third level focuses on the heterogenous returns of internet usage among
+different socioeconomic groups and, consequently, how digital technologies can
+assist in bridging or further enhancing existing socioeconomic divides.
 [@stern2009levels; @van2014digital; @van2015third].
@@
 but more importantly leads to differentiated outcomes regarding the economic
-resilience of people and places during the current pandemic.
+resilience of people and places to overcome a systemic shock such as the current
+pandemic.
 We identify this as a new digital divide, one that fundamentally alters the
@@
 and relies upon access to high quality internet services.
+**WE NEED TO CONNECT OUR FINDINGS WITH THIS AND JUSTIFY THAT THIS IS INDEED A NEW DIVIDE**
 As the quality of internet infrastructure and services, as well as variation in
 occupations are spatially dependent and clustered in space, our approach offers a
-framework for understanding which types of places,  
-are more likely to land on the right side of this new digital divide. 
-By asking how resilient broadband speeds are as experienced in different parts of the 
-UK during a time of extreme demand, 
-we interrogate which places benefit from greater the economic resilience 
-digital technologies can offer, not only during the pandemic, 
-but also into the future.
+framework for understanding which types of places, are more likely to land on the
+right side of this new digital divide.
+By asking how resilient broadband speeds are as experienced in different parts of
+the UK during a time of extreme demand, we interrogate which places benefit from
+the greater economic resilience digital technologies can offer, not only during
+the pandemic, but also into the future.
 
 The structure of this paper is as follows.
-First we will review the literature on telecommuting and digital divides to better 
+First we review the literature on telecommuting and digital divides to better
 understand the origins of the new digital divide revealed by the pandemic and its
@@
 We then describe our data and methodology.
-Our results section first describes the how internet services vary in the 
-different clusters and then assesses whether these clusters replicate or repudiate
-other socio-economic and geographic patterns of economic resilience.
+Our results section first offers classification of how internet services vary
+across the UK local authorities and then assesses whether these clusters replicate
+or repudiate other socio-economic and geographic patterns of economic resilience.
 
@@
 
-## from telecommuting to #WFH
+## From telecommuting to #WFH
 
@@
 
-## digital divides and economic resilience
+## Digital divides and economic resilience
 <!-- WFH and digital divides-->
@@
 Another study identifies a correlation between access to internet services and a
-reduction in household transport spend across 33 coutries [@bris2017ict]. 
+reduction in household transport spend across 33 countries [@bris2017ict].
 Whether the implication is less travel because of increased telecommuting,
@@
 
-## covid and working from home, cities, urban structure
-This is aligned with the regional economic resilience literature, which
+<!-- Covid-19 and working from home, cities, urban structure -->
+The multi-layered digital divides intersect with materials divides and the economic
+geography of the UK.
+Following the regional economic resilience literature, which
 underlines the differentiated capacity of cities and regions to escape or
-recover from economic crises [@martin2012regional; @kitsos2018economic]. 
-Different cities have different industrial and occupational profiles, 
-and these will affect the potential for telecommuting... 
-**Fill in more here - relate to 2nd and 3rd level digital divide, economic geography of UK**
+recover from economic crises [@martin2012regional; @kitsos2018economic],
+different places have different industrial and occupational profiles,
+and these affect the aggregated potential capacity of places for telecommuting.
+Such profiles are associated with longstanding inequalities in the UK and their
+spatial representation as a North-South divide [@martin_north_south].
+Various studies have illustrated severe inequalities between the north and the
+south part of the UK regarding, for example, skills and human capital,
+unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
+@dorling2018peak].
+Some scholars have even argued that the UK suffers some of the highest level of
+interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
+Not only all three levels of digital divides are, to a certain extend, associated
+with or shaped by the geography of the UK, but the intersection of the digital and
+material divides affects the capacity of places to overcome, at least partially,
+the economic effects of the Covid-19 pandemic.
+Importantly, this is the first time that digital technologies became an essential
+tool for economic resilience for such a great part of the population.
 
 ## other subsection?
+**I THINK WE HAVE ENOUGH OF A LIT REV**
 - circumstances where work can be carried out more flexibly in space and time
@@
 
-# Data and descriptive statistics
+# Methods and data
 
-<!-- PARA introduce broadband speeds -->
-This paper assesses the quality and reliability of upload speeds by time of day and 
-day of the week across local authorities in the UK during the time when the population
-were told to work from home if at all possible.
-The experience of upload speeds is not the same as the maximum speeds offered by an 
-ISP or possible speeds over a particular type and length of connection. 
-We use volunteered geographic information in the form of speed-checks run by users to 
-test their experienced broadband speeds, upload and download. 
-This primary data was provided by Speedchecker Ltd, 
-a private company that allows internet users to check their own broadband upload / 
-download speeds, and stores every speed-check with a timestamp and geolocation.
-Since the quality and reliability of internet services vary in time and space due to 
-both supply and demand-side influences, we use a number of different measures of 
-experienced upload speeds. 
+<!-- clustering in general -->
+The starting point of our methodological framework is cluster analysis, which can
+be defined within the modern machine learning framework as an unsupervised
+learning task, which involves partitioning unlabelled observations into homogeneous
+groups known as clusters [@montero2014tsclust].
+The key idea is that observations within clusters tend to be more similar than
+observations between clusters.
+Clustering is particularly useful for exploratory studies as it identifies
+structures within the data [@aghabozorgi2015time].
+Therefore, cluster analysis is a widely used family of techniques in geography
+[@gordon1977classification; @everitt1974cluster].
+For instance, clustering methods is the basis of *geodemographics*, which aims
+to create small area indicators or typologies of neighbourhoods based on various
+and some times diverse variables [@SINGLETON2009289; @harris2005geodemographics].
+Clustering techniques have also been employed to solve regionalisation problems
+[@niesterowicz2016].
+
+<!-- time-series clustering -->
+Common characteristic of these studies is the cross-sectional nature of the data
+they employ.
+Indeed, most clustering problems in geography deal with observations that are
+fixed in time.
+However, for this paper we are interested in creating clusters of local authorities
+in the UK with similar temporal signatures of experienced internet speeds over time.
+Hence, we deviate from the established geographical clustering tools and
+employ time-series clustering methods.
+
+Time-series clustering methods have been developed in order to deal with clustering
+problems linked to, for instance, stock or other financial data, economic,
+governmental or medical data as well as machine monitoring
+[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
+The main challenge -- and also the difference with cross-sectional clustering
+problems -- is data dimensionality given the multiplicity of data points for every
+individual object -- local authority in our case -- included in the data set.
+Time-series are dynamic data as the value of the observations change as
+a function of time [@aghabozorgi2015time].
+This high dimensionality leads to
+(i) computational and algorithmic challenges regarding handling these data and
+building algorithms to perform clustering over long time-series,
+and (ii) open questions regarding the choice of similarity measures in order to
+cluster similar times series objects together considering the whole length of the time-series
+and overcoming issues around noise, outliers and shifts [@lin2004iterative;
+@aghabozorgi2015time].
+
+Time-series clustering methods utilising the whole length of time-series can be
+grouped in three categories.
+The first -- model-based approaches -- is based on recovering the underlying model
+for each time-series and then applying clustering algorithms on the model parameters
+of each time-series [@aghabozorgi2015time].
+The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].
+The second approach is a based the formation of vectors of features based on the
+original time-series.
+These new data of reduced dimensionality is then clustered using conventional
+clustering algorithms.
+
+For this paper we utilise the third category of time-series clustering methods
+known as shape-based approaches.
+These methods match two sperate time-series objects based on the similarity of
+their shapes through the calculation of distances among these two shapes and are
+better equipped to capture similarities between short length
+time-series [@aghabozorgi2015time].
+This approach serves best this paper because (i) we aim to identify clusters of
+UK local authorities with similar temporal signatures -- i.e. shapes -- of
+experienced internet speeds and (ii) the length of our time-series is short
+(see the data discussion in this section).
+
+Another import element of time-series clustering is the actual clustering algorithm.
+Similar to clustering of cross-sectional data, we can employ partitioning
+algorithms, which lead to non-overlapping clusters, hierarchical
+clustering, which result to a hierarchy of clusters and fuzzy algorithms, which
+create overlapping clusters [@sardatime].
+Because of the simplicity of the implementation and the interpretability of the
+results, we utilise here partitioning clustering based on the widely used *k-Means*
+algorithm.
+This is an iterative algorithm, which begins with defining the desired number of
+clusters *k*.
+Then, each observation is randomly assigned to a cluster from the $[1,k]$ space.
+This is the initial cluster assignment, which is followed by iterations in order
+to minimise the distance between the centroids of the clusters and the observations
+assigned to these clusters [@james2013introduction].
+
+<!-- difference between clustering cross-sectional and time-series data -->
+There are a number of differences between the above described application of
+*k-Means* for cross-sectional data and its application for times series data.
+Instead of creating clusters around centroids, a common approach is to create
+clusters around *medoids*, which are representative time-series objects with a
+minimal distance to all other cluster objects [@sardatime].
+Also, instead of calculating the Euclidean distance between centroids and data
+points, more complex distance measures need to be employed in order to capture
+the similarity between a time-series object and a medoid.
+A common distance measure for shape-based time-series clustering in the Dynamic
+Time Warping (DTW).
+Using its underpinning dynamic programming algorithm, DTW compares two time-series
+objects to find the optimum warping path between them.
+DTW is widely used in order to overcome limitations linked to the use of Euclidean
+distance [@sardatime; @berndt1994using; @ratanamahatana2004everything].
+The `R` package `dtwclust` has been used for the time-series clustering [@dtwclust].
+
+<!-- introduction of our broadband speeds data -->
+To assess the quality and reliability of internet across local authorities
+in the UK during the time when the population were told to work from home if at
+all possible we utilise unique data regarding individual internet speed tests from
+Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
+This a private company that allows internet users to check their own broadband
+upload and download speeds, and stores every speed-check with timestamp and
+geolocation information.
+These data have been used before to assess digital divides [@riddlesden2014broadband]
+and the impact of local loop unbundling regulatory processes [@nardotto2015unbundling].
+These volunteered geographic data enable to assess the *experienced* internet
+speed by users, which may differ from the *advertised* maximum speeds provided
+by Internet Service Providers.
+We are particularly interested in upload speeds and the frequency of speed tests.
+While the former is less associated with internet-based leisure activities such
+as video streaming, which involve downloading large amounts of data, the latter
+can be linked more with work-related activities such as uploading documents or
+two-way communication and interaction systems.
+**ADD MORE. HOW ABOUT SPEED TESTS?**
+
+<!-- data transformation -->
+The first step in the workflow after dropping some outliers following
+@riddlesden2014broadband is to transform and the individual,
+geolocated and time-stamped tests to more meaningful aggregates both in terms
+of space and time.
+Regarding the temporal dimension, we aggregate all the speed-checks during the $13$
+weeks of March to May inclusive for weekdays in $2020$ by each hour of the day and
+day of the week.
+As our research aims to identify the geography of internet service resilience for
+work purposes, bank holidays and the hours between midnight and 6am were also
+excluded.
+The composite week time-series thus comprise 18 hours multiplied by 5 weekdays or
+90 time points per series.
+We aggregate these data because, we could not follow individuals or households
+and connect data points.
+Also, although this is a large data set ($241,088$ individual tests during the
+working hours of study period), there are not enough observations for each local
+and for each working hour ($631$ speed test per LAD in average).
+These time-series were calculated for each of the 382 Local Authority Districts
+(LADs) in the UK, standardised, and then a *k*-means partitioning around medoids
+clustering algorithm was applied using DTW.
+We initially run the algorithm for $k = 5, 10, 15 and 20$ and used cluster validity
+indices (CVIs) to pick the optimal solution of $k = 10$.
+Following @sardatime the following CVIs were employed: Silhouette (max), Score
+function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin (DB*, min), Dunn (max), COP (min).
+
+```{r echo=FALSE, eval=FALSE, message=FALSE}
+
+# Code to calculate the above figures about N.
+
+#upload data file
+path.data <- paste(path, "./data/temp/TSbb19_20sp.csv", sep = "")
+TSbb19_20sp <- read.csv(path.data)
+TSbb20sp <- TSbb19_20sp %>%
+  filter(dateOnly> "2019-12-31")
+
+# N. of observations for study period  
+print(nrow(TSbb20sp))
+
+# observations for study period per local authority
+print(nrow(TSbb20sp)/length(unique(TSbb20sp$LAD19NM)))
+```
+
+We then match the cluster membership to the LAD-level data to identify the
+characteristics of each cluster, including number of LADs, the descriptive
+statistics of upload speeds in that cluster, and the temporal profile by hour of
+the day and day of the composite week.
+Thus we aim to answer the first part of our research question:
+How resilient are broadband speeds as experienced in different parts of the UK
+during a time of extreme demand?
+
+<!-- metrics -->
+Since the quality and reliability of internet services vary in time and space due
+to both supply and demand-side influences, we use a number of different measures
+of experienced upload speeds.
 These include: a) mean, experienced connection speed,
 b) standard deviation or the amount of fluctuation from the mean, and
-c) the variation in speeds at particular times of day when working from home is more 
-likely to take place. 
-We take account of all three measurements in order to describe upload speeds as fully 
-as possible.
+c) the variation in speeds at particular times of day when working from home is
+more likely to take place.
+We take account of all three measurements in order to describe upload speeds as
+fully as possible.
 The following **graphs / tables** show these statistics for the sample as a whole,
@@
 
-# Time series clustering
-
-Clustering methods are a widely used family of techniques in **economic?** geography (**ADD REF**).
-It can be defined within the modern machine learning framework as an unsupervised
-learning task, which involves partitioning unlabeled observations into homogeneous
-groups known as clusters [@montero2014tsclust]. 
-The key idea is that observations within clusters tend to be more similar than 
-observations between clusters.
-Clustering is particularly useful for exploratory studies as it identifies
-structures within the data [@aghabozorgi2015time].
-
-Various applications of clustering can be found within geography: **ADD EXAMPLES**
-
-Key characteristic of these studies is the cross-sectional nature of the data they
-employed. 
-Indeed, most clustering problems in geography deal with observations 
-that are fixed in time. 
-However, for this paper we are interested in creating clusters of local authorities 
-in the UK with similar temporal signatures of experienced internet speeds over time. 
-Hence, we deviate from the established geographical clustering tools and 
-employ time series clustering methods.
-
-Time-series is a common type of dynamic data that naturally arises in many 
-different scenarios, such as stock data, medical data, and machine monitoring, 
-just to name a few **(Aghabozorgi et al., 2015; Aggarwal and Reddy, 2013)**. 
-They pose some challenging issues due to the large size and high dimensionality 
-commonly associated with time-series **(Aghabozorgi et al., 2015)**. 
-In this context, dimensionality of a series is related to time, 
-and it can be understood as the length of the series.
-
-Time series clustering methods have been developed in order to deal with clustering
-problems linked to, for instance, stock or other financial data, economic,
-governmental or medical data as well as machine monitoring
-[@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
-The main challenge -- and also the difference with cross-sectional clustering
-problems -- is data dimensionality given the multiplicity of data points for every
-individual included in the data set. 
-Time series are dynamic data as the value of the observations change as 
-a function of time [@aghabozorgi2015time].
-This high dimensionality leads to 
-(i) computational and algorithmic challenges regarding handling these data and 
-building algorithms to perform clustering over long time series, 
-and (ii) open questions regarding the choice of similarity measures in order to 
-cluster similar times series together considering the whole length of the time series 
-and overcoming issues around noise, outliers and shifts [@lin2004iterative; 
-@aghabozorgi2015time].
-
-Time series clustering methods utilising the whole length of time series can be
-grouped in three categories. 
-The first -- model-based approaches -- is based on recovering the underlying model 
-for each time series and then apply clustering algorithms on the model parameters 
-[@aghabozorgi2015time]. 
-The main criticism is the cluster accuracy for near clusters [@mitsa2010temporal]. 
-The second approach is a based the formation of vectors of features based on the 
-original time series.
-These new data of reduced dimensionality is then clustered using conventional
-clustering algorithms.
-
-For this paper we utilise the third category of time series clustering methods
-known as shape-based approaches. The main idea is to match any two sperate time
-series based on the similarity of their shapes through the calculation of
-distances among these two shapes. Such approached are better equipped to capture
-similarities between short length time series [@aghabozorgi2015time].
-This approach serves best this paper because (i) we aim to identify clusters of
-UK local authorities with similar temporal signatures -- i.e. shapes -- of
-experienced internet speeds and (ii) the length of our time series is short
-(**see also discussion in Section XX**).
-
-Other important dimensions regarding time series clustering are related with the
-clustering algorithm. Similar to clustering of cross-sectional data, we can employ
-partitioning algorithms, which lead to non-overlapping clusters, hierarchical
-clustering, which result to a hierarchy of clusters and fuzzy algorithms, which
-create overlapping clusters [@sardatime]. Because of the simplicity of the
-implementation and the interpretability of the results, we utilise here
-partitioning clustering based on the widely *k-Means* algorithm. This is an iterative
-algorithm, which begins with defining the desired number of clusters *k*. Then,
-each observation is randomly assigned to a cluster from the $[1,k]$ space. This
-is the initial cluster assignment, which is followed by iterations in order to
-minimise the distance between the centroids of the clusters and the observations
-assigned to these clusters [@james2013introduction].
-
-There are a number of differences between the above described application of
-*k-Means* for cross-sectional data and its application for times series data.
-
-k-Meidoids
-distance between points in time series
-dt
-
-comparison with density-based clustering DBSCAN [187]
-where a cluster is expanded if its neighbours are dense.
-
-- evaluation
-- standardised
-
-@sardatime
-In the context of shape-based time-series clustering, it is common to utilize the Dynamic Time
-Warping (DTW) distance as dissimilarity measure (Aghabozorgi et al., 2015). The calculation of the
-DTW distance involves a dynamic programming algorithm that tries to find the optimum warping
-path between two series under certain constraints
-
-**moved this section from data and descriptives as need to end methods of clustering with how it applies to what we did, but needs further edit after you're done**
-To apply these methods to our data, 
-we aggregate all the speed-checks during the 13 weeks of March 
-to May inclusive for weekdays in 2020 by each hour of the day and day of the week. 
-As our research aims to identify the geography of internet service resilience for 
-work purposes, bank holidays and the hours between midnight and 6am were also 
-excluded. 
-The composite week time series thus comprise 18 hours multiplied by 5 weekdays or 
-90 time points per series. 
-These time series were calculated for each of 
-the 382 Local Authority Districts (LADs) in the UK, standardised, 
-and then a k-means clustering algorithm was applied, including dynamic time warping. 
-The LADs were assigned to 10 clusters for upload speeds. 
-The cluster id for each LAD is then reattached to the speed-check dataframe to 
-identify the characteristics of each cluster, including number of LADs, and the 
-descriptives statistics of upload speeds in that cluster, 
-and the temporal profile by hour of the day and day of the week.
-Thus we aim to answer the first part of our research question: 
-How resilient are broadband speeds as experienced in different parts of the UK during 
-a time of extreme demand?
-
 <!-- HB old PARA4 - link to auxillary regressions method, or put in results to introduce results of auxillary regressions? -->
@@
 or similar quality of infrastructure.
-Our approach is also limited by potential endogeneity, as for example, better quality 
-connections with high mean speeds may enable more working from home, but greater 
-demand may cause slower speeds, less reliability and greater variability of speed at 
-different times of day or week. 
-Therefore, we avoid attributing any cause to our analysis of the experienced level of 
-quality and reliability of upload speeds. 
-Instead, we run auxilliary regressions in order to understand how the spatial and 
+Our approach is also limited by potential endogeneity, as for example, better
+quality connections with high mean speeds may enable more working from home, but
+greater demand may cause slower speeds, less reliability and greater variability
+of speed at different times of day or week.
+Therefore, we avoid attributing any cause to our analysis of the experienced level
+of quality and reliability of upload speeds.
+Instead, we run an auxiliary regression in order to understand how the spatial and
 temporal patterns of internet service relate to the economic geography of the UK.
-We discuss how the different patterns might support or undermine efforts to work from 
-home and maintain safe productivity and whether they reinforce existing spatial and 
-social inequalities. 
-From this analysis, we hope to provide a greater understanding of how telecommuting 
-and technology intersect at a time of extreme demand, and what lessons this time has 
-for a future where telecommuting is likely to remain a common means of accessing work 
-and broadband services, as well as infrastructure, must be fit for purpose.
+We discuss how the different patterns might support or undermine efforts to work
+from home and maintain safe productivity and whether they reinforce existing
+spatial and social inequalities.
+More specifically, we estimate the following multinomial model:
+
+\begin{align}
+Pr(Y_{i}=j) = \frac{exp^{X_{i}\beta_{j}}}{\sum_{i=1}^j exp^{X_{i}\beta_{j}}}
+\begin{cases}
+    i = 1, 2, ... , N \\t
+    j = 1, 2, ... , J
+\end{cases}\label{eq1}
+\end{align}
+
+The identification strategy is as follows.
+Based on the outcomes of the time-series clustering, we identify $J$ distinct
+and crisp clusters.
+We then regress this cluster membership against a vector $X_{i}$ of
+socio-economic and geographic variables, which we are discussed in details in
+**section XX**.
+This analysis enables us to provide a more nuanced understanding of how
+telecommuting and technology intersect at a time of extreme demand, and what
+lessons this time has for a future where telecommuting is likely to remain a
+common means of accessing work and broadband services, as well as infrastructure,
+must be fit for purpose.
 

### 2020-11-27 paper/v1/broadband.speed.covid.tex
after conflicts resolved, addition in intro, lit rev, methods and data.

--- 2020-11-16 paper/v1/broadband.speed.covid.tex
@@
 \begin{keywords}
-covid; internet; working from home; broadband speed; time series
+covid; internet; working from home; broadband speed; time-series
 clusters
@@
 
+During the pandemic, working from home using digital technologies,
+whether partially or exclusively, was transformed from a niche means of
+accessing work, albeit one that had been on a slow, upward trend, to a
+widespread way of life in many countries. The ability to work from home
+or telecommute meant millions retained their jobs and, to a varying
+extent, maintained productivity during periods of strict lockdown around
+the world. However, this ability has not been evenly distributed
+socially or spatially, creating a new type of digital divide. On one
+side are those who can work from home, supported by digital
+technologies, and have thus been able to enjoy both economic resilience
+and greater personal safety. On the other side, previously employed
+individuals have been forced to accept furlough or redundancy packages
+unless they are part of the cadre of essential workers, who are
+potentially at high risk of infection. Whilst the basis for this new
+digital divide has been viewed as mainly occupational, here we consider
+whether the divide is also technological.
+
+Using the UK as a case study, this paper aims to understand how the
+quality and reliability of internet service, as reflected in
+\emph{experienced} internet speeds, may reinforce or redress the spatial
+and social dimensions of the digital division exposed by the pandemic.
+To do so, we employ volunteered geographic data on individual broadband
+speed tests and state-of-the-art time-series clustering methods to
+create clusters of UK local authorities with similar temporal signatures
+of experienced internet speeds. We then associate these clusters of
+local authorities with their socioeconomic and geographic
+characteristics to explore how they overlap with or diverge from the
+existing economic and digital geography of the UK. \textbf{need to add
+urban-rural to analysis} Our analysis enables us to better understand
+how the spatial and social distribution of both occupations and online
+accessibility intersect to enable or hinder the practice of
+telecommuting at a time of extreme demand. We will also consider what
+lessons can be learned from this time for a future where telecommuting
+is likely to remain a common means of accessing work, and broadband
+services and infrastructure must be fit for purpose. \textbf{LET'S LEAVE
+IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ}
+
+The capability to work from home has previously been studied from the
+perspective of whether work tasks in a given occupation both can be and
+are allowed to be performed using digital technologies independently of
+location or co-location with colleagues, including supervisors
+\citep{allen2015effective, singh2013modeling}. However, successful
+telecommuting also requires that the quality and reliability of digital
+services, particularly home internet connection speeds, enable the
+completion of work tasks with a minimum of delay or interruption. Prior
+to the pandemic, the performance of broadband services with respect to
+telecommuters was never tested at scale, as working from home and
+connecting to colleagues and workplace resources via the internet was
+the purview of a small minority of workers. Instead, leisure use in the
+evening, when video streaming services are at their peak, has been used
+to benchmark broadband performance and service delivery by different
+Internet Service Providers (ISPs), at least in the UK \citep{ofcom2017}.
+Yet the shift towards telecommuting during various stages of lockdown
+around the world has been drastic and there are speculations that
+post-Covid, the tendency to work from home will be much higher than
+pre-Covid, raising questions around whether internet services can
+accommodate the increased demand. For example, \(47\)\% of people in
+employment in the UK worked soley from home in April \(2020\), whilst
+the same figure only reached \(5\)\% the year before
+\citep{ons2020, ons2020lm2019}. A back of the envelope calculation
+suggests that up to 40\% of the working force could work from home on an
+ongoing basis \citep{batty2020editorial}. Similar figures have been
+reported for other countries \citep{felstead2020homeworking}. For
+instance, \(37\)\% of the European workforce worked from home in April
+\(2020\) with countries like Finland reaching \(60\)\%
+\citep{eurofound2020}. In the US, almost half of the working population
+worked from home during the same period because of the pandemic
+\citep{brynjolfsson2020covid}, and a recent estimate indicated that
+\(37\)\% of all jobs in the US can be permanently performed entirely
+from home \citep{NBERw26948}.
+
+None of these changes could have happened in the absence of reliable
+information and communication technology (ICT) infrastructure -- both in
+terms of software and hardware. But while software innovations are
+easily diffused across space and society\footnote{See for example the
+  huge success of videoconferencing apps such as Zoom
+  \citep{marks2020zoom}.}, the same does not apply for ICT hardware
+infrastructure such as internet broadband connectivity. The literature
+describes first level digital divides in terms of the availability and
+quality of internet connectivity, such as that manifest in different
+geographies in the UK
+\citep{riddlesden2014broadband, philip2017digital}. Second level digital
+divides consider the presence or lack of the necessary skills to
+effectively utilise digital technologies and the internet
+\citep{blank2014dimensions, van2011internet}. The third level focuses on
+the heterogenous returns of internet usage among different socioeconomic
+groups and, consequently, how digital technologies can assist in
+bridging or further enhancing existing socioeconomic divides.
+\citep{stern2009levels, van2014digital, van2015third}. The capability to
+telecommute is related to all three levels of digital divides, but more
+importantly leads to differentiated outcomes regarding the economic
+resilience of people and places to overcome a systemic shock such as the
+current pandemic. We identify this as a new digital divide, one that
+fundamentally alters the potential returns of internet use for the user
+and wider community, assumes skills or functions that are present in
+some occupations but not in others, and relies upon access to high
+quality internet services. \textbf{WE NEED TO CONNECT OUR FINDINGS WITH
+THIS AND JUSTIFY THAT THIS IS INDEED A NEW DIVIDE} As the quality of
+internet infrastructure and services, as well as variation in
+occupations are spatially dependent and clustered in space, our approach
+offers a framework for understanding which types of places, are more
+likely to land on the right side of this new digital divide. By asking
+how resilient broadband speeds are as experienced in different parts of
+the UK during a time of extreme demand, we interrogate which places
+benefit from the greater economic resilience digital technologies can
+offer, not only during the pandemic, but also into the future.
+
+The structure of this paper is as follows. First we review the
+literature on telecommuting and digital divides to better understand the
+origins of the new digital divide revealed by the pandemic and its
+impact on the economic resilience of different places. We then describe
+our data and methodology. Our results section first offers
+classification of how internet services vary across the UK local
+authorities and then assesses whether these clusters replicate or
+repudiate other socio-economic and geographic patterns of economic
+resilience.
+
+\hypertarget{literature-review}{%
+\section{Literature review}\label{literature-review}}
+
+\hypertarget{from-telecommuting-to-wfh}{%
+\subsection{From telecommuting to
+\#WFH}\label{from-telecommuting-to-wfh}}
+
+In this analysis, the terms `telecommuting' and `working from home' are
+used interchangeably, as most remote labour during the Covid-19 crisis
+was carried out in the homes of individual employees rather than any
+other location \citep{eurofound2020}. However, it should be noted that
+previous research has explored how telecommuting can occur in other
+places, including satellite offices or on public transport
+\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
+also used a variety of definitions to measure the level of telecommuting
+within different workforces, distinguishing, for example, between those
+directly employed, indirectly employed, self-employed, full-time or
+part-time, and those who use digital technologies to work remotely
+full-days or part-days
+\citep{allen2015effective, bailey2002review, haddad2009examination}. No
+matter the definition, the option and capability to telecommute or work
+from home has never been equally distributed spatially or
+socio-economically any more than different industries and employment
+opportunities have. For example, studies from the United States, the
+Netherlands, and the UK indicate that telecommuters are most likely to
+hold professional, managerial, and technical occupations, where the
+workforce is better educated and wealthier, and that there is suppressed
+demand among women and part-time workers
+\citep{headicar2016move, peters2004employees, singh2013modeling}.
+Opportunities for working from home during the current pandemic have
+likewise not been equally spread across the workforce.
+\citet{NBERw26948} indicated that in the US, managers, educators, as
+well as those working in computer-related occupations, finance, and law
+can easily work from home, and that occupations with opportunities to
+telecommute are associated with higher earnings. This is not the case
+for the workforce occupied in more spatially fixed occupations, from
+farming, construction and manufacturing to hospitality and care
+services. In the US, these occupations tend to be lower-income,
+non-white, without a university degree, live in rental accommodation and
+lack health insurance \citep{NBERw27085}. Similar trends can be observed
+for other countries. For example, \(75\)\% of workers with tertiary
+education worked from home in Europe during spring \(2020\), whilst only
+\(34\)\% of workers with secondary education and \(14\)\% of those
+primary education did so \citep{eurofound2020}.
+
+\hypertarget{digital-divides-and-economic-resilience}{%
+\subsection{Digital divides and economic
+resilience}\label{digital-divides-and-economic-resilience}}
+
+Our understanding of telecommuting as a product of enabled occupations
+can be described as a manifestation of the third level digital divide,
+as those who are able to use digital technologies to work from home
+benefit from a high rate of return on their use of the internet in terms
+of autonomy, flexibility, and time saved from commuting
+\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
+These returns have been even greater during the Covid-19 crisis, when
+those with the ability to telecommute also have the ability to maintain
+their employment whilst protecting their health. However, the success of
+these arrangements has been dependent upon the first level digital
+divide, which is associated with access and quality of internet
+connectivity. For example, the systematic review from
+\citet{SALEMINK2017360} highlights the infrastructure quality
+differences between urban and rural areas in various advanced economies.
+Whether this variation in infrastructure quality affects the spatial
+footprint of telecommuting has not previously been measured, although
+there are indications that those who purchase high speed connections
+consume more data of all sorts and use their connections for a variety
+of purposes \citep{hauge2011consumer}. Another study identifies a
+correlation between access to internet services and a reduction in
+household transport spend across 33 countries \citep{bris2017ict}.
+Whether the implication is less travel because of increased
+telecommuting, or whether internet access enables more efficient travel,
+this finding is an indication of the potential household savings better
+internet services offer. Thus, the extreme demand during the pandemic
+provides a new opportunity to understand how infrastructure
+accessibility, quality, and reliability affects telecommuting,
+particularly as working from home during the pandemic required high
+volumes of bandwidth-intensive video conferencing in order to avoid the
+face-to-face contact that could increase the spread of infection. By
+first answering questions about internet service resilience, we can also
+refine our understanding of how this has contributed to or reduced the
+new digital divide, where economic resilience has been dependent upon
+the capability to work from home.
+
+The multi-layered digital divides intersect with materials divides and
+the economic geography of the UK. Following the regional economic
+resilience literature, which underlines the differentiated capacity of
+cities and regions to escape or recover from economic crises
+\citep{martin2012regional, kitsos2018economic}, different places have
+different industrial and occupational profiles, and these affect the
+aggregated potential capacity of places for telecommuting. Such profiles
+are associated with longstanding inequalities in the UK and their
+spatial representation as a North-South divide
+\citep{martin_north_south}. Various studies have illustrated severe
+inequalities between the north and the south part of the UK regarding,
+for example, skills and human capital, unemployment, productivity and
+prosperity \citep{lee2014grim, mccann2020perceptions, dorling2018peak}.
+Some scholars have even argued that the UK suffers some of the highest
+level of interregional inequalities in the global north
+\citep{gal2018reducing, mccann2016uk}. Not only all three levels of
+digital divides are, to a certain extend, associated with or shaped by
+the geography of the UK, but the intersection of the digital and
+material divides affects the capacity of places to overcome, at least
+partially, the economic effects of the Covid-19 pandemic. Importantly,
+this is the first time that digital technologies became an essential
+tool for economic resilience for such a great part of the population.
+
+\hypertarget{other-subsection}{%
+\subsection{other subsection?}\label{other-subsection}}
+
+\textbf{I THINK WE HAVE ENOUGH OF A LIT REV} - circumstances where work
+can be carried out more flexibly in space and time - how relate to
+digital divides - capability theory of mobility? - broadband tech stuff?
+broadband studies / divides / resilience
+
+Some new papers google recommended to me:
+
 \begin{itemize}
@@
 \item
-  \href{https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit\#heading=h.i5om1o8wpcd9}{our
-  google doc}
+  \url{https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf}
+\item
+  \url{https://link.springer.com/article/10.1007/s11116-020-10136-6}
+\item
+  \url{https://www.sciencedirect.com/science/article/pii/S0966692319311305}
+\item
+  check who cites the above and what they cite
 \end{itemize}
-
-During the pandemic, working from home using Information and
-Communication Technologies (ICT), whether partially or exclusively, was
-transformed from a niche means of accessing work, albeit one that had
-been on a slow, upward trend, to a widespread way of life in many
-countries \citep{felstead2020homeworking}. \textbf{HANNAH, DO WE REALLY
-NEED A REFERENCE IN THE PREVIOUS SENTENCE?} The ability to work from
-home or telecommute meant millions retained their jobs and, to a varying
-extent, maintained productivity during periods of strict lockdown around
-the world. However, this ability has not been evenly distributed
-socially or spatially, creating a new type of digital divide. On one
-side are those who can work from home and have been able to enjoy both
-economic resilience and greater personal safety. On the other side,
-previously employed individuals have been forced to accept furlough or
-redundancy packages unless they are part of the cadre of essential
-workers, who are potentially at high risk of infection. Using the UK as
-a case study, this paper aims to improve our understanding of the
-spatial and social dimensions of this new digital divide by assessing
-the resilience of broadband speeds in terms of both quality and
-reliability of service, and whether this reinforces or redresses prior
-digital divisions. To do so, we employ unique volunteered geographic
-data on individual broadband speed tests and state-of-the-art
-time-series clustering methods, which enable us to create clusters of UK
-local authorities with similar temporal signatures of experienced
-internet speeds. We then associate these clusters of local authorities
-with their socioeconomic and geographic characteristics to explore
-whether they are linked with existing divides and the economic geography
-of the UK. \textbf{HANNAH, HOW DOES THIS SOUND?} Our analysis enables us
-to better understand how telecommuting and technology intersect at a
-time of extreme demand, and what lessons this time has for a future
-where telecommuting is likely to remain a common means of accessing work
-and broadband services, as well as infrastructure, must be fit for
-purpose. \textbf{LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE
-MORE THE RQ}
-
-The capability to work from home has previously been studied from the
-perspective of whether work tasks in a given occupation both can be and
-are allowed to be performed independently of location or co-location
-with colleagues, including supervisors (\textbf{insert refs}). However,
-successful telecommuting also requires that the quality and reliability
-of ICT services, particularly home internet connection speeds, enable
-the completion of work tasks with a minimum of delay or interruption. In
-reality, the performance of broadband speeds with respect to
-telecommuters has never been tested before at scale, as working from
-home and connecting to colleagues and workplace resources via broadband
-has previously been the purview of a small minority of workers. This
-papers understands telecommuting as a function of the quality of ICT
-services, particularly home internet connections, and whether work tasks
-in an occupation can be performed independently of being in a particular
-location or co-locating with colleagues
-
-The shift towards telecommuting during various stages of lockdown around
-the world has been drastic and, importantly, speculations indicate that
-the post-Covid tendency to work from home will be much higher than the
-pre-Covid one. A back of the envelope calculation suggests that up to
-40\% of the working force could work from home in the UK
-\citep{batty2020editorial}. Observational data pointed an even higher
-share of people in employment in the UK who worked from home in April
-\(2020\) (\(47\)\%), while the same figure only reached \(5\)\% the year
-before \citep{ons2020, ons2020lm2019}. Similar figures have been
-reported for other countries around the world
-\citep{felstead2020homeworking}. For instance, \(37\)\% of the workforce
-worked from home in Europe in April \(2020\) with countries like Finland
-reaching \(60\)\% \citep{eurofound2020}. In the US almost half of the
-working population worked from home during the same period because of
-the pandemic \citep{brynjolfsson2020covid}. More broadly, a recent
-estimate indicated that \(37\)\% of all jobs in the US can be
-permanently performed entirely from home \citep{NBERw26948}.
-
-There is a consensus that opportunities for working from home especially
-during the current pandemic are not equally spread across the workforce.
-\citet{NBERw26948} indicated that in the US managers, educators, as well
-as those working in computer-related occupations, finance, and law can
-easily work from home. On the contrary, workers in farming,
-construction, and manufacturing do not have such opportunities. Not
-surprisingly, occupations with opportunities to telecommute are
-associated with higher earnings. This is not the case for the workforce
-occupied in less footloose occupations as they tend to be lower-income,
-non-white, without a university degree, live in rental accommodation and
-lack health insurance \citep{NBERw27085}. Although these figures refer
-to the US, similar trends can be observed for other countries. For
-example, \(75\)\% of workers with tertiary education worked from home in
-Europe during spring \(2020\), while the same share for workers with
-secondary and primary education dropped to \(34\)\% and \(14\)\%
-respectively. Moreover, employees living in cities, women and younger
-employees were have worked from home \citep{eurofound2020}.
-
-None of these changes could have happened in the absence of reliable ICT
-infrastructure -- both in terms of software and hardware. But while
-software innovations are easily diffused across space and
-society\footnote{See for example the huge success of videoconferencing
-  apps \citep{marks2020zoom}.}, the same does not apply for ICT hardware
-infrastructure such as internet broadband connectivity. The literature
-exemplifies digital divides in terms of internet access and its quality.
-For instance, \citet{riddlesden2014broadband} highlighted the broadband
-divides in the UK, while the systematic review from
-\citet{SALEMINK2017360} reinforced our understanding for the
-infrastructure quality differences between urban and rural areas.
-
-Our framework to understand telecommuting as a function of occupations
-and quality of ICT infrastructure is aligned with current debates on
-digital divides. While the, so-called, first level digital divides are
-associated with access and quality of internet connectivity, the second
-level ones are linked to the necessary skills to effectively utilise ICT
-and the internet \citep{blank2014dimensions, van2011internet}.
-Importantly, the capacity to telecommute, which to a certain extend is
-related to the first and second level digital divides, leads to
-differentiated outcomes regarding the economic resilience of people and
-places against the current pandemic. And to the extend that the quality
-of the internet infrastructure and occupation variation are spatially
-dependent and clustered in space, the spatial footprint of telecommuting
-is of great interest. In a way similar to the third level digital
-divide, which focuses on the differentiated returns of internet use
-\citep{stern2009levels, van2014digital, van2015third} places with high
-rates of telecommuting during the Covid pandemic illustrate higher
-economic resilience against the current pandemic. This is aligned with
-the regional economic resilience literature, which underlines the
-differentiated capacity of cities and regions to escape or recover from
-economic crises \citep{martin2012regional, kitsos2018economic}.
-
-The long-term effects of such drastic changes in telecommuting and
-attitudes towards working from home are difficult to predict.
-Nevertheless, they span through various aspects of economy and society:
-from changes to transportation planning due to altered commuting
-patters, to changes in land use and urban planning to accommodate people
-who work from home \citep{BUDNITZ2020102713}; and from productivity and
-innovation changes, to changes in agglomeration externalities and the
-attraction of large cities \citep{econobs} just to name a few. This
-paper is positioned to support endeavours in understanding the effects
-of increased telecommuting by exposing the spatial and social dimensions
-of telecommuting including the resilience of broadband speeds in terms
-of both quality and reliability of service, and whether this reinforces
-or redresses prior digital divisions. \textbf{HANNAH, HOW DOES THIS
-SOUND}
-
-The structure of this paper is as follows. \ldots{}
 
@@
 
-\hypertarget{literature-review}{%
-\section{Literature review}\label{literature-review}}
+\hypertarget{methods-and-data}{%
+\section{Methods and data}\label{methods-and-data}}
 
-\hypertarget{broadband-studies-divides-broadband-tech-stuff}{%
-\subsection{broadband studies, divides, broadband tech
-stuff}\label{broadband-studies-divides-broadband-tech-stuff}}
+The starting point of our methodological framework is cluster analysis,
+which can be defined within the modern machine learning framework as an
+unsupervised learning task, which involves partitioning unlabelled
+observations into homogeneous groups known as clusters
+\citep{montero2014tsclust}. The key idea is that observations within
+clusters tend to be more similar than observations between clusters.
+Clustering is particularly useful for exploratory studies as it
+identifies structures within the data \citep{aghabozorgi2015time}.
+Therefore, cluster analysis is a widely used family of techniques in
+geography \citep{gordon1977classification, everitt1974cluster}. For
+instance, clustering methods is the basis of \emph{geodemographics},
+which aims to create small area indicators or typologies of
+neighbourhoods based on various and some times diverse variables
+\citep{SINGLETON2009289, harris2005geodemographics}. Clustering
+techniques have also been employed to solve regionalisation problems
+\citep{niesterowicz2016}.
 
-\begin{itemize}
-\tightlist
-\item
-  PARA about Contention??
-\end{itemize}
+Common characteristic of these studies is the cross-sectional nature of
+the data they employ. Indeed, most clustering problems in geography deal
+with observations that are fixed in time. However, for this paper we are
+interested in creating clusters of local authorities in the UK with
+similar temporal signatures of experienced internet speeds over time.
+Hence, we deviate from the established geographical clustering tools and
+employ time-series clustering methods.
 
-The infrastructural demands of working from home on internet services
-have been minor compared to the demands of leisure users, such that the
-broadband performance offered by different Internet Service Providers
-(ISPs) has been benchmarked according to download speeds during the
-evening `primetime', when video streaming services are at their peak,
-rather than during the working day (\textbf{OfCom, 2017}). Yet the
-pandemic has fundamentally changed not only how many people work from
-home, but also their technical requirements when doing so. Whilst it may
-have been common in the past for telecommuters to complete solitary work
-tasks at home, but still visit their workplace for meetings
-(\textbf{ref?}), during the pandemic the replacement of face-to-face
-contact with bandwidth-intensive video conference calling was seen as
-essential to suppressing the spread of infection. This type of broadband
-use can particularly affect upload speeds \textbf{ref?}. This paper
-offers a framework to assess the quality and reliability of upload
-speeds by time of day and day of the week across local authorities in
-the UK during the time when the population were told to work from home
-if at all possible.
+Time-series clustering methods have been developed in order to deal with
+clustering problems linked to, for instance, stock or other financial
+data, economic, governmental or medical data as well as machine
+monitoring
+\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
+The main challenge -- and also the difference with cross-sectional
+clustering problems -- is data dimensionality given the multiplicity of
+data points for every individual object -- local authority in our case
+-- included in the data set. Time-series are dynamic data as the value
+of the observations change as a function of time
+\citep{aghabozorgi2015time}. This high dimensionality leads to (i)
+computational and algorithmic challenges regarding handling these data
+and building algorithms to perform clustering over long time-series, and
+(ii) open questions regarding the choice of similarity measures in order
+to cluster similar times series objects together considering the whole
+length of the time-series and overcoming issues around noise, outliers
+and shifts \citep{lin2004iterative, aghabozorgi2015time}.
 
-\hypertarget{from-telecommuting-to-wfh}{%
-\subsection{from telecommuting to
-\#WFH}\label{from-telecommuting-to-wfh}}
+Time-series clustering methods utilising the whole length of time-series
+can be grouped in three categories. The first -- model-based approaches
+-- is based on recovering the underlying model for each time-series and
+then applying clustering algorithms on the model parameters of each
+time-series \citep{aghabozorgi2015time}. The main criticism is the
+cluster accuracy for nearby clusters \citep{mitsa2010temporal}. The
+second approach is a based the formation of vectors of features based on
+the original time-series. These new data of reduced dimensionality is
+then clustered using conventional clustering algorithms.
 
-\begin{itemize}
-\tightlist
-\item
-  Introduce space-time geography
-\item
-  circumstances where work can be carried out more flexibly in space and
-  time
-\item
-  how relate to digital divides
-\item
-  capability theory of mobility? suppressed demand\ldots{}
-\end{itemize}
+For this paper we utilise the third category of time-series clustering
+methods known as shape-based approaches. These methods match two sperate
+time-series objects based on the similarity of their shapes through the
+calculation of distances among these two shapes and are better equipped
+to capture similarities between short length time-series
+\citep{aghabozorgi2015time}. This approach serves best this paper
+because (i) we aim to identify clusters of UK local authorities with
+similar temporal signatures -- i.e.~shapes -- of experienced internet
+speeds and (ii) the length of our time-series is short (see the data
+discussion in this section).
 
-Some new papers google recommended to me:
+Another import element of time-series clustering is the actual
+clustering algorithm. Similar to clustering of cross-sectional data, we
+can employ partitioning algorithms, which lead to non-overlapping
+clusters, hierarchical clustering, which result to a hierarchy of
+clusters and fuzzy algorithms, which create overlapping clusters
+\citep{sardatime}. Because of the simplicity of the implementation and
+the interpretability of the results, we utilise here partitioning
+clustering based on the widely used \emph{k-Means} algorithm. This is an
+iterative algorithm, which begins with defining the desired number of
+clusters \emph{k}. Then, each observation is randomly assigned to a
+cluster from the \([1,k]\) space. This is the initial cluster
+assignment, which is followed by iterations in order to minimise the
+distance between the centroids of the clusters and the observations
+assigned to these clusters \citep{james2013introduction}.
 
-\begin{itemize}
-\tightlist
-\item
-  \url{https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf}
-\item
-  \url{https://link.springer.com/article/10.1007/s11116-020-10136-6}
-\item
-  \url{https://www.sciencedirect.com/science/article/pii/S0966692319311305}
-\item
-  check who cites the above and what they cite
-\end{itemize}
+There are a number of differences between the above described
+application of \emph{k-Means} for cross-sectional data and its
+application for times series data. Instead of creating clusters around
+centroids, a common approach is to create clusters around
+\emph{medoids}, which are representative time-series objects with a
+minimal distance to all other cluster objects \citep{sardatime}. Also,
+instead of calculating the Euclidean distance between centroids and data
+points, more complex distance measures need to be employed in order to
+capture the similarity between a time-series object and a medoid. A
+common distance measure for shape-based time-series clustering in the
+Dynamic Time Warping (DTW). Using its underpinning dynamic programming
+algorithm, DTW compares two time-series objects to find the optimum
+warping path between them. DTW is widely used in order to overcome
+limitations linked to the use of Euclidean distance
+\citep{sardatime, berndt1994using, ratanamahatana2004everything}. The
+\texttt{R} package \texttt{dtwclust} has been used for the time-series
+clustering \citep{dtwclust}.
 
-In this analysis, the terms `teleworking' and `working from home' are
-used interchangeably. An appropriate definition of teleworking is `the
-remote provision of labour that would otherwise be carried out within
-company premises' (European Commission, 2020b). In practice, during the
-COVID crisis, most such work was carried out in the homes of individual
-employees rather than any other location. \citep{eurofound2020}
+To assess the quality and reliability of internet across local
+authorities in the UK during the time when the population were told to
+work from home if at all possible we utilise unique data regarding
+individual internet speed tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
+This a private company that allows internet users to check their own
+broadband upload and download speeds, and stores every speed-check with
+timestamp and geolocation information. These data have been used before
+to assess digital divides \citep{riddlesden2014broadband} and the impact
+of local loop unbundling regulatory processes
+\citep{nardotto2015unbundling}. These volunteered geographic data enable
+to assess the \emph{experienced} internet speed by users, which may
+differ from the \emph{advertised} maximum speeds provided by Internet
+Service Providers. We are particularly interested in upload speeds and
+the frequency of speed tests. While the former is less associated with
+internet-based leisure activities such as video streaming, which involve
+downloading large amounts of data, the latter can be linked more with
+work-related activities such as uploading documents or two-way
+communication and interaction systems. \textbf{ADD MORE. HOW ABOUT SPEED
+TESTS?}
 
-\hypertarget{covid-and-working-from-home-cities-urban-structure}{%
-\subsection{covid and working from home, cities, urban
-structure}\label{covid-and-working-from-home-cities-urban-structure}}
+The first step in the workflow after dropping some outliers following
+\citet{riddlesden2014broadband} is to transform and the individual,
+geolocated and time-stamped tests to more meaningful aggregates both in
+terms of space and time. Regarding the temporal dimension, we aggregate
+all the speed-checks during the \(13\) weeks of March to May inclusive
+for weekdays in \(2020\) by each hour of the day and day of the week. As
+our research aims to identify the geography of internet service
+resilience for work purposes, bank holidays and the hours between
+midnight and 6am were also excluded. The composite week time-series thus
+comprise 18 hours multiplied by 5 weekdays or 90 time points per series.
+We aggregate these data because, we could not follow individuals or
+households and connect data points. Also, although this is a large data
+set (\(241,088\) individual tests during the working hours of study
+period), there are not enough observations for each local and for each
+working hour (\(631\) speed test per LAD in average). These time-series
+were calculated for each of the 382 Local Authority Districts (LADs) in
+the UK, standardised, and then a \emph{k}-means partitioning around
+medoids clustering algorithm was applied using DTW. We initially run the
+algorithm for \(k = 5, 10, 15 and 20\) and used cluster validity indices
+(CVIs) to pick the optimal solution of \(k = 10\). Following
+\citet{sardatime} the following CVIs were employed: Silhouette (max),
+Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min),
+Modified Davies-Bouldin (DB*, min), Dunn (max), COP (min).
 
-\begin{itemize}
-\tightlist
-\item
-  economic geography of UK
-\end{itemize}
+We then match the cluster membership to the LAD-level data to identify
+the characteristics of each cluster, including number of LADs, the
+descriptive statistics of upload speeds in that cluster, and the
+temporal profile by hour of the day and day of the composite week. Thus
+we aim to answer the first part of our research question: How resilient
+are broadband speeds as experienced in different parts of the UK during
+a time of extreme demand?
 
-\hypertarget{data-and-descriptive-statistics}{%
-\section{Data and descriptive
-statistics}\label{data-and-descriptive-statistics}}
+Since the quality and reliability of internet services vary in time and
+space due to both supply and demand-side influences, we use a number of
+different measures of experienced upload speeds. These include: a) mean,
+experienced connection speed, b) standard deviation or the amount of
+fluctuation from the mean, and c) the variation in speeds at particular
+times of day when working from home is more likely to take place. We
+take account of all three measurements in order to describe upload
+speeds as fully as possible. The following \textbf{graphs / tables} show
+these statistics for the sample as a whole, as well as similar
+statistics related to frequency of testing.
 
-The experience of upload speeds is not the same as the maximum speeds
-offered by an ISP or possible speeds over a particular type and length
-of connection. Therefore, we use volunteered geographic information in
-the form of speed-checks run by users to test their experienced
-broadband speeds, upload and download. Meanwhile, the quality and
-reliability of upload speeds, like ICT more generally, vary in time and
-space due to both supply and demand-side influences, and can be measured
-in a number of ways. These include: a) mean, experienced connection
-speed, b) standard deviation or the amount of fluctuation from the mean,
-and c) the variation in speeds at particular times of day when working
-from home is more likely to take place. We take account of all three
-measurements in order to describe upload speeds as fully as possible. We
-start by calculating the mean upload speed for every local government
-district in the UK for each hour of each weekday, excluding midnight to
-6:00, weekends and bank holidays. We then cluster the local authorities
-by these temporal profiles, allowing us to identify patterns and
-describe the overall means, standard deviations, and other relevant
-statistics for each spatial cluster. Thus we aim to answer the first
-part of our research question: How resilient are broadband speeds as
-experienced in different parts of the UK during a time of extreme
-demand?
+\textbf{Data details and some figures, descriptive stats - include whole
+sample time profile for 2019 and 2020 frequency of tests run as part of
+why we chose to create the time profiles by hour of the day and day of
+the week rather than daily over the whole period.}
+
+\textbf{Just looking back, it was partly the frequency of tests, but
+that was also by geography, so do we need maps? And I feel there was
+something else I'm forgetting\ldots{}}
 
 The cause of these different experiences of broadband resilience may be
-different in different areas, as they may reflect similarities in
+different in different areas, as they may reflect either similarities in
 patterns of demand or similar quality of infrastructure. Our approach is
@@
 avoid attributing any cause to our analysis of the experienced level of
-quality and reliability of upload speeds. Instead, we run auxilliary
-regressions in order to understand how the spatial and temporal patterns
+quality and reliability of upload speeds. Instead, we run an auxiliary
+regression in order to understand how the spatial and temporal patterns
 of internet service relate to the economic geography of the UK. We
@@
 work from home and maintain safe productivity and whether they reinforce
-existing spatial and social inequalities. From this analysis, we hope to
-provide a greater understanding of how telecommuting and technology
-intersect at a time of extreme demand, and what lessons this time has
-for a future where telecommuting is likely to remain a common means of
-accessing work and broadband services, as well as infrastructure, must
-be fit for purpose.
+existing spatial and social inequalities. More specifically, we estimate
+the following multinomial model:
 
-The primary data analysed in this paper was provided by Speedchecker
-Ltd, a private company that allows internet users to check their own
-broadband upload / download speeds, and stores every speed-check with a
-timestamp and geolocation.
+\begin{align}
+Pr(Y_{i}=j) = \frac{exp^{X_{i}\beta_{j}}}{\sum_{i=1}^j exp^{X_{i}\beta_{j}}}
+\begin{cases}
+    i = 1, 2, ... , N \\t
+    j = 1, 2, ... , J
+\end{cases}\label{eq1}
+\end{align}
 
-Our approach involves aggregating all the speed-checks during the 13
-weeks of March to May inclusive for weekdays in 2020 by each hour of the
-day and day of the week. As our research aims to identify the geography
-of internet service resilience for work purposes, bank holidays and the
-hours between midnight and 6am were also excluded. The composite week
-time series thus comprise 18 hours multiplied by 5 weekdays or 90 time
-points per series. These time series were calculated for each of the 382
-Local Authority Districts (LADs) in the UK, standardised, and then a
-k-means clustering algorithm was applied, including dynamic time
-warping. The LADs were assigned to 10 clusters for upload speeds. The
-cluster id for each LAD is then reattached to the speed-check dataframe
-to identify the characteristics of each cluster, including number of
-LADs, and the descriptives statistics of upload speeds in that cluster,
-and the temporal profile by hour of the day and day of the week.
-
-\textbf{Data details and some figures, descriptive stats - include whole
-sample time profile for 2019 and 2020 frequency of tests run as part of
-why we chose to create the time profiles by hour of the day and day of
-the week rather than daily over the whole period.}
-
-\hypertarget{time-series-clustering}{%
-\section{Time series clustering}\label{time-series-clustering}}
-
-Description of the method
+The identification strategy is as follows. Based on the outcomes of the
+time-series clustering, we identify \(J\) distinct and crisp clusters.
+We then regress this cluster membership against a vector \(X_{i}\) of
+socio-economic and geographic variables, which we are discussed in
+details in \textbf{section XX}. This analysis enables us to provide a
+more nuanced understanding of how telecommuting and technology intersect
+at a time of extreme demand, and what lessons this time has for a future
+where telecommuting is likely to remain a common means of accessing work
+and broadband services, as well as infrastructure, must be fit for
+purpose.
 
@@
 of 382 local authorities, or over 62 million people, fall into cluster 6
-or cluster 9. Graph {[}.{]} shows that both of these clusters have
+or cluster 9. Graph {[}.{]} suggests that these two clusters have
 relatively similar temporal profiles, which are flatter than the other,
@@
 in most of the smaller clusters the reliability of service is worse than
-in cluster 9 in terms of the ratio of standard deviation to mean and
-worse than cluster 6 as well in terms of the ratio of upload speed in
-the morning peak compared to the evening peak. In five of the smaller
-clusters, which are home to almost 3 million people, speeds are 12-19\%
-slower between 9:00-10:59 compared to 19:00-20:59.
+in cluster 9 in terms of the ratio of standard deviation to mean. The
+smaller clusters are mainly worse than clusters 9 and 6 in terms of the
+ratio of upload speed in the morning peak compared to the evening peak.
+In five of the smaller clusters, which are home to almost 3 million
+people, speeds are 12-19\% slower between 9:00-10:59 compared to
+19:00-20:59.
 
-The quality and reliability of broadband service is thus much better in
-the 115 Local Authority areas in cluster 9, which are mostly in urban or
+The quality and reliability of broadband service is much better in the
+115 Local Authority areas in cluster 9, which are mostly in urban or
 suburban areas, compared to cluster 6. These include 13 London Boroughs
@@
 
+The long-term effects of such drastic changes in telecommuting and
+attitudes towards working from home are difficult to predict.
+Nevertheless, they span through various aspects of economy and society:
+from changes to transportation planning due to altered commuting
+patters, to changes in land use and urban planning to accommodate people
+who work from home \citep{BUDNITZ2020102713}\textbf{also 2020 Swedish
+article from JTG}; and from productivity and innovation changes, to
+changes in agglomeration externalities and the attraction of large
+cities \citep{econobs} just to name a few. This paper is positioned to
+support endeavours in understanding the effects of increased
+telecommuting by exposing the spatial and social dimensions of
+telecommuting including the resilience of broadband speeds in terms of
+both quality and reliability of service, and whether this reinforces or
+redresses prior digital divisions. \textbf{took this bit you wrote to
+put in the discussion at the end?}
+
 \hypertarget{acknowledgements}{%

### 2020-12-01 paper/v1/broadband.speed.covid.Rmd
Paper edits, esp tests descriptives in Methods section.

@@
 and geographic characteristics to explore how they overlap with or diverge from
-the existing economic and digital geography of the UK. **need to add urban-rural to analysis**
+the existing economic and digital geography of the UK. 
 Our analysis enables us to better understand how the spatial and social distribution
@@
 We then describe our data and methodology.
-Our results section first offers classification of how internet services vary
+Our results section first offers a classification of how internet services vary
 across the UK local authorities and then assesses whether these clusters replicate
@@
 different industries and employment opportunities have.
-For example, studies from the United States, the Netherlands, and the UK indicate
+For example, studies from the United States, the Netherlands, and the UK found
 that telecommuters are most likely to hold professional, managerial, and technical
-occupations, where the workforce is better educated and wealthier, and that there is
+occupations where the workforce is better educated and wealthier, and that there is
 suppressed demand among women and part-time workers
@@
 level digital divide, which is associated with access and quality of internet
-connectivity.
-For example, the systematic review from @SALEMINK2017360 highlights the
-infrastructure quality differences between urban and rural areas
+connectivity at a time of extremely high demand.
+@SALEMINK2017360 provides a systematic review of the pre-pandemic, 
+first level digital divide in infrastructure quality between urban and rural areas 
 in various advanced economies.
-Whether this variation in infrastructure quality affects the spatial footprint of
-telecommuting has not previously been measured, although there are indications that
-those who purchase high speed connections consume more data of all sorts and use their
-connections for a variety of purposes [@hauge2011consumer].
-Another study identifies a correlation between access to internet services and a
-reduction in household transport spend across 33 countries [@bris2017ict].
-Whether the implication is less travel because of increased telecommuting,
-or whether internet access enables more efficient travel, this finding is
-an indication of the potential household savings better internet services offer.
-Thus, the extreme demand during the pandemic provides a new opportunity to
-understand how infrastructure accessibility, quality, and reliability
-affects telecommuting, particularly as working from home during the pandemic required
-high volumes of bandwidth-intensive video conferencing in order to avoid the
-face-to-face contact that could increase the spread of infection.
-By first answering questions about internet service resilience, we can also refine our
-understanding of how this has contributed to or reduced the new digital divide,
-where economic resilience has been dependent upon the capability to work from home.
+Yet whether this variation in infrastructure quality affects the spatial footprint of
+telecommuting has not previously been measured.
+There are indications that those who purchase high speed connections consume 
+more data of all sorts and use their connections for a variety of purposes 
+[@hauge2011consumer], 
+and that there is a correlation between access to internet services and a
+reduction in household transport spend [@bris2017ict].
+Whether the implication is more internet use and less travel 
+because of increased telecommuting, these studies suggest that better internet 
+services enable households to make savings and efficiencies, 
+an example of the first level digital divide reinforcing the third level.
+Yet did the purchase of high speed connections and increased internet access also  
+prepare households for long-term home-working, enforced by government restrictions?
+The extreme demand during the pandemic provides a new opportunity to understand 
+how infrastructure accessibility, quality, and reliability affects telecommuting, 
+particularly in light of the high volumes of bandwidth-intensive video conferencing 
+required in order to avoid the face-to-face contact that could increase 
+the spread of infection. 
+We seek to answer how internet service resilience contributes to or reduces 
+economic resilience 
+when the latter is dependent upon the capability to work from home, 
+and thus improve our understanding of this new digital divide.
+**RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL DIVIDES?**
 
 <!-- Covid-19 and working from home, cities, urban structure -->
-The multi-layered digital divides intersect with materials divides and the economic
+The multi-layered digital divides intersect with material divides and the economic
 geography of the UK.
@@
 interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
-Not only all three levels of digital divides are, to a certain extend, associated
+Not only are all three levels of digital divides associated to a certain extent 
 with or shaped by the geography of the UK, but the intersection of the digital and
-material divides affects the capacity of places to overcome, at least partially,
-the economic effects of the Covid-19 pandemic.
+material divides affects the capacity of places to overcome
+some of the economic effects of the Covid-19 pandemic.
 Importantly, this is the first time that digital technologies became an essential
@@
 be defined within the modern machine learning framework as an unsupervised
-learning task, which involves partitioning unlabelled observations into homogeneous
-groups known as clusters [@montero2014tsclust].
+learning task, partitioning unlabelled observations into homogeneous groups 
+known as clusters [@montero2014tsclust].
 The key idea is that observations within clusters tend to be more similar than
@@
 [@gordon1977classification; @everitt1974cluster].
-For instance, clustering methods is the basis of *geodemographics*, which aims
-to create small area indicators or typologies of neighbourhoods based on various
-and some times diverse variables [@SINGLETON2009289; @harris2005geodemographics].
+For instance, clustering methods are the basis of *geodemographics*, a discipline 
+which aims to create small area indicators or typologies of neighbourhoods 
+based on various and sometimes diverse variables 
+[@SINGLETON2009289; @harris2005geodemographics].
 Clustering techniques have also been employed to solve regionalisation problems
@@
 <!-- time-series clustering -->
-Common characteristic of these studies is the cross-sectional nature of the data
+Common characteristics of these studies are the cross-sectional nature of the data
 they employ.
@@
 fixed in time.
-However, for this paper we are interested in creating clusters of local authorities
-in the UK with similar temporal signatures of experienced internet speeds over time.
+However, for this paper we are interested in internet speeds, which vary over time. 
+We aim to create clusters of local authorities in the UK with similar temporal 
+signatures of experienced internet speeds.
 Hence, we deviate from the established geographical clustering tools and
@@
 (i) computational and algorithmic challenges regarding handling these data and
-building algorithms to perform clustering over long time-series,
-and (ii) open questions regarding the choice of similarity measures in order to
-cluster similar times series objects together considering the whole length of the time-series
+building algorithms to perform clustering over long time-series, and 
+(ii) open questions regarding the choice of similarity measures in order to cluster 
+similar times series objects together considering the whole length of the time-series
 and overcoming issues around noise, outliers and shifts [@lin2004iterative;
@@
 The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].
-The second approach is a based the formation of vectors of features based on the
+The second approach is based the formation of vectors of features derived from the
 original time-series.
-These new data of reduced dimensionality is then clustered using conventional
+These new data of reduced dimensionality are then clustered using conventional
 clustering algorithms.
@@
 These methods match two sperate time-series objects based on the similarity of
-their shapes through the calculation of distances among these two shapes and are
-better equipped to capture similarities between short length
+their shapes through the calculation of distances between the shapes, 
+and are thus better equipped to capture similarities between short length
 time-series [@aghabozorgi2015time].
@@
 Another import element of time-series clustering is the actual clustering algorithm.
-Similar to clustering of cross-sectional data, we can employ partitioning
-algorithms, which lead to non-overlapping clusters, hierarchical
-clustering, which result to a hierarchy of clusters and fuzzy algorithms, which
+Similar to the clustering of cross-sectional data, we can employ partitioning
+algorithms, which lead to non-overlapping clusters, hierarchical clustering, 
+which classifies clusters at different levels, and fuzzy algorithms, which
 create overlapping clusters [@sardatime].
@@
 clusters *k*.
-Then, each observation is randomly assigned to a cluster from the $[1,k]$ space.
-This is the initial cluster assignment, which is followed by iterations in order
-to minimise the distance between the centroids of the clusters and the observations
+Then each observation is randomly assigned to a cluster from the $[1,k]$ space.
+This initial cluster assignment is followed by iterations in order to minimise 
+the distance between the centroids of the clusters and the observations
 assigned to these clusters [@james2013introduction].
@@
 the similarity between a time-series object and a medoid.
-A common distance measure for shape-based time-series clustering in the Dynamic
-Time Warping (DTW).
+A common distance measure for shape-based time-series clustering is 
+Dynamic Time Warping (DTW).
 Using its underpinning dynamic programming algorithm, DTW compares two time-series
@@
 in the UK during the time when the population were told to work from home if at
-all possible we utilise unique data regarding individual internet speed tests from
+all possible we utilise unique data comprising individual internet speed tests from
 Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
@@
 We are particularly interested in upload speeds and the frequency of speed tests.
-While the former is less associated with internet-based leisure activities such
-as video streaming, which involve downloading large amounts of data, the latter
+Upload speeds are less associated with internet-based leisure activities such
+as video streaming, which involve downloading large amounts of data, but they
 can be linked more with work-related activities such as uploading documents or
 two-way communication and interaction systems.
-**ADD MORE. HOW ABOUT SPEED TESTS?**
+Meanwhile, the frequency of speed tests was important in identifying 
+the temporal profile which woud give us most insight into 
+experienced internet service and resilience.
+Although we know from government statements the start and end of the period of
+interest, namely, March to May 2020, the frequency of speed tests within that period 
+provides an indication of the volume of experience over any particular units of time.
 
 <!-- data transformation -->
-The first step in the workflow after dropping some outliers following
-@riddlesden2014broadband is to transform and the individual,
-geolocated and time-stamped tests to more meaningful aggregates both in terms
-of space and time.
-Regarding the temporal dimension, we aggregate all the speed-checks during the $13$
-weeks of March to May inclusive for weekdays in $2020$ by each hour of the day and
-day of the week.
+Thus, the first step in the workflow after dropping some outliers following
+@riddlesden2014broadband was to transform the individual, geolocated and time-stamped 
+tests to more meaningful aggregates both in terms of space and time.
+The frequency of testing indicates that whilst there is an overall trend of increased
+testing or increased experience of internet service from March to April and then a 
+slight reduction from April to May, this trend masks substantial variation 
+by not only the day of the week, but also time of day, as can be seen in **[graph]**.
+Thus, a daily aggregation of upload speeds would also mask this pattern.
+Furthermore, the importance of this pattern is highlighted by a comparison 
+with the same period in 2019 **see graph**, 
+when the volume of testing and thus of experience of internet service quality
+was much more tied to the evening leisure peak and presumably to download speeds.
+Since the increase in testing during the working day in 2020 is an indication that 
+users have a greater perception of the variable quality of internet service,
+particularly during a new morning peak of testing for service reliability,
+we decided to include a measure of hourly variation in our temporal profiles.
+However, although this is a large data set ($241,088$ individual tests during the
+working hours of study period), there are not enough observations for each
+Local Authority District (LAD) and for each working hour of each working day
+($631$ speed tests per LAD on average) to profile speeds at that level of detail.
+Therefore, we aggregate all the speed-checks during the $13$ weeks of March to May 
+inclusive for weekdays in $2020$ by each hour of the day and day of the week.
 As our research aims to identify the geography of internet service resilience for
-work purposes, bank holidays and the hours between midnight and 6am were also
-excluded.
+work purposes, bank holidays and the hours between midnight and 6am were excluded,
+as well as weekend days.
 The composite week time-series thus comprise 18 hours multiplied by 5 weekdays or
 90 time points per series.
-We aggregate these data because, we could not follow individuals or households
-and connect data points.
-Also, although this is a large data set ($241,088$ individual tests during the
-working hours of study period), there are not enough observations for each local
-and for each working hour ($631$ speed test per LAD in average).
-These time-series were calculated for each of the 382 Local Authority Districts
-(LADs) in the UK, standardised, and then a *k*-means partitioning around medoids
-clustering algorithm was applied using DTW.
+We also aggregate these data spatially because we could not follow individuals or 
+households and connect data points.
+The time-series were calculated for each of the 382 LADs in the UK, standardised, 
+and then a *k*-means partitioning around medoids clustering algorithm 
+was applied using DTW.
 We initially run the algorithm for $k = 5, 10, 15 and 20$ and used cluster validity
@@
 
-We then match the cluster membership to the LAD-level data to identify the
-characteristics of each cluster, including number of LADs, the descriptive
-statistics of upload speeds in that cluster, and the temporal profile by hour of
-the day and day of the composite week.
-Thus we aim to answer the first part of our research question:
-How resilient are broadband speeds as experienced in different parts of the UK
-during a time of extreme demand?
-
 <!-- metrics -->
+We then review the characteristics of each cluster, including number of LADs, 
+and the temporal profile of upload speed by hour of the day 
+and day of the composite week.
 Since the quality and reliability of internet services vary in time and space due
 to both supply and demand-side influences, we use a number of different measures
-of experienced upload speeds.
+to then describe experienced upload speeds per cluster.
 These include: a) mean, experienced connection speed,
@@
 more likely to take place.
-We take account of all three measurements in order to describe upload speeds as
-fully as possible.
-The following **graphs / tables** show these statistics for the sample as a whole,
-as well as similar statistics related to frequency of testing.
+We take account of all three measurements in our descriptives statistics of
+upload speeds in order to determine how resilient broadband speeds are 
+as experienced in different parts of the UK during a time of extreme demand.
 
@@
 We then regress this cluster membership against a vector $X_{i}$ of
-socio-economic and geographic variables, which we are discussed in details in
+socio-economic and geographic variables, which are discussed in detail in
 **section XX**.

### 2020-12-02 paper/v1/broadband.speed.covid.Rmd
new clusters and aux based on k=[5:15]. Plus a minor change in the method section to address the above. No urban/rural and IMD were added. Let's talk today about it.

@@
 was applied using DTW.
-We initially run the algorithm for $k = 5, 10, 15 and 20$ and used cluster validity
-indices (CVIs) to pick the optimal solution of $k = 10$.
-Following @sardatime the following CVIs were employed: Silhouette (max), Score
-function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin (DB*, min), Dunn (max), COP (min).
+We initially run the algorithm for [5, 15] $k \in \mathbb{N} \bigcap [5,15]$  
+and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
+Following @sardatime the majority vote for the following CVIs was used: Silhouette 
+(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified 
+Davies-Bouldin (DB*, min), Dunn (max), COP (min).
 

### 2020-12-02 src/LAs_Clusters.Rmd
new clusters and aux based on k=[5:15]. Plus a minor change in the method section to address the above. No urban/rural and IMD were added. Let's talk today about it.

@@
 
+## Urban/rural
+
+```{r eval=F}
+# not complete
+url <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/137663/la-class-dataset-post0409.xls"
+GET(url, write_disk(tf <- tempfile(fileext = ".xls")))
+urban.rural <- read_excel(tf, 1L, range = "A17:CP1877", col_names = T)
+
+```
+
 ## merge
@@
 data$cluster.up <- as.factor(data$cluster.up)
-data$cluster.up <- relevel(data$cluster.up, "3") # base category: cluster 3
+data$cluster.up <- relevel(data$cluster.up, "4") # base category: cluster 4
 
@@
                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3)),
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3)
+                             , round(r2[1],3), round(r2[1],3)),
                            (c("N",
@@
                             n.obs, n.obs, n.obs,
-                            n.obs, n.obs, n.obs))))
+                            n.obs, n.obs, n.obs
+                            , n.obs, n.obs))))
 
@@
 data$cluster.tests <- as.factor(data$cluster.tests)
-data$cluster.tests <- relevel(data$cluster.tests, "5") # base category: cluster 5
+data$cluster.tests <- relevel(data$cluster.tests, "4") # base category
 
@@
                              round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3)),
+                             round(r2[1],3), round(r2[1],3), round(r2[1],3)),
                            (c("N",
                             n.obs, n.obs, n.obs,
-                            n.obs))))
+                            n.obs, n.obs, n.obs))))
 
@@
 data$cluster.down <- as.factor(data$cluster.down)
-data$cluster.down <- relevel(data$cluster.down, "2") # base category: cluster 2
+data$cluster.down <- relevel(data$cluster.down, "2") # base category
 

### 2020-12-02 src/ts_clusters.Rmd
new clusters and aux based on k=[5:15]. Plus a minor change in the method section to address the above. No urban/rural and IMD were added. Let's talk today about it.

@@
 ts.wide <- ts.wide %>%
-  select(help.date.time, LAD19NM, mean.up) #%>%
+  dplyr::select(help.date.time, LAD19NM, mean.up) #%>%
   #filter(date > "2020-2-15")
@@
 # 5 - 20 cluster sollutions
-pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
+pc_k <- tsclust(ts.wide, k = c(5:15), seed = 94L,
                 type="partitional",
@@
 # decide k, see RJ-2019-023 for max/min
-names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+names(pc_k) <- paste0("k_", c("5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15"))
+cvis <- sapply(pc_k, cvi, type = "internal")
+cvis
+# max Sil
+# max SF
+# max CH
+# min DB
+# min DBstar
+# max D
+# min COP
+
+# loop for k decision
+pref.cvi <- NULL
+
+max.cvis <- c(1:3,6)
+for (i in max.cvis){
+  a <- which.max(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+min.cvis <- c(4:5,7)
+for (i in min.cvis){
+  a <- which.min(cvis[i,])
+  pref.cvi <- c(pref.cvi, a)
+}
+
+print(pref.cvi) # cvis[2,] is always 0
+
+pc_k <- tsclust(ts.wide, k = 13, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+plot(pc_k)
+
+cluster_up <- data.frame(LAD=ts.wide.names,
+                           cluster=pc_k@cluster)
+```
+
+### Download speed
+
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+ts.wide <- merge(ts.down2020, help.dates, by = "weekday")
+ts.wide$hour <- paste(ts.wide$hour, ":00", sep="")
+ts.wide$help.date.time <- with(ts.wide, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+
+ts.wide <- ts.wide %>%
+  dplyr::select(help.date.time, LAD19NM, mean.down) #%>%
+  #filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide <- ts_wide(ts.wide)
+
+# impute with means of each TS
+# TODO check sensitivity
+ts.wide <- na_mean(ts.wide)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# extract names
+ts.wide.names <- names(ts.wide[,-1])
+ts.wide <- ts.wide[,-1]
+
+# standardise
+ts.wide <- zscore(ts.wide)
+
+# convert to list for tsclust
+ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
+
+# 5 - 20 cluster sollutions
+pc_k <- tsclust(ts.wide, k = c(5:15), seed = 94L,
+                type="partitional",
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+# decide k, see RJ-2019-023 for max/min
+names(pc_k) <- paste0("k_", c("5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15"))
 cvis <- sapply(pc_k, cvi, type = "internal")
@@
 
-cluster_up <- data.frame(LAD=ts.wide.names,
+cluster_down <- data.frame(LAD=ts.wide.names,
                          cluster=pc_k@cluster)
@@
 
-### Download speed
+### N. of speed tests
 
@@
 
-ts.wide <- merge(ts.down2020, help.dates, by = "weekday")
+# tests 
+ts.wide <- merge(ts.tests2020, help.dates, by = "weekday")
 ts.wide$hour <- paste(ts.wide$hour, ":00", sep="")
@@
 ts.wide <- ts.wide %>%
-  select(help.date.time, LAD19NM, mean.down) #%>%
+  dplyr::select(help.date.time, LAD19NM, n.tests) #%>%
   #filter(date > "2020-2-15")
@@
 # decide k, see RJ-2019-023 for max/min
-names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+names(pc_k) <- paste0("k_", c("5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15"))
 cvis <- sapply(pc_k, cvi, type = "internal")
@@
 
-pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
+pc_k <- tsclust(ts.wide, k = 7, seed = 94L, 
                 type="partitional", 
@@
 
-cluster_down <- data.frame(LAD=ts.wide.names,
+cluster_tests <- data.frame(LAD=ts.wide.names,
                            cluster=pc_k@cluster)
@@
 
-### N. of speed tests
+### Upload speed diff 2019-2020
 
@@
 
-# tests 
-ts.wide <- merge(ts.tests2020, help.dates, by = "weekday")
+# 2020 copied from above
+
+# create a 'fake' week for a ts id
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2020, month = 3, day = 2),
+                make_datetime(year = 2020, month = 3, day = 3),
+                make_datetime(year = 2020, month = 3, day = 4),
+                make_datetime(year = 2020, month = 3, day = 5),
+                make_datetime(year = 2020, month = 3, day = 6)),
+                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+ts.wide <- merge(ts.up2020, help.dates, by = "weekday")
 ts.wide$hour <- paste(ts.wide$hour, ":00", sep="")
@@
 ts.wide <- ts.wide %>%
-  select(help.date.time, LAD19NM, n.tests) #%>%
+  dplyr::select(help.date.time, LAD19NM, mean.up) #%>%
   #filter(date > "2020-2-15")
@@
 ts.wide <- na_mean(ts.wide)
+
+# 2019
+
+# create a 'fake' week for a ts id
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2019, month = 3, day = 4),
+                make_datetime(year = 2019, month = 3, day = 5),
+                make_datetime(year = 2019, month = 3, day = 6),
+                make_datetime(year = 2019, month = 3, day = 7),
+                make_datetime(year = 2019, month = 3, day = 8)),
+                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+ts.wide2019 <- merge(ts.up2019, help.dates, by = "weekday")
+ts.wide2019$hour <- paste(ts.wide2019$hour, ":00", sep="")
+ts.wide2019$help.date.time <- with(ts.wide2019, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+
+ts.wide2019 <- ts.wide2019 %>%
+  dplyr::select(help.date.time, LAD19NM, mean.up) #%>%
+  #filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide2019 <- ts_wide(ts.wide2019)
+
+# impute with means of each TS
+# TODO check sensitivity
+ts.wide2019 <- na_mean(ts.wide2019)
 #sapply(ts.wide, function(x) sum(is.na(x)))
+
+# diff 2020-2019
+ts.wide <- mapply('-', ts.wide, ts.wide2019, SIMPLIFY = FALSE)
+ts.wide <- as.data.frame(ts.wide)
 
@@
 
-pc_k <- tsclust(ts.wide, k = 5, seed = 94L, 
-                type="partitional", 
-                distance = "dtw_basic", centroid = "pam", trace = T, 
-                args = tsclust_args(dist = list(window.size = 20L)))
-plot(pc_k)
-
-cluster_tests <- data.frame(LAD=ts.wide.names,
-                           cluster=pc_k@cluster)
-```
-
-### Upload speed diff 2019-2020
-
-```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
-
-# 2020 copied from above
-
-# create a 'fake' week for a ts id
-help.dates <- tibble(help.date.time = c(make_datetime(year = 2020, month = 3, day = 2),
-                make_datetime(year = 2020, month = 3, day = 3),
-                make_datetime(year = 2020, month = 3, day = 4),
-                make_datetime(year = 2020, month = 3, day = 5),
-                make_datetime(year = 2020, month = 3, day = 6)),
-                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
-ts.wide <- merge(ts.up2020, help.dates, by = "weekday")
-ts.wide$hour <- paste(ts.wide$hour, ":00", sep="")
-ts.wide$help.date.time <- with(ts.wide, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
-
-ts.wide <- ts.wide %>%
-  select(help.date.time, LAD19NM, mean.up) #%>%
-  #filter(date > "2020-2-15")
-
-# turn to a wide format
-ts.wide <- ts_wide(ts.wide)
-
-# impute with means of each TS
-# TODO check sensitivity
-ts.wide <- na_mean(ts.wide)
-
-# 2019
-
-# create a 'fake' week for a ts id
-help.dates <- tibble(help.date.time = c(make_datetime(year = 2019, month = 3, day = 4),
-                make_datetime(year = 2019, month = 3, day = 5),
-                make_datetime(year = 2019, month = 3, day = 6),
-                make_datetime(year = 2019, month = 3, day = 7),
-                make_datetime(year = 2019, month = 3, day = 8)),
-                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
-ts.wide2019 <- merge(ts.up2019, help.dates, by = "weekday")
-ts.wide2019$hour <- paste(ts.wide2019$hour, ":00", sep="")
-ts.wide2019$help.date.time <- with(ts.wide2019, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
-
-ts.wide2019 <- ts.wide2019 %>%
-  select(help.date.time, LAD19NM, mean.up) #%>%
-  #filter(date > "2020-2-15")
-
-# turn to a wide format
-ts.wide2019 <- ts_wide(ts.wide2019)
-
-# impute with means of each TS
-# TODO check sensitivity
-ts.wide2019 <- na_mean(ts.wide2019)
-#sapply(ts.wide, function(x) sum(is.na(x)))
-
-# diff 2020-2019
-ts.wide <- mapply('-', ts.wide, ts.wide2019, SIMPLIFY = FALSE)
-ts.wide <- as.data.frame(ts.wide)
-
-# extract names
-ts.wide.names <- names(ts.wide[,-1])
-ts.wide <- ts.wide[,-1]
-
-# standardise
-ts.wide <- zscore(ts.wide)
-
-# convert to list for tsclust
-ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
-
-# 5 - 20 cluster sollutions
-pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
-                type="partitional",
-                distance = "dtw_basic", centroid = "pam", trace = T, 
-                args = tsclust_args(dist = list(window.size = 20L)))
-
-# decide k, see RJ-2019-023 for max/min
-names(pc_k) <- paste0("k_", c("5","10","15", "20"))
-cvis <- sapply(pc_k, cvi, type = "internal")
-cvis
-# max Sil
-# max SF
-# max CH
-# min DB
-# min DBstar
-# max D
-# min COP
-
-# loop for k decision
-pref.cvi <- NULL
-
-max.cvis <- c(1:3,6)
-for (i in max.cvis){
-  a <- which.max(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-min.cvis <- c(4:5,7)
-for (i in min.cvis){
-  a <- which.min(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-print(pref.cvi) # cvis[2,] is always 0
-
-pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
+pc_k <- tsclust(ts.wide, k = 6, seed = 94L, 
                 type="partitional", 
@@
 ts.wide <- ts.wide %>%
-  select(help.date.time, LAD19NM, mean.down) #%>%
+  dplyr::select(help.date.time, LAD19NM, mean.down) #%>%
   #filter(date > "2020-2-15")
@@
 ts.wide2019 <- ts.wide2019 %>%
-  select(help.date.time, LAD19NM, mean.down) #%>%
+  dplyr::select(help.date.time, LAD19NM, mean.down) #%>%
   #filter(date > "2020-2-15")
@@
 # 5 - 20 cluster sollutions
-pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
+pc_k <- tsclust(ts.wide, k = c(5:15), seed = 94L, #HERE
                 type="partitional",
@@
 
-pc_k <- tsclust(ts.wide, k = 10, seed = 94L, 
+pc_k <- tsclust(ts.wide, k = 15, seed = 94L, 
                 type="partitional", 

### 2020-12-02 src/LAs_Clusters.Rmd
regressions with nvq4

@@
   rename(NVQ4 = `% with NVQ4+ - aged 16-64`) %>%
-  rename(econ.act = `Economic activity rate - aged 16-64`)
+  rename(econ.act = `Economic activity rate - aged 16-64`) %>% # No NVQ data here as this is for Jul 2019-Jun 2020. NVQ is only  
+  dplyr::select(-NVQ3, -NVQ4)                                  # available for Dec - Dec. I download these data seperetely.
 
@@
 # sapply(test, function(x) sum(is.na(x)))
+```
+
+## Labour supply March 2020
+
+```{r}
+
+nvq4 <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_17_5.data.csv?geography=1811939329...1811939332,1811939334...1811939336,1811939338...1811939497,1811939499...1811939501,1811939503,1811939505...1811939507,1811939509...1811939517,1811939519,1811939520,1811939524...1811939570,1811939575...1811939599,1811939601...1811939628,1811939630...1811939634,1811939636...1811939647,1811939649,1811939655...1811939664,1811939667...1811939680,1811939682,1811939683,1811939685,1811939687...1811939704,1811939707,1811939708,1811939710,1811939712...1811939717,1811939719,1811939720,1811939722...1811939730,1807745025...1807745028,1807745030...1807745032,1807745034...1807745155,1807745157...1807745164,1807745166...1807745170,1807745172...1807745177,1807745179...1807745194,1807745196,1807745197,1807745199,1807745201...1807745218,1807745221,1807745222,1807745224,1807745226...1807745231,1807745233,1807745234,1807745236...1807745244&date=latestMINUS2&variable=290&measures=20599,21001,21002,21003")
+
+nvq4 <- nvq4 %>%
+  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, MEASURES_NAME, OBS_VALUE) %>%
+  filter(MEASURES_NAME=="Variable") %>%
+  dplyr::select(-MEASURES_NAME) %>%
+  distinct(GEOGRAPHY_CODE, OBS_VALUE, .keep_all = T) %>% 
+  rename(lad19cd = GEOGRAPHY_CODE) %>%
+  rename(nvq4 = OBS_VALUE)
 ```
@@
 ```{r}
-data <- list(lad, ladDist, job.dens, pop, pop.dens, labour, earnings, busi) # ns 
+data <- list(lad, ladDist, job.dens, pop, pop.dens, labour, earnings, busi, nvq4) # ns 
 sapply(data, function(x) dim(x))
@@
                   managers + tech + skilled + prof + admin + caring + plant +
-                  earnings + I(total.busi/pop) +
+                  earnings + I(total.busi/pop) + nvq4 +
                   I(LAD.tests.am20/pop) + # non sensitive to addition of other tests pm
@@
                              "n. business est. per hab., 2019",
+                             "% of NVQ4+",
                              "n. of AM broadband tests per hab., 2019",
@@
                   job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
-                  distMet + distLondon + south +
+                  distMet + distLondon + south + nvq4 +
                   managers + tech + skilled + prof + admin + caring + plant +

### 2020-12-03 src/descriptive_Clusters.Rmd

@@
 TS2019 <- selectByDate(TSbb19_20sp, year = 2019)
+
+#create time variation plots for speedtests for each year
+Tests19 <- count(TS2019, date)
+TS2019 <- left_join(TS2019, Tests19)
+names(TS2019)[13] <- "test frequency 2019"
+timeVariation(TS2019, pollutant = "test frequency 2019", statistic = "mean")
+Tests20 <- count(TS2020, date)
+TS2020 <- left_join(TS2020, Tests20)
+names(TS2020)[13] <- "test frequency 2020"
+timeVariation(TS2020, pollutant = "test frequency 2020", statistic = "mean")
 
@@
 #summarise upload speed clusters by hour of each weekday and plot
+#plot largest 5 clusters first - all over 1 million pop
 cluster.up <- TS2020 %>%
@@
 cluster.up <- left_join(cluster.up, LAD.cluster.up)
-cluster.up <- cluster.up[which(cluster.up$up.LADs > 5),]
+cluster.up <- cluster.up[which(cluster.up$up.LADs > 9),]
 cluster.up$clusterID <- as.factor(cluster.up$clusterID)
@@
 #plot cluster statistic for mean upload speed
-UpCluster <- ggplot(cluster.up, aes(help.date.time, mean.up, 
+UpClusterL <- ggplot(cluster.up, aes(help.date.time, mean.up, 
                                          colour = clusterID))
-UpCluster + geom_line() + labs(x = "Weekdays March-May") + 
+UpClusterL + geom_line() + labs(x = "Weekdays March-May", y = "mean upload speed") + 
+  scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
+                    "2020-03-03 15:00:00", "2020-03-04 15:00:00",
+                    "2020-03-05 15:00:00", "2020-03-06 15:00:00"), 
+                    labels = "%a"), labels = 
+                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + geom_point()
+#plot next 6 smaller clusters, but exclude 2 clusters with 1 LAD and under 100k pop
+cluster.upS <- TS2020 %>%
+  group_by(cluster.up, weekday, hour) %>%
+  summarise(mean.up = mean(speedup))
+names(cluster.upS)[1] <- "clusterID"
+names(LAD.cluster.up)[1] <- "clusterID"
+cluster.upS <- left_join(cluster.upS, LAD.cluster.up)
+cluster.upS <- cluster.upS[which(cluster.upS$up.LADs < 10 & cluster.upS$up.LADs > 1),]
+cluster.upS$clusterID <- as.factor(cluster.upS$clusterID)
+#add fake dates for plotting
+help.dates <- tibble(help.date.time = c(
+  make_datetime(year = 2020, month = 3, day = 2),
+  make_datetime(year = 2020, month = 3, day = 3),
+  make_datetime(year = 2020, month = 3, day = 4),
+  make_datetime(year = 2020, month = 3, day = 5),
+  make_datetime(year = 2020, month = 3, day = 6)),
+                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+cluster.upS <- merge(cluster.upS, help.dates, by = "weekday")
+cluster.upS$hour <- paste(cluster.upS$hour, ":00", sep="")
+cluster.upS$help.date.time <- with(cluster.upS, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+#plot cluster statistic for mean upload speed
+UpClusterS <- ggplot(cluster.upS, aes(help.date.time, mean.up, 
+                                         colour = clusterID))
+UpClusterS + geom_line() + labs(x = "Weekdays March-May", y = "mean upload speed") + 
   scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 

### 2020-12-04 paper/v1/broadband.speed.covid.Rmd
Updated paper - a little in data and methods section, mostly in first section of results.

@@
 and the impact of local loop unbundling regulatory processes [@nardotto2015unbundling].
-These volunteered geographic data enable to assess the *experienced* internet
-speed by users, which may differ from the *advertised* maximum speeds provided
-by Internet Service Providers.
-We are particularly interested in upload speeds and the frequency of speed tests.
-Upload speeds are less associated with internet-based leisure activities such
-as video streaming, which involve downloading large amounts of data, but they
-can be linked more with work-related activities such as uploading documents or
-two-way communication and interaction systems.
+By using volunteered geographic data, we are able to assess the *experienced* 
+internet speed by users, which may differ from the *advertised* maximum speeds of 
+Internet Service Providers (ISPs).
+We are particularly interested in upload speeds and the frequency of speed tests
+over the period from March to May 2020, as government statements indicate this is
+when UK workers were told to work from home if at all possible **[need ref?]**.
+Average upload speeds are slower than average download speeds, at $9.3$Mb/s 
+mean upload speed for the whole sample during the period of interest, 
+compared to $29.6$Mb/s for download speeds, but they are also less associated 
+with internet-based, high-demand, leisure activities such as video streaming.
+Therefore, upload speeds are more relevant to work-related activities such as 
+uploading documents or two-way audio, video, and text-based communication systems.
 Meanwhile, the frequency of speed tests was important in identifying 
 the temporal profile which woud give us most insight into 
-experienced internet service and resilience.
-Although we know from government statements the start and end of the period of
-interest, namely, March to May 2020, the frequency of speed tests within that period 
-provides an indication of the volume of experience over any particular units of time.
+experienced internet service and resilience, and provide 
+an indication of the volume of experience over particular units of time.
 
 <!-- data transformation -->
-Thus, the first step in the workflow after dropping some outliers following
+The first step in the workflow after dropping some outliers following
 @riddlesden2014broadband was to transform the individual, geolocated and time-stamped 
@@
 slight reduction from April to May, this trend masks substantial variation 
-by not only the day of the week, but also time of day, as can be seen in **[graph]**.
-Thus, a daily aggregation of upload speeds would also mask this pattern.
-Furthermore, the importance of this pattern is highlighted by a comparison 
-with the same period in 2019 **see graph**, 
+by not only the day of the week, but also time of day, as can be seen in Figure $1$**add graph of 2020 Time Variation**.
+Thus, a daily aggregation of upload speeds would mask the variation in experience
+over the course of each weekday.
+Furthermore, the importance of this variation is highlighted by a comparison 
+with the same period in $2019$, as in Figure $2$ **add graph of 2019 TimeVar**, 
 when the volume of testing and thus of experience of internet service quality
 was much more tied to the evening leisure peak and presumably to download speeds.
-Since the increase in testing during the working day in 2020 is an indication that 
+Since the increase in testing during the working day in $2020$ is an indication that 
 users have a greater perception of the variable quality of internet service,
@@
 As our research aims to identify the geography of internet service resilience for
-work purposes, bank holidays and the hours between midnight and 6am were excluded,
+work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
 as well as weekend days.
-The composite week time-series thus comprise 18 hours multiplied by 5 weekdays or
-90 time points per series.
+The composite week time-series thus comprise $18$ hours multiplied by $5$ weekdays or
+$90$ time points per series.
 We also aggregate these data spatially because we could not follow individuals or 
 households and connect data points.
-The time-series were calculated for each of the 382 LADs in the UK, standardised, 
+The time-series were calculated for each of the $382$ LADs in the UK, standardised, 
 and then a *k*-means partitioning around medoids clustering algorithm 
@@
 
-**Data details and some figures, descriptive stats - include whole sample time profile for 2019 and 2020 frequency of tests run as part of why we chose to create the time profiles by hour of the day and day of the week rather than daily over the whole period.**
+**More data details and some figures, descriptive stats**
 
-**Just looking back, it was partly the frequency of tests, but that was also by geography, so do we need maps? And I feel there was something else I'm forgetting...**
-
-<!-- HB old PARA4 - link to auxillary regressions method, or put in results to introduce results of auxillary regressions? -->
+<!-- link to auxillary regressions method -->
 The cause of these different experiences of broadband resilience may be different in
@@
 The temporal profiles used to cluster the local authorities have been summarised in
-graph [.], which
-shows a composite profile of mean upload speeds per hour per day for each cluster.
-For upload speeds, 345 of 382 local authorities, or over 62 million people,
-fall into cluster 6 or cluster 9.
-Graph [.] suggests that these two clusters have relatively similar temporal profiles,
-which are flatter than the other, smaller clusters,
-suggesting better reliability of service.
-However, the upload speeds at all times for cluster 9 are substantially higher
-than for cluster 6, which is an indication of better quality of service.
-This difference is reflected not only in the mean speeds for these clusters for the
-whole sample, but also the mean upload speeds in the morning peak from 9:00-10:59,
-as well as the evening peak period from 19:00-20:59.
-In comparison, the time profile in graph [.] shows upload speeds in cluster 1 are
-on average lower at certain times of day during the study period than any other
-cluster, whilst the profile for cluster 3 appears to show speeds fluctuating as much
-as cluster 1, but at levels usually higher even than cluster 9.
+Figure $3$ **add Graphs as 3a and 3b** and Table $1$ **Do we want all columns? Is this the first table?**.
+The graphs in Figure $3$ show a composite profile of mean upload speeds per hour 
+per day for each cluster, with the largest five clusters in Figure $3$a, 
+and the next six in Figure $3$b.
+Figure $3$ and Table $1$ provide a comprehensive overview of the quality and
+reliability of experienced broadband in different parts of the UK, 
+the temporal clusters offering a novel approach to understanding spatial disparity.
 
-The variability of the smaller clusters may be related to the fewer speed tests from
-fewer local authorities that have been averaged, whilst averaging greater numbers of
-speed tests could artificially flatten the profile.
-This appears not to be the case for cluster 9, as there is negligible difference
-between morning and evening upload speeds, at 1% slower in the morning, confirming a
-high level of reliability of service.
-In comparison, upload speeds in cluster 6 are 4% slower in the morning than in the
-evening and experienced the joint highest ratio of standard deviation to mean across
-the time period under assessment.
-This suggests that although the time profile is relatively flat in graph [.],
-the experience is one of speeds that fluctuate from a lower mean,
-and therefore might more often impact on online activities.
-Still, in most of the smaller clusters the reliability of service is worse than in
-cluster 9 in terms of the ratio of standard deviation to mean.
-The smaller clusters are mainly worse than clusters 9 and 6 in terms of
-the ratio of upload speed in the morning peak compared to the evening peak.
-In five of the smaller clusters, which are home to almost 3 million people,
-speeds are 12-19% slower between 9:00-10:59 compared to 19:00-20:59.
+The largest two clusters, comprising $281$ local authorities and almost $50$ million 
+people, fall into clusters $6$ and cluster $13$.
+Cluster $13$ has one of the lowest aggregate mean upload speed of any of the clusters,
+and the second highest ratio of the standard deviation to the mean.
+This suggests that those living in local authorities in this cluster experienced 
+some of the lowest quality broadband services in terms of upload speeds in the UK.
+However, the variation in upload speeds in cluster $13$, 
+which can be an indication of its reliability, does not seem to disproportionately 
+affect the morning peak from $9:00-10:59$, as upload speeds are, on average, 
+only $2.6$% slower than in the evening peak period between $19:00$ and $20:59$.
+Although cluster $13$ does include some rural, remote areas of the country, 
+such as Northwest Scotland, Cornwall and Powys in Wales as shown on Figure $4$,
+**insert map**it also includes the major metropolitan centres of Bristol, Liverpool, 
+and Newcastle, as well as nine (of $32$) London Boroughs and plenty of home county 
+and suburban areas.
+Meanwhile, those living in cluster $6$ experience aggregate mean upload speeds of 
+about $1$Mb/s faster than those in cluster $13$, but still lower than the other three
+large clusters and most of the smaller clusters, 
+suggesting a middling quality of service.
+However, the time profile for cluster $6$ in Figure $3$a shows that upload speeds 
+drop quickly from $6:00$ to $7:00$ on a Monday and peak between $23:00$ and midnight 
+on Wednesday and Thursday, but tend to be lower during the working day.
+Indeed, those in cluster $6$, including people living in Birmingham, Leeds and 
+Sheffield, in $12$ London Boroughs, and many suburban areas and smaller cities like
+Cardiff, Oxford and Cambridge experience mean upload speeds in the morning that are
+$4.4$% lower than in the evening.
+This lower level of reliability may be as a result of increased demand, 
+e.g. for telecommuting, especially as there are few truly remote areas 
+in this cluster.
 
-The quality and reliability of broadband service is much better in
-the 115 Local Authority areas in cluster 9,
-which are mostly in urban or suburban areas, compared to cluster 6.
-These include 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation.
-There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, and many suburban districts from the South East of England to South Tyneside.
-Meanwhile, the 230 local authorities in cluster 6, which have lower speeds on average and more variation in service still include major urban areas, such as Bristol, Liverpool and Leeds, and many suburban areas, but also include some of the most rural areas in the country.
-Meanwhile, Cluster 1, with 10 local authorities that are home to over 1 million people has the second slowest speeds in the morning compared to the evening 'peaks' and the second highest ratio of standard deviation to mean.
-This cluster's most populous area is Westminster in central London.
+However, whilst the difference in experienced mean upload speeds between $9:00-10:59$ 
+and $19:00-20:59$ in cluster $6$ is greater than in any of the other large clusters, 
+it is smaller than this difference in all the clusters shown in Figure $3$b.
+Whilst this may suggest a lack of reliability, especially considering the large
+spikes and dips in these clusters, the statistics in Table $1$ show 
+a more varied picture.
+The experienced upload speeds in cluster $1$ are the second slowest of any cluster,
+the ratio of standard deviation to mean is the highest, as is the ratio of speeds 
+in the morning compared to the evening peak hours analysed.
+The five, mainly rural authorities in this cluster clearly lack broadband resilience,
+for telecommuting or otherwise.
+Home to almost $3.5$ thousand people, the districts are scattered around the country:
+Ceredigion in West Wales, Darlington in the Northeast, Eden between the Lake District
+and the North Penines, Rossendale in the Northwest, and Rutland in the East Midlands.
+In comparison, clusters $3$ and $10$ have relatively high mean upload speeds, and 
+low standard deviations, suggesting reliability and resilience, 
+as well as quality broadband services, despite slower speeds in the morning peak.
+Like cluster $6$, is this also a sign of increased telecommuting?
+The eight LADs in clusters $3$ and $10$ are home to over one million people, a
+and are all more urban and suburban than cluster $1$, including the London Borough 
+of Newham, the Southeast commuter towns of Harlow and Luton, the Birmingham suburbs 
+of Bromsgrove and Cannock Chase, Dundee City, and the towns of Eastbourne and Corby.
 
-**further description of temporal profiles of other clusters - include all in graph?** Indeed, the only cluster where upload speeds were slower in the evening than in the morning was cluster 5, made up of 2 local authorities with less than 200,000 people: Three Rivers, a suburban district north of London, and Fylde, a seaside suburb of Blackpool.
-However, these are likely to be outliers and may not have many tests from which the clusters are calcualted.
-**Take out smaller clusters?** The exceptions can be found in seven of the eight much smaller clusters including 35 local authority districts, where AM peak upload speeds are between 6% and 18% slower than PM peak upload speeds, although the mean speeds for each cluster are higher than cluster 6.
-Indeed, in 25 local authorities with a combined population of almost 3 million, speeds are 13% or more slower in the morning than in the evening.
-Included in this latter group are central London borough of Westminster and the London Borough of Newham, rural authorities like Eden and West Devon, and small cities like Dundee and Carlisle.
+Meanwhile, the most resilient broadband service in terms of both quality and
+reliability appears to be found in clusters $9$ and $11$, home to over $12$ million people. 
+The mean speeds for these clusters are second and third highest, 
+the ratio of the standard deviation to the mean are the two lowest, and
+the experienced upload speeds in the morning peak are not that different 
+from those experienced in the evening peak.
+In Figure $3$a, cluster $11$ shows more noticeable peaks and troughs, 
+but the lowest points are not at the peak times described in Table $1$. 
+Rather, the slowest upload speeds on average occur between $6:00-7:00$ on 
+Monday morning, $14:00-15:00$ on Friday, and $16:00-17:00$ on Thursday. 
+These slowest times are still mostly faster than the average hourly upload speeds
+in cluster $13$.
+Although not all of the local authorities in these two clusters are urban, 
+among the $65$ LADs are Manchester and three of the nine other boroughs of 
+Greater Manchester, Glasgow, Nottingham, both Portsmouth and Southampton, 
+eight London Boroughs, four of the seven constituent authorities of the 
+West Midlands Combined Authority.
+There are also a number of other tightly bounded urban areas, such as Aberdeen, 
+Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at the centre
+of less confined districts like Burnley, Milton Keynes, and Northampton.
+The local authorities in these two clusters clearly experienced resilient
+broadband that could support high levels of telecommuting.
+They are not on the wrong side of the first level digital divide, 
+but how likely are they to be able to take advantage of their resilient ICT
+infrastructure and services?
 
@@
 
-Auxiliary regressions indicate that the speed tests in cluster 9 authorities are also more likely to have been run on services provided by Virgin Media, suggesting they are in the half of the country with the most lucrative ICT market, which originally attracted the cable TV provider (OfCom...).
-For example, although auxiliary regressions show that Cluster 6 local authorities are more likely to be in the South of the UK than Cluster 9, the cluster notably includes Southern rural districts from Cornwall to North Norfolk.
-Slower speeds could reflect the lower quality of service in rural areas compared to urban and suburban areas.
-There is frustration, however, as auxiliary regressions show that those living in cluster 6 had the highest probability of testing their broadband between 9:00 and 11:00 of any of the 10 upload speed clusters.
+Clusters $3$, $9$, $10$ and $11$ seem to benefit most from high quality and 
+resilient broadband services.
+The dips in mean upload speeds in clusters $6$, $3$ and $10$ during the morning
+peak are suggestive of more use during the working day, and potentially more
+telecommuting.
+Using auxiliary regressions, we test whether the clusters that have higher mean
+speeds and more reliable services are more urban, more wealthy, and / or more likely 
+to benefit from a choice of high quality internet services.
+We also measure which of our clusters are more likely to have a higher proportion of 
+occupations which enable telecommuting because of the nature of the work.
+The results of these auxiliary regressions are presented in Table $2$.
+The reference case is cluster $4$, which includes the local authority of Hambleton
+in North Yorkshire, a rural area of just over ninety thousand people.
+Mean, experienced upload speeds in cluster $4$ are close to both the average speeds
+for the $13$ clusters ($9.9$Mb/s) and the pre-clustered average for the whole sample
+($9.3$Mb/s).
+The standard deviation for cluster $4$ and the ratio of evening to morning peak speeds
+are indications of worse reliability than many of the other clusters, 
+but this is only one more reason to view the results in Table $2$ as relative
+rather than absolute probabilities.
 
-the finding in the auxiliary regressions that those in cluster 9 ran the fewest speed tests per person during the morning period of any cluster.
-Now this may be an indication of fewer people working from home, less contention, and less resultant frustration.
-Cluster 9 comprises many central urban areas and has the lowest number of established businesses per inhabitant, which could be interpreted as a dominance of large employers.
-However, the job density is lower than cluster 6, meaning there are not as many jobs per resident in these areas.
-As the cluster includes many suburban areas too, which may be largely residential, could the quality and reliability of internet service be reinforcing patterns of telecommuting by those in wealthier suburbs who can work from home?
-Earnings in cluster 9 are second highest of all the clusters, with only Cluster 2 (comprising just North East Lincolnshire and East Lothian, population 265k) earning more per person.
-
-The auxiliary regressions show that compared to the two authorities in cluster 5, all the other clusters had a lower percentage of working people in managerial, professional and administrative jobs.???
-Yet the auxiliary regression suggests that there are not many tests being run during the am peak in cluster 1.
-This may be because there are fewer people working at home checking their broadband than in most other clusters.
-Indeed, the auxiliary regressions indicate that cluster 1 has the highest job density or proportion of jobs to working-age population, which is likely to due to the presence of Westminster, central London, cluster 1's most populous local authority.
-Westminster not only has more workplaces than residents, but it is reasonable to presume that many who would normally work in Westminster, but be able to work from home during lockdown are likely to live outside central London and not be subject to the fluctuating speeds there.
-Workplaces, meanwhile, some of which would still have been open, could be running programmes that cause the slowdown and variation, but would be more likely to have their own in-house diagnostics, rather than using a service like Speedchecker Ltd.
+First, the number of speed tests run per cluster inhabitant between $9:00-10:59$
+is an indication of satisfaction or at least a lack of concern over broadband speeds
+and quality of service. 
+People in clusters $3$, $9$, $10$ and $11$ all ran fewer speed tests at this time of
+day than people in clusters $6$ and $13$.
+Those in clusters $3$, $9$, $10$ and $11$ also benefit from a higher proportion of
+fast Virgin cable connections than those in clusters $6$ and $13$. 
+Since services provided by Virgin Media are only available to $45$% of premises
+in the UK [@ofcom2016], 
+where the more lucrative ICT market originally attracted the cable TV provider, 
+this is an indication that people in these clusters are more likely to live on the 
+right side of the first level, infrastructure-based digital divide.
+Another indication is if people in these clusters are more likely to be closer to
+London or a metropolitan area.
+The broadband speed tests run in the authorities in cluster $3$ are more likely to 
+be taking place close to London than those run in any of the other clusters, 
+a result that makes sense considering two of the four authorities in cluster $3$ 
+are the London commuter towns of Harlow and Luton.
+However, even though London was also included in the variable calculating distance 
+to one of the ten major metropolitan areas in England, or Glasgow or Cardiff, 
+tests run in cluster $3$ are likely to be furthest away. 
+**Did you include London in distMet as well? - code suggests so**
+**Do we need to re-think this variable (e.g. remove Cardiff, add Leicester or...)**
 

### 2020-12-11 paper/v1/broadband.speed.covid.Rmd
updates in text, tables, figures

@@
   \usepackage[utf8]{inputenc}
+  \usepackage{dcolumn}
   \def\tightlist{}
@@
 library(hrbrthemes)
+library(stargazer)
+library(png)
 
@@
 
-# Introduction
+# Introduction {#sec:1}
 
@@
 to a varying extent, maintained productivity during periods of strict
-lockdown around the world.
+lockdowns around the world.
 However, this ability has not been evenly distributed socially or spatially,
@@
 We will also consider what lessons can be learned from this time for a future where
-telecommuting is likely to remain a common means of accessing work, and broadband
-services and infrastructure must be fit for purpose.
+telecommuting is likely to remain a more common means of accessing work, 
+at least in comparison to the pre-Covid era, 
+and broadband services and infrastructure must be fit for purpose.
 **LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**
@@
 
-# Literature review
+# Literature review {#sec:2}
 
-## From telecommuting to #WFH
+## From telecommuting to #WFH {#sec:2.1}
 
@@
 
-## Digital divides and economic resilience
+## Digital divides and economic resilience {#sec:2.2}
 <!-- WFH and digital divides-->
@@
 and thus improve our understanding of this new digital divide.
-**RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL DIVIDES?**
+**RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL DIVIDES?
+ET: OR REINFORCING THE 3RD LEVEL DIGITAL DIVIDE**
 
@@
 
-## other subsection?
-**I THINK WE HAVE ENOUGH OF A LIT REV**
-- circumstances where work can be carried out more flexibly in space and time
-- how relate to digital divides
-- capability theory of mobility?
-- broadband tech stuff? broadband studies / divides / resilience
+<!-- ## other subsection? -->
+<!-- **I THINK WE HAVE ENOUGH OF A LIT REV** -->
+<!-- - circumstances where work can be carried out more flexibly in space and time -->
+<!-- - how relate to digital divides -->
+<!-- - capability theory of mobility? -->
+<!-- - broadband tech stuff? broadband studies / divides / resilience -->
 
-Some new papers google recommended to me:
+<!-- Some new papers google recommended to me: -->
 
-- https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf
-- https://link.springer.com/article/10.1007/s11116-020-10136-6
-- https://www.sciencedirect.com/science/article/pii/S0966692319311305
-- check who cites the above and what they cite
+<!-- - https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf -->
+<!-- - https://link.springer.com/article/10.1007/s11116-020-10136-6 -->
+<!-- - https://www.sciencedirect.com/science/article/pii/S0966692319311305 -->
+<!-- - check who cites the above and what they cite -->
 
@@
 
-**MORE SOURCES:**
+<!-- **MORE SOURCES:** -->
 
-- https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas
-- [EPB commentaries](https://journals.sagepub.com/toc/EPB/current)
-- https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home
-- https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity
-- https://www.coronavirusandtheeconomy.com/question/how-will-economic-effects-coronavirus-vary-across-areas-uk
-- https://www.coronavirusandtheeconomy.com/which-parts-uk-have-been-hit-hardest-covid-19-crisis
-- https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas
+<!-- - https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas -->
+<!-- - [EPB commentaries](https://journals.sagepub.com/toc/EPB/current) -->
+<!-- - https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home -->
+<!-- - https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity -->
+<!-- - https://www.coronavirusandtheeconomy.com/question/how-will-economic-effects-coronavirus-vary-across-areas-uk -->
+<!-- - https://www.coronavirusandtheeconomy.com/which-parts-uk-have-been-hit-hardest-covid-19-crisis -->
+<!-- - https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas -->
 
-# Methods and data
+# Methods and data {#sec:3}
 
@@
 [@gordon1977classification; @everitt1974cluster].
-For instance, clustering methods are the basis of *geodemographics*, a discipline 
+For instance, clustering methods are the basis of *geodemographics*, a research domain 
 which aims to create small area indicators or typologies of neighbourhoods 
@@
 [@SINGLETON2009289; @harris2005geodemographics].
-Clustering techniques have also been employed to solve regionalisation problems
+Clustering techniques have also been employed to solve *regionalisation* problems
 [@niesterowicz2016].
@@
 fixed in time.
-However, for this paper we are interested in internet speeds, which vary over time. 
-We aim to create clusters of local authorities in the UK with similar temporal 
+However, for this paper we are interested in internet speeds, which vary over time 
+and, therefore, create clusters of local authorities in the UK with similar temporal 
 signatures of experienced internet speeds.
@@
 problems -- is data dimensionality given the multiplicity of data points for every
-individual object -- local authority in our case -- included in the data set.
+individual object -- local authorities in our case -- included in the data set.
 Time-series are dynamic data as the value of the observations change as
@@
 time-series [@aghabozorgi2015time].
-This approach serves best this paper because (i) we aim to identify clusters of
+This approach serves best this paper because (i) we identify clusters of
 UK local authorities with similar temporal signatures -- i.e. shapes -- of
@@
 The frequency of testing indicates that whilst there is an overall trend of increased
-testing or increased experience of internet service from March to April and then a 
+testing **or increased experience of internet service NOT SURE WHAT THIS MEAND** from March to April and then a 
 slight reduction from April to May, this trend masks substantial variation 
-by not only the day of the week, but also time of day, as can be seen in Figure $1$**add graph of 2020 Time Variation**.
+by not only the day of the week, but also time of day, as can be seen in Figure \ref{test2020}.
 Thus, a daily aggregation of upload speeds would mask the variation in experience
@@
 Furthermore, the importance of this variation is highlighted by a comparison 
-with the same period in $2019$, as in Figure $2$ **add graph of 2019 TimeVar**, 
+with the same period in $2019$, as in Figure \ref{test2019}, 
 when the volume of testing and thus of experience of internet service quality
@@
 As our research aims to identify the geography of internet service resilience for
-work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
+work purposes, bank holidays and the hours between midnight and 6:00 were excluded,
 as well as weekend days.
@@
 was applied using DTW.
-We initially run the algorithm for [5, 15] $k \in \mathbb{N} \bigcap [5,15]$  
+We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
 and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
@@
 Davies-Bouldin (DB*, min), Dunn (max), COP (min).
+
+![Speed tests over time, 2020 \label{test2020}](figures/time.var.plot2020.png){height=40% width=75%}
+
+![Speed tests over time, 2019 \label{test2019}](figures/time.var.plot2019.png){height=40% width=75%}
 
@@
 \begin{cases}
-    i = 1, 2, ... , N \\t
+    i = 1, 2, ... , N \\  
     j = 1, 2, ... , J
@@
 socio-economic and geographic variables, which are discussed in detail in
-**section XX**.
+Section \ref{#sec:4.2}.
 This analysis enables us to provide a more nuanced understanding of how
@@
 
-# Results
+# Results {#sec:4}
 
-## Upload Clusters / cluster description
+## Upload Clusters / cluster description {#sec:4.1}
 
-The temporal profiles used to cluster the local authorities have been summarised in
-Figure $3$ **add Graphs as 3a and 3b** and Table $1$ **Do we want all columns? Is this the first table?**.
-The graphs in Figure $3$ show a composite profile of mean upload speeds per hour 
-per day for each cluster, with the largest five clusters in Figure $3$a, 
-and the next six in Figure $3$b.
-Figure $3$ and Table $1$ provide a comprehensive overview of the quality and
+The temporal profiles of the local authority clusters have been summarised in
+Figures \ref{UpClusterL} and \ref{UpClusterS} and Table \ref{up.cluster.descr}.
+The graphs show a composite profile of mean upload speeds per hour 
+per day for each cluster, with the largest, in terms of the LAD membership, 
+five clusters in Figure \ref{UpClusterL}, 
+and the next six in Figure \ref{UpClusterS}.
+These figures and table provide a comprehensive overview of the quality and
 reliability of experienced broadband in different parts of the UK, 
@@
 
+```{r echo=FALSE, out.width = "95%",fig.cap="\\label{UpClusterL}Temporal profilies for upload speed large clusters"}
+img1_path <- "figures/upClusterL.png"
+include_graphics(img1_path)
+```
+
+```{r echo=FALSE, out.width = "95%",fig.cap="\\label{UpClusterS}Temporal profilies for upload speed small clusters"}
+img1_path <- "figures/upClusterS.png"
+include_graphics(img1_path)
+```
+
+``````{r echo=FALSE, message=FALSE, results = "asis"}
+path.clustersLADup <- paste(path, "./data/temp/LAD.cluster.up.csv", sep = "")
+
+clustersLADup <- read_csv(path.clustersLADup)
+
+# clustersLADup <- clustersLADup %>%
+#   dplyr::rename('N. of LADs' = up.LADs,
+#          'mean upload speed' = upMean,
+#          'SD upload speed' = upSD,
+#          'mean AM upload speed' = upAMmean)
+
+clustersLADup <- clustersLADup %>% 
+  mutate(across(4:7, round, 3))
+
+stargazer(clustersLADup[1:7],
+          summary = F,
+          font.size = "footnotesize",
+          type = "latex",
+          header = F, 
+          covariate.labels=c("Cluster",
+                             "N. of LADs",
+                             "LAD population",
+                             "mean speed",
+                             "SD speed",
+                             "mean AM speed",
+                             "mean PM speed"),
+                             #"AM_slower",
+                             #"SDtoMean"),
+          rownames = F,
+          no.space=TRUE,
+          column.sep.width = "0pt", 
+          notes = "Note: All speed measures are upload speeds",
+          title = "Upload speed cluster characteristics\\label{up.cluster.descr}")
+```
 The largest two clusters, comprising $281$ local authorities and almost $50$ million 
@@
 which can be an indication of its reliability, does not seem to disproportionately 
-affect the morning peak from $9:00-10:59$, as upload speeds are, on average, 
-only $2.6$% slower than in the evening peak period between $19:00$ and $20:59$.
+affect the morning peak from 9:00-10:59, as upload speeds are, on average, 
+only $2.6$% slower than in the evening peak period between 19:00 and 20:59.
 Although cluster $13$ does include some rural, remote areas of the country, 
-such as Northwest Scotland, Cornwall and Powys in Wales as shown on Figure $4$,
-**insert map**it also includes the major metropolitan centres of Bristol, Liverpool, 
+such as Northwest Scotland, Cornwall and Powys in Wales as shown on Figure \ref{map.up.clusters},
+it also includes the major metropolitan centres of Bristol, Liverpool, 
 and Newcastle, as well as nine (of $32$) London Boroughs and plenty of home county 
@@
 However, the time profile for cluster $6$ in Figure $3$a shows that upload speeds 
-drop quickly from $6:00$ to $7:00$ on a Monday and peak between $23:00$ and midnight 
+drop quickly from $6:00$ to 7:00 on a Monday and peak between 23:00 and midnight 
 on Wednesday and Thursday, but tend to be lower during the working day.
@@
 
-However, whilst the difference in experienced mean upload speeds between $9:00-10:59$ 
-and $19:00-20:59$ in cluster $6$ is greater than in any of the other large clusters, 
-it is smaller than this difference in all the clusters shown in Figure $3$b.
+```{r echo=FALSE, out.width = "95%",fig.cap="\\label{map.up.clusters}Upload speed clusters for LADs"}
+img1_path <- "figures/map.up.clusters.png"
+include_graphics(img1_path)
+```
+
+However, whilst the difference in experienced mean upload speeds between 9:00-10:59
+and 19:00-20:59 in cluster $6$ is greater than in any of the other large clusters, 
+it is smaller than this difference in all the clusters shown in Figure \ref{UpClusterL}.
 Whilst this may suggest a lack of reliability, especially considering the large
@@
 Like cluster $6$, is this also a sign of increased telecommuting?
-The eight LADs in clusters $3$ and $10$ are home to over one million people, a
+The eight LADs in clusters $3$ and $10$ are home to over one million people,
 and are all more urban and suburban than cluster $1$, including the London Borough 
@@
 from those experienced in the evening peak.
-In Figure $3$a, cluster $11$ shows more noticeable peaks and troughs, 
+In Figure \ref{UpClusterL}, cluster $11$ shows more noticeable peaks and troughs, 
 but the lowest points are not at the peak times described in Table $1$. 
-Rather, the slowest upload speeds on average occur between $6:00-7:00$ on 
-Monday morning, $14:00-15:00$ on Friday, and $16:00-17:00$ on Thursday. 
+Rather, the slowest upload speeds on average occur between 6:00-7:00 on 
+Monday morning, 14:00-15:00 on Friday, and $16:00-17:00$ on Thursday. 
 These slowest times are still mostly faster than the average hourly upload speeds
@@
 broadband that could support high levels of telecommuting.
-They are not on the wrong side of the first level digital divide, 
+They are *not* on the wrong side of the first level digital divide, 
 but how likely are they to be able to take advantage of their resilient ICT
@@
 
-## aux regressions
+## aux regressions {#sec:4.2}
+
+```{r echo=FALSE, message=FALSE, results = "asis"}
+path.LAs_Clusters <- paste0(path, "/data/temp/LAs_Clusters.RData")
+load(path.LAs_Clusters)
+
+stargazer(aux.up, 
+          type = "latex",
+          nobs = T,
+          digits = 3,
+          font.size = "tiny",
+          float.env = "sidewaystable",
+          dep.var.caption = "",
+          title = "Auxiliary multinomial regression of upload speed clusters on socio-economic and geographic LAD variables\\label{aux}",
+          header = F,
+          covariate.labels=c(#"pop. 16-64, 2018",
+                             "pop, 2018",
+                             #"pop. densuty, 2018",
+                             "job density, 2018",
+                             "distance to nearest met. area",
+                             "distance to London",
+                             "South of the UK",
+                             "managerial jobs, 2020",
+                             "tech jobs, 2020",
+                             "skilled trade jobs, 2020",
+                             "professional jobs, 2020",
+                             "administrative jobs, 2020",
+                             "leisure jobs, 2020",
+                             "machine operation jobs, 2020",
+                             "earnings, 2019",
+                             "n. business est. per hab., 2019",
+                             "NVQ4+",
+                             "AM tests per hab., 2020",
+                             "% of Virgin Media connections"),
+          add.lines = list(c("McFadden's R squared",
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
+                             , round(r2.up[1],3), round(r2.up[1],3)),
+                           (c("N",
+                            n.obs, n.obs, n.obs,
+                            n.obs, n.obs, n.obs,
+                            n.obs, n.obs, n.obs
+                            , n.obs, n.obs))))
+```
 
@@
 occupations which enable telecommuting because of the nature of the work.
-The results of these auxiliary regressions are presented in Table $2$.
+The results of these auxiliary regressions are presented in Table \ref{aux}.
+The dependent variable is the LAD cluster membership as described in Section \ref{#sec:3}
+and equation \ref{eq1} and each column represent a different cluster.
 The reference case is cluster $4$, which includes the local authority of Hambleton
@@
 The standard deviation for cluster $4$ and the ratio of evening to morning peak speeds
-are indications of worse reliability than many of the other clusters, 
-but this is only one more reason to view the results in Table $2$ as relative
+are indications of worse reliability than many of the other clusters.
+Hence, the results in Table \ref{aux} should be seens as relative
 rather than absolute probabilities.
 
-First, the number of speed tests run per cluster inhabitant between $9:00-10:59$
+First, the number of speed tests run per cluster inhabitant between 9:00-10:59
 is an indication of satisfaction or at least a lack of concern over broadband speeds
 and quality of service. 
-People in clusters $3$, $9$, $10$ and $11$ all ran fewer speed tests at this time of
-day than people in clusters $6$ and $13$.
+People in clusters $3$, $9$, $10$ and $11$ all ran fewer speed tests (at a per capita 
+basis) at this time of day than people in clusters $6$ and $13$.
 Those in clusters $3$, $9$, $10$ and $11$ also benefit from a higher proportion of
@@
 
-# Conclusions
+# Conclusions {#sec:5}
 

### 2020-12-11 paper/v1/broadband.speed.covid.tex
updates in text, tables, figures

@@
 \usepackage[utf8]{inputenc}
+\usepackage{dcolumn}
 \def\tightlist{}
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\hypertarget{sec:1}{%
+\section{Introduction}\label{sec:1}}
 
@@
 or telecommute meant millions retained their jobs and, to a varying
-extent, maintained productivity during periods of strict lockdown around
-the world. However, this ability has not been evenly distributed
+extent, maintained productivity during periods of strict lockdowns
+around the world. However, this ability has not been evenly distributed
 socially or spatially, creating a new type of digital divide. On one
@@
 characteristics to explore how they overlap with or diverge from the
-existing economic and digital geography of the UK. \textbf{need to add
-urban-rural to analysis} Our analysis enables us to better understand
-how the spatial and social distribution of both occupations and online
-accessibility intersect to enable or hinder the practice of
-telecommuting at a time of extreme demand. We will also consider what
-lessons can be learned from this time for a future where telecommuting
-is likely to remain a common means of accessing work, and broadband
-services and infrastructure must be fit for purpose. \textbf{LET'S LEAVE
-IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ}
+existing economic and digital geography of the UK. Our analysis enables
+us to better understand how the spatial and social distribution of both
+occupations and online accessibility intersect to enable or hinder the
+practice of telecommuting at a time of extreme demand. We will also
+consider what lessons can be learned from this time for a future where
+telecommuting is likely to remain a more common means of accessing work,
+at least in comparison to the pre-Covid era, and broadband services and
+infrastructure must be fit for purpose. \textbf{LET'S LEAVE IT FOR NOW,
+BUT I THINK WE CAN CRYSTALISE MORE THE RQ}
 
@@
 impact on the economic resilience of different places. We then describe
-our data and methodology. Our results section first offers
+our data and methodology. Our results section first offers a
 classification of how internet services vary across the UK local
@@
 
-\hypertarget{literature-review}{%
-\section{Literature review}\label{literature-review}}
+\hypertarget{sec:2}{%
+\section{Literature review}\label{sec:2}}
 
-\hypertarget{from-telecommuting-to-wfh}{%
-\subsection{From telecommuting to
-\#WFH}\label{from-telecommuting-to-wfh}}
+\hypertarget{sec:2.1}{%
+\subsection{From telecommuting to \#WFH}\label{sec:2.1}}
 
@@
 opportunities have. For example, studies from the United States, the
-Netherlands, and the UK indicate that telecommuters are most likely to
-hold professional, managerial, and technical occupations, where the
-workforce is better educated and wealthier, and that there is suppressed
-demand among women and part-time workers
+Netherlands, and the UK found that telecommuters are most likely to hold
+professional, managerial, and technical occupations where the workforce
+is better educated and wealthier, and that there is suppressed demand
+among women and part-time workers
 \citep{headicar2016move, peters2004employees, singh2013modeling}.
@@
 
-\hypertarget{digital-divides-and-economic-resilience}{%
-\subsection{Digital divides and economic
-resilience}\label{digital-divides-and-economic-resilience}}
+\hypertarget{sec:2.2}{%
+\subsection{Digital divides and economic resilience}\label{sec:2.2}}
 
@@
 divide, which is associated with access and quality of internet
-connectivity. For example, the systematic review from
-\citet{SALEMINK2017360} highlights the infrastructure quality
-differences between urban and rural areas in various advanced economies.
-Whether this variation in infrastructure quality affects the spatial
-footprint of telecommuting has not previously been measured, although
-there are indications that those who purchase high speed connections
-consume more data of all sorts and use their connections for a variety
-of purposes \citep{hauge2011consumer}. Another study identifies a
-correlation between access to internet services and a reduction in
-household transport spend across 33 countries \citep{bris2017ict}.
-Whether the implication is less travel because of increased
-telecommuting, or whether internet access enables more efficient travel,
-this finding is an indication of the potential household savings better
-internet services offer. Thus, the extreme demand during the pandemic
-provides a new opportunity to understand how infrastructure
-accessibility, quality, and reliability affects telecommuting,
-particularly as working from home during the pandemic required high
-volumes of bandwidth-intensive video conferencing in order to avoid the
-face-to-face contact that could increase the spread of infection. By
-first answering questions about internet service resilience, we can also
-refine our understanding of how this has contributed to or reduced the
-new digital divide, where economic resilience has been dependent upon
-the capability to work from home.
+connectivity at a time of extremely high demand. \citet{SALEMINK2017360}
+provides a systematic review of the pre-pandemic, first level digital
+divide in infrastructure quality between urban and rural areas in
+various advanced economies. Yet whether this variation in infrastructure
+quality affects the spatial footprint of telecommuting has not
+previously been measured. There are indications that those who purchase
+high speed connections consume more data of all sorts and use their
+connections for a variety of purposes \citep{hauge2011consumer}, and
+that there is a correlation between access to internet services and a
+reduction in household transport spend \citep{bris2017ict}. Whether the
+implication is more internet use and less travel because of increased
+telecommuting, these studies suggest that better internet services
+enable households to make savings and efficiencies, an example of the
+first level digital divide reinforcing the third level. Yet did the
+purchase of high speed connections and increased internet access also
+prepare households for long-term home-working, enforced by government
+restrictions? The extreme demand during the pandemic provides a new
+opportunity to understand how infrastructure accessibility, quality, and
+reliability affects telecommuting, particularly in light of the high
+volumes of bandwidth-intensive video conferencing required in order to
+avoid the face-to-face contact that could increase the spread of
+infection. We seek to answer how internet service resilience contributes
+to or reduces economic resilience when the latter is dependent upon the
+capability to work from home, and thus improve our understanding of this
+new digital divide. \textbf{RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE
+THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL
+QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL
+DIVIDES? ET: OR REINFORCING THE 3RD LEVEL DIGITAL DIVIDE}
 
-The multi-layered digital divides intersect with materials divides and
+The multi-layered digital divides intersect with material divides and
 the economic geography of the UK. Following the regional economic
@@
 level of interregional inequalities in the global north
-\citep{gal2018reducing, mccann2016uk}. Not only all three levels of
-digital divides are, to a certain extend, associated with or shaped by
-the geography of the UK, but the intersection of the digital and
-material divides affects the capacity of places to overcome, at least
-partially, the economic effects of the Covid-19 pandemic. Importantly,
-this is the first time that digital technologies became an essential
-tool for economic resilience for such a great part of the population.
+\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
+digital divides associated to a certain extent with or shaped by the
+geography of the UK, but the intersection of the digital and material
+divides affects the capacity of places to overcome some of the economic
+effects of the Covid-19 pandemic. Importantly, this is the first time
+that digital technologies became an essential tool for economic
+resilience for such a great part of the population.
 
-\hypertarget{other-subsection}{%
-\subsection{other subsection?}\label{other-subsection}}
-
-\textbf{I THINK WE HAVE ENOUGH OF A LIT REV} - circumstances where work
-can be carried out more flexibly in space and time - how relate to
-digital divides - capability theory of mobility? - broadband tech stuff?
-broadband studies / divides / resilience
-
-Some new papers google recommended to me:
-
-\begin{itemize}
-\tightlist
-\item
-  \url{https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf}
-\item
-  \url{https://link.springer.com/article/10.1007/s11116-020-10136-6}
-\item
-  \url{https://www.sciencedirect.com/science/article/pii/S0966692319311305}
-\item
-  check who cites the above and what they cite
-\end{itemize}
-
-\textbf{MORE SOURCES:}
-
-\begin{itemize}
-\tightlist
-\item
-  \url{https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas}
-\item
-  \href{https://journals.sagepub.com/toc/EPB/current}{EPB commentaries}
-\item
-  \url{https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home}
-\item
-  \url{https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity}
-\item
-  \url{https://www.coronavirusandtheeconomy.com/question/how-will-economic-effects-coronavirus-vary-across-areas-uk}
-\item
-  \url{https://www.coronavirusandtheeconomy.com/which-parts-uk-have-been-hit-hardest-covid-19-crisis}
-\item
-  \url{https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas}
-\end{itemize}
-
-\hypertarget{methods-and-data}{%
-\section{Methods and data}\label{methods-and-data}}
+\hypertarget{sec:3}{%
+\section{Methods and data}\label{sec:3}}
 
@@
 which can be defined within the modern machine learning framework as an
-unsupervised learning task, which involves partitioning unlabelled
-observations into homogeneous groups known as clusters
-\citep{montero2014tsclust}. The key idea is that observations within
-clusters tend to be more similar than observations between clusters.
-Clustering is particularly useful for exploratory studies as it
-identifies structures within the data \citep{aghabozorgi2015time}.
-Therefore, cluster analysis is a widely used family of techniques in
-geography \citep{gordon1977classification, everitt1974cluster}. For
-instance, clustering methods is the basis of \emph{geodemographics},
-which aims to create small area indicators or typologies of
-neighbourhoods based on various and some times diverse variables
+unsupervised learning task, partitioning unlabelled observations into
+homogeneous groups known as clusters \citep{montero2014tsclust}. The key
+idea is that observations within clusters tend to be more similar than
+observations between clusters. Clustering is particularly useful for
+exploratory studies as it identifies structures within the data
+\citep{aghabozorgi2015time}. Therefore, cluster analysis is a widely
+used family of techniques in geography
+\citep{gordon1977classification, everitt1974cluster}. For instance,
+clustering methods are the basis of \emph{geodemographics}, a research
+domain which aims to create small area indicators or typologies of
+neighbourhoods based on various and sometimes diverse variables
 \citep{SINGLETON2009289, harris2005geodemographics}. Clustering
-techniques have also been employed to solve regionalisation problems
-\citep{niesterowicz2016}.
+techniques have also been employed to solve \emph{regionalisation}
+problems \citep{niesterowicz2016}.
 
-Common characteristic of these studies is the cross-sectional nature of
-the data they employ. Indeed, most clustering problems in geography deal
-with observations that are fixed in time. However, for this paper we are
-interested in creating clusters of local authorities in the UK with
-similar temporal signatures of experienced internet speeds over time.
-Hence, we deviate from the established geographical clustering tools and
-employ time-series clustering methods.
+Common characteristics of these studies are the cross-sectional nature
+of the data they employ. Indeed, most clustering problems in geography
+deal with observations that are fixed in time. However, for this paper
+we are interested in internet speeds, which vary over time and,
+therefore, create clusters of local authorities in the UK with similar
+temporal signatures of experienced internet speeds. Hence, we deviate
+from the established geographical clustering tools and employ
+time-series clustering methods.
 
@@
 clustering problems -- is data dimensionality given the multiplicity of
-data points for every individual object -- local authority in our case
+data points for every individual object -- local authorities in our case
 -- included in the data set. Time-series are dynamic data as the value
@@
 cluster accuracy for nearby clusters \citep{mitsa2010temporal}. The
-second approach is a based the formation of vectors of features based on
-the original time-series. These new data of reduced dimensionality is
-then clustered using conventional clustering algorithms.
+second approach is based the formation of vectors of features derived
+from the original time-series. These new data of reduced dimensionality
+are then clustered using conventional clustering algorithms.
 
@@
 time-series objects based on the similarity of their shapes through the
-calculation of distances among these two shapes and are better equipped
-to capture similarities between short length time-series
+calculation of distances between the shapes, and are thus better
+equipped to capture similarities between short length time-series
 \citep{aghabozorgi2015time}. This approach serves best this paper
-because (i) we aim to identify clusters of UK local authorities with
-similar temporal signatures -- i.e.~shapes -- of experienced internet
-speeds and (ii) the length of our time-series is short (see the data
-discussion in this section).
+because (i) we identify clusters of UK local authorities with similar
+temporal signatures -- i.e.~shapes -- of experienced internet speeds and
+(ii) the length of our time-series is short (see the data discussion in
+this section).
 
 Another import element of time-series clustering is the actual
-clustering algorithm. Similar to clustering of cross-sectional data, we
-can employ partitioning algorithms, which lead to non-overlapping
-clusters, hierarchical clustering, which result to a hierarchy of
-clusters and fuzzy algorithms, which create overlapping clusters
-\citep{sardatime}. Because of the simplicity of the implementation and
-the interpretability of the results, we utilise here partitioning
-clustering based on the widely used \emph{k-Means} algorithm. This is an
-iterative algorithm, which begins with defining the desired number of
-clusters \emph{k}. Then, each observation is randomly assigned to a
-cluster from the \([1,k]\) space. This is the initial cluster
-assignment, which is followed by iterations in order to minimise the
+clustering algorithm. Similar to the clustering of cross-sectional data,
+we can employ partitioning algorithms, which lead to non-overlapping
+clusters, hierarchical clustering, which classifies clusters at
+different levels, and fuzzy algorithms, which create overlapping
+clusters \citep{sardatime}. Because of the simplicity of the
+implementation and the interpretability of the results, we utilise here
+partitioning clustering based on the widely used \emph{k-Means}
+algorithm. This is an iterative algorithm, which begins with defining
+the desired number of clusters \emph{k}. Then each observation is
+randomly assigned to a cluster from the \([1,k]\) space. This initial
+cluster assignment is followed by iterations in order to minimise the
 distance between the centroids of the clusters and the observations
@@
 capture the similarity between a time-series object and a medoid. A
-common distance measure for shape-based time-series clustering in the
+common distance measure for shape-based time-series clustering is
 Dynamic Time Warping (DTW). Using its underpinning dynamic programming
@@
 authorities in the UK during the time when the population were told to
-work from home if at all possible we utilise unique data regarding
+work from home if at all possible we utilise unique data comprising
 individual internet speed tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
@@
 of local loop unbundling regulatory processes
-\citep{nardotto2015unbundling}. These volunteered geographic data enable
-to assess the \emph{experienced} internet speed by users, which may
-differ from the \emph{advertised} maximum speeds provided by Internet
-Service Providers. We are particularly interested in upload speeds and
-the frequency of speed tests. While the former is less associated with
-internet-based leisure activities such as video streaming, which involve
-downloading large amounts of data, the latter can be linked more with
-work-related activities such as uploading documents or two-way
-communication and interaction systems. \textbf{ADD MORE. HOW ABOUT SPEED
-TESTS?}
+\citep{nardotto2015unbundling}. By using volunteered geographic data, we
+are able to assess the \emph{experienced} internet speed by users, which
+may differ from the \emph{advertised} maximum speeds of Internet Service
+Providers (ISPs). We are particularly interested in upload speeds and
+the frequency of speed tests over the period from March to May 2020, as
+government statements indicate this is when UK workers were told to work
+from home if at all possible \textbf{{[}need ref?{]}}. Average upload
+speeds are slower than average download speeds, at \(9.3\)Mb/s mean
+upload speed for the whole sample during the period of interest,
+compared to \(29.6\)Mb/s for download speeds, but they are also less
+associated with internet-based, high-demand, leisure activities such as
+video streaming. Therefore, upload speeds are more relevant to
+work-related activities such as uploading documents or two-way audio,
+video, and text-based communication systems. Meanwhile, the frequency of
+speed tests was important in identifying the temporal profile which woud
+give us most insight into experienced internet service and resilience,
+and provide an indication of the volume of experience over particular
+units of time.
 
 The first step in the workflow after dropping some outliers following
-\citet{riddlesden2014broadband} is to transform and the individual,
+\citet{riddlesden2014broadband} was to transform the individual,
 geolocated and time-stamped tests to more meaningful aggregates both in
-terms of space and time. Regarding the temporal dimension, we aggregate
-all the speed-checks during the \(13\) weeks of March to May inclusive
-for weekdays in \(2020\) by each hour of the day and day of the week. As
-our research aims to identify the geography of internet service
-resilience for work purposes, bank holidays and the hours between
-midnight and 6am were also excluded. The composite week time-series thus
-comprise 18 hours multiplied by 5 weekdays or 90 time points per series.
-We aggregate these data because, we could not follow individuals or
-households and connect data points. Also, although this is a large data
-set (\(241,088\) individual tests during the working hours of study
-period), there are not enough observations for each local and for each
-working hour (\(631\) speed test per LAD in average). These time-series
-were calculated for each of the 382 Local Authority Districts (LADs) in
-the UK, standardised, and then a \emph{k}-means partitioning around
-medoids clustering algorithm was applied using DTW. We initially run the
-algorithm for \(k = 5, 10, 15 and 20\) and used cluster validity indices
-(CVIs) to pick the optimal solution of \(k = 10\). Following
-\citet{sardatime} the following CVIs were employed: Silhouette (max),
-Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min),
-Modified Davies-Bouldin (DB*, min), Dunn (max), COP (min).
+terms of space and time. The frequency of testing indicates that whilst
+there is an overall trend of increased testing \textbf{or increased
+experience of internet service NOT SURE WHAT THIS MEAND} from March to
+April and then a slight reduction from April to May, this trend masks
+substantial variation by not only the day of the week, but also time of
+day, as can be seen in Figure \ref{test2020}. Thus, a daily aggregation
+of upload speeds would mask the variation in experience over the course
+of each weekday. Furthermore, the importance of this variation is
+highlighted by a comparison with the same period in \(2019\), as in
+Figure \ref{test2019}, when the volume of testing and thus of experience
+of internet service quality was much more tied to the evening leisure
+peak and presumably to download speeds. Since the increase in testing
+during the working day in \(2020\) is an indication that users have a
+greater perception of the variable quality of internet service,
+particularly during a new morning peak of testing for service
+reliability, we decided to include a measure of hourly variation in our
+temporal profiles. However, although this is a large data set
+(\(241,088\) individual tests during the working hours of study period),
+there are not enough observations for each Local Authority District
+(LAD) and for each working hour of each working day (\(631\) speed tests
+per LAD on average) to profile speeds at that level of detail.
+Therefore, we aggregate all the speed-checks during the \(13\) weeks of
+March to May inclusive for weekdays in \(2020\) by each hour of the day
+and day of the week. As our research aims to identify the geography of
+internet service resilience for work purposes, bank holidays and the
+hours between midnight and 6:00 were excluded, as well as weekend days.
+The composite week time-series thus comprise \(18\) hours multiplied by
+\(5\) weekdays or \(90\) time points per series. We also aggregate these
+data spatially because we could not follow individuals or households and
+connect data points. The time-series were calculated for each of the
+\(382\) LADs in the UK, standardised, and then a \emph{k}-means
+partitioning around medoids clustering algorithm was applied using DTW.
+We initially run the algorithm for \(k \in \mathbb{N} \bigcap [5,15]\)
+and used cluster validity indices (CVIs) to pick the optimal solution of
+\(k = 13\). Following \citet{sardatime} the majority vote for the
+following CVIs was used: Silhouette (max), Score function (max),
+Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin
+(DB*, min), Dunn (max), COP (min).
 
-We then match the cluster membership to the LAD-level data to identify
-the characteristics of each cluster, including number of LADs, the
-descriptive statistics of upload speeds in that cluster, and the
-temporal profile by hour of the day and day of the composite week. Thus
-we aim to answer the first part of our research question: How resilient
-are broadband speeds as experienced in different parts of the UK during
-a time of extreme demand?
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,height=0.4\textheight]{figures/time.var.plot2020.png}
+\caption{Speed tests over time, 2020 \label{test2020}}
+\end{figure}
 
-Since the quality and reliability of internet services vary in time and
-space due to both supply and demand-side influences, we use a number of
-different measures of experienced upload speeds. These include: a) mean,
+\begin{figure}
+\centering
+\includegraphics[width=0.75\textwidth,height=0.4\textheight]{figures/time.var.plot2019.png}
+\caption{Speed tests over time, 2019 \label{test2019}}
+\end{figure}
+
+We then review the characteristics of each cluster, including number of
+LADs, and the temporal profile of upload speed by hour of the day and
+day of the composite week. Since the quality and reliability of internet
+services vary in time and space due to both supply and demand-side
+influences, we use a number of different measures to then describe
+experienced upload speeds per cluster. These include: a) mean,
 experienced connection speed, b) standard deviation or the amount of
@@
 times of day when working from home is more likely to take place. We
-take account of all three measurements in order to describe upload
-speeds as fully as possible. The following \textbf{graphs / tables} show
-these statistics for the sample as a whole, as well as similar
-statistics related to frequency of testing.
+take account of all three measurements in our descriptives statistics of
+upload speeds in order to determine how resilient broadband speeds are
+as experienced in different parts of the UK during a time of extreme
+demand.
 
-\textbf{Data details and some figures, descriptive stats - include whole
-sample time profile for 2019 and 2020 frequency of tests run as part of
-why we chose to create the time profiles by hour of the day and day of
-the week rather than daily over the whole period.}
-
-\textbf{Just looking back, it was partly the frequency of tests, but
-that was also by geography, so do we need maps? And I feel there was
-something else I'm forgetting\ldots{}}
+\textbf{More data details and some figures, descriptive stats}
 
@@
 We then regress this cluster membership against a vector \(X_{i}\) of
-socio-economic and geographic variables, which we are discussed in
-details in \textbf{section XX}. This analysis enables us to provide a
-more nuanced understanding of how telecommuting and technology intersect
-at a time of extreme demand, and what lessons this time has for a future
+socio-economic and geographic variables, which are discussed in detail
+in Section \ref{#sec:4.2}. This analysis enables us to provide a more
+nuanced understanding of how telecommuting and technology intersect at a
+time of extreme demand, and what lessons this time has for a future
 where telecommuting is likely to remain a common means of accessing work
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
+\hypertarget{sec:4}{%
+\section{Results}\label{sec:4}}
 
-\hypertarget{upload-clusters-cluster-description}{%
-\subsection{Upload Clusters / cluster
-description}\label{upload-clusters-cluster-description}}
+\hypertarget{sec:4.1}{%
+\subsection{Upload Clusters / cluster description}\label{sec:4.1}}
 
-The temporal profiles used to cluster the local authorities have been
-summarised in graph {[}.{]}, which shows a composite profile of mean
-upload speeds per hour per day for each cluster. For upload speeds, 345
-of 382 local authorities, or over 62 million people, fall into cluster 6
-or cluster 9. Graph {[}.{]} suggests that these two clusters have
-relatively similar temporal profiles, which are flatter than the other,
-smaller clusters, suggesting better reliability of service. However, the
-upload speeds at all times for cluster 9 are substantially higher than
-for cluster 6, which is an indication of better quality of service. This
-difference is reflected not only in the mean speeds for these clusters
-for the whole sample, but also the mean upload speeds in the morning
-peak from 9:00-10:59, as well as the evening peak period from
-19:00-20:59. In comparison, the time profile in graph {[}.{]} shows
-upload speeds in cluster 1 are on average lower at certain times of day
-during the study period than any other cluster, whilst the profile for
-cluster 3 appears to show speeds fluctuating as much as cluster 1, but
-at levels usually higher even than cluster 9.
+The temporal profiles of the local authority clusters have been
+summarised in Figures \ref{UpClusterL} and \ref{UpClusterS} and Table
+\ref{up.cluster.descr}. The graphs show a composite profile of mean
+upload speeds per hour per day for each cluster, with the largest, in
+terms of the LAD membership, five clusters in Figure \ref{UpClusterL},
+and the next six in Figure \ref{UpClusterS}. These figures and table
+provide a comprehensive overview of the quality and reliability of
+experienced broadband in different parts of the UK, the temporal
+clusters offering a novel approach to understanding spatial disparity.
 
-The variability of the smaller clusters may be related to the fewer
-speed tests from fewer local authorities that have been averaged, whilst
-averaging greater numbers of speed tests could artificially flatten the
-profile. This appears not to be the case for cluster 9, as there is
-negligible difference between morning and evening upload speeds, at 1\%
-slower in the morning, confirming a high level of reliability of
-service. In comparison, upload speeds in cluster 6 are 4\% slower in the
-morning than in the evening and experienced the joint highest ratio of
-standard deviation to mean across the time period under assessment. This
-suggests that although the time profile is relatively flat in graph
-{[}.{]}, the experience is one of speeds that fluctuate from a lower
-mean, and therefore might more often impact on online activities. Still,
-in most of the smaller clusters the reliability of service is worse than
-in cluster 9 in terms of the ratio of standard deviation to mean. The
-smaller clusters are mainly worse than clusters 9 and 6 in terms of the
-ratio of upload speed in the morning peak compared to the evening peak.
-In five of the smaller clusters, which are home to almost 3 million
-people, speeds are 12-19\% slower between 9:00-10:59 compared to
-19:00-20:59.
+\begin{figure}
+\includegraphics[width=0.95\linewidth]{figures/upClusterL} \caption{\label{UpClusterL}Temporal profilies for upload speed large clusters}\label{fig:unnamed-chunk-2}
+\end{figure}
 
-The quality and reliability of broadband service is much better in the
-115 Local Authority areas in cluster 9, which are mostly in urban or
-suburban areas, compared to cluster 6. These include 13 London Boroughs
-(of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7
-constituent authorities of the West Midlands Combined Authority, as well
-as cities like Glasgow, Leicester, Nottingham, Sheffield, and the
-Portsmouth and Southampton conurbation. There are also some notable
-medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes,
-and York, and many suburban districts from the South East of England to
-South Tyneside. Meanwhile, the 230 local authorities in cluster 6, which
-have lower speeds on average and more variation in service still include
-major urban areas, such as Bristol, Liverpool and Leeds, and many
-suburban areas, but also include some of the most rural areas in the
-country. Meanwhile, Cluster 1, with 10 local authorities that are home
-to over 1 million people has the second slowest speeds in the morning
-compared to the evening `peaks' and the second highest ratio of standard
-deviation to mean. This cluster's most populous area is Westminster in
-central London.
+\begin{figure}
+\includegraphics[width=0.95\linewidth]{figures/upClusterS} \caption{\label{UpClusterS}Temporal profilies for upload speed small clusters}\label{fig:unnamed-chunk-3}
+\end{figure}
 
-\textbf{further description of temporal profiles of other clusters -
-include all in graph?} Indeed, the only cluster where upload speeds were
-slower in the evening than in the morning was cluster 5, made up of 2
-local authorities with less than 200,000 people: Three Rivers, a
-suburban district north of London, and Fylde, a seaside suburb of
-Blackpool. However, these are likely to be outliers and may not have
-many tests from which the clusters are calcualted. \textbf{Take out
-smaller clusters?} The exceptions can be found in seven of the eight
-much smaller clusters including 35 local authority districts, where AM
-peak upload speeds are between 6\% and 18\% slower than PM peak upload
-speeds, although the mean speeds for each cluster are higher than
-cluster 6. Indeed, in 25 local authorities with a combined population of
-almost 3 million, speeds are 13\% or more slower in the morning than in
-the evening. Included in this latter group are central London borough of
-Westminster and the London Borough of Newham, rural authorities like
-Eden and West Devon, and small cities like Dundee and Carlisle.
+\begin{table}[!htbp] \centering 
+  \caption{Upload speed cluster characteristics\label{up.cluster.descr}} 
+  \label{} 
+\footnotesize 
+\begin{tabular}{@{\extracolsep{0pt}} ccccccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+Cluster & N. of LADs & LAD population & mean speed & SD speed & mean AM speed & mean PM speed \\ 
+\hline \\[-1.8ex] 
+1 & 5 & 343100 & 8557.389 & 6138.998 & 7746.892 & 9562.508 \\ 
+2 & 2 & 265600 & 10921.69 & 6686.872 & 9674.043 & 10645.43 \\ 
+3 & 4 & 474700 & 10200.753 & 5658.418 & 9470.446 & 11236.237 \\ 
+4 & 1 & 91100 & 9688.899 & 6121.587 & 7816.234 & 9689.224 \\ 
+5 & 1 & 79800 & 10126.506 & 6024.124 & 9030.22 & 11100.605 \\ 
+6 & 155 & 29535700 & 9396.565 & 5839.097 & 9161.33 & 9580.422 \\ 
+7 & 4 & 559800 & 10119.495 & 6102.311 & 9813.353 & 11070.093 \\ 
+8 & 5 & 436300 & 9428.931 & 6254.047 & 8681.534 & 10434.339 \\ 
+9 & 32 & 6355500 & 10878.131 & 5957.156 & 10831.557 & 11070.88 \\ 
+10 & 4 & 699600 & 10795.469 & 6005.111 & 9257.855 & 10697.492 \\ 
+11 & 33 & 5771400 & 10845.045 & 5936.36 & 10780.877 & 10987.827 \\ 
+12 & 10 & 1544900 & 9550.756 & 6165.524 & 9254.436 & 9047.503 \\ 
+13 & 126 & 20277700 & 8392.451 & 5849.227 & 8298.513 & 8521.828 \\ 
+\hline \\[-1.8ex] 
+\multicolumn{7}{l}{Note: All speed measures are upload speeds} \\ 
+\end{tabular} 
+\end{table}
 
-\hypertarget{aux-regressions}{%
-\subsection{aux regressions}\label{aux-regressions}}
+The largest two clusters, comprising \(281\) local authorities and
+almost \(50\) million people, fall into clusters \(6\) and cluster
+\(13\). Cluster \(13\) has one of the lowest aggregate mean upload speed
+of any of the clusters, and the second highest ratio of the standard
+deviation to the mean. This suggests that those living in local
+authorities in this cluster experienced some of the lowest quality
+broadband services in terms of upload speeds in the UK. However, the
+variation in upload speeds in cluster \(13\), which can be an indication
+of its reliability, does not seem to disproportionately affect the
+morning peak from 9:00-10:59, as upload speeds are, on average, only
+\(2.6\)\% slower than in the evening peak period between 19:00 and
+20:59. Although cluster \(13\) does include some rural, remote areas of
+the country, such as Northwest Scotland, Cornwall and Powys in Wales as
+shown on Figure \ref{map.up.clusters}, it also includes the major
+metropolitan centres of Bristol, Liverpool, and Newcastle, as well as
+nine (of \(32\)) London Boroughs and plenty of home county and suburban
+areas. Meanwhile, those living in cluster \(6\) experience aggregate
+mean upload speeds of about \(1\)Mb/s faster than those in cluster
+\(13\), but still lower than the other three large clusters and most of
+the smaller clusters, suggesting a middling quality of service. However,
+the time profile for cluster \(6\) in Figure \(3\)a shows that upload
+speeds drop quickly from \(6:00\) to 7:00 on a Monday and peak between
+23:00 and midnight on Wednesday and Thursday, but tend to be lower
+during the working day. Indeed, those in cluster \(6\), including people
+living in Birmingham, Leeds and Sheffield, in \(12\) London Boroughs,
+and many suburban areas and smaller cities like Cardiff, Oxford and
+Cambridge experience mean upload speeds in the morning that are
+\(4.4\)\% lower than in the evening. This lower level of reliability may
+be as a result of increased demand, e.g.~for telecommuting, especially
+as there are few truly remote areas in this cluster.
 
-Auxiliary regressions indicate that the speed tests in cluster 9
-authorities are also more likely to have been run on services provided
-by Virgin Media, suggesting they are in the half of the country with the
-most lucrative ICT market, which originally attracted the cable TV
-provider (OfCom\ldots). For example, although auxiliary regressions show
-that Cluster 6 local authorities are more likely to be in the South of
-the UK than Cluster 9, the cluster notably includes Southern rural
-districts from Cornwall to North Norfolk. Slower speeds could reflect
-the lower quality of service in rural areas compared to urban and
-suburban areas. There is frustration, however, as auxiliary regressions
-show that those living in cluster 6 had the highest probability of
-testing their broadband between 9:00 and 11:00 of any of the 10 upload
-speed clusters.
+\begin{figure}
+\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-5}
+\end{figure}
 
-the finding in the auxiliary regressions that those in cluster 9 ran the
-fewest speed tests per person during the morning period of any cluster.
-Now this may be an indication of fewer people working from home, less
-contention, and less resultant frustration. Cluster 9 comprises many
-central urban areas and has the lowest number of established businesses
-per inhabitant, which could be interpreted as a dominance of large
-employers. However, the job density is lower than cluster 6, meaning
-there are not as many jobs per resident in these areas. As the cluster
-includes many suburban areas too, which may be largely residential,
-could the quality and reliability of internet service be reinforcing
-patterns of telecommuting by those in wealthier suburbs who can work
-from home? Earnings in cluster 9 are second highest of all the clusters,
-with only Cluster 2 (comprising just North East Lincolnshire and East
-Lothian, population 265k) earning more per person.
+However, whilst the difference in experienced mean upload speeds between
+9:00-10:59 and 19:00-20:59 in cluster \(6\) is greater than in any of
+the other large clusters, it is smaller than this difference in all the
+clusters shown in Figure \ref{UpClusterL}. Whilst this may suggest a
+lack of reliability, especially considering the large spikes and dips in
+these clusters, the statistics in Table \(1\) show a more varied
+picture. The experienced upload speeds in cluster \(1\) are the second
+slowest of any cluster, the ratio of standard deviation to mean is the
+highest, as is the ratio of speeds in the morning compared to the
+evening peak hours analysed. The five, mainly rural authorities in this
+cluster clearly lack broadband resilience, for telecommuting or
+otherwise. Home to almost \(3.5\) thousand people, the districts are
+scattered around the country: Ceredigion in West Wales, Darlington in
+the Northeast, Eden between the Lake District and the North Penines,
+Rossendale in the Northwest, and Rutland in the East Midlands. In
+comparison, clusters \(3\) and \(10\) have relatively high mean upload
+speeds, and low standard deviations, suggesting reliability and
+resilience, as well as quality broadband services, despite slower speeds
+in the morning peak. Like cluster \(6\), is this also a sign of
+increased telecommuting? The eight LADs in clusters \(3\) and \(10\) are
+home to over one million people, and are all more urban and suburban
+than cluster \(1\), including the London Borough of Newham, the
+Southeast commuter towns of Harlow and Luton, the Birmingham suburbs of
+Bromsgrove and Cannock Chase, Dundee City, and the towns of Eastbourne
+and Corby.
 
-The auxiliary regressions show that compared to the two authorities in
-cluster 5, all the other clusters had a lower percentage of working
-people in managerial, professional and administrative jobs.??? Yet the
-auxiliary regression suggests that there are not many tests being run
-during the am peak in cluster 1. This may be because there are fewer
-people working at home checking their broadband than in most other
-clusters. Indeed, the auxiliary regressions indicate that cluster 1 has
-the highest job density or proportion of jobs to working-age population,
-which is likely to due to the presence of Westminster, central London,
-cluster 1's most populous local authority. Westminster not only has more
-workplaces than residents, but it is reasonable to presume that many who
-would normally work in Westminster, but be able to work from home during
-lockdown are likely to live outside central London and not be subject to
-the fluctuating speeds there. Workplaces, meanwhile, some of which would
-still have been open, could be running programmes that cause the
-slowdown and variation, but would be more likely to have their own
-in-house diagnostics, rather than using a service like Speedchecker Ltd.
+Meanwhile, the most resilient broadband service in terms of both quality
+and reliability appears to be found in clusters \(9\) and \(11\), home
+to over \(12\) million people. The mean speeds for these clusters are
+second and third highest, the ratio of the standard deviation to the
+mean are the two lowest, and the experienced upload speeds in the
+morning peak are not that different from those experienced in the
+evening peak. In Figure \ref{UpClusterL}, cluster \(11\) shows more
+noticeable peaks and troughs, but the lowest points are not at the peak
+times described in Table \(1\). Rather, the slowest upload speeds on
+average occur between 6:00-7:00 on Monday morning, 14:00-15:00 on
+Friday, and \(16:00-17:00\) on Thursday. These slowest times are still
+mostly faster than the average hourly upload speeds in cluster \(13\).
+Although not all of the local authorities in these two clusters are
+urban, among the \(65\) LADs are Manchester and three of the nine other
+boroughs of Greater Manchester, Glasgow, Nottingham, both Portsmouth and
+Southampton, eight London Boroughs, four of the seven constituent
+authorities of the West Midlands Combined Authority. There are also a
+number of other tightly bounded urban areas, such as Aberdeen,
+Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at
+the centre of less confined districts like Burnley, Milton Keynes, and
+Northampton. The local authorities in these two clusters clearly
+experienced resilient broadband that could support high levels of
+telecommuting. They are \emph{not} on the wrong side of the first level
+digital divide, but how likely are they to be able to take advantage of
+their resilient ICT infrastructure and services?
 
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
+\hypertarget{sec:4.2}{%
+\subsection{aux regressions}\label{sec:4.2}}
+
+\begin{sidewaystable}[!htbp] \centering 
+  \caption{Auxiliary multinomial regression of upload speed clusters on socio-economic and geographic LAD variables\label{aux}} 
+  \label{} 
+\tiny 
+\begin{tabular}{@{\extracolsep{5pt}}lccccccccccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+\\[-1.8ex] & 1 & 2 & 3 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 \\ 
+\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11)\\ 
+\hline \\[-1.8ex] 
+ pop, 2018 & $-$0.00004$^{***}$ & 0.00002$^{*}$ & 0.00001 & 0.00002$^{***}$ & 0.00002$^{***}$ & 0.00000 & 0.00002$^{***}$ & 0.00002$^{***}$ & 0.00002$^{***}$ & 0.00002$^{**}$ & 0.00002$^{***}$ \\ 
+  & (0.00002) & (0.00001) & (0.00001) & (0.00001) & (0.00001) & (0.00002) & (0.00001) & (0.00001) & (0.00001) & (0.00001) & (0.00001) \\ 
+  & & & & & & & & & & & \\ 
+ job density, 2018 & $-$0.536$^{***}$ & $-$1.834$^{***}$ & $-$0.132$^{***}$ & $-$0.925$^{***}$ & $-$1.208$^{***}$ & $-$0.299$^{***}$ & $-$1.746$^{***}$ & $-$1.436$^{***}$ & 3.350$^{***}$ & 3.400$^{***}$ & 0.630$^{***}$ \\
+  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
+  & & & & & & & & & & & \\ 
+ distance to nearest met.
+area & $-$0.034$^{***}$ & $-$0.014$^{***}$ & 0.002$^{***}$ & $-$0.020$^{***}$ & $-$0.074$^{***}$ & $-$0.044$^{***}$ & $-$0.013$^{***}$ & $-$0.036$^{***}$ & $-$0.031$^{***}$ & $-$0.036$^{***}$ & $-$0.024$^{***}$ \\
+  & (0.0005) & (0.001) & (0.0002) & (0.002) & (0.0001) & (0.0002) & (0.002) & (0.0005) & (0.0003) & (0.0002) & (0.002) \\ 
+  & & & & & & & & & & & \\ 
+ distance to London & 0.007$^{***}$ & 0.002 & $-$0.016$^{***}$ & 0.001 & 0.004$^{***}$ & 0.004$^{***}$ & $-$0.002$^{*}$ & 0.005$^{**}$ & $-$0.002 & 0.003 & 0.006$^{***}$ \\ 
+  & (0.001) & (0.002) & (0.0004) & (0.001) & (0.001) & (0.001) & (0.001) & (0.002) & (0.002) & (0.002) & (0.001) \\ 
+  & & & & & & & & & & & \\ 
+ South of the UK & $-$0.410$^{***}$ & $-$1.451$^{***}$ & $-$0.039$^{***}$ & $-$0.111$^{***}$ & $-$0.048$^{***}$ & $-$0.841$^{***}$ & $-$0.798$^{***}$ & 1.492$^{***}$ & 0.610$^{***}$ & 0.798$^{***}$ & 2.403$^{***}$ \\
+  & (0.00000) & (0.00000) & (0.00000) & (0.00001) & (0.00000) & (0.00000) & (0.00001) & (0.00000) & (0.00001) & (0.00001) & (0.00001) \\ 
+  & & & & & & & & & & & \\ 
+ managerial jobs, 2020 & 0.939$^{***}$ & 0.704$^{***}$ & 0.435$^{***}$ & 0.704$^{***}$ & 0.316$^{***}$ & 0.786$^{***}$ & 0.576$^{***}$ & 0.311$^{***}$ & 0.476$^{***}$ & 0.594$^{***}$ & 0.615$^{***}$ \\
+  & (0.00004) & (0.0001) & (0.00004) & (0.00004) & (0.00002) & (0.0001) & (0.00003) & (0.00003) & (0.00002) & (0.00004) & (0.00003) \\ 
+  & & & & & & & & & & & \\ 
+ tech jobs, 2020 & 0.096$^{***}$ & $-$0.257$^{***}$ & $-$0.071$^{***}$ & $-$0.111$^{***}$ & $-$0.206$^{***}$ & 0.199$^{***}$ & $-$0.126$^{***}$ & $-$0.606$^{***}$ & $-$0.180$^{***}$ & $-$0.398$^{***}$ & $-$0.112$^{***}$ \\
+  & (0.00004) & (0.00004) & (0.00004) & (0.00003) & (0.00003) & (0.0001) & (0.00003) & (0.00003) & (0.00002) & (0.00004) & (0.00003) \\ 
+  & & & & & & & & & & & \\ 
+ skilled trade jobs, 2020 & 0.651$^{***}$ & 0.160$^{***}$ & $-$0.191$^{***}$ & 0.236$^{***}$ & 0.604$^{***}$ & $-$0.184$^{***}$ & 0.205$^{***}$ & 0.597$^{***}$ & 0.108$^{***}$ & $-$0.022$^{***}$ & 0.295$^{***}$ \\
+  & (0.00004) & (0.00004) & (0.00003) & (0.00003) & (0.00003) & (0.00004) & (0.00002) & (0.00003) & (0.00003) & (0.00004) & (0.00003) \\ 
+  & & & & & & & & & & & \\ 
+ professional jobs, 2020 & $-$0.118$^{***}$ & $-$0.234$^{***}$ & $-$0.121$^{***}$ & $-$0.172$^{***}$ & $-$0.514$^{***}$ & $-$0.172$^{***}$ & $-$0.349$^{***}$ & $-$0.351$^{***}$ & $-$0.245$^{***}$ & $-$0.344$^{***}$ & $-$0.229$^{***}$ \\
+  & (0.00005) & (0.0001) & (0.0001) & (0.00005) & (0.00003) & (0.0001) & (0.00004) & (0.0001) & (0.00003) & (0.00005) & (0.0001) \\ 
+  & & & & & & & & & & & \\ 
+ administrative jobs, 2020 & 0.019$^{***}$ & $-$0.836$^{***}$ & $-$0.040$^{***}$ & $-$0.117$^{***}$ & $-$0.139$^{***}$ & 0.206$^{***}$ & $-$0.058$^{***}$ & $-$0.200$^{***}$ & $-$0.055$^{***}$ & $-$0.168$^{***}$ & $-$0.177$^{***}$ \\
+  & (0.00003) & (0.00002) & (0.00004) & (0.00001) & (0.00003) & (0.00003) & (0.00001) & (0.00002) & (0.00002) & (0.00002) & (0.00002) \\ 
+  & & & & & & & & & & & \\ 
+ leisure jobs, 2020 & $-$0.198$^{***}$ & $-$0.180$^{***}$ & $-$0.225$^{***}$ & $-$0.476$^{***}$ & $-$0.654$^{***}$ & $-$0.820$^{***}$ & $-$0.537$^{***}$ & $-$0.935$^{***}$ & $-$0.353$^{***}$ & $-$0.625$^{***}$ & $-$0.491$^{***}$ \\
+  & (0.00002) & (0.00004) & (0.00004) & (0.00002) & (0.00002) & (0.00003) & (0.00002) & (0.00001) & (0.00002) & (0.00003) & (0.00002) \\ 
+  & & & & & & & & & & & \\ 
+ machine operation jobs, 2020 & $-$0.336$^{***}$ & 0.207$^{***}$ & 0.392$^{***}$ & 0.010$^{***}$ & $-$0.433$^{***}$ & 0.139$^{***}$ & $-$0.099$^{***}$ & $-$0.139$^{***}$ & $-$0.144$^{***}$ & 0.098$^{***}$ & $-$0.179$^{***}$ \\
+  & (0.00002) & (0.00003) & (0.00003) & (0.00002) & (0.00001) & (0.00003) & (0.00002) & (0.00001) & (0.00001) & (0.00002) & (0.00001) \\ 
+  & & & & & & & & & & & \\ 
+ earnings, 2019 & $-$0.003$^{*}$ & 0.010$^{***}$ & 0.012$^{***}$ & 0.020$^{***}$ & 0.027$^{***}$ & 0.001 & 0.020$^{***}$ & 0.016$^{***}$ & 0.015$^{***}$ & 0.025$^{***}$ & 0.014$^{***}$ \\ 
+  & (0.002) & (0.002) & (0.002) & (0.001) & (0.001) & (0.003) & (0.001) & (0.002) & (0.001) & (0.001) & (0.001) \\ 
+  & & & & & & & & & & & \\ 
+ n.
+business est.
+per hab., 2019 & 0.126$^{***}$ & $-$0.120$^{***}$ & $-$0.094$^{***}$ & $-$0.133$^{***}$ & 0.123$^{***}$ & $-$0.051$^{***}$ & $-$0.334$^{***}$ & $-$0.133$^{***}$ & $-$0.150$^{***}$ & 0.289$^{***}$ & 0.377$^{***}$ \\
+  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
+  & & & & & & & & & & & \\ 
+ NVQ4+ & $-$0.141$^{***}$ & 0.064$^{***}$ & $-$0.091$^{***}$ & $-$0.070$^{***}$ & $-$0.010$^{***}$ & 0.004$^{***}$ & 0.016$^{***}$ & 0.170$^{***}$ & $-$0.110$^{***}$ & $-$0.038$^{***}$ & $-$0.035$^{***}$ \\
+  & (0.0001) & (0.0001) & (0.0001) & (0.0001) & (0.0001) & (0.0002) & (0.0001) & (0.0001) & (0.0001) & (0.0001) & (0.0001) \\ 
+  & & & & & & & & & & & \\ 
+ AM tests per hab., 2020 & 0.0005$^{***}$ & $-$0.002$^{***}$ & $-$0.005$^{***}$ & 0.010$^{***}$ & 0.0004$^{***}$ & $-$0.001$^{***}$ & $-$0.002$^{***}$ & $-$0.005$^{***}$ & $-$0.013$^{***}$ & $-$0.001$^{***}$ & 0.016$^{***}$ \\
+  & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) \\ 
+  & & & & & & & & & & & \\ 
+ % of Virgin Media connections & $-$0.044$^{***}$ & 1.578$^{***}$ & 1.210$^{***}$ & 0.248$^{***}$ & $-$1.724$^{***}$ & $-$0.242$^{***}$ & 3.109$^{***}$ & $-$0.085$^{***}$ & 1.214$^{***}$ & $-$3.889$^{***}$ & $-$0.745$^{***}$ \\
+  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
+  & & & & & & & & & & & \\ 
+ Constant & 0.321$^{***}$ & $-$0.436$^{***}$ & 0.199$^{***}$ & $-$2.953$^{***}$ & 0.278$^{***}$ & 0.002$^{***}$ & $-$0.866$^{***}$ & 0.788$^{***}$ & 2.600$^{***}$ & 0.017$^{***}$ & 0.022$^{***}$ \\ 
+  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
+  & & & & & & & & & & & \\ 
+\hline \\[-1.8ex] 
+McFadden's R squared & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 \\ 
+N & 323 & 323 & 323 & 323 & 323 & 323 & 323 & 323 & 323 & 323 & 323 \\ 
+Akaike Inf. Crit. & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 \\ 
+\hline 
+\hline \\[-1.8ex] 
+\textit{Note:}  & \multicolumn{11}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
+\end{tabular} 
+\end{sidewaystable}
+
+Clusters \(3\), \(9\), \(10\) and \(11\) seem to benefit most from high
+quality and resilient broadband services. The dips in mean upload speeds
+in clusters \(6\), \(3\) and \(10\) during the morning peak are
+suggestive of more use during the working day, and potentially more
+telecommuting. Using auxiliary regressions, we test whether the clusters
+that have higher mean speeds and more reliable services are more urban,
+more wealthy, and / or more likely to benefit from a choice of high
+quality internet services. We also measure which of our clusters are
+more likely to have a higher proportion of occupations which enable
+telecommuting because of the nature of the work. The results of these
+auxiliary regressions are presented in Table \ref{aux}. The dependent
+variable is the LAD cluster membership as described in Section
+\ref{#sec:3} and equation \ref{eq1} and each column represent a
+different cluster. The reference case is cluster \(4\), which includes
+the local authority of Hambleton in North Yorkshire, a rural area of
+just over ninety thousand people. Mean, experienced upload speeds in
+cluster \(4\) are close to both the average speeds for the \(13\)
+clusters (\(9.9\)Mb/s) and the pre-clustered average for the whole
+sample (\(9.3\)Mb/s). The standard deviation for cluster \(4\) and the
+ratio of evening to morning peak speeds are indications of worse
+reliability than many of the other clusters. Hence, the results in Table
+\ref{aux} should be seens as relative rather than absolute
+probabilities.
+
+First, the number of speed tests run per cluster inhabitant between
+9:00-10:59 is an indication of satisfaction or at least a lack of
+concern over broadband speeds and quality of service. People in clusters
+\(3\), \(9\), \(10\) and \(11\) all ran fewer speed tests (at a per
+capita basis) at this time of day than people in clusters \(6\) and
+\(13\). Those in clusters \(3\), \(9\), \(10\) and \(11\) also benefit
+from a higher proportion of fast Virgin cable connections than those in
+clusters \(6\) and \(13\). Since services provided by Virgin Media are
+only available to \(45\)\% of premises in the UK \citep{ofcom2016},
+where the more lucrative ICT market originally attracted the cable TV
+provider, this is an indication that people in these clusters are more
+likely to live on the right side of the first level,
+infrastructure-based digital divide. Another indication is if people in
+these clusters are more likely to be closer to London or a metropolitan
+area. The broadband speed tests run in the authorities in cluster \(3\)
+are more likely to be taking place close to London than those run in any
+of the other clusters, a result that makes sense considering two of the
+four authorities in cluster \(3\) are the London commuter towns of
+Harlow and Luton. However, even though London was also included in the
+variable calculating distance to one of the ten major metropolitan areas
+in England, or Glasgow or Cardiff, tests run in cluster \(3\) are likely
+to be furthest away. \textbf{Did you include London in distMet as well?
+- code suggests so} \textbf{Do we need to re-think this variable
+(e.g.~remove Cardiff, add Leicester or\ldots)}
+
+\hypertarget{sec:5}{%
+\section{Conclusions}\label{sec:5}}
 

### 2020-12-11 src/descriptive_Clusters.Rmd
updates in text, tables, figures

--- 2020-10-21 src/descriptive_Clusters.Rmd
@@
 TS2019 <- selectByDate(TSbb19_20sp, year = 2019)
+
+#create time variation plots for speedtests for each year
+Tests19 <- count(TS2019, date)
+TS2019 <- left_join(TS2019, Tests19)
+names(TS2019)[13] <- "test frequency 2019"
+time.var.plot2019 <- timeVariation(TS2019, pollutant = "test frequency 2019", statistic = "mean")
+Tests20 <- count(TS2020, date)
+TS2020 <- left_join(TS2020, Tests20)
+names(TS2020)[13] <- "test frequency 2020"
+time.var.plot2020 <- timeVariation(TS2020, pollutant = "test frequency 2020", statistic = "mean")
+
+#export time variation plots for paper
+out.plot <- paste(path, "/paper/v1/figures/time.var.plot2019.png", sep = "")
+png(out.plot)
+time.var.plot2019
+dev.off()
+
+out.plot <- paste(path, "/paper/v1/figures/time.var.plot2020.png", sep = "")
+png(out.plot)
+time.var.plot2020
+dev.off()
 
@@
 #plot cluster statistic for mean upload speed
-UpCluster <- ggplot(cluster.up, aes(help.date.time, mean.up, 
-                                         colour = clusterID))
-UpCluster + geom_line() + labs(x = "Weekdays March-May") + 
+UpClusterL <- ggplot(cluster.up, aes(help.date.time, mean.up, 
+                                      colour = clusterID)) +
+  theme_minimal() +
+  geom_line() + 
+  labs(x = "Weekdays March-May", y = "mean upload speed", col = "Cluster ID") + 
   scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
@@
                     labels = "%a"), labels = 
-                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + geom_point()
+                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + 
+  geom_point() +
+  theme(legend.position = c(0.7, 0.08),
+        legend.direction = "horizontal")
+
+#export UpClusterL plot for paper
+out.plot <- paste(path, "/paper/v1/figures/UpClusterL.png", sep = "")
+png(out.plot)
+UpClusterL
+dev.off()
+
+#plot next 6 smaller clusters, but exclude 2 clusters with 1 LAD and under 100k pop
+cluster.upS <- TS2020 %>%
+  group_by(cluster.up, weekday, hour) %>%
+  summarise(mean.up = mean(speedup))
+names(cluster.upS)[1] <- "clusterID"
+names(LAD.cluster.up)[1] <- "clusterID"
+cluster.upS <- left_join(cluster.upS, LAD.cluster.up)
+cluster.upS <- cluster.upS[which(cluster.upS$up.LADs < 10 & cluster.upS$up.LADs > 1),]
+cluster.upS$clusterID <- as.factor(cluster.upS$clusterID)
+#add fake dates for plotting
+help.dates <- tibble(help.date.time = c(
+  make_datetime(year = 2020, month = 3, day = 2),
+  make_datetime(year = 2020, month = 3, day = 3),
+  make_datetime(year = 2020, month = 3, day = 4),
+  make_datetime(year = 2020, month = 3, day = 5),
+  make_datetime(year = 2020, month = 3, day = 6)),
+                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+cluster.upS <- merge(cluster.upS, help.dates, by = "weekday")
+cluster.upS$hour <- paste(cluster.upS$hour, ":00", sep="")
+cluster.upS$help.date.time <- with(cluster.upS, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+#plot cluster statistic for mean upload speed
+UpClusterS <- ggplot(cluster.upS, aes(help.date.time, mean.up, 
+                                      colour = clusterID)) +
+  theme_minimal() +
+  geom_line() + 
+  labs(x = "Weekdays March-May", y = "mean upload speed", col = "Cluster ID") + 
+  scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
+                    "2020-03-03 15:00:00", "2020-03-04 15:00:00",
+                    "2020-03-05 15:00:00", "2020-03-06 15:00:00"), 
+                    labels = "%a"), labels = 
+                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + 
+  geom_point() +
+  theme(legend.position = c(0.8, 0.08),
+        legend.direction = "horizontal")
+
+#export UpClusterS plot for paper
+out.plot <- paste(path, "/paper/v1/figures/UpClusterS.png", sep = "")
+png(out.plot)
+UpClusterS
+dev.off()
 

### 2020-12-15 paper/v1/broadband.speed.covid.Rmd
minor changes

@@
 socio-economic and geographic variables, which are discussed in detail in
-Section \ref{#sec:4.2}.
+the relevant [section](#sec:4.2).
 This analysis enables us to provide a more nuanced understanding of how
@@
 clustersLADup <- clustersLADup %>% 
-  mutate(across(4:7, round, 3))
+  mutate(across(4:7, round, 0))
 
@@
           header = F,
+          digit.separator = ",",
+          digit.separate = c(3,3),
           covariate.labels=c("Cluster",
@@
 Whilst this may suggest a lack of reliability, especially considering the large
-spikes and dips in these clusters, the statistics in Table $1$ show 
+spikes and dips in these clusters, the statistics in Table \ref{up.cluster.descr} show 
 a more varied picture.
@@
                              "AM tests per hab., 2020",
-                             "% of Virgin Media connections"),
+                             "Virgin Media \\%, 2020"),
           add.lines = list(c("McFadden's R squared",
@@
 The results of these auxiliary regressions are presented in Table \ref{aux}.
-The dependent variable is the LAD cluster membership as described in Section \ref{#sec:3}
+The dependent variable is the LAD cluster membership as described in the 
+[Methods and data](#sec:3) section
 and equation \ref{eq1} and each column represent a different cluster.

### 2020-12-15 paper/v1/broadband.speed.covid.tex
minor changes

@@
 socio-economic and geographic variables, which are discussed in detail
-in Section \ref{#sec:4.2}. This analysis enables us to provide a more
-nuanced understanding of how telecommuting and technology intersect at a
-time of extreme demand, and what lessons this time has for a future
-where telecommuting is likely to remain a common means of accessing work
-and broadband services, as well as infrastructure, must be fit for
-purpose.
+in the relevant \protect\hyperlink{sec:4.2}{section}. This analysis
+enables us to provide a more nuanced understanding of how telecommuting
+and technology intersect at a time of extreme demand, and what lessons
+this time has for a future where telecommuting is likely to remain a
+common means of accessing work and broadband services, as well as
+infrastructure, must be fit for purpose.
 
@@
 \hline \\[-1.8ex] 
-1 & 5 & 343100 & 8557.389 & 6138.998 & 7746.892 & 9562.508 \\ 
-2 & 2 & 265600 & 10921.69 & 6686.872 & 9674.043 & 10645.43 \\ 
-3 & 4 & 474700 & 10200.753 & 5658.418 & 9470.446 & 11236.237 \\ 
-4 & 1 & 91100 & 9688.899 & 6121.587 & 7816.234 & 9689.224 \\ 
-5 & 1 & 79800 & 10126.506 & 6024.124 & 9030.22 & 11100.605 \\ 
-6 & 155 & 29535700 & 9396.565 & 5839.097 & 9161.33 & 9580.422 \\ 
-7 & 4 & 559800 & 10119.495 & 6102.311 & 9813.353 & 11070.093 \\ 
-8 & 5 & 436300 & 9428.931 & 6254.047 & 8681.534 & 10434.339 \\ 
-9 & 32 & 6355500 & 10878.131 & 5957.156 & 10831.557 & 11070.88 \\ 
-10 & 4 & 699600 & 10795.469 & 6005.111 & 9257.855 & 10697.492 \\ 
-11 & 33 & 5771400 & 10845.045 & 5936.36 & 10780.877 & 10987.827 \\ 
-12 & 10 & 1544900 & 9550.756 & 6165.524 & 9254.436 & 9047.503 \\ 
-13 & 126 & 20277700 & 8392.451 & 5849.227 & 8298.513 & 8521.828 \\ 
+1 & 5 & 343100 & 8557 & 6139 & 7747 & 9563 \\ 
+2 & 2 & 265600 & 10922 & 6687 & 9674 & 10645 \\ 
+3 & 4 & 474700 & 10201 & 5658 & 9470 & 11236 \\ 
+4 & 1 & 91100 & 9689 & 6122 & 7816 & 9689 \\ 
+5 & 1 & 79800 & 10127 & 6024 & 9030 & 11101 \\ 
+6 & 155 & 29535700 & 9397 & 5839 & 9161 & 9580 \\ 
+7 & 4 & 559800 & 10119 & 6102 & 9813 & 11070 \\ 
+8 & 5 & 436300 & 9429 & 6254 & 8682 & 10434 \\ 
+9 & 32 & 6355500 & 10878 & 5957 & 10832 & 11071 \\ 
+10 & 4 & 699600 & 10795 & 6005 & 9258 & 10697 \\ 
+11 & 33 & 5771400 & 10845 & 5936 & 10781 & 10988 \\ 
+12 & 10 & 1544900 & 9551 & 6166 & 9254 & 9048 \\ 
+13 & 126 & 20277700 & 8392 & 5849 & 8299 & 8522 \\ 
 \hline \\[-1.8ex] 
@@
 lack of reliability, especially considering the large spikes and dips in
-these clusters, the statistics in Table \(1\) show a more varied
-picture. The experienced upload speeds in cluster \(1\) are the second
-slowest of any cluster, the ratio of standard deviation to mean is the
-highest, as is the ratio of speeds in the morning compared to the
-evening peak hours analysed. The five, mainly rural authorities in this
-cluster clearly lack broadband resilience, for telecommuting or
+these clusters, the statistics in Table \ref{up.cluster.descr} show a
+more varied picture. The experienced upload speeds in cluster \(1\) are
+the second slowest of any cluster, the ratio of standard deviation to
+mean is the highest, as is the ratio of speeds in the morning compared
+to the evening peak hours analysed. The five, mainly rural authorities
+in this cluster clearly lack broadband resilience, for telecommuting or
 otherwise. Home to almost \(3.5\) thousand people, the districts are
@@
   & & & & & & & & & & & \\ 
- % of Virgin Media connections & $-$0.044$^{***}$ & 1.578$^{***}$ & 1.210$^{***}$ & 0.248$^{***}$ & $-$1.724$^{***}$ & $-$0.242$^{***}$ & 3.109$^{***}$ & $-$0.085$^{***}$ & 1.214$^{***}$ & $-$3.889$^{***}$ & $-$0.745$^{***}$ \\
+ Virgin Media \%, 2020 & $-$0.044$^{***}$ & 1.578$^{***}$ & 1.210$^{***}$ & 0.248$^{***}$ & $-$1.724$^{***}$ & $-$0.242$^{***}$ & 3.109$^{***}$ & $-$0.085$^{***}$ & 1.214$^{***}$ & $-$3.889$^{***}$ & $-$0.745$^{***}$ \\
   & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
@@
 auxiliary regressions are presented in Table \ref{aux}. The dependent
-variable is the LAD cluster membership as described in Section
-\ref{#sec:3} and equation \ref{eq1} and each column represent a
-different cluster. The reference case is cluster \(4\), which includes
-the local authority of Hambleton in North Yorkshire, a rural area of
-just over ninety thousand people. Mean, experienced upload speeds in
-cluster \(4\) are close to both the average speeds for the \(13\)
-clusters (\(9.9\)Mb/s) and the pre-clustered average for the whole
-sample (\(9.3\)Mb/s). The standard deviation for cluster \(4\) and the
-ratio of evening to morning peak speeds are indications of worse
-reliability than many of the other clusters. Hence, the results in Table
-\ref{aux} should be seens as relative rather than absolute
-probabilities.
+variable is the LAD cluster membership as described in the
+\protect\hyperlink{sec:3}{Methods and data} section and equation
+\ref{eq1} and each column represent a different cluster. The reference
+case is cluster \(4\), which includes the local authority of Hambleton
+in North Yorkshire, a rural area of just over ninety thousand people.
+Mean, experienced upload speeds in cluster \(4\) are close to both the
+average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
+pre-clustered average for the whole sample (\(9.3\)Mb/s). The standard
+deviation for cluster \(4\) and the ratio of evening to morning peak
+speeds are indications of worse reliability than many of the other
+clusters. Hence, the results in Table \ref{aux} should be seens as
+relative rather than absolute probabilities.
 

### 2020-12-18 paper/v1/broadband.speed.covid.Rmd

@@
 Internet Service Providers (ISPs), at least in the UK [@ofcom2017].
+
 Yet the shift towards telecommuting during various stages of lockdown around the world
@@
 pandemic.
-We identify this as a new digital divide, one that fundamentally alters the
+The extreme level of demand for telecommuting fundamentally alters the
 potential returns of internet use for the user and wider community,
@@
 and relies upon access to high quality internet services.
-**WE NEED TO CONNECT OUR FINDINGS WITH THIS AND JUSTIFY THAT THIS IS INDEED A NEW DIVIDE**
+
 As the quality of internet infrastructure and services, as well as variation in
 occupations are spatially dependent and clustered in space, our approach offers a
-framework for understanding which types of places, are more likely to land on the
-right side of this new digital divide.
-By asking how resilient broadband speeds are as experienced in different parts of
-the UK during a time of extreme demand, we interrogate which places benefit from
-the greater economic resilience digital technologies can offer, not only during
-the pandemic, but also into the future.
-
+framework for understanding the impact of and interactions between 
+the different levels of digital division in different places 
+with different characteristics.
+By asking how resilient broadband speeds, and particularly upload speeds are 
+as experienced in different parts of the UK during a time of extreme demand, 
+we interrogate which places benefit from the greater economic resilience 
+digital technologies can offer, not only during the pandemic, 
+but also into the future.
 The structure of this paper is as follows.
 First we review the literature on telecommuting and digital divides to better
-understand the origins of the new digital divide revealed by the pandemic and its
-impact on the economic resilience of different places.
+understand the structural and spatial development of these practices pre-pandemic,
+and thus their importance to the economic resilience of different places.
 We then describe our data and methodology.
@@
 [@headicar2016move; @peters2004employees; @singh2013modeling].
+
 Opportunities for working from home during the current pandemic have likewise not
@@
 in various advanced economies.
+Rural areas, predictably, fare worse.
 Yet whether this variation in infrastructure quality affects the spatial footprint of
-telecommuting has not previously been measured.
+telecommuting has not previously been measured, in part because telecommuting has
+not previously been the cause of greatest demand and pressure on internet services.
+
 There are indications that those who purchase high speed connections consume 
@@
 We seek to answer how internet service resilience contributes to or reduces 
-economic resilience 
-when the latter is dependent upon the capability to work from home, 
-and thus improve our understanding of this new digital divide.
-**RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL DIVIDES?
-ET: OR REINFORCING THE 3RD LEVEL DIGITAL DIVIDE**
+economic resilience when the latter is dependent upon 
+the capability to work from home. 
+We also aim to improve our understanding of the impact 
+of first level digital division on telecommuting, and whether this results in 
+much more fundamental third level digital division than has previously been perceived.
 
 <!-- Covid-19 and working from home, cities, urban structure -->
-The multi-layered digital divides intersect with material divides and the economic
-geography of the UK.
+Furthermore, these multi-layered digital divides intersect with material divides 
+and the economic geography of the UK.
 Following the regional economic resilience literature, which
@@
 Various studies have illustrated severe inequalities between the north and the
-south part of the UK regarding, for example, skills and human capital,
+south regions of England **replace 'UK'** regarding, for example, skills and human capital,
 unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@@
 
+##Time-Series Clustering {#sec:3.1}
 <!-- clustering in general -->
@@
 problems -- is data dimensionality given the multiplicity of data points for every
-individual object -- local authorities in our case -- included in the data set.
+individual object, local authorities in our case, included in the data set.
 Time-series are dynamic data as the value of the observations change as
@@
 
-Time-series clustering methods utilising the whole length of time-series can be
-grouped in three categories.
-The first -- model-based approaches -- is based on recovering the underlying model
-for each time-series and then applying clustering algorithms on the model parameters
-of each time-series [@aghabozorgi2015time].
-The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].
-The second approach is based the formation of vectors of features derived from the
-original time-series.
-These new data of reduced dimensionality are then clustered using conventional
-clustering algorithms.
+**REMOVE THIS PARAGRAPH - IRRELEVANT DETAIL: Time-series clustering methods utilising the whole length of time-series can be grouped in three categories.
+The first -- model-based approaches -- is based on recovering the underlying model for each time-series and then applying clustering algorithms on the model parameters of each time-series [@aghabozorgi2015time].
+The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].The second approach is based the formation of vectors of features derived from the original time-series.
+These new data of reduced dimensionality are then clustered using conventional clustering algorithms.**
 
-For this paper we utilise the third category of time-series clustering methods
+For this paper we utilise a category of time-series clustering methods
 known as shape-based approaches.
@@
 
+##Experienced Broadband Speeds {#sec:3.2}
 <!-- introduction of our broadband speeds data -->
@@
 Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
-This a private company that allows internet users to check their own broadband
+This is a private company that allows internet users to check their own broadband
 upload and download speeds, and stores every speed-check with timestamp and
@@
 Internet Service Providers (ISPs).
+
 We are particularly interested in upload speeds and the frequency of speed tests
 over the period from March to May 2020, as government statements indicate this is
-when UK workers were told to work from home if at all possible **[need ref?]**.
+when UK workers were told to work from home if at all possible [@GovUK2020].
 Average upload speeds are slower than average download speeds, at $9.3$Mb/s 
@@
 Meanwhile, the frequency of speed tests was important in identifying 
-the temporal profile which woud give us most insight into 
+the temporal profile which would give us most insight into 
 experienced internet service and resilience, and provide 
@@
 The frequency of testing indicates that whilst there is an overall trend of increased
-testing **or increased experience of internet service NOT SURE WHAT THIS MEAND** from March to April and then a 
-slight reduction from April to May, this trend masks substantial variation 
-by not only the day of the week, but also time of day, as can be seen in Figure \ref{test2020}.
-Thus, a daily aggregation of upload speeds would mask the variation in experience
-over the course of each weekday.
+testing from March to April and then a slight reduction from April to May, 
+this trend masks substantial variation by not only the day of the week, 
+but also time of day, as can be seen in Figure \ref{test2020}.
+Thus, a daily aggregation of upload speeds would mask the variation in experienced
+service over the course of each weekday.
 Furthermore, the importance of this variation is highlighted by a comparison 
@@
 we decided to include a measure of hourly variation in our temporal profiles.
-However, although this is a large data set ($241,088$ individual tests during the
-working hours of study period), there are not enough observations for each
-Local Authority District (LAD) and for each working hour of each working day
-($631$ speed tests per LAD on average) to profile speeds at that level of detail.
+
+However, although this is a large data set -- $241,088$ individual tests performed 
+during weekday hours across the study period -- 
+there are not enough observations for each Local Authority District (LAD) 
+and for each working hour of each working day -- $631$ speed tests per LAD on average
+-- to profile speeds at that level of detail.
 Therefore, we aggregate all the speed-checks during the $13$ weeks of March to May 
@@
 <!-- metrics -->
-We then review the characteristics of each cluster, including number of LADs, 
-and the temporal profile of upload speed by hour of the day 
-and day of the composite week.
+In [section](#sec:4.1), we review the temporal profile of upload speed 
+by hour of the day and day of the composite week, as well as 
+the experienced speed characteristics of each cluster.
 Since the quality and reliability of internet services vary in time and space due
 to both supply and demand-side influences, we use a number of different measures
-to then describe experienced upload speeds per cluster.
+to describe experienced upload speeds per cluster.
 These include: a) mean, experienced connection speed,
@@
 as experienced in different parts of the UK during a time of extreme demand.
-
-**More data details and some figures, descriptive stats**
 
@@
 temporal patterns of internet service relate to the economic geography of the UK.
-We discuss how the different patterns might support or undermine efforts to work
-from home and maintain safe productivity and whether they reinforce existing
-spatial and social inequalities.
 More specifically, we estimate the following multinomial model:
@@
 
-The identification strategy is as follows.
 Based on the outcomes of the time-series clustering, we identify $J$ distinct
@@
 socio-economic and geographic variables, which are discussed in detail in
-the relevant [section](#sec:4.2).
+the relevant [section](#sec:4.2), in order to explore how the different patterns 
+might support or undermine efforts to work from home and maintain safe productivity 
+and whether they reinforce existing spatial and social inequalities.
 This analysis enables us to provide a more nuanced understanding of how
@@
 The graphs show a composite profile of mean upload speeds per hour 
-per day for each cluster, with the largest, in terms of the LAD membership, 
-five clusters in Figure \ref{UpClusterL}, 
+per day for each cluster, with the largest, in terms of the LAD membership and
+population, five clusters in Figure \ref{UpClusterL}, 
 and the next six in Figure \ref{UpClusterS}.
@@
 ```
-The largest two clusters, comprising $281$ local authorities and almost $50$ million 
-people, fall into clusters $6$ and cluster $13$.
-Cluster $13$ has one of the lowest aggregate mean upload speed of any of the clusters,
+The second largest cluster, comprising $126$ local authorities and over $20$ million 
+people, is cluster $13$.
+Cluster $13$ has the slowest aggregate mean upload speed of any of the clusters,
 and the second highest ratio of the standard deviation to the mean.
@@
 which can be an indication of its reliability, does not seem to disproportionately 
-affect the morning peak from 9:00-10:59, as upload speeds are, on average, 
-only $2.6$% slower than in the evening peak period between 19:00 and 20:59.
-Although cluster $13$ does include some rural, remote areas of the country, 
-such as Northwest Scotland, Cornwall and Powys in Wales as shown on Figure \ref{map.up.clusters},
-it also includes the major metropolitan centres of Bristol, Liverpool, 
-and Newcastle, as well as nine (of $32$) London Boroughs and plenty of home county 
-and suburban areas.
-Meanwhile, those living in cluster $6$ experience aggregate mean upload speeds of 
-about $1$Mb/s faster than those in cluster $13$, but still lower than the other three
-large clusters and most of the smaller clusters, 
-suggesting a middling quality of service.
-However, the time profile for cluster $6$ in Figure $3$a shows that upload speeds 
-drop quickly from $6:00$ to 7:00 on a Monday and peak between 23:00 and midnight 
-on Wednesday and Thursday, but tend to be lower during the working day.
-Indeed, those in cluster $6$, including people living in Birmingham, Leeds and 
-Sheffield, in $12$ London Boroughs, and many suburban areas and smaller cities like
-Cardiff, Oxford and Cambridge experience mean upload speeds in the morning that are
-$4.4$% lower than in the evening.
-This lower level of reliability may be as a result of increased demand, 
-e.g. for telecommuting, especially as there are few truly remote areas 
-in this cluster.
+affect the morning peak from $9:00$-$10:59$, as upload speeds are, on average, 
+only $2.6$% slower than in the evening peak period between $19:00$ and $20:59$,
+when entertainment purposes are likely to be using the most bandwidth.
+In comparison, the five LADs that are home to $343$ thousand people in cluster $1$ 
+not only experience the second slowest mean upload speeds 
+and the highest ratio of standard deviation to the mean, 
+but are also much more affected during the morning peak.
+
+Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs home to $29.5$
+million people, experience aggregate mean upload speeds of about $1$Mb/s faster 
+than those in cluster $13$, but still lower than the other three large clusters 
+and most of the smaller clusters, suggesting a middling quality of service.
+However, the time profile for cluster $6$ in Figure \ref{UpClusterL} shows 
+that upload speeds drop quickly from $6:00$ to $7:00$ on a Monday 
+and peak between $23:00$ and midnight on Wednesday and Thursday, 
+but tend to be lower during the working day.
+Indeed, experienced mean upload speeds in the morning peak are $4.4$% lower 
+than in the evening peak -- a greater, more noticeable change than any of the other
+large clusters experience, but smaller than any of the clusters with temporal
+profiles shown in Figure \ref{UpClusterS}.
+Clusters $8$ and $12$ also have mean upload speeds under $10$Mb/s, but higher than
+clusters $1$ and $13$, and their standard deviation is not dissimilar. 
+However, this masks great variation in when lower speeds are experienced, 
+with the mean upload speeds much lower between $9:00$-$10:59$ than 
+between $19:00$-$20:59$ in cluster $8$, but slightly faster in the morning in 
+cluster $12$.
 
@@
 
-However, whilst the difference in experienced mean upload speeds between 9:00-10:59
-and 19:00-20:59 in cluster $6$ is greater than in any of the other large clusters, 
-it is smaller than this difference in all the clusters shown in Figure \ref{UpClusterL}.
-Whilst this may suggest a lack of reliability, especially considering the large
-spikes and dips in these clusters, the statistics in Table \ref{up.cluster.descr} show 
-a more varied picture.
-The experienced upload speeds in cluster $1$ are the second slowest of any cluster,
-the ratio of standard deviation to mean is the highest, as is the ratio of speeds 
-in the morning compared to the evening peak hours analysed.
-The five, mainly rural authorities in this cluster clearly lack broadband resilience,
-for telecommuting or otherwise.
-Home to almost $3.5$ thousand people, the districts are scattered around the country:
-Ceredigion in West Wales, Darlington in the Northeast, Eden between the Lake District
-and the North Penines, Rossendale in the Northwest, and Rutland in the East Midlands.
-In comparison, clusters $3$ and $10$ have relatively high mean upload speeds, and 
-low standard deviations, suggesting reliability and resilience, 
-as well as quality broadband services, despite slower speeds in the morning peak.
-Like cluster $6$, is this also a sign of increased telecommuting?
-The eight LADs in clusters $3$ and $10$ are home to over one million people,
-and are all more urban and suburban than cluster $1$, including the London Borough 
-of Newham, the Southeast commuter towns of Harlow and Luton, the Birmingham suburbs 
-of Bromsgrove and Cannock Chase, Dundee City, and the towns of Eastbourne and Corby.
+Indeed cluster $12$ is the only cluster to experience higher speeds in the evening
+peak, compared to the morning peak, suggesting that widespread telecommuting
+has generally changed the temporal profile of internet activity throughout the UK.
+Yet even if all but one cluster is showing slower speeds in the morning than 
+the evening, the reliability of internet services in different clusters 
+during the working day still varies considerably.
+Interpreting this variation from the large spikes and dips shown on 
+Figures \ref{UpClusterL} and \ref{UpClusterS} is difficult, 
+but the statistics in Table \ref{up.cluster.descr} show that clusters $9$ and $11$ 
+have the most reliable internet services.
+The ratio of standard deviation to mean in both these clusters is below 55%, 
+and the ratio of upload speeds in the peak periods indicates that speeds are
+only about 2% slower in the morning.
+Mean speeds are also higher than in any other cluster, excluding cluster $2$,
+where measures of reliability suggests poorer performance.
 
-Meanwhile, the most resilient broadband service in terms of both quality and
-reliability appears to be found in clusters $9$ and $11$, home to over $12$ million people. 
-The mean speeds for these clusters are second and third highest, 
-the ratio of the standard deviation to the mean are the two lowest, and
-the experienced upload speeds in the morning peak are not that different 
-from those experienced in the evening peak.
+Thus, broadband services in clusters $9$ and $11$, home to over twelve million people 
+performed the best during the study period, in terms of both quality and reliability.
 In Figure \ref{UpClusterL}, cluster $11$ shows more noticeable peaks and troughs, 
-but the lowest points are not at the peak times described in Table $1$. 
+but the lowest points are not at the peak times described in Table \ref{up.cluster.descr}. 
 Rather, the slowest upload speeds on average occur between 6:00-7:00 on 
@@
 in cluster $13$.
-Although not all of the local authorities in these two clusters are urban, 
-among the $65$ LADs are Manchester and three of the nine other boroughs of 
-Greater Manchester, Glasgow, Nottingham, both Portsmouth and Southampton, 
-eight London Boroughs, four of the seven constituent authorities of the 
-West Midlands Combined Authority.
-There are also a number of other tightly bounded urban areas, such as Aberdeen, 
-Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at the centre
-of less confined districts like Burnley, Milton Keynes, and Northampton.
-The local authorities in these two clusters clearly experienced resilient
+Finally, ignoring the smallest clusters in terms of population, 
+clusters $3$, $7$ and $10$ also have relatively high mean upload speeds.
+Clusters $3$ and $10$ pair high mean speeds with low standard deviations 
+relative to the mean speeds, suggesting reliability and resilience, 
+as well as quality broadband services.
+Cluster $7$ has a higher ratio of standard deviation to mean, but there is less 
+difference in average speeds between the morning and evening peaks than in clusters
+$3$ and $10$.
+
+In summary, the local authorities in clusters $9$ and $11$ experienced resilient
 broadband that could support high levels of telecommuting.
-They are *not* on the wrong side of the first level digital divide, 
+Those in clusters $2$, $3$, $7$, and $10$ also experience higher than average 
+mean speeds and rank high to middle on measures of service realiability.
+These LADs are *not* on the wrong side of the first level digital divide, 
 but how likely are they to be able to take advantage of their resilient ICT
 infrastructure and services?
+Meanwhile, cluster $6$ is not only the largest in terms of number of LADs 
+and population, it has the closest mean upload speed to the pre-clustered average
+for the whole sample.
+As well as average quality internet services, those in cluster $6$ also experience 
+average reliability for work purposes, ranking fifth behind the four other clusters
+with populations over one million, but ahead of the smaller clusters.
+Clusters $8$ and $12$ are also close to average mean upload speeds, but show very 
+different patterns in terms of reliability, whilst clusters $1$ and $13$ appear to
+suffer most from a lack of quality internet services, with slow speeds and
+high standard deviations.
+With those in cluster $1$ in particular more likely to experience that poor 
+reliability during the morning peak, is this first level digital divide occurring
+in areas where few are occupationally able to telecommute anyway, and what are the
+implications for economic resilience?
 
@@
 
-Clusters $3$, $9$, $10$ and $11$ seem to benefit most from high quality and 
-resilient broadband services.
-The dips in mean upload speeds in clusters $6$, $3$ and $10$ during the morning
-peak are suggestive of more use during the working day, and potentially more
-telecommuting.
 Using auxiliary regressions, we test whether the clusters that have higher mean
@@
 and equation \ref{eq1} and each column represent a different cluster.
-The reference case is cluster $4$, which includes the local authority of Hambleton
-in North Yorkshire, a rural area of just over ninety thousand people.
+The reference case is cluster $4$, which includes only the local authority of 
+Hambleton in North Yorkshire, a rural area of just over ninety thousand people.
 Mean, experienced upload speeds in cluster $4$ are close to both the average speeds
 for the $13$ clusters ($9.9$Mb/s) and the pre-clustered average for the whole sample
-($9.3$Mb/s).
-The standard deviation for cluster $4$ and the ratio of evening to morning peak speeds
-are indications of worse reliability than many of the other clusters.
+($9.3$Mb/s) as shown in \ref{up.cluster.descr}.
+However, the standard deviation for cluster $4$ and the difference between average 
+speeds in the morning compared to the evening peak periods are indications of 
+worse reliability than many of the other clusters.
 Hence, the results in Table \ref{aux} should be seens as relative
@@
 and quality of service. 
-People in clusters $3$, $9$, $10$ and $11$ all ran fewer speed tests (at a per capita 
-basis) at this time of day than people in clusters $6$ and $13$.
-Those in clusters $3$, $9$, $10$ and $11$ also benefit from a higher proportion of
-fast Virgin cable connections than those in clusters $6$ and $13$. 
-Since services provided by Virgin Media are only available to $45$% of premises
-in the UK [@ofcom2016], 
-where the more lucrative ICT market originally attracted the cable TV provider, 
-this is an indication that people in these clusters are more likely to live on the 
-right side of the first level, infrastructure-based digital divide.
-Another indication is if people in these clusters are more likely to be closer to
-London or a metropolitan area.
+People in clusters $6$ and $13$ ran the most speed tests (at a per capita basis) 
+at this time of day, whilst people in cluster $11$ ran the fewest.
+This high level of satisfaction matches the high quality and reliability of service 
+we would expect from the analysis in [section](#sec:4.1).
+Those in clusters $2$, $3$, $9$ and $10$ also ran fewer speed tests per capita than
+those in the other six multi-authority clusters.
+Cluster $7$ was the only cluster with higher mean speeds and middle to high measures
+of reliability which was more likely to run more speed tests per capita.
+This might be a reflection of the choice of ISP and connection options in those
+LADs, as, along with cluster $12$, those in cluster $7$ were least likely to have
+Virgin Media connections.
+Fast Virgin Media cable connections have historically only been available 
+to $45$% of premises in the UK [@ofcom2016], where the more lucrative and 
+competative market originally attracted the cable TV provider. 
+Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
+Virgin connection, which is an indication that people in these clusters are 
+more likely to live in urban areas, with more choice of broadband services, 
+or in other words, are more likely to be on the right side of the first level, 
+infrastructure-based digital divide.
+
+We measure distance from London and from London or another metropolitan areas as 
+two more indicators of rurality and potentially first level digital divides. 
 The broadband speed tests run in the authorities in cluster $3$ are more likely to 
@@
 However, even though London was also included in the variable calculating distance 
-to one of the ten major metropolitan areas in England, or Glasgow or Cardiff, 
-tests run in cluster $3$ are likely to be furthest away. 
-**Did you include London in distMet as well? - code suggests so**
-**Do we need to re-think this variable (e.g. remove Cardiff, add Leicester or...)**
+from the centre of one of either the ten largest metropolitan areas in England, 
+Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away. 
+Still, whilst significant, the effect is tiny and the other members of cluster $3$
+are Corby and Eastbourne, two towns which, whilst not part of metropolitan areas,
+are also not in remote parts of the country, and are home to an estimated population 
+of $175$ thousand people between them. **move maps to here?**
+Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large 
+metropolitan area, even though the four local authorities of cluster $7$ include
+no central urban boroughs, and only one LAD that is part of a metropolitan area of 
+governance - Tameside in Greater Manchester.
+This may be why those in cluster $7$ are not served by Virgin Media.
+It also is a demonstration of the complexity of experienced broadband upload speeds
+as captured by time-series clustering, and their likely interaction with first level 
+digital divides.
+Cluster $1$, which our analysis suggests lacks broadband resilience, contains five, 
+mainly rural authorities. 
+However, they are scattered around the country --
+Ceredigion in West Wales, Darlington in the Northeast, Eden between the Lake District
+and the North Penines, Rossendale in the Northwest, and Rutland in the East Midlands
+-- and therefore, the results in Table \ref{aux} indicate that these authorities
+are not as far from the centre of a metropolitan area as authorities in cluster $3$, 
+but furthest from London.
+
+Internet resilience as measured here does not seem to follow the North-South 
+economic divide either, as broadly defined in our regression as being part of the
+Greater London, Southeast, Southwest and East of England regions.
+The authorities most likely to be in the South are those in cluster $13$,
+which was identified as having the slowest mean upload speeds of any of the clusters, 
+and thus a low level of service.
+However, cluster $13$ does include some rural, remote areas of the country, 
+such as Northwest Scotland, Cornwall and Powys in Wales as shown on Figure \ref{map.up.clusters}
+It also includes the major metropolitan centres of Bristol, Liverpool, 
+and Newcastle, as well as nine (of $32$) London Boroughs and plenty of home county 
+and suburban areas.
+Although by the standard deviation measure cluster $13$ has unreliable internet 
+services, speed variation was less during the morning peak.
 
 # Conclusions {#sec:5}
+
+Upload speeds are not used as benchmarks in the UK, nor have been seen as integral to 
+universal service, considering there has never before been such extreme demand for
+telecommuting and operations such as video calls.
+
+those in cluster $6$, including people living in Birmingham, Leeds and 
+Sheffield, in twelve London Boroughs, and many suburban areas and smaller cities like
+Cardiff, Oxford and Cambridge 
+This lower level of reliability may be as a result of increased demand, 
+e.g. for telecommuting, especially as there are few truly remote areas 
+in this cluster.
+
+The eight LADs in clusters $3$ and $10$ are home to over one million people,
+and are all more urban and suburban than cluster $1$, including the London Borough 
+of Newham, the Southeast commuter towns of Harlow and Luton, the Birmingham suburbs 
+of Bromsgrove and Cannock Chase, Dundee City, and the towns of Eastbourne and Corby.
+
+Although not all of the local authorities in these clusters $9$ and $11$ are urban, 
+among the $65$ LADs are Manchester and three of the nine other boroughs of 
+Greater Manchester, Glasgow, Nottingham, both Portsmouth and Southampton, 
+eight London Boroughs, four of the seven constituent authorities of the 
+West Midlands Combined Authority.
+There are also a number of other tightly bounded urban areas, such as Aberdeen, 
+Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at the centre
+of less confined districts like Burnley, Milton Keynes, and Northampton.
+
+Clusters $3$, $9$, $10$ and $11$ seem to benefit most from high quality and 
+resilient broadband services.
+The dips in mean upload speeds in clusters $6$, $3$ and $10$ during the morning
+peak are suggestive of more use during the working day, and potentially more
+telecommuting.
 

### 2020-12-21 paper/v1_sage/broadband.speed.covid.Rmd
new sage template. I need to change the path for the figures to lead to the new sage folder.

--- 2020-12-18 paper/v1/broadband.speed.covid.Rmd
@@
 ---
-title: |
-  WFH and broadband speed (title needs rework)
-type: ARTICLE TEMPLATE
+title: WFH and broadband speed (title needs rework)
 author:
-  - name: A. N. Author
-    affil: a
-    email: latex.helpdesk@tandf.co.uk
-  - name: John Smith
-    affil: b
-    email: john.smith@uni-freiburg.de
-affiliation:
-  - num: a
-    address: |
-      Taylor & Francis, 4 Park Square, Milton Park, Abingdon, UK
-  - num: b
-    address: |
-      Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany
-bibliography: bibliography.bib
-appendix: appendix.tex
-abstract: |
-  TBC
-keywords: |
-  covid; internet; working from home; broadband speed; time-series clusters
+- name: A. Uthor*
+  num: 1,2
+- name: O. Tro
+  num: 2
+- name: O. Vriga
+  num: 3
+address:
+- num: 1
+  org: Department of Incredible Research, University A, City A, Country A
+- num: 2
+  org: Department of Applied Things, University B, City B, Country B
+- num: 3
+  org: Very Important Stuff Committee, Institute C, City C, Country C
+corrauth: "Corresponding author name, This is sample corresponding address."
+email: correspondingauthor@protonmail.com
+abstract: "TBC"
+keywords:   covid; internet; working from home; broadband speed; time-series clusters;
+classoption:
+  - Royal
+  - times
+bibliography: bibliography
+bibliographystyle: sageh
 header-includes: |
@@
   \def\tightlist{}
-output: rticles::tf_article
+  \newcommand{\beginappendix}{\setcounter{table}{0} \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
+output:
+  rticles::sage_article:
+    keep_tex: yes
 ---
@@
 services can accommodate the increased demand.
-For example, $47$% of people in employment in the UK worked soley from home in April
+For example, $47$% of people in employment in the UK worked solely from home in April
 $2020$, whilst the same figure only reached $5$% the year before [@ons2020;
@@
 or repudiate other socio-economic and geographic patterns of economic resilience.
+The paper ends with a conclusions section.
 
@@
 interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
-Not only are all three levels of digital divides associated to a certain extent 
+Not only are all three levels of digital divides associated, to a certain extent,
 with or shaped by the geography of the UK, but the intersection of the digital and
@@
 
-##Time-Series Clustering {#sec:3.1}
+## Time-Series Clustering {#sec:3.1}
 <!-- clustering in general -->
@@
 
-**REMOVE THIS PARAGRAPH - IRRELEVANT DETAIL: Time-series clustering methods utilising the whole length of time-series can be grouped in three categories.
+<!-- Time-series clustering methods utilising the whole length of time-series can be grouped in three categories.
 The first -- model-based approaches -- is based on recovering the underlying model for each time-series and then applying clustering algorithms on the model parameters of each time-series [@aghabozorgi2015time].
 The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].The second approach is based the formation of vectors of features derived from the original time-series.
-These new data of reduced dimensionality are then clustered using conventional clustering algorithms.**
+These new data of reduced dimensionality are then clustered using conventional clustering algorithms.
+-->
 
@@
 Because of the simplicity of the implementation and the interpretability of the
-results, we utilise here partitioning clustering based on the widely used *k-Means*
+results, we utilise here partitioning clustering based on the widely used *k*-means
 algorithm.
@@
 There are a number of differences between the above described application of
-*k-Means* for cross-sectional data and its application for times series data.
+*k*-means for cross-sectional data and its application for times series data.
 Instead of creating clusters around centroids, a common approach is to create
@@
 
-##Experienced Broadband Speeds {#sec:3.2}
+## Experienced Broadband Speeds {#sec:3.2}
 <!-- introduction of our broadband speeds data -->
@@
 By using volunteered geographic data, we are able to assess the *experienced*
-internet speed by users, which may differ from the *advertised* maximum speeds of 
-Internet Service Providers (ISPs).
+internet speed by users, which may differ from the *advertised* maximum speeds of ISPs.
 
 We are particularly interested in upload speeds and the frequency of speed tests
-over the period from March to May 2020, as government statements indicate this is
+over the period from March to May $2020$, as government statements indicate this is
 when UK workers were told to work from home if at all possible [@GovUK2020].
@@
 As our research aims to identify the geography of internet service resilience for
-work purposes, bank holidays and the hours between midnight and 6:00 were excluded,
+work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
 as well as weekend days.
@@
 <!-- metrics -->
-In [section](#sec:4.1), we review the temporal profile of upload speed 
+In Section [4.1](#sec:4.1), we review the temporal profile of upload speed
 by hour of the day and day of the composite week, as well as
@@
 more likely to take place.
-We take account of all three measurements in our descriptives statistics of
+We take account of all three measurements in our descriptive statistics of
 upload speeds in order to determine how resilient broadband speeds are
@@
 socio-economic and geographic variables, which are discussed in detail in
-the relevant [section](#sec:4.2), in order to explore how the different patterns 
+the relevant Section [4.2]](#sec:4.2), in order to explore how the different patterns
 might support or undermine efforts to work from home and maintain safe productivity
@@
 However, this masks great variation in when lower speeds are experienced,
-with the mean upload speeds much lower between $9:00$-$10:59$ than 
-between $19:00$-$20:59$ in cluster $8$, but slightly faster in the morning in 
+with the mean upload speeds much lower between $9:00-10:59$ than
+between $19:00-20:59$ in cluster $8$, but slightly faster in the morning in
 cluster $12$.
@@
 but the lowest points are not at the peak times described in Table \ref{up.cluster.descr}.
-Rather, the slowest upload speeds on average occur between 6:00-7:00 on 
-Monday morning, 14:00-15:00 on Friday, and $16:00-17:00$ on Thursday. 
+Rather, the slowest upload speeds on average occur between $6:00-7:00$ on
+Monday morning, $14:00-15:00$ on Friday, and $16:00-17:00$ on Thursday.
 These slowest times are still mostly faster than the average hourly upload speeds
@@
 Finally, ignoring the smallest clusters in terms of population,
+that is clusters $4$ and $5$,
 clusters $3$, $7$ and $10$ also have relatively high mean upload speeds.
@@
 In summary, the local authorities in clusters $9$ and $11$ experienced resilient
-broadband that could support high levels of telecommuting.
+broadband internet that could support high levels of telecommuting.
 Those in clusters $2$, $3$, $7$, and $10$ also experience higher than average
-mean speeds and rank high to middle on measures of service realiability.
+mean speeds and rank high to middle on measures of service reliability.
 These LADs are *not* on the wrong side of the first level digital divide,
@@
 
-## aux regressions {#sec:4.2}
+## Post-clustering regression analysis {#sec:4.2}
 
@@
 Using auxiliary regressions, we test whether the clusters that have higher mean
-speeds and more reliable services are more urban, more wealthy, and / or more likely 
+speeds and more reliable services consist of LADs that are more urban, affluent,
+and / or more likely
 to benefit from a choice of high quality internet services.
-We also measure which of our clusters are more likely to have a higher proportion of 
+We also estimate which of our clusters are more likely to have a higher proportion of
 occupations which enable telecommuting because of the nature of the work.
@@
 The dependent variable is the LAD cluster membership as described in the
-[Methods and data](#sec:3) section
+[methods and data](#sec:3) section
 and equation \ref{eq1} and each column represent a different cluster.
@@
 worse reliability than many of the other clusters.
-Hence, the results in Table \ref{aux} should be seens as relative
+Hence, the results in Table \ref{aux} should be seen as relative
 rather than absolute probabilities.
 
-First, the number of speed tests run per cluster inhabitant between 9:00-10:59
+First, the number of speed tests run per cluster inhabitant between $9:00$-$10:59$
 is an indication of satisfaction or at least a lack of concern over broadband speeds
 and quality of service.
-People in clusters $6$ and $13$ ran the most speed tests (at a per capita basis) 
-at this time of day, whilst people in cluster $11$ ran the fewest.
+People in LADs assigned to clusters $6$ and $13$ were more likely to ran speed tests
+(at a per capita basis)
+at this time of day, whilst people in cluster $11$ were the least likely ones.
 This high level of satisfaction matches the high quality and reliability of service
-we would expect from the analysis in [section](#sec:4.1).
-Those in clusters $2$, $3$, $9$ and $10$ also ran fewer speed tests per capita than
-those in the other six multi-authority clusters.
+we would expect from the analysis in Section [4.1]](#sec:4.1).
+Those in clusters $2$, $3$, $9$ and $10$ were also less likely to ran speed tests per
+capita than those in the other six multi-authority clusters.
 Cluster $7$ was the only cluster with higher mean speeds and middle to high measures
@@
 to $45$% of premises in the UK [@ofcom2016], where the more lucrative and
-competative market originally attracted the cable TV provider. 
+competitive market originally attracted the cable TV provider.
 Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
-Virgin connection, which is an indication that people in these clusters are 
+Virgin connections, which is an indication that people in these clusters are
 more likely to live in urban areas, with more choice of broadband services,
@@
 
-We measure distance from London and from London or another metropolitan areas as 
-two more indicators of rurality and potentially first level digital divides. 
+We employ distance from London and for the nearest metropolitan area (including
+London) as two more variables depicting peripherality, urban structure and,
+potentially, first level digital divides.
 The broadband speed tests run in the authorities in cluster $3$ are more likely to
@@
 
-Internet resilience as measured here does not seem to follow the North-South 
-economic divide either, as broadly defined in our regression as being part of the
-Greater London, Southeast, Southwest and East of England regions.
+Internet resilience, as measured here, seems to be more nuanced than our --
+arguably crude -- dummy variable depicting the North-South economic divide either,
+which assigns $1$ to LADs locates in Greater London, Southeast, Southwest and East
+of England regions.
 The authorities most likely to be in the South are those in cluster $13$,
@@
 However, cluster $13$ does include some rural, remote areas of the country,
-such as Northwest Scotland, Cornwall and Powys in Wales as shown on Figure \ref{map.up.clusters}
+such as Northwest Scotland, Cornwall and Powys in Wales as shown on Figure
+\ref{map.up.clusters}.
 It also includes the major metropolitan centres of Bristol, Liverpool,
@@
 services, speed variation was less during the morning peak.
+More broadly, the estimates for this variable are not consistent between the
+two clubs of LAD clusters: those who are in the wrong and the right side of
+digital divides.
+
+Regarding the rest of the socio-economic variables we see a varied
+picture. Clusters $9$ and $11$, which are the ones benefiting by the most resilient
+internet connections are likely to consist of LADs with higher share of managerial
+and highly skilled occupations as well as higher average earnings. Interestingly,
+both clusters are characterised by lower share of occupations that are not usually
+associated with teleworking (professional, associated professional and technical,
+administrative, leisure and machine operation jobs), but also by lower number of
+business per habitant and. However, the likelihood of hosting highly skilled
+residents is not consistent with cluster $11$ having a negative coefficient for the
+NVQ4 variable. This might be an indication of the higher returns that resilient
+broadband infrastructure generates for cluster $9$ versus cluster $11$. While both
+of these cluster consist more likely by large LADs in terms of population, job
+density if higher for LADs in cluster $11$ indicating more labour intensive
+economic structures contrary to LADs in cluster $9$.
+
+A similar divided picture can be built for the next tier of clusters: $2$, $3$,
+$7$, $10$. LADs in these clusters experience higher than average mean speeds and
+rank high to middle on measures of service reliability. As mentioned before, LADs
+in these clusters are on the right side of the first level digital divide. However,
+while all of them are more likely to be populated by people employed in managerial
+jobs, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
+work force, which can more easily take advantage of working from home opportunities.
+A more mixed picture can be built for LADs in clusters $3$ and $7$, which are
+associated, on the one hand, with lower skills and, on the other hand, with higher
+likelihood of individuals in managerial occupations. A similar contradicting
+pattern can be illustrated for cluster $6$, which is the largest one with average 
+quality of internet service. Arguably, LADs in clusters $2$ and $10$ are the ones
+benefiting more by resilient internet infrastructure in an era that working
+from home became a vehicle for economic resilience.
+
+Moving on to the next tier of clusters $8$ and $12$, which are characterised by
+LADs enjoying close to average mean upload speeds, but different patterns of internet
+reliability, we see a differentiated picture. On the one hand, both of them are
+likely to host individuals employed in managerial occupations while the opposite
+applies on skilled, professional and leisure occupations. On the other hand,
+LADs in cluster $8$ are characterised by a very small positive and LADs in cluster
+$12$ by a negative likelihood to host highly skilled individuals. Moreover, LADs
+in these two clusters have a rather
+similar profile regarding the occupational variables, but different tendencies
+regarding the stock of businesses. In total, LADs in these clusters do not
+benefit by great internet services and could only marginally benefit by working
+from home opportunities.
+
+Lastly, clusters $1$ and $13$, which appear to suffer most from a lack of quality
+internet services are more likely to also suffer from low skilled workforce. LADs
+in cluster $13$. Contradictory, the likelihood of workforce employed in managerial
+occupations is positive for LADs in both clusters, but only individuals in LADs
+in cluster $13$ are more likely to enjoy high earnings.
 
@@
 use and urban planning to accommodate people who work from home
-[@BUDNITZ2020102713]**also 2020 Swedish article from JTG**; and from productivity and innovation changes, to changes
+[@BUDNITZ2020102713]**also 2020 Swedish article from JTG**; and from productivity 
+and innovation changes, to changes
 in agglomeration externalities and the attraction of large cities [@econobs]
@@
 
-# Funding {-}
+# Appendix {#sec:appendix}
 
-An unnumbered section, e.g.\ `\section*{Funding}`, may be used for grant details, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.
+```{r echo=FALSE, message=FALSE, results = "asis"}
 
-# References
+# Cluster table
+clusters <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
+clusters <- read_csv(clusters)
+clusters <- clusters %>%
+  dplyr::select(LAD, cluster.up) %>%
+  rename('cluster membership' = cluster.up) 
+  #group_by('cluster membership') %>%
+  #dplyr::arrange()
+  #sort('cluster membership', decreasing = F)
+
+stargazer(clusters,
+          summary = F,
+          font.size = "footnotesize",
+          type = "latex",
+          header = F,
+          title = "LAD cluster membership for upload speed timeseries\\label{cluster.membership}")  
+```
+
+```{r echo=FALSE, message=FALSE, results = "asis"}
+
+# Descriptive statistics table
+path.LAs_Clusters <- paste0(path, "/data/temp/LAs_Clusters.RData")
+load(path.LAs_Clusters)
+
+for.descr <- data %>%
+  dplyr::select(pop, job.dens2018, distMet,
+                distLondon, south, 
+                managers, tech, skilled, prof,
+                admin, caring, plant, 
+                earnings, nvq4,
+                virgin,
+                total.busi, LAD.tests.am20, pop) %>%
+  mutate('total.busi/pop' = total.busi/pop,
+         'LAD.tests.am20/pop' = LAD.tests.am20/pop) %>%
+  dplyr::select(-total.busi, -pop, -LAD.tests.am20)
+
+stargazer(for.descr,
+          summary = T,
+          font.size = "footnotesize",
+          type = "latex",
+          header = F,
+          digit.separator = ",",
+          digit.separate = c(3,3),
+          covariate.labels=c("pop, 2018",
+                             "job density, 2018",
+                             "distance to nearest met. area",
+                             "distance to London",
+                             "South of the UK",
+                             "managerial jobs, 2020",
+                             "tech jobs, 2020",
+                             "skilled trade jobs, 2020",
+                             "professional jobs, 2020",
+                             "administrative jobs, 2020",
+                             "leisure jobs, 2020",
+                             "machine operation jobs, 2020",
+                             "earnings, 2019",
+                             "NVQ4+",
+                             "Virgin Media \\%, 2020",
+                             "n. business est. per hab., 2019",
+                             "AM tests per hab., 2020"),
+          rownames = F,
+          no.space=TRUE,
+          column.sep.width = "0pt",
+          #notes = "",
+          title = "Descriptive statistics for the auxiliary regression explanatory variables\\label{descr.aux}")
+
+```

### 2020-12-21 paper/v1_sage/broadband.speed.covid.tex
new sage template. I need to change the path for the figures to lead to the new sage folder.

--- 2020-12-15 paper/v1/broadband.speed.covid.tex
@@
-% interactcadsample.tex
-% v1.03 - April 2017
-
-\documentclass[]{interact}
-
-\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
-\usepackage{subfigure}% Support for small, `sub' figures and tables
-%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
-
-\usepackage{natbib}% Citation support using natbib.sty
-\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
-\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
-
-\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{example}[theorem]{Example}
-
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-\newtheorem{notation}{Notation}
-
-% see https://stackoverflow.com/a/47122900
+\documentclass[Royal,times,sageh]{sagej}
+
+\usepackage{moreverb,url,natbib, multirow, tabularx}
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 
@@
 \def\tightlist{}
+\newcommand{\beginappendix}{\setcounter{table}{0} \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
 
@@
 
-\articletype{ARTICLE TEMPLATE}
-
 \title{WFH and broadband speed (title needs rework)}
 
-\author{\name{A. N. Author$^{a}$, John Smith$^{b}$}
-\affil{$^{a}$Taylor \& Francis, 4 Park Square, Milton Park, Abingdon, UK; $^{b}$Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany}
-}
-
-\thanks{CONTACT A.
-N.
-Author.
-Email: \href{mailto:latex.helpdesk@tandf.co.uk}{\nolinkurl{latex.helpdesk@tandf.co.uk}}, John Smith.
-Email: \href{mailto:john.smith@uni-freiburg.de}{\nolinkurl{john.smith@uni-freiburg.de}}}
-
-\maketitle
+\runninghead{}
+
+\author{A. Uthor*\affilnum{1,2}, O. Tro\affilnum{2}, O. Vriga\affilnum{3}}
+
+\affiliation{\affilnum{1}{Department of Incredible Research, University A, City A, Country A}\\\affilnum{2}{Department of Applied Things, University B, City B, Country B}\\\affilnum{3}{Very Important Stuff Committee, Institute C, City C, Country C}}
+
+\corrauth{Corresponding author name, This is sample corresponding address.}
+
+\email{\href{mailto:correspondingauthor@protonmail.com}{\nolinkurl{correspondingauthor@protonmail.com}}}
 
@@
 
-\begin{keywords}
-covid; internet; working from home; broadband speed; time-series
-clusters
-\end{keywords}
+\keywords{covid; internet; working from home; broadband speed; time-series
+clusters;}
+
+\maketitle
 
@@
 Internet Service Providers (ISPs), at least in the UK \citep{ofcom2017}.
+
 Yet the shift towards telecommuting during various stages of lockdown
@@
 accommodate the increased demand. For example, \(47\)\% of people in
-employment in the UK worked soley from home in April \(2020\), whilst
+employment in the UK worked solely from home in April \(2020\), whilst
 the same figure only reached \(5\)\% the year before
@@
 resilience of people and places to overcome a systemic shock such as the
-current pandemic. We identify this as a new digital divide, one that
+current pandemic. The extreme level of demand for telecommuting
 fundamentally alters the potential returns of internet use for the user
@@
 some occupations but not in others, and relies upon access to high
-quality internet services. \textbf{WE NEED TO CONNECT OUR FINDINGS WITH
-THIS AND JUSTIFY THAT THIS IS INDEED A NEW DIVIDE} As the quality of
-internet infrastructure and services, as well as variation in
-occupations are spatially dependent and clustered in space, our approach
-offers a framework for understanding which types of places, are more
-likely to land on the right side of this new digital divide. By asking
-how resilient broadband speeds are as experienced in different parts of
-the UK during a time of extreme demand, we interrogate which places
-benefit from the greater economic resilience digital technologies can
-offer, not only during the pandemic, but also into the future.
-
-The structure of this paper is as follows. First we review the
-literature on telecommuting and digital divides to better understand the
-origins of the new digital divide revealed by the pandemic and its
-impact on the economic resilience of different places. We then describe
-our data and methodology. Our results section first offers a
-classification of how internet services vary across the UK local
-authorities and then assesses whether these clusters replicate or
-repudiate other socio-economic and geographic patterns of economic
-resilience.
+quality internet services.
+
+As the quality of internet infrastructure and services, as well as
+variation in occupations are spatially dependent and clustered in space,
+our approach offers a framework for understanding the impact of and
+interactions between the different levels of digital division in
+different places with different characteristics. By asking how resilient
+broadband speeds, and particularly upload speeds are as experienced in
+different parts of the UK during a time of extreme demand, we
+interrogate which places benefit from the greater economic resilience
+digital technologies can offer, not only during the pandemic, but also
+into the future. The structure of this paper is as follows. First we
+review the literature on telecommuting and digital divides to better
+understand the structural and spatial development of these practices
+pre-pandemic, and thus their importance to the economic resilience of
+different places. We then describe our data and methodology. Our results
+section first offers a classification of how internet services vary
+across the UK local authorities and then assesses whether these clusters
+replicate or repudiate other socio-economic and geographic patterns of
+economic resilience. The paper ends with a conclusions section.
 
@@
 \citep{headicar2016move, peters2004employees, singh2013modeling}.
+
 Opportunities for working from home during the current pandemic have
@@
 divide in infrastructure quality between urban and rural areas in
-various advanced economies. Yet whether this variation in infrastructure
-quality affects the spatial footprint of telecommuting has not
-previously been measured. There are indications that those who purchase
-high speed connections consume more data of all sorts and use their
-connections for a variety of purposes \citep{hauge2011consumer}, and
-that there is a correlation between access to internet services and a
-reduction in household transport spend \citep{bris2017ict}. Whether the
-implication is more internet use and less travel because of increased
-telecommuting, these studies suggest that better internet services
-enable households to make savings and efficiencies, an example of the
-first level digital divide reinforcing the third level. Yet did the
-purchase of high speed connections and increased internet access also
-prepare households for long-term home-working, enforced by government
-restrictions? The extreme demand during the pandemic provides a new
-opportunity to understand how infrastructure accessibility, quality, and
-reliability affects telecommuting, particularly in light of the high
-volumes of bandwidth-intensive video conferencing required in order to
-avoid the face-to-face contact that could increase the spread of
-infection. We seek to answer how internet service resilience contributes
-to or reduces economic resilience when the latter is dependent upon the
-capability to work from home, and thus improve our understanding of this
-new digital divide. \textbf{RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE
-THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL
-QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL
-DIVIDES? ET: OR REINFORCING THE 3RD LEVEL DIGITAL DIVIDE}
-
-The multi-layered digital divides intersect with material divides and
-the economic geography of the UK. Following the regional economic
-resilience literature, which underlines the differentiated capacity of
-cities and regions to escape or recover from economic crises
+various advanced economies. Rural areas, predictably, fare worse. Yet
+whether this variation in infrastructure quality affects the spatial
+footprint of telecommuting has not previously been measured, in part
+because telecommuting has not previously been the cause of greatest
+demand and pressure on internet services.
+
+There are indications that those who purchase high speed connections
+consume more data of all sorts and use their connections for a variety
+of purposes \citep{hauge2011consumer}, and that there is a correlation
+between access to internet services and a reduction in household
+transport spend \citep{bris2017ict}. Whether the implication is more
+internet use and less travel because of increased telecommuting, these
+studies suggest that better internet services enable households to make
+savings and efficiencies, an example of the first level digital divide
+reinforcing the third level. Yet did the purchase of high speed
+connections and increased internet access also prepare households for
+long-term home-working, enforced by government restrictions? The extreme
+demand during the pandemic provides a new opportunity to understand how
+infrastructure accessibility, quality, and reliability affects
+telecommuting, particularly in light of the high volumes of
+bandwidth-intensive video conferencing required in order to avoid the
+face-to-face contact that could increase the spread of infection. We
+seek to answer how internet service resilience contributes to or reduces
+economic resilience when the latter is dependent upon the capability to
+work from home. We also aim to improve our understanding of the impact
+of first level digital division on telecommuting, and whether this
+results in much more fundamental third level digital division than has
+previously been perceived.
+
+Furthermore, these multi-layered digital divides intersect with material
+divides and the economic geography of the UK. Following the regional
+economic resilience literature, which underlines the differentiated
+capacity of cities and regions to escape or recover from economic crises
 \citep{martin2012regional, kitsos2018economic}, different places have
@@
 \citep{martin_north_south}. Various studies have illustrated severe
-inequalities between the north and the south part of the UK regarding,
-for example, skills and human capital, unemployment, productivity and
-prosperity \citep{lee2014grim, mccann2020perceptions, dorling2018peak}.
-Some scholars have even argued that the UK suffers some of the highest
-level of interregional inequalities in the global north
+inequalities between the north and the south regions of England
+\textbf{replace `UK'} regarding, for example, skills and human capital,
+unemployment, productivity and prosperity
+\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
+scholars have even argued that the UK suffers some of the highest level
+of interregional inequalities in the global north
 \citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
-digital divides associated to a certain extent with or shaped by the
+digital divides associated, to a certain extent, with or shaped by the
 geography of the UK, but the intersection of the digital and material
@@
 \section{Methods and data}\label{sec:3}}
+
+\hypertarget{sec:3.1}{%
+\subsection{Time-Series Clustering}\label{sec:3.1}}
 
@@
 clustering problems -- is data dimensionality given the multiplicity of
-data points for every individual object -- local authorities in our case
--- included in the data set. Time-series are dynamic data as the value
-of the observations change as a function of time
+data points for every individual object, local authorities in our case,
+included in the data set. Time-series are dynamic data as the value of
+the observations change as a function of time
 \citep{aghabozorgi2015time}. This high dimensionality leads to (i)
@@
 
-Time-series clustering methods utilising the whole length of time-series
-can be grouped in three categories. The first -- model-based approaches
--- is based on recovering the underlying model for each time-series and
-then applying clustering algorithms on the model parameters of each
-time-series \citep{aghabozorgi2015time}. The main criticism is the
-cluster accuracy for nearby clusters \citep{mitsa2010temporal}. The
-second approach is based the formation of vectors of features derived
-from the original time-series. These new data of reduced dimensionality
-are then clustered using conventional clustering algorithms.
-
-For this paper we utilise the third category of time-series clustering
-methods known as shape-based approaches. These methods match two sperate
+For this paper we utilise a category of time-series clustering methods
+known as shape-based approaches. These methods match two sperate
 time-series objects based on the similarity of their shapes through the
@@
 implementation and the interpretability of the results, we utilise here
-partitioning clustering based on the widely used \emph{k-Means}
+partitioning clustering based on the widely used \emph{k}-means
 algorithm. This is an iterative algorithm, which begins with defining
@@
 There are a number of differences between the above described
-application of \emph{k-Means} for cross-sectional data and its
+application of \emph{k}-means for cross-sectional data and its
 application for times series data. Instead of creating clusters around
@@
 
+\hypertarget{sec:3.2}{%
+\subsection{Experienced Broadband Speeds}\label{sec:3.2}}
+
 To assess the quality and reliability of internet across local
@@
 individual internet speed tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
-This a private company that allows internet users to check their own
+This is a private company that allows internet users to check their own
 broadband upload and download speeds, and stores every speed-check with
@@
 are able to assess the \emph{experienced} internet speed by users, which
-may differ from the \emph{advertised} maximum speeds of Internet Service
-Providers (ISPs). We are particularly interested in upload speeds and
-the frequency of speed tests over the period from March to May 2020, as
-government statements indicate this is when UK workers were told to work
-from home if at all possible \textbf{{[}need ref?{]}}. Average upload
-speeds are slower than average download speeds, at \(9.3\)Mb/s mean
-upload speed for the whole sample during the period of interest,
-compared to \(29.6\)Mb/s for download speeds, but they are also less
-associated with internet-based, high-demand, leisure activities such as
-video streaming. Therefore, upload speeds are more relevant to
-work-related activities such as uploading documents or two-way audio,
-video, and text-based communication systems. Meanwhile, the frequency of
-speed tests was important in identifying the temporal profile which woud
-give us most insight into experienced internet service and resilience,
-and provide an indication of the volume of experience over particular
-units of time.
+may differ from the \emph{advertised} maximum speeds of ISPs.
+
+We are particularly interested in upload speeds and the frequency of
+speed tests over the period from March to May \(2020\), as government
+statements indicate this is when UK workers were told to work from home
+if at all possible \citep{GovUK2020}. Average upload speeds are slower
+than average download speeds, at \(9.3\)Mb/s mean upload speed for the
+whole sample during the period of interest, compared to \(29.6\)Mb/s for
+download speeds, but they are also less associated with internet-based,
+high-demand, leisure activities such as video streaming. Therefore,
+upload speeds are more relevant to work-related activities such as
+uploading documents or two-way audio, video, and text-based
+communication systems. Meanwhile, the frequency of speed tests was
+important in identifying the temporal profile which would give us most
+insight into experienced internet service and resilience, and provide an
+indication of the volume of experience over particular units of time.
 
@@
 terms of space and time. The frequency of testing indicates that whilst
-there is an overall trend of increased testing \textbf{or increased
-experience of internet service NOT SURE WHAT THIS MEAND} from March to
-April and then a slight reduction from April to May, this trend masks
-substantial variation by not only the day of the week, but also time of
-day, as can be seen in Figure \ref{test2020}. Thus, a daily aggregation
-of upload speeds would mask the variation in experience over the course
+there is an overall trend of increased testing from March to April and
+then a slight reduction from April to May, this trend masks substantial
+variation by not only the day of the week, but also time of day, as can
+be seen in Figure \ref{test2020}. Thus, a daily aggregation of upload
+speeds would mask the variation in experienced service over the course
 of each weekday. Furthermore, the importance of this variation is
@@
 reliability, we decided to include a measure of hourly variation in our
-temporal profiles. However, although this is a large data set
-(\(241,088\) individual tests during the working hours of study period),
-there are not enough observations for each Local Authority District
-(LAD) and for each working hour of each working day (\(631\) speed tests
-per LAD on average) to profile speeds at that level of detail.
-Therefore, we aggregate all the speed-checks during the \(13\) weeks of
-March to May inclusive for weekdays in \(2020\) by each hour of the day
-and day of the week. As our research aims to identify the geography of
-internet service resilience for work purposes, bank holidays and the
-hours between midnight and 6:00 were excluded, as well as weekend days.
+temporal profiles.
+
+However, although this is a large data set -- \(241,088\) individual
+tests performed during weekday hours across the study period -- there
+are not enough observations for each Local Authority District (LAD) and
+for each working hour of each working day -- \(631\) speed tests per LAD
+on average -- to profile speeds at that level of detail. Therefore, we
+aggregate all the speed-checks during the \(13\) weeks of March to May
+inclusive for weekdays in \(2020\) by each hour of the day and day of
+the week. As our research aims to identify the geography of internet
+service resilience for work purposes, bank holidays and the hours
+between midnight and \(6:00\) were excluded, as well as weekend days.
 The composite week time-series thus comprise \(18\) hours multiplied by
@@
 
-We then review the characteristics of each cluster, including number of
-LADs, and the temporal profile of upload speed by hour of the day and
-day of the composite week. Since the quality and reliability of internet
-services vary in time and space due to both supply and demand-side
-influences, we use a number of different measures to then describe
-experienced upload speeds per cluster. These include: a) mean,
-experienced connection speed, b) standard deviation or the amount of
-fluctuation from the mean, and c) the variation in speeds at particular
-times of day when working from home is more likely to take place. We
-take account of all three measurements in our descriptives statistics of
-upload speeds in order to determine how resilient broadband speeds are
-as experienced in different parts of the UK during a time of extreme
-demand.
-
-\textbf{More data details and some figures, descriptive stats}
+In Section \protect\hyperlink{sec:4.1}{4.1}, we review the temporal
+profile of upload speed by hour of the day and day of the composite
+week, as well as the experienced speed characteristics of each cluster.
+Since the quality and reliability of internet services vary in time and
+space due to both supply and demand-side influences, we use a number of
+different measures to describe experienced upload speeds per cluster.
+These include: a) mean, experienced connection speed, b) standard
+deviation or the amount of fluctuation from the mean, and c) the
+variation in speeds at particular times of day when working from home is
+more likely to take place. We take account of all three measurements in
+our descriptive statistics of upload speeds in order to determine how
+resilient broadband speeds are as experienced in different parts of the
+UK during a time of extreme demand.
 
@@
 regression in order to understand how the spatial and temporal patterns
-of internet service relate to the economic geography of the UK. We
-discuss how the different patterns might support or undermine efforts to
-work from home and maintain safe productivity and whether they reinforce
-existing spatial and social inequalities. More specifically, we estimate
-the following multinomial model:
+of internet service relate to the economic geography of the UK. More
+specifically, we estimate the following multinomial model:
 
@@
 
-The identification strategy is as follows. Based on the outcomes of the
-time-series clustering, we identify \(J\) distinct and crisp clusters.
-We then regress this cluster membership against a vector \(X_{i}\) of
-socio-economic and geographic variables, which are discussed in detail
-in the relevant \protect\hyperlink{sec:4.2}{section}. This analysis
-enables us to provide a more nuanced understanding of how telecommuting
-and technology intersect at a time of extreme demand, and what lessons
-this time has for a future where telecommuting is likely to remain a
-common means of accessing work and broadband services, as well as
-infrastructure, must be fit for purpose.
+Based on the outcomes of the time-series clustering, we identify \(J\)
+distinct and crisp clusters. We then regress this cluster membership
+against a vector \(X_{i}\) of socio-economic and geographic variables,
+which are discussed in detail in the relevant Section
+{[}4.2{]}{]}(\#sec:4.2), in order to explore how the different patterns
+might support or undermine efforts to work from home and maintain safe
+productivity and whether they reinforce existing spatial and social
+inequalities. This analysis enables us to provide a more nuanced
+understanding of how telecommuting and technology intersect at a time of
+extreme demand, and what lessons this time has for a future where
+telecommuting is likely to remain a common means of accessing work and
+broadband services, as well as infrastructure, must be fit for purpose.
 
@@
 upload speeds per hour per day for each cluster, with the largest, in
-terms of the LAD membership, five clusters in Figure \ref{UpClusterL},
-and the next six in Figure \ref{UpClusterS}. These figures and table
-provide a comprehensive overview of the quality and reliability of
-experienced broadband in different parts of the UK, the temporal
-clusters offering a novel approach to understanding spatial disparity.
+terms of the LAD membership and population, five clusters in Figure
+\ref{UpClusterL}, and the next six in Figure \ref{UpClusterS}. These
+figures and table provide a comprehensive overview of the quality and
+reliability of experienced broadband in different parts of the UK, the
+temporal clusters offering a novel approach to understanding spatial
+disparity.
 
@@
 
-The largest two clusters, comprising \(281\) local authorities and
-almost \(50\) million people, fall into clusters \(6\) and cluster
-\(13\). Cluster \(13\) has one of the lowest aggregate mean upload speed
-of any of the clusters, and the second highest ratio of the standard
-deviation to the mean. This suggests that those living in local
-authorities in this cluster experienced some of the lowest quality
-broadband services in terms of upload speeds in the UK. However, the
-variation in upload speeds in cluster \(13\), which can be an indication
-of its reliability, does not seem to disproportionately affect the
-morning peak from 9:00-10:59, as upload speeds are, on average, only
-\(2.6\)\% slower than in the evening peak period between 19:00 and
-20:59. Although cluster \(13\) does include some rural, remote areas of
-the country, such as Northwest Scotland, Cornwall and Powys in Wales as
-shown on Figure \ref{map.up.clusters}, it also includes the major
-metropolitan centres of Bristol, Liverpool, and Newcastle, as well as
-nine (of \(32\)) London Boroughs and plenty of home county and suburban
-areas. Meanwhile, those living in cluster \(6\) experience aggregate
-mean upload speeds of about \(1\)Mb/s faster than those in cluster
-\(13\), but still lower than the other three large clusters and most of
-the smaller clusters, suggesting a middling quality of service. However,
-the time profile for cluster \(6\) in Figure \(3\)a shows that upload
-speeds drop quickly from \(6:00\) to 7:00 on a Monday and peak between
-23:00 and midnight on Wednesday and Thursday, but tend to be lower
-during the working day. Indeed, those in cluster \(6\), including people
-living in Birmingham, Leeds and Sheffield, in \(12\) London Boroughs,
-and many suburban areas and smaller cities like Cardiff, Oxford and
-Cambridge experience mean upload speeds in the morning that are
-\(4.4\)\% lower than in the evening. This lower level of reliability may
-be as a result of increased demand, e.g.~for telecommuting, especially
-as there are few truly remote areas in this cluster.
+The second largest cluster, comprising \(126\) local authorities and
+over \(20\) million people, is cluster \(13\). Cluster \(13\) has the
+slowest aggregate mean upload speed of any of the clusters, and the
+second highest ratio of the standard deviation to the mean. This
+suggests that those living in local authorities in this cluster
+experienced some of the lowest quality broadband services in terms of
+upload speeds in the UK. However, the variation in upload speeds in
+cluster \(13\), which can be an indication of its reliability, does not
+seem to disproportionately affect the morning peak from
+\(9:00\)-\(10:59\), as upload speeds are, on average, only \(2.6\)\%
+slower than in the evening peak period between \(19:00\) and \(20:59\),
+when entertainment purposes are likely to be using the most bandwidth.
+In comparison, the five LADs that are home to \(343\) thousand people in
+cluster \(1\) not only experience the second slowest mean upload speeds
+and the highest ratio of standard deviation to the mean, but are also
+much more affected during the morning peak.
+
+Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
+LADs home to \(29.5\) million people, experience aggregate mean upload
+speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
+lower than the other three large clusters and most of the smaller
+clusters, suggesting a middling quality of service. However, the time
+profile for cluster \(6\) in Figure \ref{UpClusterL} shows that upload
+speeds drop quickly from \(6:00\) to \(7:00\) on a Monday and peak
+between \(23:00\) and midnight on Wednesday and Thursday, but tend to be
+lower during the working day. Indeed, experienced mean upload speeds in
+the morning peak are \(4.4\)\% lower than in the evening peak -- a
+greater, more noticeable change than any of the other large clusters
+experience, but smaller than any of the clusters with temporal profiles
+shown in Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have
+mean upload speeds under \(10\)Mb/s, but higher than clusters \(1\) and
+\(13\), and their standard deviation is not dissimilar. However, this
+masks great variation in when lower speeds are experienced, with the
+mean upload speeds much lower between \(9:00-10:59\) than between
+\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning in
+cluster \(12\).
 
@@
 
-However, whilst the difference in experienced mean upload speeds between
-9:00-10:59 and 19:00-20:59 in cluster \(6\) is greater than in any of
-the other large clusters, it is smaller than this difference in all the
-clusters shown in Figure \ref{UpClusterL}. Whilst this may suggest a
-lack of reliability, especially considering the large spikes and dips in
-these clusters, the statistics in Table \ref{up.cluster.descr} show a
-more varied picture. The experienced upload speeds in cluster \(1\) are
-the second slowest of any cluster, the ratio of standard deviation to
-mean is the highest, as is the ratio of speeds in the morning compared
-to the evening peak hours analysed. The five, mainly rural authorities
-in this cluster clearly lack broadband resilience, for telecommuting or
-otherwise. Home to almost \(3.5\) thousand people, the districts are
-scattered around the country: Ceredigion in West Wales, Darlington in
-the Northeast, Eden between the Lake District and the North Penines,
-Rossendale in the Northwest, and Rutland in the East Midlands. In
-comparison, clusters \(3\) and \(10\) have relatively high mean upload
-speeds, and low standard deviations, suggesting reliability and
-resilience, as well as quality broadband services, despite slower speeds
-in the morning peak. Like cluster \(6\), is this also a sign of
-increased telecommuting? The eight LADs in clusters \(3\) and \(10\) are
-home to over one million people, and are all more urban and suburban
-than cluster \(1\), including the London Borough of Newham, the
-Southeast commuter towns of Harlow and Luton, the Birmingham suburbs of
-Bromsgrove and Cannock Chase, Dundee City, and the towns of Eastbourne
-and Corby.
-
-Meanwhile, the most resilient broadband service in terms of both quality
-and reliability appears to be found in clusters \(9\) and \(11\), home
-to over \(12\) million people. The mean speeds for these clusters are
-second and third highest, the ratio of the standard deviation to the
-mean are the two lowest, and the experienced upload speeds in the
-morning peak are not that different from those experienced in the
-evening peak. In Figure \ref{UpClusterL}, cluster \(11\) shows more
-noticeable peaks and troughs, but the lowest points are not at the peak
-times described in Table \(1\). Rather, the slowest upload speeds on
-average occur between 6:00-7:00 on Monday morning, 14:00-15:00 on
+Indeed cluster \(12\) is the only cluster to experience higher speeds in
+the evening peak, compared to the morning peak, suggesting that
+widespread telecommuting has generally changed the temporal profile of
+internet activity throughout the UK. Yet even if all but one cluster is
+showing slower speeds in the morning than the evening, the reliability
+of internet services in different clusters during the working day still
+varies considerably. Interpreting this variation from the large spikes
+and dips shown on Figures \ref{UpClusterL} and \ref{UpClusterS} is
+difficult, but the statistics in Table \ref{up.cluster.descr} show that
+clusters \(9\) and \(11\) have the most reliable internet services. The
+ratio of standard deviation to mean in both these clusters is below
+55\%, and the ratio of upload speeds in the peak periods indicates that
+speeds are only about 2\% slower in the morning. Mean speeds are also
+higher than in any other cluster, excluding cluster \(2\), where
+measures of reliability suggests poorer performance.
+
+Thus, broadband services in clusters \(9\) and \(11\), home to over
+twelve million people performed the best during the study period, in
+terms of both quality and reliability. In Figure \ref{UpClusterL},
+cluster \(11\) shows more noticeable peaks and troughs, but the lowest
+points are not at the peak times described in Table
+\ref{up.cluster.descr}. Rather, the slowest upload speeds on average
+occur between \(6:00-7:00\) on Monday morning, \(14:00-15:00\) on
 Friday, and \(16:00-17:00\) on Thursday. These slowest times are still
 mostly faster than the average hourly upload speeds in cluster \(13\).
-Although not all of the local authorities in these two clusters are
-urban, among the \(65\) LADs are Manchester and three of the nine other
-boroughs of Greater Manchester, Glasgow, Nottingham, both Portsmouth and
-Southampton, eight London Boroughs, four of the seven constituent
-authorities of the West Midlands Combined Authority. There are also a
-number of other tightly bounded urban areas, such as Aberdeen,
-Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at
-the centre of less confined districts like Burnley, Milton Keynes, and
-Northampton. The local authorities in these two clusters clearly
-experienced resilient broadband that could support high levels of
-telecommuting. They are \emph{not} on the wrong side of the first level
-digital divide, but how likely are they to be able to take advantage of
-their resilient ICT infrastructure and services?
+Finally, ignoring the smallest clusters in terms of population, that is
+clusters \(4\) and \(5\), clusters \(3\), \(7\) and \(10\) also have
+relatively high mean upload speeds. Clusters \(3\) and \(10\) pair high
+mean speeds with low standard deviations relative to the mean speeds,
+suggesting reliability and resilience, as well as quality broadband
+services. Cluster \(7\) has a higher ratio of standard deviation to
+mean, but there is less difference in average speeds between the morning
+and evening peaks than in clusters \(3\) and \(10\).
+
+In summary, the local authorities in clusters \(9\) and \(11\)
+experienced resilient broadband internet that could support high levels
+of telecommuting. Those in clusters \(2\), \(3\), \(7\), and \(10\) also
+experience higher than average mean speeds and rank high to middle on
+measures of service reliability. These LADs are \emph{not} on the wrong
+side of the first level digital divide, but how likely are they to be
+able to take advantage of their resilient ICT infrastructure and
+services? Meanwhile, cluster \(6\) is not only the largest in terms of
+number of LADs and population, it has the closest mean upload speed to
+the pre-clustered average for the whole sample. As well as average
+quality internet services, those in cluster \(6\) also experience
+average reliability for work purposes, ranking fifth behind the four
+other clusters with populations over one million, but ahead of the
+smaller clusters. Clusters \(8\) and \(12\) are also close to average
+mean upload speeds, but show very different patterns in terms of
+reliability, whilst clusters \(1\) and \(13\) appear to suffer most from
+a lack of quality internet services, with slow speeds and high standard
+deviations. With those in cluster \(1\) in particular more likely to
+experience that poor reliability during the morning peak, is this first
+level digital divide occurring in areas where few are occupationally
+able to telecommute anyway, and what are the implications for economic
+resilience?
 
 \hypertarget{sec:4.2}{%
-\subsection{aux regressions}\label{sec:4.2}}
+\subsection{Post-clustering regression analysis}\label{sec:4.2}}
 
@@
 
+Using auxiliary regressions, we test whether the clusters that have
+higher mean speeds and more reliable services consist of LADs that are
+more urban, affluent, and / or more likely to benefit from a choice of
+high quality internet services. We also estimate which of our clusters
+are more likely to have a higher proportion of occupations which enable
+telecommuting because of the nature of the work. The results of these
+auxiliary regressions are presented in Table \ref{aux}. The dependent
+variable is the LAD cluster membership as described in the
+\protect\hyperlink{sec:3}{methods and data} section and equation
+\ref{eq1} and each column represent a different cluster. The reference
+case is cluster \(4\), which includes only the local authority of
+Hambleton in North Yorkshire, a rural area of just over ninety thousand
+people. Mean, experienced upload speeds in cluster \(4\) are close to
+both the average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
+pre-clustered average for the whole sample (\(9.3\)Mb/s) as shown in
+\ref{up.cluster.descr}. However, the standard deviation for cluster
+\(4\) and the difference between average speeds in the morning compared
+to the evening peak periods are indications of worse reliability than
+many of the other clusters. Hence, the results in Table \ref{aux} should
+be seen as relative rather than absolute probabilities.
+
+First, the number of speed tests run per cluster inhabitant between
+\(9:00\)-\(10:59\) is an indication of satisfaction or at least a lack
+of concern over broadband speeds and quality of service. People in LADs
+assigned to clusters \(6\) and \(13\) were more likely to ran speed
+tests (at a per capita basis) at this time of day, whilst people in
+cluster \(11\) were the least likely ones. This high level of
+satisfaction matches the high quality and reliability of service we
+would expect from the analysis in Section {[}4.1{]}{]}(\#sec:4.1). Those
+in clusters \(2\), \(3\), \(9\) and \(10\) were also less likely to ran
+speed tests per capita than those in the other six multi-authority
+clusters. Cluster \(7\) was the only cluster with higher mean speeds and
+middle to high measures of reliability which was more likely to run more
+speed tests per capita. This might be a reflection of the choice of ISP
+and connection options in those LADs, as, along with cluster \(12\),
+those in cluster \(7\) were least likely to have Virgin Media
+connections. Fast Virgin Media cable connections have historically only
+been available to \(45\)\% of premises in the UK \citep{ofcom2016},
+where the more lucrative and competitive market originally attracted the
+cable TV provider. Those in clusters \(2\), \(3\), \(9\) and \(11\)
+benefit from a higher proportion of Virgin connections, which is an
+indication that people in these clusters are more likely to live in
+urban areas, with more choice of broadband services, or in other words,
+are more likely to be on the right side of the first level,
+infrastructure-based digital divide.
+
+We employ distance from London and for the nearest metropolitan area
+(including London) as two more variables depicting peripherality, urban
+structure and, potentially, first level digital divides. The broadband
+speed tests run in the authorities in cluster \(3\) are more likely to
+be taking place close to London than those run in any of the other
+clusters, a result that makes sense considering two of the four
+authorities in cluster \(3\) are the London commuter towns of Harlow and
+Luton. However, even though London was also included in the variable
+calculating distance from the centre of one of either the ten largest
+metropolitan areas in England, Glasgow or Cardiff, tests run in cluster
+\(3\) are likely to be furthest away. Still, whilst significant, the
+effect is tiny and the other members of cluster \(3\) are Corby and
+Eastbourne, two towns which, whilst not part of metropolitan areas, are
+also not in remote parts of the country, and are home to an estimated
+population of \(175\) thousand people between them. \textbf{move maps to
+here?} Meanwhile, LADs in cluster \(7\) are most likely to be near the
+centre of a large metropolitan area, even though the four local
+authorities of cluster \(7\) include no central urban boroughs, and only
+one LAD that is part of a metropolitan area of governance - Tameside in
+Greater Manchester. This may be why those in cluster \(7\) are not
+served by Virgin Media. It also is a demonstration of the complexity of
+experienced broadband upload speeds as captured by time-series
+clustering, and their likely interaction with first level digital
+divides. Cluster \(1\), which our analysis suggests lacks broadband
+resilience, contains five, mainly rural authorities. However, they are
+scattered around the country -- Ceredigion in West Wales, Darlington in
+the Northeast, Eden between the Lake District and the North Penines,
+Rossendale in the Northwest, and Rutland in the East Midlands -- and
+therefore, the results in Table \ref{aux} indicate that these
+authorities are not as far from the centre of a metropolitan area as
+authorities in cluster \(3\), but furthest from London.
+
+Internet resilience, as measured here, seems to be more nuanced than our
+-- arguably crude -- dummy variable depicting the North-South economic
+divide either, which assigns \(1\) to LADs locates in Greater London,
+Southeast, Southwest and East of England regions. The authorities most
+likely to be in the South are those in cluster \(13\), which was
+identified as having the slowest mean upload speeds of any of the
+clusters, and thus a low level of service. However, cluster \(13\) does
+include some rural, remote areas of the country, such as Northwest
+Scotland, Cornwall and Powys in Wales as shown on Figure
+\ref{map.up.clusters}. It also includes the major metropolitan centres
+of Bristol, Liverpool, and Newcastle, as well as nine (of \(32\)) London
+Boroughs and plenty of home county and suburban areas. Although by the
+standard deviation measure cluster \(13\) has unreliable internet
+services, speed variation was less during the morning peak. More
+broadly, the estimates for this variable are not consistent between the
+two clubs of LAD clusters: those who are in the wrong and the right side
+of digital divides.
+
+Regarding the rest of the socio-economic variables we see a varied
+picture. Clusters \(9\) and \(11\), which are the ones benefiting by the
+most resilient internet connections are likely to consist of LADs with
+higher share of managerial and highly skilled occupations as well as
+higher average earnings. Interestingly, both clusters are characterised
+by lower share of occupations that are not usually associated with
+teleworking (professional, associated professional and technical,
+administrative, leisure and machine operation jobs), but also by lower
+number of business per habitant and. However, the likelihood of hosting
+highly skilled residents is not consistent with cluster \(11\) having a
+negative coefficient for the NVQ4 variable. This might be an indication
+of the higher returns that resilient broadband infrastructure generates
+for cluster \(9\) versus cluster \(11\). While both of these cluster
+consist more likely by large LADs in terms of population, job density if
+higher for LADs in cluster \(11\) indicating more labour intensive
+economic structures contrary to LADs in cluster \(9\).
+
+A similar divided picture can be built for the next tier of clusters:
+\(2\), \(3\), \(7\), \(10\). LADs in these clusters experience higher
+than average mean speeds and rank high to middle on measures of service
+reliability. As mentioned before, LADs in these clusters are on the
+right side of the first level digital divide. However, while all of
+them are more likely to be populated by people employed in managerial
+jobs, only LADs in cluster \(2\) and \(10\) are more likely to have
+highly skilled work force, which can more easily take advantage of
+working from home opportunities. A more mixed picture can be built for
+LADs in clusters \(3\) and \(7\), which are associated, on the one hand,
+with lower skills and, on the other hand, with higher likelihood of
+individuals in managerial occupations. A similar contradicting pattern
+can be illustrated for cluster \(6\), which is the largest one with
+average quality of internet service. Arguably, LADs in clusters \(2\)
+and \(10\) are the ones benefiting more by resilient internet
+infrastructure in an era that working from home became a vehicle for
+economic resilience.
+
+Moving on to the next tier of clusters \(8\) and \(12\), which are
+characterised by LADs enjoying close to average mean upload speeds, but
+different patterns of internet reliability, we see a differentiated
+picture. On the one hand, both of them are likely to host individuals
+employed in managerial occupations while the opposite applies on
+skilled, professional and leisure occupations. On the other hand, LADs
+in cluster \(8\) are characterised by a very small positive and LADs in
+cluster \(12\) by a negative likelihood to host highly skilled
+individuals. Moreover, LADs in these two clusters have a rather similar
+profile regarding the occupational variables, but different tendencies
+regarding the stock of businesses. In total, LADs in these clusters do
+not benefit by great internet services and could only marginally benefit
+by working from home opportunities.
+
+Lastly, clusters \(1\) and \(13\), which appear to suffer most from a
+lack of quality internet services are more likely to also suffer from
+low skilled workforce. LADs in cluster \(13\). Contradictory, the
+likelihood of workforce employed in managerial occupations is positive
+for LADs in both clusters, but only individuals in LADs in cluster
+\(13\) are more likely to enjoy high earnings.
+
+\hypertarget{sec:5}{%
+\section{Conclusions}\label{sec:5}}
+
+Upload speeds are not used as benchmarks in the UK, nor have been seen
+as integral to universal service, considering there has never before
+been such extreme demand for telecommuting and operations such as video
+calls.
+
+those in cluster \(6\), including people living in Birmingham, Leeds and
+Sheffield, in twelve London Boroughs, and many suburban areas and
+smaller cities like Cardiff, Oxford and Cambridge This lower level of
+reliability may be as a result of increased demand, e.g.~for
+telecommuting, especially as there are few truly remote areas in this
+cluster.
+
+The eight LADs in clusters \(3\) and \(10\) are home to over one million
+people, and are all more urban and suburban than cluster \(1\),
+including the London Borough of Newham, the Southeast commuter towns of
+Harlow and Luton, the Birmingham suburbs of Bromsgrove and Cannock
+Chase, Dundee City, and the towns of Eastbourne and Corby.
+
+Although not all of the local authorities in these clusters \(9\) and
+\(11\) are urban, among the \(65\) LADs are Manchester and three of the
+nine other boroughs of Greater Manchester, Glasgow, Nottingham, both
+Portsmouth and Southampton, eight London Boroughs, four of the seven
+constituent authorities of the West Midlands Combined Authority. There
+are also a number of other tightly bounded urban areas, such as
+Aberdeen, Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban
+areas at the centre of less confined districts like Burnley, Milton
+Keynes, and Northampton.
+
 Clusters \(3\), \(9\), \(10\) and \(11\) seem to benefit most from high
@@
 suggestive of more use during the working day, and potentially more
-telecommuting. Using auxiliary regressions, we test whether the clusters
-that have higher mean speeds and more reliable services are more urban,
-more wealthy, and / or more likely to benefit from a choice of high
-quality internet services. We also measure which of our clusters are
-more likely to have a higher proportion of occupations which enable
-telecommuting because of the nature of the work. The results of these
-auxiliary regressions are presented in Table \ref{aux}. The dependent
-variable is the LAD cluster membership as described in the
-\protect\hyperlink{sec:3}{Methods and data} section and equation
-\ref{eq1} and each column represent a different cluster. The reference
-case is cluster \(4\), which includes the local authority of Hambleton
-in North Yorkshire, a rural area of just over ninety thousand people.
-Mean, experienced upload speeds in cluster \(4\) are close to both the
-average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
-pre-clustered average for the whole sample (\(9.3\)Mb/s). The standard
-deviation for cluster \(4\) and the ratio of evening to morning peak
-speeds are indications of worse reliability than many of the other
-clusters. Hence, the results in Table \ref{aux} should be seens as
-relative rather than absolute probabilities.
-
-First, the number of speed tests run per cluster inhabitant between
-9:00-10:59 is an indication of satisfaction or at least a lack of
-concern over broadband speeds and quality of service. People in clusters
-\(3\), \(9\), \(10\) and \(11\) all ran fewer speed tests (at a per
-capita basis) at this time of day than people in clusters \(6\) and
-\(13\). Those in clusters \(3\), \(9\), \(10\) and \(11\) also benefit
-from a higher proportion of fast Virgin cable connections than those in
-clusters \(6\) and \(13\). Since services provided by Virgin Media are
-only available to \(45\)\% of premises in the UK \citep{ofcom2016},
-where the more lucrative ICT market originally attracted the cable TV
-provider, this is an indication that people in these clusters are more
-likely to live on the right side of the first level,
-infrastructure-based digital divide. Another indication is if people in
-these clusters are more likely to be closer to London or a metropolitan
-area. The broadband speed tests run in the authorities in cluster \(3\)
-are more likely to be taking place close to London than those run in any
-of the other clusters, a result that makes sense considering two of the
-four authorities in cluster \(3\) are the London commuter towns of
-Harlow and Luton. However, even though London was also included in the
-variable calculating distance to one of the ten major metropolitan areas
-in England, or Glasgow or Cardiff, tests run in cluster \(3\) are likely
-to be furthest away. \textbf{Did you include London in distMet as well?
-- code suggests so} \textbf{Do we need to re-think this variable
-(e.g.~remove Cardiff, add Leicester or\ldots)}
-
-\hypertarget{sec:5}{%
-\section{Conclusions}\label{sec:5}}
+telecommuting.
 
@@
 
-\hypertarget{funding}{%
-\section*{Funding}\label{funding}}
-\addcontentsline{toc}{section}{Funding}
-
-An unnumbered section,
-e.g.~\texttt{\textbackslash{}section*\{Funding\}}, may be used for grant
-details, etc.~if required and included \emph{in the non-anonymous
-version} before any Notes or References.
-
-\bibliographystyle{tfcad}
-\bibliography{bibliography.bib}
-
-\input{"appendix.tex"}
+\hypertarget{sec:appendix}{%
+\section{Appendix}\label{sec:appendix}}
+
+\begin{table}[!htbp] \centering 
+  \caption{LAD cluster membership for upload speed timeseries\label{cluster.membership}} 
+  \label{} 
+\footnotesize 
+\begin{tabular}{@{\extracolsep{5pt}} ccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+ & LAD & cluster membership \\ 
+\hline \\[-1.8ex] 
+1 & Aberdeen City & 9 \\ 
+2 & Aberdeenshire & 13 \\ 
+3 & Adur & 13 \\ 
+4 & Allerdale & 6 \\ 
+5 & Amber Valley & 6 \\ 
+6 & Angus & 6 \\ 
+7 & Antrim and Newtownabbey & 13 \\ 
+8 & Ards and North Down & 12 \\ 
+9 & Argyll and Bute & 13 \\ 
+10 & Armagh City, Banbridge and Craigavon & 13 \\ 
+11 & Arun & 13 \\ 
+12 & Ashfield & 6 \\ 
+13 & Ashford & 6 \\ 
+14 & Aylesbury Vale & 6 \\ 
+15 & Babergh & 13 \\ 
+16 & Barking and Dagenham & 11 \\ 
+17 & Barnet & 6 \\ 
+18 & Barnsley & 9 \\ 
+19 & Barrow-in-Furness & 13 \\ 
+20 & Basildon & 13 \\ 
+21 & Basingstoke and Deane & 6 \\ 
+22 & Bassetlaw & 13 \\ 
+23 & Bath and North East Somerset & 6 \\ 
+24 & Bedford & 13 \\ 
+25 & Belfast & 6 \\ 
+26 & Bexley & 6 \\ 
+27 & Birmingham & 6 \\ 
+28 & Blaby & 11 \\ 
+29 & Blackburn with Darwen & 13 \\ 
+30 & Blackpool & 11 \\ 
+31 & Blaenau Gwent & 6 \\ 
+32 & Bolsover & 11 \\ 
+33 & Bolton & 13 \\ 
+34 & Boston & 13 \\ 
+35 & Bournemouth, Christchurch and Poole & 6 \\ 
+36 & Bracknell Forest & 13 \\ 
+37 & Bradford & 6 \\ 
+38 & Braintree & 6 \\ 
+39 & Breckland & 13 \\ 
+40 & Brent & 11 \\ 
+41 & Brentwood & 6 \\ 
+42 & Bridgend & 6 \\ 
+43 & Brighton and Hove & 13 \\ 
+44 & Bristol, City of & 13 \\ 
+45 & Broadland & 13 \\ 
+46 & Bromley & 6 \\ 
+47 & Bromsgrove & 10 \\ 
+48 & Broxbourne & 9 \\ 
+49 & Broxtowe & 6 \\ 
+50 & Burnley & 11 \\ 
+51 & Bury & 6 \\ 
+52 & Caerphilly & 11 \\ 
+53 & Calderdale & 6 \\ 
+54 & Cambridge & 6 \\ 
+55 & Camden & 13 \\ 
+56 & Cannock Chase & 10 \\ 
+57 & Canterbury & 6 \\ 
+58 & Cardiff & 6 \\ 
+59 & Carlisle & 11 \\ 
+60 & Carmarthenshire & 13 \\ 
+61 & Castle Point & 6 \\ 
+62 & Causeway Coast and Glens & 11 \\ 
+63 & Central Bedfordshire & 13 \\ 
+64 & Ceredigion & 1 \\ 
+65 & Charnwood & 9 \\ 
+66 & Chelmsford & 6 \\ 
+67 & Cheltenham & 6 \\ 
+68 & Cherwell & 6 \\ 
+69 & Cheshire East & 13 \\ 
+70 & Cheshire West and Chester & 13 \\ 
+71 & Chesterfield & 6 \\ 
+72 & Chichester & 13 \\ 
+73 & Chiltern & 13 \\ 
+74 & Chorley & 9 \\ 
+75 & City of Edinburgh & 6 \\ 
+76 & City of London & 13 \\ 
+77 & Clackmannanshire & 6 \\ 
+78 & Colchester & 6 \\ 
+79 & Conwy & 12 \\ 
+80 & Copeland & 6 \\ 
+81 & Corby & 3 \\ 
+82 & Cornwall & 13 \\ 
+83 & Cotswold & 13 \\ 
+84 & County Durham & 6 \\ 
+85 & Coventry & 6 \\ 
+86 & Craven & 13 \\ 
+87 & Crawley & 11 \\ 
+88 & Croydon & 6 \\ 
+89 & Dacorum & 13 \\ 
+90 & Darlington & 1 \\ 
+91 & Dartford & 6 \\ 
+92 & Daventry & 6 \\ 
+93 & Denbighshire & 6 \\ 
+94 & Derby & 6 \\ 
+95 & Derbyshire Dales & 13 \\ 
+96 & Derry City and Strabane & 6 \\ 
+97 & Doncaster & 11 \\ 
+98 & Dorset & 6 \\ 
+99 & Dover & 13 \\ 
+100 & Dudley & 11 \\ 
+101 & Dumfries and Galloway & 13 \\ 
+102 & Dundee City & 10 \\ 
+103 & Ealing & 6 \\ 
+104 & East Ayrshire & 6 \\ 
+105 & East Cambridgeshire & 13 \\ 
+106 & East Devon & 13 \\ 
+107 & East Dunbartonshire & 13 \\ 
+108 & East Hampshire & 6 \\ 
+109 & East Hertfordshire & 13 \\ 
+110 & East Lindsey & 6 \\ 
+111 & East Lothian & 2 \\ 
+112 & East Northamptonshire & 6 \\ 
+113 & East Renfrewshire & 6 \\ 
+114 & East Riding of Yorkshire & 6 \\ 
+115 & East Staffordshire & 12 \\ 
+116 & East Suffolk & 6 \\ 
+117 & Eastbourne & 3 \\ 
+118 & Eastleigh & 6 \\ 
+119 & Eden & 1 \\ 
+120 & Elmbridge & 6 \\ 
+121 & Enfield & 6 \\ 
+122 & Epping Forest & 12 \\ 
+123 & Epsom and Ewell & 13 \\ 
+124 & Erewash & 9 \\ 
+125 & Exeter & 13 \\ 
+126 & Falkirk & 6 \\ 
+127 & Fareham & 6 \\ 
+128 & Fenland & 12 \\ 
+129 & Fermanagh and Omagh & 13 \\ 
+130 & Fife & 13 \\ 
+131 & Flintshire & 13 \\ 
+132 & Folkestone and Hythe & 13 \\ 
+133 & Forest of Dean & 13 \\ 
+134 & Fylde & 5 \\ 
+135 & Gateshead & 6 \\ 
+136 & Gedling & 6 \\ 
+137 & Glasgow City & 9 \\ 
+138 & Gloucester & 13 \\ 
+139 & Gosport & 6 \\ 
+140 & Gravesham & 6 \\ 
+141 & Great Yarmouth & 6 \\ 
+142 & Greenwich & 9 \\ 
+143 & Guildford & 6 \\ 
+144 & Gwynedd & 13 \\ 
+145 & Hackney & 13 \\ 
+146 & Halton & 9 \\ 
+147 & Hambleton & 4 \\ 
+148 & Hammersmith and Fulham & 12 \\ 
+149 & Harborough & 6 \\ 
+150 & Haringey & 6 \\ 
+151 & Harlow & 3 \\ 
+152 & Harrogate & 6 \\ 
+153 & Harrow & 6 \\ 
+154 & Hart & 6 \\ 
+155 & Hartlepool & 6 \\ 
+156 & Hastings & 13 \\ 
+157 & Havant & 9 \\ 
+158 & Havering & 6 \\ 
+159 & Herefordshire, County of & 6 \\ 
+160 & Hertsmere & 11 \\ 
+161 & High Peak & 6 \\ 
+162 & Highland & 13 \\ 
+163 & Hillingdon & 13 \\ 
+164 & Hinckley and Bosworth & 6 \\ 
+165 & Horsham & 6 \\ 
+166 & Hounslow & 11 \\ 
+167 & Huntingdonshire & 13 \\ 
+168 & Hyndburn & 11 \\ 
+169 & Inverclyde & 13 \\ 
+170 & Ipswich & 11 \\ 
+171 & Isle of Anglesey & 13 \\ 
+172 & Isle of Wight & 13 \\ 
+173 & Isles of Scilly & 11 \\ 
+174 & Islington & 6 \\ 
+175 & Kensington and Chelsea & 11 \\ 
+176 & Kettering & 6 \\ 
+177 & King's Lynn and West Norfolk & 6 \\ 
+178 & Kingston upon Hull, City of & 13 \\ 
+179 & Kingston upon Thames & 6 \\ 
+180 & Kirklees & 6 \\ 
+181 & Knowsley & 9 \\ 
+182 & Lambeth & 13 \\ 
+183 & Lancaster & 13 \\ 
+184 & Leeds & 6 \\ 
+185 & Leicester & 6 \\ 
+186 & Lewes & 11 \\ 
+187 & Lewisham & 9 \\ 
+188 & Lichfield & 8 \\ 
+189 & Lincoln & 6 \\ 
+190 & Lisburn and Castlereagh & 7 \\ 
+191 & Liverpool & 13 \\ 
+192 & Luton & 3 \\ 
+193 & Maidstone & 6 \\ 
+194 & Maldon & 6 \\ 
+195 & Malvern Hills & 13 \\ 
+196 & Manchester & 11 \\ 
+197 & Mansfield & 6 \\ 
+198 & Medway & 6 \\ 
+199 & Melton & 13 \\ 
+200 & Mendip & 6 \\ 
+201 & Merthyr Tydfil & 13 \\ 
+202 & Merton & 9 \\ 
+203 & Mid and East Antrim & 9 \\ 
+204 & Mid Devon & 13 \\ 
+205 & Mid Suffolk & 13 \\ 
+206 & Mid Sussex & 6 \\ 
+207 & Mid Ulster & 13 \\ 
+208 & Middlesbrough & 6 \\ 
+209 & Midlothian & 13 \\ 
+210 & Milton Keynes & 9 \\ 
+211 & Mole Valley & 13 \\ 
+212 & Monmouthshire & 6 \\ 
+213 & Moray & 6 \\ 
+214 & Na h-Eileanan Siar & 13 \\ 
+215 & Neath Port Talbot & 13 \\ 
+216 & New Forest & 6 \\ 
+217 & Newark and Sherwood & 9 \\ 
+218 & Newcastle-under-Lyme & 6 \\ 
+219 & Newcastle upon Tyne & 13 \\ 
+220 & Newham & 10 \\ 
+221 & Newport & 6 \\ 
+222 & Newry, Mourne and Down & 13 \\ 
+223 & North Ayrshire & 6 \\ 
+224 & North Devon & 13 \\ 
+225 & North East Derbyshire & 6 \\ 
+226 & North East Lincolnshire & 2 \\ 
+227 & North Hertfordshire & 6 \\ 
+228 & North Kesteven & 6 \\ 
+229 & North Lanarkshire & 6 \\ 
+230 & North Lincolnshire & 6 \\ 
+231 & North Norfolk & 6 \\ 
+232 & North Somerset & 13 \\ 
+233 & North Tyneside & 6 \\ 
+234 & North Warwickshire & 11 \\ 
+235 & North West Leicestershire & 6 \\ 
+236 & Northampton & 9 \\ 
+237 & Northumberland & 6 \\ 
+238 & Norwich & 11 \\ 
+239 & Nottingham & 11 \\ 
+240 & Nuneaton and Bedworth & 6 \\ 
+241 & Oadby and Wigston & 8 \\ 
+242 & Oldham & 9 \\ 
+243 & Orkney Islands & 13 \\ 
+244 & Oxford & 6 \\ 
+245 & Pembrokeshire & 6 \\ 
+246 & Pendle & 6 \\ 
+247 & Perth and Kinross & 13 \\ 
+248 & Peterborough & 11 \\ 
+249 & Plymouth & 13 \\ 
+250 & Portsmouth & 9 \\ 
+251 & Powys & 13 \\ 
+252 & Preston & 12 \\ 
+253 & Reading & 13 \\ 
+254 & Redbridge & 13 \\ 
+255 & Redcar and Cleveland & 13 \\ 
+256 & Redditch & 11 \\ 
+257 & Reigate and Banstead & 13 \\ 
+258 & Renfrewshire & 6 \\ 
+259 & Rhondda Cynon Taf & 12 \\ 
+260 & Ribble Valley & 6 \\ 
+261 & Richmond upon Thames & 9 \\ 
+262 & Richmondshire & 13 \\ 
+263 & Rochdale & 11 \\ 
+264 & Rochford & 6 \\ 
+265 & Rossendale & 1 \\ 
+266 & Rother & 13 \\ 
+267 & Rotherham & 13 \\ 
+268 & Rugby & 9 \\ 
+269 & Runnymede & 6 \\ 
+270 & Rushcliffe & 6 \\ 
+271 & Rushmoor & 13 \\ 
+272 & Rutland & 1 \\ 
+273 & Ryedale & 6 \\ 
+274 & Salford & 6 \\ 
+275 & Sandwell & 9 \\ 
+276 & Scarborough & 11 \\ 
+277 & Scottish Borders & 13 \\ 
+278 & Sedgemoor & 13 \\ 
+279 & Sefton & 6 \\ 
+280 & Selby & 7 \\ 
+281 & Sevenoaks & 13 \\ 
+282 & Sheffield & 6 \\ 
+283 & Shetland Islands & 13 \\ 
+284 & Shropshire & 6 \\ 
+285 & Slough & 11 \\ 
+286 & Solihull & 6 \\ 
+287 & Somerset West and Taunton & 13 \\ 
+288 & South Ayrshire & 6 \\ 
+289 & South Bucks & 13 \\ 
+290 & South Cambridgeshire & 13 \\ 
+291 & South Derbyshire & 9 \\ 
+292 & South Gloucestershire & 13 \\ 
+293 & South Hams & 6 \\ 
+294 & South Holland & 6 \\ 
+295 & South Kesteven & 9 \\ 
+296 & South Lakeland & 13 \\ 
+297 & South Lanarkshire & 6 \\ 
+298 & South Norfolk & 13 \\ 
+299 & South Northamptonshire & 9 \\ 
+300 & South Oxfordshire & 6 \\ 
+301 & South Ribble & 13 \\ 
+302 & South Somerset & 13 \\ 
+303 & South Staffordshire & 6 \\ 
+304 & South Tyneside & 13 \\ 
+305 & Southampton & 9 \\ 
+306 & Southend-on-Sea & 13 \\ 
+307 & Southwark & 13 \\ 
+308 & Spelthorne & 9 \\ 
+309 & St Albans & 6 \\ 
+310 & St. Helens & 11 \\ 
+311 & Stafford & 13 \\ 
+312 & Staffordshire Moorlands & 6 \\ 
+313 & Stevenage & 11 \\ 
+314 & Stirling & 13 \\ 
+315 & Stockport & 6 \\ 
+316 & Stockton-on-Tees & 9 \\ 
+317 & Stoke-on-Trent & 6 \\ 
+318 & Stratford-on-Avon & 13 \\ 
+319 & Stroud & 13 \\ 
+320 & Sunderland & 11 \\ 
+321 & Surrey Heath & 6 \\ 
+322 & Sutton & 6 \\ 
+323 & Swale & 6 \\ 
+324 & Swansea & 13 \\ 
+325 & Swindon & 13 \\ 
+326 & Tameside & 7 \\ 
+327 & Tamworth & 6 \\ 
+328 & Tandridge & 8 \\ 
+329 & Teignbridge & 13 \\ 
+330 & Telford and Wrekin & 9 \\ 
+331 & Tendring & 6 \\ 
+332 & Test Valley & 6 \\ 
+333 & Tewkesbury & 13 \\ 
+334 & Thanet & 13 \\ 
+335 & Three Rivers & 12 \\ 
+336 & Thurrock & 6 \\ 
+337 & Tonbridge and Malling & 6 \\ 
+338 & Torbay & 13 \\ 
+339 & Torfaen & 6 \\ 
+340 & Torridge & 13 \\ 
+341 & Tower Hamlets & 13 \\ 
+342 & Trafford & 9 \\ 
+343 & Tunbridge Wells & 13 \\ 
+344 & Uttlesford & 13 \\ 
+345 & Vale of Glamorgan & 8 \\ 
+346 & Vale of White Horse & 11 \\ 
+347 & Wakefield & 6 \\ 
+348 & Walsall & 9 \\ 
+349 & Waltham Forest & 13 \\ 
+350 & Wandsworth & 13 \\ 
+351 & Warrington & 6 \\ 
+352 & Warwick & 6 \\ 
+353 & Watford & 13 \\ 
+354 & Waverley & 13 \\ 
+355 & Wealden & 6 \\ 
+356 & Wellingborough & 6 \\ 
+357 & Welwyn Hatfield & 9 \\ 
+358 & West Berkshire & 6 \\ 
+359 & West Devon & 8 \\ 
+360 & West Dunbartonshire & 6 \\ 
+361 & West Lancashire & 6 \\ 
+362 & West Lindsey & 13 \\ 
+363 & West Lothian & 6 \\ 
+364 & West Oxfordshire & 6 \\ 
+365 & West Suffolk & 6 \\ 
+366 & Westminster & 12 \\ 
+367 & Wigan & 6 \\ 
+368 & Wiltshire & 6 \\ 
+369 & Winchester & 13 \\ 
+370 & Windsor and Maidenhead & 13 \\ 
+371 & Wirral & 13 \\ 
+372 & Woking & 6 \\ 
+373 & Wokingham & 13 \\ 
+374 & Wolverhampton & 11 \\ 
+375 & Worcester & 6 \\ 
+376 & Worthing & 13 \\ 
+377 & Wrexham & 6 \\ 
+378 & Wychavon & 13 \\ 
+379 & Wycombe & 6 \\ 
+380 & Wyre & 13 \\ 
+381 & Wyre Forest & 7 \\ 
+382 & York & 6 \\ 
+\hline \\[-1.8ex] 
+\end{tabular} 
+\end{table}
+
+\begin{table}[!htbp] \centering 
+  \caption{Descriptive statistics for the auxiliary regression explanatory variables\label{descr.aux}} 
+  \label{} 
+\footnotesize 
+\begin{tabular}{@{\extracolsep{0pt}}lccccccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St.
+Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Pctl(25)} & \multicolumn{1}{c}{Pctl(75)} & \multicolumn{1}{c}{Max} \\
+\hline \\[-1.8ex] 
+pop, 2018 & 365 & 1.137 & 5.726 & 0.400 & 0.700 & 0.930 & 110.110 \\ 
+job density, 2018 & 365 & 53.269 & 57.700 & 0.150 & 22.050 & 69.290 & 544.090 \\ 
+distance to nearest met. area & 365 & 201.558 & 173.634 & 0.150 & 76.180 & 278.880 & 1,003.950 \\ 
+distance to London & 365 & 0.463 & 0.499 & 0 & 0 & 1 & 1 \\ 
+South of the UK & 363 & 12.009 & 4.013 & 3.600 & 9.000 & 14.300 & 27.900 \\ 
+managerial jobs, 2020 & 364 & 14.505 & 4.057 & 3.500 & 11.800 & 16.900 & 29.600 \\ 
+tech jobs, 2020 & 358 & 10.513 & 3.764 & 1.000 & 8.025 & 12.500 & 21.600 \\ 
+skilled trade jobs, 2020 & 364 & 21.223 & 6.902 & 4.400 & 16.775 & 24.850 & 71.600 \\ 
+professional jobs, 2020 & 359 & 9.965 & 2.738 & 3.200 & 8.100 & 11.400 & 21.300 \\ 
+administrative jobs, 2020 & 362 & 9.261 & 2.827 & 2.800 & 7.300 & 11.400 & 17.800 \\ 
+leisure jobs, 2020 & 337 & 6.339 & 2.847 & 1.200 & 4.400 & 7.900 & 19.800 \\ 
+machine operation jobs, 2020 & 360 & 592.184 & 81.129 & 437.600 & 534.625 & 633.875 & 893.200 \\ 
+earnings, 2019 & 365 & 39.329 & 11.076 & 15.000 & 31.800 & 45.300 & 100.000 \\ 
+NVQ4+ & 365 & 0.152 & 0.141 & 0.000 & 0.018 & 0.241 & 0.753 \\ 
+Virgin Media \%, 2020 & 365 & 0.057 & 0.164 & 0.023 & 0.038 & 0.056 & 3.174 \\ 
+n. business est. per hab., 2019 & 365 & 0.0005 & 0.0002 & 0.0001 & 0.0003 & 0.001 & 0.001 \\ 
+\hline \\[-1.8ex] 
+\end{tabular} 
+\end{table}
+
+\bibliographystyle{sageh}
+\bibliography{bibliography}
 

### 2020-12-22 paper/v1_sage/broadband.speed.covid.Rmd

@@
 Various studies have illustrated severe inequalities between the north and the
-south regions of England **replace 'UK'** regarding, for example, skills and human capital,
+south regions of England regarding, for example, skills and human capital,
 unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@@
 However, the time profile for cluster $6$ in Figure \ref{UpClusterL} shows
-that upload speeds drop quickly from $6:00$ to $7:00$ on a Monday
+that upload speeds are highest at $6:00$ on a Monday morning
 and peak between $23:00$ and midnight on Wednesday and Thursday,
@@
 and the ratio of upload speeds in the peak periods indicates that speeds are
-only about 2% slower in the morning.
+only about 2% slower in the morning than the evening.
 Mean speeds are also higher than in any other cluster, excluding cluster $2$,
@@
 Those in clusters $2$, $3$, $7$, and $10$ also experience higher than average
-mean speeds and rank high to middle on measures of service reliability.
+mean speeds and rank middle to high on measures of service reliability.
 These LADs are *not* on the wrong side of the first level digital divide,
@@
 for the $13$ clusters ($9.9$Mb/s) and the pre-clustered average for the whole sample
-($9.3$Mb/s) as shown in \ref{up.cluster.descr}.
+($9.3$Mb/s) as shown in Table \ref{up.cluster.descr}.
 However, the standard deviation for cluster $4$ and the difference between average
@@
 and quality of service.
-People in LADs assigned to clusters $6$ and $13$ were more likely to ran speed tests
+People in LADs assigned to clusters $6$ and $13$ were more likely to run speed tests
 (at a per capita basis)
-at this time of day, whilst people in cluster $11$ were the least likely ones.
+at this time of day, whilst people in cluster $11$ were the least likely to do so.
 This high level of satisfaction matches the high quality and reliability of service
@@
 from the centre of one of either the ten largest metropolitan areas in England,
-Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
+or Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
 Still, whilst significant, the effect is tiny and the other members of cluster $3$
@@
 of $175$ thousand people between them. **move maps to here?**
+
 Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large
 metropolitan area, even though the four local authorities of cluster $7$ include
-no central urban boroughs, and only one LAD that is part of a metropolitan area of
+no central urban boroughs and only one LAD that is part of a metropolitan area of
 governance - Tameside in Greater Manchester.
@@
 which assigns $1$ to LADs locates in Greater London, Southeast, Southwest and East
-of England regions.
+of England regions, and $0$ to the rest.
 The authorities most likely to be in the South are those in cluster $13$,
@@
 It also includes the major metropolitan centres of Bristol, Liverpool,
-and Newcastle, as well as nine (of $32$) London Boroughs and plenty of home county
-and suburban areas.
+and Newcastle, as well as nine (of $32$) London Boroughs, 
+including the City of London and plenty of home county and suburban areas.
 Although by the standard deviation measure cluster $13$ has unreliable internet
-services, speed variation was less during the morning peak.
-More broadly, the estimates for this variable are not consistent between the
-two clubs of LAD clusters: those who are in the wrong and the right side of
-digital divides.
-
-Regarding the rest of the socio-economic variables we see a varied
-picture. Clusters $9$ and $11$, which are the ones benefiting by the most resilient
-internet connections are likely to consist of LADs with higher share of managerial
-and highly skilled occupations as well as higher average earnings. Interestingly,
-both clusters are characterised by lower share of occupations that are not usually
-associated with teleworking (professional, associated professional and technical,
-administrative, leisure and machine operation jobs), but also by lower number of
-business per habitant and. However, the likelihood of hosting highly skilled
-residents is not consistent with cluster $11$ having a negative coefficient for the
-NVQ4 variable. This might be an indication of the higher returns that resilient
-broadband infrastructure generates for cluster $9$ versus cluster $11$. While both
-of these cluster consist more likely by large LADs in terms of population, job
-density if higher for LADs in cluster $11$ indicating more labour intensive
-economic structures contrary to LADs in cluster $9$.
-
-A similar divided picture can be built for the next tier of clusters: $2$, $3$,
+services, speed variation was less during the morning peak, suggesting that 
+the estimates for reliability are inconsistent. 
+Considering that this is one of the largest clusters, and thus the averages 
+incorporate more noise than some of the smaller clusters, 
+it may be that the LADs in this cluster do not all suffer equally from 
+a first level digital divide.
+
+Yet we need to consider the results for other variables in Table \ref{aux} to better
+determine whether clusters $1$ and $13$, which appear to suffer most from a lack 
+of quality internet services, are also more likely to have a low skilled workforce, 
+less able to benefit from telecommuting. 
+Cluster $1$ has the lowest propotion of educated people of any of the clusters,
+and the lowest earnings, despite recording the highest proportion of managerial, 
+professional jobs, and the second highest proportion of tech jobs.
+Rural areas such as those in cluster $1$ are often home to many retired people,
+which might explain these results or perhaps, as these figures are relative,
+we could note that cluster $1$ also has a greater proportion of skilled trades
+than other clusters.
+Either way, it seems likely that the first level digital divide reinforces 
+other inequities in cluster $1$.
+In comparison, those cluster $13$ are more likely to earn more and have a better
+education.
+Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the 
+middle in terms of job density.
+If this is an indication of a high number of SMEs, it might give some explanation
+for the variable internet quality, considering that small businesses have not been
+seen as the most valuable customers for higher speed broadband services **insert ref**.
+
+Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more 
+noise within the averages we use to measure internet quality and reliability,
+and the result indicates an average level of service, even if better than LADs
+in cluster $13$.
+Cluster $6$ also falls towards the middle of the clusters on many of the socio-
+economic variables in Table \ref{aux}. 
+It ranks third or fourth out of the eleven clusters in the table in terms of
+having a higher proportion of managerial, professional, and tech jobs, as well as 
+higher earnings, but is fourth from bottom in terms of education.
+The LADs in cluster $6$ are also diverse, with few truly remote areas, 
+but many urban areas in both the North, including Birmingham, Leeds and Sheffield, 
+and South, including twelve London Boroughs, and many suburban areas and 
+smaller cities like Oxford and Cambridge.
+The capital cities of the other UK nations, Belfast, Cardiff and Edinburgh, are
+also in this cluster, suggesting perhaps that the lower level of reliability 
+discussed in Section [4.1](#sec:4.1) may be as a result of increased demand,
+e.g. for telecommuting.
+
+In comparison, clusters $9$ and $11$, which benefit from the most resilient
+internet connections, are characterised by lower share of occupations that are  
+may benefit from the ability to telecommute (e.g. managerial and technical). 
+Cluster $9$ is more highly educated and higher earning than cluster $11$, 
+has fewer businesses per inhabitant and less job density.
+Indeed, cluster $11$ has a negative coefficient for the NVQ4 variable, whilst
+for cluster $9$ the coefficient is positive, and cluster $11$ has the second
+highest job density and cluster $9$ the second lowest. 
+The former might be an indication of the higher returns that resilient
+broadband infrastructure generates for cluster $9$ versus cluster $11$. 
+Indeed, we did detect slightly more slowdown in the AM peak in cluster $9$ than in
+cluster $11$ (see Table \ref{up.cluster.descr}).
+Meanwhile, a propensity for a high job density might indicate more labour intensive 
+economic structures in cluster $11$ than in the LADs in cluster $9$.
+
+The LADs in clusters $9$ and $11$ also have opposite signs for their likelihood 
+of being in the South of England.
+Those in cluster $9$ less likely to be in the South and those in cluster $11$ 
+more likely, although the coefficient for cluster $9$ might be skewed by 
+the presence of the Scottish cities of Glasgow and Aberdeen, 
+as Scotland has a different economic profile than England.
+Still, both clusters consist of larger LADs in terms of population, 
+including the urban centre of Manchester and three of the nine other districts of
+Greater Manchester, Nottingham, both Portsmouth and Southampton,
+eight London Boroughs, and four of the seven constituent authorities of the
+West Midlands Combined Authority.
+Among the $65$ LADs are also a number of other tightly bounded urban areas, 
+such as Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas 
+at the centre of less confined districts like Burnley, 
+the digital hub of Milton Keynes, and Northampton.
+Thus, our findings confirm the importance of geography and urban locations 
+for being on the right side of the first level digital divide, and suggest that 
+internet resilience cuts across other digital and economic divides.
+
+A similarly divided picture can be built for the next tier of clusters: $2$, $3$,
 $7$, $10$. LADs in these clusters experience higher than average mean speeds and
-rank high to middle on measures of service reliability. As mentioned before, LADs
-in these clusters are on the right side of the first level digital divide. However,
-while all of them are more likely to be populated by people employed in managerial
-jobs, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
-work force, which can more easily take advantage of working from home opportunities.
-A more mixed picture can be built for LADs in clusters $3$ and $7$, which are
-associated, on the one hand, with lower skills and, on the other hand, with higher
-likelihood of individuals in managerial occupations. A similar contradicting
-pattern can be illustrated for cluster $6$, which is the largest one with average 
-quality of internet service. Arguably, LADs in clusters $2$ and $10$ are the ones
-benefiting more by resilient internet infrastructure in an era that working
-from home became a vehicle for economic resilience.
-
-Moving on to the next tier of clusters $8$ and $12$, which are characterised by
-LADs enjoying close to average mean upload speeds, but different patterns of internet
-reliability, we see a differentiated picture. On the one hand, both of them are
-likely to host individuals employed in managerial occupations while the opposite
-applies on skilled, professional and leisure occupations. On the other hand,
-LADs in cluster $8$ are characterised by a very small positive and LADs in cluster
-$12$ by a negative likelihood to host highly skilled individuals. Moreover, LADs
-in these two clusters have a rather
-similar profile regarding the occupational variables, but different tendencies
-regarding the stock of businesses. In total, LADs in these clusters do not
-benefit by great internet services and could only marginally benefit by working
-from home opportunities.
-
-Lastly, clusters $1$ and $13$, which appear to suffer most from a lack of quality
-internet services are more likely to also suffer from low skilled workforce. LADs
-in cluster $13$. Contradictory, the likelihood of workforce employed in managerial
-occupations is positive for LADs in both clusters, but only individuals in LADs
-in cluster $13$ are more likely to enjoy high earnings.
+rank high to middle on measures of service reliability. 
+LADs in these clusters are on the right side of the first level digital divide. 
+However, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
+work force, which might more easily take advantage of working from home 
+opportunities.
+Cluster $2$ is comprised of just two LADs, with the lowest job density of any
+cluster -- these are East Lothian near Edinburgh, and North East Lincolnshire, 
+home to Grimsby.
+Cluster $10$ is comprised of four LADs, including two peripheral suburban areas 
+of Birmingham, the East London Borough of Newham, and Dundee.
+Suburbs are considered the most likely urban form in which telecommuters live, 
+and Dundee has a reputation for tech startups **insert ref**.
+Yet although those in clusters $2$ and $10$ may be making the most of their 
+resilient internet, those in cluster $7$ have the highest earnings.
+The four LADs in cluster $7$ include the suburbs South of Belfast, a suburban 
+district of Greater Manchester, the villages between York and Leeds, and a
+couple of small West Midlands towns.
+LADs in cluster $3$ are associated more with lower skills and have the highest
+proportion of individuals in machine operation jobs. 
+They include the Southeast commuter towns of Harlow and Luton and the slightly more
+distant towns of Eastbourne and Corby.
+Arguably, LADs in clusters $3$ benefit the least from their resilient internet 
+infrastructure in an era when working from home became a vehicle for economic 
+resilience.
+
+Finally, clusters $8$ and $12$, are characterised by LADs enjoying close to average 
+mean upload speeds, but different patterns of internet reliability. 
+LADs in cluster $8$ are likely to host more individuals employed in tech and
+administrative occupations than any other cluster, as well as many in managerial 
+occupations, whilst the opposite applies to skilled trade and leisure occupations. 
+These LADs are characterised by a very small positive likelihood of more educated 
+residents, but are not significantly likely to earn more than other clusters.
+Including suburbs of Leicester, Cardiff, Birmingham, the green belt area south of 
+London straddling the M25 motorway, and rural West Devon, the LADs in cluster $8$
+seem to likely to have the skills and occupations that would benefit from 
+quality internet services, but suffer from poor internet resilience and reliability.
+
+In comparison, LADs in cluster $12$ are less likely to be able to benefit from
+quality internet services if they had them, with fewer individuals achieving 
+NVQ4 or better and lower levels of occupations that would benefit from homeworking. 
+Yet LADs in this cluster benefit from the second highest level of earnings, and
+have the largest stock of businesses per inhabitant and the highest job density. 
+This density of businesses could be why cluster $12$ is the only cluster with higher
+speeds in the morning peak during the study period.
+If people are at home, business premises might well be abandoned.
+Cluster $12$ is home to $1.5$million people spread across twelve LADs, including
+the City of Westminster and the London Borough of Hammersmith.
+However, it also includes a LAD in Northern Ireland, two in Wales, two suburban areas
+North of London, Preston in the Northwest, East Staffordshire in the Midlands, and
+Fenland in the East of England.
+This spatial diversity demonstrates that the temporal clustering of internet 
+resilience does not create spatial clusters, and digital divides do not necessarily
+overlap with economic ones.
 
@@
 telecommuting and operations such as video calls.
-
-those in cluster $6$, including people living in Birmingham, Leeds and
-Sheffield, in twelve London Boroughs, and many suburban areas and smaller cities like
-Cardiff, Oxford and Cambridge
-This lower level of reliability may be as a result of increased demand,
-e.g. for telecommuting, especially as there are few truly remote areas
-in this cluster.
-
-The eight LADs in clusters $3$ and $10$ are home to over one million people,
-and are all more urban and suburban than cluster $1$, including the London Borough
-of Newham, the Southeast commuter towns of Harlow and Luton, the Birmingham suburbs
-of Bromsgrove and Cannock Chase, Dundee City, and the towns of Eastbourne and Corby.
-
-Although not all of the local authorities in these clusters $9$ and $11$ are urban,
-among the $65$ LADs are Manchester and three of the nine other boroughs of
-Greater Manchester, Glasgow, Nottingham, both Portsmouth and Southampton,
-eight London Boroughs, four of the seven constituent authorities of the
-West Midlands Combined Authority.
-There are also a number of other tightly bounded urban areas, such as Aberdeen,
-Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at the centre
-of less confined districts like Burnley, Milton Keynes, and Northampton.
 

### 2020-12-22 paper/v1_sage/broadband.speed.covid.Rmd
with amended appendix

@@
 
-# Appendix {#sec:appendix}
+\pagebreak
+
+# Appendix 1 {#appendix1}
+
+This is the LAD cluster membership for the upload speed timeseries.
 
@@
   dplyr::select(LAD, cluster.up) %>%
-  rename('cluster membership' = cluster.up) 
+  #rename('cluster membership' = cluster.up) %>%
   #group_by('cluster membership') %>%
+  group_split(cluster.up, .keep = F)
   #dplyr::arrange()
@@
 
-stargazer(clusters,
-          summary = F,
-          font.size = "footnotesize",
-          type = "latex",
-          header = F,
-          title = "LAD cluster membership for upload speed timeseries\\label{cluster.membership}")  
-```
+cluster1 <- as.vector(unlist(clusters[1]))
+cluster2 <- as.vector(unlist(clusters[2]))
+cluster3 <- as.vector(unlist(clusters[3]))
+cluster4 <- as.vector(unlist(clusters[4]))
+cluster5 <- as.vector(unlist(clusters[5]))
+cluster6 <- as.vector(unlist(clusters[6]))
+cluster7 <- as.vector(unlist(clusters[7]))
+cluster8 <- as.vector(unlist(clusters[8]))
+cluster9 <- as.vector(unlist(clusters[9]))
+cluster10 <- as.vector(unlist(clusters[10]))
+cluster11 <- as.vector(unlist(clusters[11]))
+cluster12 <- as.vector(unlist(clusters[12]))
+cluster13 <- as.vector(unlist(clusters[13]))
+```
+
+**Cluster 1: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster1, type="cmd2"), collapse=", "))
+```
+
+**Cluster 2: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster2, type="cmd2"), collapse=", "))
+```
+
+**Cluster 3: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster3, type="cmd2"), collapse=", "))
+```
+
+**Cluster 4: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster4, type="cmd2"), collapse=", "))
+```
+
+**Cluster 5: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster5, type="cmd2"), collapse=", "))
+```
+
+**Cluster 6: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster6, type="cmd2"), collapse=", "))
+```
+
+**Cluster 7: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster7, type="cmd2"), collapse=", "))
+```
+
+**Cluster 8: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster8, type="cmd2"), collapse=", "))
+```
+
+**Cluster 9: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster9, type="cmd2"), collapse=", "))
+```
+
+**Cluster 10: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster10, type="cmd2"), collapse=", "))
+```
+
+**Cluster 11: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster11, type="cmd2"), collapse=", "))
+```
+
+**Cluster 12: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster12, type="cmd2"), collapse=", "))
+```
+
+**Cluster 13: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster13, type="cmd2"), collapse=", "))
+```
+
+```{r echo=FALSE, message=FALSE, results = "asis", eval=FALSE}
+# LADs <- do.call(cbindX, clusters) # this is with library(gdata)
+# LADs1 <- LADs[,1:7]
+# LADs2 <- LADs[,8:13]
+# 
+# names(LADs1) <- c("Cluster 1", "Cluster 2", "Cluster 3",
+#                      "Cluster 4", "Cluster 5", "Cluster 6",
+#                      "Cluster 7")
+# names(LADs2) <- c("Cluster 8", "Cluster 9", "Cluster 10",
+#                      "Cluster 11", "Cluster 12")
+# 
+# opts <- options(knitr.kable.NA = "") # kableExtra.auto_format = FALSE
+# 
+# # LADs1 %>%
+# # kable(caption = "LAD cluster membership for upload speed timeseries (1 out of 2)\\label{cluster.membership1}") %>%
+# #   kable_styling(font_size = 5,fixed_thead = T)
+# 
+# LADs1 %>%
+# kbl(caption = "LAD cluster membership for upload speed timeseries (1 out of 2)\\label{cluster.membership1}") %>%
+#    kable_classic(font_size = 5,
+#                 fixed_thead = T,
+#                 full_width = F
+#                 )
+# 
+# 
+# # stargazer(LADs1,
+# #           summary = F,
+# #           font.size = "footnotesize",
+# #           type = "latex",
+# #           header = F,
+# #           title = "LAD cluster membership for upload speed timeseries (1 out of 2)\\label{cluster.membership1}")  
+# 
+# stargazer(LADs2,
+#           summary = F,
+#           font.size = "footnotesize",
+#           type = "latex",
+#           header = F,
+#           title = "LAD cluster membership for upload speed timeseries (2 out of 2)\\label{cluster.membership2}")  
+```
+
+# Appendix 2 {#appendix2}
 
@@
 ```
+
+\pagebreak

### 2020-12-22 paper/v1_sage/broadband.speed.covid.tex
with amended appendix

@@
 inequalities between the north and the south regions of England
-\textbf{replace `UK'} regarding, for example, skills and human capital,
-unemployment, productivity and prosperity
+regarding, for example, skills and human capital, unemployment,
+productivity and prosperity
 \citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
@@
 profile for cluster \(6\) in Figure \ref{UpClusterL} shows that upload
-speeds drop quickly from \(6:00\) to \(7:00\) on a Monday and peak
-between \(23:00\) and midnight on Wednesday and Thursday, but tend to be
-lower during the working day. Indeed, experienced mean upload speeds in
-the morning peak are \(4.4\)\% lower than in the evening peak -- a
-greater, more noticeable change than any of the other large clusters
-experience, but smaller than any of the clusters with temporal profiles
-shown in Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have
-mean upload speeds under \(10\)Mb/s, but higher than clusters \(1\) and
-\(13\), and their standard deviation is not dissimilar. However, this
-masks great variation in when lower speeds are experienced, with the
-mean upload speeds much lower between \(9:00-10:59\) than between
-\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning in
-cluster \(12\).
+speeds are highest at \(6:00\) on a Monday morning and peak between
+\(23:00\) and midnight on Wednesday and Thursday, but tend to be lower
+during the working day. Indeed, experienced mean upload speeds in the
+morning peak are \(4.4\)\% lower than in the evening peak -- a greater,
+more noticeable change than any of the other large clusters experience,
+but smaller than any of the clusters with temporal profiles shown in
+Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have mean upload
+speeds under \(10\)Mb/s, but higher than clusters \(1\) and \(13\), and
+their standard deviation is not dissimilar. However, this masks great
+variation in when lower speeds are experienced, with the mean upload
+speeds much lower between \(9:00-10:59\) than between \(19:00-20:59\) in
+cluster \(8\), but slightly faster in the morning in cluster \(12\).
 
@@
 55\%, and the ratio of upload speeds in the peak periods indicates that
-speeds are only about 2\% slower in the morning. Mean speeds are also
-higher than in any other cluster, excluding cluster \(2\), where
-measures of reliability suggests poorer performance.
+speeds are only about 2\% slower in the morning than the evening. Mean
+speeds are also higher than in any other cluster, excluding cluster
+\(2\), where measures of reliability suggests poorer performance.
 
@@
 of telecommuting. Those in clusters \(2\), \(3\), \(7\), and \(10\) also
-experience higher than average mean speeds and rank high to middle on
+experience higher than average mean speeds and rank middle to high on
 measures of service reliability. These LADs are \emph{not} on the wrong
@@
 pre-clustered average for the whole sample (\(9.3\)Mb/s) as shown in
-\ref{up.cluster.descr}. However, the standard deviation for cluster
-\(4\) and the difference between average speeds in the morning compared
-to the evening peak periods are indications of worse reliability than
-many of the other clusters. Hence, the results in Table \ref{aux} should
-be seen as relative rather than absolute probabilities.
+Table \ref{up.cluster.descr}. However, the standard deviation for
+cluster \(4\) and the difference between average speeds in the morning
+compared to the evening peak periods are indications of worse
+reliability than many of the other clusters. Hence, the results in Table
+\ref{aux} should be seen as relative rather than absolute probabilities.
 
@@
 of concern over broadband speeds and quality of service. People in LADs
-assigned to clusters \(6\) and \(13\) were more likely to ran speed
+assigned to clusters \(6\) and \(13\) were more likely to run speed
 tests (at a per capita basis) at this time of day, whilst people in
-cluster \(11\) were the least likely ones. This high level of
+cluster \(11\) were the least likely to do so. This high level of
 satisfaction matches the high quality and reliability of service we
@@
 calculating distance from the centre of one of either the ten largest
-metropolitan areas in England, Glasgow or Cardiff, tests run in cluster
-\(3\) are likely to be furthest away. Still, whilst significant, the
-effect is tiny and the other members of cluster \(3\) are Corby and
+metropolitan areas in England, or Glasgow or Cardiff, tests run in
+cluster \(3\) are likely to be furthest away. Still, whilst significant,
+the effect is tiny and the other members of cluster \(3\) are Corby and
 Eastbourne, two towns which, whilst not part of metropolitan areas, are
@@
 population of \(175\) thousand people between them. \textbf{move maps to
-here?} Meanwhile, LADs in cluster \(7\) are most likely to be near the
-centre of a large metropolitan area, even though the four local
-authorities of cluster \(7\) include no central urban boroughs, and only
-one LAD that is part of a metropolitan area of governance - Tameside in
-Greater Manchester. This may be why those in cluster \(7\) are not
-served by Virgin Media. It also is a demonstration of the complexity of
+here?}
+
+Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
+of a large metropolitan area, even though the four local authorities of
+cluster \(7\) include no central urban boroughs and only one LAD that is
+part of a metropolitan area of governance - Tameside in Greater
+Manchester. This may be why those in cluster \(7\) are not served by
+Virgin Media. It also is a demonstration of the complexity of
 experienced broadband upload speeds as captured by time-series
@@
 divide either, which assigns \(1\) to LADs locates in Greater London,
-Southeast, Southwest and East of England regions. The authorities most
-likely to be in the South are those in cluster \(13\), which was
-identified as having the slowest mean upload speeds of any of the
-clusters, and thus a low level of service. However, cluster \(13\) does
-include some rural, remote areas of the country, such as Northwest
-Scotland, Cornwall and Powys in Wales as shown on Figure
+Southeast, Southwest and East of England regions, and \(0\) to the rest.
+The authorities most likely to be in the South are those in cluster
+\(13\), which was identified as having the slowest mean upload speeds of
+any of the clusters, and thus a low level of service. However, cluster
+\(13\) does include some rural, remote areas of the country, such as
+Northwest Scotland, Cornwall and Powys in Wales as shown on Figure
 \ref{map.up.clusters}. It also includes the major metropolitan centres
 of Bristol, Liverpool, and Newcastle, as well as nine (of \(32\)) London
-Boroughs and plenty of home county and suburban areas. Although by the
-standard deviation measure cluster \(13\) has unreliable internet
-services, speed variation was less during the morning peak. More
-broadly, the estimates for this variable are not consistent between the
-two clubs of LAD clusters: those who are in the wrong and the right side
-of digital divides.
-
-Regarding the rest of the socio-economic variables we see a varied
-picture. Clusters \(9\) and \(11\), which are the ones benefiting by the
-most resilient internet connections are likely to consist of LADs with
-higher share of managerial and highly skilled occupations as well as
-higher average earnings. Interestingly, both clusters are characterised
-by lower share of occupations that are not usually associated with
-teleworking (professional, associated professional and technical,
-administrative, leisure and machine operation jobs), but also by lower
-number of business per habitant and. However, the likelihood of hosting
-highly skilled residents is not consistent with cluster \(11\) having a
-negative coefficient for the NVQ4 variable. This might be an indication
-of the higher returns that resilient broadband infrastructure generates
-for cluster \(9\) versus cluster \(11\). While both of these cluster
-consist more likely by large LADs in terms of population, job density if
-higher for LADs in cluster \(11\) indicating more labour intensive
-economic structures contrary to LADs in cluster \(9\).
-
-A similar divided picture can be built for the next tier of clusters:
+Boroughs, including the City of London and plenty of home county and
+suburban areas. Although by the standard deviation measure cluster
+\(13\) has unreliable internet services, speed variation was less during
+the morning peak, suggesting that the estimates for reliability are
+inconsistent. Considering that this is one of the largest clusters, and
+thus the averages incorporate more noise than some of the smaller
+clusters, it may be that the LADs in this cluster do not all suffer
+equally from a first level digital divide.
+
+Yet we need to consider the results for other variables in Table
+\ref{aux} to better determine whether clusters \(1\) and \(13\), which
+appear to suffer most from a lack of quality internet services, are also
+more likely to have a low skilled workforce, less able to benefit from
+telecommuting. Cluster \(1\) has the lowest propotion of educated people
+of any of the clusters, and the lowest earnings, despite recording the
+highest proportion of managerial, professional jobs, and the second
+highest proportion of tech jobs. Rural areas such as those in cluster
+\(1\) are often home to many retired people, which might explain these
+results or perhaps, as these figures are relative, we could note that
+cluster \(1\) also has a greater proportion of skilled trades than other
+clusters. Either way, it seems likely that the first level digital
+divide reinforces other inequities in cluster \(1\). In comparison,
+those cluster \(13\) are more likely to earn more and have a better
+education. Cluster \(13\) also has the most businesses per inhabitant,
+but is somewhere in the middle in terms of job density. If this is an
+indication of a high number of SMEs, it might give some explanation for
+the variable internet quality, considering that small businesses have
+not been seen as the most valuable customers for higher speed broadband
+services \textbf{insert ref}.
+
+Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
+there is more noise within the averages we use to measure internet
+quality and reliability, and the result indicates an average level of
+service, even if better than LADs in cluster \(13\). Cluster \(6\) also
+falls towards the middle of the clusters on many of the socio- economic
+variables in Table \ref{aux}. It ranks third or fourth out of the eleven
+clusters in the table in terms of having a higher proportion of
+managerial, professional, and tech jobs, as well as higher earnings, but
+is fourth from bottom in terms of education. The LADs in cluster \(6\)
+are also diverse, with few truly remote areas, but many urban areas in
+both the North, including Birmingham, Leeds and Sheffield, and South,
+including twelve London Boroughs, and many suburban areas and smaller
+cities like Oxford and Cambridge. The capital cities of the other UK
+nations, Belfast, Cardiff and Edinburgh, are also in this cluster,
+suggesting perhaps that the lower level of reliability discussed in
+Section \protect\hyperlink{sec:4.1}{4.1} may be as a result of increased
+demand, e.g.~for telecommuting.
+
+In comparison, clusters \(9\) and \(11\), which benefit from the most
+resilient internet connections, are characterised by lower share of
+occupations that are\\
+may benefit from the ability to telecommute (e.g.~managerial and
+technical). Cluster \(9\) is more highly educated and higher earning
+than cluster \(11\), has fewer businesses per inhabitant and less job
+density. Indeed, cluster \(11\) has a negative coefficient for the NVQ4
+variable, whilst for cluster \(9\) the coefficient is positive, and
+cluster \(11\) has the second highest job density and cluster \(9\) the
+second lowest. The former might be an indication of the higher returns
+that resilient broadband infrastructure generates for cluster \(9\)
+versus cluster \(11\). Indeed, we did detect slightly more slowdown in
+the AM peak in cluster \(9\) than in cluster \(11\) (see Table
+\ref{up.cluster.descr}). Meanwhile, a propensity for a high job density
+might indicate more labour intensive economic structures in cluster
+\(11\) than in the LADs in cluster \(9\).
+
+The LADs in clusters \(9\) and \(11\) also have opposite signs for their
+likelihood of being in the South of England. Those in cluster \(9\) less
+likely to be in the South and those in cluster \(11\) more likely,
+although the coefficient for cluster \(9\) might be skewed by the
+presence of the Scottish cities of Glasgow and Aberdeen, as Scotland has
+a different economic profile than England. Still, both clusters consist
+of larger LADs in terms of population, including the urban centre of
+Manchester and three of the nine other districts of Greater Manchester,
+Nottingham, both Portsmouth and Southampton, eight London Boroughs, and
+four of the seven constituent authorities of the West Midlands Combined
+Authority. Among the \(65\) LADs are also a number of other tightly
+bounded urban areas, such as Blackpool, Ipswich, Norwich, Slough, and
+Stevenage, and urban areas at the centre of less confined districts like
+Burnley, the digital hub of Milton Keynes, and Northampton. Thus, our
+findings confirm the importance of geography and urban locations for
+being on the right side of the first level digital divide, and suggest
+that internet resilience cuts across other digital and economic divides.
+
+A similarly divided picture can be built for the next tier of clusters:
 \(2\), \(3\), \(7\), \(10\). LADs in these clusters experience higher
 than average mean speeds and rank high to middle on measures of service
-reliability. As mentioned before, LADs in these clusters are on the
-right side of the first level digital divide. However, while all of
-them are more likely to be populated by people employed in managerial
-jobs, only LADs in cluster \(2\) and \(10\) are more likely to have
-highly skilled work force, which can more easily take advantage of
-working from home opportunities. A more mixed picture can be built for
-LADs in clusters \(3\) and \(7\), which are associated, on the one hand,
-with lower skills and, on the other hand, with higher likelihood of
-individuals in managerial occupations. A similar contradicting pattern
-can be illustrated for cluster \(6\), which is the largest one with
-average quality of internet service. Arguably, LADs in clusters \(2\)
-and \(10\) are the ones benefiting more by resilient internet
-infrastructure in an era that working from home became a vehicle for
+reliability. LADs in these clusters are on the right side of the first
+level digital divide. However, only LADs in cluster \(2\) and \(10\) are
+more likely to have highly skilled work force, which might more easily
+take advantage of working from home opportunities. Cluster \(2\) is
+comprised of just two LADs, with the lowest job density of any cluster
+-- these are East Lothian near Edinburgh, and North East Lincolnshire,
+home to Grimsby. Cluster \(10\) is comprised of four LADs, including two
+peripheral suburban areas of Birmingham, the East London Borough of
+Newham, and Dundee. Suburbs are considered the most likely urban form in
+which telecommuters live, and Dundee has a reputation for tech startups
+\textbf{insert ref}. Yet although those in clusters \(2\) and \(10\) may
+be making the most of their resilient internet, those in cluster \(7\)
+have the highest earnings. The four LADs in cluster \(7\) include the
+suburbs South of Belfast, a suburban district of Greater Manchester, the
+villages between York and Leeds, and a couple of small West Midlands
+towns. LADs in cluster \(3\) are associated more with lower skills and
+have the highest proportion of individuals in machine operation jobs.
+They include the Southeast commuter towns of Harlow and Luton and the
+slightly more distant towns of Eastbourne and Corby. Arguably, LADs in
+clusters \(3\) benefit the least from their resilient internet
+infrastructure in an era when working from home became a vehicle for
 economic resilience.
 
-Moving on to the next tier of clusters \(8\) and \(12\), which are
-characterised by LADs enjoying close to average mean upload speeds, but
-different patterns of internet reliability, we see a differentiated
-picture. On the one hand, both of them are likely to host individuals
-employed in managerial occupations while the opposite applies on
-skilled, professional and leisure occupations. On the other hand, LADs
-in cluster \(8\) are characterised by a very small positive and LADs in
-cluster \(12\) by a negative likelihood to host highly skilled
-individuals. Moreover, LADs in these two clusters have a rather similar
-profile regarding the occupational variables, but different tendencies
-regarding the stock of businesses. In total, LADs in these clusters do
-not benefit by great internet services and could only marginally benefit
-by working from home opportunities.
-
-Lastly, clusters \(1\) and \(13\), which appear to suffer most from a
-lack of quality internet services are more likely to also suffer from
-low skilled workforce. LADs in cluster \(13\). Contradictory, the
-likelihood of workforce employed in managerial occupations is positive
-for LADs in both clusters, but only individuals in LADs in cluster
-\(13\) are more likely to enjoy high earnings.
+Finally, clusters \(8\) and \(12\), are characterised by LADs enjoying
+close to average mean upload speeds, but different patterns of internet
+reliability. LADs in cluster \(8\) are likely to host more individuals
+employed in tech and administrative occupations than any other cluster,
+as well as many in managerial occupations, whilst the opposite applies
+to skilled trade and leisure occupations. These LADs are characterised
+by a very small positive likelihood of more educated residents, but are
+not significantly likely to earn more than other clusters. Including
+suburbs of Leicester, Cardiff, Birmingham, the green belt area south of
+London straddling the M25 motorway, and rural West Devon, the LADs in
+cluster \(8\) seem to likely to have the skills and occupations that
+would benefit from quality internet services, but suffer from poor
+internet resilience and reliability.
+
+In comparison, LADs in cluster \(12\) are less likely to be able to
+benefit from quality internet services if they had them, with fewer
+individuals achieving NVQ4 or better and lower levels of occupations
+that would benefit from homeworking. Yet LADs in this cluster benefit
+from the second highest level of earnings, and have the largest stock of
+businesses per inhabitant and the highest job density. This density of
+businesses could be why cluster \(12\) is the only cluster with higher
+speeds in the morning peak during the study period. If people are at
+home, business premises might well be abandoned. Cluster \(12\) is home
+to \(1.5\)million people spread across twelve LADs, including the City
+of Westminster and the London Borough of Hammersmith. However, it also
+includes a LAD in Northern Ireland, two in Wales, two suburban areas
+North of London, Preston in the Northwest, East Staffordshire in the
+Midlands, and Fenland in the East of England. This spatial diversity
+demonstrates that the temporal clustering of internet resilience does
+not create spatial clusters, and digital divides do not necessarily
+overlap with economic ones.
 
@@
 calls.
-
-those in cluster \(6\), including people living in Birmingham, Leeds and
-Sheffield, in twelve London Boroughs, and many suburban areas and
-smaller cities like Cardiff, Oxford and Cambridge This lower level of
-reliability may be as a result of increased demand, e.g.~for
-telecommuting, especially as there are few truly remote areas in this
-cluster.
-
-The eight LADs in clusters \(3\) and \(10\) are home to over one million
-people, and are all more urban and suburban than cluster \(1\),
-including the London Borough of Newham, the Southeast commuter towns of
-Harlow and Luton, the Birmingham suburbs of Bromsgrove and Cannock
-Chase, Dundee City, and the towns of Eastbourne and Corby.
-
-Although not all of the local authorities in these clusters \(9\) and
-\(11\) are urban, among the \(65\) LADs are Manchester and three of the
-nine other boroughs of Greater Manchester, Glasgow, Nottingham, both
-Portsmouth and Southampton, eight London Boroughs, four of the seven
-constituent authorities of the West Midlands Combined Authority. There
-are also a number of other tightly bounded urban areas, such as
-Aberdeen, Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban
-areas at the centre of less confined districts like Burnley, Milton
-Keynes, and Northampton.
 
@@
 
-\hypertarget{sec:appendix}{%
-\section{Appendix}\label{sec:appendix}}
-
-\begin{table}[!htbp] \centering 
-  \caption{LAD cluster membership for upload speed timeseries\label{cluster.membership}} 
-  \label{} 
-\footnotesize 
-\begin{tabular}{@{\extracolsep{5pt}} ccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
- & LAD & cluster membership \\ 
-\hline \\[-1.8ex] 
-1 & Aberdeen City & 9 \\ 
-2 & Aberdeenshire & 13 \\ 
-3 & Adur & 13 \\ 
-4 & Allerdale & 6 \\ 
-5 & Amber Valley & 6 \\ 
-6 & Angus & 6 \\ 
-7 & Antrim and Newtownabbey & 13 \\ 
-8 & Ards and North Down & 12 \\ 
-9 & Argyll and Bute & 13 \\ 
-10 & Armagh City, Banbridge and Craigavon & 13 \\ 
-11 & Arun & 13 \\ 
-12 & Ashfield & 6 \\ 
-13 & Ashford & 6 \\ 
-14 & Aylesbury Vale & 6 \\ 
-15 & Babergh & 13 \\ 
-16 & Barking and Dagenham & 11 \\ 
-17 & Barnet & 6 \\ 
-18 & Barnsley & 9 \\ 
-19 & Barrow-in-Furness & 13 \\ 
-20 & Basildon & 13 \\ 
-21 & Basingstoke and Deane & 6 \\ 
-22 & Bassetlaw & 13 \\ 
-23 & Bath and North East Somerset & 6 \\ 
-24 & Bedford & 13 \\ 
-25 & Belfast & 6 \\ 
-26 & Bexley & 6 \\ 
-27 & Birmingham & 6 \\ 
-28 & Blaby & 11 \\ 
-29 & Blackburn with Darwen & 13 \\ 
-30 & Blackpool & 11 \\ 
-31 & Blaenau Gwent & 6 \\ 
-32 & Bolsover & 11 \\ 
-33 & Bolton & 13 \\ 
-34 & Boston & 13 \\ 
-35 & Bournemouth, Christchurch and Poole & 6 \\ 
-36 & Bracknell Forest & 13 \\ 
-37 & Bradford & 6 \\ 
-38 & Braintree & 6 \\ 
-39 & Breckland & 13 \\ 
-40 & Brent & 11 \\ 
-41 & Brentwood & 6 \\ 
-42 & Bridgend & 6 \\ 
-43 & Brighton and Hove & 13 \\ 
-44 & Bristol, City of & 13 \\ 
-45 & Broadland & 13 \\ 
-46 & Bromley & 6 \\ 
-47 & Bromsgrove & 10 \\ 
-48 & Broxbourne & 9 \\ 
-49 & Broxtowe & 6 \\ 
-50 & Burnley & 11 \\ 
-51 & Bury & 6 \\ 
-52 & Caerphilly & 11 \\ 
-53 & Calderdale & 6 \\ 
-54 & Cambridge & 6 \\ 
-55 & Camden & 13 \\ 
-56 & Cannock Chase & 10 \\ 
-57 & Canterbury & 6 \\ 
-58 & Cardiff & 6 \\ 
-59 & Carlisle & 11 \\ 
-60 & Carmarthenshire & 13 \\ 
-61 & Castle Point & 6 \\ 
-62 & Causeway Coast and Glens & 11 \\ 
-63 & Central Bedfordshire & 13 \\ 
-64 & Ceredigion & 1 \\ 
-65 & Charnwood & 9 \\ 
-66 & Chelmsford & 6 \\ 
-67 & Cheltenham & 6 \\ 
-68 & Cherwell & 6 \\ 
-69 & Cheshire East & 13 \\ 
-70 & Cheshire West and Chester & 13 \\ 
-71 & Chesterfield & 6 \\ 
-72 & Chichester & 13 \\ 
-73 & Chiltern & 13 \\ 
-74 & Chorley & 9 \\ 
-75 & City of Edinburgh & 6 \\ 
-76 & City of London & 13 \\ 
-77 & Clackmannanshire & 6 \\ 
-78 & Colchester & 6 \\ 
-79 & Conwy & 12 \\ 
-80 & Copeland & 6 \\ 
-81 & Corby & 3 \\ 
-82 & Cornwall & 13 \\ 
-83 & Cotswold & 13 \\ 
-84 & County Durham & 6 \\ 
-85 & Coventry & 6 \\ 
-86 & Craven & 13 \\ 
-87 & Crawley & 11 \\ 
-88 & Croydon & 6 \\ 
-89 & Dacorum & 13 \\ 
-90 & Darlington & 1 \\ 
-91 & Dartford & 6 \\ 
-92 & Daventry & 6 \\ 
-93 & Denbighshire & 6 \\ 
-94 & Derby & 6 \\ 
-95 & Derbyshire Dales & 13 \\ 
-96 & Derry City and Strabane & 6 \\ 
-97 & Doncaster & 11 \\ 
-98 & Dorset & 6 \\ 
-99 & Dover & 13 \\ 
-100 & Dudley & 11 \\ 
-101 & Dumfries and Galloway & 13 \\ 
-102 & Dundee City & 10 \\ 
-103 & Ealing & 6 \\ 
-104 & East Ayrshire & 6 \\ 
-105 & East Cambridgeshire & 13 \\ 
-106 & East Devon & 13 \\ 
-107 & East Dunbartonshire & 13 \\ 
-108 & East Hampshire & 6 \\ 
-109 & East Hertfordshire & 13 \\ 
-110 & East Lindsey & 6 \\ 
-111 & East Lothian & 2 \\ 
-112 & East Northamptonshire & 6 \\ 
-113 & East Renfrewshire & 6 \\ 
-114 & East Riding of Yorkshire & 6 \\ 
-115 & East Staffordshire & 12 \\ 
-116 & East Suffolk & 6 \\ 
-117 & Eastbourne & 3 \\ 
-118 & Eastleigh & 6 \\ 
-119 & Eden & 1 \\ 
-120 & Elmbridge & 6 \\ 
-121 & Enfield & 6 \\ 
-122 & Epping Forest & 12 \\ 
-123 & Epsom and Ewell & 13 \\ 
-124 & Erewash & 9 \\ 
-125 & Exeter & 13 \\ 
-126 & Falkirk & 6 \\ 
-127 & Fareham & 6 \\ 
-128 & Fenland & 12 \\ 
-129 & Fermanagh and Omagh & 13 \\ 
-130 & Fife & 13 \\ 
-131 & Flintshire & 13 \\ 
-132 & Folkestone and Hythe & 13 \\ 
-133 & Forest of Dean & 13 \\ 
-134 & Fylde & 5 \\ 
-135 & Gateshead & 6 \\ 
-136 & Gedling & 6 \\ 
-137 & Glasgow City & 9 \\ 
-138 & Gloucester & 13 \\ 
-139 & Gosport & 6 \\ 
-140 & Gravesham & 6 \\ 
-141 & Great Yarmouth & 6 \\ 
-142 & Greenwich & 9 \\ 
-143 & Guildford & 6 \\ 
-144 & Gwynedd & 13 \\ 
-145 & Hackney & 13 \\ 
-146 & Halton & 9 \\ 
-147 & Hambleton & 4 \\ 
-148 & Hammersmith and Fulham & 12 \\ 
-149 & Harborough & 6 \\ 
-150 & Haringey & 6 \\ 
-151 & Harlow & 3 \\ 
-152 & Harrogate & 6 \\ 
-153 & Harrow & 6 \\ 
-154 & Hart & 6 \\ 
-155 & Hartlepool & 6 \\ 
-156 & Hastings & 13 \\ 
-157 & Havant & 9 \\ 
-158 & Havering & 6 \\ 
-159 & Herefordshire, County of & 6 \\ 
-160 & Hertsmere & 11 \\ 
-161 & High Peak & 6 \\ 
-162 & Highland & 13 \\ 
-163 & Hillingdon & 13 \\ 
-164 & Hinckley and Bosworth & 6 \\ 
-165 & Horsham & 6 \\ 
-166 & Hounslow & 11 \\ 
-167 & Huntingdonshire & 13 \\ 
-168 & Hyndburn & 11 \\ 
-169 & Inverclyde & 13 \\ 
-170 & Ipswich & 11 \\ 
-171 & Isle of Anglesey & 13 \\ 
-172 & Isle of Wight & 13 \\ 
-173 & Isles of Scilly & 11 \\ 
-174 & Islington & 6 \\ 
-175 & Kensington and Chelsea & 11 \\ 
-176 & Kettering & 6 \\ 
-177 & King's Lynn and West Norfolk & 6 \\ 
-178 & Kingston upon Hull, City of & 13 \\ 
-179 & Kingston upon Thames & 6 \\ 
-180 & Kirklees & 6 \\ 
-181 & Knowsley & 9 \\ 
-182 & Lambeth & 13 \\ 
-183 & Lancaster & 13 \\ 
-184 & Leeds & 6 \\ 
-185 & Leicester & 6 \\ 
-186 & Lewes & 11 \\ 
-187 & Lewisham & 9 \\ 
-188 & Lichfield & 8 \\ 
-189 & Lincoln & 6 \\ 
-190 & Lisburn and Castlereagh & 7 \\ 
-191 & Liverpool & 13 \\ 
-192 & Luton & 3 \\ 
-193 & Maidstone & 6 \\ 
-194 & Maldon & 6 \\ 
-195 & Malvern Hills & 13 \\ 
-196 & Manchester & 11 \\ 
-197 & Mansfield & 6 \\ 
-198 & Medway & 6 \\ 
-199 & Melton & 13 \\ 
-200 & Mendip & 6 \\ 
-201 & Merthyr Tydfil & 13 \\ 
-202 & Merton & 9 \\ 
-203 & Mid and East Antrim & 9 \\ 
-204 & Mid Devon & 13 \\ 
-205 & Mid Suffolk & 13 \\ 
-206 & Mid Sussex & 6 \\ 
-207 & Mid Ulster & 13 \\ 
-208 & Middlesbrough & 6 \\ 
-209 & Midlothian & 13 \\ 
-210 & Milton Keynes & 9 \\ 
-211 & Mole Valley & 13 \\ 
-212 & Monmouthshire & 6 \\ 
-213 & Moray & 6 \\ 
-214 & Na h-Eileanan Siar & 13 \\ 
-215 & Neath Port Talbot & 13 \\ 
-216 & New Forest & 6 \\ 
-217 & Newark and Sherwood & 9 \\ 
-218 & Newcastle-under-Lyme & 6 \\ 
-219 & Newcastle upon Tyne & 13 \\ 
-220 & Newham & 10 \\ 
-221 & Newport & 6 \\ 
-222 & Newry, Mourne and Down & 13 \\ 
-223 & North Ayrshire & 6 \\ 
-224 & North Devon & 13 \\ 
-225 & North East Derbyshire & 6 \\ 
-226 & North East Lincolnshire & 2 \\ 
-227 & North Hertfordshire & 6 \\ 
-228 & North Kesteven & 6 \\ 
-229 & North Lanarkshire & 6 \\ 
-230 & North Lincolnshire & 6 \\ 
-231 & North Norfolk & 6 \\ 
-232 & North Somerset & 13 \\ 
-233 & North Tyneside & 6 \\ 
-234 & North Warwickshire & 11 \\ 
-235 & North West Leicestershire & 6 \\ 
-236 & Northampton & 9 \\ 
-237 & Northumberland & 6 \\ 
-238 & Norwich & 11 \\ 
-239 & Nottingham & 11 \\ 
-240 & Nuneaton and Bedworth & 6 \\ 
-241 & Oadby and Wigston & 8 \\ 
-242 & Oldham & 9 \\ 
-243 & Orkney Islands & 13 \\ 
-244 & Oxford & 6 \\ 
-245 & Pembrokeshire & 6 \\ 
-246 & Pendle & 6 \\ 
-247 & Perth and Kinross & 13 \\ 
-248 & Peterborough & 11 \\ 
-249 & Plymouth & 13 \\ 
-250 & Portsmouth & 9 \\ 
-251 & Powys & 13 \\ 
-252 & Preston & 12 \\ 
-253 & Reading & 13 \\ 
-254 & Redbridge & 13 \\ 
-255 & Redcar and Cleveland & 13 \\ 
-256 & Redditch & 11 \\ 
-257 & Reigate and Banstead & 13 \\ 
-258 & Renfrewshire & 6 \\ 
-259 & Rhondda Cynon Taf & 12 \\ 
-260 & Ribble Valley & 6 \\ 
-261 & Richmond upon Thames & 9 \\ 
-262 & Richmondshire & 13 \\ 
-263 & Rochdale & 11 \\ 
-264 & Rochford & 6 \\ 
-265 & Rossendale & 1 \\ 
-266 & Rother & 13 \\ 
-267 & Rotherham & 13 \\ 
-268 & Rugby & 9 \\ 
-269 & Runnymede & 6 \\ 
-270 & Rushcliffe & 6 \\ 
-271 & Rushmoor & 13 \\ 
-272 & Rutland & 1 \\ 
-273 & Ryedale & 6 \\ 
-274 & Salford & 6 \\ 
-275 & Sandwell & 9 \\ 
-276 & Scarborough & 11 \\ 
-277 & Scottish Borders & 13 \\ 
-278 & Sedgemoor & 13 \\ 
-279 & Sefton & 6 \\ 
-280 & Selby & 7 \\ 
-281 & Sevenoaks & 13 \\ 
-282 & Sheffield & 6 \\ 
-283 & Shetland Islands & 13 \\ 
-284 & Shropshire & 6 \\ 
-285 & Slough & 11 \\ 
-286 & Solihull & 6 \\ 
-287 & Somerset West and Taunton & 13 \\ 
-288 & South Ayrshire & 6 \\ 
-289 & South Bucks & 13 \\ 
-290 & South Cambridgeshire & 13 \\ 
-291 & South Derbyshire & 9 \\ 
-292 & South Gloucestershire & 13 \\ 
-293 & South Hams & 6 \\ 
-294 & South Holland & 6 \\ 
-295 & South Kesteven & 9 \\ 
-296 & South Lakeland & 13 \\ 
-297 & South Lanarkshire & 6 \\ 
-298 & South Norfolk & 13 \\ 
-299 & South Northamptonshire & 9 \\ 
-300 & South Oxfordshire & 6 \\ 
-301 & South Ribble & 13 \\ 
-302 & South Somerset & 13 \\ 
-303 & South Staffordshire & 6 \\ 
-304 & South Tyneside & 13 \\ 
-305 & Southampton & 9 \\ 
-306 & Southend-on-Sea & 13 \\ 
-307 & Southwark & 13 \\ 
-308 & Spelthorne & 9 \\ 
-309 & St Albans & 6 \\ 
-310 & St. Helens & 11 \\ 
-311 & Stafford & 13 \\ 
-312 & Staffordshire Moorlands & 6 \\ 
-313 & Stevenage & 11 \\ 
-314 & Stirling & 13 \\ 
-315 & Stockport & 6 \\ 
-316 & Stockton-on-Tees & 9 \\ 
-317 & Stoke-on-Trent & 6 \\ 
-318 & Stratford-on-Avon & 13 \\ 
-319 & Stroud & 13 \\ 
-320 & Sunderland & 11 \\ 
-321 & Surrey Heath & 6 \\ 
-322 & Sutton & 6 \\ 
-323 & Swale & 6 \\ 
-324 & Swansea & 13 \\ 
-325 & Swindon & 13 \\ 
-326 & Tameside & 7 \\ 
-327 & Tamworth & 6 \\ 
-328 & Tandridge & 8 \\ 
-329 & Teignbridge & 13 \\ 
-330 & Telford and Wrekin & 9 \\ 
-331 & Tendring & 6 \\ 
-332 & Test Valley & 6 \\ 
-333 & Tewkesbury & 13 \\ 
-334 & Thanet & 13 \\ 
-335 & Three Rivers & 12 \\ 
-336 & Thurrock & 6 \\ 
-337 & Tonbridge and Malling & 6 \\ 
-338 & Torbay & 13 \\ 
-339 & Torfaen & 6 \\ 
-340 & Torridge & 13 \\ 
-341 & Tower Hamlets & 13 \\ 
-342 & Trafford & 9 \\ 
-343 & Tunbridge Wells & 13 \\ 
-344 & Uttlesford & 13 \\ 
-345 & Vale of Glamorgan & 8 \\ 
-346 & Vale of White Horse & 11 \\ 
-347 & Wakefield & 6 \\ 
-348 & Walsall & 9 \\ 
-349 & Waltham Forest & 13 \\ 
-350 & Wandsworth & 13 \\ 
-351 & Warrington & 6 \\ 
-352 & Warwick & 6 \\ 
-353 & Watford & 13 \\ 
-354 & Waverley & 13 \\ 
-355 & Wealden & 6 \\ 
-356 & Wellingborough & 6 \\ 
-357 & Welwyn Hatfield & 9 \\ 
-358 & West Berkshire & 6 \\ 
-359 & West Devon & 8 \\ 
-360 & West Dunbartonshire & 6 \\ 
-361 & West Lancashire & 6 \\ 
-362 & West Lindsey & 13 \\ 
-363 & West Lothian & 6 \\ 
-364 & West Oxfordshire & 6 \\ 
-365 & West Suffolk & 6 \\ 
-366 & Westminster & 12 \\ 
-367 & Wigan & 6 \\ 
-368 & Wiltshire & 6 \\ 
-369 & Winchester & 13 \\ 
-370 & Windsor and Maidenhead & 13 \\ 
-371 & Wirral & 13 \\ 
-372 & Woking & 6 \\ 
-373 & Wokingham & 13 \\ 
-374 & Wolverhampton & 11 \\ 
-375 & Worcester & 6 \\ 
-376 & Worthing & 13 \\ 
-377 & Wrexham & 6 \\ 
-378 & Wychavon & 13 \\ 
-379 & Wycombe & 6 \\ 
-380 & Wyre & 13 \\ 
-381 & Wyre Forest & 7 \\ 
-382 & York & 6 \\ 
-\hline \\[-1.8ex] 
-\end{tabular} 
-\end{table}
+\pagebreak
+
+\hypertarget{appendix1}{%
+\section{Appendix 1}\label{appendix1}}
+
+This is the LAD cluster membership for the upload speed timeseries.
+
+\textbf{Cluster 1: } Ceredigion, Darlington, Eden, Rossendale, Rutland
+
+\textbf{Cluster 2: } East Lothian, North East Lincolnshire
+
+\textbf{Cluster 3: } Corby, Eastbourne, Harlow, Luton
+
+\textbf{Cluster 4: } Hambleton
+
+\textbf{Cluster 5: } Fylde
+
+\textbf{Cluster 6: } Allerdale, Amber Valley, Angus, Ashfield, Ashford,
+Aylesbury Vale, Barnet, Basingstoke and Deane, Bath and North East
+Somerset, Belfast, Bexley, Birmingham, Blaenau Gwent, Bournemouth,
+Christchurch and Poole, Bradford, Braintree, Brentwood, Bridgend,
+Bromley, Broxtowe, Bury, Calderdale, Cambridge, Canterbury, Cardiff,
+Castle Point, Chelmsford, Cheltenham, Cherwell, Chesterfield, City of
+Edinburgh, Clackmannanshire, Colchester, Copeland, County Durham,
+Coventry, Croydon, Dartford, Daventry, Denbighshire, Derby, Derry City
+and Strabane, Dorset, Ealing, East Ayrshire, East Hampshire, East
+Lindsey, East Northamptonshire, East Renfrewshire, East Riding of
+Yorkshire, East Suffolk, Eastleigh, Elmbridge, Enfield, Falkirk,
+Fareham, Gateshead, Gedling, Gosport, Gravesham, Great Yarmouth,
+Guildford, Harborough, Haringey, Harrogate, Harrow, Hart, Hartlepool,
+Havering, Herefordshire, County of, High Peak, Hinckley and Bosworth,
+Horsham, Islington, Kettering, King's Lynn and West Norfolk, Kingston
+upon Thames, Kirklees, Leeds, Leicester, Lincoln, Maidstone, Maldon,
+Mansfield, Medway, Mendip, Mid Sussex, Middlesbrough, Monmouthshire,
+Moray, New Forest, Newcastle-under-Lyme, Newport, North Ayrshire, North
+East Derbyshire, North Hertfordshire, North Kesteven, North Lanarkshire,
+North Lincolnshire, North Norfolk, North Tyneside, North West
+Leicestershire, Northumberland, Nuneaton and Bedworth, Oxford,
+Pembrokeshire, Pendle, Renfrewshire, Ribble Valley, Rochford, Runnymede,
+Rushcliffe, Ryedale, Salford, Sefton, Sheffield, Shropshire, Solihull,
+South Ayrshire, South Hams, South Holland, South Lanarkshire, South
+Oxfordshire, South Staffordshire, St Albans, Staffordshire Moorlands,
+Stockport, Stoke-on-Trent, Surrey Heath, Sutton, Swale, Tamworth,
+Tendring, Test Valley, Thurrock, Tonbridge and Malling, Torfaen,
+Wakefield, Warrington, Warwick, Wealden, Wellingborough, West Berkshire,
+West Dunbartonshire, West Lancashire, West Lothian, West Oxfordshire,
+West Suffolk, Wigan, Wiltshire, Woking, Worcester, Wrexham, Wycombe,
+York
+
+\textbf{Cluster 7: } Lisburn and Castlereagh, Selby, Tameside, Wyre
+Forest
+
+\textbf{Cluster 8: } Lichfield, Oadby and Wigston, Tandridge, Vale of
+Glamorgan, West Devon
+
+\textbf{Cluster 9: } Aberdeen City, Barnsley, Broxbourne, Charnwood,
+Chorley, Erewash, Glasgow City, Greenwich, Halton, Havant, Knowsley,
+Lewisham, Merton, Mid and East Antrim, Milton Keynes, Newark and
+Sherwood, Northampton, Oldham, Portsmouth, Richmond upon Thames, Rugby,
+Sandwell, South Derbyshire, South Kesteven, South Northamptonshire,
+Southampton, Spelthorne, Stockton-on-Tees, Telford and Wrekin, Trafford,
+Walsall, Welwyn Hatfield
+
+\textbf{Cluster 10: } Bromsgrove, Cannock Chase, Dundee City, Newham
+
+\textbf{Cluster 11: } Barking and Dagenham, Blaby, Blackpool, Bolsover,
+Brent, Burnley, Caerphilly, Carlisle, Causeway Coast and Glens, Crawley,
+Doncaster, Dudley, Hertsmere, Hounslow, Hyndburn, Ipswich, Isles of
+Scilly, Kensington and Chelsea, Lewes, Manchester, North Warwickshire,
+Norwich, Nottingham, Peterborough, Redditch, Rochdale, Scarborough,
+Slough, St.~Helens, Stevenage, Sunderland, Vale of White Horse,
+Wolverhampton
+
+\textbf{Cluster 12: } Ards and North Down, Conwy, East Staffordshire,
+Epping Forest, Fenland, Hammersmith and Fulham, Preston, Rhondda Cynon
+Taf, Three Rivers, Westminster
+
+\textbf{Cluster 13: } Aberdeenshire, Adur, Antrim and Newtownabbey,
+Argyll and Bute, Armagh City, Banbridge and Craigavon, Arun, Babergh,
+Barrow-in-Furness, Basildon, Bassetlaw, Bedford, Blackburn with Darwen,
+Bolton, Boston, Bracknell Forest, Breckland, Brighton and Hove, Bristol,
+City of, Broadland, Camden, Carmarthenshire, Central Bedfordshire,
+Cheshire East, Cheshire West and Chester, Chichester, Chiltern, City of
+London, Cornwall, Cotswold, Craven, Dacorum, Derbyshire Dales, Dover,
+Dumfries and Galloway, East Cambridgeshire, East Devon, East
+Dunbartonshire, East Hertfordshire, Epsom and Ewell, Exeter, Fermanagh
+and Omagh, Fife, Flintshire, Folkestone and Hythe, Forest of Dean,
+Gloucester, Gwynedd, Hackney, Hastings, Highland, Hillingdon,
+Huntingdonshire, Inverclyde, Isle of Anglesey, Isle of Wight, Kingston
+upon Hull, City of, Lambeth, Lancaster, Liverpool, Malvern Hills,
+Melton, Merthyr Tydfil, Mid Devon, Mid Suffolk, Mid Ulster, Midlothian,
+Mole Valley, Na h-Eileanan Siar, Neath Port Talbot, Newcastle upon Tyne,
+Newry, Mourne and Down, North Devon, North Somerset, Orkney Islands,
+Perth and Kinross, Plymouth, Powys, Reading, Redbridge, Redcar and
+Cleveland, Reigate and Banstead, Richmondshire, Rother, Rotherham,
+Rushmoor, Scottish Borders, Sedgemoor, Sevenoaks, Shetland Islands,
+Somerset West and Taunton, South Bucks, South Cambridgeshire, South
+Gloucestershire, South Lakeland, South Norfolk, South Ribble, South
+Somerset, South Tyneside, Southend-on-Sea, Southwark, Stafford,
+Stirling, Stratford-on-Avon, Stroud, Swansea, Swindon, Teignbridge,
+Tewkesbury, Thanet, Torbay, Torridge, Tower Hamlets, Tunbridge Wells,
+Uttlesford, Waltham Forest, Wandsworth, Watford, Waverley, West Lindsey,
+Winchester, Windsor and Maidenhead, Wirral, Wokingham, Worthing,
+Wychavon, Wyre
+
+\hypertarget{appendix2}{%
+\section{Appendix 2}\label{appendix2}}
 
@@
 
+\pagebreak
+
 \bibliographystyle{sageh}

### 2020-12-24 paper/v1_sage/broadband.speed.covid.Rmd
conclusions

@@
 ---
-title: WFH and broadband speed (title needs rework)
+title: "Working from home and digital divides: resilience in the time of a pandemic"
 author:
@@
 and broadband services and infrastructure must be fit for purpose.
-**LET'S LEAVE IT FOR NOW, BUT I THINK WE CAN CRYSTALISE MORE THE RQ**
 
@@
 <!-- Who works from home-->
-In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used
+In this analysis, the terms ‘telecommuting’ and ‘working from home^
+[See also the popular social media hashtag #WFM]’ are used
 interchangeably, as most remote labour during the Covid-19 crisis was carried out
@@
 socio-economic and geographic variables, which are discussed in detail in
-the relevant Section [4.2]](#sec:4.2), in order to explore how the different patterns
+the relevant Section [4.2](#sec:4.2), in order to explore how the different patterns
 might support or undermine efforts to work from home and maintain safe productivity
@@
 
+<!-- limitations-->
+Our analysis is not free of limitations as, for instance, although we do have
+individual data we cannot identify individuals or households and, therefore, we
+need to aggregate our data at the LAD level. However, such individual data could
+only be found though surveys which by definition result in limited samples
+and, importantly, less detailed data about the quality of internet connectivity.
+
 # Results {#sec:4}
@@
 These figures and table provide a comprehensive overview of the quality and
-reliability of experienced broadband in different parts of the UK,
-the temporal clusters offering a novel approach to understanding spatial disparity.
+reliability of experienced broadband in different parts of the UK.
 
@@
 The second largest cluster, comprising $126$ local authorities and over $20$ million
-people, is cluster $13$.
-Cluster $13$ has the slowest aggregate mean upload speed of any of the clusters,
+people, is cluster $13$, which has the slowest aggregate mean upload speed
+of any of the clusters,
 and the second highest ratio of the standard deviation to the mean.
@@
 Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs home to $29.5$
-million people, experience aggregate mean upload speeds of about $1$Mb/s faster
+million people, experienced aggregate mean upload speeds of about $1$Mb/s faster
 than those in cluster $13$, but still lower than the other three large clusters
 and most of the smaller clusters, suggesting a middling quality of service.
-However, the time profile for cluster $6$ in Figure \ref{UpClusterL} shows
+However, the temporal profile for cluster $6$ in Figure \ref{UpClusterL} shows
 that upload speeds are highest at $6:00$ on a Monday morning
@@
 cluster $12$.
-
-```{r echo=FALSE, out.width = "95%",fig.cap="\\label{map.up.clusters}Upload speed clusters for LADs"}
-img1_path <- "figures/map.up.clusters.png"
-include_graphics(img1_path)
-```
 
@@
 
-Using auxiliary regressions, we test whether the clusters that have higher mean
+Using an auxiliary multinomial regression, we test whether the clusters that
+have higher mean
 speeds and more reliable services consist of LADs that are more urban, affluent,
@@
 occupations which enable telecommuting because of the nature of the work.
-The results of these auxiliary regressions are presented in Table \ref{aux}.
+The results of the auxiliary regression are presented in Table \ref{aux}.
 The dependent variable is the LAD cluster membership as described in the
@@
 
-First, the number of speed tests run per cluster inhabitant between $9:00$-$10:59$
-is an indication of satisfaction or at least a lack of concern over broadband speeds
-and quality of service.
-People in LADs assigned to clusters $6$ and $13$ were more likely to run speed tests
-(at a per capita basis)
-at this time of day, whilst people in cluster $11$ were the least likely to do so.
-This high level of satisfaction matches the high quality and reliability of service
-we would expect from the analysis in Section [4.1]](#sec:4.1).
-Those in clusters $2$, $3$, $9$ and $10$ were also less likely to ran speed tests per
-capita than those in the other six multi-authority clusters.
-Cluster $7$ was the only cluster with higher mean speeds and middle to high measures
-of reliability which was more likely to run more speed tests per capita.
-This might be a reflection of the choice of ISP and connection options in those
-LADs, as, along with cluster $12$, those in cluster $7$ were least likely to have
-Virgin Media connections.
-Fast Virgin Media cable connections have historically only been available
-to $45$% of premises in the UK [@ofcom2016], where the more lucrative and
-competitive market originally attracted the cable TV provider.
+First, we control for the number of speed tests run per cluster inhabitant
+between $9:00$-$10:59$ as well as the share of fast Virgin Media internet
+connections. Regarding the former, we expect people in LADs with more unreliable
+connections to test more often their internet speeds and the results validate our
+priors. Regarding the former, fast Virgin Media cable connections have historically
+only been available to $45$% of premises in the UK [@ofcom2016], where the more
+lucrative and competitive market originally attracted the cable TV provider.
 Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
@@
 
-We employ distance from London and for the nearest metropolitan area (including
-London) as two more variables depicting peripherality, urban structure and,
+<!-- is an indication of satisfaction or at least a lack of concern over broadband speeds -->
+<!-- and quality of service. -->
+<!-- People in LADs assigned to clusters $6$ and $13$ were more likely to run speed tests -->
+<!-- (at a per capita basis) -->
+<!-- at this time of day, whilst people in cluster $11$ were the least likely to do so. -->
+<!-- This high level of satisfaction matches the high quality and reliability of service -->
+<!-- we would expect from the analysis in Section [4.1](#sec:4.1). -->
+<!-- Those in clusters $2$, $3$, $9$ and $10$ were also less likely to ran speed tests per -->
+<!-- capita than those in the other six multi-authority clusters. -->
+<!-- Cluster $7$ was the only cluster with higher mean speeds and middle to high measures -->
+<!-- of reliability which was more likely to run more speed tests per capita. -->
+<!-- This might be a reflection of the choice of ISP and connection options in those -->
+<!-- LADs, as, along with cluster $12$, those in cluster $7$ were least likely to have -->
+<!-- Virgin Media connections. -->
+
+Then, we employ distance from London and for the nearest metropolitan area (including
+London) as two variables depicting peripherality, urban structure and,
 potentially, first level digital divides.
@@
 are also not in remote parts of the country, and are home to an estimated population
-of $175$ thousand people between them. **move maps to here?**
+of $175$ thousand people between them.
+
+```{r echo=FALSE, out.width = "95%",fig.cap="\\label{map.up.clusters}Upload speed clusters for LADs"}
+img1_path <- "figures/map.up.clusters.png"
+include_graphics(img1_path)
+```
 
@@
 governance - Tameside in Greater Manchester.
-This may be why those in cluster $7$ are not served by Virgin Media.
+This may explain why those in cluster $7$ are not served by Virgin Media.
 It also is a demonstration of the complexity of experienced broadband upload speeds
@@
 other inequities in cluster $1$.
-In comparison, those cluster $13$ are more likely to earn more and have a better
+In comparison, those in cluster $13$ are more likely to earn more and have a better
 education.
@@
 middle in terms of job density.
-If this is an indication of a high number of SMEs, it might give some explanation
-for the variable internet quality, considering that small businesses have not been
+If this is an indication of a high number of SMEs, it can explain
+the variable internet quality, considering that small businesses have not been
 seen as the most valuable customers for higher speed broadband services **insert ref**.
@@
 economic variables in Table \ref{aux}.
-It ranks third or fourth out of the eleven clusters in the table in terms of
+It ranks third or fourth out of the eleven clusters in the table in terms of the
+likelihood of
 having a higher proportion of managerial, professional, and tech jobs, as well as
@@
 
-In comparison, clusters $9$ and $11$, which benefit from the most resilient
-internet connections, are characterised by lower share of occupations that are  
-may benefit from the ability to telecommute (e.g. managerial and technical). 
-Cluster $9$ is more highly educated and higher earning than cluster $11$, 
-has fewer businesses per inhabitant and less job density.
+In comparison, clusters $9$ and $11$, which enjoy the most resilient
+internet connections, are characterised by higher share of occupations that
+may benefit from the ability to telecommute (e.g. managerial).
+Also, LADs in Cluster $9$ are more likely to host highly educated individuals with
+higher earning than cluster $11$, which has fewer businesses per inhabitant and
+lower job density.
 Indeed, cluster $11$ has a negative coefficient for the NVQ4 variable, whilst
@@
 The former might be an indication of the higher returns that resilient
-broadband infrastructure generates for cluster $9$ versus cluster $11$. 
-Indeed, we did detect slightly more slowdown in the AM peak in cluster $9$ than in
-cluster $11$ (see Table \ref{up.cluster.descr}).
-Meanwhile, a propensity for a high job density might indicate more labour intensive 
-economic structures in cluster $11$ than in the LADs in cluster $9$.
-
-The LADs in clusters $9$ and $11$ also have opposite signs for their likelihood 
-of being in the South of England.
-Those in cluster $9$ less likely to be in the South and those in cluster $11$ 
+broadband infrastructure can generate for cluster $9$ versus cluster $11$.
+Indeed, we did detect slightly more slowdown in the AM peak in cluster $9$ than
+in cluster $11$ (see Table \ref{up.cluster.descr}).
+Meanwhile, a propensity for a higher job density indicate higher urbanization
+rate in LADs in cluster $11$ than those in cluster $9$.
+
+LADs in cluster $9$ are also less likely to be in the South and those in cluster $11$
 more likely, although the coefficient for cluster $9$ might be skewed by
@@
 proportion of individuals in machine operation jobs.
-They include the Southeast commuter towns of Harlow and Luton and the slightly more
-distant towns of Eastbourne and Corby.
+They include the Southeast commuter towns of Harlow and Luton and the slightly
+more distant towns of Eastbourne and Corby.
 Arguably, LADs in clusters $3$ benefit the least from their resilient internet
@@
 This spatial diversity demonstrates that the temporal clustering of internet
-resilience does not create spatial clusters, and digital divides do not necessarily
-overlap with economic ones.
+resilience is not necessarily spatially dependent, and digital divides do not
+necessarily overlap with economic ones.
 
@@
 
-Upload speeds are not used as benchmarks in the UK, nor have been seen as integral to
-universal service, considering there has never before been such extreme demand for
-telecommuting and operations such as video calls.
+This paper offers a new perspective on telecommuting from the viewpoint of the
+complex web of digital divides. Good and reliable internet connections are
+necessary for the population to be able to work from home, but at the same time
+not every place hosts individuals in occupations which allow for telecommuting and
+not every place hosts individuals with the necessary skills to effectively use
+the internet and work form home. Importantly, not every place can utilise good
+internet connectivity -- even if they have them -- in order to achieve economic
+resilience in a period like the current pandemic, but also for a future when
+working from home might be a more common activity.
+
+Contrary to the previous literature, we employ experienced upload speeds as a
+benchmark of internet resilience concerning the capacity to work from home.
+Upload speeds are not used as benchmarks in the UK, nor have been seen as integral
+to universal service, considering there has never before been such extreme demand
+for telecommuting and operations such as video calls. While download speeds are
+associated more with video streaming services, fast upload internet speed are
+necessary for seamless video calls and uploading large files, activities associated
+with work. Given the granular nature of our data, we are able to consider upload
+internet speeds during working hours. To analyse these data we employ time series
+clustering methods, a family of unsupervised machine learning techniques which
+are not often utilised in geographical research.
+
+Using the UK as a case study, our analysis illustrated a nuanced picture, which
+is the outcome of the intersection between the different digital divides and the
+potential to telecommute. While some LADs appear to benefit both from reliable
+internet connections and population which can work from home and, in essence,
+capitalise the digital infrastructure, other LADs are suffering by both unreliable
+internet and low share of individuals who can potentially telecommute.
+
+- different places have different levels of preparedness, some places have some ground to cover
+
+- benefits
 
@@
 use and urban planning to accommodate people who work from home
-[@BUDNITZ2020102713]**also 2020 Swedish article from JTG**; and from productivity 
-and innovation changes, to changes
-in agglomeration externalities and the attraction of large cities [@econobs]
-just to name a few. This paper is positioned to support endeavours in
-understanding the effects of increased telecommuting by
-exposing the spatial and social dimensions of telecommuting including the
-resilience of broadband speeds in terms of both quality and reliability of
-service, and whether this reinforces or redresses prior digital divisions.
-**took this bit you wrote to put in the discussion at the end?**
+[@BUDNITZ2020102713; @ELLDER2020102777] and from productivity
+and innovation changes, to changes in agglomeration externalities and the
+attraction of large cities [@econobs] just to name a few.
+The take home message from this paper is that working form home and, consequently,
+its long term effects cannot be understood, set aside plan our cities for, without
+considering the underpinning digital divides and how telecommuting intersects
+with them.
 
@@
          'LAD.tests.am20/pop' = LAD.tests.am20/pop) %>%
-  dplyr::select(-total.busi, -pop, -LAD.tests.am20)
+  dplyr::select(-total.busi, -LAD.tests.am20)
 
@@
                              "distance to London",
-                             "South of the UK",
+                             "south of the UK",
                              "managerial jobs, 2020",

### 2020-12-24 paper/v1_sage/broadband.speed.covid.tex
conclusions

@@
 
-\title{WFH and broadband speed (title needs rework)}
+\title{Working from home and digital divides: resilience in the time of a
+pandemic}
 
@@
 at least in comparison to the pre-Covid era, and broadband services and
-infrastructure must be fit for purpose. \textbf{LET'S LEAVE IT FOR NOW,
-BUT I THINK WE CAN CRYSTALISE MORE THE RQ}
+infrastructure must be fit for purpose.
 
@@
 
-In this analysis, the terms `telecommuting' and `working from home' are
-used interchangeably, as most remote labour during the Covid-19 crisis
-was carried out in the homes of individual employees rather than any
-other location \citep{eurofound2020}. However, it should be noted that
+In this analysis, the terms `telecommuting' and `working from home\^{}
+{[}See also the popular social media hashtag \#WFM{]}' are used
+interchangeably, as most remote labour during the Covid-19 crisis was
+carried out in the homes of individual employees rather than any other
+location \citep{eurofound2020}. However, it should be noted that
 previous research has explored how telecommuting can occur in other
@@
 which are discussed in detail in the relevant Section
-{[}4.2{]}{]}(\#sec:4.2), in order to explore how the different patterns
-might support or undermine efforts to work from home and maintain safe
-productivity and whether they reinforce existing spatial and social
-inequalities. This analysis enables us to provide a more nuanced
-understanding of how telecommuting and technology intersect at a time of
-extreme demand, and what lessons this time has for a future where
-telecommuting is likely to remain a common means of accessing work and
-broadband services, as well as infrastructure, must be fit for purpose.
+\protect\hyperlink{sec:4.2}{4.2}, in order to explore how the different
+patterns might support or undermine efforts to work from home and
+maintain safe productivity and whether they reinforce existing spatial
+and social inequalities. This analysis enables us to provide a more
+nuanced understanding of how telecommuting and technology intersect at a
+time of extreme demand, and what lessons this time has for a future
+where telecommuting is likely to remain a common means of accessing work
+and broadband services, as well as infrastructure, must be fit for
+purpose.
+
+Our analysis is not free of limitations as, for instance, although we do
+have individual data we cannot identify individuals or households and,
+therefore, we need to aggregate our data at the LAD level. However, such
+individual data could only be found though surveys which by definition
+result in limited samples and, importantly, less detailed data about the
+quality of internet connectivity.
 
@@
 figures and table provide a comprehensive overview of the quality and
-reliability of experienced broadband in different parts of the UK, the
-temporal clusters offering a novel approach to understanding spatial
-disparity.
+reliability of experienced broadband in different parts of the UK.
 
@@
 The second largest cluster, comprising \(126\) local authorities and
-over \(20\) million people, is cluster \(13\). Cluster \(13\) has the
-slowest aggregate mean upload speed of any of the clusters, and the
-second highest ratio of the standard deviation to the mean. This
-suggests that those living in local authorities in this cluster
-experienced some of the lowest quality broadband services in terms of
-upload speeds in the UK. However, the variation in upload speeds in
-cluster \(13\), which can be an indication of its reliability, does not
-seem to disproportionately affect the morning peak from
-\(9:00\)-\(10:59\), as upload speeds are, on average, only \(2.6\)\%
-slower than in the evening peak period between \(19:00\) and \(20:59\),
-when entertainment purposes are likely to be using the most bandwidth.
-In comparison, the five LADs that are home to \(343\) thousand people in
-cluster \(1\) not only experience the second slowest mean upload speeds
-and the highest ratio of standard deviation to the mean, but are also
-much more affected during the morning peak.
+over \(20\) million people, is cluster \(13\), which has the slowest
+aggregate mean upload speed of any of the clusters, and the second
+highest ratio of the standard deviation to the mean. This suggests that
+those living in local authorities in this cluster experienced some of
+the lowest quality broadband services in terms of upload speeds in the
+UK. However, the variation in upload speeds in cluster \(13\), which can
+be an indication of its reliability, does not seem to disproportionately
+affect the morning peak from \(9:00\)-\(10:59\), as upload speeds are,
+on average, only \(2.6\)\% slower than in the evening peak period
+between \(19:00\) and \(20:59\), when entertainment purposes are likely
+to be using the most bandwidth. In comparison, the five LADs that are
+home to \(343\) thousand people in cluster \(1\) not only experience the
+second slowest mean upload speeds and the highest ratio of standard
+deviation to the mean, but are also much more affected during the
+morning peak.
 
 Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
-LADs home to \(29.5\) million people, experience aggregate mean upload
+LADs home to \(29.5\) million people, experienced aggregate mean upload
 speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
 lower than the other three large clusters and most of the smaller
-clusters, suggesting a middling quality of service. However, the time
-profile for cluster \(6\) in Figure \ref{UpClusterL} shows that upload
-speeds are highest at \(6:00\) on a Monday morning and peak between
-\(23:00\) and midnight on Wednesday and Thursday, but tend to be lower
-during the working day. Indeed, experienced mean upload speeds in the
-morning peak are \(4.4\)\% lower than in the evening peak -- a greater,
-more noticeable change than any of the other large clusters experience,
-but smaller than any of the clusters with temporal profiles shown in
-Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have mean upload
-speeds under \(10\)Mb/s, but higher than clusters \(1\) and \(13\), and
-their standard deviation is not dissimilar. However, this masks great
-variation in when lower speeds are experienced, with the mean upload
-speeds much lower between \(9:00-10:59\) than between \(19:00-20:59\) in
-cluster \(8\), but slightly faster in the morning in cluster \(12\).
-
-\begin{figure}
-\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-5}
-\end{figure}
+clusters, suggesting a middling quality of service. However, the
+temporal profile for cluster \(6\) in Figure \ref{UpClusterL} shows that
+upload speeds are highest at \(6:00\) on a Monday morning and peak
+between \(23:00\) and midnight on Wednesday and Thursday, but tend to be
+lower during the working day. Indeed, experienced mean upload speeds in
+the morning peak are \(4.4\)\% lower than in the evening peak -- a
+greater, more noticeable change than any of the other large clusters
+experience, but smaller than any of the clusters with temporal profiles
+shown in Figure \ref{UpClusterS}. Clusters \(8\) and \(12\) also have
+mean upload speeds under \(10\)Mb/s, but higher than clusters \(1\) and
+\(13\), and their standard deviation is not dissimilar. However, this
+masks great variation in when lower speeds are experienced, with the
+mean upload speeds much lower between \(9:00-10:59\) than between
+\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning in
+cluster \(12\).
 
@@
 
-Using auxiliary regressions, we test whether the clusters that have
-higher mean speeds and more reliable services consist of LADs that are
-more urban, affluent, and / or more likely to benefit from a choice of
-high quality internet services. We also estimate which of our clusters
-are more likely to have a higher proportion of occupations which enable
-telecommuting because of the nature of the work. The results of these
-auxiliary regressions are presented in Table \ref{aux}. The dependent
-variable is the LAD cluster membership as described in the
+Using an auxiliary multinomial regression, we test whether the clusters
+that have higher mean speeds and more reliable services consist of LADs
+that are more urban, affluent, and / or more likely to benefit from a
+choice of high quality internet services. We also estimate which of our
+clusters are more likely to have a higher proportion of occupations
+which enable telecommuting because of the nature of the work. The
+results of the auxiliary regression are presented in Table \ref{aux}.
+The dependent variable is the LAD cluster membership as described in the
 \protect\hyperlink{sec:3}{methods and data} section and equation
@@
 
-First, the number of speed tests run per cluster inhabitant between
-\(9:00\)-\(10:59\) is an indication of satisfaction or at least a lack
-of concern over broadband speeds and quality of service. People in LADs
-assigned to clusters \(6\) and \(13\) were more likely to run speed
-tests (at a per capita basis) at this time of day, whilst people in
-cluster \(11\) were the least likely to do so. This high level of
-satisfaction matches the high quality and reliability of service we
-would expect from the analysis in Section {[}4.1{]}{]}(\#sec:4.1). Those
-in clusters \(2\), \(3\), \(9\) and \(10\) were also less likely to ran
-speed tests per capita than those in the other six multi-authority
-clusters. Cluster \(7\) was the only cluster with higher mean speeds and
-middle to high measures of reliability which was more likely to run more
-speed tests per capita. This might be a reflection of the choice of ISP
-and connection options in those LADs, as, along with cluster \(12\),
-those in cluster \(7\) were least likely to have Virgin Media
-connections. Fast Virgin Media cable connections have historically only
-been available to \(45\)\% of premises in the UK \citep{ofcom2016},
-where the more lucrative and competitive market originally attracted the
-cable TV provider. Those in clusters \(2\), \(3\), \(9\) and \(11\)
-benefit from a higher proportion of Virgin connections, which is an
-indication that people in these clusters are more likely to live in
-urban areas, with more choice of broadband services, or in other words,
-are more likely to be on the right side of the first level,
-infrastructure-based digital divide.
-
-We employ distance from London and for the nearest metropolitan area
-(including London) as two more variables depicting peripherality, urban
+First, we control for the number of speed tests run per cluster
+inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
+Virgin Media internet connections. Regarding the former, we expect
+people in LADs with more unreliable connections to test more often their
+internet speeds and the results validate our priors. Regarding the
+former, fast Virgin Media cable connections have historically only been
+available to \(45\)\% of premises in the UK \citep{ofcom2016}, where the
+more lucrative and competitive market originally attracted the cable TV
+provider. Those in clusters \(2\), \(3\), \(9\) and \(11\) benefit from
+a higher proportion of Virgin connections, which is an indication that
+people in these clusters are more likely to live in urban areas, with
+more choice of broadband services, or in other words, are more likely to
+be on the right side of the first level, infrastructure-based digital
+divide.
+
+Then, we employ distance from London and for the nearest metropolitan
+area (including London) as two variables depicting peripherality, urban
 structure and, potentially, first level digital divides. The broadband
@@
 also not in remote parts of the country, and are home to an estimated
-population of \(175\) thousand people between them. \textbf{move maps to
-here?}
+population of \(175\) thousand people between them.
+
+\begin{figure}
+\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-6}
+\end{figure}
 
@@
 part of a metropolitan area of governance - Tameside in Greater
-Manchester. This may be why those in cluster \(7\) are not served by
-Virgin Media. It also is a demonstration of the complexity of
+Manchester. This may explain why those in cluster \(7\) are not served
+by Virgin Media. It also is a demonstration of the complexity of
 experienced broadband upload speeds as captured by time-series
@@
 divide reinforces other inequities in cluster \(1\). In comparison,
-those cluster \(13\) are more likely to earn more and have a better
+those in cluster \(13\) are more likely to earn more and have a better
 education. Cluster \(13\) also has the most businesses per inhabitant,
 but is somewhere in the middle in terms of job density. If this is an
-indication of a high number of SMEs, it might give some explanation for
-the variable internet quality, considering that small businesses have
-not been seen as the most valuable customers for higher speed broadband
-services \textbf{insert ref}.
+indication of a high number of SMEs, it can explain the variable
+internet quality, considering that small businesses have not been seen
+as the most valuable customers for higher speed broadband services
+\textbf{insert ref}.
 
@@
 variables in Table \ref{aux}. It ranks third or fourth out of the eleven
-clusters in the table in terms of having a higher proportion of
-managerial, professional, and tech jobs, as well as higher earnings, but
-is fourth from bottom in terms of education. The LADs in cluster \(6\)
-are also diverse, with few truly remote areas, but many urban areas in
-both the North, including Birmingham, Leeds and Sheffield, and South,
-including twelve London Boroughs, and many suburban areas and smaller
-cities like Oxford and Cambridge. The capital cities of the other UK
-nations, Belfast, Cardiff and Edinburgh, are also in this cluster,
-suggesting perhaps that the lower level of reliability discussed in
-Section \protect\hyperlink{sec:4.1}{4.1} may be as a result of increased
-demand, e.g.~for telecommuting.
-
-In comparison, clusters \(9\) and \(11\), which benefit from the most
-resilient internet connections, are characterised by lower share of
-occupations that are\\
-may benefit from the ability to telecommute (e.g.~managerial and
-technical). Cluster \(9\) is more highly educated and higher earning
-than cluster \(11\), has fewer businesses per inhabitant and less job
-density. Indeed, cluster \(11\) has a negative coefficient for the NVQ4
-variable, whilst for cluster \(9\) the coefficient is positive, and
-cluster \(11\) has the second highest job density and cluster \(9\) the
-second lowest. The former might be an indication of the higher returns
-that resilient broadband infrastructure generates for cluster \(9\)
-versus cluster \(11\). Indeed, we did detect slightly more slowdown in
-the AM peak in cluster \(9\) than in cluster \(11\) (see Table
-\ref{up.cluster.descr}). Meanwhile, a propensity for a high job density
-might indicate more labour intensive economic structures in cluster
-\(11\) than in the LADs in cluster \(9\).
-
-The LADs in clusters \(9\) and \(11\) also have opposite signs for their
-likelihood of being in the South of England. Those in cluster \(9\) less
-likely to be in the South and those in cluster \(11\) more likely,
-although the coefficient for cluster \(9\) might be skewed by the
-presence of the Scottish cities of Glasgow and Aberdeen, as Scotland has
-a different economic profile than England. Still, both clusters consist
-of larger LADs in terms of population, including the urban centre of
-Manchester and three of the nine other districts of Greater Manchester,
-Nottingham, both Portsmouth and Southampton, eight London Boroughs, and
-four of the seven constituent authorities of the West Midlands Combined
-Authority. Among the \(65\) LADs are also a number of other tightly
-bounded urban areas, such as Blackpool, Ipswich, Norwich, Slough, and
-Stevenage, and urban areas at the centre of less confined districts like
-Burnley, the digital hub of Milton Keynes, and Northampton. Thus, our
-findings confirm the importance of geography and urban locations for
-being on the right side of the first level digital divide, and suggest
-that internet resilience cuts across other digital and economic divides.
+clusters in the table in terms of the likelihood of having a higher
+proportion of managerial, professional, and tech jobs, as well as higher
+earnings, but is fourth from bottom in terms of education. The LADs in
+cluster \(6\) are also diverse, with few truly remote areas, but many
+urban areas in both the North, including Birmingham, Leeds and
+Sheffield, and South, including twelve London Boroughs, and many
+suburban areas and smaller cities like Oxford and Cambridge. The capital
+cities of the other UK nations, Belfast, Cardiff and Edinburgh, are also
+in this cluster, suggesting perhaps that the lower level of reliability
+discussed in Section \protect\hyperlink{sec:4.1}{4.1} may be as a result
+of increased demand, e.g.~for telecommuting.
+
+In comparison, clusters \(9\) and \(11\), which enjoy the most resilient
+internet connections, are characterised by higher share of occupations
+that may benefit from the ability to telecommute (e.g.~managerial).
+Also, LADs in Cluster \(9\) are more likely to host highly educated
+individuals with higher earning than cluster \(11\), which has fewer
+businesses per inhabitant and lower job density. Indeed, cluster \(11\)
+has a negative coefficient for the NVQ4 variable, whilst for cluster
+\(9\) the coefficient is positive, and cluster \(11\) has the second
+highest job density and cluster \(9\) the second lowest. The former
+might be an indication of the higher returns that resilient broadband
+infrastructure can generate for cluster \(9\) versus cluster \(11\).
+Indeed, we did detect slightly more slowdown in the AM peak in cluster
+\(9\) than in cluster \(11\) (see Table \ref{up.cluster.descr}).
+Meanwhile, a propensity for a higher job density indicate higher
+urbanization rate in LADs in cluster \(11\) than those in cluster \(9\).
+
+LADs in cluster \(9\) are also less likely to be in the South and those
+in cluster \(11\) more likely, although the coefficient for cluster
+\(9\) might be skewed by the presence of the Scottish cities of Glasgow
+and Aberdeen, as Scotland has a different economic profile than England.
+Still, both clusters consist of larger LADs in terms of population,
+including the urban centre of Manchester and three of the nine other
+districts of Greater Manchester, Nottingham, both Portsmouth and
+Southampton, eight London Boroughs, and four of the seven constituent
+authorities of the West Midlands Combined Authority. Among the \(65\)
+LADs are also a number of other tightly bounded urban areas, such as
+Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at
+the centre of less confined districts like Burnley, the digital hub of
+Milton Keynes, and Northampton. Thus, our findings confirm the
+importance of geography and urban locations for being on the right side
+of the first level digital divide, and suggest that internet resilience
+cuts across other digital and economic divides.
 
@@
 Midlands, and Fenland in the East of England. This spatial diversity
-demonstrates that the temporal clustering of internet resilience does
-not create spatial clusters, and digital divides do not necessarily
+demonstrates that the temporal clustering of internet resilience is not
+necessarily spatially dependent, and digital divides do not necessarily
 overlap with economic ones.
@@
 
-Upload speeds are not used as benchmarks in the UK, nor have been seen
-as integral to universal service, considering there has never before
-been such extreme demand for telecommuting and operations such as video
-calls.
+This paper offers a new perspective on telecommuting from the viewpoint
+of the complex web of digital divides. Good and reliable internet
+connections are necessary for the population to be able to work from
+home, but at the same time not every place hosts individuals in
+occupations which allow for telecommuting and not every place hosts
+individuals with the necessary skills to effectively use the internet
+and work form home. Importantly, not every place can utilise good
+internet connectivity -- even if they have them -- in order to achieve
+economic resilience in a period like the current pandemic, but also for
+a future when working from home might be a more common activity.
+
+Contrary to the previous literature, we employ experienced upload speeds
+as a benchmark of internet resilience concerning the capacity to work
+from home. Upload speeds are not used as benchmarks in the UK, nor have
+been seen as integral to universal service, considering there has never
+before been such extreme demand for telecommuting and operations such as
+video calls. While download speeds are associated more with video
+streaming services, fast upload internet speed are necessary for
+seamless video calls and uploading large files, activities associated
+with work. Given the granular nature of our data, we are able to
+consider upload internet speeds during working hours. To analyse these
+data we employ time series clustering methods, a family of unsupervised
+machine learning techniques which are not often utilised in geographical
+research.
+
+Using the UK as a case study, our analysis illustrated a nuanced
+picture, which is the outcome of the intersection between the different
+digital divides and the potential to telecommute. While some LADs appear
+to benefit both from reliable internet connections and population which
+can work from home and, in essence, capitalise the digital
+infrastructure, other LADs are suffering by both unreliable internet and
+low share of individuals who can potentially telecommute.
+
+\begin{itemize}
+\item
+  different places have different levels of preparedness, some places
+  have some ground to cover
+\item
+  benefits
+\end{itemize}
 
@@
 patters, to changes in land use and urban planning to accommodate people
-who work from home \citep{BUDNITZ2020102713}\textbf{also 2020 Swedish
-article from JTG}; and from productivity and innovation changes, to
-changes in agglomeration externalities and the attraction of large
-cities \citep{econobs} just to name a few. This paper is positioned to
-support endeavours in understanding the effects of increased
-telecommuting by exposing the spatial and social dimensions of
-telecommuting including the resilience of broadband speeds in terms of
-both quality and reliability of service, and whether this reinforces or
-redresses prior digital divisions. \textbf{took this bit you wrote to
-put in the discussion at the end?}
+who work from home \citep{BUDNITZ2020102713, ELLDER2020102777} and from
+productivity and innovation changes, to changes in agglomeration
+externalities and the attraction of large cities \citep{econobs} just to
+name a few. The take home message from this paper is that working form
+home and, consequently, its long term effects cannot be understood, set
+aside plan our cities for, without considering the underpinning digital
+divides and how telecommuting intersects with them.
 
@@
 \hline \\[-1.8ex] 
-pop, 2018 & 365 & 1.137 & 5.726 & 0.400 & 0.700 & 0.930 & 110.110 \\ 
-job density, 2018 & 365 & 53.269 & 57.700 & 0.150 & 22.050 & 69.290 & 544.090 \\ 
-distance to nearest met. area & 365 & 201.558 & 173.634 & 0.150 & 76.180 & 278.880 & 1,003.950 \\ 
-distance to London & 365 & 0.463 & 0.499 & 0 & 0 & 1 & 1 \\ 
-South of the UK & 363 & 12.009 & 4.013 & 3.600 & 9.000 & 14.300 & 27.900 \\ 
-managerial jobs, 2020 & 364 & 14.505 & 4.057 & 3.500 & 11.800 & 16.900 & 29.600 \\ 
-tech jobs, 2020 & 358 & 10.513 & 3.764 & 1.000 & 8.025 & 12.500 & 21.600 \\ 
-skilled trade jobs, 2020 & 364 & 21.223 & 6.902 & 4.400 & 16.775 & 24.850 & 71.600 \\ 
-professional jobs, 2020 & 359 & 9.965 & 2.738 & 3.200 & 8.100 & 11.400 & 21.300 \\ 
-administrative jobs, 2020 & 362 & 9.261 & 2.827 & 2.800 & 7.300 & 11.400 & 17.800 \\ 
-leisure jobs, 2020 & 337 & 6.339 & 2.847 & 1.200 & 4.400 & 7.900 & 19.800 \\ 
-machine operation jobs, 2020 & 360 & 592.184 & 81.129 & 437.600 & 534.625 & 633.875 & 893.200 \\ 
-earnings, 2019 & 365 & 39.329 & 11.076 & 15.000 & 31.800 & 45.300 & 100.000 \\ 
-NVQ4+ & 365 & 0.152 & 0.141 & 0.000 & 0.018 & 0.241 & 0.753 \\ 
-Virgin Media \%, 2020 & 365 & 0.057 & 0.164 & 0.023 & 0.038 & 0.056 & 3.174 \\ 
-n. business est. per hab., 2019 & 365 & 0.0005 & 0.0002 & 0.0001 & 0.0003 & 0.001 & 0.001 \\ 
+pop, 2018 & 365 & 174,952.100 & 119,557.100 & 8,700 & 100,400 & 214,900 & 1,141,400 \\ 
+job density, 2018 & 365 & 1.137 & 5.726 & 0.400 & 0.700 & 0.930 & 110.110 \\ 
+distance to nearest met. area & 365 & 53.269 & 57.700 & 0.150 & 22.050 & 69.290 & 544.090 \\ 
+distance to London & 365 & 201.558 & 173.634 & 0.150 & 76.180 & 278.880 & 1,003.950 \\ 
+south of the UK & 365 & 0.463 & 0.499 & 0 & 0 & 1 & 1 \\ 
+managerial jobs, 2020 & 363 & 12.009 & 4.013 & 3.600 & 9.000 & 14.300 & 27.900 \\ 
+tech jobs, 2020 & 364 & 14.505 & 4.057 & 3.500 & 11.800 & 16.900 & 29.600 \\ 
+skilled trade jobs, 2020 & 358 & 10.513 & 3.764 & 1.000 & 8.025 & 12.500 & 21.600 \\ 
+professional jobs, 2020 & 364 & 21.223 & 6.902 & 4.400 & 16.775 & 24.850 & 71.600 \\ 
+administrative jobs, 2020 & 359 & 9.965 & 2.738 & 3.200 & 8.100 & 11.400 & 21.300 \\ 
+leisure jobs, 2020 & 362 & 9.261 & 2.827 & 2.800 & 7.300 & 11.400 & 17.800 \\ 
+machine operation jobs, 2020 & 337 & 6.339 & 2.847 & 1.200 & 4.400 & 7.900 & 19.800 \\ 
+earnings, 2019 & 360 & 592.184 & 81.129 & 437.600 & 534.625 & 633.875 & 893.200 \\ 
+NVQ4+ & 365 & 39.329 & 11.076 & 15.000 & 31.800 & 45.300 & 100.000 \\ 
+Virgin Media \%, 2020 & 365 & 0.152 & 0.141 & 0.000 & 0.018 & 0.241 & 0.753 \\ 
+n. business est. per hab., 2019 & 365 & 0.057 & 0.164 & 0.023 & 0.038 & 0.056 & 3.174 \\ 
+AM tests per hab., 2020 & 365 & 0.0005 & 0.0002 & 0.0001 & 0.0003 & 0.001 & 0.001 \\ 
 \hline \\[-1.8ex] 

### 2020-12-28 paper/v1_sage/broadband.speed.covid.Rmd
Lots of edits, but not sure if it's shorter!

@@
 However, this ability has not been evenly distributed socially or spatially,
-creating a new type of digital divide.
+creating new intersections of economic and digital divisions, never tested before.
 On one side are those who can work from home, supported by digital technologies,
@@
 workers, who are potentially at high risk of infection.
-Whilst the basis for this new digital divide has been viewed as mainly occupational,
-here we consider whether the divide is also technological.
+Whilst the basis for this pandemic-generated divide has been viewed as mainly 
+occupational, here we consider whether it is also technological.
 
@@
 reliability of internet service, as reflected in *experienced* internet speeds,
-may reinforce or redress the spatial and social dimensions of the digital division
-exposed by the pandemic.
+may reinforce or redress the spatial and social dimensions of digital divisions.
 To do so, we employ volunteered geographic data on individual broadband speed tests
@@
 @ons2020lm2019].
-A back of the envelope calculation suggests that up to 40% of the working force
+A back of the envelope calculation suggests that up to 40% of the labour force
 could work from home on an ongoing basis [@batty2020editorial].
 Similar figures have been reported for other countries [@felstead2020homeworking].
-For instance, $37$% of the European workforce worked from home in April $2020$ with
+Approximately $37$% of the European workforce worked from home in April $2020$ with
 countries like Finland reaching $60$% [@eurofound2020].
@@
 performed entirely from home [@NBERw26948].
-
-<!-- ICT infrastructure and digital divides-->
 None of these changes could have happened in the absence of reliable information
@@
 society^[See for example the huge success of videoconferencing apps such as Zoom
-[@marks2020zoom].],
-the same does not apply for ICT hardware infrastructure such as internet
-broadband connectivity.
+[@marks2020zoom].], the same does not apply for ICT hardware infrastructure such as 
+internet broadband connectivity.
+
+<!-- ICT infrastructure and digital divides-->
 The literature describes first level digital divides in terms of the
@@
 assumes skills or functions that are present in some occupations but not in others,
-and relies upon access to high quality internet services.
-
-As the quality of internet infrastructure and services, as well as variation in
-occupations are spatially dependent and clustered in space, our approach offers a
-framework for understanding the impact of and interactions between
-the different levels of digital division in different places
-with different characteristics.
-By asking how resilient broadband speeds, and particularly upload speeds are
-as experienced in different parts of the UK during a time of extreme demand,
+and relies upon access to high quality and reliable internet services.
+
+As the quality of internet infrastructure and services, as well as the concentration
+of different occupations are spatially dependent and clustered in space, 
+our approach offers a framework for understanding the impact of and interactions 
+between digital divisions geographically and socioeconomically.
+By asking how resilient broadband speeds, and particularly upload speeds, are
+experienced in different parts of the UK during a time of extreme demand,
 we interrogate which places benefit from the greater economic resilience
@@
 First we review the literature on telecommuting and digital divides to better
-understand the structural and spatial development of these practices pre-pandemic,
+understand their structural and spatial development pre-pandemic,
 and thus their importance to the economic resilience of different places.
@@
 Our results section first offers a classification of how internet services vary
-across the UK local authorities and then assesses whether these clusters replicate
-or repudiate other socio-economic and geographic patterns of economic resilience.
-The paper ends with a conclusions section.
+across clusters of UK local authorities and then assesses whether these clusters 
+replicate or repudiate other socio-economic and geographic patterns of economic 
+resilience.
+We conclude with a discussion of the insights we've gained from our new perspective 
+on digital divisions.
 
@@
 These returns have been even greater during the Covid-19 crisis,
-when those with the ability to telecommute also have the ability to maintain
+when those with the capability to telecommute also have the ability to maintain
 their employment whilst protecting their health.
 However, the success of these arrangements has been dependent upon the first
-level digital divide, which is associated with access and quality of internet
-connectivity at a time of extremely high demand.
+level digital divide, which is associated with access to and quality of internet
+connectivity.
 @SALEMINK2017360 provides a systematic review of the pre-pandemic,
@@
 more data of all sorts and use their connections for a variety of purposes
-[@hauge2011consumer],
-and that there is a correlation between access to internet services and a
+[@hauge2011consumer].
+There is also a correlation between access to internet services and a
 reduction in household transport spend [@bris2017ict].
-Whether the implication is more internet use and less travel
+Independently of whether the additional internet use and reduced travel is
 because of increased telecommuting, these studies suggest that better internet
@@
 the spread of infection.
-We seek to answer how internet service resilience contributes to or reduces
-economic resilience when the latter is dependent upon
-the capability to work from home.
-We also aim to improve our understanding of the impact
-of first level digital division on telecommuting, and whether this results in
-much more fundamental third level digital division than has previously been perceived.
+We seek to answer how internet service resilience contributes to or reduces economic 
+resilience when the latter is dependent upon the capability to work from home.
+We also aim to improve our understanding of the impact of first level digital 
+division on telecommuting, and whether this results in much more fundamental third 
+level digital division than has previously been perceived.
 
@@
 and the economic geography of the UK.
-Following the regional economic resilience literature, which
-underlines the differentiated capacity of cities and regions to escape or
-recover from economic crises [@martin2012regional; @kitsos2018economic],
-different places have different industrial and occupational profiles,
-and these affect the aggregated potential capacity of places for telecommuting.
+Following the regional economic resilience literature, which underlines the 
+differentiated capacity of cities and regions to escape or recover from economic 
+crises [@martin2012regional; @kitsos2018economic], different places have different 
+industrial and occupational profiles, and these affect the aggregated potential 
+capacity of places for telecommuting.
 Such profiles are associated with longstanding inequalities in the UK and their
@@
 interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
-Not only are all three levels of digital divides associated, to a certain extent,
-with or shaped by the geography of the UK, but the intersection of the digital and
-material divides affects the capacity of places to overcome
-some of the economic effects of the Covid-19 pandemic.
+Not only are all three levels of digital divides associated with or shaped by the 
+geography of the UK, but the intersection of the digital and material divides 
+affects the capacity of places to overcome some of the economic effects of the 
+Covid-19 pandemic.
 Importantly, this is the first time that digital technologies became an essential
@@
 <!-- clustering in general -->
-The starting point of our methodological framework is cluster analysis, which can
+First we identified our methodological framework as cluster analysis, which can
 be defined within the modern machine learning framework as an unsupervised
@@
 [@gordon1977classification; @everitt1974cluster].
-For instance, clustering methods are the basis of *geodemographics*, a research domain
-which aims to create small area indicators or typologies of neighbourhoods
-based on various and sometimes diverse variables
-[@SINGLETON2009289; @harris2005geodemographics].
+For instance, clustering methods are the basis of *geodemographics*, a research 
+domain which aims to create small area indicators or typologies of neighbourhoods
+based on diverse variables [@SINGLETON2009289; @harris2005geodemographics].
 Clustering techniques have also been employed to solve *regionalisation* problems
 [@niesterowicz2016].
-
-<!-- time-series clustering -->
 Common characteristics of these studies are the cross-sectional nature of the data
@@
 fixed in time.
-However, for this paper we are interested in internet speeds, which vary over time
-and, therefore, create clusters of local authorities in the UK with similar temporal
-signatures of experienced internet speeds.
-Hence, we deviate from the established geographical clustering tools and
-employ time-series clustering methods.
-
+However, for this paper we are interested in internet speeds, which vary over time.
+We therefore create clusters of local authorities in the UK with similar temporal
+signatures of experienced internet speeds, and employ time-series clustering methods.
+
+<!-- time-series clustering -->
 Time-series clustering methods have been developed in order to deal with clustering
@@
 
-<!-- Time-series clustering methods utilising the whole length of time-series can be grouped in three categories.
-The first -- model-based approaches -- is based on recovering the underlying model for each time-series and then applying clustering algorithms on the model parameters of each time-series [@aghabozorgi2015time].
-The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].The second approach is based the formation of vectors of features derived from the original time-series.
-These new data of reduced dimensionality are then clustered using conventional clustering algorithms.
--->
-
 For this paper we utilise a category of time-series clustering methods
@@
 their shapes through the calculation of distances between the shapes,
-and are thus better equipped to capture similarities between short length
-time-series [@aghabozorgi2015time].
-This approach serves best this paper because (i) we identify clusters of
-UK local authorities with similar temporal signatures -- i.e. shapes -- of
-experienced internet speeds and (ii) the length of our time-series is short
-(see the data discussion in this section).
-
-Another import element of time-series clustering is the actual clustering algorithm.
-Similar to the clustering of cross-sectional data, we can employ partitioning
-algorithms, which lead to non-overlapping clusters, hierarchical clustering,
-which classifies clusters at different levels, and fuzzy algorithms, which
-create overlapping clusters [@sardatime].
-Because of the simplicity of the implementation and the interpretability of the
-results, we utilise here partitioning clustering based on the widely used *k*-means
-algorithm.
+and are better equipped to capture similarities between short length time-series 
+[@aghabozorgi2015time], such as our data.
+We thus identify clusters of UK local authorities with similar temporal signatures --
+i.e. shapes -- of experienced internet speeds.
+The clusters are identified using a partitioning algorithm, which leads to 
+non-overlapping clusters, and particularly the widely use *k*-means algorithm because
+of the simplicity of the implementation and the interpretability of the
+results.
 This is an iterative algorithm, which begins with defining the desired number of
@@
 <!-- difference between clustering cross-sectional and time-series data -->
-There are a number of differences between the above described application of
-*k*-means for cross-sectional data and its application for times series data.
+There are a number of differences between the application of *k*-means for 
+cross-sectional data and its application for times series data.
 Instead of creating clusters around centroids, a common approach is to create
@@
 A common distance measure for shape-based time-series clustering is
-Dynamic Time Warping (DTW).
-Using its underpinning dynamic programming algorithm, DTW compares two time-series
-objects to find the optimum warping path between them.
+Dynamic Time Warping (DTW), an algorithm comparing two time-series objects to find 
+the optimum warping path between them.
 DTW is widely used in order to overcome limitations linked to the use of Euclidean
@@
 <!-- introduction of our broadband speeds data -->
-To assess the quality and reliability of internet across local authorities
-in the UK during the time when the population were told to work from home if at
-all possible we utilise unique data comprising individual internet speed tests from
+To assess the internet quality and reliability across local authorities in the UK, 
+we utilise unique data comprising individual internet speed tests from
 Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
@@
 These data have been used before to assess digital divides [@riddlesden2014broadband]
-and the impact of local loop unbundling regulatory processes [@nardotto2015unbundling].
-By using volunteered geographic data, we are able to assess the *experienced*
-internet speed by users, which may differ from the *advertised* maximum speeds of ISPs.
-
+and the impact of local loop unbundling regulatory processes 
+[@nardotto2015unbundling], and we followed the former's approach to remove outliers. 
+By using this volunteered geographic data, we are able to assess the internet speed 
+*experienced* by users, which may differ from the maximum speeds *advertised* by 
+ISPs.
 We are particularly interested in upload speeds and the frequency of speed tests
 over the period from March to May $2020$, as government statements indicate this is
-when UK workers were told to work from home if at all possible [@GovUK2020].
+when UK workers were first told to work from home if at all possible [@GovUK2020].
 Average upload speeds are slower than average download speeds, at $9.3$Mb/s
@@
 uploading documents or two-way audio, video, and text-based communication systems.
-Meanwhile, the frequency of speed tests was important in identifying
-the temporal profile which would give us most insight into
-experienced internet service and resilience, and provide
-an indication of the volume of experience over particular units of time.
 
 <!-- data transformation -->
-The first step in the workflow after dropping some outliers following
-@riddlesden2014broadband was to transform the individual, geolocated and time-stamped
-tests to more meaningful aggregates both in terms of space and time.
-The frequency of testing indicates that whilst there is an overall trend of increased
-testing from March to April and then a slight reduction from April to May,
+The frequency of speed tests was important in identifying the temporal profile which 
+would give us most insight into experienced  internet service and resilience over 
+units of time.
+Our analysis indicates that whilst there is an overall trend of increased testing 
+from March to April and then a slight reduction from April to May,
 this trend masks substantial variation by not only the day of the week,
@@
 Furthermore, the importance of this variation is highlighted by a comparison
-with the same period in $2019$, as in Figure \ref{test2019},
-when the volume of testing and thus of experience of internet service quality
-was much more tied to the evening leisure peak and presumably to download speeds.
-Since the increase in testing during the working day in $2020$ is an indication that
-users have a greater perception of the variable quality of internet service,
-particularly during a new morning peak of testing for service reliability,
-we decided to include a measure of hourly variation in our temporal profiles.
-
-However, although this is a large data set -- $241,088$ individual tests performed
-during weekday hours across the study period --
-there are not enough observations for each Local Authority District (LAD)
-and for each working hour of each working day -- $631$ speed tests per LAD on average
--- to profile speeds at that level of detail.
-Therefore, we aggregate all the speed-checks during the $13$ weeks of March to May
-inclusive for weekdays in $2020$ by each hour of the day and day of the week.
+with the same period in $2019$, as in Figure \ref{test2019}, when the volume of 
+testing and thus of experience of internet service quality peaked in the evening, 
+presumably in response to demand for leisure activities and download speeds.
+In contrast, the majority of the increase in testing in $2020$ is during the working
+day, creating a new morning peak in Figure \ref{test2020}.
+Therefore, we include a measure of hourly variation in our temporal profiles to
+reflect the change in users' perception of the workday reliability of internet 
+services.
+
+However, we judged that there were insufficient observations -- only $631$ speed 
+tests per LAD on average -- for each for each working hour of each working day in 
+each Local Authority District (LAD) to profile speeds at that level of detail.
+Spatial aggregation was also necessary because we could not follow individuals or
+households and connect data points.
+Therefore, we aggregate the $241,088$ individual, geolocated and time-stamped 
+speed-checks during the $13$ weeks of March to May inclusive for weekdays in $2020$ 
+by each hour of the day and day of the week.
 As our research aims to identify the geography of internet service resilience for
@@
 $90$ time points per series.
-We also aggregate these data spatially because we could not follow individuals or
-households and connect data points.
 The time-series were calculated for each of the $382$ LADs in the UK, standardised,
@@
 In Section [4.1](#sec:4.1), we review the temporal profile of upload speed
-by hour of the day and day of the composite week, as well as
-the experienced speed characteristics of each cluster.
+by hour of the day and day of the composite week for each cluster.
 Since the quality and reliability of internet services vary in time and space due
-to both supply and demand-side influences, we use a number of different measures
+to both supply and demand-side influences, we also use a number of different measures
 to describe experienced upload speeds per cluster.
@@
 b) standard deviation or the amount of fluctuation from the mean, and
-c) the variation in speeds at particular times of day when working from home is
-more likely to take place.
+c) the variation in speeds during the new morning peak of testing when working 
+from home is more likely to take place.
 We take account of all three measurements in our descriptive statistics of
@@
 <!-- link to auxillary regressions method -->
-The cause of these different experiences of broadband resilience may be different in
+The cause of these different experiences of broadband resilience may vary across
 different areas, as they may reflect either similarities in patterns of demand
@@
 temporal patterns of internet service relate to the economic geography of the UK.
-More specifically, we estimate the following multinomial model:
+More specifically, we estimate the following multinomial logit model:
 
@@
 socio-economic and geographic variables, which are discussed in detail in
-the relevant Section [4.2](#sec:4.2), in order to explore how the different patterns
-might support or undermine efforts to work from home and maintain safe productivity
-and whether they reinforce existing spatial and social inequalities.
-This analysis enables us to provide a more nuanced understanding of how
-telecommuting and technology intersect at a time of extreme demand, and what
-lessons this time has for a future where telecommuting is likely to remain a
-common means of accessing work and broadband services, as well as infrastructure,
-must be fit for purpose.
-
+the relevant Section [4.2](#sec:4.2).
 <!-- limitations-->
-Our analysis is not free of limitations as, for instance, although we do have
-individual data we cannot identify individuals or households and, therefore, we
-need to aggregate our data at the LAD level. However, such individual data could
-only be found though surveys which by definition result in limited samples
-and, importantly, less detailed data about the quality of internet connectivity.
+Although we do have individual data, we cannot identify individuals or households 
+and, therefore, we need to aggregate our data at the LAD level. 
+Therefore, our results offer only correlations between the socioeconomic 
+characteristics of certain geographic locations and internet service quality,
+not a record of who was using the internet services to telecommute and maintain
+safe productivity.
+Such individual data could be found though surveys, but these offer less 
+detailed data about the experience of internet resilience due to enforced, 
+rather than leisure demand, which is the main contribution of this paper.
+Our auxiliary regression, therefore, provides no more than an indication of how 
+internet connectivity can reinforce or redress existing spatial and social 
+inequalities in different places.
+However, it opens a path to future research by highlighting the importance of  
+understanding of how telecommuting capabilities and digital infrastructure divisions 
+intersect.
 
@@
 some of the lowest quality broadband services in terms of upload speeds in the UK.
-However, the variation in upload speeds in cluster $13$,
-which can be an indication of its reliability, does not seem to disproportionately
-affect the morning peak from $9:00$-$10:59$, as upload speeds are, on average,
+However, Figure \ref{UpClusterL} shows that the high standard deviation, 
+which is one indication of unreliable internet, did not disproportionately affect 
+the morning peak from $9:00$-$10:59$, when upload speeds were, on average, 
 only $2.6$% slower than in the evening peak period between $19:00$ and $20:59$, 
 when entertainment purposes are likely to be using the most bandwidth.
-In comparison, the five LADs that are home to $343$ thousand people in cluster $1$
-not only experience the second slowest mean upload speeds
-and the highest ratio of standard deviation to the mean,
-but are also much more affected during the morning peak.
-
-Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs home to $29.5$
+In comparison, the five LADs in cluster $1$, home to $343$ thousand people, not only 
+experience the second slowest mean upload speeds and the highest ratio of standard 
+deviation to the mean, but are also much more affected during the morning peak.
+
+Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs and $29.5$
 million people, experienced aggregate mean upload speeds of about $1$Mb/s faster
@@
 and most of the smaller clusters, suggesting a middling quality of service.
-However, the temporal profile for cluster $6$ in Figure \ref{UpClusterL} shows
-that upload speeds are highest at $6:00$ on a Monday morning
-and peak between $23:00$ and midnight on Wednesday and Thursday,
-but tend to be lower during the working day.
-Indeed, experienced mean upload speeds in the morning peak are $4.4$% lower
+The temporal profile for cluster $6$ in Figure \ref{UpClusterL} shows that upload 
+speeds are highest at $6:00$ on a Monday morning and peak between $23:00$ and 
+midnight on Wednesday and Thursday, but tend to be lower during the working day.
+Furthermore, experienced mean upload speeds in the morning peak are $4.4$% lower
 than in the evening peak -- a greater, more noticeable change than any of the other
-large clusters experience, but smaller than any of the clusters with temporal
+large clusters experience, suggesting poorer reliability during the working day.
+This difference is less, however than any of the clusters with temporal
 profiles shown in Figure \ref{UpClusterS}.
+
 Clusters $8$ and $12$ also have mean upload speeds under $10$Mb/s, but higher than
-clusters $1$ and $13$, and their standard deviation is not dissimilar.
-However, this masks great variation in when lower speeds are experienced,
-with the mean upload speeds much lower between $9:00-10:59$ than
-between $19:00-20:59$ in cluster $8$, but slightly faster in the morning in
-cluster $12$.
-
-Indeed cluster $12$ is the only cluster to experience higher speeds in the evening
-peak, compared to the morning peak, suggesting that widespread telecommuting
-has generally changed the temporal profile of internet activity throughout the UK.
-Yet even if all but one cluster is showing slower speeds in the morning than
-the evening, the reliability of internet services in different clusters
+clusters $1$, $13$, and $6$. However, mean upload speeds are much lower between 
+$9:00-10:59$ than between $19:00-20:59$ in cluster $8$, but slightly faster in the 
+morning than in the evening in cluster $12$.
+Indeed, cluster $12$ is the only cluster to experience higher speeds in the morning
+peak compared to the evening, and thus the only cluster where the temporal profile 
+of internet use is closer to what might have been expected pre-pandemic.
+And yet even among the other clusters, the reliability of internet services
 during the working day still varies considerably.
@@
 have the most reliable internet services.
-The ratio of standard deviation to mean in both these clusters is below 55%,
-and the ratio of upload speeds in the peak periods indicates that speeds are
-only about 2% slower in the morning than the evening.
+The ratio of standard deviation to mean in both these clusters is lowest,
+and speeds are only about 2% slower in the morning than the evening.
 Mean speeds are also higher than in any other cluster, excluding cluster $2$,
@@
 In Figure \ref{UpClusterL}, cluster $11$ shows more noticeable peaks and troughs,
-but the lowest points are not at the peak times described in Table \ref{up.cluster.descr}.
-Rather, the slowest upload speeds on average occur between $6:00-7:00$ on
-Monday morning, $14:00-15:00$ on Friday, and $16:00-17:00$ on Thursday.
-These slowest times are still mostly faster than the average hourly upload speeds
-in cluster $13$.
-Finally, ignoring the smallest clusters in terms of population,
-that is clusters $4$ and $5$,
-clusters $3$, $7$ and $10$ also have relatively high mean upload speeds.
-Clusters $3$ and $10$ pair high mean speeds with low standard deviations
-relative to the mean speeds, suggesting reliability and resilience,
-as well as quality broadband services.
-Cluster $7$ has a higher ratio of standard deviation to mean, but there is less
-difference in average speeds between the morning and evening peaks than in clusters
-$3$ and $10$.
+but the lowest points are not during the morning peak, but occur, on average, 
+between $6:00-7:00$ on Monday morning, $14:00-15:00$ on Friday, and $16:00-17:00$ on 
+Thursday. These slow periods still offer better speeds than the average hourly 
+profile of cluster $13$. Finally, ignoring the smallest clusters in terms of 
+population, that is clusters $4$ and $5$, clusters $3$, $7$ and $10$ also have 
+relatively high mean upload speeds. Clusters $3$ and $10$ pair high mean speeds with 
+low standard deviations relative to the mean speeds, suggesting reliability and 
+resilience, as well as quality broadband services. Cluster $7$ has a higher ratio of 
+standard deviation to mean, but there is less difference in average speeds between 
+the morning and evening peaks than in clusters $3$ and $10$.
 
@@
 high standard deviations.
-With those in cluster $1$ in particular more likely to experience that poor
-reliability during the morning peak, is this first level digital divide occurring
-in areas where few are occupationally able to telecommute anyway, and what are the
-implications for economic resilience?
+With those in cluster $1$ more likely to experience that poor reliability during the 
+morning peak, is this first level digital divide occurring in areas where few are 
+occupationally able to telecommute anyway, and what are the implications for 
+economic resilience?
 
@@
 Using an auxiliary multinomial regression, we test whether the clusters that
-have higher mean
-speeds and more reliable services consist of LADs that are more urban, affluent,
-and / or more likely
-to benefit from a choice of high quality internet services.
+have higher mean speeds and more reliable services consist of LADs that are more 
+urban, affluent, and / or more likely to benefit from a choice of high quality 
+internet services.
 We also estimate which of our clusters are more likely to have a higher proportion of
@@
 Hambleton in North Yorkshire, a rural area of just over ninety thousand people.
-Mean, experienced upload speeds in cluster $4$ are close to both the average speeds
-for the $13$ clusters ($9.9$Mb/s) and the pre-clustered average for the whole sample
-($9.3$Mb/s) as shown in Table \ref{up.cluster.descr}.
+Mean, experienced upload speeds in cluster $4$ (see Table \ref{up.cluster.descr}) 
+are close to both the average speeds for the $13$ clusters ($9.9$Mb/s) and the 
+pre-clustered average for the whole sample ($9.3$Mb/s).
 However, the standard deviation for cluster $4$ and the difference between average
 speeds in the morning compared to the evening peak periods are indications of
-worse reliability than many of the other clusters.
-Hence, the results in Table \ref{aux} should be seen as relative
-rather than absolute probabilities.
+worse reliability than many of the other clusters. Hence, the results in Table 
+\ref{aux} should be seen as relative rather than absolute probabilities.
 
@@
 connections. Regarding the former, we expect people in LADs with more unreliable
-connections to test more often their internet speeds and the results validate our
-priors. Regarding the former, fast Virgin Media cable connections have historically
+connections to test their internet speeds more often, and the results validate our
+priors. Meanwhile, fast Virgin Media cable connections have historically
 only been available to $45$% of premises in the UK [@ofcom2016], where the more
@@
 or in other words, are more likely to be on the right side of the first level,
-infrastructure-based digital divide.
+infrastructure-based digital divide, as we expected from the analysis in Section 
+[4.1](#sec:4.1).
 
@@
 <!-- This high level of satisfaction matches the high quality and reliability of service -->
-<!-- we would expect from the analysis in Section [4.1](#sec:4.1). -->
 <!-- Those in clusters $2$, $3$, $9$ and $10$ were also less likely to ran speed tests per -->
@@
 
-Then, we employ distance from London and for the nearest metropolitan area (including
-London) as two variables depicting peripherality, urban structure and,
+Then, we employ distance from London and from the nearest metropolitan area 
+(including London) as two variables depicting peripherality, urban structure and,
 potentially, first level digital divides.
 The broadband speed tests run in the authorities in cluster $3$ are more likely to
-be taking place close to London than those run in any of the other clusters,
-a result that makes sense considering two of the four authorities in cluster $3$
-are the London commuter towns of Harlow and Luton.
+be taking place close to London than those run in any of the other clusters, and
+two of the four authorities in cluster $3$ are the London commuter towns of Harlow 
+and Luton.
 However, even though London was also included in the variable calculating distance
@@
 or Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
-Still, whilst significant, the effect is tiny and the other members of cluster $3$
-are Corby and Eastbourne, two towns which, whilst not part of metropolitan areas,
-are also not in remote parts of the country, and are home to an estimated population
-of $175$ thousand people between them.
+Thus, it is important to consider the membership of each cluster as well as the
+regression results. Corby and Eastbourne, the other two authorities in cluster $3$ 
+are, whilst not part of metropolitan areas, are also large, accessible towns.
 
@@
 no central urban boroughs and only one LAD that is part of a metropolitan area of
-governance - Tameside in Greater Manchester.
-This may explain why those in cluster $7$ are not served by Virgin Media.
+governance - Tameside in Greater Manchester. This may explain why those in cluster 
+$7$ are the second least likely to be served by Virgin Media.
 It also is a demonstration of the complexity of experienced broadband upload speeds
-as captured by time-series clustering, and their likely interaction with first level
-digital divides.
+as captured by time-series clustering, and how quality and reliability, not only 
+availability, engender first level digital divides.
 Cluster $1$, which our analysis suggests lacks broadband resilience, contains five,
 mainly rural authorities.
-However, they are scattered around the country --
-Ceredigion in West Wales, Darlington in the Northeast, Eden between the Lake District
-and the North Penines, Rossendale in the Northwest, and Rutland in the East Midlands
--- and therefore, the results in Table \ref{aux} indicate that these authorities
-are not as far from the centre of a metropolitan area as authorities in cluster $3$,
+However, they are scattered around the country -- Ceredigion in West Wales, 
+Darlington in the Northeast, Eden and Rossendale in the Northwest, and Rutland in 
+the East Midlands -- and therefore, the results in Table \ref{aux} indicate that 
+these authorities are closer to a metropolitan area than authorities in cluster $3$,
 but furthest from London.
 
-Internet resilience, as measured here, seems to be more nuanced than our --
-arguably crude -- dummy variable depicting the North-South economic divide either,
-which assigns $1$ to LADs locates in Greater London, Southeast, Southwest and East
-of England regions, and $0$ to the rest.
+Internet resilience as measured here is also more nuanced than our -- arguably crude 
+-- dummy variable depicting the North-South economic divide, which assigns $1$ to 
+LADs located in Greater London, Southeast, Southwest and East of England regions, 
+and $0$ to the rest.
 The authorities most likely to be in the South are those in cluster $13$,
@@
 and thus a low level of service.
-However, cluster $13$ does include some rural, remote areas of the country,
-such as Northwest Scotland, Cornwall and Powys in Wales as shown on Figure
+However, cluster $13$ includes some rural, remote areas of the country, such as 
+Northwest Scotland, Cornwall and Powys in Wales as shown on Figure 
 \ref{map.up.clusters}. 
-It also includes the major metropolitan centres of Bristol, Liverpool,
-and Newcastle, as well as nine (of $32$) London Boroughs,
-including the City of London and plenty of home county and suburban areas.
-Although by the standard deviation measure cluster $13$ has unreliable internet
-services, speed variation was less during the morning peak, suggesting that
-the estimates for reliability are inconsistent.
-Considering that this is one of the largest clusters, and thus the averages
-incorporate more noise than some of the smaller clusters,
+It also includes major metropolitan centres in the North -- Liverpool, Newcastle -- 
+and South -- Bristol, nine (of $32$) London Boroughs.
+There are also plenty of Southern home county and suburban areas.
+The standard deviation measure for cluster $13$ is high, but speed variation is 
+low during the morning peak, suggesting that the estimates for reliability are 
+inconsistent. Considering that this is one of the largest clusters, and thus the 
+averages incorporate more noise than some of the smaller clusters,
 it may be that the LADs in this cluster do not all suffer equally from
@@
 Yet we need to consider the results for other variables in Table \ref{aux} to better
-determine whether clusters $1$ and $13$, which appear to suffer most from a lack
-of quality internet services, are also more likely to have a low skilled workforce,
+determine whether clusters $1$ and $13$, which appear to suffer most from poor
+quality internet services, are also more likely to have a low skilled workforce,
 less able to benefit from telecommuting.
@@
 we could note that cluster $1$ also has a greater proportion of skilled trades
-than other clusters.
-Either way, it seems likely that the first level digital divide reinforces
-other inequities in cluster $1$.
-In comparison, those in cluster $13$ are more likely to earn more and have a better
-education.
+than other clusters. It seems likely that the first level digital divide reinforces
+other inequities in cluster $1$, whilst those in cluster $13$ are more likely 
+to earn more and have a better education despite poor internet services.
 Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the
-middle in terms of job density.
-If this is an indication of a high number of SMEs, it can explain
-the variable internet quality, considering that small businesses have not been
-seen as the most valuable customers for higher speed broadband services **insert ref**.
+middle in terms of job density. If this is an indication of a high number of SMEs, 
+it could explain the variable internet quality, considering that small businesses 
+have not been seen as the most valuable customers for higher speed broadband services
+if they are not located near residential customers [@ofcom2016].
 
 Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more
-noise within the averages we use to measure internet quality and reliability,
-and the result indicates an average level of service, even if better than LADs
-in cluster $13$.
-Cluster $6$ also falls towards the middle of the clusters on many of the socio-
-economic variables in Table \ref{aux}.
-It ranks third or fourth out of the eleven clusters in the table in terms of the
-likelihood of
-having a higher proportion of managerial, professional, and tech jobs, as well as
-higher earnings, but is fourth from bottom in terms of education.
-The LADs in cluster $6$ are also diverse, with few truly remote areas,
-but many urban areas in both the North, including Birmingham, Leeds and Sheffield,
-and South, including twelve London Boroughs, and many suburban areas and
-smaller cities like Oxford and Cambridge.
-The capital cities of the other UK nations, Belfast, Cardiff and Edinburgh, are
-also in this cluster, suggesting perhaps that the lower level of reliability
-discussed in Section [4.1](#sec:4.1) may be as a result of increased demand,
-e.g. for telecommuting.
-
-In comparison, clusters $9$ and $11$, which enjoy the most resilient
-internet connections, are characterised by higher share of occupations that
-may benefit from the ability to telecommute (e.g. managerial).
-Also, LADs in Cluster $9$ are more likely to host highly educated individuals with
-higher earning than cluster $11$, which has fewer businesses per inhabitant and
-lower job density.
-Indeed, cluster $11$ has a negative coefficient for the NVQ4 variable, whilst
-for cluster $9$ the coefficient is positive, and cluster $11$ has the second
-highest job density and cluster $9$ the second lowest.
-The former might be an indication of the higher returns that resilient
-broadband infrastructure can generate for cluster $9$ versus cluster $11$.
-Indeed, we did detect slightly more slowdown in the AM peak in cluster $9$ than
-in cluster $11$ (see Table \ref{up.cluster.descr}).
-Meanwhile, a propensity for a higher job density indicate higher urbanization
-rate in LADs in cluster $11$ than those in cluster $9$.
-
+noise within the averages we use to measure internet quality and reliability.
+Our results indicate average mean speeds, and Table \ref{aux} shows that 
+cluster $6$ also falls towards the middle of the clusters on many of the 
+socioeconomic variables. It ranks third or fourth out of the eleven in terms of 
+the likelihood of having a higher proportion of managerial, professional, and 
+tech jobs, as well as higher earnings, but is fourth from bottom for educational
+attainment. The LADs in cluster $6$ are also diverse, with few truly remote areas,
+but urban areas throughout England, including Birmingham, Leeds, Sheffield, 
+twelve London Boroughs, and many suburban areas and smaller cities like Oxford 
+and Cambridge. The capital cities of the other UK nations, Belfast, Cardiff and 
+Edinburgh, are also in this cluster, suggesting perhaps that the lower level of  
+reliability discussed in Section [4.1](#sec:4.1) may be due to increased demand, 
+e.g. for telecommuting, when it is not supported by the most resilient internet
+connections.
+
+In contrast, Clusters $9$ and $11$ enjoy resilient internet connections and are 
+characterised by higher share of occupations that may benefit from the ability 
+to telecommute (e.g. managerial). LADs in Cluster $9$ are more likely to host 
+highly educated individuals with higher earnings than cluster $11$, which has a 
+negative coefficient for the NVQ4 variable. Cluster $11$ has fewer businesses per 
+inhabitant and the second highest job density and cluster $9$ the second lowest.
+These coefficients might indicate that resilient broadband infrastructure 
+generates higher returns for those in cluster $9$, where slightly more slowdown in 
+the AM peak was detected (see Table \ref{up.cluster.descr}).
+<!--Meanwhile, a propensity for a higher job density could indicate a higher urbanization rate in LADs in cluster $11$ than those in cluster $9$. I DON't
+QUITE UNDERSTAND THIS SENTENCE, SO SINCE WE NEED TO CUT WORDS...?-->
 LADs in cluster $9$ are also less likely to be in the South and those in cluster $11$
 more likely, although the coefficient for cluster $9$ might be skewed by
-the presence of the Scottish cities of Glasgow and Aberdeen,
-as Scotland has a different economic profile than England.
-Still, both clusters consist of larger LADs in terms of population,
-including the urban centre of Manchester and three of the nine other districts of
-Greater Manchester, Nottingham, both Portsmouth and Southampton,
-eight London Boroughs, and four of the seven constituent authorities of the
-West Midlands Combined Authority.
-Among the $65$ LADs are also a number of other tightly bounded urban areas,
-such as Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas
-at the centre of less confined districts like Burnley,
+the presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a 
+different economic profile than England. Still, both clusters consist of larger 
+LADs in terms of population, including the city centre among four of the ten 
+districts of Greater Manchester, eight London Boroughs, four of the seven 
+constituent authorities of the West Midlands Combined Authority, Nottingham, and 
+both Portsmouth and Southampton. Among the $65$ LADs are also a number of other 
+tightly bounded urban areas, such as Blackpool, Ipswich, Norwich, Slough, and 
+Stevenage, and urban areas at the centre of less confined districts like Burnley, 
 the digital hub of Milton Keynes, and Northampton.
-Thus, our findings confirm the importance of geography and urban locations
-for being on the right side of the first level digital divide, and suggest that
-internet resilience cuts across other digital and economic divides.
-
-A similarly divided picture can be built for the next tier of clusters: $2$, $3$,
-$7$, $10$. LADs in these clusters experience higher than average mean speeds and
-rank high to middle on measures of service reliability.
-LADs in these clusters are on the right side of the first level digital divide.
+These urban locations are on the right side of the first level digital divide, 
+and the regression results suggest that internet resilience is supporting a wide
+range of small and large urban economies.
+
+LADs in clusters $2$, $3$, $7$, $10$ are also on the right side of the first level 
+digital divide, experiencing higher than average mean speeds and ranking high to 
+middle on measures of service reliability.
 However, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
-work force, which might more easily take advantage of working from home
-opportunities.
+work force to take advantage of working from home opportunities. 
 Cluster $2$ is comprised of just two LADs, with the lowest job density of any 
 cluster -- these are East Lothian near Edinburgh, and North East Lincolnshire, 
-home to Grimsby.
-Cluster $10$ is comprised of four LADs, including two peripheral suburban areas
-of Birmingham, the East London Borough of Newham, and Dundee.
-Suburbs are considered the most likely urban form in which telecommuters live,
-and Dundee has a reputation for tech startups **insert ref**.
-Yet although those in clusters $2$ and $10$ may be making the most of their
-resilient internet, those in cluster $7$ have the highest earnings.
-The four LADs in cluster $7$ include the suburbs South of Belfast, a suburban
-district of Greater Manchester, the villages between York and Leeds, and a
-couple of small West Midlands towns.
-LADs in cluster $3$ are associated more with lower skills and have the highest
-proportion of individuals in machine operation jobs.
-They include the Southeast commuter towns of Harlow and Luton and the slightly
-more distant towns of Eastbourne and Corby.
-Arguably, LADs in clusters $3$ benefit the least from their resilient internet
-infrastructure in an era when working from home became a vehicle for economic
-resilience.
+home to the town of Grimsby. Cluster $10$ is comprised of four LADs, including two 
+peripheral suburban areas of Birmingham, the East London Borough of Newham, 
+and Dundee. Suburbs are considered the most likely urban form in which telecommuters 
+live [@e2018does], and Dundee has a reputation for tech startups [@technation2017].
+LADs in cluster $7$, which include the suburbs South of Belfast, a suburban
+district of Greater Manchester, the villages between York and Leeds, and a couple of 
+small West Midlands towns have the highest earnings, and thus may also be making the 
+most of their resilient internet. In contrast, LADs in cluster $3$ are associated 
+more with lower skills and have the highest proportion of individuals in machine 
+operation jobs. Arguably, LADs in clusters $3$ benefit the least from their 
+resilient internet infrastructure in an era when working from home became a vehicle 
+for economic resilience.
 
 Finally, clusters $8$ and $12$, are characterised by LADs enjoying close to average
-mean upload speeds, but different patterns of internet reliability.
-LADs in cluster $8$ are likely to host more individuals employed in tech and
+mean upload speeds, but opposing patterns of internet reliability during the morning
+peak. LADs in cluster $8$ are likely to host more individuals employed in tech and
 administrative occupations than any other cluster, as well as many in managerial
@@
 residents, but are not significantly likely to earn more than other clusters.
-Including suburbs of Leicester, Cardiff, Birmingham, the green belt area south of
+Including suburbs near Leicester, Cardiff, Birmingham, the green belt area south of
 London straddling the M25 motorway, and rural West Devon, the LADs in cluster $8$
@@
 This density of businesses could be why cluster $12$ is the only cluster with higher
-speeds in the morning peak during the study period.
-If people are at home, business premises might well be abandoned.
-Cluster $12$ is home to $1.5$million people spread across twelve LADs, including
-the City of Westminster and the London Borough of Hammersmith.
-However, it also includes a LAD in Northern Ireland, two in Wales, two suburban areas
-North of London, Preston in the Northwest, East Staffordshire in the Midlands, and
-Fenland in the East of England.
-This spatial diversity demonstrates that the temporal clustering of internet
+speeds in the morning peak during the study period. If people are at home, business 
+premises might well have been abandoned. Cluster $12$ is home to $1.5$million people 
+spread across twelve LADs, including the City of Westminster but also a LAD in 
+Northern Ireland, two in Wales, Preston in the Northwest, and Fenland in the East of 
+England. This spatial diversity demonstrates that the temporal clustering of internet
 resilience is not necessarily spatially dependent, and digital divides do not
@@
 This paper offers a new perspective on telecommuting from the viewpoint of the
-complex web of digital divides. Good and reliable internet connections are
-necessary for the population to be able to work from home, but at the same time
-not every place hosts individuals in occupations which allow for telecommuting and
-not every place hosts individuals with the necessary skills to effectively use
-the internet and work form home. Importantly, not every place can utilise good
-internet connectivity -- even if they have them -- in order to achieve economic
-resilience in a period like the current pandemic, but also for a future when
-working from home might be a more common activity.
-
-Contrary to the previous literature, we employ experienced upload speeds as a
-benchmark of internet resilience concerning the capacity to work from home.
-Upload speeds are not used as benchmarks in the UK, nor have been seen as integral
-to universal service, considering there has never before been such extreme demand
-for telecommuting and operations such as video calls. While download speeds are
-associated more with video streaming services, fast upload internet speed are
-necessary for seamless video calls and uploading large files, activities associated
-with work. Given the granular nature of our data, we are able to consider upload
-internet speeds during working hours. To analyse these data we employ time series
-clustering methods, a family of unsupervised machine learning techniques which
-are not often utilised in geographical research.
-
-Using the UK as a case study, our analysis illustrated a nuanced picture, which
-is the outcome of the intersection between the different digital divides and the
-potential to telecommute. While some LADs appear to benefit both from reliable
-internet connections and population which can work from home and, in essence,
-capitalise the digital infrastructure, other LADs are suffering by both unreliable
-internet and low share of individuals who can potentially telecommute.
-
-- different places have different levels of preparedness, some places have some ground to cover
-
-- benefits
-
-Clusters $3$, $9$, $10$ and $11$ seem to benefit most from high quality and
-resilient broadband services.
-The dips in mean upload speeds in clusters $6$, $3$ and $10$ during the morning
-peak are suggestive of more use during the working day, and potentially more
-telecommuting.
+complex web of digital divides, and employs time series clustering methods, 
+a family of unsupervised machine learning techniques which are not often utilised 
+in geographical research. Fast, reliable internet connections are
+necessary for the population to be able to work from home.
+Although not every place hosts individuals in occupations which allow for 
+telecommuting nor with the necessary skills to effectively use the internet to 
+telecommute, this paper raises the issue that without good internet connectivity,
+these places would struggle more to achieve economic resilience in a period like 
+the current pandemic in any case. 
+Indeed, our analysis demonstrated that the temporal profiles of twelve of our 
+thirteen clusters had slower upload speeds in the morning than in the evening.
+The opposite is likely to have been the norm prior to the pandemic, as the level of 
+demand and bandwidth management is the most common cause of temporal variation in 
+experienced speeds, and why evening download speeds, rather than daytime upload 
+speeds have been used to benchmark the performance of internet services. 
+Thus, the new patterns can be taken as evidence of widespread telecommuting 
+and other daytime internet use which changed the temporal profile of 
+internet activity throughout the UK, not just in areas with more digital industry 
+or better skills.
+
+Upload speeds have not previously been seen as integral to universal service, 
+considering there has never before been such extreme demand for telecommuting 
+and operations such as video calls. 
+This may be why the average upload speeds for the largest two clusters 
+-- $6$ and $13$ -- contain so many LADs that are centres of the knowledge economy, 
+from Oxford and Cambridge in the former, to areas like Reading, which has the 
+highest concentration of digital businesses in the country [@technation2017]. 
+Thus, whilst some areas suffer from an intersection of digital and economic divides, 
+such as those in cluster $1$, others found their digital infrastructure to be less
+reliable when confronted with the sudden change in the timing and type of demand.
+Still other LADs, such as Milton Keynes in cluster $9$ were able to benefit both 
+from reliable internet connections and populations which were familiar with working 
+from home and could capitalise on their digital infrastructure.
+Yet the nuanced picture we gained through our analysis of the UK case study suggests 
+that quality internet connectivity would also enabled other LADs to gain ground
+during the pandemic despite lower skills or earnings currently, especially in a
+future where telecommuting might be a more common means of accessing work 
+and broadband services, as well as infrastructure, must be fit for purpose..
 
 <!-- WFH broader consequences-->
-The long-term effects of such drastic changes in telecommuting and attitudes
+Indeed, the long-term effects of such drastic changes in telecommuting and attitudes
 towards working from home are difficult to predict. Nevertheless,
 they span through various aspects of economy and society: from changes to
-transportation planning due to altered commuting patters, to changes in land
+transportation planning due to altered commuting patterns to changes in land
 use and urban planning to accommodate people who work from home
-[@BUDNITZ2020102713; @ELLDER2020102777] and from productivity
-and innovation changes, to changes in agglomeration externalities and the
-attraction of large cities [@econobs] just to name a few.
-The take home message from this paper is that working form home and, consequently,
-its long term effects cannot be understood, set aside plan our cities for, without
-considering the underpinning digital divides and how telecommuting intersects
-with them.
+[@BUDNITZ2020102713; @ELLDER2020102777], and from productivity
+and innovation changes to changes in agglomeration externalities and the
+attraction of large cities [@econobs].
+Further research may be able to measure the economic resilience of the different 
+clusters of places discussed in this paper once this pandemic is firmly past.
+However, our analysis demonstrates that the economic resilience made possible by
+working from home cannot be understood without considering the underpinning digital 
+divides and cannot be achieved without planning for how the levels of digital,
+social and economic divides might intersect.
 

### 2020-12-29 paper/v1/broadband.speed.covid.Rmd
abstract + directories

@@
 ---
-title: |
-  WFH and broadband speed (title needs rework)
-type: ARTICLE TEMPLATE
+title: "Working from home and digital divides: resilience in the time of a pandemic"
 author:
-  - name: A. N. Author
-    affil: a
-    email: latex.helpdesk@tandf.co.uk
-  - name: John Smith
-    affil: b
-    email: john.smith@uni-freiburg.de
-affiliation:
-  - num: a
-    address: |
-      Taylor & Francis, 4 Park Square, Milton Park, Abingdon, UK
-  - num: b
-    address: |
-      Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany
-bibliography: bibliography.bib
-appendix: appendix.tex
-abstract: |
-  TBC
-keywords: |
-  covid; internet; working from home; broadband speed; time-series clusters
+- name: A. Uthor*
+  num: 1,2
+- name: O. Tro
+  num: 2
+- name: O. Vriga
+  num: 3
+address:
+- num: 1
+  org: Department of Incredible Research, University A, City A, Country A
+- num: 2
+  org: Department of Applied Things, University B, City B, Country B
+- num: 3
+  org: Very Important Stuff Committee, Institute C, City C, Country C
+corrauth: "Corresponding author name, This is sample corresponding address."
+email: correspondingauthor@protonmail.com
+abstract: "This paper offers a new perspective on telecommuting from the viewpoint of the complex web of digital divides.
+Using the UK as a case study, this paper studies how the quality and reliability of internet service, as reflected in *experienced* vis-à-vis *advertised* upload internet speeds during the spring 2020 lockdown, may reinforce or redress the spatial and social dimensions of digital divisions.
+Fast, reliable internet connections are necessary for the population to be able to work from home.
+Although not every place hosts individuals in occupations which allow for telecommuting nor with the necessary skills to effectively use the internet to telecommute, we raise the issue that without good internet connectivity, these places would struggle more to achieve economic resilience in a period like the current pandemic.
+We employ data on individual broadband speed tests and state-of-the-art time-series clustering methods to create clusters of UK local authorities with similar temporal signatures of experienced internet speeds.
+We then associate these clusters of local authorities with their socioeconomic and geographic characteristics to explore how they overlap with or diverge from the existing economic and digital geography of the UK.
+Our analysis enables us to better understand how the spatial and social distribution of both occupations and online accessibility intersect to enable or hinder the practice of telecommuting at a time of extreme demand."
+keywords:   covid; internet; working from home; broadband speed; time-series clusters;
+classoption:
+  - Royal
+  - times
+bibliography: bibliography
+bibliographystyle: sageh
 header-includes: |
@@
 However, this ability has not been evenly distributed socially or spatially,
-creating a new type of digital divide.
+creating new intersections of economic and digital divisions.
 On one side are those who can work from home, supported by digital technologies,
@@
 workers, who are potentially at high risk of infection.
-Whilst the basis for this new digital divide has been viewed as mainly occupational,
-here we consider whether the divide is also technological.
+Whilst the basis for this pandemic-generated divide has been viewed as mainly
+occupational, here we consider whether it is also technological and, consequently, geographical.
 
@@
 Using the UK as a case study, this paper aims to understand how the quality and
-reliability of internet service, as reflected in *experienced* internet speeds,
-may reinforce or redress the spatial and social dimensions of the digital division
-exposed by the pandemic.
-To do so, we employ volunteered geographic data on individual broadband speed tests
+reliability of internet service, as reflected in *experienced* internet speeds
+during the spring 2020 lockdown,
+may reinforce or redress the spatial and social dimensions of digital divisions.
+We employ volunteered geographic data on individual broadband speed tests
 and state-of-the-art time-series clustering methods to create clusters of UK local
@@
 practice of telecommuting at a time of extreme demand.
-We will also consider what lessons can be learned from this time for a future where
+We also consider what lessons can be learned from this time for a future where
 telecommuting is likely to remain a more common means of accessing work,
@@
 has been used to benchmark broadband performance and service delivery by different
-Internet Service Providers (ISPs), at least in the UK [@ofcom2017].
+Internet Service Providers (ISPs) [@ofcom2017].
 
@@
 has been drastic and there are speculations that post-Covid, the tendency to work from
-home will be much higher than pre-Covid, raising questions around whether internet
+home will be much higher, raising questions around whether internet
 services can accommodate the increased demand.
-For example, $47$% of people in employment in the UK worked soley from home in April
+For example, $47$% of people in employment in the UK worked solely from home in April
 $2020$, whilst the same figure only reached $5$% the year before [@ons2020;
 @ons2020lm2019].
-A back of the envelope calculation suggests that up to 40% of the working force
+A back of the envelope calculation suggests that up to 40% of the labour force
 could work from home on an ongoing basis [@batty2020editorial].
 Similar figures have been reported for other countries [@felstead2020homeworking].
-For instance, $37$% of the European workforce worked from home in April $2020$ with
+Approximately $37$% of the European workforce worked from home in April $2020$ with
 countries like Finland reaching $60$% [@eurofound2020].
@@
 society^[See for example the huge success of videoconferencing apps such as Zoom
-[@marks2020zoom].],
-the same does not apply for ICT hardware infrastructure such as internet
-broadband connectivity.
+[@marks2020zoom].], the same does not apply for ICT hardware infrastructure such as
+internet broadband connectivity.
+
+<!-- ICT infrastructure and digital divides-->
 The literature describes first level digital divides in terms of the
@@
 pandemic.
-The extreme level of demand for telecommuting fundamentally alters the
+<!--  The extreme level of demand for telecommuting fundamentally alters the
 potential returns of internet use for the user and wider community,
 assumes skills or functions that are present in some occupations but not in others,
-and relies upon access to high quality internet services.
-
-As the quality of internet infrastructure and services, as well as variation in
-occupations are spatially dependent and clustered in space, our approach offers a
-framework for understanding the impact of and interactions between 
-the different levels of digital division in different places 
-with different characteristics.
-By asking how resilient broadband speeds, and particularly upload speeds are 
-as experienced in different parts of the UK during a time of extreme demand, 
+and relies upon access to high quality and reliable internet services.-->
+
+As the quality of internet infrastructure and services, as well as the concentration
+of different occupations are spatially dependent and clustered in space,
+our approach offers a framework for understanding the impact of and interactions
+between digital divisions geographically and socioeconomically.
+By asking how resilient broadband speeds, and particularly upload speeds, are
+experienced in different parts of the UK during a time of extreme demand,
 we interrogate which places benefit from the greater economic resilience
-digital technologies can offer, not only during the pandemic, 
-but also into the future.
+digital technologies can offer.
 The structure of this paper is as follows.
 First we review the literature on telecommuting and digital divides to better
-understand the structural and spatial development of these practices pre-pandemic,
+understand their structural and spatial development pre-pandemic,
 and thus their importance to the economic resilience of different places.
@@
 Our results section first offers a classification of how internet services vary
-across the UK local authorities and then assesses whether these clusters replicate
-or repudiate other socio-economic and geographic patterns of economic resilience.
+across clusters of UK local authorities and then assesses whether these clusters
+replicate or repudiate other socio-economic and geographic patterns of economic
+resilience.
+We conclude with a discussion of the insights we have gained from our new
+perspective on digital divisions.
 
@@
 <!-- Who works from home-->
-In this analysis, the terms ‘telecommuting’ and ‘working from home’ are used
+In this analysis, the terms ‘telecommuting’ and ‘working from home^[See also the popular social media hashtag #WFM]’ are used
 interchangeably, as most remote labour during the Covid-19 crisis was carried out
@@
 Yet whether this variation in infrastructure quality affects the spatial footprint of
-telecommuting has not previously been measured, in part because telecommuting has
+telecommuting has not previously been discussed, in part because telecommuting has
 not previously been the cause of greatest demand and pressure on internet services.
@@
 more data of all sorts and use their connections for a variety of purposes
-[@hauge2011consumer], 
-and that there is a correlation between access to internet services and a
+[@hauge2011consumer].
+There is also a correlation between access to internet services and a
 reduction in household transport spend [@bris2017ict].
-Whether the implication is more internet use and less travel 
+Independently of whether the additional internet use and reduced travel is
 because of increased telecommuting, these studies suggest that better internet
@@
 an example of the first level digital divide reinforcing the third level.
-Yet did the purchase of high speed connections and increased internet access also
-prepare households for long-term home-working, enforced by government restrictions?
+
+<!-- Yet did the purchase of high speed connections and increased internet access also
+prepare households for long-term home-working, enforced by government restrictions?-->
 The extreme demand during the pandemic provides a new opportunity to understand
@@
 the spread of infection.
-We seek to answer how internet service resilience contributes to or reduces 
-economic resilience when the latter is dependent upon 
-the capability to work from home. 
-We also aim to improve our understanding of the impact 
-of first level digital division on telecommuting, and whether this results in 
-much more fundamental third level digital division than has previously been perceived.
+We seek to answer how internet service resilience contributes to or reduces economic
+resilience when the latter is dependent upon the capability to work from home.
+We also aim to improve our understanding of the impact of first level digital
+division on telecommuting, and whether this results in much more fundamental third
+level digital division than has previously been perceived.
 
@@
 and the economic geography of the UK.
-Following the regional economic resilience literature, which
-underlines the differentiated capacity of cities and regions to escape or
-recover from economic crises [@martin2012regional; @kitsos2018economic],
-different places have different industrial and occupational profiles,
-and these affect the aggregated potential capacity of places for telecommuting.
+Following the regional economic resilience literature, which underlines the
+differentiated capacity of cities and regions to escape or recover from economic
+crises [@martin2012regional; @kitsos2018economic], different places have different
+industrial and occupational profiles, and these affect the aggregated potential
+capacity of places for telecommuting.
 Such profiles are associated with longstanding inequalities in the UK and their
@@
 interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
-Not only are all three levels of digital divides associated to a certain extent 
-with or shaped by the geography of the UK, but the intersection of the digital and
-material divides affects the capacity of places to overcome
-some of the economic effects of the Covid-19 pandemic.
+Not only are all three levels of digital divides associated with or shaped by the
+geography of the UK, but the intersection of the digital and material divides
+affects the capacity of places to overcome some of the economic effects of the
+Covid-19 pandemic.
 Importantly, this is the first time that digital technologies became an essential
@@
 
-<!-- ## other subsection? -->
-<!-- **I THINK WE HAVE ENOUGH OF A LIT REV** -->
-<!-- - circumstances where work can be carried out more flexibly in space and time -->
-<!-- - how relate to digital divides -->
-<!-- - capability theory of mobility? -->
-<!-- - broadband tech stuff? broadband studies / divides / resilience -->
-
-<!-- Some new papers google recommended to me: -->
-
-<!-- - https://urbanstudies.uva.nl/binaries/content/assets/subsites/centre-for-urban-studies/working-paper-series/wps_43.pdf -->
-<!-- - https://link.springer.com/article/10.1007/s11116-020-10136-6 -->
-<!-- - https://www.sciencedirect.com/science/article/pii/S0966692319311305 -->
-<!-- - check who cites the above and what they cite -->
-
-<!-- How can we observe observe WFM? -->
-<!-- removed for now -->
-<!-- Directly through surveys (@felstead2020homeworking).
-Add survey limitations (cost, representativeness, time lag between survey and data availability)
-Data gap.
-Passive data collection through internet speeds
-Another option could be mobility data.
-How is internet speed data a better source than mobility?-->
-
-<!-- **MORE SOURCES:** -->
-
-<!-- - https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas -->
-<!-- - [EPB commentaries](https://journals.sagepub.com/toc/EPB/current) -->
-<!-- - https://www.coronavirusandtheeconomy.com/question/what-has-coronavirus-taught-us-about-working-home -->
-<!-- - https://www.coronavirusandtheeconomy.com/question/who-can-work-home-and-how-does-it-affect-their-productivity -->
-<!-- - https://www.coronavirusandtheeconomy.com/question/how-will-economic-effects-coronavirus-vary-across-areas-uk -->
-<!-- - https://www.coronavirusandtheeconomy.com/which-parts-uk-have-been-hit-hardest-covid-19-crisis -->
-<!-- - https://www.coronavirusandtheeconomy.com/question/why-has-coronavirus-affected-cities-more-rural-areas -->
-
 # Methods and data {#sec:3}
 
-##Time-Series Clustering {#sec:3.1}
+## Time-Series clustering {#sec:3.1}
 <!-- clustering in general -->
-The starting point of our methodological framework is cluster analysis, which can
+First we identified our methodological framework as cluster analysis, which can
 be defined within the modern machine learning framework as an unsupervised
@@
 structures within the data [@aghabozorgi2015time].
-Therefore, cluster analysis is a widely used family of techniques in geography
+Therefore, cluster analysis is a widely used in geography
 [@gordon1977classification; @everitt1974cluster].
-For instance, clustering methods are the basis of *geodemographics*, a research domain 
-which aims to create small area indicators or typologies of neighbourhoods 
-based on various and sometimes diverse variables 
-[@SINGLETON2009289; @harris2005geodemographics].
-Clustering techniques have also been employed to solve *regionalisation* problems
+For instance, clustering methods are the basis of *geodemographics*, a research
+domain which aims to create small area indicators or typologies of neighbourhoods
+based on diverse variables [@SINGLETON2009289; @harris2005geodemographics].
+They have also been employed to solve *regionalisation* problems
 [@niesterowicz2016].
-
-<!-- time-series clustering -->
 Common characteristics of these studies are the cross-sectional nature of the data
@@
 However, for this paper we are interested in internet speeds, which vary over time
-and, therefore, create clusters of local authorities in the UK with similar temporal 
-signatures of experienced internet speeds.
-Hence, we deviate from the established geographical clustering tools and
-employ time-series clustering methods.
-
-Time-series clustering methods have been developed in order to deal with clustering
-problems linked to, for instance, stock or other financial data, economic,
-governmental or medical data as well as machine monitoring
+and, therefore, we create clusters of local authorities in the UK with similar
+temporal signatures of experienced internet speeds.
+
+<!-- time-series clustering -->
+To do so, we employ time-series clustering methods, which have been developed
+to deal with clustering problems linked to, for instance, stock or other financial
+data, economic, governmental or medical data as well as machine monitoring
 [@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
@@
 (ii) open questions regarding the choice of similarity measures in order to cluster
-similar times series objects together considering the whole length of the time-series
+similar times-series objects together considering the whole length of the time-series
 and overcoming issues around noise, outliers and shifts [@lin2004iterative;
 @aghabozorgi2015time].
-
-**REMOVE THIS PARAGRAPH - IRRELEVANT DETAIL: Time-series clustering methods utilising the whole length of time-series can be grouped in three categories.
-The first -- model-based approaches -- is based on recovering the underlying model for each time-series and then applying clustering algorithms on the model parameters of each time-series [@aghabozorgi2015time].
-The main criticism is the cluster accuracy for nearby clusters [@mitsa2010temporal].The second approach is based the formation of vectors of features derived from the original time-series.
-These new data of reduced dimensionality are then clustered using conventional clustering algorithms.**
 
@@
 their shapes through the calculation of distances between the shapes,
-and are thus better equipped to capture similarities between short length
-time-series [@aghabozorgi2015time].
-This approach serves best this paper because (i) we identify clusters of
-UK local authorities with similar temporal signatures -- i.e. shapes -- of
-experienced internet speeds and (ii) the length of our time-series is short
-(see the data discussion in this section).
-
-Another import element of time-series clustering is the actual clustering algorithm.
-Similar to the clustering of cross-sectional data, we can employ partitioning
-algorithms, which lead to non-overlapping clusters, hierarchical clustering, 
-which classifies clusters at different levels, and fuzzy algorithms, which
-create overlapping clusters [@sardatime].
-Because of the simplicity of the implementation and the interpretability of the
-results, we utilise here partitioning clustering based on the widely used *k-Means*
-algorithm.
+and are better equipped to capture similarities between short length time-series
+[@aghabozorgi2015time], such as our data.
+We thus identify clusters of UK local authorities with similar temporal signatures --
+i.e. shapes -- of experienced internet speeds.
+The clusters are identified using a partitioning algorithm, which leads to
+non-overlapping clusters, and particularly the widely use *k*-means algorithm because
+of the simplicity of the implementation and the interpretability of the
+results.
 This is an iterative algorithm, which begins with defining the desired number of
@@
 <!-- difference between clustering cross-sectional and time-series data -->
-There are a number of differences between the above described application of
-*k-Means* for cross-sectional data and its application for times series data.
+There are a number of differences between the application of *k*-means for
+cross-sectional and times-series data.
 Instead of creating clusters around centroids, a common approach is to create
@@
 Also, instead of calculating the Euclidean distance between centroids and data
-points, more complex distance measures need to be employed in order to capture
+points, more complex distance measures need to be employed to capture
 the similarity between a time-series object and a medoid.
 A common distance measure for shape-based time-series clustering is
-Dynamic Time Warping (DTW).
-Using its underpinning dynamic programming algorithm, DTW compares two time-series
-objects to find the optimum warping path between them.
+Dynamic Time Warping (DTW), an algorithm comparing two time-series objects to find
+the optimum warping path between them.
 DTW is widely used in order to overcome limitations linked to the use of Euclidean
@@
 
-##Experienced Broadband Speeds {#sec:3.2}
+## Experienced Broadband Speeds {#sec:3.2}
 <!-- introduction of our broadband speeds data -->
-To assess the quality and reliability of internet across local authorities
-in the UK during the time when the population were told to work from home if at
-all possible we utilise unique data comprising individual internet speed tests from
+To assess the internet quality and reliability across local authorities in the UK,
+we utilise unique data comprising individual internet speed tests from
 Speedchecker Ltd^[[https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)].
@@
 These data have been used before to assess digital divides [@riddlesden2014broadband]
-and the impact of local loop unbundling regulatory processes [@nardotto2015unbundling].
-By using volunteered geographic data, we are able to assess the *experienced* 
-internet speed by users, which may differ from the *advertised* maximum speeds of 
-Internet Service Providers (ISPs).
-
+and the impact of local loop unbundling regulatory processes
+[@nardotto2015unbundling], and we followed the former's approach to remove outliers.
+By using this volunteered geographic data, we are able to assess the internet speed
+*experienced* by users, which may differ from the maximum speeds *advertised* by
+ISPs.
 We are particularly interested in upload speeds and the frequency of speed tests
-over the period from March to May 2020, as government statements indicate this is
-when UK workers were told to work from home if at all possible [@GovUK2020].
+over the period from March to May $2020$, as government statements indicate this is
+when UK workers were first told to work from home if at all possible [@GovUK2020].
 Average upload speeds are slower than average download speeds, at $9.3$Mb/s
-mean upload speed for the whole sample during the period of interest, 
+mean upload speed for the whole sample,
 compared to $29.6$Mb/s for download speeds, but they are also less associated
@@
 uploading documents or two-way audio, video, and text-based communication systems.
-Meanwhile, the frequency of speed tests was important in identifying 
-the temporal profile which would give us most insight into 
-experienced internet service and resilience, and provide 
-an indication of the volume of experience over particular units of time.
 
 <!-- data transformation -->
-The first step in the workflow after dropping some outliers following
-@riddlesden2014broadband was to transform the individual, geolocated and time-stamped 
-tests to more meaningful aggregates both in terms of space and time.
-The frequency of testing indicates that whilst there is an overall trend of increased
-testing from March to April and then a slight reduction from April to May, 
+The frequency of speed tests was important in identifying the temporal profile which
+would give us most insight into experienced  internet service and resilience over
+units of time.
+Whilst there is an overall trend of increased testing
+from March to April and then a slight reduction from April to May,
 this trend masks substantial variation by not only the day of the week,
@@
 Furthermore, the importance of this variation is highlighted by a comparison
-with the same period in $2019$, as in Figure \ref{test2019}, 
-when the volume of testing and thus of experience of internet service quality
-was much more tied to the evening leisure peak and presumably to download speeds.
-Since the increase in testing during the working day in $2020$ is an indication that 
-users have a greater perception of the variable quality of internet service,
-particularly during a new morning peak of testing for service reliability,
-we decided to include a measure of hourly variation in our temporal profiles.
-
-However, although this is a large data set -- $241,088$ individual tests performed 
-during weekday hours across the study period -- 
-there are not enough observations for each Local Authority District (LAD) 
-and for each working hour of each working day -- $631$ speed tests per LAD on average
--- to profile speeds at that level of detail.
-Therefore, we aggregate all the speed-checks during the $13$ weeks of March to May 
-inclusive for weekdays in $2020$ by each hour of the day and day of the week.
+with the same period in $2019$, as in Figure \ref{test2019}, when the volume of
+testing and thus of experience of internet service quality peaked in the evening,
+presumably in response to demand for leisure activities and download speeds.
+In contrast, the majority of the increase in testing in $2020$ is during the working
+day, creating a new morning peak in Figure \ref{test2020}.
+Therefore, we include a measure of hourly variation in our temporal profiles to
+reflect the change in users' perception of the workday reliability of internet
+services.
+
+However, there were insufficient observations -- only $631$ speed
+tests per LAD on average -- for each for each working hour of each working day in
+each Local Authority District (LAD) to profile speeds at that level of detail.
+Spatial aggregation was also necessary because we could not follow individuals or
+households and connect data points.
+Therefore, we aggregate the $241,088$ individual, geolocated and time-stamped
+speed-checks during the $13$ weeks of March to May inclusive for weekdays in $2020$
+by each hour of the day and day of the week.
 As our research aims to identify the geography of internet service resilience for
-work purposes, bank holidays and the hours between midnight and 6:00 were excluded,
+work purposes, bank holidays and the hours between midnight and $6:00$ were excluded,
 as well as weekend days.
@@
 $90$ time points per series.
-We also aggregate these data spatially because we could not follow individuals or 
-households and connect data points.
 The time-series were calculated for each of the $382$ LADs in the UK, standardised,
@@
 <!-- metrics -->
-In [section](#sec:4.1), we review the temporal profile of upload speed 
-by hour of the day and day of the composite week, as well as 
-the experienced speed characteristics of each cluster.
+In Section [4.1](#sec:4.1), we review the temporal profile of upload speed
+by hour of the day and day of the composite week for each cluster.
 Since the quality and reliability of internet services vary in time and space due
-to both supply and demand-side influences, we use a number of different measures
+to both supply and demand-side influences, we also use a number of different measures
 to describe experienced upload speeds per cluster.
-These include: a) mean, experienced connection speed,
-b) standard deviation or the amount of fluctuation from the mean, and
-c) the variation in speeds at particular times of day when working from home is
-more likely to take place.
-We take account of all three measurements in our descriptives statistics of
-upload speeds in order to determine how resilient broadband speeds are 
+These include: i) mean, experienced connection speed,
+ii) standard deviation or the amount of fluctuation from the mean, and
+iii) the variation in speeds during the new morning peak of testing when working
+from home is more likely to take place.
+We take account of all three measurements
+in order to determine how resilient broadband speeds are
 as experienced in different parts of the UK during a time of extreme demand.
@@
 of quality and reliability of upload speeds.
-Instead, we run an auxiliary regression in order to understand how the spatial and
+Instead, we run an auxiliary regression to understand how the spatial and
 temporal patterns of internet service relate to the economic geography of the UK.
-More specifically, we estimate the following multinomial model:
+More specifically, we estimate the following multinomial logit model:
 
@@
 socio-economic and geographic variables, which are discussed in detail in
-the relevant [section](#sec:4.2), in order to explore how the different patterns 
-might support or undermine efforts to work from home and maintain safe productivity 
-and whether they reinforce existing spatial and social inequalities.
-This analysis enables us to provide a more nuanced understanding of how
-telecommuting and technology intersect at a time of extreme demand, and what
-lessons this time has for a future where telecommuting is likely to remain a
-common means of accessing work and broadband services, as well as infrastructure,
-must be fit for purpose.
+the relevant Section [4.2](#sec:4.2).
+<!-- limitations-->
+Because we cannot identify individuals or households
+and, consequently, we aggregated our data at the LAD level,
+our results offer correlations between the socioeconomic
+characteristics of certain geographic locations and internet service quality,
+not a record of who was telecommuting.
+Such individual data could be found though surveys, but these offer less
+detailed information about the experience of internet resilience due to enforced
+demand, which is the main contribution of this paper.
+Our auxiliary regression, therefore, provides an indication of how
+internet connectivity can reinforce or redress existing spatial and social
+inequalities in different places.
+However, it opens a path to future research by highlighting the importance of  
+understanding of how telecommuting capabilities and digital infrastructure divisions
+intersect.
 
@@
 These figures and table provide a comprehensive overview of the quality and
-reliability of experienced broadband in different parts of the UK, 
-the temporal clusters offering a novel approach to understanding spatial disparity.
+reliability of experienced broadband in different parts of the UK.
 
@@
 The second largest cluster, comprising $126$ local authorities and over $20$ million
-people, is cluster $13$.
-Cluster $13$ has the slowest aggregate mean upload speed of any of the clusters,
+people, is cluster $13$, which has the slowest aggregate mean upload speed,
 and the second highest ratio of the standard deviation to the mean.
@@
 some of the lowest quality broadband services in terms of upload speeds in the UK.
-However, the variation in upload speeds in cluster $13$, 
-which can be an indication of its reliability, does not seem to disproportionately 
-affect the morning peak from $9:00$-$10:59$, as upload speeds are, on average, 
+However, Figure \ref{UpClusterL} shows that the high standard deviation,
+which is one indication of unreliable internet, did not disproportionately affect
+the morning peak from $9:00$-$10:59$, when upload speeds were, on average,
 only $2.6$% slower than in the evening peak period between $19:00$ and $20:59$,
 when entertainment purposes are likely to be using the most bandwidth.
-In comparison, the five LADs that are home to $343$ thousand people in cluster $1$ 
-not only experience the second slowest mean upload speeds 
-and the highest ratio of standard deviation to the mean, 
-but are also much more affected during the morning peak.
-
-Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs home to $29.5$
-million people, experience aggregate mean upload speeds of about $1$Mb/s faster 
+In comparison, the five LADs in cluster $1$, home to $343$ thousand people, not only
+experience the second slowest mean upload speeds and the highest ratio of standard
+deviation to the mean, but are also much more affected during the morning peak.
+
+Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs and $29.5$
+million people -- experienced aggregate mean upload speeds of about $1$Mb/s faster
 than those in cluster $13$, but still lower than the other three large clusters
 and most of the smaller clusters, suggesting a middling quality of service.
-However, the time profile for cluster $6$ in Figure \ref{UpClusterL} shows 
-that upload speeds drop quickly from $6:00$ to $7:00$ on a Monday 
-and peak between $23:00$ and midnight on Wednesday and Thursday, 
-but tend to be lower during the working day.
-Indeed, experienced mean upload speeds in the morning peak are $4.4$% lower 
+The temporal profile for cluster $6$ in Figure \ref{UpClusterL} shows that upload
+speeds are highest at $6:00$ on a Monday morning and peak between $23:00$ and
+midnight on Wednesday and Thursday, but tend to be lower during the working day.
+Furthermore, experienced mean upload speeds in the morning peak are $4.4$% lower
 than in the evening peak -- a greater, more noticeable change than any of the other
-large clusters experience, but smaller than any of the clusters with temporal
-profiles shown in Figure \ref{UpClusterS}.
+large clusters experience, suggesting poorer reliability during the working day.
+This difference is less, however than any of the clusters
+as shown in Figure \ref{UpClusterS}.
+
 Clusters $8$ and $12$ also have mean upload speeds under $10$Mb/s, but higher than
-clusters $1$ and $13$, and their standard deviation is not dissimilar. 
-However, this masks great variation in when lower speeds are experienced, 
-with the mean upload speeds much lower between $9:00$-$10:59$ than 
-between $19:00$-$20:59$ in cluster $8$, but slightly faster in the morning in 
-cluster $12$.
-
-```{r echo=FALSE, out.width = "95%",fig.cap="\\label{map.up.clusters}Upload speed clusters for LADs"}
-img1_path <- "figures/map.up.clusters.png"
-include_graphics(img1_path)
-```
-
-Indeed cluster $12$ is the only cluster to experience higher speeds in the evening
-peak, compared to the morning peak, suggesting that widespread telecommuting
-has generally changed the temporal profile of internet activity throughout the UK.
-Yet even if all but one cluster is showing slower speeds in the morning than 
-the evening, the reliability of internet services in different clusters 
+clusters $1$, $13$, and $6$. However, mean upload speeds are much lower between
+$9:00-10:59$ than between $19:00-20:59$ in cluster $8$, but slightly faster in the
+morning than in the evening in cluster $12$.
+Indeed, cluster $12$ is the only cluster to experience higher speeds in the morning
+peak compared to the evening, and thus the only cluster where the temporal profile
+of internet use is closer to what might have been expected pre-pandemic.
+And yet even among the other clusters, the reliability of internet services
 during the working day still varies considerably.
@@
 have the most reliable internet services.
-The ratio of standard deviation to mean in both these clusters is below 55%, 
-and the ratio of upload speeds in the peak periods indicates that speeds are
-only about 2% slower in the morning.
+The ratio of standard deviation to mean in both these clusters is lowest,
+and speeds are only about 2% slower in the morning than the evening.
 Mean speeds are also higher than in any other cluster, excluding cluster $2$,
@@
 In Figure \ref{UpClusterL}, cluster $11$ shows more noticeable peaks and troughs,
-but the lowest points are not at the peak times described in Table \ref{up.cluster.descr}. 
-Rather, the slowest upload speeds on average occur between 6:00-7:00 on 
-Monday morning, 14:00-15:00 on Friday, and $16:00-17:00$ on Thursday. 
-These slowest times are still mostly faster than the average hourly upload speeds
-in cluster $13$.
-Finally, ignoring the smallest clusters in terms of population, 
-clusters $3$, $7$ and $10$ also have relatively high mean upload speeds.
-Clusters $3$ and $10$ pair high mean speeds with low standard deviations 
-relative to the mean speeds, suggesting reliability and resilience, 
-as well as quality broadband services.
-Cluster $7$ has a higher ratio of standard deviation to mean, but there is less 
-difference in average speeds between the morning and evening peaks than in clusters
-$3$ and $10$.
-
-In summary, the local authorities in clusters $9$ and $11$ experienced resilient
-broadband that could support high levels of telecommuting.
+but the lowest points are not during the morning peak, but occur, on average,
+between $6:00-7:00$ on Monday morning, $14:00-15:00$ on Friday, and $16:00-17:00$ on
+Thursday. These slow periods still offer better speeds than the average hourly
+profile of cluster $13$. Finally, ignoring the smallest clusters in terms of
+population, that is clusters $4$ and $5$, clusters $3$, $7$ and $10$ also have
+relatively high mean upload speeds. Clusters $3$ and $10$ pair high mean speeds with
+low standard deviations relative to the mean speeds, suggesting reliability and
+resilience, as well as quality broadband services. Cluster $7$ has a higher ratio of
+standard deviation to mean, but there is less difference in average speeds between
+the morning and evening peaks than in clusters $3$ and $10$.
+
+In summary, LADs in clusters $9$ and $11$ experienced resilient
+broadband internet that could support high levels of telecommuting.
 Those in clusters $2$, $3$, $7$, and $10$ also experience higher than average
-mean speeds and rank high to middle on measures of service realiability.
+mean speeds and rank middle to high on measures of service reliability.
 These LADs are *not* on the wrong side of the first level digital divide,
@@
 high standard deviations.
-With those in cluster $1$ in particular more likely to experience that poor 
-reliability during the morning peak, is this first level digital divide occurring
-in areas where few are occupationally able to telecommute anyway, and what are the
-implications for economic resilience?
-
-## aux regressions {#sec:4.2}
+With those in cluster $1$ more likely to experience that poor reliability during the
+morning peak, is this first level digital divide occurring in areas where few are
+occupationally able to telecommute anyway, and what are the implications for
+economic resilience?
+
+## Post-clustering regression analysis {#sec:4.2}
 
@@
 
-Using auxiliary regressions, we test whether the clusters that have higher mean
-speeds and more reliable services are more urban, more wealthy, and / or more likely 
-to benefit from a choice of high quality internet services.
-We also measure which of our clusters are more likely to have a higher proportion of 
+Using an auxiliary multinomial regression, we test whether the clusters that
+have higher mean speeds and more reliable services consist of LADs that are more
+urban, affluent, and / or more likely to benefit from a choice of high quality
+internet services.
+We also estimate which of our clusters are more likely to have a higher proportion of
 occupations which enable telecommuting because of the nature of the work.
-The results of these auxiliary regressions are presented in Table \ref{aux}.
+The results of the auxiliary regression are presented in Table \ref{aux}.
 The dependent variable is the LAD cluster membership as described in the
-[Methods and data](#sec:3) section
+[methods and data](#sec:3) section
 and equation \ref{eq1} and each column represent a different cluster.
@@
 Hambleton in North Yorkshire, a rural area of just over ninety thousand people.
-Mean, experienced upload speeds in cluster $4$ are close to both the average speeds
-for the $13$ clusters ($9.9$Mb/s) and the pre-clustered average for the whole sample
-($9.3$Mb/s) as shown in \ref{up.cluster.descr}.
+Mean, experienced upload speeds in cluster $4$ (see Table \ref{up.cluster.descr})
+are close to both the average speeds for the $13$ clusters ($9.9$Mb/s) and the
+pre-clustered average for the whole sample ($9.3$Mb/s).
 However, the standard deviation for cluster $4$ and the difference between average
 speeds in the morning compared to the evening peak periods are indications of
-worse reliability than many of the other clusters.
-Hence, the results in Table \ref{aux} should be seens as relative
-rather than absolute probabilities.
-
-First, the number of speed tests run per cluster inhabitant between 9:00-10:59
-is an indication of satisfaction or at least a lack of concern over broadband speeds
-and quality of service. 
-People in clusters $6$ and $13$ ran the most speed tests (at a per capita basis) 
-at this time of day, whilst people in cluster $11$ ran the fewest.
-This high level of satisfaction matches the high quality and reliability of service 
-we would expect from the analysis in [section](#sec:4.1).
-Those in clusters $2$, $3$, $9$ and $10$ also ran fewer speed tests per capita than
-those in the other six multi-authority clusters.
-Cluster $7$ was the only cluster with higher mean speeds and middle to high measures
-of reliability which was more likely to run more speed tests per capita.
-This might be a reflection of the choice of ISP and connection options in those
-LADs, as, along with cluster $12$, those in cluster $7$ were least likely to have
-Virgin Media connections.
-Fast Virgin Media cable connections have historically only been available 
-to $45$% of premises in the UK [@ofcom2016], where the more lucrative and 
-competative market originally attracted the cable TV provider. 
+worse reliability than many of the other clusters. Hence, the results in Table
+\ref{aux} should be seen as relative rather than absolute probabilities.
+
+First, we control for the number of speed tests run per cluster inhabitant
+between $9:00$-$10:59$ as well as the share of fast Virgin Media internet
+connections. Regarding the former, we expect people in LADs with more unreliable
+connections to test their internet speeds more often, and the results validate our
+priors. Meanwhile, fast Virgin Media cable connections have historically
+only been available to $45$% of premises in the UK [@ofcom2016], where the more
+lucrative and competitive market originally attracted the cable TV provider.
 Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
-Virgin connection, which is an indication that people in these clusters are 
+Virgin connections, which is an indication that people in these clusters are
 more likely to live in urban areas, with more choice of broadband services,
 or in other words, are more likely to be on the right side of the first level,
-infrastructure-based digital divide.
-
-We measure distance from London and from London or another metropolitan areas as 
-two more indicators of rurality and potentially first level digital divides. 
+infrastructure-based digital divide, as we expected from the analysis in Section
+[4.1](#sec:4.1).
+
+<!-- is an indication of satisfaction or at least a lack of concern over broadband speeds -->
+<!-- and quality of service. -->
+<!-- People in LADs assigned to clusters $6$ and $13$ were more likely to run speed tests -->
+<!-- (at a per capita basis) -->
+<!-- at this time of day, whilst people in cluster $11$ were the least likely to do so. -->
+<!-- This high level of satisfaction matches the high quality and reliability of service -->
+<!-- Those in clusters $2$, $3$, $9$ and $10$ were also less likely to ran speed tests per -->
+<!-- capita than those in the other six multi-authority clusters. -->
+<!-- Cluster $7$ was the only cluster with higher mean speeds and middle to high measures -->
+<!-- of reliability which was more likely to run more speed tests per capita. -->
+<!-- This might be a reflection of the choice of ISP and connection options in those -->
+<!-- LADs, as, along with cluster $12$, those in cluster $7$ were least likely to have -->
+<!-- Virgin Media connections. -->
+
+Then, we employ distance from London and from the nearest metropolitan area
+(including London) as two variables depicting peripherality, urban structure and,
+potentially, first level digital divides.
 The broadband speed tests run in the authorities in cluster $3$ are more likely to
-be taking place close to London than those run in any of the other clusters, 
-a result that makes sense considering two of the four authorities in cluster $3$ 
-are the London commuter towns of Harlow and Luton.
+be taking place close to London than those run in any of the other clusters, and
+two of the four authorities in cluster $3$ are the London commuter towns of Harlow
+and Luton.
 However, even though London was also included in the variable calculating distance
 from the centre of one of either the ten largest metropolitan areas in England,
-Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away. 
-Still, whilst significant, the effect is tiny and the other members of cluster $3$
-are Corby and Eastbourne, two towns which, whilst not part of metropolitan areas,
-are also not in remote parts of the country, and are home to an estimated population 
-of $175$ thousand people between them. **move maps to here?**
+or Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
+Thus, it is important to consider the membership of each cluster as well as the
+regression results. Corby and Eastbourne, the other two authorities in cluster $3$
+are, whilst not part of metropolitan areas, are also large, accessible towns.
+
+```{r echo=FALSE, out.width = "95%",fig.cap="\\label{map.up.clusters}Upload speed clusters for LADs"}
+img1_path <- "figures/map.up.clusters.png"
+include_graphics(img1_path)
+```
+
 Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large
 metropolitan area, even though the four local authorities of cluster $7$ include
-no central urban boroughs, and only one LAD that is part of a metropolitan area of 
-governance - Tameside in Greater Manchester.
-This may be why those in cluster $7$ are not served by Virgin Media.
-It also is a demonstration of the complexity of experienced broadband upload speeds
-as captured by time-series clustering, and their likely interaction with first level 
-digital divides.
+no central urban boroughs and only one LAD that is part of a metropolitan area of
+governance - Tameside in Greater Manchester. This may explain why those in cluster
+$7$ are the second least likely to be served by Virgin Media.
+It is also a demonstration of the complexity of experienced broadband upload speeds
+as captured by time-series clustering, and how quality and reliability, not only
+availability, engender first level digital divides.
 Cluster $1$, which our analysis suggests lacks broadband resilience, contains five,
 mainly rural authorities.
-However, they are scattered around the country --
-Ceredigion in West Wales, Darlington in the Northeast, Eden between the Lake District
-and the North Penines, Rossendale in the Northwest, and Rutland in the East Midlands
--- and therefore, the results in Table \ref{aux} indicate that these authorities
-are not as far from the centre of a metropolitan area as authorities in cluster $3$, 
+However, they are scattered around the country -- Ceredigion in West Wales,
+Darlington in the Northeast, Eden and Rossendale in the Northwest, and Rutland in
+the East Midlands -- and therefore, the results in Table \ref{aux} indicate that
+these authorities are closer to a metropolitan area than authorities in cluster $3$,
 but furthest from London.
 
-Internet resilience as measured here does not seem to follow the North-South 
-economic divide either, as broadly defined in our regression as being part of the
-Greater London, Southeast, Southwest and East of England regions.
+Internet resilience is also more nuanced than our -- arguably crude
+-- dummy variable depicting the North-South economic divide, which assigns $1$ to
+LADs located in Greater London, Southeast, Southwest and East of England regions,
+and $0$ to the rest.
 The authorities most likely to be in the South are those in cluster $13$,
@@
 and thus a low level of service.
-However, cluster $13$ does include some rural, remote areas of the country, 
-such as Northwest Scotland, Cornwall and Powys in Wales as shown on Figure \ref{map.up.clusters}
-It also includes the major metropolitan centres of Bristol, Liverpool, 
-and Newcastle, as well as nine (of $32$) London Boroughs and plenty of home county 
-and suburban areas.
-Although by the standard deviation measure cluster $13$ has unreliable internet 
-services, speed variation was less during the morning peak.
+However, cluster $13$ includes some rural, remote areas of the country, such as
+Northwest Scotland, Cornwall and Powys in Wales as shown on Figure
+\ref{map.up.clusters}.
+It also includes major metropolitan centres in the North -- Liverpool, Newcastle --
+and South -- Bristol, nine (of $32$) London Boroughs.
+There are also plenty of Southern home county and suburban areas.
+The standard deviation measure for cluster $13$ is high, but speed variation is
+low during the morning peak, suggesting that the estimates for reliability are
+inconsistent. Considering that this is one of the largest clusters, and thus the
+averages incorporate more noise than some of the smaller clusters,
+it may be that the LADs in this cluster do not all suffer equally from
+a first level digital divide.
+
+Yet we need to consider the results for other variables in Table \ref{aux} to better
+determine whether clusters $1$ and $13$, which appear to suffer most from poor
+quality internet services, are also more likely to have a low skilled workforce,
+less able to benefit from telecommuting.
+Cluster $1$ has the lowest proportion of educated people,
+and the lowest earnings, despite recording the highest proportion of managerial,
+professional jobs, and the second highest proportion of tech jobs.
+Rural areas such as those in cluster $1$ are often home to many retired people,
+which might explain these results or perhaps, as these figures are relative,
+we could note that cluster $1$ also has a greater proportion of skilled trades
+than other clusters. It seems likely that the first level digital divide reinforces
+other inequities in cluster $1$, whilst those in cluster $13$ are more likely
+to earn more and have a better education despite poor internet services.
+Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the
+middle in terms of job density. If this is an indication of a high number of SMEs,
+it could explain the variable internet quality, considering that small businesses
+have not been seen as the most valuable customers for higher speed broadband services
+if they are not located near residential customers [@ofcom2016].
+
+Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more
+noise within the averages we use to measure internet quality and reliability.
+Our results indicate average mean speeds, and Table \ref{aux} shows that
+cluster $6$ also falls towards the middle of the clusters on many of the
+socioeconomic variables. It ranks third or fourth out of the eleven in terms of
+the likelihood of having a higher proportion of managerial, professional, and
+tech jobs, as well as higher earnings, but is fourth from bottom for educational
+attainment. The LADs in cluster $6$ are also diverse, with few truly remote areas,
+but urban areas throughout England, including Birmingham, Leeds, Sheffield,
+twelve London Boroughs, and many suburban areas and smaller cities like Oxford
+and Cambridge. The capital cities of the other UK nations, Belfast, Cardiff and
+Edinburgh, are also in this cluster, suggesting perhaps that the lower level of
+reliability discussed in Section [4.1](#sec:4.1) may be due to increased demand,
+e.g. for telecommuting, when it is not supported by the most resilient internet
+connections.
+
+In contrast, Clusters $9$ and $11$ enjoy resilient internet connections and are
+characterised by higher share of occupations that may benefit from the ability
+to telecommute (e.g. managerial). LADs in Cluster $9$ are more likely to host
+highly educated individuals with higher earnings than cluster $11$, which has a
+negative coefficient for the NVQ4 variable. Cluster $11$ has fewer businesses per
+inhabitant and the second highest job density and cluster $9$ the second lowest.
+These coefficients might indicate that resilient broadband infrastructure
+generates higher returns for those in cluster $9$, where slightly more slowdown in
+the AM peak was detected (see Table \ref{up.cluster.descr}).
+LADs in cluster $9$ are also less likely to be in the South and those in cluster $11$
+more likely, although the coefficient for cluster $9$ might be skewed by
+the presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
+different economic profile than England. Still, both clusters consist of larger
+LADs in terms of population, including the city centre among four of the ten
+districts of Greater Manchester, eight London Boroughs, four of the seven
+constituent authorities of the West Midlands Combined Authority, Nottingham, and
+both Portsmouth and Southampton. Among the $65$ LADs are also a number of other
+tightly bounded urban areas, such as Blackpool, Ipswich, Norwich, Slough, and
+Stevenage, and urban areas at the centre of less confined districts like Burnley,
+the digital hub of Milton Keynes, and Northampton.
+These urban locations are on the right side of the first level digital divide,
+and the regression results suggest that internet resilience is supporting a wide
+range of small and large urban economies.
+
+LADs in clusters $2$, $3$, $7$, $10$ are also on the right side of the first level
+digital divide, experiencing higher than average mean speeds and ranking high to
+middle on measures of service reliability.
+However, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
+work force to take advantage of working from home opportunities.
+Cluster $2$ is comprised of just two LADs, with the lowest job density of any
+cluster -- these are East Lothian near Edinburgh, and North East Lincolnshire,
+home to the town of Grimsby. Cluster $10$ is comprised of four LADs, including two
+peripheral suburban areas of Birmingham, the East London Borough of Newham,
+and Dundee. Suburbs are considered the most likely urban form in which telecommuters
+live [@e2018does], and Dundee has a reputation for tech startups [@technation2017].
+LADs in cluster $7$, which include the suburbs South of Belfast, a suburban
+district of Greater Manchester, the villages between York and Leeds, and a couple of
+small West Midlands towns have the highest earnings, and thus may also be making the
+most of their resilient internet. In contrast, LADs in cluster $3$ are associated
+more with lower skills and have the highest proportion of individuals in machine
+operation jobs. Arguably, LADs in clusters $3$ benefit the least from their
+resilient internet infrastructure in an era when working from home became a vehicle
+for economic resilience.
+
+Finally, clusters $8$ and $12$ consist by LADs enjoying close to average
+mean upload speeds, but opposing patterns of internet reliability during the morning
+peak. LADs in cluster $8$ are likely to host more individuals employed in tech and
+administrative occupations than any other cluster, as well as many in managerial
+occupations, whilst the opposite applies to skilled trade and leisure occupations.
+These LADs are characterised by a very small positive likelihood of more educated
+residents, but are not significantly likely to earn more than other clusters.
+Including suburbs near Leicester, Cardiff, Birmingham, the green belt area south of
+London straddling the M25 motorway, and rural West Devon, the LADs in cluster $8$
+seem to likely to have the skills and occupations that would benefit from
+quality internet services, but suffer from poor internet resilience and reliability.
+
+In comparison, LADs in cluster $12$ are less likely to be able to benefit from
+quality internet services if they had them, with fewer individuals achieving
+NVQ4 or better and lower levels of occupations that would benefit from homeworking.
+Yet LADs in this cluster benefit from the second highest level of earnings, and
+have the largest stock of businesses per inhabitant and the highest job density.
+This density of businesses could be why cluster $12$ is the only cluster with higher
+speeds in the morning peak during the study period. If people are at home, business
+premises might well have been abandoned. Cluster $12$ is home to $1.5$ million people
+spread across twelve LADs, including the City of Westminster but also a LAD in
+Northern Ireland, two in Wales, Preston in the Northwest, and Fenland in the East of
+England. This spatial diversity demonstrates that the temporal clustering of internet
+resilience is not necessarily spatially dependent, and digital divides do not
+necessarily overlap with economic ones.
 
@@
 
-Upload speeds are not used as benchmarks in the UK, nor have been seen as integral to 
-universal service, considering there has never before been such extreme demand for
-telecommuting and operations such as video calls.
-
-those in cluster $6$, including people living in Birmingham, Leeds and 
-Sheffield, in twelve London Boroughs, and many suburban areas and smaller cities like
-Cardiff, Oxford and Cambridge 
-This lower level of reliability may be as a result of increased demand, 
-e.g. for telecommuting, especially as there are few truly remote areas 
-in this cluster.
-
-The eight LADs in clusters $3$ and $10$ are home to over one million people,
-and are all more urban and suburban than cluster $1$, including the London Borough 
-of Newham, the Southeast commuter towns of Harlow and Luton, the Birmingham suburbs 
-of Bromsgrove and Cannock Chase, Dundee City, and the towns of Eastbourne and Corby.
-
-Although not all of the local authorities in these clusters $9$ and $11$ are urban, 
-among the $65$ LADs are Manchester and three of the nine other boroughs of 
-Greater Manchester, Glasgow, Nottingham, both Portsmouth and Southampton, 
-eight London Boroughs, four of the seven constituent authorities of the 
-West Midlands Combined Authority.
-There are also a number of other tightly bounded urban areas, such as Aberdeen, 
-Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at the centre
-of less confined districts like Burnley, Milton Keynes, and Northampton.
-
-Clusters $3$, $9$, $10$ and $11$ seem to benefit most from high quality and 
-resilient broadband services.
-The dips in mean upload speeds in clusters $6$, $3$ and $10$ during the morning
-peak are suggestive of more use during the working day, and potentially more
-telecommuting.
+This paper offers a new perspective on telecommuting from the viewpoint of the
+complex web of digital divides. To do so, we employ novel data regarding
+experienced upload speeds and time-series clustering methods,
+a family of unsupervised machine learning techniques which are rarely utilised
+in geographical research. Fast, reliable internet connections are
+necessary for the population to be able to work from home.
+Although not every place hosts individuals in occupations which allow for
+telecommuting nor with the necessary skills to effectively use the internet to
+telecommute, this paper raises the issue that without good internet connectivity,
+these places would struggle more to achieve economic resilience in a period like
+the current pandemic in any case.
+Indeed, our analysis demonstrated that the temporal profiles of twelve of our
+thirteen clusters had slower upload speeds in the morning than in the evening.
+The opposite is likely to have been the norm prior to the pandemic, as the level of
+demand and bandwidth management is the most common cause of temporal variation in
+experienced speeds, and why evening download speeds, rather than daytime upload
+speeds have been used to benchmark the performance of internet services.
+Thus, the new patterns can be taken as evidence of widespread telecommuting
+and other daytime internet use which changed the temporal profile of
+internet activity throughout the UK, not just in areas with more digital industry
+or better skills.
+
+Upload speeds have not previously been seen as integral to universal service,
+considering there has never before been such extreme demand for telecommuting
+and operations such as video calls.
+This may be why the average upload speeds for the largest two clusters
+-- $6$ and $13$ -- contain so many LADs that are centres of the knowledge economy,
+from Oxford and Cambridge in the former, to areas like Reading, which has the
+highest concentration of digital businesses in the country [@technation2017].
+Thus, whilst some areas suffer from an intersection of digital and economic divides,
+such as those in cluster $1$, others found their digital infrastructure to be less
+reliable when confronted with the sudden change in the timing and type of demand.
+Still other LADs, such as Milton Keynes in cluster $9$ were able to benefit both
+from reliable internet connections and populations which were familiar with working
+from home and could capitalise on their digital infrastructure.
+Yet the nuanced picture we gained through our analysis of the UK case study suggests
+that quality internet connectivity would also enable other LADs to gain ground
+during the pandemic despite lower skills or earnings currently, especially in a
+future where telecommuting might be a more common means of accessing work
+and broadband services must be fit for purpose.
 
 <!-- WFH broader consequences-->
-The long-term effects of such drastic changes in telecommuting and attitudes
+Indeed, the long-term effects of such drastic changes in telecommuting and attitudes
 towards working from home are difficult to predict. Nevertheless,
 they span through various aspects of economy and society: from changes to
-transportation planning due to altered commuting patters, to changes in land
+transportation planning due to altered commuting patterns to changes in land
 use and urban planning to accommodate people who work from home
-[@BUDNITZ2020102713]**also 2020 Swedish article from JTG**; and from productivity and innovation changes, to changes
-in agglomeration externalities and the attraction of large cities [@econobs]
-just to name a few. This paper is positioned to support endeavours in
-understanding the effects of increased telecommuting by
-exposing the spatial and social dimensions of telecommuting including the
-resilience of broadband speeds in terms of both quality and reliability of
-service, and whether this reinforces or redresses prior digital divisions.
-**took this bit you wrote to put in the discussion at the end?**
-
-# Acknowledgement(s) {-}
-
-An unnumbered section, e.g.\ `\section*{Acknowledgements}`, may be used for thanks, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.
-
-# Funding {-}
-
-An unnumbered section, e.g.\ `\section*{Funding}`, may be used for grant details, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.
-
-# References
+[@BUDNITZ2020102713; @ELLDER2020102777], and from productivity
+and innovation changes to changes in agglomeration externalities and the
+attraction of large cities [@econobs].
+Further research may be able to measure the economic resilience of the different
+clusters of places discussed in this paper once this pandemic is firmly past.
+However, our analysis demonstrates that the economic resilience made possible by
+working from home cannot be understood without considering the underpinning digital
+divides and cannot be achieved without planning for how the levels of digital,
+social and economic divides might intersect.
+
+<!-- # Acknowledgement(s) {-} -->
+
+<!-- An unnumbered section, e.g.\ `\section*{Acknowledgements}`, may be used for thanks, etc.\ if required and included _in the non-anonymous version_ before any Notes or References. -->
+
+<!-- \pagebreak -->
+
+# Appendix 1 {#appendix1}
+
+This is the LAD cluster membership for the upload speed timeseries.
+
+```{r echo=FALSE, message=FALSE, results = "asis"}
+
+# Cluster table
+clusters <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
+clusters <- read_csv(clusters)
+clusters <- clusters %>%
+  dplyr::select(LAD, cluster.up) %>%
+  #rename('cluster membership' = cluster.up) %>%
+  #group_by('cluster membership') %>%
+  group_split(cluster.up, .keep = F)
+  #dplyr::arrange()
+  #sort('cluster membership', decreasing = F)
+
+cluster1 <- as.vector(unlist(clusters[1]))
+cluster2 <- as.vector(unlist(clusters[2]))
+cluster3 <- as.vector(unlist(clusters[3]))
+cluster4 <- as.vector(unlist(clusters[4]))
+cluster5 <- as.vector(unlist(clusters[5]))
+cluster6 <- as.vector(unlist(clusters[6]))
+cluster7 <- as.vector(unlist(clusters[7]))
+cluster8 <- as.vector(unlist(clusters[8]))
+cluster9 <- as.vector(unlist(clusters[9]))
+cluster10 <- as.vector(unlist(clusters[10]))
+cluster11 <- as.vector(unlist(clusters[11]))
+cluster12 <- as.vector(unlist(clusters[12]))
+cluster13 <- as.vector(unlist(clusters[13]))
+```
+
+**Cluster 1: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster1, type="cmd2"), collapse=", "))
+```
+
+**Cluster 2: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster2, type="cmd2"), collapse=", "))
+```
+
+**Cluster 3: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster3, type="cmd2"), collapse=", "))
+```
+
+**Cluster 4: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster4, type="cmd2"), collapse=", "))
+```
+
+**Cluster 5: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster5, type="cmd2"), collapse=", "))
+```
+
+**Cluster 6: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster6, type="cmd2"), collapse=", "))
+```
+
+**Cluster 7: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster7, type="cmd2"), collapse=", "))
+```
+
+**Cluster 8: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster8, type="cmd2"), collapse=", "))
+```
+
+**Cluster 9: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster9, type="cmd2"), collapse=", "))
+```
+
+**Cluster 10: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster10, type="cmd2"), collapse=", "))
+```
+
+**Cluster 11: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster11, type="cmd2"), collapse=", "))
+```
+
+**Cluster 12: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster12, type="cmd2"), collapse=", "))
+```
+
+**Cluster 13: **
+```{r echo=FALSE, message=FALSE, results = "asis"}
+cat(paste(shQuote(cluster13, type="cmd2"), collapse=", "))
+```
+
+```{r echo=FALSE, message=FALSE, results = "asis", eval=FALSE}
+# LADs <- do.call(cbindX, clusters) # this is with library(gdata)
+# LADs1 <- LADs[,1:7]
+# LADs2 <- LADs[,8:13]
+#
+# names(LADs1) <- c("Cluster 1", "Cluster 2", "Cluster 3",
+#                      "Cluster 4", "Cluster 5", "Cluster 6",
+#                      "Cluster 7")
+# names(LADs2) <- c("Cluster 8", "Cluster 9", "Cluster 10",
+#                      "Cluster 11", "Cluster 12")
+#
+# opts <- options(knitr.kable.NA = "") # kableExtra.auto_format = FALSE
+#
+# # LADs1 %>%
+# # kable(caption = "LAD cluster membership for upload speed timeseries (1 out of 2)\\label{cluster.membership1}") %>%
+# #   kable_styling(font_size = 5,fixed_thead = T)
+#
+# LADs1 %>%
+# kbl(caption = "LAD cluster membership for upload speed timeseries (1 out of 2)\\label{cluster.membership1}") %>%
+#    kable_classic(font_size = 5,
+#                 fixed_thead = T,
+#                 full_width = F
+#                 )
+#
+#
+# # stargazer(LADs1,
+# #           summary = F,
+# #           font.size = "footnotesize",
+# #           type = "latex",
+# #           header = F,
+# #           title = "LAD cluster membership for upload speed timeseries (1 out of 2)\\label{cluster.membership1}")  
+#
+# stargazer(LADs2,
+#           summary = F,
+#           font.size = "footnotesize",
+#           type = "latex",
+#           header = F,
+#           title = "LAD cluster membership for upload speed timeseries (2 out of 2)\\label{cluster.membership2}")  
+```
+
+# Appendix 2 {#appendix2}
+
+```{r echo=FALSE, message=FALSE, results = "asis"}
+
+# Descriptive statistics table
+path.LAs_Clusters <- paste0(path, "/data/temp/LAs_Clusters.RData")
+load(path.LAs_Clusters)
+
+for.descr <- data %>%
+  dplyr::select(pop, job.dens2018, distMet,
+                distLondon, south,
+                managers, tech, skilled, prof,
+                admin, caring, plant,
+                earnings, nvq4,
+                virgin,
+                total.busi, LAD.tests.am20, pop) %>%
+  mutate('total.busi/pop' = total.busi/pop,
+         'LAD.tests.am20/pop' = LAD.tests.am20/pop) %>%
+  dplyr::select(-total.busi, -LAD.tests.am20)
+
+stargazer(for.descr,
+          summary = T,
+          font.size = "footnotesize",
+          type = "latex",
+          header = F,
+          digit.separator = ",",
+          digit.separate = c(3,3),
+          covariate.labels=c("pop, 2018",
+                             "job density, 2018",
+                             "distance to nearest met. area",
+                             "distance to London",
+                             "south of the UK",
+                             "managerial jobs, 2020",
+                             "tech jobs, 2020",
+                             "skilled trade jobs, 2020",
+                             "professional jobs, 2020",
+                             "administrative jobs, 2020",
+                             "leisure jobs, 2020",
+                             "machine operation jobs, 2020",
+                             "earnings, 2019",
+                             "NVQ4+",
+                             "Virgin Media \\%, 2020",
+                             "n. business est. per hab., 2019",
+                             "AM tests per hab., 2020"),
+          rownames = F,
+          no.space=TRUE,
+          column.sep.width = "0pt",
+          #notes = "",
+          title = "Descriptive statistics for the auxiliary regression explanatory variables\\label{descr.aux}")
+
+```
+
+\pagebreak

### 2020-12-29 paper/v1/broadband.speed.covid.tex
abstract + directories

@@
-% interactcadsample.tex
-% v1.03 - April 2017
+\documentclass[Royal,times,sageh]{sagej}
 
-\documentclass[]{interact}
-
-\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
-\usepackage{subfigure}% Support for small, `sub' figures and tables
-%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
-
-\usepackage{natbib}% Citation support using natbib.sty
-\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
-\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
-
-\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{example}[theorem]{Example}
-
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-\newtheorem{notation}{Notation}
-
-% see https://stackoverflow.com/a/47122900
+\usepackage{moreverb,url,natbib, multirow, tabularx}
+\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 
@@
 
-\articletype{ARTICLE TEMPLATE}
+\title{Working from home and digital divides: resilience in the time of a
+pandemic}
 
-\title{WFH and broadband speed (title needs rework)}
+\runninghead{}
 
-\author{\name{A. N. Author$^{a}$, John Smith$^{b}$}
-\affil{$^{a}$Taylor \& Francis, 4 Park Square, Milton Park, Abingdon, UK; $^{b}$Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany}
-}
+\author{A. Uthor*\affilnum{1,2}, O. Tro\affilnum{2}, O. Vriga\affilnum{3}}
 
-\thanks{CONTACT A.
-N.
-Author.
-Email: \href{mailto:latex.helpdesk@tandf.co.uk}{\nolinkurl{latex.helpdesk@tandf.co.uk}}, John Smith.
-Email: \href{mailto:john.smith@uni-freiburg.de}{\nolinkurl{john.smith@uni-freiburg.de}}}
+\affiliation{\affilnum{1}{Department of Incredible Research, University A, City A, Country A}\\\affilnum{2}{Department of Applied Things, University B, City B, Country B}\\\affilnum{3}{Very Important Stuff Committee, Institute C, City C, Country C}}
+
+\corrauth{Corresponding author name, This is sample corresponding address.}
+
+\email{\href{mailto:correspondingauthor@protonmail.com}{\nolinkurl{correspondingauthor@protonmail.com}}}
+
+\begin{abstract}
+This paper offers a new perspective on telecommuting from the viewpoint
+of the complex web of digital divides. Using the UK as a case study,
+this paper studies how the quality and reliability of internet service,
+as reflected in \emph{experienced} vis-à-vis \emph{advertised} upload
+internet speeds during the spring 2020 lockdown, may reinforce or
+redress the spatial and social dimensions of digital divisions. Fast,
+reliable internet connections are necessary for the population to be
+able to work from home. Although not every place hosts individuals in
+occupations which allow for telecommuting nor with the necessary skills
+to effectively use the internet to telecommute, we raise the issue that
+without good internet connectivity, these places would struggle more to
+achieve economic resilience in a period like the current pandemic. We
+employ data on individual broadband speed tests and state-of-the-art
+time-series clustering methods to create clusters of UK local
+authorities with similar temporal signatures of experienced internet
+speeds. We then associate these clusters of local authorities with their
+socioeconomic and geographic characteristics to explore how they overlap
+with or diverge from the existing economic and digital geography of the
+UK. Our analysis enables us to better understand how the spatial and
+social distribution of both occupations and online accessibility
+intersect to enable or hinder the practice of telecommuting at a time of
+extreme demand.
+\end{abstract}
+
+\keywords{covid; internet; working from home; broadband speed; time-series
+clusters;}
 
 \maketitle
-
-\begin{abstract}
-TBC
-\end{abstract}
-
-\begin{keywords}
-covid; internet; working from home; broadband speed; time-series
-clusters
-\end{keywords}
 
@@
 around the world. However, this ability has not been evenly distributed
-socially or spatially, creating a new type of digital divide. On one
-side are those who can work from home, supported by digital
-technologies, and have thus been able to enjoy both economic resilience
-and greater personal safety. On the other side, previously employed
-individuals have been forced to accept furlough or redundancy packages
-unless they are part of the cadre of essential workers, who are
-potentially at high risk of infection. Whilst the basis for this new
-digital divide has been viewed as mainly occupational, here we consider
-whether the divide is also technological.
+socially or spatially, creating new intersections of economic and
+digital divisions. On one side are those who can work from home,
+supported by digital technologies, and have thus been able to enjoy both
+economic resilience and greater personal safety. On the other side,
+previously employed individuals have been forced to accept furlough or
+redundancy packages unless they are part of the cadre of essential
+workers, who are potentially at high risk of infection. Whilst the basis
+for this pandemic-generated divide has been viewed as mainly
+occupational, here we consider whether it is also technological and,
+consequently, geographical.
 
@@
 quality and reliability of internet service, as reflected in
-\emph{experienced} internet speeds, may reinforce or redress the spatial
-and social dimensions of the digital division exposed by the pandemic.
-To do so, we employ volunteered geographic data on individual broadband
+\emph{experienced} internet speeds during the spring 2020 lockdown, may
+reinforce or redress the spatial and social dimensions of digital
+divisions. We employ volunteered geographic data on individual broadband
 speed tests and state-of-the-art time-series clustering methods to
@@
 occupations and online accessibility intersect to enable or hinder the
-practice of telecommuting at a time of extreme demand. We will also
-consider what lessons can be learned from this time for a future where
+practice of telecommuting at a time of extreme demand. We also consider
+what lessons can be learned from this time for a future where
 telecommuting is likely to remain a more common means of accessing work,
 at least in comparison to the pre-Covid era, and broadband services and
-infrastructure must be fit for purpose. \textbf{LET'S LEAVE IT FOR NOW,
-BUT I THINK WE CAN CRYSTALISE MORE THE RQ}
+infrastructure must be fit for purpose.
 
@@
 to benchmark broadband performance and service delivery by different
-Internet Service Providers (ISPs), at least in the UK \citep{ofcom2017}.
+Internet Service Providers (ISPs) \citep{ofcom2017}.
+
 Yet the shift towards telecommuting during various stages of lockdown
 around the world has been drastic and there are speculations that
-post-Covid, the tendency to work from home will be much higher than
-pre-Covid, raising questions around whether internet services can
-accommodate the increased demand. For example, \(47\)\% of people in
-employment in the UK worked soley from home in April \(2020\), whilst
-the same figure only reached \(5\)\% the year before
-\citep{ons2020, ons2020lm2019}. A back of the envelope calculation
-suggests that up to 40\% of the working force could work from home on an
-ongoing basis \citep{batty2020editorial}. Similar figures have been
-reported for other countries \citep{felstead2020homeworking}. For
-instance, \(37\)\% of the European workforce worked from home in April
-\(2020\) with countries like Finland reaching \(60\)\%
-\citep{eurofound2020}. In the US, almost half of the working population
-worked from home during the same period because of the pandemic
-\citep{brynjolfsson2020covid}, and a recent estimate indicated that
-\(37\)\% of all jobs in the US can be permanently performed entirely
-from home \citep{NBERw26948}.
+post-Covid, the tendency to work from home will be much higher, raising
+questions around whether internet services can accommodate the increased
+demand. For example, \(47\)\% of people in employment in the UK worked
+solely from home in April \(2020\), whilst the same figure only reached
+\(5\)\% the year before \citep{ons2020, ons2020lm2019}. A back of the
+envelope calculation suggests that up to 40\% of the labour force could
+work from home on an ongoing basis \citep{batty2020editorial}. Similar
+figures have been reported for other countries
+\citep{felstead2020homeworking}. Approximately \(37\)\% of the European
+workforce worked from home in April \(2020\) with countries like Finland
+reaching \(60\)\% \citep{eurofound2020}. In the US, almost half of the
+working population worked from home during the same period because of
+the pandemic \citep{brynjolfsson2020covid}, and a recent estimate
+indicated that \(37\)\% of all jobs in the US can be permanently
+performed entirely from home \citep{NBERw26948}. None of these changes
+could have happened in the absence of reliable information and
+communication technology (ICT) infrastructure -- both in terms of
+software and hardware. But while software innovations are easily
+diffused across space and society\footnote{See for example the huge
+  success of videoconferencing apps such as Zoom \citep{marks2020zoom}.},
+the same does not apply for ICT hardware infrastructure such as internet
+broadband connectivity.
 
-None of these changes could have happened in the absence of reliable
-information and communication technology (ICT) infrastructure -- both in
-terms of software and hardware. But while software innovations are
-easily diffused across space and society\footnote{See for example the
-  huge success of videoconferencing apps such as Zoom
-  \citep{marks2020zoom}.}, the same does not apply for ICT hardware
-infrastructure such as internet broadband connectivity. The literature
-describes first level digital divides in terms of the availability and
-quality of internet connectivity, such as that manifest in different
-geographies in the UK
+The literature describes first level digital divides in terms of the
+availability and quality of internet connectivity, such as that manifest
+in different geographies in the UK
 \citep{riddlesden2014broadband, philip2017digital}. Second level digital
@@
 resilience of people and places to overcome a systemic shock such as the
-current pandemic. We identify this as a new digital divide, one that
-fundamentally alters the potential returns of internet use for the user
-and wider community, assumes skills or functions that are present in
-some occupations but not in others, and relies upon access to high
-quality internet services. \textbf{WE NEED TO CONNECT OUR FINDINGS WITH
-THIS AND JUSTIFY THAT THIS IS INDEED A NEW DIVIDE} As the quality of
-internet infrastructure and services, as well as variation in
-occupations are spatially dependent and clustered in space, our approach
-offers a framework for understanding which types of places, are more
-likely to land on the right side of this new digital divide. By asking
-how resilient broadband speeds are as experienced in different parts of
-the UK during a time of extreme demand, we interrogate which places
-benefit from the greater economic resilience digital technologies can
-offer, not only during the pandemic, but also into the future.
+current pandemic.
 
-The structure of this paper is as follows. First we review the
-literature on telecommuting and digital divides to better understand the
-origins of the new digital divide revealed by the pandemic and its
-impact on the economic resilience of different places. We then describe
-our data and methodology. Our results section first offers a
-classification of how internet services vary across the UK local
-authorities and then assesses whether these clusters replicate or
-repudiate other socio-economic and geographic patterns of economic
-resilience.
+As the quality of internet infrastructure and services, as well as the
+concentration of different occupations are spatially dependent and
+clustered in space, our approach offers a framework for understanding
+the impact of and interactions between digital divisions geographically
+and socioeconomically. By asking how resilient broadband speeds, and
+particularly upload speeds, are experienced in different parts of the UK
+during a time of extreme demand, we interrogate which places benefit
+from the greater economic resilience digital technologies can offer. The
+structure of this paper is as follows. First we review the literature on
+telecommuting and digital divides to better understand their structural
+and spatial development pre-pandemic, and thus their importance to the
+economic resilience of different places. We then describe our data and
+methodology. Our results section first offers a classification of how
+internet services vary across clusters of UK local authorities and then
+assesses whether these clusters replicate or repudiate other
+socio-economic and geographic patterns of economic resilience. We
+conclude with a discussion of the insights we have gained from our new
+perspective on digital divisions.
 
@@
 
-In this analysis, the terms `telecommuting' and `working from home' are
-used interchangeably, as most remote labour during the Covid-19 crisis
-was carried out in the homes of individual employees rather than any
-other location \citep{eurofound2020}. However, it should be noted that
+In this analysis, the terms `telecommuting' and `working from
+home\footnote{See also the popular social media hashtag \#WFM}' are used
+interchangeably, as most remote labour during the Covid-19 crisis was
+carried out in the homes of individual employees rather than any other
+location \citep{eurofound2020}. However, it should be noted that
 previous research has explored how telecommuting can occur in other
@@
 These returns have been even greater during the Covid-19 crisis, when
-those with the ability to telecommute also have the ability to maintain
-their employment whilst protecting their health. However, the success of
-these arrangements has been dependent upon the first level digital
-divide, which is associated with access and quality of internet
-connectivity at a time of extremely high demand. \citet{SALEMINK2017360}
-provides a systematic review of the pre-pandemic, first level digital
-divide in infrastructure quality between urban and rural areas in
-various advanced economies. Yet whether this variation in infrastructure
-quality affects the spatial footprint of telecommuting has not
-previously been measured. There are indications that those who purchase
-high speed connections consume more data of all sorts and use their
-connections for a variety of purposes \citep{hauge2011consumer}, and
-that there is a correlation between access to internet services and a
-reduction in household transport spend \citep{bris2017ict}. Whether the
-implication is more internet use and less travel because of increased
+those with the capability to telecommute also have the ability to
+maintain their employment whilst protecting their health. However, the
+success of these arrangements has been dependent upon the first level
+digital divide, which is associated with access to and quality of
+internet connectivity. \citet{SALEMINK2017360} provides a systematic
+review of the pre-pandemic, first level digital divide in infrastructure
+quality between urban and rural areas in various advanced economies.
+Rural areas, predictably, fare worse. Yet whether this variation in
+infrastructure quality affects the spatial footprint of telecommuting
+has not previously been discussed, in part because telecommuting has not
+previously been the cause of greatest demand and pressure on internet
+services.
+
+There are indications that those who purchase high speed connections
+consume more data of all sorts and use their connections for a variety
+of purposes \citep{hauge2011consumer}. There is also a correlation
+between access to internet services and a reduction in household
+transport spend \citep{bris2017ict}. Independently of whether the
+additional internet use and reduced travel is because of increased
 telecommuting, these studies suggest that better internet services
 enable households to make savings and efficiencies, an example of the
-first level digital divide reinforcing the third level. Yet did the
-purchase of high speed connections and increased internet access also
-prepare households for long-term home-working, enforced by government
-restrictions? The extreme demand during the pandemic provides a new
-opportunity to understand how infrastructure accessibility, quality, and
-reliability affects telecommuting, particularly in light of the high
-volumes of bandwidth-intensive video conferencing required in order to
-avoid the face-to-face contact that could increase the spread of
-infection. We seek to answer how internet service resilience contributes
-to or reduces economic resilience when the latter is dependent upon the
-capability to work from home, and thus improve our understanding of this
-new digital divide. \textbf{RELATED TO YOUR POINT ABOVE, WE MAY CONCLUDE
-THIS IS NOT A NEW DIGITAL DIVIDE, BUT THAT IS STILL ANSWERING A USEFUL
-QUESTION - MIGHT END UP REWORDING AS INTERACTION BETWEEN DIGITAL
-DIVIDES? ET: OR REINFORCING THE 3RD LEVEL DIGITAL DIVIDE}
+first level digital divide reinforcing the third level.
 
-The multi-layered digital divides intersect with material divides and
-the economic geography of the UK. Following the regional economic
-resilience literature, which underlines the differentiated capacity of
-cities and regions to escape or recover from economic crises
+The extreme demand during the pandemic provides a new opportunity to
+understand how infrastructure accessibility, quality, and reliability
+affects telecommuting, particularly in light of the high volumes of
+bandwidth-intensive video conferencing required in order to avoid the
+face-to-face contact that could increase the spread of infection. We
+seek to answer how internet service resilience contributes to or reduces
+economic resilience when the latter is dependent upon the capability to
+work from home. We also aim to improve our understanding of the impact
+of first level digital division on telecommuting, and whether this
+results in much more fundamental third level digital division than has
+previously been perceived.
+
+Furthermore, these multi-layered digital divides intersect with material
+divides and the economic geography of the UK. Following the regional
+economic resilience literature, which underlines the differentiated
+capacity of cities and regions to escape or recover from economic crises
 \citep{martin2012regional, kitsos2018economic}, different places have
@@
 \citep{martin_north_south}. Various studies have illustrated severe
-inequalities between the north and the south part of the UK regarding,
-for example, skills and human capital, unemployment, productivity and
-prosperity \citep{lee2014grim, mccann2020perceptions, dorling2018peak}.
-Some scholars have even argued that the UK suffers some of the highest
-level of interregional inequalities in the global north
+inequalities between the north and the south regions of England
+regarding, for example, skills and human capital, unemployment,
+productivity and prosperity
+\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
+scholars have even argued that the UK suffers some of the highest level
+of interregional inequalities in the global north
 \citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
-digital divides associated to a certain extent with or shaped by the
-geography of the UK, but the intersection of the digital and material
-divides affects the capacity of places to overcome some of the economic
-effects of the Covid-19 pandemic. Importantly, this is the first time
-that digital technologies became an essential tool for economic
-resilience for such a great part of the population.
+digital divides associated with or shaped by the geography of the UK,
+but the intersection of the digital and material divides affects the
+capacity of places to overcome some of the economic effects of the
+Covid-19 pandemic. Importantly, this is the first time that digital
+technologies became an essential tool for economic resilience for such a
+great part of the population.
 
@@
 
-The starting point of our methodological framework is cluster analysis,
+\hypertarget{sec:3.1}{%
+\subsection{Time-Series clustering}\label{sec:3.1}}
+
+First we identified our methodological framework as cluster analysis,
 which can be defined within the modern machine learning framework as an
@@
 \citep{aghabozorgi2015time}. Therefore, cluster analysis is a widely
-used family of techniques in geography
-\citep{gordon1977classification, everitt1974cluster}. For instance,
-clustering methods are the basis of \emph{geodemographics}, a research
-domain which aims to create small area indicators or typologies of
-neighbourhoods based on various and sometimes diverse variables
-\citep{SINGLETON2009289, harris2005geodemographics}. Clustering
-techniques have also been employed to solve \emph{regionalisation}
-problems \citep{niesterowicz2016}.
+used in geography \citep{gordon1977classification, everitt1974cluster}.
+For instance, clustering methods are the basis of
+\emph{geodemographics}, a research domain which aims to create small
+area indicators or typologies of neighbourhoods based on diverse
+variables \citep{SINGLETON2009289, harris2005geodemographics}. They have
+also been employed to solve \emph{regionalisation} problems
+\citep{niesterowicz2016}. Common characteristics of these studies are
+the cross-sectional nature of the data they employ. Indeed, most
+clustering problems in geography deal with observations that are fixed
+in time. However, for this paper we are interested in internet speeds,
+which vary over time and, therefore, we create clusters of local
+authorities in the UK with similar temporal signatures of experienced
+internet speeds.
 
-Common characteristics of these studies are the cross-sectional nature
-of the data they employ. Indeed, most clustering problems in geography
-deal with observations that are fixed in time. However, for this paper
-we are interested in internet speeds, which vary over time and,
-therefore, create clusters of local authorities in the UK with similar
-temporal signatures of experienced internet speeds. Hence, we deviate
-from the established geographical clustering tools and employ
-time-series clustering methods.
-
-Time-series clustering methods have been developed in order to deal with
-clustering problems linked to, for instance, stock or other financial
-data, economic, governmental or medical data as well as machine
-monitoring
+To do so, we employ time-series clustering methods, which have been
+developed to deal with clustering problems linked to, for instance,
+stock or other financial data, economic, governmental or medical data as
+well as machine monitoring
 \citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
@@
 (ii) open questions regarding the choice of similarity measures in order
-to cluster similar times series objects together considering the whole
+to cluster similar times-series objects together considering the whole
 length of the time-series and overcoming issues around noise, outliers
@@
 
-Time-series clustering methods utilising the whole length of time-series
-can be grouped in three categories. The first -- model-based approaches
--- is based on recovering the underlying model for each time-series and
-then applying clustering algorithms on the model parameters of each
-time-series \citep{aghabozorgi2015time}. The main criticism is the
-cluster accuracy for nearby clusters \citep{mitsa2010temporal}. The
-second approach is based the formation of vectors of features derived
-from the original time-series. These new data of reduced dimensionality
-are then clustered using conventional clustering algorithms.
+For this paper we utilise a category of time-series clustering methods
+known as shape-based approaches. These methods match two sperate
+time-series objects based on the similarity of their shapes through the
+calculation of distances between the shapes, and are better equipped to
+capture similarities between short length time-series
+\citep{aghabozorgi2015time}, such as our data. We thus identify clusters
+of UK local authorities with similar temporal signatures -- i.e.~shapes
+-- of experienced internet speeds. The clusters are identified using a
+partitioning algorithm, which leads to non-overlapping clusters, and
+particularly the widely use \emph{k}-means algorithm because of the
+simplicity of the implementation and the interpretability of the
+results. This is an iterative algorithm, which begins with defining the
+desired number of clusters \emph{k}. Then each observation is randomly
+assigned to a cluster from the \([1,k]\) space. This initial cluster
+assignment is followed by iterations in order to minimise the distance
+between the centroids of the clusters and the observations assigned to
+these clusters \citep{james2013introduction}.
 
-For this paper we utilise the third category of time-series clustering
-methods known as shape-based approaches. These methods match two sperate
-time-series objects based on the similarity of their shapes through the
-calculation of distances between the shapes, and are thus better
-equipped to capture similarities between short length time-series
-\citep{aghabozorgi2015time}. This approach serves best this paper
-because (i) we identify clusters of UK local authorities with similar
-temporal signatures -- i.e.~shapes -- of experienced internet speeds and
-(ii) the length of our time-series is short (see the data discussion in
-this section).
-
-Another import element of time-series clustering is the actual
-clustering algorithm. Similar to the clustering of cross-sectional data,
-we can employ partitioning algorithms, which lead to non-overlapping
-clusters, hierarchical clustering, which classifies clusters at
-different levels, and fuzzy algorithms, which create overlapping
-clusters \citep{sardatime}. Because of the simplicity of the
-implementation and the interpretability of the results, we utilise here
-partitioning clustering based on the widely used \emph{k-Means}
-algorithm. This is an iterative algorithm, which begins with defining
-the desired number of clusters \emph{k}. Then each observation is
-randomly assigned to a cluster from the \([1,k]\) space. This initial
-cluster assignment is followed by iterations in order to minimise the
-distance between the centroids of the clusters and the observations
-assigned to these clusters \citep{james2013introduction}.
-
-There are a number of differences between the above described
-application of \emph{k-Means} for cross-sectional data and its
-application for times series data. Instead of creating clusters around
-centroids, a common approach is to create clusters around
-\emph{medoids}, which are representative time-series objects with a
-minimal distance to all other cluster objects \citep{sardatime}. Also,
-instead of calculating the Euclidean distance between centroids and data
-points, more complex distance measures need to be employed in order to
-capture the similarity between a time-series object and a medoid. A
-common distance measure for shape-based time-series clustering is
-Dynamic Time Warping (DTW). Using its underpinning dynamic programming
-algorithm, DTW compares two time-series objects to find the optimum
-warping path between them. DTW is widely used in order to overcome
-limitations linked to the use of Euclidean distance
+There are a number of differences between the application of
+\emph{k}-means for cross-sectional and times-series data. Instead of
+creating clusters around centroids, a common approach is to create
+clusters around \emph{medoids}, which are representative time-series
+objects with a minimal distance to all other cluster objects
+\citep{sardatime}. Also, instead of calculating the Euclidean distance
+between centroids and data points, more complex distance measures need
+to be employed to capture the similarity between a time-series object
+and a medoid. A common distance measure for shape-based time-series
+clustering is Dynamic Time Warping (DTW), an algorithm comparing two
+time-series objects to find the optimum warping path between them. DTW
+is widely used in order to overcome limitations linked to the use of
+Euclidean distance
 \citep{sardatime, berndt1994using, ratanamahatana2004everything}. The
@@
 
-To assess the quality and reliability of internet across local
-authorities in the UK during the time when the population were told to
-work from home if at all possible we utilise unique data comprising
-individual internet speed tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
-This a private company that allows internet users to check their own
+\hypertarget{sec:3.2}{%
+\subsection{Experienced Broadband Speeds}\label{sec:3.2}}
+
+To assess the internet quality and reliability across local authorities
+in the UK, we utilise unique data comprising individual internet speed
+tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
+This is a private company that allows internet users to check their own
 broadband upload and download speeds, and stores every speed-check with
@@
 of local loop unbundling regulatory processes
-\citep{nardotto2015unbundling}. By using volunteered geographic data, we
-are able to assess the \emph{experienced} internet speed by users, which
-may differ from the \emph{advertised} maximum speeds of Internet Service
-Providers (ISPs). We are particularly interested in upload speeds and
-the frequency of speed tests over the period from March to May 2020, as
-government statements indicate this is when UK workers were told to work
-from home if at all possible \textbf{{[}need ref?{]}}. Average upload
-speeds are slower than average download speeds, at \(9.3\)Mb/s mean
-upload speed for the whole sample during the period of interest,
-compared to \(29.6\)Mb/s for download speeds, but they are also less
-associated with internet-based, high-demand, leisure activities such as
-video streaming. Therefore, upload speeds are more relevant to
-work-related activities such as uploading documents or two-way audio,
-video, and text-based communication systems. Meanwhile, the frequency of
-speed tests was important in identifying the temporal profile which woud
-give us most insight into experienced internet service and resilience,
-and provide an indication of the volume of experience over particular
-units of time.
+\citep{nardotto2015unbundling}, and we followed the former's approach to
+remove outliers. By using this volunteered geographic data, we are able
+to assess the internet speed \emph{experienced} by users, which may
+differ from the maximum speeds \emph{advertised} by ISPs. We are
+particularly interested in upload speeds and the frequency of speed
+tests over the period from March to May \(2020\), as government
+statements indicate this is when UK workers were first told to work from
+home if at all possible \citep{GovUK2020}. Average upload speeds are
+slower than average download speeds, at \(9.3\)Mb/s mean upload speed
+for the whole sample, compared to \(29.6\)Mb/s for download speeds, but
+they are also less associated with internet-based, high-demand, leisure
+activities such as video streaming. Therefore, upload speeds are more
+relevant to work-related activities such as uploading documents or
+two-way audio, video, and text-based communication systems.
 
-The first step in the workflow after dropping some outliers following
-\citet{riddlesden2014broadband} was to transform the individual,
-geolocated and time-stamped tests to more meaningful aggregates both in
-terms of space and time. The frequency of testing indicates that whilst
-there is an overall trend of increased testing \textbf{or increased
-experience of internet service NOT SURE WHAT THIS MEAND} from March to
-April and then a slight reduction from April to May, this trend masks
-substantial variation by not only the day of the week, but also time of
-day, as can be seen in Figure \ref{test2020}. Thus, a daily aggregation
-of upload speeds would mask the variation in experience over the course
-of each weekday. Furthermore, the importance of this variation is
-highlighted by a comparison with the same period in \(2019\), as in
-Figure \ref{test2019}, when the volume of testing and thus of experience
-of internet service quality was much more tied to the evening leisure
-peak and presumably to download speeds. Since the increase in testing
-during the working day in \(2020\) is an indication that users have a
-greater perception of the variable quality of internet service,
-particularly during a new morning peak of testing for service
-reliability, we decided to include a measure of hourly variation in our
-temporal profiles. However, although this is a large data set
-(\(241,088\) individual tests during the working hours of study period),
-there are not enough observations for each Local Authority District
-(LAD) and for each working hour of each working day (\(631\) speed tests
-per LAD on average) to profile speeds at that level of detail.
-Therefore, we aggregate all the speed-checks during the \(13\) weeks of
-March to May inclusive for weekdays in \(2020\) by each hour of the day
-and day of the week. As our research aims to identify the geography of
-internet service resilience for work purposes, bank holidays and the
-hours between midnight and 6:00 were excluded, as well as weekend days.
+The frequency of speed tests was important in identifying the temporal
+profile which would give us most insight into experienced internet
+service and resilience over units of time. Whilst there is an overall
+trend of increased testing from March to April and then a slight
+reduction from April to May, this trend masks substantial variation by
+not only the day of the week, but also time of day, as can be seen in
+Figure \ref{test2020}. Thus, a daily aggregation of upload speeds would
+mask the variation in experienced service over the course of each
+weekday. Furthermore, the importance of this variation is highlighted by
+a comparison with the same period in \(2019\), as in Figure
+\ref{test2019}, when the volume of testing and thus of experience of
+internet service quality peaked in the evening, presumably in response
+to demand for leisure activities and download speeds. In contrast, the
+majority of the increase in testing in \(2020\) is during the working
+day, creating a new morning peak in Figure \ref{test2020}. Therefore, we
+include a measure of hourly variation in our temporal profiles to
+reflect the change in users' perception of the workday reliability of
+internet services.
+
+However, there were insufficient observations -- only \(631\) speed
+tests per LAD on average -- for each for each working hour of each
+working day in each Local Authority District (LAD) to profile speeds at
+that level of detail. Spatial aggregation was also necessary because we
+could not follow individuals or households and connect data points.
+Therefore, we aggregate the \(241,088\) individual, geolocated and
+time-stamped speed-checks during the \(13\) weeks of March to May
+inclusive for weekdays in \(2020\) by each hour of the day and day of
+the week. As our research aims to identify the geography of internet
+service resilience for work purposes, bank holidays and the hours
+between midnight and \(6:00\) were excluded, as well as weekend days.
 The composite week time-series thus comprise \(18\) hours multiplied by
-\(5\) weekdays or \(90\) time points per series. We also aggregate these
-data spatially because we could not follow individuals or households and
-connect data points. The time-series were calculated for each of the
-\(382\) LADs in the UK, standardised, and then a \emph{k}-means
-partitioning around medoids clustering algorithm was applied using DTW.
-We initially run the algorithm for \(k \in \mathbb{N} \bigcap [5,15]\)
-and used cluster validity indices (CVIs) to pick the optimal solution of
-\(k = 13\). Following \citet{sardatime} the majority vote for the
-following CVIs was used: Silhouette (max), Score function (max),
-Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin
-(DB*, min), Dunn (max), COP (min).
+\(5\) weekdays or \(90\) time points per series. The time-series were
+calculated for each of the \(382\) LADs in the UK, standardised, and
+then a \emph{k}-means partitioning around medoids clustering algorithm
+was applied using DTW. We initially run the algorithm for
+\(k \in \mathbb{N} \bigcap [5,15]\) and used cluster validity indices
+(CVIs) to pick the optimal solution of \(k = 13\). Following
+\citet{sardatime} the majority vote for the following CVIs was used:
+Silhouette (max), Score function (max), Calinski-Harabasz (max),
+Davies-Bouldin (min), Modified Davies-Bouldin (DB*, min), Dunn (max),
+COP (min).
 
@@
 
-We then review the characteristics of each cluster, including number of
-LADs, and the temporal profile of upload speed by hour of the day and
-day of the composite week. Since the quality and reliability of internet
-services vary in time and space due to both supply and demand-side
-influences, we use a number of different measures to then describe
-experienced upload speeds per cluster. These include: a) mean,
-experienced connection speed, b) standard deviation or the amount of
-fluctuation from the mean, and c) the variation in speeds at particular
-times of day when working from home is more likely to take place. We
-take account of all three measurements in our descriptives statistics of
-upload speeds in order to determine how resilient broadband speeds are
-as experienced in different parts of the UK during a time of extreme
-demand.
+In Section \protect\hyperlink{sec:4.1}{4.1}, we review the temporal
+profile of upload speed by hour of the day and day of the composite week
+for each cluster. Since the quality and reliability of internet services
+vary in time and space due to both supply and demand-side influences, we
+also use a number of different measures to describe experienced upload
+speeds per cluster. These include: i) mean, experienced connection
+speed, ii) standard deviation or the amount of fluctuation from the
+mean, and iii) the variation in speeds during the new morning peak of
+testing when working from home is more likely to take place. We take
+account of all three measurements in order to determine how resilient
+broadband speeds are as experienced in different parts of the UK during
+a time of extreme demand.
 
-\textbf{More data details and some figures, descriptive stats}
-
-The cause of these different experiences of broadband resilience may be
-different in different areas, as they may reflect either similarities in
+The cause of these different experiences of broadband resilience may
+vary across different areas, as they may reflect either similarities in
 patterns of demand or similar quality of infrastructure. Our approach is
@@
 quality and reliability of upload speeds. Instead, we run an auxiliary
-regression in order to understand how the spatial and temporal patterns
-of internet service relate to the economic geography of the UK. We
-discuss how the different patterns might support or undermine efforts to
-work from home and maintain safe productivity and whether they reinforce
-existing spatial and social inequalities. More specifically, we estimate
-the following multinomial model:
+regression to understand how the spatial and temporal patterns of
+internet service relate to the economic geography of the UK. More
+specifically, we estimate the following multinomial logit model:
 
@@
 
-The identification strategy is as follows. Based on the outcomes of the
-time-series clustering, we identify \(J\) distinct and crisp clusters.
-We then regress this cluster membership against a vector \(X_{i}\) of
-socio-economic and geographic variables, which are discussed in detail
-in the relevant \protect\hyperlink{sec:4.2}{section}. This analysis
-enables us to provide a more nuanced understanding of how telecommuting
-and technology intersect at a time of extreme demand, and what lessons
-this time has for a future where telecommuting is likely to remain a
-common means of accessing work and broadband services, as well as
-infrastructure, must be fit for purpose.
+Based on the outcomes of the time-series clustering, we identify \(J\)
+distinct and crisp clusters. We then regress this cluster membership
+against a vector \(X_{i}\) of socio-economic and geographic variables,
+which are discussed in detail in the relevant Section
+\protect\hyperlink{sec:4.2}{4.2}. Because we cannot identify individuals
+or households and, consequently, we aggregated our data at the LAD
+level, our results offer correlations between the socioeconomic
+characteristics of certain geographic locations and internet service
+quality, not a record of who was telecommuting. Such individual data
+could be found though surveys, but these offer less detailed information
+about the experience of internet resilience due to enforced demand,
+which is the main contribution of this paper. Our auxiliary regression,
+therefore, provides an indication of how internet connectivity can
+reinforce or redress existing spatial and social inequalities in
+different places. However, it opens a path to future research by
+highlighting the importance of\\
+understanding of how telecommuting capabilities and digital
+infrastructure divisions intersect.
 
@@
 upload speeds per hour per day for each cluster, with the largest, in
-terms of the LAD membership, five clusters in Figure \ref{UpClusterL},
-and the next six in Figure \ref{UpClusterS}. These figures and table
-provide a comprehensive overview of the quality and reliability of
-experienced broadband in different parts of the UK, the temporal
-clusters offering a novel approach to understanding spatial disparity.
+terms of the LAD membership and population, five clusters in Figure
+\ref{UpClusterL}, and the next six in Figure \ref{UpClusterS}. These
+figures and table provide a comprehensive overview of the quality and
+reliability of experienced broadband in different parts of the UK.
 
@@
 
-The largest two clusters, comprising \(281\) local authorities and
-almost \(50\) million people, fall into clusters \(6\) and cluster
-\(13\). Cluster \(13\) has one of the lowest aggregate mean upload speed
-of any of the clusters, and the second highest ratio of the standard
-deviation to the mean. This suggests that those living in local
+The second largest cluster, comprising \(126\) local authorities and
+over \(20\) million people, is cluster \(13\), which has the slowest
+aggregate mean upload speed, and the second highest ratio of the
+standard deviation to the mean. This suggests that those living in local
 authorities in this cluster experienced some of the lowest quality
-broadband services in terms of upload speeds in the UK. However, the
-variation in upload speeds in cluster \(13\), which can be an indication
-of its reliability, does not seem to disproportionately affect the
-morning peak from 9:00-10:59, as upload speeds are, on average, only
-\(2.6\)\% slower than in the evening peak period between 19:00 and
-20:59. Although cluster \(13\) does include some rural, remote areas of
-the country, such as Northwest Scotland, Cornwall and Powys in Wales as
-shown on Figure \ref{map.up.clusters}, it also includes the major
-metropolitan centres of Bristol, Liverpool, and Newcastle, as well as
-nine (of \(32\)) London Boroughs and plenty of home county and suburban
-areas. Meanwhile, those living in cluster \(6\) experience aggregate
-mean upload speeds of about \(1\)Mb/s faster than those in cluster
-\(13\), but still lower than the other three large clusters and most of
-the smaller clusters, suggesting a middling quality of service. However,
-the time profile for cluster \(6\) in Figure \(3\)a shows that upload
-speeds drop quickly from \(6:00\) to 7:00 on a Monday and peak between
-23:00 and midnight on Wednesday and Thursday, but tend to be lower
-during the working day. Indeed, those in cluster \(6\), including people
-living in Birmingham, Leeds and Sheffield, in \(12\) London Boroughs,
-and many suburban areas and smaller cities like Cardiff, Oxford and
-Cambridge experience mean upload speeds in the morning that are
-\(4.4\)\% lower than in the evening. This lower level of reliability may
-be as a result of increased demand, e.g.~for telecommuting, especially
-as there are few truly remote areas in this cluster.
+broadband services in terms of upload speeds in the UK. However, Figure
+\ref{UpClusterL} shows that the high standard deviation, which is one
+indication of unreliable internet, did not disproportionately affect the
+morning peak from \(9:00\)-\(10:59\), when upload speeds were, on
+average, only \(2.6\)\% slower than in the evening peak period between
+\(19:00\) and \(20:59\), when entertainment purposes are likely to be
+using the most bandwidth. In comparison, the five LADs in cluster \(1\),
+home to \(343\) thousand people, not only experience the second slowest
+mean upload speeds and the highest ratio of standard deviation to the
+mean, but are also much more affected during the morning peak.
 
-\begin{figure}
-\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-5}
-\end{figure}
+Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
+LADs and \(29.5\) million people -- experienced aggregate mean upload
+speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
+lower than the other three large clusters and most of the smaller
+clusters, suggesting a middling quality of service. The temporal profile
+for cluster \(6\) in Figure \ref{UpClusterL} shows that upload speeds
+are highest at \(6:00\) on a Monday morning and peak between \(23:00\)
+and midnight on Wednesday and Thursday, but tend to be lower during the
+working day. Furthermore, experienced mean upload speeds in the morning
+peak are \(4.4\)\% lower than in the evening peak -- a greater, more
+noticeable change than any of the other large clusters experience,
+suggesting poorer reliability during the working day. This difference is
+less, however than any of the clusters as shown in Figure
+\ref{UpClusterS}.
 
-However, whilst the difference in experienced mean upload speeds between
-9:00-10:59 and 19:00-20:59 in cluster \(6\) is greater than in any of
-the other large clusters, it is smaller than this difference in all the
-clusters shown in Figure \ref{UpClusterL}. Whilst this may suggest a
-lack of reliability, especially considering the large spikes and dips in
-these clusters, the statistics in Table \ref{up.cluster.descr} show a
-more varied picture. The experienced upload speeds in cluster \(1\) are
-the second slowest of any cluster, the ratio of standard deviation to
-mean is the highest, as is the ratio of speeds in the morning compared
-to the evening peak hours analysed. The five, mainly rural authorities
-in this cluster clearly lack broadband resilience, for telecommuting or
-otherwise. Home to almost \(3.5\) thousand people, the districts are
-scattered around the country: Ceredigion in West Wales, Darlington in
-the Northeast, Eden between the Lake District and the North Penines,
-Rossendale in the Northwest, and Rutland in the East Midlands. In
-comparison, clusters \(3\) and \(10\) have relatively high mean upload
-speeds, and low standard deviations, suggesting reliability and
-resilience, as well as quality broadband services, despite slower speeds
-in the morning peak. Like cluster \(6\), is this also a sign of
-increased telecommuting? The eight LADs in clusters \(3\) and \(10\) are
-home to over one million people, and are all more urban and suburban
-than cluster \(1\), including the London Borough of Newham, the
-Southeast commuter towns of Harlow and Luton, the Birmingham suburbs of
-Bromsgrove and Cannock Chase, Dundee City, and the towns of Eastbourne
-and Corby.
+Clusters \(8\) and \(12\) also have mean upload speeds under \(10\)Mb/s,
+but higher than clusters \(1\), \(13\), and \(6\). However, mean upload
+speeds are much lower between \(9:00-10:59\) than between
+\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning
+than in the evening in cluster \(12\). Indeed, cluster \(12\) is the
+only cluster to experience higher speeds in the morning peak compared to
+the evening, and thus the only cluster where the temporal profile of
+internet use is closer to what might have been expected pre-pandemic.
+And yet even among the other clusters, the reliability of internet
+services during the working day still varies considerably. Interpreting
+this variation from the large spikes and dips shown on Figures
+\ref{UpClusterL} and \ref{UpClusterS} is difficult, but the statistics
+in Table \ref{up.cluster.descr} show that clusters \(9\) and \(11\) have
+the most reliable internet services. The ratio of standard deviation to
+mean in both these clusters is lowest, and speeds are only about 2\%
+slower in the morning than the evening. Mean speeds are also higher than
+in any other cluster, excluding cluster \(2\), where measures of
+reliability suggests poorer performance.
 
-Meanwhile, the most resilient broadband service in terms of both quality
-and reliability appears to be found in clusters \(9\) and \(11\), home
-to over \(12\) million people. The mean speeds for these clusters are
-second and third highest, the ratio of the standard deviation to the
-mean are the two lowest, and the experienced upload speeds in the
-morning peak are not that different from those experienced in the
-evening peak. In Figure \ref{UpClusterL}, cluster \(11\) shows more
-noticeable peaks and troughs, but the lowest points are not at the peak
-times described in Table \(1\). Rather, the slowest upload speeds on
-average occur between 6:00-7:00 on Monday morning, 14:00-15:00 on
-Friday, and \(16:00-17:00\) on Thursday. These slowest times are still
-mostly faster than the average hourly upload speeds in cluster \(13\).
-Although not all of the local authorities in these two clusters are
-urban, among the \(65\) LADs are Manchester and three of the nine other
-boroughs of Greater Manchester, Glasgow, Nottingham, both Portsmouth and
-Southampton, eight London Boroughs, four of the seven constituent
-authorities of the West Midlands Combined Authority. There are also a
-number of other tightly bounded urban areas, such as Aberdeen,
-Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at
-the centre of less confined districts like Burnley, Milton Keynes, and
-Northampton. The local authorities in these two clusters clearly
-experienced resilient broadband that could support high levels of
-telecommuting. They are \emph{not} on the wrong side of the first level
-digital divide, but how likely are they to be able to take advantage of
-their resilient ICT infrastructure and services?
+Thus, broadband services in clusters \(9\) and \(11\), home to over
+twelve million people performed the best during the study period, in
+terms of both quality and reliability. In Figure \ref{UpClusterL},
+cluster \(11\) shows more noticeable peaks and troughs, but the lowest
+points are not during the morning peak, but occur, on average, between
+\(6:00-7:00\) on Monday morning, \(14:00-15:00\) on Friday, and
+\(16:00-17:00\) on Thursday. These slow periods still offer better
+speeds than the average hourly profile of cluster \(13\). Finally,
+ignoring the smallest clusters in terms of population, that is clusters
+\(4\) and \(5\), clusters \(3\), \(7\) and \(10\) also have relatively
+high mean upload speeds. Clusters \(3\) and \(10\) pair high mean speeds
+with low standard deviations relative to the mean speeds, suggesting
+reliability and resilience, as well as quality broadband services.
+Cluster \(7\) has a higher ratio of standard deviation to mean, but
+there is less difference in average speeds between the morning and
+evening peaks than in clusters \(3\) and \(10\).
+
+In summary, LADs in clusters \(9\) and \(11\) experienced resilient
+broadband internet that could support high levels of telecommuting.
+Those in clusters \(2\), \(3\), \(7\), and \(10\) also experience higher
+than average mean speeds and rank middle to high on measures of service
+reliability. These LADs are \emph{not} on the wrong side of the first
+level digital divide, but how likely are they to be able to take
+advantage of their resilient ICT infrastructure and services? Meanwhile,
+cluster \(6\) is not only the largest in terms of number of LADs and
+population, it has the closest mean upload speed to the pre-clustered
+average for the whole sample. As well as average quality internet
+services, those in cluster \(6\) also experience average reliability for
+work purposes, ranking fifth behind the four other clusters with
+populations over one million, but ahead of the smaller clusters.
+Clusters \(8\) and \(12\) are also close to average mean upload speeds,
+but show very different patterns in terms of reliability, whilst
+clusters \(1\) and \(13\) appear to suffer most from a lack of quality
+internet services, with slow speeds and high standard deviations. With
+those in cluster \(1\) more likely to experience that poor reliability
+during the morning peak, is this first level digital divide occurring in
+areas where few are occupationally able to telecommute anyway, and what
+are the implications for economic resilience?
 
 \hypertarget{sec:4.2}{%
-\subsection{aux regressions}\label{sec:4.2}}
+\subsection{Post-clustering regression analysis}\label{sec:4.2}}
 
@@
 
-Clusters \(3\), \(9\), \(10\) and \(11\) seem to benefit most from high
-quality and resilient broadband services. The dips in mean upload speeds
-in clusters \(6\), \(3\) and \(10\) during the morning peak are
-suggestive of more use during the working day, and potentially more
-telecommuting. Using auxiliary regressions, we test whether the clusters
-that have higher mean speeds and more reliable services are more urban,
-more wealthy, and / or more likely to benefit from a choice of high
-quality internet services. We also measure which of our clusters are
-more likely to have a higher proportion of occupations which enable
-telecommuting because of the nature of the work. The results of these
-auxiliary regressions are presented in Table \ref{aux}. The dependent
-variable is the LAD cluster membership as described in the
-\protect\hyperlink{sec:3}{Methods and data} section and equation
+Using an auxiliary multinomial regression, we test whether the clusters
+that have higher mean speeds and more reliable services consist of LADs
+that are more urban, affluent, and / or more likely to benefit from a
+choice of high quality internet services. We also estimate which of our
+clusters are more likely to have a higher proportion of occupations
+which enable telecommuting because of the nature of the work. The
+results of the auxiliary regression are presented in Table \ref{aux}.
+The dependent variable is the LAD cluster membership as described in the
+\protect\hyperlink{sec:3}{methods and data} section and equation
 \ref{eq1} and each column represent a different cluster. The reference
-case is cluster \(4\), which includes the local authority of Hambleton
-in North Yorkshire, a rural area of just over ninety thousand people.
-Mean, experienced upload speeds in cluster \(4\) are close to both the
-average speeds for the \(13\) clusters (\(9.9\)Mb/s) and the
-pre-clustered average for the whole sample (\(9.3\)Mb/s). The standard
-deviation for cluster \(4\) and the ratio of evening to morning peak
-speeds are indications of worse reliability than many of the other
-clusters. Hence, the results in Table \ref{aux} should be seens as
-relative rather than absolute probabilities.
+case is cluster \(4\), which includes only the local authority of
+Hambleton in North Yorkshire, a rural area of just over ninety thousand
+people. Mean, experienced upload speeds in cluster \(4\) (see Table
+\ref{up.cluster.descr}) are close to both the average speeds for the
+\(13\) clusters (\(9.9\)Mb/s) and the pre-clustered average for the
+whole sample (\(9.3\)Mb/s). However, the standard deviation for cluster
+\(4\) and the difference between average speeds in the morning compared
+to the evening peak periods are indications of worse reliability than
+many of the other clusters. Hence, the results in Table \ref{aux} should
+be seen as relative rather than absolute probabilities.
 
-First, the number of speed tests run per cluster inhabitant between
-9:00-10:59 is an indication of satisfaction or at least a lack of
-concern over broadband speeds and quality of service. People in clusters
-\(3\), \(9\), \(10\) and \(11\) all ran fewer speed tests (at a per
-capita basis) at this time of day than people in clusters \(6\) and
-\(13\). Those in clusters \(3\), \(9\), \(10\) and \(11\) also benefit
-from a higher proportion of fast Virgin cable connections than those in
-clusters \(6\) and \(13\). Since services provided by Virgin Media are
-only available to \(45\)\% of premises in the UK \citep{ofcom2016},
-where the more lucrative ICT market originally attracted the cable TV
-provider, this is an indication that people in these clusters are more
-likely to live on the right side of the first level,
-infrastructure-based digital divide. Another indication is if people in
-these clusters are more likely to be closer to London or a metropolitan
-area. The broadband speed tests run in the authorities in cluster \(3\)
-are more likely to be taking place close to London than those run in any
-of the other clusters, a result that makes sense considering two of the
-four authorities in cluster \(3\) are the London commuter towns of
-Harlow and Luton. However, even though London was also included in the
-variable calculating distance to one of the ten major metropolitan areas
-in England, or Glasgow or Cardiff, tests run in cluster \(3\) are likely
-to be furthest away. \textbf{Did you include London in distMet as well?
-- code suggests so} \textbf{Do we need to re-think this variable
-(e.g.~remove Cardiff, add Leicester or\ldots)}
+First, we control for the number of speed tests run per cluster
+inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
+Virgin Media internet connections. Regarding the former, we expect
+people in LADs with more unreliable connections to test their internet
+speeds more often, and the results validate our priors. Meanwhile, fast
+Virgin Media cable connections have historically only been available to
+\(45\)\% of premises in the UK \citep{ofcom2016}, where the more
+lucrative and competitive market originally attracted the cable TV
+provider. Those in clusters \(2\), \(3\), \(9\) and \(11\) benefit from
+a higher proportion of Virgin connections, which is an indication that
+people in these clusters are more likely to live in urban areas, with
+more choice of broadband services, or in other words, are more likely to
+be on the right side of the first level, infrastructure-based digital
+divide, as we expected from the analysis in Section
+\protect\hyperlink{sec:4.1}{4.1}.
+
+Then, we employ distance from London and from the nearest metropolitan
+area (including London) as two variables depicting peripherality, urban
+structure and, potentially, first level digital divides. The broadband
+speed tests run in the authorities in cluster \(3\) are more likely to
+be taking place close to London than those run in any of the other
+clusters, and two of the four authorities in cluster \(3\) are the
+London commuter towns of Harlow and Luton. However, even though London
+was also included in the variable calculating distance from the centre
+of one of either the ten largest metropolitan areas in England, or
+Glasgow or Cardiff, tests run in cluster \(3\) are likely to be furthest
+away. Thus, it is important to consider the membership of each cluster
+as well as the regression results. Corby and Eastbourne, the other two
+authorities in cluster \(3\) are, whilst not part of metropolitan areas,
+are also large, accessible towns.
+
+\begin{figure}
+\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-6}
+\end{figure}
+
+Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
+of a large metropolitan area, even though the four local authorities of
+cluster \(7\) include no central urban boroughs and only one LAD that is
+part of a metropolitan area of governance - Tameside in Greater
+Manchester. This may explain why those in cluster \(7\) are the second
+least likely to be served by Virgin Media. It is also a demonstration of
+the complexity of experienced broadband upload speeds as captured by
+time-series clustering, and how quality and reliability, not only
+availability, engender first level digital divides. Cluster \(1\), which
+our analysis suggests lacks broadband resilience, contains five, mainly
+rural authorities. However, they are scattered around the country --
+Ceredigion in West Wales, Darlington in the Northeast, Eden and
+Rossendale in the Northwest, and Rutland in the East Midlands -- and
+therefore, the results in Table \ref{aux} indicate that these
+authorities are closer to a metropolitan area than authorities in
+cluster \(3\), but furthest from London.
+
+Internet resilience is also more nuanced than our -- arguably crude --
+dummy variable depicting the North-South economic divide, which assigns
+\(1\) to LADs located in Greater London, Southeast, Southwest and East
+of England regions, and \(0\) to the rest. The authorities most likely
+to be in the South are those in cluster \(13\), which was identified as
+having the slowest mean upload speeds of any of the clusters, and thus a
+low level of service. However, cluster \(13\) includes some rural,
+remote areas of the country, such as Northwest Scotland, Cornwall and
+Powys in Wales as shown on Figure \ref{map.up.clusters}. It also
+includes major metropolitan centres in the North -- Liverpool, Newcastle
+-- and South -- Bristol, nine (of \(32\)) London Boroughs. There are
+also plenty of Southern home county and suburban areas. The standard
+deviation measure for cluster \(13\) is high, but speed variation is low
+during the morning peak, suggesting that the estimates for reliability
+are inconsistent. Considering that this is one of the largest clusters,
+and thus the averages incorporate more noise than some of the smaller
+clusters, it may be that the LADs in this cluster do not all suffer
+equally from a first level digital divide.
+
+Yet we need to consider the results for other variables in Table
+\ref{aux} to better determine whether clusters \(1\) and \(13\), which
+appear to suffer most from poor quality internet services, are also more
+likely to have a low skilled workforce, less able to benefit from
+telecommuting. Cluster \(1\) has the lowest proportion of educated
+people, and the lowest earnings, despite recording the highest
+proportion of managerial, professional jobs, and the second highest
+proportion of tech jobs. Rural areas such as those in cluster \(1\) are
+often home to many retired people, which might explain these results or
+perhaps, as these figures are relative, we could note that cluster \(1\)
+also has a greater proportion of skilled trades than other clusters. It
+seems likely that the first level digital divide reinforces other
+inequities in cluster \(1\), whilst those in cluster \(13\) are more
+likely to earn more and have a better education despite poor internet
+services. Cluster \(13\) also has the most businesses per inhabitant,
+but is somewhere in the middle in terms of job density. If this is an
+indication of a high number of SMEs, it could explain the variable
+internet quality, considering that small businesses have not been seen
+as the most valuable customers for higher speed broadband services if
+they are not located near residential customers \citep{ofcom2016}.
+
+Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
+there is more noise within the averages we use to measure internet
+quality and reliability. Our results indicate average mean speeds, and
+Table \ref{aux} shows that cluster \(6\) also falls towards the middle
+of the clusters on many of the socioeconomic variables. It ranks third
+or fourth out of the eleven in terms of the likelihood of having a
+higher proportion of managerial, professional, and tech jobs, as well as
+higher earnings, but is fourth from bottom for educational attainment.
+The LADs in cluster \(6\) are also diverse, with few truly remote areas,
+but urban areas throughout England, including Birmingham, Leeds,
+Sheffield, twelve London Boroughs, and many suburban areas and smaller
+cities like Oxford and Cambridge. The capital cities of the other UK
+nations, Belfast, Cardiff and Edinburgh, are also in this cluster,
+suggesting perhaps that the lower level of reliability discussed in
+Section \protect\hyperlink{sec:4.1}{4.1} may be due to increased demand,
+e.g.~for telecommuting, when it is not supported by the most resilient
+internet connections.
+
+In contrast, Clusters \(9\) and \(11\) enjoy resilient internet
+connections and are characterised by higher share of occupations that
+may benefit from the ability to telecommute (e.g.~managerial). LADs in
+Cluster \(9\) are more likely to host highly educated individuals with
+higher earnings than cluster \(11\), which has a negative coefficient
+for the NVQ4 variable. Cluster \(11\) has fewer businesses per
+inhabitant and the second highest job density and cluster \(9\) the
+second lowest. These coefficients might indicate that resilient
+broadband infrastructure generates higher returns for those in cluster
+\(9\), where slightly more slowdown in the AM peak was detected (see
+Table \ref{up.cluster.descr}). LADs in cluster \(9\) are also less
+likely to be in the South and those in cluster \(11\) more likely,
+although the coefficient for cluster \(9\) might be skewed by the
+presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
+different economic profile than England. Still, both clusters consist of
+larger LADs in terms of population, including the city centre among four
+of the ten districts of Greater Manchester, eight London Boroughs, four
+of the seven constituent authorities of the West Midlands Combined
+Authority, Nottingham, and both Portsmouth and Southampton. Among the
+\(65\) LADs are also a number of other tightly bounded urban areas, such
+as Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas
+at the centre of less confined districts like Burnley, the digital hub
+of Milton Keynes, and Northampton. These urban locations are on the
+right side of the first level digital divide, and the regression results
+suggest that internet resilience is supporting a wide range of small and
+large urban economies.
+
+LADs in clusters \(2\), \(3\), \(7\), \(10\) are also on the right side
+of the first level digital divide, experiencing higher than average mean
+speeds and ranking high to middle on measures of service reliability.
+However, only LADs in cluster \(2\) and \(10\) are more likely to have
+highly skilled work force to take advantage of working from home
+opportunities. Cluster \(2\) is comprised of just two LADs, with the
+lowest job density of any cluster -- these are East Lothian near
+Edinburgh, and North East Lincolnshire, home to the town of Grimsby.
+Cluster \(10\) is comprised of four LADs, including two peripheral
+suburban areas of Birmingham, the East London Borough of Newham, and
+Dundee. Suburbs are considered the most likely urban form in which
+telecommuters live \citep{e2018does}, and Dundee has a reputation for
+tech startups \citep{technation2017}. LADs in cluster \(7\), which
+include the suburbs South of Belfast, a suburban district of Greater
+Manchester, the villages between York and Leeds, and a couple of small
+West Midlands towns have the highest earnings, and thus may also be
+making the most of their resilient internet. In contrast, LADs in
+cluster \(3\) are associated more with lower skills and have the highest
+proportion of individuals in machine operation jobs. Arguably, LADs in
+clusters \(3\) benefit the least from their resilient internet
+infrastructure in an era when working from home became a vehicle for
+economic resilience.
+
+Finally, clusters \(8\) and \(12\) consist by LADs enjoying close to
+average mean upload speeds, but opposing patterns of internet
+reliability during the morning peak. LADs in cluster \(8\) are likely to
+host more individuals employed in tech and administrative occupations
+than any other cluster, as well as many in managerial occupations,
+whilst the opposite applies to skilled trade and leisure occupations.
+These LADs are characterised by a very small positive likelihood of more
+educated residents, but are not significantly likely to earn more than
+other clusters. Including suburbs near Leicester, Cardiff, Birmingham,
+the green belt area south of London straddling the M25 motorway, and
+rural West Devon, the LADs in cluster \(8\) seem to likely to have the
+skills and occupations that would benefit from quality internet
+services, but suffer from poor internet resilience and reliability.
+
+In comparison, LADs in cluster \(12\) are less likely to be able to
+benefit from quality internet services if they had them, with fewer
+individuals achieving NVQ4 or better and lower levels of occupations
+that would benefit from homeworking. Yet LADs in this cluster benefit
+from the second highest level of earnings, and have the largest stock of
+businesses per inhabitant and the highest job density. This density of
+businesses could be why cluster \(12\) is the only cluster with higher
+speeds in the morning peak during the study period. If people are at
+home, business premises might well have been abandoned. Cluster \(12\)
+is home to \(1.5\) million people spread across twelve LADs, including
+the City of Westminster but also a LAD in Northern Ireland, two in
+Wales, Preston in the Northwest, and Fenland in the East of England.
+This spatial diversity demonstrates that the temporal clustering of
+internet resilience is not necessarily spatially dependent, and digital
+divides do not necessarily overlap with economic ones.
 
@@
 
-The long-term effects of such drastic changes in telecommuting and
-attitudes towards working from home are difficult to predict.
+This paper offers a new perspective on telecommuting from the viewpoint
+of the complex web of digital divides. To do so, we employ novel data
+regarding experienced upload speeds and time-series clustering methods,
+a family of unsupervised machine learning techniques which are rarely
+utilised in geographical research. Fast, reliable internet connections
+are necessary for the population to be able to work from home. Although
+not every place hosts individuals in occupations which allow for
+telecommuting nor with the necessary skills to effectively use the
+internet to telecommute, this paper raises the issue that without good
+internet connectivity, these places would struggle more to achieve
+economic resilience in a period like the current pandemic in any case.
+Indeed, our analysis demonstrated that the temporal profiles of twelve
+of our thirteen clusters had slower upload speeds in the morning than in
+the evening. The opposite is likely to have been the norm prior to the
+pandemic, as the level of demand and bandwidth management is the most
+common cause of temporal variation in experienced speeds, and why
+evening download speeds, rather than daytime upload speeds have been
+used to benchmark the performance of internet services. Thus, the new
+patterns can be taken as evidence of widespread telecommuting and other
+daytime internet use which changed the temporal profile of internet
+activity throughout the UK, not just in areas with more digital industry
+or better skills.
+
+Upload speeds have not previously been seen as integral to universal
+service, considering there has never before been such extreme demand for
+telecommuting and operations such as video calls. This may be why the
+average upload speeds for the largest two clusters -- \(6\) and \(13\)
+-- contain so many LADs that are centres of the knowledge economy, from
+Oxford and Cambridge in the former, to areas like Reading, which has the
+highest concentration of digital businesses in the country
+\citep{technation2017}. Thus, whilst some areas suffer from an
+intersection of digital and economic divides, such as those in cluster
+\(1\), others found their digital infrastructure to be less reliable
+when confronted with the sudden change in the timing and type of demand.
+Still other LADs, such as Milton Keynes in cluster \(9\) were able to
+benefit both from reliable internet connections and populations which
+were familiar with working from home and could capitalise on their
+digital infrastructure. Yet the nuanced picture we gained through our
+analysis of the UK case study suggests that quality internet
+connectivity would also enable other LADs to gain ground during the
+pandemic despite lower skills or earnings currently, especially in a
+future where telecommuting might be a more common means of accessing
+work and broadband services must be fit for purpose.
+
+Indeed, the long-term effects of such drastic changes in telecommuting
+and attitudes towards working from home are difficult to predict.
 Nevertheless, they span through various aspects of economy and society:
 from changes to transportation planning due to altered commuting
-patters, to changes in land use and urban planning to accommodate people
-who work from home \citep{BUDNITZ2020102713}\textbf{also 2020 Swedish
-article from JTG}; and from productivity and innovation changes, to
-changes in agglomeration externalities and the attraction of large
-cities \citep{econobs} just to name a few. This paper is positioned to
-support endeavours in understanding the effects of increased
-telecommuting by exposing the spatial and social dimensions of
-telecommuting including the resilience of broadband speeds in terms of
-both quality and reliability of service, and whether this reinforces or
-redresses prior digital divisions. \textbf{took this bit you wrote to
-put in the discussion at the end?}
+patterns to changes in land use and urban planning to accommodate people
+who work from home \citep{BUDNITZ2020102713, ELLDER2020102777}, and from
+productivity and innovation changes to changes in agglomeration
+externalities and the attraction of large cities \citep{econobs}.
+Further research may be able to measure the economic resilience of the
+different clusters of places discussed in this paper once this pandemic
+is firmly past. However, our analysis demonstrates that the economic
+resilience made possible by working from home cannot be understood
+without considering the underpinning digital divides and cannot be
+achieved without planning for how the levels of digital, social and
+economic divides might intersect.
 
-\hypertarget{acknowledgements}{%
-\section*{Acknowledgement(s)}\label{acknowledgements}}
-\addcontentsline{toc}{section}{Acknowledgement(s)}
+\hypertarget{appendix1}{%
+\section{Appendix 1}\label{appendix1}}
 
-An unnumbered section,
-e.g.~\texttt{\textbackslash{}section*\{Acknowledgements\}}, may be used
-for thanks, etc.~if required and included \emph{in the non-anonymous
-version} before any Notes or References.
+This is the LAD cluster membership for the upload speed timeseries.
 
-\hypertarget{funding}{%
-\section*{Funding}\label{funding}}
-\addcontentsline{toc}{section}{Funding}
+\textbf{Cluster 1: } Ceredigion, Darlington, Eden, Rossendale, Rutland
 
-An unnumbered section,
-e.g.~\texttt{\textbackslash{}section*\{Funding\}}, may be used for grant
-details, etc.~if required and included \emph{in the non-anonymous
-version} before any Notes or References.
+\textbf{Cluster 2: } East Lothian, North East Lincolnshire
 
-\bibliographystyle{tfcad}
-\bibliography{bibliography.bib}
+\textbf{Cluster 3: } Corby, Eastbourne, Harlow, Luton
 
-\input{"appendix.tex"}
+\textbf{Cluster 4: } Hambleton
+
+\textbf{Cluster 5: } Fylde
+
+\textbf{Cluster 6: } Allerdale, Amber Valley, Angus, Ashfield, Ashford,
+Aylesbury Vale, Barnet, Basingstoke and Deane, Bath and North East
+Somerset, Belfast, Bexley, Birmingham, Blaenau Gwent, Bournemouth,
+Christchurch and Poole, Bradford, Braintree, Brentwood, Bridgend,
+Bromley, Broxtowe, Bury, Calderdale, Cambridge, Canterbury, Cardiff,
+Castle Point, Chelmsford, Cheltenham, Cherwell, Chesterfield, City of
+Edinburgh, Clackmannanshire, Colchester, Copeland, County Durham,
+Coventry, Croydon, Dartford, Daventry, Denbighshire, Derby, Derry City
+and Strabane, Dorset, Ealing, East Ayrshire, East Hampshire, East
+Lindsey, East Northamptonshire, East Renfrewshire, East Riding of
+Yorkshire, East Suffolk, Eastleigh, Elmbridge, Enfield, Falkirk,
+Fareham, Gateshead, Gedling, Gosport, Gravesham, Great Yarmouth,
+Guildford, Harborough, Haringey, Harrogate, Harrow, Hart, Hartlepool,
+Havering, Herefordshire, County of, High Peak, Hinckley and Bosworth,
+Horsham, Islington, Kettering, King's Lynn and West Norfolk, Kingston
+upon Thames, Kirklees, Leeds, Leicester, Lincoln, Maidstone, Maldon,
+Mansfield, Medway, Mendip, Mid Sussex, Middlesbrough, Monmouthshire,
+Moray, New Forest, Newcastle-under-Lyme, Newport, North Ayrshire, North
+East Derbyshire, North Hertfordshire, North Kesteven, North Lanarkshire,
+North Lincolnshire, North Norfolk, North Tyneside, North West
+Leicestershire, Northumberland, Nuneaton and Bedworth, Oxford,
+Pembrokeshire, Pendle, Renfrewshire, Ribble Valley, Rochford, Runnymede,
+Rushcliffe, Ryedale, Salford, Sefton, Sheffield, Shropshire, Solihull,
+South Ayrshire, South Hams, South Holland, South Lanarkshire, South
+Oxfordshire, South Staffordshire, St Albans, Staffordshire Moorlands,
+Stockport, Stoke-on-Trent, Surrey Heath, Sutton, Swale, Tamworth,
+Tendring, Test Valley, Thurrock, Tonbridge and Malling, Torfaen,
+Wakefield, Warrington, Warwick, Wealden, Wellingborough, West Berkshire,
+West Dunbartonshire, West Lancashire, West Lothian, West Oxfordshire,
+West Suffolk, Wigan, Wiltshire, Woking, Worcester, Wrexham, Wycombe,
+York
+
+\textbf{Cluster 7: } Lisburn and Castlereagh, Selby, Tameside, Wyre
+Forest
+
+\textbf{Cluster 8: } Lichfield, Oadby and Wigston, Tandridge, Vale of
+Glamorgan, West Devon
+
+\textbf{Cluster 9: } Aberdeen City, Barnsley, Broxbourne, Charnwood,
+Chorley, Erewash, Glasgow City, Greenwich, Halton, Havant, Knowsley,
+Lewisham, Merton, Mid and East Antrim, Milton Keynes, Newark and
+Sherwood, Northampton, Oldham, Portsmouth, Richmond upon Thames, Rugby,
+Sandwell, South Derbyshire, South Kesteven, South Northamptonshire,
+Southampton, Spelthorne, Stockton-on-Tees, Telford and Wrekin, Trafford,
+Walsall, Welwyn Hatfield
+
+\textbf{Cluster 10: } Bromsgrove, Cannock Chase, Dundee City, Newham
+
+\textbf{Cluster 11: } Barking and Dagenham, Blaby, Blackpool, Bolsover,
+Brent, Burnley, Caerphilly, Carlisle, Causeway Coast and Glens, Crawley,
+Doncaster, Dudley, Hertsmere, Hounslow, Hyndburn, Ipswich, Isles of
+Scilly, Kensington and Chelsea, Lewes, Manchester, North Warwickshire,
+Norwich, Nottingham, Peterborough, Redditch, Rochdale, Scarborough,
+Slough, St.~Helens, Stevenage, Sunderland, Vale of White Horse,
+Wolverhampton
+
+\textbf{Cluster 12: } Ards and North Down, Conwy, East Staffordshire,
+Epping Forest, Fenland, Hammersmith and Fulham, Preston, Rhondda Cynon
+Taf, Three Rivers, Westminster
+
+\textbf{Cluster 13: } Aberdeenshire, Adur, Antrim and Newtownabbey,
+Argyll and Bute, Armagh City, Banbridge and Craigavon, Arun, Babergh,
+Barrow-in-Furness, Basildon, Bassetlaw, Bedford, Blackburn with Darwen,
+Bolton, Boston, Bracknell Forest, Breckland, Brighton and Hove, Bristol,
+City of, Broadland, Camden, Carmarthenshire, Central Bedfordshire,
+Cheshire East, Cheshire West and Chester, Chichester, Chiltern, City of
+London, Cornwall, Cotswold, Craven, Dacorum, Derbyshire Dales, Dover,
+Dumfries and Galloway, East Cambridgeshire, East Devon, East
+Dunbartonshire, East Hertfordshire, Epsom and Ewell, Exeter, Fermanagh
+and Omagh, Fife, Flintshire, Folkestone and Hythe, Forest of Dean,
+Gloucester, Gwynedd, Hackney, Hastings, Highland, Hillingdon,
+Huntingdonshire, Inverclyde, Isle of Anglesey, Isle of Wight, Kingston
+upon Hull, City of, Lambeth, Lancaster, Liverpool, Malvern Hills,
+Melton, Merthyr Tydfil, Mid Devon, Mid Suffolk, Mid Ulster, Midlothian,
+Mole Valley, Na h-Eileanan Siar, Neath Port Talbot, Newcastle upon Tyne,
+Newry, Mourne and Down, North Devon, North Somerset, Orkney Islands,
+Perth and Kinross, Plymouth, Powys, Reading, Redbridge, Redcar and
+Cleveland, Reigate and Banstead, Richmondshire, Rother, Rotherham,
+Rushmoor, Scottish Borders, Sedgemoor, Sevenoaks, Shetland Islands,
+Somerset West and Taunton, South Bucks, South Cambridgeshire, South
+Gloucestershire, South Lakeland, South Norfolk, South Ribble, South
+Somerset, South Tyneside, Southend-on-Sea, Southwark, Stafford,
+Stirling, Stratford-on-Avon, Stroud, Swansea, Swindon, Teignbridge,
+Tewkesbury, Thanet, Torbay, Torridge, Tower Hamlets, Tunbridge Wells,
+Uttlesford, Waltham Forest, Wandsworth, Watford, Waverley, West Lindsey,
+Winchester, Windsor and Maidenhead, Wirral, Wokingham, Worthing,
+Wychavon, Wyre
+
+\hypertarget{appendix2}{%
+\section{Appendix 2}\label{appendix2}}
+
+\begin{table}[!htbp] \centering 
+  \caption{Descriptive statistics for the auxiliary regression explanatory variables\label{descr.aux}} 
+  \label{} 
+\footnotesize 
+\begin{tabular}{@{\extracolsep{0pt}}lccccccc} 
+\\[-1.8ex]\hline 
+\hline \\[-1.8ex] 
+Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St.
+Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Pctl(25)} & \multicolumn{1}{c}{Pctl(75)} & \multicolumn{1}{c}{Max} \\
+\hline \\[-1.8ex] 
+pop, 2018 & 365 & 174,952.100 & 119,557.100 & 8,700 & 100,400 & 214,900 & 1,141,400 \\ 
+job density, 2018 & 365 & 1.137 & 5.726 & 0.400 & 0.700 & 0.930 & 110.110 \\ 
+distance to nearest met. area & 365 & 53.269 & 57.700 & 0.150 & 22.050 & 69.290 & 544.090 \\ 
+distance to London & 365 & 201.558 & 173.634 & 0.150 & 76.180 & 278.880 & 1,003.950 \\ 
+south of the UK & 365 & 0.463 & 0.499 & 0 & 0 & 1 & 1 \\ 
+managerial jobs, 2020 & 363 & 12.009 & 4.013 & 3.600 & 9.000 & 14.300 & 27.900 \\ 
+tech jobs, 2020 & 364 & 14.505 & 4.057 & 3.500 & 11.800 & 16.900 & 29.600 \\ 
+skilled trade jobs, 2020 & 358 & 10.513 & 3.764 & 1.000 & 8.025 & 12.500 & 21.600 \\ 
+professional jobs, 2020 & 364 & 21.223 & 6.902 & 4.400 & 16.775 & 24.850 & 71.600 \\ 
+administrative jobs, 2020 & 359 & 9.965 & 2.738 & 3.200 & 8.100 & 11.400 & 21.300 \\ 
+leisure jobs, 2020 & 362 & 9.261 & 2.827 & 2.800 & 7.300 & 11.400 & 17.800 \\ 
+machine operation jobs, 2020 & 337 & 6.339 & 2.847 & 1.200 & 4.400 & 7.900 & 19.800 \\ 
+earnings, 2019 & 360 & 592.184 & 81.129 & 437.600 & 534.625 & 633.875 & 893.200 \\ 
+NVQ4+ & 365 & 39.329 & 11.076 & 15.000 & 31.800 & 45.300 & 100.000 \\ 
+Virgin Media \%, 2020 & 365 & 0.152 & 0.141 & 0.000 & 0.018 & 0.241 & 0.753 \\ 
+n. business est. per hab., 2019 & 365 & 0.057 & 0.164 & 0.023 & 0.038 & 0.056 & 3.174 \\ 
+AM tests per hab., 2020 & 365 & 0.0005 & 0.0002 & 0.0001 & 0.0003 & 0.001 & 0.001 \\ 
+\hline \\[-1.8ex] 
+\end{tabular} 
+\end{table}
+
+\pagebreak
+
+\bibliographystyle{sageh}
+\bibliography{bibliography}
 

### 2020-12-29 paper/v1_old/broadband.speed.covid.Rmd
abstract + directories

--- 2020-12-22 paper/v1_sage/broadband.speed.covid.Rmd
@@
 ---
-title: WFH and broadband speed (title needs rework)
+title: |
+  WFH and broadband speed (title needs rework)
+type: ARTICLE TEMPLATE
 author:
-- name: A. Uthor*
-  num: 1,2
-- name: O. Tro
-  num: 2
-- name: O. Vriga
-  num: 3
-address:
-- num: 1
-  org: Department of Incredible Research, University A, City A, Country A
-- num: 2
-  org: Department of Applied Things, University B, City B, Country B
-- num: 3
-  org: Very Important Stuff Committee, Institute C, City C, Country C
-corrauth: "Corresponding author name, This is sample corresponding address."
-email: correspondingauthor@protonmail.com
-abstract: "TBC"
-keywords:   covid; internet; working from home; broadband speed; time-series clusters;
-classoption:
-  - Royal
-  - times
-bibliography: bibliography
-bibliographystyle: sageh
+  - name: A. N. Author
+    affil: a
+    email: latex.helpdesk@tandf.co.uk
+  - name: John Smith
+    affil: b
+    email: john.smith@uni-freiburg.de
+affiliation:
+  - num: a
+    address: |
+      Taylor & Francis, 4 Park Square, Milton Park, Abingdon, UK
+  - num: b
+    address: |
+      Institut für Informatik, Albert-Ludwigs-Universität, Freiburg, Germany
+bibliography: bibliography.bib
+appendix: appendix.tex
+abstract: |
+  TBC
+keywords: |
+  covid; internet; working from home; broadband speed; time-series clusters
 header-includes: |
@@
   \def\tightlist{}
-  \newcommand{\beginappendix}{\setcounter{table}{0} \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
-output:
-  rticles::sage_article:
-    keep_tex: yes
+output: rticles::tf_article
 ---
@@
 Various studies have illustrated severe inequalities between the north and the
-south regions of England regarding, for example, skills and human capital,
+south regions of England **replace 'UK'** regarding, for example, skills and human capital,
 unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
@@
 However, the time profile for cluster $6$ in Figure \ref{UpClusterL} shows
-that upload speeds are highest at $6:00$ on a Monday morning
+that upload speeds drop quickly from $6:00$ to $7:00$ on a Monday
 and peak between $23:00$ and midnight on Wednesday and Thursday,
@@
 and the ratio of upload speeds in the peak periods indicates that speeds are
-only about 2% slower in the morning than the evening.
+only about 2% slower in the morning.
 Mean speeds are also higher than in any other cluster, excluding cluster $2$,
@@
 Those in clusters $2$, $3$, $7$, and $10$ also experience higher than average
-mean speeds and rank middle to high on measures of service reliability.
+mean speeds and rank high to middle on measures of service reliability.
 These LADs are *not* on the wrong side of the first level digital divide,
@@
 for the $13$ clusters ($9.9$Mb/s) and the pre-clustered average for the whole sample
-($9.3$Mb/s) as shown in Table \ref{up.cluster.descr}.
+($9.3$Mb/s) as shown in \ref{up.cluster.descr}.
 However, the standard deviation for cluster $4$ and the difference between average
@@
 and quality of service.
-People in LADs assigned to clusters $6$ and $13$ were more likely to run speed tests
+People in LADs assigned to clusters $6$ and $13$ were more likely to ran speed tests
 (at a per capita basis)
-at this time of day, whilst people in cluster $11$ were the least likely to do so.
+at this time of day, whilst people in cluster $11$ were the least likely ones.
 This high level of satisfaction matches the high quality and reliability of service
@@
 from the centre of one of either the ten largest metropolitan areas in England,
-or Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
+Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
 Still, whilst significant, the effect is tiny and the other members of cluster $3$
@@
 of $175$ thousand people between them. **move maps to here?**
-
 Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large
 metropolitan area, even though the four local authorities of cluster $7$ include
-no central urban boroughs and only one LAD that is part of a metropolitan area of
+no central urban boroughs, and only one LAD that is part of a metropolitan area of
 governance - Tameside in Greater Manchester.
@@
 which assigns $1$ to LADs locates in Greater London, Southeast, Southwest and East
-of England regions, and $0$ to the rest.
+of England regions.
 The authorities most likely to be in the South are those in cluster $13$,
@@
 It also includes the major metropolitan centres of Bristol, Liverpool,
-and Newcastle, as well as nine (of $32$) London Boroughs, 
-including the City of London and plenty of home county and suburban areas.
+and Newcastle, as well as nine (of $32$) London Boroughs and plenty of home county
+and suburban areas.
 Although by the standard deviation measure cluster $13$ has unreliable internet
-services, speed variation was less during the morning peak, suggesting that 
-the estimates for reliability are inconsistent. 
-Considering that this is one of the largest clusters, and thus the averages 
-incorporate more noise than some of the smaller clusters, 
-it may be that the LADs in this cluster do not all suffer equally from 
-a first level digital divide.
-
-Yet we need to consider the results for other variables in Table \ref{aux} to better
-determine whether clusters $1$ and $13$, which appear to suffer most from a lack 
-of quality internet services, are also more likely to have a low skilled workforce, 
-less able to benefit from telecommuting. 
-Cluster $1$ has the lowest propotion of educated people of any of the clusters,
-and the lowest earnings, despite recording the highest proportion of managerial, 
-professional jobs, and the second highest proportion of tech jobs.
-Rural areas such as those in cluster $1$ are often home to many retired people,
-which might explain these results or perhaps, as these figures are relative,
-we could note that cluster $1$ also has a greater proportion of skilled trades
-than other clusters.
-Either way, it seems likely that the first level digital divide reinforces 
-other inequities in cluster $1$.
-In comparison, those cluster $13$ are more likely to earn more and have a better
-education.
-Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the 
-middle in terms of job density.
-If this is an indication of a high number of SMEs, it might give some explanation
-for the variable internet quality, considering that small businesses have not been
-seen as the most valuable customers for higher speed broadband services **insert ref**.
-
-Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more 
-noise within the averages we use to measure internet quality and reliability,
-and the result indicates an average level of service, even if better than LADs
-in cluster $13$.
-Cluster $6$ also falls towards the middle of the clusters on many of the socio-
-economic variables in Table \ref{aux}. 
-It ranks third or fourth out of the eleven clusters in the table in terms of
-having a higher proportion of managerial, professional, and tech jobs, as well as 
-higher earnings, but is fourth from bottom in terms of education.
-The LADs in cluster $6$ are also diverse, with few truly remote areas, 
-but many urban areas in both the North, including Birmingham, Leeds and Sheffield, 
-and South, including twelve London Boroughs, and many suburban areas and 
-smaller cities like Oxford and Cambridge.
-The capital cities of the other UK nations, Belfast, Cardiff and Edinburgh, are
-also in this cluster, suggesting perhaps that the lower level of reliability 
-discussed in Section [4.1](#sec:4.1) may be as a result of increased demand,
-e.g. for telecommuting.
-
-In comparison, clusters $9$ and $11$, which benefit from the most resilient
-internet connections, are characterised by lower share of occupations that are  
-may benefit from the ability to telecommute (e.g. managerial and technical). 
-Cluster $9$ is more highly educated and higher earning than cluster $11$, 
-has fewer businesses per inhabitant and less job density.
-Indeed, cluster $11$ has a negative coefficient for the NVQ4 variable, whilst
-for cluster $9$ the coefficient is positive, and cluster $11$ has the second
-highest job density and cluster $9$ the second lowest. 
-The former might be an indication of the higher returns that resilient
-broadband infrastructure generates for cluster $9$ versus cluster $11$. 
-Indeed, we did detect slightly more slowdown in the AM peak in cluster $9$ than in
-cluster $11$ (see Table \ref{up.cluster.descr}).
-Meanwhile, a propensity for a high job density might indicate more labour intensive 
-economic structures in cluster $11$ than in the LADs in cluster $9$.
-
-The LADs in clusters $9$ and $11$ also have opposite signs for their likelihood 
-of being in the South of England.
-Those in cluster $9$ less likely to be in the South and those in cluster $11$ 
-more likely, although the coefficient for cluster $9$ might be skewed by 
-the presence of the Scottish cities of Glasgow and Aberdeen, 
-as Scotland has a different economic profile than England.
-Still, both clusters consist of larger LADs in terms of population, 
-including the urban centre of Manchester and three of the nine other districts of
-Greater Manchester, Nottingham, both Portsmouth and Southampton,
-eight London Boroughs, and four of the seven constituent authorities of the
-West Midlands Combined Authority.
-Among the $65$ LADs are also a number of other tightly bounded urban areas, 
-such as Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas 
-at the centre of less confined districts like Burnley, 
-the digital hub of Milton Keynes, and Northampton.
-Thus, our findings confirm the importance of geography and urban locations 
-for being on the right side of the first level digital divide, and suggest that 
-internet resilience cuts across other digital and economic divides.
-
-A similarly divided picture can be built for the next tier of clusters: $2$, $3$,
+services, speed variation was less during the morning peak.
+More broadly, the estimates for this variable are not consistent between the
+two clubs of LAD clusters: those who are in the wrong and the right side of
+digital divides.
+
+Regarding the rest of the socio-economic variables we see a varied
+picture. Clusters $9$ and $11$, which are the ones benefiting by the most resilient
+internet connections are likely to consist of LADs with higher share of managerial
+and highly skilled occupations as well as higher average earnings. Interestingly,
+both clusters are characterised by lower share of occupations that are not usually
+associated with teleworking (professional, associated professional and technical,
+administrative, leisure and machine operation jobs), but also by lower number of
+business per habitant and. However, the likelihood of hosting highly skilled
+residents is not consistent with cluster $11$ having a negative coefficient for the
+NVQ4 variable. This might be an indication of the higher returns that resilient
+broadband infrastructure generates for cluster $9$ versus cluster $11$. While both
+of these cluster consist more likely by large LADs in terms of population, job
+density if higher for LADs in cluster $11$ indicating more labour intensive
+economic structures contrary to LADs in cluster $9$.
+
+A similar divided picture can be built for the next tier of clusters: $2$, $3$,
 $7$, $10$. LADs in these clusters experience higher than average mean speeds and
-rank high to middle on measures of service reliability. 
-LADs in these clusters are on the right side of the first level digital divide. 
-However, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
-work force, which might more easily take advantage of working from home 
-opportunities.
-Cluster $2$ is comprised of just two LADs, with the lowest job density of any
-cluster -- these are East Lothian near Edinburgh, and North East Lincolnshire, 
-home to Grimsby.
-Cluster $10$ is comprised of four LADs, including two peripheral suburban areas 
-of Birmingham, the East London Borough of Newham, and Dundee.
-Suburbs are considered the most likely urban form in which telecommuters live, 
-and Dundee has a reputation for tech startups **insert ref**.
-Yet although those in clusters $2$ and $10$ may be making the most of their 
-resilient internet, those in cluster $7$ have the highest earnings.
-The four LADs in cluster $7$ include the suburbs South of Belfast, a suburban 
-district of Greater Manchester, the villages between York and Leeds, and a
-couple of small West Midlands towns.
-LADs in cluster $3$ are associated more with lower skills and have the highest
-proportion of individuals in machine operation jobs. 
-They include the Southeast commuter towns of Harlow and Luton and the slightly more
-distant towns of Eastbourne and Corby.
-Arguably, LADs in clusters $3$ benefit the least from their resilient internet 
-infrastructure in an era when working from home became a vehicle for economic 
-resilience.
-
-Finally, clusters $8$ and $12$, are characterised by LADs enjoying close to average 
-mean upload speeds, but different patterns of internet reliability. 
-LADs in cluster $8$ are likely to host more individuals employed in tech and
-administrative occupations than any other cluster, as well as many in managerial 
-occupations, whilst the opposite applies to skilled trade and leisure occupations. 
-These LADs are characterised by a very small positive likelihood of more educated 
-residents, but are not significantly likely to earn more than other clusters.
-Including suburbs of Leicester, Cardiff, Birmingham, the green belt area south of 
-London straddling the M25 motorway, and rural West Devon, the LADs in cluster $8$
-seem to likely to have the skills and occupations that would benefit from 
-quality internet services, but suffer from poor internet resilience and reliability.
-
-In comparison, LADs in cluster $12$ are less likely to be able to benefit from
-quality internet services if they had them, with fewer individuals achieving 
-NVQ4 or better and lower levels of occupations that would benefit from homeworking. 
-Yet LADs in this cluster benefit from the second highest level of earnings, and
-have the largest stock of businesses per inhabitant and the highest job density. 
-This density of businesses could be why cluster $12$ is the only cluster with higher
-speeds in the morning peak during the study period.
-If people are at home, business premises might well be abandoned.
-Cluster $12$ is home to $1.5$million people spread across twelve LADs, including
-the City of Westminster and the London Borough of Hammersmith.
-However, it also includes a LAD in Northern Ireland, two in Wales, two suburban areas
-North of London, Preston in the Northwest, East Staffordshire in the Midlands, and
-Fenland in the East of England.
-This spatial diversity demonstrates that the temporal clustering of internet 
-resilience does not create spatial clusters, and digital divides do not necessarily
-overlap with economic ones.
+rank high to middle on measures of service reliability. As mentioned before, LADs
+in these clusters are on the right side of the first level digital divide. However,
+while all of them are more likely to be populated by people employed in managerial
+jobs, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
+work force, which can more easily take advantage of working from home opportunities.
+A more mixed picture can be built for LADs in clusters $3$ and $7$, which are
+associated, on the one hand, with lower skills and, on the other hand, with higher
+likelihood of individuals in managerial occupations. A similar contradicting
+pattern can be illustrated for cluster $6$, which is the largest one with average 
+quality of internet service. Arguably, LADs in clusters $2$ and $10$ are the ones
+benefiting more by resilient internet infrastructure in an era that working
+from home became a vehicle for economic resilience.
+
+Moving on to the next tier of clusters $8$ and $12$, which are characterised by
+LADs enjoying close to average mean upload speeds, but different patterns of internet
+reliability, we see a differentiated picture. On the one hand, both of them are
+likely to host individuals employed in managerial occupations while the opposite
+applies on skilled, professional and leisure occupations. On the other hand,
+LADs in cluster $8$ are characterised by a very small positive and LADs in cluster
+$12$ by a negative likelihood to host highly skilled individuals. Moreover, LADs
+in these two clusters have a rather
+similar profile regarding the occupational variables, but different tendencies
+regarding the stock of businesses. In total, LADs in these clusters do not
+benefit by great internet services and could only marginally benefit by working
+from home opportunities.
+
+Lastly, clusters $1$ and $13$, which appear to suffer most from a lack of quality
+internet services are more likely to also suffer from low skilled workforce. LADs
+in cluster $13$. Contradictory, the likelihood of workforce employed in managerial
+occupations is positive for LADs in both clusters, but only individuals in LADs
+in cluster $13$ are more likely to enjoy high earnings.
+
+[Appendix](#sec:appendix)
 
@@
 telecommuting and operations such as video calls.
+
+those in cluster $6$, including people living in Birmingham, Leeds and
+Sheffield, in twelve London Boroughs, and many suburban areas and smaller cities like
+Cardiff, Oxford and Cambridge
+This lower level of reliability may be as a result of increased demand,
+e.g. for telecommuting, especially as there are few truly remote areas
+in this cluster.
+
+The eight LADs in clusters $3$ and $10$ are home to over one million people,
+and are all more urban and suburban than cluster $1$, including the London Borough
+of Newham, the Southeast commuter towns of Harlow and Luton, the Birmingham suburbs
+of Bromsgrove and Cannock Chase, Dundee City, and the towns of Eastbourne and Corby.
+
+Although not all of the local authorities in these clusters $9$ and $11$ are urban,
+among the $65$ LADs are Manchester and three of the nine other boroughs of
+Greater Manchester, Glasgow, Nottingham, both Portsmouth and Southampton,
+eight London Boroughs, four of the seven constituent authorities of the
+West Midlands Combined Authority.
+There are also a number of other tightly bounded urban areas, such as Aberdeen,
+Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas at the centre
+of less confined districts like Burnley, Milton Keynes, and Northampton.
 
@@
 An unnumbered section, e.g.\ `\section*{Acknowledgements}`, may be used for thanks, etc.\ if required and included _in the non-anonymous version_ before any Notes or References.
-
-\pagebreak
-
-# Appendix 1 {#appendix1}
-
-This is the LAD cluster membership for the upload speed timeseries.
-
-```{r echo=FALSE, message=FALSE, results = "asis"}
-
-# Cluster table
-clusters <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
-clusters <- read_csv(clusters)
-clusters <- clusters %>%
-  dplyr::select(LAD, cluster.up) %>%
-  #rename('cluster membership' = cluster.up) %>%
-  #group_by('cluster membership') %>%
-  group_split(cluster.up, .keep = F)
-  #dplyr::arrange()
-  #sort('cluster membership', decreasing = F)
-
-cluster1 <- as.vector(unlist(clusters[1]))
-cluster2 <- as.vector(unlist(clusters[2]))
-cluster3 <- as.vector(unlist(clusters[3]))
-cluster4 <- as.vector(unlist(clusters[4]))
-cluster5 <- as.vector(unlist(clusters[5]))
-cluster6 <- as.vector(unlist(clusters[6]))
-cluster7 <- as.vector(unlist(clusters[7]))
-cluster8 <- as.vector(unlist(clusters[8]))
-cluster9 <- as.vector(unlist(clusters[9]))
-cluster10 <- as.vector(unlist(clusters[10]))
-cluster11 <- as.vector(unlist(clusters[11]))
-cluster12 <- as.vector(unlist(clusters[12]))
-cluster13 <- as.vector(unlist(clusters[13]))
-```
-
-**Cluster 1: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster1, type="cmd2"), collapse=", "))
-```
-
-**Cluster 2: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster2, type="cmd2"), collapse=", "))
-```
-
-**Cluster 3: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster3, type="cmd2"), collapse=", "))
-```
-
-**Cluster 4: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster4, type="cmd2"), collapse=", "))
-```
-
-**Cluster 5: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster5, type="cmd2"), collapse=", "))
-```
-
-**Cluster 6: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster6, type="cmd2"), collapse=", "))
-```
-
-**Cluster 7: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster7, type="cmd2"), collapse=", "))
-```
-
-**Cluster 8: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster8, type="cmd2"), collapse=", "))
-```
-
-**Cluster 9: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster9, type="cmd2"), collapse=", "))
-```
-
-**Cluster 10: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster10, type="cmd2"), collapse=", "))
-```
-
-**Cluster 11: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster11, type="cmd2"), collapse=", "))
-```
-
-**Cluster 12: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster12, type="cmd2"), collapse=", "))
-```
-
-**Cluster 13: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster13, type="cmd2"), collapse=", "))
-```
-
-```{r echo=FALSE, message=FALSE, results = "asis", eval=FALSE}
-# LADs <- do.call(cbindX, clusters) # this is with library(gdata)
-# LADs1 <- LADs[,1:7]
-# LADs2 <- LADs[,8:13]
-# 
-# names(LADs1) <- c("Cluster 1", "Cluster 2", "Cluster 3",
-#                      "Cluster 4", "Cluster 5", "Cluster 6",
-#                      "Cluster 7")
-# names(LADs2) <- c("Cluster 8", "Cluster 9", "Cluster 10",
-#                      "Cluster 11", "Cluster 12")
-# 
-# opts <- options(knitr.kable.NA = "") # kableExtra.auto_format = FALSE
-# 
-# # LADs1 %>%
-# # kable(caption = "LAD cluster membership for upload speed timeseries (1 out of 2)\\label{cluster.membership1}") %>%
-# #   kable_styling(font_size = 5,fixed_thead = T)
-# 
-# LADs1 %>%
-# kbl(caption = "LAD cluster membership for upload speed timeseries (1 out of 2)\\label{cluster.membership1}") %>%
-#    kable_classic(font_size = 5,
-#                 fixed_thead = T,
-#                 full_width = F
-#                 )
-# 
-# 
-# # stargazer(LADs1,
-# #           summary = F,
-# #           font.size = "footnotesize",
-# #           type = "latex",
-# #           header = F,
-# #           title = "LAD cluster membership for upload speed timeseries (1 out of 2)\\label{cluster.membership1}")  
-# 
-# stargazer(LADs2,
-#           summary = F,
-#           font.size = "footnotesize",
-#           type = "latex",
-#           header = F,
-#           title = "LAD cluster membership for upload speed timeseries (2 out of 2)\\label{cluster.membership2}")  
-```
-
-# Appendix 2 {#appendix2}
-
-```{r echo=FALSE, message=FALSE, results = "asis"}
-
-# Descriptive statistics table
-path.LAs_Clusters <- paste0(path, "/data/temp/LAs_Clusters.RData")
-load(path.LAs_Clusters)
-
-for.descr <- data %>%
-  dplyr::select(pop, job.dens2018, distMet,
-                distLondon, south, 
-                managers, tech, skilled, prof,
-                admin, caring, plant, 
-                earnings, nvq4,
-                virgin,
-                total.busi, LAD.tests.am20, pop) %>%
-  mutate('total.busi/pop' = total.busi/pop,
-         'LAD.tests.am20/pop' = LAD.tests.am20/pop) %>%
-  dplyr::select(-total.busi, -pop, -LAD.tests.am20)
-
-stargazer(for.descr,
-          summary = T,
-          font.size = "footnotesize",
-          type = "latex",
-          header = F,
-          digit.separator = ",",
-          digit.separate = c(3,3),
-          covariate.labels=c("pop, 2018",
-                             "job density, 2018",
-                             "distance to nearest met. area",
-                             "distance to London",
-                             "South of the UK",
-                             "managerial jobs, 2020",
-                             "tech jobs, 2020",
-                             "skilled trade jobs, 2020",
-                             "professional jobs, 2020",
-                             "administrative jobs, 2020",
-                             "leisure jobs, 2020",
-                             "machine operation jobs, 2020",
-                             "earnings, 2019",
-                             "NVQ4+",
-                             "Virgin Media \\%, 2020",
-                             "n. business est. per hab., 2019",
-                             "AM tests per hab., 2020"),
-          rownames = F,
-          no.space=TRUE,
-          column.sep.width = "0pt",
-          #notes = "",
-          title = "Descriptive statistics for the auxiliary regression explanatory variables\\label{descr.aux}")
-
-```
-
-\pagebreak

### 2020-12-29 paper/v1_old/broadband.speed.covid.tex
abstract + directories

--- 2020-12-21 paper/v1_sage/broadband.speed.covid.tex
@@
-\documentclass[Royal,times,sageh]{sagej}
+\documentclass[,]{sagej}
 
@@
 \def\tightlist{}
-\newcommand{\beginappendix}{\setcounter{table}{0} \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
 
@@
 
-\author{A. Uthor*\affilnum{1,2}, O. Tro\affilnum{2}, O. Vriga\affilnum{3}}
-
-\affiliation{\affilnum{1}{Department of Incredible Research, University A, City A, Country A}\\\affilnum{2}{Department of Applied Things, University B, City B, Country B}\\\affilnum{3}{Very Important Stuff Committee, Institute C, City C, Country C}}
-
-\corrauth{Corresponding author name, This is sample corresponding address.}
-
-\email{\href{mailto:correspondingauthor@protonmail.com}{\nolinkurl{correspondingauthor@protonmail.com}}}
+\author{A. N. Author\affilnum{}, John Smith\affilnum{}}
+
+\affiliation{}
 
@@
 \keywords{covid; internet; working from home; broadband speed; time-series
-clusters;}
+clusters}
 
@@
 
+\protect\hyperlink{sec:appendix}{Appendix}
+
 \hypertarget{sec:5}{%
@@
 
-\hypertarget{sec:appendix}{%
-\section{Appendix}\label{sec:appendix}}
-
-\begin{table}[!htbp] \centering 
-  \caption{LAD cluster membership for upload speed timeseries\label{cluster.membership}} 
-  \label{} 
-\footnotesize 
-\begin{tabular}{@{\extracolsep{5pt}} ccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
- & LAD & cluster membership \\ 
-\hline \\[-1.8ex] 
-1 & Aberdeen City & 9 \\ 
-2 & Aberdeenshire & 13 \\ 
-3 & Adur & 13 \\ 
-4 & Allerdale & 6 \\ 
-5 & Amber Valley & 6 \\ 
-6 & Angus & 6 \\ 
-7 & Antrim and Newtownabbey & 13 \\ 
-8 & Ards and North Down & 12 \\ 
-9 & Argyll and Bute & 13 \\ 
-10 & Armagh City, Banbridge and Craigavon & 13 \\ 
-11 & Arun & 13 \\ 
-12 & Ashfield & 6 \\ 
-13 & Ashford & 6 \\ 
-14 & Aylesbury Vale & 6 \\ 
-15 & Babergh & 13 \\ 
-16 & Barking and Dagenham & 11 \\ 
-17 & Barnet & 6 \\ 
-18 & Barnsley & 9 \\ 
-19 & Barrow-in-Furness & 13 \\ 
-20 & Basildon & 13 \\ 
-21 & Basingstoke and Deane & 6 \\ 
-22 & Bassetlaw & 13 \\ 
-23 & Bath and North East Somerset & 6 \\ 
-24 & Bedford & 13 \\ 
-25 & Belfast & 6 \\ 
-26 & Bexley & 6 \\ 
-27 & Birmingham & 6 \\ 
-28 & Blaby & 11 \\ 
-29 & Blackburn with Darwen & 13 \\ 
-30 & Blackpool & 11 \\ 
-31 & Blaenau Gwent & 6 \\ 
-32 & Bolsover & 11 \\ 
-33 & Bolton & 13 \\ 
-34 & Boston & 13 \\ 
-35 & Bournemouth, Christchurch and Poole & 6 \\ 
-36 & Bracknell Forest & 13 \\ 
-37 & Bradford & 6 \\ 
-38 & Braintree & 6 \\ 
-39 & Breckland & 13 \\ 
-40 & Brent & 11 \\ 
-41 & Brentwood & 6 \\ 
-42 & Bridgend & 6 \\ 
-43 & Brighton and Hove & 13 \\ 
-44 & Bristol, City of & 13 \\ 
-45 & Broadland & 13 \\ 
-46 & Bromley & 6 \\ 
-47 & Bromsgrove & 10 \\ 
-48 & Broxbourne & 9 \\ 
-49 & Broxtowe & 6 \\ 
-50 & Burnley & 11 \\ 
-51 & Bury & 6 \\ 
-52 & Caerphilly & 11 \\ 
-53 & Calderdale & 6 \\ 
-54 & Cambridge & 6 \\ 
-55 & Camden & 13 \\ 
-56 & Cannock Chase & 10 \\ 
-57 & Canterbury & 6 \\ 
-58 & Cardiff & 6 \\ 
-59 & Carlisle & 11 \\ 
-60 & Carmarthenshire & 13 \\ 
-61 & Castle Point & 6 \\ 
-62 & Causeway Coast and Glens & 11 \\ 
-63 & Central Bedfordshire & 13 \\ 
-64 & Ceredigion & 1 \\ 
-65 & Charnwood & 9 \\ 
-66 & Chelmsford & 6 \\ 
-67 & Cheltenham & 6 \\ 
-68 & Cherwell & 6 \\ 
-69 & Cheshire East & 13 \\ 
-70 & Cheshire West and Chester & 13 \\ 
-71 & Chesterfield & 6 \\ 
-72 & Chichester & 13 \\ 
-73 & Chiltern & 13 \\ 
-74 & Chorley & 9 \\ 
-75 & City of Edinburgh & 6 \\ 
-76 & City of London & 13 \\ 
-77 & Clackmannanshire & 6 \\ 
-78 & Colchester & 6 \\ 
-79 & Conwy & 12 \\ 
-80 & Copeland & 6 \\ 
-81 & Corby & 3 \\ 
-82 & Cornwall & 13 \\ 
-83 & Cotswold & 13 \\ 
-84 & County Durham & 6 \\ 
-85 & Coventry & 6 \\ 
-86 & Craven & 13 \\ 
-87 & Crawley & 11 \\ 
-88 & Croydon & 6 \\ 
-89 & Dacorum & 13 \\ 
-90 & Darlington & 1 \\ 
-91 & Dartford & 6 \\ 
-92 & Daventry & 6 \\ 
-93 & Denbighshire & 6 \\ 
-94 & Derby & 6 \\ 
-95 & Derbyshire Dales & 13 \\ 
-96 & Derry City and Strabane & 6 \\ 
-97 & Doncaster & 11 \\ 
-98 & Dorset & 6 \\ 
-99 & Dover & 13 \\ 
-100 & Dudley & 11 \\ 
-101 & Dumfries and Galloway & 13 \\ 
-102 & Dundee City & 10 \\ 
-103 & Ealing & 6 \\ 
-104 & East Ayrshire & 6 \\ 
-105 & East Cambridgeshire & 13 \\ 
-106 & East Devon & 13 \\ 
-107 & East Dunbartonshire & 13 \\ 
-108 & East Hampshire & 6 \\ 
-109 & East Hertfordshire & 13 \\ 
-110 & East Lindsey & 6 \\ 
-111 & East Lothian & 2 \\ 
-112 & East Northamptonshire & 6 \\ 
-113 & East Renfrewshire & 6 \\ 
-114 & East Riding of Yorkshire & 6 \\ 
-115 & East Staffordshire & 12 \\ 
-116 & East Suffolk & 6 \\ 
-117 & Eastbourne & 3 \\ 
-118 & Eastleigh & 6 \\ 
-119 & Eden & 1 \\ 
-120 & Elmbridge & 6 \\ 
-121 & Enfield & 6 \\ 
-122 & Epping Forest & 12 \\ 
-123 & Epsom and Ewell & 13 \\ 
-124 & Erewash & 9 \\ 
-125 & Exeter & 13 \\ 
-126 & Falkirk & 6 \\ 
-127 & Fareham & 6 \\ 
-128 & Fenland & 12 \\ 
-129 & Fermanagh and Omagh & 13 \\ 
-130 & Fife & 13 \\ 
-131 & Flintshire & 13 \\ 
-132 & Folkestone and Hythe & 13 \\ 
-133 & Forest of Dean & 13 \\ 
-134 & Fylde & 5 \\ 
-135 & Gateshead & 6 \\ 
-136 & Gedling & 6 \\ 
-137 & Glasgow City & 9 \\ 
-138 & Gloucester & 13 \\ 
-139 & Gosport & 6 \\ 
-140 & Gravesham & 6 \\ 
-141 & Great Yarmouth & 6 \\ 
-142 & Greenwich & 9 \\ 
-143 & Guildford & 6 \\ 
-144 & Gwynedd & 13 \\ 
-145 & Hackney & 13 \\ 
-146 & Halton & 9 \\ 
-147 & Hambleton & 4 \\ 
-148 & Hammersmith and Fulham & 12 \\ 
-149 & Harborough & 6 \\ 
-150 & Haringey & 6 \\ 
-151 & Harlow & 3 \\ 
-152 & Harrogate & 6 \\ 
-153 & Harrow & 6 \\ 
-154 & Hart & 6 \\ 
-155 & Hartlepool & 6 \\ 
-156 & Hastings & 13 \\ 
-157 & Havant & 9 \\ 
-158 & Havering & 6 \\ 
-159 & Herefordshire, County of & 6 \\ 
-160 & Hertsmere & 11 \\ 
-161 & High Peak & 6 \\ 
-162 & Highland & 13 \\ 
-163 & Hillingdon & 13 \\ 
-164 & Hinckley and Bosworth & 6 \\ 
-165 & Horsham & 6 \\ 
-166 & Hounslow & 11 \\ 
-167 & Huntingdonshire & 13 \\ 
-168 & Hyndburn & 11 \\ 
-169 & Inverclyde & 13 \\ 
-170 & Ipswich & 11 \\ 
-171 & Isle of Anglesey & 13 \\ 
-172 & Isle of Wight & 13 \\ 
-173 & Isles of Scilly & 11 \\ 
-174 & Islington & 6 \\ 
-175 & Kensington and Chelsea & 11 \\ 
-176 & Kettering & 6 \\ 
-177 & King's Lynn and West Norfolk & 6 \\ 
-178 & Kingston upon Hull, City of & 13 \\ 
-179 & Kingston upon Thames & 6 \\ 
-180 & Kirklees & 6 \\ 
-181 & Knowsley & 9 \\ 
-182 & Lambeth & 13 \\ 
-183 & Lancaster & 13 \\ 
-184 & Leeds & 6 \\ 
-185 & Leicester & 6 \\ 
-186 & Lewes & 11 \\ 
-187 & Lewisham & 9 \\ 
-188 & Lichfield & 8 \\ 
-189 & Lincoln & 6 \\ 
-190 & Lisburn and Castlereagh & 7 \\ 
-191 & Liverpool & 13 \\ 
-192 & Luton & 3 \\ 
-193 & Maidstone & 6 \\ 
-194 & Maldon & 6 \\ 
-195 & Malvern Hills & 13 \\ 
-196 & Manchester & 11 \\ 
-197 & Mansfield & 6 \\ 
-198 & Medway & 6 \\ 
-199 & Melton & 13 \\ 
-200 & Mendip & 6 \\ 
-201 & Merthyr Tydfil & 13 \\ 
-202 & Merton & 9 \\ 
-203 & Mid and East Antrim & 9 \\ 
-204 & Mid Devon & 13 \\ 
-205 & Mid Suffolk & 13 \\ 
-206 & Mid Sussex & 6 \\ 
-207 & Mid Ulster & 13 \\ 
-208 & Middlesbrough & 6 \\ 
-209 & Midlothian & 13 \\ 
-210 & Milton Keynes & 9 \\ 
-211 & Mole Valley & 13 \\ 
-212 & Monmouthshire & 6 \\ 
-213 & Moray & 6 \\ 
-214 & Na h-Eileanan Siar & 13 \\ 
-215 & Neath Port Talbot & 13 \\ 
-216 & New Forest & 6 \\ 
-217 & Newark and Sherwood & 9 \\ 
-218 & Newcastle-under-Lyme & 6 \\ 
-219 & Newcastle upon Tyne & 13 \\ 
-220 & Newham & 10 \\ 
-221 & Newport & 6 \\ 
-222 & Newry, Mourne and Down & 13 \\ 
-223 & North Ayrshire & 6 \\ 
-224 & North Devon & 13 \\ 
-225 & North East Derbyshire & 6 \\ 
-226 & North East Lincolnshire & 2 \\ 
-227 & North Hertfordshire & 6 \\ 
-228 & North Kesteven & 6 \\ 
-229 & North Lanarkshire & 6 \\ 
-230 & North Lincolnshire & 6 \\ 
-231 & North Norfolk & 6 \\ 
-232 & North Somerset & 13 \\ 
-233 & North Tyneside & 6 \\ 
-234 & North Warwickshire & 11 \\ 
-235 & North West Leicestershire & 6 \\ 
-236 & Northampton & 9 \\ 
-237 & Northumberland & 6 \\ 
-238 & Norwich & 11 \\ 
-239 & Nottingham & 11 \\ 
-240 & Nuneaton and Bedworth & 6 \\ 
-241 & Oadby and Wigston & 8 \\ 
-242 & Oldham & 9 \\ 
-243 & Orkney Islands & 13 \\ 
-244 & Oxford & 6 \\ 
-245 & Pembrokeshire & 6 \\ 
-246 & Pendle & 6 \\ 
-247 & Perth and Kinross & 13 \\ 
-248 & Peterborough & 11 \\ 
-249 & Plymouth & 13 \\ 
-250 & Portsmouth & 9 \\ 
-251 & Powys & 13 \\ 
-252 & Preston & 12 \\ 
-253 & Reading & 13 \\ 
-254 & Redbridge & 13 \\ 
-255 & Redcar and Cleveland & 13 \\ 
-256 & Redditch & 11 \\ 
-257 & Reigate and Banstead & 13 \\ 
-258 & Renfrewshire & 6 \\ 
-259 & Rhondda Cynon Taf & 12 \\ 
-260 & Ribble Valley & 6 \\ 
-261 & Richmond upon Thames & 9 \\ 
-262 & Richmondshire & 13 \\ 
-263 & Rochdale & 11 \\ 
-264 & Rochford & 6 \\ 
-265 & Rossendale & 1 \\ 
-266 & Rother & 13 \\ 
-267 & Rotherham & 13 \\ 
-268 & Rugby & 9 \\ 
-269 & Runnymede & 6 \\ 
-270 & Rushcliffe & 6 \\ 
-271 & Rushmoor & 13 \\ 
-272 & Rutland & 1 \\ 
-273 & Ryedale & 6 \\ 
-274 & Salford & 6 \\ 
-275 & Sandwell & 9 \\ 
-276 & Scarborough & 11 \\ 
-277 & Scottish Borders & 13 \\ 
-278 & Sedgemoor & 13 \\ 
-279 & Sefton & 6 \\ 
-280 & Selby & 7 \\ 
-281 & Sevenoaks & 13 \\ 
-282 & Sheffield & 6 \\ 
-283 & Shetland Islands & 13 \\ 
-284 & Shropshire & 6 \\ 
-285 & Slough & 11 \\ 
-286 & Solihull & 6 \\ 
-287 & Somerset West and Taunton & 13 \\ 
-288 & South Ayrshire & 6 \\ 
-289 & South Bucks & 13 \\ 
-290 & South Cambridgeshire & 13 \\ 
-291 & South Derbyshire & 9 \\ 
-292 & South Gloucestershire & 13 \\ 
-293 & South Hams & 6 \\ 
-294 & South Holland & 6 \\ 
-295 & South Kesteven & 9 \\ 
-296 & South Lakeland & 13 \\ 
-297 & South Lanarkshire & 6 \\ 
-298 & South Norfolk & 13 \\ 
-299 & South Northamptonshire & 9 \\ 
-300 & South Oxfordshire & 6 \\ 
-301 & South Ribble & 13 \\ 
-302 & South Somerset & 13 \\ 
-303 & South Staffordshire & 6 \\ 
-304 & South Tyneside & 13 \\ 
-305 & Southampton & 9 \\ 
-306 & Southend-on-Sea & 13 \\ 
-307 & Southwark & 13 \\ 
-308 & Spelthorne & 9 \\ 
-309 & St Albans & 6 \\ 
-310 & St. Helens & 11 \\ 
-311 & Stafford & 13 \\ 
-312 & Staffordshire Moorlands & 6 \\ 
-313 & Stevenage & 11 \\ 
-314 & Stirling & 13 \\ 
-315 & Stockport & 6 \\ 
-316 & Stockton-on-Tees & 9 \\ 
-317 & Stoke-on-Trent & 6 \\ 
-318 & Stratford-on-Avon & 13 \\ 
-319 & Stroud & 13 \\ 
-320 & Sunderland & 11 \\ 
-321 & Surrey Heath & 6 \\ 
-322 & Sutton & 6 \\ 
-323 & Swale & 6 \\ 
-324 & Swansea & 13 \\ 
-325 & Swindon & 13 \\ 
-326 & Tameside & 7 \\ 
-327 & Tamworth & 6 \\ 
-328 & Tandridge & 8 \\ 
-329 & Teignbridge & 13 \\ 
-330 & Telford and Wrekin & 9 \\ 
-331 & Tendring & 6 \\ 
-332 & Test Valley & 6 \\ 
-333 & Tewkesbury & 13 \\ 
-334 & Thanet & 13 \\ 
-335 & Three Rivers & 12 \\ 
-336 & Thurrock & 6 \\ 
-337 & Tonbridge and Malling & 6 \\ 
-338 & Torbay & 13 \\ 
-339 & Torfaen & 6 \\ 
-340 & Torridge & 13 \\ 
-341 & Tower Hamlets & 13 \\ 
-342 & Trafford & 9 \\ 
-343 & Tunbridge Wells & 13 \\ 
-344 & Uttlesford & 13 \\ 
-345 & Vale of Glamorgan & 8 \\ 
-346 & Vale of White Horse & 11 \\ 
-347 & Wakefield & 6 \\ 
-348 & Walsall & 9 \\ 
-349 & Waltham Forest & 13 \\ 
-350 & Wandsworth & 13 \\ 
-351 & Warrington & 6 \\ 
-352 & Warwick & 6 \\ 
-353 & Watford & 13 \\ 
-354 & Waverley & 13 \\ 
-355 & Wealden & 6 \\ 
-356 & Wellingborough & 6 \\ 
-357 & Welwyn Hatfield & 9 \\ 
-358 & West Berkshire & 6 \\ 
-359 & West Devon & 8 \\ 
-360 & West Dunbartonshire & 6 \\ 
-361 & West Lancashire & 6 \\ 
-362 & West Lindsey & 13 \\ 
-363 & West Lothian & 6 \\ 
-364 & West Oxfordshire & 6 \\ 
-365 & West Suffolk & 6 \\ 
-366 & Westminster & 12 \\ 
-367 & Wigan & 6 \\ 
-368 & Wiltshire & 6 \\ 
-369 & Winchester & 13 \\ 
-370 & Windsor and Maidenhead & 13 \\ 
-371 & Wirral & 13 \\ 
-372 & Woking & 6 \\ 
-373 & Wokingham & 13 \\ 
-374 & Wolverhampton & 11 \\ 
-375 & Worcester & 6 \\ 
-376 & Worthing & 13 \\ 
-377 & Wrexham & 6 \\ 
-378 & Wychavon & 13 \\ 
-379 & Wycombe & 6 \\ 
-380 & Wyre & 13 \\ 
-381 & Wyre Forest & 7 \\ 
-382 & York & 6 \\ 
-\hline \\[-1.8ex] 
-\end{tabular} 
-\end{table}
-
-\begin{table}[!htbp] \centering 
-  \caption{Descriptive statistics for the auxiliary regression explanatory variables\label{descr.aux}} 
-  \label{} 
-\footnotesize 
-\begin{tabular}{@{\extracolsep{0pt}}lccccccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St.
-Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Pctl(25)} & \multicolumn{1}{c}{Pctl(75)} & \multicolumn{1}{c}{Max} \\
-\hline \\[-1.8ex] 
-pop, 2018 & 365 & 1.137 & 5.726 & 0.400 & 0.700 & 0.930 & 110.110 \\ 
-job density, 2018 & 365 & 53.269 & 57.700 & 0.150 & 22.050 & 69.290 & 544.090 \\ 
-distance to nearest met. area & 365 & 201.558 & 173.634 & 0.150 & 76.180 & 278.880 & 1,003.950 \\ 
-distance to London & 365 & 0.463 & 0.499 & 0 & 0 & 1 & 1 \\ 
-South of the UK & 363 & 12.009 & 4.013 & 3.600 & 9.000 & 14.300 & 27.900 \\ 
-managerial jobs, 2020 & 364 & 14.505 & 4.057 & 3.500 & 11.800 & 16.900 & 29.600 \\ 
-tech jobs, 2020 & 358 & 10.513 & 3.764 & 1.000 & 8.025 & 12.500 & 21.600 \\ 
-skilled trade jobs, 2020 & 364 & 21.223 & 6.902 & 4.400 & 16.775 & 24.850 & 71.600 \\ 
-professional jobs, 2020 & 359 & 9.965 & 2.738 & 3.200 & 8.100 & 11.400 & 21.300 \\ 
-administrative jobs, 2020 & 362 & 9.261 & 2.827 & 2.800 & 7.300 & 11.400 & 17.800 \\ 
-leisure jobs, 2020 & 337 & 6.339 & 2.847 & 1.200 & 4.400 & 7.900 & 19.800 \\ 
-machine operation jobs, 2020 & 360 & 592.184 & 81.129 & 437.600 & 534.625 & 633.875 & 893.200 \\ 
-earnings, 2019 & 365 & 39.329 & 11.076 & 15.000 & 31.800 & 45.300 & 100.000 \\ 
-NVQ4+ & 365 & 0.152 & 0.141 & 0.000 & 0.018 & 0.241 & 0.753 \\ 
-Virgin Media \%, 2020 & 365 & 0.057 & 0.164 & 0.023 & 0.038 & 0.056 & 3.174 \\ 
-n. business est. per hab., 2019 & 365 & 0.0005 & 0.0002 & 0.0001 & 0.0003 & 0.001 & 0.001 \\ 
-\hline \\[-1.8ex] 
-\end{tabular} 
-\end{table}
-
-\bibliographystyle{sageh}
-\bibliography{bibliography}
+\bibliographystyle{}
+\bibliography{bibliography.bib}
 

### 2020-12-30 paper/v1/broadband.speed.covid.Rmd

@@
 abstract: "This paper offers a new perspective on telecommuting from the viewpoint of the complex web of digital divides.
-Using the UK as a case study, this paper studies how the quality and reliability of internet service, as reflected in *experienced* vis-à-vis *advertised* upload internet speeds during the spring 2020 lockdown, may reinforce or redress the spatial and social dimensions of digital divisions.
+Using the UK as a case study, this paper studies how the quality and reliability of internet services, as reflected in *experienced* internet upload speeds during the spring 2020 lockdown, may reinforce or redress the spatial and social dimensions of digital divisions.
 Fast, reliable internet connections are necessary for the population to be able to work from home.
-Although not every place hosts individuals in occupations which allow for telecommuting nor with the necessary skills to effectively use the internet to telecommute, we raise the issue that without good internet connectivity, these places would struggle more to achieve economic resilience in a period like the current pandemic.
-We employ data on individual broadband speed tests and state-of-the-art time-series clustering methods to create clusters of UK local authorities with similar temporal signatures of experienced internet speeds.
+Although not every place hosts individuals in occupations which allow for telecommuting nor with the necessary skills to effectively use the internet to telecommute, good internet connectivity is also essential to local economic resilience in a period like the current pandemic.
+Employing data on individual broadband speed tests and state-of-the-art time-series clustering methods, we create clusters of UK local authorities with similar temporal signatures of experienced upload speeds.
 We then associate these clusters of local authorities with their socioeconomic and geographic characteristics to explore how they overlap with or diverge from the existing economic and digital geography of the UK.
@@
 pandemic.
-<!--  The extreme level of demand for telecommuting fundamentally alters the
-potential returns of internet use for the user and wider community,
-assumes skills or functions that are present in some occupations but not in others,
-and relies upon access to high quality and reliable internet services.-->
 
@@
 in the homes of individual employees rather than any other location [@eurofound2020].
-However, it should be noted that previous research has explored how telecommuting
+However, previous research has explored how telecommuting
 can occur in other places, including satellite offices or on public transport
@@
 Previous research has also used a variety of definitions to measure the level of
-telecommuting within different workforces, distinguishing, for example, between those
+telecommuting within different workforces, distinguishing between those
 directly employed, indirectly employed, self-employed, full-time or part-time, and
@@
 different industries and employment opportunities have.
-For example, studies from the United States, the Netherlands, and the UK found
+Studies from the United States, the Netherlands, and the UK found
 that telecommuters are most likely to hold professional, managerial, and technical
@@
 been equally spread across the workforce.
-@NBERw26948 indicated that in the US, managers, educators, as well as those
+@NBERw26948 indicated that in the US, managers, educators, those
 working in computer-related occupations, finance, and law can easily work from home,
@@
 Yet whether this variation in infrastructure quality affects the spatial footprint of
-telecommuting has not previously been discussed, in part because telecommuting has
-not previously been the cause of greatest demand and pressure on internet services.
+telecommuting has not previously been investigated, in part because telecommuting has
+not previously been a cause of concentrated demand and pressure on internet services.
 
 There are indications that those who purchase high speed connections consume
-more data of all sorts and use their connections for a variety of purposes
+more data of all sorts and use their connections for a greater variety of purposes
 [@hauge2011consumer].
@@
 
-<!-- Yet did the purchase of high speed connections and increased internet access also
-prepare households for long-term home-working, enforced by government restrictions?-->
-The extreme demand during the pandemic provides a new opportunity to understand
-how infrastructure accessibility, quality, and reliability affects telecommuting,
-particularly in light of the high volumes of bandwidth-intensive video conferencing
-required in order to avoid the face-to-face contact that could increase
-the spread of infection.
+<!-- Covid-19 and working from home, cities, urban structure -->
+Multi-layered digital divides may also intersect with material divides
+and the economic geography of the UK.
+The regional economic resilience literature underlines the differentiated capacity 
+of cities and regions to escape or recover from economic crises 
+[@martin2012regional; @kitsos2018economic].
+As different places have different industrial and occupational profiles, 
+these affect their aggregated potential capacity for telecommuting.
+Such profiles are associated with longstanding inequalities in the UK and their
+spatial representation as a North-South divide [@martin_north_south].
+Various studies have illustrated severe inequalities between the north and the
+south regions of England in terms of skills and human capital,
+unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
+@dorling2018peak].
+Some scholars have even argued that the UK suffers from some of the highest levels of
+interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
+All three levels of digital divides are associated with or shaped by the
+geography of the UK. Yet this is the first time that the intersection of digital 
+and material divides is relevant to understanding the economic resilience of places 
+and large swathes of the population, as digital technologies became an essential
+tool of productivity during the Covid-19 pandemic.
+
+The extreme demand during the pandemic thus provides a new opportunity to understand
+how internet infrastructure quality, and reliability affects telecommuting, 
+particularly in light of the high volumes of bandwidth-intensive 
+video conferencing required in order to avoid the face-to-face contact that could 
+increase the spread of infection.
 We seek to answer how internet service resilience contributes to or reduces economic
@@
 We also aim to improve our understanding of the impact of first level digital
-division on telecommuting, and whether this results in much more fundamental third
-level digital division than has previously been perceived.
-
-<!-- Covid-19 and working from home, cities, urban structure -->
-Furthermore, these multi-layered digital divides intersect with material divides
-and the economic geography of the UK.
-Following the regional economic resilience literature, which underlines the
-differentiated capacity of cities and regions to escape or recover from economic
-crises [@martin2012regional; @kitsos2018economic], different places have different
-industrial and occupational profiles, and these affect the aggregated potential
-capacity of places for telecommuting.
-Such profiles are associated with longstanding inequalities in the UK and their
-spatial representation as a North-South divide [@martin_north_south].
-Various studies have illustrated severe inequalities between the north and the
-south regions of England regarding, for example, skills and human capital,
-unemployment, productivity and prosperity [@lee2014grim; @mccann2020perceptions;
-@dorling2018peak].
-Some scholars have even argued that the UK suffers some of the highest level of
-interregional inequalities in the global north [@gal2018reducing; @mccann2016uk].
-Not only are all three levels of digital divides associated with or shaped by the
-geography of the UK, but the intersection of the digital and material divides
-affects the capacity of places to overcome some of the economic effects of the
-Covid-19 pandemic.
-Importantly, this is the first time that digital technologies became an essential
-tool for economic resilience for such a great part of the population.
+division on telecommuting, and whether this results in more fundamental third
+level digital divisions than has previously been perceived.
 
@@
 <!-- clustering in general -->
-First we identified our methodological framework as cluster analysis, which can
-be defined within the modern machine learning framework as an unsupervised
+Our chosen methodological framework is cluster analysis, which can
+be defined within machine learning approaches as an unsupervised
 learning task, partitioning unlabelled observations into homogeneous groups
@@
 structures within the data [@aghabozorgi2015time].
-Therefore, cluster analysis is a widely used in geography
-[@gordon1977classification; @everitt1974cluster].
-For instance, clustering methods are the basis of *geodemographics*, a research
+Cluster analysis is a widely used in geography
+[@gordon1977classification; @everitt1974cluster], for example to solve 
+*regionalisation* problems [@niesterowicz2016].
+Clustering methods are also the basis of *geodemographics*, a research
 domain which aims to create small area indicators or typologies of neighbourhoods
 based on diverse variables [@SINGLETON2009289; @harris2005geodemographics].
-They have also been employed to solve *regionalisation* problems
-[@niesterowicz2016].
-Common characteristics of these studies are the cross-sectional nature of the data
-they employ.
-Indeed, most clustering problems in geography deal with observations that are
-fixed in time.
-However, for this paper we are interested in internet speeds, which vary over time
-and, therefore, we create clusters of local authorities in the UK with similar
+These studies usually employ cross-sectional data, and most clustering problems in 
+geography deal with observations that are fixed in time.
+However, for this paper we are interested in internet speeds, which vary over time.
+Therefore, we create clusters of local authorities in the UK with similar
 temporal signatures of experienced internet speeds.
@@
 [@aggarwal2013time; @aggarwal2001surprising; @hyndman2015large; @WARRENLIAO20051857].
-The main challenge -- and also the difference with cross-sectional clustering
-problems -- is data dimensionality given the multiplicity of data points for every
-individual object, local authorities in our case, included in the data set.
-Time-series are dynamic data as the value of the observations change as
-a function of time [@aghabozorgi2015time].
+The main challenge, which does not apply to cross-sectional clustering
+problems, is data dimensionality, with a multiplicity of data points for every
+individual object included in the data set, and how their value changes dynamically
+as a function of time [@aghabozorgi2015time].
 This high dimensionality leads to
@@
 known as shape-based approaches.
-These methods match two sperate time-series objects based on the similarity of
+These methods match two separate time-series objects based on the similarity of
 their shapes through the calculation of distances between the shapes,
@@
 i.e. shapes -- of experienced internet speeds.
-The clusters are identified using a partitioning algorithm, which leads to
-non-overlapping clusters, and particularly the widely use *k*-means algorithm because
-of the simplicity of the implementation and the interpretability of the
-results.
-This is an iterative algorithm, which begins with defining the desired number of
-clusters *k*.
+The clusters are identified using the common partitioning algorithm, 
+where no clusters overlap, known as *k*-means.
+This iterative algorithm is popular because of the simplicity of the implementation 
+and the interpretability of the results.
+It begins with defining the desired number of clusters: *k*.
 Then each observation is randomly assigned to a cluster from the $[1,k]$ space.
@@
 with the same period in $2019$, as in Figure \ref{test2019}, when the volume of
-testing and thus of experience of internet service quality peaked in the evening,
+testing and thus of experienced internet service quality peaked in the evening,
 presumably in response to demand for leisure activities and download speeds.
@@
 <!-- link to auxillary regressions method -->
-The cause of these different experiences of broadband resilience may vary across
-different areas, as they may reflect either similarities in patterns of demand
-or similar quality of infrastructure.
+The cause of these different experiences of broadband resilience may vary between
+and within clusters, as they may reflect either patterns of demand or quality 
+of infrastructure.
 Our approach is also limited by potential endogeneity, as for example, better
 quality connections with high mean speeds may enable more working from home, but
-greater demand may cause slower speeds, less reliability and greater variability
+greater demand can cause slower speeds, less reliability, or greater variability
 of speed at different times of day or week.
@@
 <!-- limitations-->
-Because we cannot identify individuals or households
-and, consequently, we aggregated our data at the LAD level,
-our results offer correlations between the socioeconomic
+Because we cannot identify individuals or households and consequently aggregated our 
+data at the LAD level, our results offer correlations between the socioeconomic
 characteristics of certain geographic locations and internet service quality,
@@
 The graphs show a composite profile of mean upload speeds per hour
-per day for each cluster, with the largest, in terms of the LAD membership and
-population, five clusters in Figure \ref{UpClusterL},
+per day for each cluster, with the largest five clusters, 
+in terms of the LAD membership and population, in Figure \ref{UpClusterL},
 and the next six in Figure \ref{UpClusterS}.
@@
 The temporal profile for cluster $6$ in Figure \ref{UpClusterL} shows that upload
-speeds are highest at $6:00$ on a Monday morning and peak between $23:00$ and
-midnight on Wednesday and Thursday, but tend to be lower during the working day.
+speeds are highest at $6:00$ on a Monday morning and between $23:00$ and midnight 
+on Wednesday and Thursday, but tend to be lower during the working day.
 Furthermore, experienced mean upload speeds in the morning peak are $4.4$% lower
@@
 This difference is less, however than any of the clusters
-as shown in Figure \ref{UpClusterS}.
+included in Figure \ref{UpClusterS}.
 
@@
 Mean speeds are also higher than in any other cluster, excluding cluster $2$,
-where measures of reliability suggests poorer performance.
-
-Thus, broadband services in clusters $9$ and $11$, home to over twelve million people
-performed the best during the study period, in terms of both quality and reliability.
+where measures of reliability suggest poorer performance.
+
+Thus, broadband services in clusters $9$ and $11$, 
+home to over twelve million people, performed the best during the study period, 
+in terms of both quality and reliability.
 In Figure \ref{UpClusterL}, cluster $11$ shows more noticeable peaks and troughs,
-but the lowest points are not during the morning peak, but occur, on average,
-between $6:00-7:00$ on Monday morning, $14:00-15:00$ on Friday, and $16:00-17:00$ on
-Thursday. These slow periods still offer better speeds than the average hourly
+with the lowest points occurring, on average, between $6:00-7:00$ on Monday morning, 
+$14:00-15:00$ on Friday, and $16:00-17:00$ on Thursday, not the workday peak. 
+Furthermore, these slow periods offer better speeds than the average hourly
 profile of cluster $13$. Finally, ignoring the smallest clusters in terms of
@@
 
-Using an auxiliary multinomial regression, we test whether the clusters that
+Using an auxiliary multinomial logit regression, we test whether the clusters that
 have higher mean speeds and more reliable services consist of LADs that are more
@@
 We also estimate which of our clusters are more likely to have a higher proportion of
-occupations which enable telecommuting because of the nature of the work.
+occupations where the nature of the work enables telecommuting.
 The results of the auxiliary regression are presented in Table \ref{aux}.
@@
 [methods and data](#sec:3) section
-and equation \ref{eq1} and each column represent a different cluster.
+and equation \ref{eq1}. Each column represents a different cluster.
 The reference case is cluster $4$, which includes only the local authority of
@@
 Mean, experienced upload speeds in cluster $4$ (see Table \ref{up.cluster.descr})
-are close to both the average speeds for the $13$ clusters ($9.9$Mb/s) and the
+are between the average speeds for the $13$ clusters ($9.9$Mb/s) and the
 pre-clustered average for the whole sample ($9.3$Mb/s).
@@
 Virgin connections, which is an indication that people in these clusters are
-more likely to live in urban areas, with more choice of broadband services,
-or in other words, are more likely to be on the right side of the first level,
+more likely to live in urban areas, with more choice of broadband services.
+In other words, they are more likely to be on the right side of the first level,
 infrastructure-based digital divide, as we expected from the analysis in Section
@@
 
-<!-- is an indication of satisfaction or at least a lack of concern over broadband speeds -->
-<!-- and quality of service. -->
-<!-- People in LADs assigned to clusters $6$ and $13$ were more likely to run speed tests -->
-<!-- (at a per capita basis) -->
-<!-- at this time of day, whilst people in cluster $11$ were the least likely to do so. -->
-<!-- This high level of satisfaction matches the high quality and reliability of service -->
-<!-- Those in clusters $2$, $3$, $9$ and $10$ were also less likely to ran speed tests per -->
-<!-- capita than those in the other six multi-authority clusters. -->
-<!-- Cluster $7$ was the only cluster with higher mean speeds and middle to high measures -->
-<!-- of reliability which was more likely to run more speed tests per capita. -->
-<!-- This might be a reflection of the choice of ISP and connection options in those -->
-<!-- LADs, as, along with cluster $12$, those in cluster $7$ were least likely to have -->
-<!-- Virgin Media connections. -->
-
-Then, we employ distance from London and from the nearest metropolitan area
+We employ distance from London and from the nearest metropolitan area
 (including London) as two variables depicting peripherality, urban structure and,
@@
 regression results. Corby and Eastbourne, the other two authorities in cluster $3$
-are, whilst not part of metropolitan areas, are also large, accessible towns.
+are large, accessible towns, although not part of metropolitan areas.
 
@@
 $7$ are the second least likely to be served by Virgin Media.
-It is also a demonstration of the complexity of experienced broadband upload speeds
-as captured by time-series clustering, and how quality and reliability, not only
-availability, engender first level digital divides.
-Cluster $1$, which our analysis suggests lacks broadband resilience, contains five,
-mainly rural authorities.
-However, they are scattered around the country -- Ceredigion in West Wales,
-Darlington in the Northeast, Eden and Rossendale in the Northwest, and Rutland in
-the East Midlands -- and therefore, the results in Table \ref{aux} indicate that
-these authorities are closer to a metropolitan area than authorities in cluster $3$,
-but furthest from London.
+It is also a demonstration of the complexity of both experienced broadband 
+upload speeds as captured by time-series clustering, and the geography of 
+first level digital divides as a product of quality as well as availability.
+In comparison, cluster $1$, which our analysis suggests lacks broadband resilience, 
+contains five, mainly rural authorities, but they are closer to a metropolitan area 
+than authorities in cluster $3$, although furthest from London, perhaps because 
+they are scattered around the country.
 
@@
 It also includes major metropolitan centres in the North -- Liverpool, Newcastle --
-and South -- Bristol, nine (of $32$) London Boroughs.
+and South -- Bristol, and nine (of $32$) London Boroughs.
 There are also plenty of Southern home county and suburban areas.
@@
 Cluster $1$ has the lowest proportion of educated people,
-and the lowest earnings, despite recording the highest proportion of managerial,
+and the lowest earnings, despite recording the highest proportion of managerial and
 professional jobs, and the second highest proportion of tech jobs.
-Rural areas such as those in cluster $1$ are often home to many retired people,
+Rural areas such as those in cluster $1$ are home to many retired people,
 which might explain these results or perhaps, as these figures are relative,
@@
 
-In contrast, Clusters $9$ and $11$ enjoy resilient internet connections and are
-characterised by higher share of occupations that may benefit from the ability
-to telecommute (e.g. managerial). LADs in Cluster $9$ are more likely to host
-highly educated individuals with higher earnings than cluster $11$, which has a
-negative coefficient for the NVQ4 variable. Cluster $11$ has fewer businesses per
-inhabitant and the second highest job density and cluster $9$ the second lowest.
+In contrast, Clusters $9$ and $11$ enjoy resilient internet connections, 
+but LADs in Cluster $9$ are more likely to host highly educated individuals 
+with higher earnings than cluster $11$, which has a negative coefficient for 
+the NVQ4 variable. Cluster $11$ has fewer businesses per inhabitant and the 
+second highest job density and cluster $9$ the second lowest.
 These coefficients might indicate that resilient broadband infrastructure
@@
 However, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
-work force to take advantage of working from home opportunities.
+workers to take advantage of working from home opportunities.
 Cluster $2$ is comprised of just two LADs, with the lowest job density of any
-cluster -- these are East Lothian near Edinburgh, and North East Lincolnshire,
-home to the town of Grimsby. Cluster $10$ is comprised of four LADs, including two
+cluster. Cluster $10$ is comprised of four LADs, including two
 peripheral suburban areas of Birmingham, the East London Borough of Newham,
@@
 
-Finally, clusters $8$ and $12$ consist by LADs enjoying close to average
+Finally, clusters $8$ and $12$ consist of LADs enjoying close to average
 mean upload speeds, but opposing patterns of internet reliability during the morning
@@
 premises might well have been abandoned. Cluster $12$ is home to $1.5$ million people
-spread across twelve LADs, including the City of Westminster but also a LAD in
+spread across twelve LADs, including the City of Westminster, 
+another London Borough, and two others outside London, but also a LAD in
 Northern Ireland, two in Wales, Preston in the Northwest, and Fenland in the East of
@@
 This paper offers a new perspective on telecommuting from the viewpoint of the
-complex web of digital divides. To do so, we employ novel data regarding
+complex web of digital divides. We employ novel data regarding
 experienced upload speeds and time-series clustering methods,
@@
 experienced speeds, and why evening download speeds, rather than daytime upload
-speeds have been used to benchmark the performance of internet services.
+speeds, have been used to benchmark the performance of internet services.
 Thus, the new patterns can be taken as evidence of widespread telecommuting
@@
 -- $6$ and $13$ -- contain so many LADs that are centres of the knowledge economy,
-from Oxford and Cambridge in the former, to areas like Reading, which has the
-highest concentration of digital businesses in the country [@technation2017].
+from Oxford and Cambridge in the former, to areas like Reading in the latter, 
+which has the highest concentration of digital businesses in the country 
+[@technation2017].
 Thus, whilst some areas suffer from an intersection of digital and economic divides,
@@
 towards working from home are difficult to predict. Nevertheless,
-they span through various aspects of economy and society: from changes to
+they span various aspects of economy and society: from changes to
 transportation planning due to altered commuting patterns to changes in land

### 2020-12-30 paper/v1/broadband.speed.covid.tex
resizing images

@@
 of the complex web of digital divides. Using the UK as a case study,
-this paper studies how the quality and reliability of internet service,
-as reflected in \emph{experienced} vis-à-vis \emph{advertised} upload
-internet speeds during the spring 2020 lockdown, may reinforce or
-redress the spatial and social dimensions of digital divisions. Fast,
-reliable internet connections are necessary for the population to be
-able to work from home. Although not every place hosts individuals in
-occupations which allow for telecommuting nor with the necessary skills
-to effectively use the internet to telecommute, we raise the issue that
-without good internet connectivity, these places would struggle more to
-achieve economic resilience in a period like the current pandemic. We
-employ data on individual broadband speed tests and state-of-the-art
-time-series clustering methods to create clusters of UK local
-authorities with similar temporal signatures of experienced internet
+this paper studies how the quality and reliability of internet services,
+as reflected in \emph{experienced} internet upload speeds during the
+spring 2020 lockdown, may reinforce or redress the spatial and social
+dimensions of digital divisions. Fast, reliable internet connections are
+necessary for the population to be able to work from home. Although not
+every place hosts individuals in occupations which allow for
+telecommuting nor with the necessary skills to effectively use the
+internet to telecommute, good internet connectivity is also essential to
+local economic resilience in a period like the current pandemic.
+Employing data on individual broadband speed tests and state-of-the-art
+time-series clustering methods, we create clusters of UK local
+authorities with similar temporal signatures of experienced upload
 speeds. We then associate these clusters of local authorities with their
@@
 carried out in the homes of individual employees rather than any other
-location \citep{eurofound2020}. However, it should be noted that
-previous research has explored how telecommuting can occur in other
-places, including satellite offices or on public transport
-\citep{felstead2012rapid, siha2006telecommuting}. Previous research has
-also used a variety of definitions to measure the level of telecommuting
-within different workforces, distinguishing, for example, between those
-directly employed, indirectly employed, self-employed, full-time or
-part-time, and those who use digital technologies to work remotely
-full-days or part-days
+location \citep{eurofound2020}. However, previous research has explored
+how telecommuting can occur in other places, including satellite offices
+or on public transport \citep{felstead2012rapid, siha2006telecommuting}.
+Previous research has also used a variety of definitions to measure the
+level of telecommuting within different workforces, distinguishing
+between those directly employed, indirectly employed, self-employed,
+full-time or part-time, and those who use digital technologies to work
+remotely full-days or part-days
 \citep{allen2015effective, bailey2002review, haddad2009examination}. No
@@
 socio-economically any more than different industries and employment
-opportunities have. For example, studies from the United States, the
-Netherlands, and the UK found that telecommuters are most likely to hold
-professional, managerial, and technical occupations where the workforce
-is better educated and wealthier, and that there is suppressed demand
-among women and part-time workers
+opportunities have. Studies from the United States, the Netherlands, and
+the UK found that telecommuters are most likely to hold professional,
+managerial, and technical occupations where the workforce is better
+educated and wealthier, and that there is suppressed demand among women
+and part-time workers
 \citep{headicar2016move, peters2004employees, singh2013modeling}.
@@
 likewise not been equally spread across the workforce.
-\citet{NBERw26948} indicated that in the US, managers, educators, as
-well as those working in computer-related occupations, finance, and law
-can easily work from home, and that occupations with opportunities to
-telecommute are associated with higher earnings. This is not the case
-for the workforce occupied in more spatially fixed occupations, from
-farming, construction and manufacturing to hospitality and care
-services. In the US, these occupations tend to be lower-income,
-non-white, without a university degree, live in rental accommodation and
-lack health insurance \citep{NBERw27085}. Similar trends can be observed
-for other countries. For example, \(75\)\% of workers with tertiary
-education worked from home in Europe during spring \(2020\), whilst only
-\(34\)\% of workers with secondary education and \(14\)\% of those
-primary education did so \citep{eurofound2020}.
+\citet{NBERw26948} indicated that in the US, managers, educators, those
+working in computer-related occupations, finance, and law can easily
+work from home, and that occupations with opportunities to telecommute
+are associated with higher earnings. This is not the case for the
+workforce occupied in more spatially fixed occupations, from farming,
+construction and manufacturing to hospitality and care services. In the
+US, these occupations tend to be lower-income, non-white, without a
+university degree, live in rental accommodation and lack health
+insurance \citep{NBERw27085}. Similar trends can be observed for other
+countries. For example, \(75\)\% of workers with tertiary education
+worked from home in Europe during spring \(2020\), whilst only \(34\)\%
+of workers with secondary education and \(14\)\% of those primary
+education did so \citep{eurofound2020}.
 
@@
 infrastructure quality affects the spatial footprint of telecommuting
-has not previously been discussed, in part because telecommuting has not
-previously been the cause of greatest demand and pressure on internet
-services.
+has not previously been investigated, in part because telecommuting has
+not previously been a cause of concentrated demand and pressure on
+internet services.
 
 There are indications that those who purchase high speed connections
-consume more data of all sorts and use their connections for a variety
-of purposes \citep{hauge2011consumer}. There is also a correlation
-between access to internet services and a reduction in household
-transport spend \citep{bris2017ict}. Independently of whether the
-additional internet use and reduced travel is because of increased
+consume more data of all sorts and use their connections for a greater
+variety of purposes \citep{hauge2011consumer}. There is also a
+correlation between access to internet services and a reduction in
+household transport spend \citep{bris2017ict}. Independently of whether
+the additional internet use and reduced travel is because of increased
 telecommuting, these studies suggest that better internet services
@@
 
-The extreme demand during the pandemic provides a new opportunity to
-understand how infrastructure accessibility, quality, and reliability
+Multi-layered digital divides may also intersect with material divides
+and the economic geography of the UK. The regional economic resilience
+literature underlines the differentiated capacity of cities and regions
+to escape or recover from economic crises
+\citep{martin2012regional, kitsos2018economic}. As different places have
+different industrial and occupational profiles, these affect their
+aggregated potential capacity for telecommuting. Such profiles are
+associated with longstanding inequalities in the UK and their spatial
+representation as a North-South divide \citep{martin_north_south}.
+Various studies have illustrated severe inequalities between the north
+and the south regions of England in terms of skills and human capital,
+unemployment, productivity and prosperity
+\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
+scholars have even argued that the UK suffers from some of the highest
+levels of interregional inequalities in the global north
+\citep{gal2018reducing, mccann2016uk}. All three levels of digital
+divides are associated with or shaped by the geography of the UK. Yet
+this is the first time that the intersection of digital and material
+divides is relevant to understanding the economic resilience of places
+and large swathes of the population, as digital technologies became an
+essential tool of productivity during the Covid-19 pandemic.
+
+The extreme demand during the pandemic thus provides a new opportunity
+to understand how internet infrastructure quality, and reliability
 affects telecommuting, particularly in light of the high volumes of
@@
 of first level digital division on telecommuting, and whether this
-results in much more fundamental third level digital division than has
+results in more fundamental third level digital divisions than has
 previously been perceived.
-
-Furthermore, these multi-layered digital divides intersect with material
-divides and the economic geography of the UK. Following the regional
-economic resilience literature, which underlines the differentiated
-capacity of cities and regions to escape or recover from economic crises
-\citep{martin2012regional, kitsos2018economic}, different places have
-different industrial and occupational profiles, and these affect the
-aggregated potential capacity of places for telecommuting. Such profiles
-are associated with longstanding inequalities in the UK and their
-spatial representation as a North-South divide
-\citep{martin_north_south}. Various studies have illustrated severe
-inequalities between the north and the south regions of England
-regarding, for example, skills and human capital, unemployment,
-productivity and prosperity
-\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
-scholars have even argued that the UK suffers some of the highest level
-of interregional inequalities in the global north
-\citep{gal2018reducing, mccann2016uk}. Not only are all three levels of
-digital divides associated with or shaped by the geography of the UK,
-but the intersection of the digital and material divides affects the
-capacity of places to overcome some of the economic effects of the
-Covid-19 pandemic. Importantly, this is the first time that digital
-technologies became an essential tool for economic resilience for such a
-great part of the population.
 
@@
 
-First we identified our methodological framework as cluster analysis,
-which can be defined within the modern machine learning framework as an
-unsupervised learning task, partitioning unlabelled observations into
-homogeneous groups known as clusters \citep{montero2014tsclust}. The key
-idea is that observations within clusters tend to be more similar than
-observations between clusters. Clustering is particularly useful for
-exploratory studies as it identifies structures within the data
-\citep{aghabozorgi2015time}. Therefore, cluster analysis is a widely
-used in geography \citep{gordon1977classification, everitt1974cluster}.
-For instance, clustering methods are the basis of
+Our chosen methodological framework is cluster analysis, which can be
+defined within machine learning approaches as an unsupervised learning
+task, partitioning unlabelled observations into homogeneous groups known
+as clusters \citep{montero2014tsclust}. The key idea is that
+observations within clusters tend to be more similar than observations
+between clusters. Clustering is particularly useful for exploratory
+studies as it identifies structures within the data
+\citep{aghabozorgi2015time}. Cluster analysis is a widely used in
+geography \citep{gordon1977classification, everitt1974cluster}, for
+example to solve \emph{regionalisation} problems
+\citep{niesterowicz2016}. Clustering methods are also the basis of
 \emph{geodemographics}, a research domain which aims to create small
 area indicators or typologies of neighbourhoods based on diverse
-variables \citep{SINGLETON2009289, harris2005geodemographics}. They have
-also been employed to solve \emph{regionalisation} problems
-\citep{niesterowicz2016}. Common characteristics of these studies are
-the cross-sectional nature of the data they employ. Indeed, most
-clustering problems in geography deal with observations that are fixed
-in time. However, for this paper we are interested in internet speeds,
-which vary over time and, therefore, we create clusters of local
-authorities in the UK with similar temporal signatures of experienced
-internet speeds.
+variables \citep{SINGLETON2009289, harris2005geodemographics}. These
+studies usually employ cross-sectional data, and most clustering
+problems in geography deal with observations that are fixed in time.
+However, for this paper we are interested in internet speeds, which vary
+over time. Therefore, we create clusters of local authorities in the UK
+with similar temporal signatures of experienced internet speeds.
 
@@
 \citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
-The main challenge -- and also the difference with cross-sectional
-clustering problems -- is data dimensionality given the multiplicity of
-data points for every individual object, local authorities in our case,
-included in the data set. Time-series are dynamic data as the value of
-the observations change as a function of time
-\citep{aghabozorgi2015time}. This high dimensionality leads to (i)
-computational and algorithmic challenges regarding handling these data
-and building algorithms to perform clustering over long time-series, and
-(ii) open questions regarding the choice of similarity measures in order
-to cluster similar times-series objects together considering the whole
-length of the time-series and overcoming issues around noise, outliers
-and shifts \citep{lin2004iterative, aghabozorgi2015time}.
+The main challenge, which does not apply to cross-sectional clustering
+problems, is data dimensionality, with a multiplicity of data points for
+every individual object included in the data set, and how their value
+changes dynamically as a function of time \citep{aghabozorgi2015time}.
+This high dimensionality leads to (i) computational and algorithmic
+challenges regarding handling these data and building algorithms to
+perform clustering over long time-series, and (ii) open questions
+regarding the choice of similarity measures in order to cluster similar
+times-series objects together considering the whole length of the
+time-series and overcoming issues around noise, outliers and shifts
+\citep{lin2004iterative, aghabozorgi2015time}.
 
 For this paper we utilise a category of time-series clustering methods
-known as shape-based approaches. These methods match two sperate
+known as shape-based approaches. These methods match two separate
 time-series objects based on the similarity of their shapes through the
@@
 of UK local authorities with similar temporal signatures -- i.e.~shapes
--- of experienced internet speeds. The clusters are identified using a
-partitioning algorithm, which leads to non-overlapping clusters, and
-particularly the widely use \emph{k}-means algorithm because of the
+-- of experienced internet speeds. The clusters are identified using the
+common partitioning algorithm, where no clusters overlap, known as
+\emph{k}-means. This iterative algorithm is popular because of the
 simplicity of the implementation and the interpretability of the
-results. This is an iterative algorithm, which begins with defining the
-desired number of clusters \emph{k}. Then each observation is randomly
-assigned to a cluster from the \([1,k]\) space. This initial cluster
-assignment is followed by iterations in order to minimise the distance
-between the centroids of the clusters and the observations assigned to
-these clusters \citep{james2013introduction}.
+results. It begins with defining the desired number of clusters:
+\emph{k}. Then each observation is randomly assigned to a cluster from
+the \([1,k]\) space. This initial cluster assignment is followed by
+iterations in order to minimise the distance between the centroids of
+the clusters and the observations assigned to these clusters
+\citep{james2013introduction}.
 
@@
 a comparison with the same period in \(2019\), as in Figure
-\ref{test2019}, when the volume of testing and thus of experience of
+\ref{test2019}, when the volume of testing and thus of experienced
 internet service quality peaked in the evening, presumably in response
@@
 \centering
-\includegraphics[width=0.75\textwidth,height=0.4\textheight]{figures/time.var.plot2020.png}
+\includegraphics[width=0.85\textwidth,height=0.5\textheight]{figures/time.var.plot2020.png}
 \caption{Speed tests over time, 2020 \label{test2020}}
@@
 The cause of these different experiences of broadband resilience may
-vary across different areas, as they may reflect either similarities in
-patterns of demand or similar quality of infrastructure. Our approach is
-also limited by potential endogeneity, as for example, better quality
-connections with high mean speeds may enable more working from home, but
-greater demand may cause slower speeds, less reliability and greater
-variability of speed at different times of day or week. Therefore, we
-avoid attributing any cause to our analysis of the experienced level of
-quality and reliability of upload speeds. Instead, we run an auxiliary
-regression to understand how the spatial and temporal patterns of
-internet service relate to the economic geography of the UK. More
-specifically, we estimate the following multinomial logit model:
+vary between and within clusters, as they may reflect either patterns of
+demand or quality of infrastructure. Our approach is also limited by
+potential endogeneity, as for example, better quality connections with
+high mean speeds may enable more working from home, but greater demand
+can cause slower speeds, less reliability, or greater variability of
+speed at different times of day or week. Therefore, we avoid attributing
+any cause to our analysis of the experienced level of quality and
+reliability of upload speeds. Instead, we run an auxiliary regression to
+understand how the spatial and temporal patterns of internet service
+relate to the economic geography of the UK. More specifically, we
+estimate the following multinomial logit model:
 
@@
 \protect\hyperlink{sec:4.2}{4.2}. Because we cannot identify individuals
-or households and, consequently, we aggregated our data at the LAD
-level, our results offer correlations between the socioeconomic
-characteristics of certain geographic locations and internet service
-quality, not a record of who was telecommuting. Such individual data
-could be found though surveys, but these offer less detailed information
-about the experience of internet resilience due to enforced demand,
-which is the main contribution of this paper. Our auxiliary regression,
-therefore, provides an indication of how internet connectivity can
-reinforce or redress existing spatial and social inequalities in
-different places. However, it opens a path to future research by
-highlighting the importance of\\
-understanding of how telecommuting capabilities and digital
-infrastructure divisions intersect.
+or households and consequently aggregated our data at the LAD level, our
+results offer correlations between the socioeconomic characteristics of
+certain geographic locations and internet service quality, not a record
+of who was telecommuting. Such individual data could be found though
+surveys, but these offer less detailed information about the experience
+of internet resilience due to enforced demand, which is the main
+contribution of this paper. Our auxiliary regression, therefore,
+provides an indication of how internet connectivity can reinforce or
+redress existing spatial and social inequalities in different places.
+However, it opens a path to future research by highlighting the
+importance of understanding of how telecommuting capabilities and
+digital infrastructure divisions intersect.
 
@@
 \ref{up.cluster.descr}. The graphs show a composite profile of mean
-upload speeds per hour per day for each cluster, with the largest, in
-terms of the LAD membership and population, five clusters in Figure
+upload speeds per hour per day for each cluster, with the largest five
+clusters, in terms of the LAD membership and population, in Figure
 \ref{UpClusterL}, and the next six in Figure \ref{UpClusterS}. These
@@
 for cluster \(6\) in Figure \ref{UpClusterL} shows that upload speeds
-are highest at \(6:00\) on a Monday morning and peak between \(23:00\)
-and midnight on Wednesday and Thursday, but tend to be lower during the
+are highest at \(6:00\) on a Monday morning and between \(23:00\) and
+midnight on Wednesday and Thursday, but tend to be lower during the
 working day. Furthermore, experienced mean upload speeds in the morning
@@
 suggesting poorer reliability during the working day. This difference is
-less, however than any of the clusters as shown in Figure
+less, however than any of the clusters included in Figure
 \ref{UpClusterS}.
@@
 in any other cluster, excluding cluster \(2\), where measures of
-reliability suggests poorer performance.
+reliability suggest poorer performance.
 
 Thus, broadband services in clusters \(9\) and \(11\), home to over
-twelve million people performed the best during the study period, in
+twelve million people, performed the best during the study period, in
 terms of both quality and reliability. In Figure \ref{UpClusterL},
-cluster \(11\) shows more noticeable peaks and troughs, but the lowest
-points are not during the morning peak, but occur, on average, between
-\(6:00-7:00\) on Monday morning, \(14:00-15:00\) on Friday, and
-\(16:00-17:00\) on Thursday. These slow periods still offer better
-speeds than the average hourly profile of cluster \(13\). Finally,
-ignoring the smallest clusters in terms of population, that is clusters
-\(4\) and \(5\), clusters \(3\), \(7\) and \(10\) also have relatively
-high mean upload speeds. Clusters \(3\) and \(10\) pair high mean speeds
-with low standard deviations relative to the mean speeds, suggesting
-reliability and resilience, as well as quality broadband services.
-Cluster \(7\) has a higher ratio of standard deviation to mean, but
-there is less difference in average speeds between the morning and
-evening peaks than in clusters \(3\) and \(10\).
+cluster \(11\) shows more noticeable peaks and troughs, with the lowest
+points occurring, on average, between \(6:00-7:00\) on Monday morning,
+\(14:00-15:00\) on Friday, and \(16:00-17:00\) on Thursday, not the
+workday peak. Furthermore, these slow periods offer better speeds than
+the average hourly profile of cluster \(13\). Finally, ignoring the
+smallest clusters in terms of population, that is clusters \(4\) and
+\(5\), clusters \(3\), \(7\) and \(10\) also have relatively high mean
+upload speeds. Clusters \(3\) and \(10\) pair high mean speeds with low
+standard deviations relative to the mean speeds, suggesting reliability
+and resilience, as well as quality broadband services. Cluster \(7\) has
+a higher ratio of standard deviation to mean, but there is less
+difference in average speeds between the morning and evening peaks than
+in clusters \(3\) and \(10\).
 
@@
 
-Using an auxiliary multinomial regression, we test whether the clusters
-that have higher mean speeds and more reliable services consist of LADs
-that are more urban, affluent, and / or more likely to benefit from a
-choice of high quality internet services. We also estimate which of our
-clusters are more likely to have a higher proportion of occupations
-which enable telecommuting because of the nature of the work. The
+Using an auxiliary multinomial logit regression, we test whether the
+clusters that have higher mean speeds and more reliable services consist
+of LADs that are more urban, affluent, and / or more likely to benefit
+from a choice of high quality internet services. We also estimate which
+of our clusters are more likely to have a higher proportion of
+occupations where the nature of the work enables telecommuting. The
 results of the auxiliary regression are presented in Table \ref{aux}.
@@
 \protect\hyperlink{sec:3}{methods and data} section and equation
-\ref{eq1} and each column represent a different cluster. The reference
+\ref{eq1}. Each column represents a different cluster. The reference
 case is cluster \(4\), which includes only the local authority of
@@
 people. Mean, experienced upload speeds in cluster \(4\) (see Table
-\ref{up.cluster.descr}) are close to both the average speeds for the
-\(13\) clusters (\(9.9\)Mb/s) and the pre-clustered average for the
-whole sample (\(9.3\)Mb/s). However, the standard deviation for cluster
-\(4\) and the difference between average speeds in the morning compared
-to the evening peak periods are indications of worse reliability than
-many of the other clusters. Hence, the results in Table \ref{aux} should
-be seen as relative rather than absolute probabilities.
+\ref{up.cluster.descr}) are between the average speeds for the \(13\)
+clusters (\(9.9\)Mb/s) and the pre-clustered average for the whole
+sample (\(9.3\)Mb/s). However, the standard deviation for cluster \(4\)
+and the difference between average speeds in the morning compared to the
+evening peak periods are indications of worse reliability than many of
+the other clusters. Hence, the results in Table \ref{aux} should be seen
+as relative rather than absolute probabilities.
 
@@
 people in these clusters are more likely to live in urban areas, with
-more choice of broadband services, or in other words, are more likely to
-be on the right side of the first level, infrastructure-based digital
+more choice of broadband services. In other words, they are more likely
+to be on the right side of the first level, infrastructure-based digital
 divide, as we expected from the analysis in Section
@@
 
-Then, we employ distance from London and from the nearest metropolitan
-area (including London) as two variables depicting peripherality, urban
+We employ distance from London and from the nearest metropolitan area
+(including London) as two variables depicting peripherality, urban
 structure and, potentially, first level digital divides. The broadband
@@
 as well as the regression results. Corby and Eastbourne, the other two
-authorities in cluster \(3\) are, whilst not part of metropolitan areas,
-are also large, accessible towns.
+authorities in cluster \(3\) are large, accessible towns, although not
+part of metropolitan areas.
 
@@
 least likely to be served by Virgin Media. It is also a demonstration of
-the complexity of experienced broadband upload speeds as captured by
-time-series clustering, and how quality and reliability, not only
-availability, engender first level digital divides. Cluster \(1\), which
-our analysis suggests lacks broadband resilience, contains five, mainly
-rural authorities. However, they are scattered around the country --
-Ceredigion in West Wales, Darlington in the Northeast, Eden and
-Rossendale in the Northwest, and Rutland in the East Midlands -- and
-therefore, the results in Table \ref{aux} indicate that these
-authorities are closer to a metropolitan area than authorities in
-cluster \(3\), but furthest from London.
+the complexity of both experienced broadband upload speeds as captured
+by time-series clustering, and the geography of first level digital
+divides as a product of quality as well as availability. In comparison,
+cluster \(1\), which our analysis suggests lacks broadband resilience,
+contains five, mainly rural authorities, but they are closer to a
+metropolitan area than authorities in cluster \(3\), although furthest
+from London, perhaps because they are scattered around the country.
 
@@
 includes major metropolitan centres in the North -- Liverpool, Newcastle
--- and South -- Bristol, nine (of \(32\)) London Boroughs. There are
+-- and South -- Bristol, and nine (of \(32\)) London Boroughs. There are
 also plenty of Southern home county and suburban areas. The standard
@@
 people, and the lowest earnings, despite recording the highest
-proportion of managerial, professional jobs, and the second highest
+proportion of managerial and professional jobs, and the second highest
 proportion of tech jobs. Rural areas such as those in cluster \(1\) are
-often home to many retired people, which might explain these results or
+home to many retired people, which might explain these results or
 perhaps, as these figures are relative, we could note that cluster \(1\)
@@
 In contrast, Clusters \(9\) and \(11\) enjoy resilient internet
-connections and are characterised by higher share of occupations that
-may benefit from the ability to telecommute (e.g.~managerial). LADs in
-Cluster \(9\) are more likely to host highly educated individuals with
-higher earnings than cluster \(11\), which has a negative coefficient
-for the NVQ4 variable. Cluster \(11\) has fewer businesses per
-inhabitant and the second highest job density and cluster \(9\) the
-second lowest. These coefficients might indicate that resilient
-broadband infrastructure generates higher returns for those in cluster
-\(9\), where slightly more slowdown in the AM peak was detected (see
-Table \ref{up.cluster.descr}). LADs in cluster \(9\) are also less
+connections, but LADs in Cluster \(9\) are more likely to host highly
+educated individuals with higher earnings than cluster \(11\), which has
+a negative coefficient for the NVQ4 variable. Cluster \(11\) has fewer
+businesses per inhabitant and the second highest job density and cluster
+\(9\) the second lowest. These coefficients might indicate that
+resilient broadband infrastructure generates higher returns for those in
+cluster \(9\), where slightly more slowdown in the AM peak was detected
+(see Table \ref{up.cluster.descr}). LADs in cluster \(9\) are also less
 likely to be in the South and those in cluster \(11\) more likely,
@@
 However, only LADs in cluster \(2\) and \(10\) are more likely to have
-highly skilled work force to take advantage of working from home
+highly skilled workers to take advantage of working from home
 opportunities. Cluster \(2\) is comprised of just two LADs, with the
-lowest job density of any cluster -- these are East Lothian near
-Edinburgh, and North East Lincolnshire, home to the town of Grimsby.
-Cluster \(10\) is comprised of four LADs, including two peripheral
-suburban areas of Birmingham, the East London Borough of Newham, and
-Dundee. Suburbs are considered the most likely urban form in which
-telecommuters live \citep{e2018does}, and Dundee has a reputation for
-tech startups \citep{technation2017}. LADs in cluster \(7\), which
-include the suburbs South of Belfast, a suburban district of Greater
-Manchester, the villages between York and Leeds, and a couple of small
-West Midlands towns have the highest earnings, and thus may also be
-making the most of their resilient internet. In contrast, LADs in
-cluster \(3\) are associated more with lower skills and have the highest
-proportion of individuals in machine operation jobs. Arguably, LADs in
-clusters \(3\) benefit the least from their resilient internet
-infrastructure in an era when working from home became a vehicle for
-economic resilience.
-
-Finally, clusters \(8\) and \(12\) consist by LADs enjoying close to
+lowest job density of any cluster. Cluster \(10\) is comprised of four
+LADs, including two peripheral suburban areas of Birmingham, the East
+London Borough of Newham, and Dundee. Suburbs are considered the most
+likely urban form in which telecommuters live \citep{e2018does}, and
+Dundee has a reputation for tech startups \citep{technation2017}. LADs
+in cluster \(7\), which include the suburbs South of Belfast, a suburban
+district of Greater Manchester, the villages between York and Leeds, and
+a couple of small West Midlands towns have the highest earnings, and
+thus may also be making the most of their resilient internet. In
+contrast, LADs in cluster \(3\) are associated more with lower skills
+and have the highest proportion of individuals in machine operation
+jobs. Arguably, LADs in clusters \(3\) benefit the least from their
+resilient internet infrastructure in an era when working from home
+became a vehicle for economic resilience.
+
+Finally, clusters \(8\) and \(12\) consist of LADs enjoying close to
 average mean upload speeds, but opposing patterns of internet
@@
 is home to \(1.5\) million people spread across twelve LADs, including
-the City of Westminster but also a LAD in Northern Ireland, two in
-Wales, Preston in the Northwest, and Fenland in the East of England.
-This spatial diversity demonstrates that the temporal clustering of
-internet resilience is not necessarily spatially dependent, and digital
-divides do not necessarily overlap with economic ones.
+the City of Westminster, another London Borough, and two others outside
+London, but also a LAD in Northern Ireland, two in Wales, Preston in the
+Northwest, and Fenland in the East of England. This spatial diversity
+demonstrates that the temporal clustering of internet resilience is not
+necessarily spatially dependent, and digital divides do not necessarily
+overlap with economic ones.
 
@@
 This paper offers a new perspective on telecommuting from the viewpoint
-of the complex web of digital divides. To do so, we employ novel data
-regarding experienced upload speeds and time-series clustering methods,
-a family of unsupervised machine learning techniques which are rarely
-utilised in geographical research. Fast, reliable internet connections
-are necessary for the population to be able to work from home. Although
-not every place hosts individuals in occupations which allow for
-telecommuting nor with the necessary skills to effectively use the
-internet to telecommute, this paper raises the issue that without good
-internet connectivity, these places would struggle more to achieve
-economic resilience in a period like the current pandemic in any case.
-Indeed, our analysis demonstrated that the temporal profiles of twelve
-of our thirteen clusters had slower upload speeds in the morning than in
-the evening. The opposite is likely to have been the norm prior to the
+of the complex web of digital divides. We employ novel data regarding
+experienced upload speeds and time-series clustering methods, a family
+of unsupervised machine learning techniques which are rarely utilised in
+geographical research. Fast, reliable internet connections are necessary
+for the population to be able to work from home. Although not every
+place hosts individuals in occupations which allow for telecommuting nor
+with the necessary skills to effectively use the internet to
+telecommute, this paper raises the issue that without good internet
+connectivity, these places would struggle more to achieve economic
+resilience in a period like the current pandemic in any case. Indeed,
+our analysis demonstrated that the temporal profiles of twelve of our
+thirteen clusters had slower upload speeds in the morning than in the
+evening. The opposite is likely to have been the norm prior to the
 pandemic, as the level of demand and bandwidth management is the most
 common cause of temporal variation in experienced speeds, and why
-evening download speeds, rather than daytime upload speeds have been
+evening download speeds, rather than daytime upload speeds, have been
 used to benchmark the performance of internet services. Thus, the new
@@
 -- contain so many LADs that are centres of the knowledge economy, from
-Oxford and Cambridge in the former, to areas like Reading, which has the
-highest concentration of digital businesses in the country
+Oxford and Cambridge in the former, to areas like Reading in the latter,
+which has the highest concentration of digital businesses in the country
 \citep{technation2017}. Thus, whilst some areas suffer from an
@@
 and attitudes towards working from home are difficult to predict.
-Nevertheless, they span through various aspects of economy and society:
-from changes to transportation planning due to altered commuting
-patterns to changes in land use and urban planning to accommodate people
-who work from home \citep{BUDNITZ2020102713, ELLDER2020102777}, and from
+Nevertheless, they span various aspects of economy and society: from
+changes to transportation planning due to altered commuting patterns to
+changes in land use and urban planning to accommodate people who work
+from home \citep{BUDNITZ2020102713, ELLDER2020102777}, and from
 productivity and innovation changes to changes in agglomeration

### 2020-12-30 paper/v1/broadband.speed.covid.Rmd
resizing images
resizing again

@@
 ---
-title: "Working from home and digital divides: resilience in the time of a pandemic"
+title: "Working from home and digital divides: resilience during the pandemic"
 author:
@@
 
-![Speed tests over time, 2020 \label{test2020}](figures/time.var.plot2020.png){height=40% width=75%}
+![Speed tests over time, 2020 \label{test2020}](figures/time.var.plot2020.png){height=50% width=85%}
 

### 2020-12-30 paper/v1/broadband.speed.covid.Rmd
Added reference, figure dimensions, further edits to results, etc.

@@
 Our analysis enables us to better understand how the spatial and social distribution
-of both occupations and online accessibility intersect to enable or hinder the
+of education, occupation and online accessibility intersect to enable or hinder the
 practice of telecommuting at a time of extreme demand.
@@
 telecommuting is likely to remain a more common means of accessing work,
-at least in comparison to the pre-Covid era,
-and broadband services and infrastructure must be fit for purpose.
+at least in comparison to the pre-Covid era.
 
@@
 location or co-location with colleagues, including supervisors
-[@allen2015effective; @singh2013modeling].
-However, successful telecommuting also requires that the quality and
-reliability of digital services, particularly home internet connection speeds,
-enable the completion of work tasks with a minimum of delay or interruption.
-Prior to the pandemic, the performance of broadband services with respect to
-telecommuters was never tested at scale, as working from home and connecting to
-colleagues and workplace resources via the internet was the purview of a small minority of workers.
+[@allen2015effective; @singh2013modeling]. However, successful telecommuting also 
+requires that the quality and reliability of digital services, particularly home 
+internet connection speeds, enable the completion of work tasks with a minimum of 
+delay or interruption. Prior to the pandemic, the performance of broadband services 
+with respect to telecommuters was never tested at scale, as working from home and 
+connecting to colleagues and workplace resources via the internet was the purview of 
+a small minority of workers.
 Instead, leisure use in the evening, when video streaming services are at their peak,
@@
 between digital divisions geographically and socioeconomically.
-By asking how resilient broadband speeds, and particularly upload speeds, are
+By asking how resilient broadband speeds, and particularly upload speeds, are as
 experienced in different parts of the UK during a time of extreme demand,
@@
 in various advanced economies.
-Rural areas, predictably, fare worse.
+Rural areas, predictably, fare worse, yet as @blank2018local highlight, variation
+in individual internet uptake and use is a product of more complex spatial and 
+demographic characteristics than simple rurality or urbanisation.
 Yet whether this variation in infrastructure quality affects the spatial footprint of
@@
 increase the spread of infection.
-We seek to answer how internet service resilience contributes to or reduces economic
-resilience when the latter is dependent upon the capability to work from home.
+We seek to answer how internet service resilience could contributes to or reduce 
+economic resilience when the latter is dependent upon the capability to 
+work from home.
 We also aim to improve our understanding of the impact of first level digital
-division on telecommuting, and whether this results in more fundamental third
+divisions on telecommuting, and whether this results in more fundamental third
 level digital divisions than has previously been perceived.
@@
 
-![Speed tests over time, 2019 \label{test2019}](figures/time.var.plot2019.png){height=40% width=75%}
+![Speed tests over time, 2019 \label{test2019}](figures/time.var.plot2019.png){height=50% width=85%}
 
@@
 of internet use is closer to what might have been expected pre-pandemic.
-And yet even among the other clusters, the reliability of internet services
+Among the other clusters, however, the reliability of internet services
 during the working day still varies considerably.
@@
 Furthermore, these slow periods offer better speeds than the average hourly
-profile of cluster $13$. Finally, ignoring the smallest clusters in terms of
-population, that is clusters $4$ and $5$, clusters $3$, $7$ and $10$ also have
-relatively high mean upload speeds. Clusters $3$ and $10$ pair high mean speeds with
+profile of cluster $13$. Clusters $3$, $7$ and $10$ also have relatively high 
+mean upload speeds. Clusters $3$ and $10$ pair high mean speeds with
 low standard deviations relative to the mean speeds, suggesting reliability and
@@
 than authorities in cluster $3$, although furthest from London, perhaps because 
-they are scattered around the country.
+they are scattered around the country -- see Figure \ref{map.up.clusters}.
 
@@
 professional jobs, and the second highest proportion of tech jobs.
-Rural areas such as those in cluster $1$ are home to many retired people,
-which might explain these results or perhaps, as these figures are relative,
-we could note that cluster $1$ also has a greater proportion of skilled trades
-than other clusters. It seems likely that the first level digital divide reinforces
-other inequities in cluster $1$, whilst those in cluster $13$ are more likely
-to earn more and have a better education despite poor internet services.
+Rural areas such as those in cluster $1$ are home to many older, retired people 
+[@blank2018local], which might explain these results or perhaps, 
+as these figures are relative, we could note that cluster $1$ also has a greater 
+proportion of skilled trades than other clusters. 
+In any case, it appears that the first level digital divide 
+reinforces other inequities in cluster $1$, where the even slower than average 
+morning upload speeds shown in Table \ref{up.cluster.descr} suggest that internet
+users were more active during the working day. Meanwhile, those in cluster $13$ are 
+more likely to earn more and have a better education despite poor internet services.
 Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the
@@
 reliability discussed in Section [4.1](#sec:4.1) may be due to increased demand,
-e.g. for telecommuting, when it is not supported by the most resilient internet
-connections.
+e.g. for telecommuting, despite lacking the most resilient internet connections.
 
@@
 generates higher returns for those in cluster $9$, where slightly more slowdown in
-the AM peak was detected (see Table \ref{up.cluster.descr}).
-LADs in cluster $9$ are also less likely to be in the South and those in cluster $11$
+the AM peak was detected -- see Table \ref{up.cluster.descr}.
+LADs in cluster $9$ are less likely to be in the South and those in cluster $11$
 more likely, although the coefficient for cluster $9$ might be skewed by
@@
 different economic profile than England. Still, both clusters consist of larger
-LADs in terms of population, including the city centre among four of the ten
-districts of Greater Manchester, eight London Boroughs, four of the seven
-constituent authorities of the West Midlands Combined Authority, Nottingham, and
-both Portsmouth and Southampton. Among the $65$ LADs are also a number of other
-tightly bounded urban areas, such as Blackpool, Ipswich, Norwich, Slough, and
-Stevenage, and urban areas at the centre of less confined districts like Burnley,
-the digital hub of Milton Keynes, and Northampton.
+LADs in terms of population, including districts within five of England's ten
+largest cities, and a number of other stand-alone urban areas and large market towns 
+-- see Appendix [1](#appendix1).
 These urban locations are on the right side of the first level digital divide,
@@
 LADs in clusters $2$, $3$, $7$, $10$ are also on the right side of the first level
-digital divide, experiencing higher than average mean speeds and ranking high to
-middle on measures of service reliability.
-However, only LADs in cluster $2$ and $10$ are more likely to have highly skilled
+digital divide, experiencing higher than average mean upload speeds and ranking 
+high to middle on measures of service reliability.
+LADs in clusters $2$ and $10$ are also more likely to have highly skilled
 workers to take advantage of working from home opportunities.
@@
 LADs in cluster $7$, which include the suburbs South of Belfast, a suburban
-district of Greater Manchester, the villages between York and Leeds, and a couple of
-small West Midlands towns have the highest earnings, and thus may also be making the
+district of Greater Manchester, and some other Midlands and North Yorkshire towns 
+and villages have the highest earnings, and thus may also be making the
 most of their resilient internet. In contrast, LADs in cluster $3$ are associated
@@
 residents, but are not significantly likely to earn more than other clusters.
-Including suburbs near Leicester, Cardiff, Birmingham, the green belt area south of
-London straddling the M25 motorway, and rural West Devon, the LADs in cluster $8$
-seem to likely to have the skills and occupations that would benefit from
-quality internet services, but suffer from poor internet resilience and reliability.
+Including suburbs near Leicester, Cardiff, and Birmingham and south of London, 
+the LADs in cluster $8$ seem to likely to have the skills and occupations that 
+would benefit from quality internet services, but suffer from poor internet 
+resilience and reliability.
 
@@
 premises might well have been abandoned. Cluster $12$ is home to $1.5$ million people
-spread across twelve LADs, including the City of Westminster, 
-another London Borough, and two others outside London, but also a LAD in
-Northern Ireland, two in Wales, Preston in the Northwest, and Fenland in the East of
-England. This spatial diversity demonstrates that the temporal clustering of internet
+spread across twelve LADs, from London to Wales as listed in Appendix [1](#appendix1). 
+This spatial diversity demonstrates that the temporal clustering of internet
 resilience is not necessarily spatially dependent, and digital divides do not
@@
 telecommuting nor with the necessary skills to effectively use the internet to
-telecommute, this paper raises the issue that without good internet connectivity,
-these places would struggle more to achieve economic resilience in a period like
-the current pandemic in any case.
+telecommute, this paper raises the issue that places without good internet 
+connectivity will still struggle more than other places to achieve economic 
+resilience in a period like the current pandemic when internet resilience is 
+so vital.
 Indeed, our analysis demonstrated that the temporal profiles of twelve of our
 thirteen clusters had slower upload speeds in the morning than in the evening.
-The opposite is likely to have been the norm prior to the pandemic, as the level of
+The opposite is likely to have been the norm prior to the pandemic, as level of
 demand and bandwidth management is the most common cause of temporal variation in
@@
 Thus, whilst some areas suffer from an intersection of digital and economic divides,
-such as those in cluster $1$, others found their digital infrastructure to be less
-reliable when confronted with the sudden change in the timing and type of demand.
-Still other LADs, such as Milton Keynes in cluster $9$ were able to benefit both
-from reliable internet connections and populations which were familiar with working
-from home and could capitalise on their digital infrastructure.
+such as those in cluster $1$, other areas, including cluster $8$ as well as $6$ and
+$13$ also found their digital infrastructure to be less than reliable when 
+confronted with the sudden change in the timing and type of demand.
+In contrast, LADs such as the digital hub, Milton Keynes, in cluster $9$ were able 
+to benefit both from reliable internet connections and populations which were 
+familiar with working from home and could capitalise on their digital infrastructure.
 Yet the nuanced picture we gained through our analysis of the UK case study suggests
-that quality internet connectivity would also enable other LADs to gain ground
-during the pandemic despite lower skills or earnings currently, especially in a
-future where telecommuting might be a more common means of accessing work
-and broadband services must be fit for purpose.
+that quality internet connectivity may also have enabled other LADs, such as those in
+cluster $3$ or $11$ to gain ground during the pandemic despite lower skill levels.
 
 <!-- WFH broader consequences-->
-Indeed, the long-term effects of such drastic changes in telecommuting and attitudes
-towards working from home are difficult to predict. Nevertheless,
-they span various aspects of economy and society: from changes to
+Digital infrastructure which considers upload speeds and working day reliability as 
+well as availability are likely to be particularly important in a future where
+telecommuting might be a more common means of accessing work and broadband services 
+must be fit for purpose, although the long-term effects of such drastic changes in 
+telecommuting and attitudes towards working from home are difficult to predict. 
+Nevertheless, they span various aspects of economy and society: from changes to
 transportation planning due to altered commuting patterns to changes in land

### 2021-01-02 paper/v1/broadband.speed.covid_anonymised.tex
submission

--- 2020-12-30 paper/v1/broadband.speed.covid.tex
@@
 
-\title{Working from home and digital divides: resilience in the time of a
-pandemic}
+\title{Working from home and digital divides: resilience during the pandemic}
 
@@
 
-\author{A. Uthor*\affilnum{1,2}, O. Tro\affilnum{2}, O. Vriga\affilnum{3}}
-
-\affiliation{\affilnum{1}{Department of Incredible Research, University A, City A, Country A}\\\affilnum{2}{Department of Applied Things, University B, City B, Country B}\\\affilnum{3}{Very Important Stuff Committee, Institute C, City C, Country C}}
-
-\corrauth{Corresponding author name, This is sample corresponding address.}
-
-\email{\href{mailto:correspondingauthor@protonmail.com}{\nolinkurl{correspondingauthor@protonmail.com}}}
+\author{}
+
+\affiliation{}
 
@@
 existing economic and digital geography of the UK. Our analysis enables
-us to better understand how the spatial and social distribution of both
-occupations and online accessibility intersect to enable or hinder the
-practice of telecommuting at a time of extreme demand. We also consider
-what lessons can be learned from this time for a future where
-telecommuting is likely to remain a more common means of accessing work,
-at least in comparison to the pre-Covid era, and broadband services and
-infrastructure must be fit for purpose.
+us to better understand how the spatial and social distribution of
+education, occupation and online accessibility intersect to enable or
+hinder the practice of telecommuting at a time of extreme demand. We
+also consider what lessons can be learned from this time for a future
+where telecommuting is likely to remain a more common means of accessing
+work, at least in comparison to the pre-Covid era.
 
@@
 and socioeconomically. By asking how resilient broadband speeds, and
-particularly upload speeds, are experienced in different parts of the UK
-during a time of extreme demand, we interrogate which places benefit
+particularly upload speeds, are as experienced in different parts of the
+UK during a time of extreme demand, we interrogate which places benefit
 from the greater economic resilience digital technologies can offer. The
@@
 quality between urban and rural areas in various advanced economies.
-Rural areas, predictably, fare worse. Yet whether this variation in
-infrastructure quality affects the spatial footprint of telecommuting
-has not previously been investigated, in part because telecommuting has
-not previously been a cause of concentrated demand and pressure on
-internet services.
+Rural areas, predictably, fare worse, yet as \citet{blank2018local}
+highlight, variation in individual internet uptake and use is a product
+of more complex spatial and demographic characteristics than simple
+rurality or urbanisation. Yet whether this variation in infrastructure
+quality affects the spatial footprint of telecommuting has not
+previously been investigated, in part because telecommuting has not
+previously been a cause of concentrated demand and pressure on internet
+services.
 
@@
 face-to-face contact that could increase the spread of infection. We
-seek to answer how internet service resilience contributes to or reduces
-economic resilience when the latter is dependent upon the capability to
-work from home. We also aim to improve our understanding of the impact
-of first level digital division on telecommuting, and whether this
-results in more fundamental third level digital divisions than has
-previously been perceived.
+seek to answer how internet service resilience could contributes to or
+reduce economic resilience when the latter is dependent upon the
+capability to work from home. We also aim to improve our understanding
+of the impact of first level digital divisions on telecommuting, and
+whether this results in more fundamental third level digital divisions
+than has previously been perceived.
 
@@
 \centering
-\includegraphics[width=0.75\textwidth,height=0.4\textheight]{figures/time.var.plot2019.png}
+\includegraphics[width=0.85\textwidth,height=0.5\textheight]{figures/time.var.plot2019.png}
 \caption{Speed tests over time, 2019 \label{test2019}}
@@
 internet use is closer to what might have been expected pre-pandemic.
-And yet even among the other clusters, the reliability of internet
-services during the working day still varies considerably. Interpreting
-this variation from the large spikes and dips shown on Figures
+Among the other clusters, however, the reliability of internet services
+during the working day still varies considerably. Interpreting this
+variation from the large spikes and dips shown on Figures
 \ref{UpClusterL} and \ref{UpClusterS} is difficult, but the statistics
@@
 workday peak. Furthermore, these slow periods offer better speeds than
-the average hourly profile of cluster \(13\). Finally, ignoring the
-smallest clusters in terms of population, that is clusters \(4\) and
-\(5\), clusters \(3\), \(7\) and \(10\) also have relatively high mean
-upload speeds. Clusters \(3\) and \(10\) pair high mean speeds with low
-standard deviations relative to the mean speeds, suggesting reliability
-and resilience, as well as quality broadband services. Cluster \(7\) has
-a higher ratio of standard deviation to mean, but there is less
-difference in average speeds between the morning and evening peaks than
-in clusters \(3\) and \(10\).
+the average hourly profile of cluster \(13\). Clusters \(3\), \(7\) and
+\(10\) also have relatively high mean upload speeds. Clusters \(3\) and
+\(10\) pair high mean speeds with low standard deviations relative to
+the mean speeds, suggesting reliability and resilience, as well as
+quality broadband services. Cluster \(7\) has a higher ratio of standard
+deviation to mean, but there is less difference in average speeds
+between the morning and evening peaks than in clusters \(3\) and \(10\).
 
@@
 metropolitan area than authorities in cluster \(3\), although furthest
-from London, perhaps because they are scattered around the country.
+from London, perhaps because they are scattered around the country --
+see Figure \ref{map.up.clusters}.
 
@@
 proportion of tech jobs. Rural areas such as those in cluster \(1\) are
-home to many retired people, which might explain these results or
-perhaps, as these figures are relative, we could note that cluster \(1\)
-also has a greater proportion of skilled trades than other clusters. It
-seems likely that the first level digital divide reinforces other
-inequities in cluster \(1\), whilst those in cluster \(13\) are more
+home to many older, retired people \citep{blank2018local}, which might
+explain these results or perhaps, as these figures are relative, we
+could note that cluster \(1\) also has a greater proportion of skilled
+trades than other clusters. In any case, it appears that the first level
+digital divide reinforces other inequities in cluster \(1\), where the
+even slower than average morning upload speeds shown in Table
+\ref{up.cluster.descr} suggest that internet users were more active
+during the working day. Meanwhile, those in cluster \(13\) are more
 likely to earn more and have a better education despite poor internet
@@
 Section \protect\hyperlink{sec:4.1}{4.1} may be due to increased demand,
-e.g.~for telecommuting, when it is not supported by the most resilient
-internet connections.
+e.g.~for telecommuting, despite lacking the most resilient internet
+connections.
 
@@
 cluster \(9\), where slightly more slowdown in the AM peak was detected
-(see Table \ref{up.cluster.descr}). LADs in cluster \(9\) are also less
+-- see Table \ref{up.cluster.descr}. LADs in cluster \(9\) are less
 likely to be in the South and those in cluster \(11\) more likely,
@@
 different economic profile than England. Still, both clusters consist of
-larger LADs in terms of population, including the city centre among four
-of the ten districts of Greater Manchester, eight London Boroughs, four
-of the seven constituent authorities of the West Midlands Combined
-Authority, Nottingham, and both Portsmouth and Southampton. Among the
-\(65\) LADs are also a number of other tightly bounded urban areas, such
-as Blackpool, Ipswich, Norwich, Slough, and Stevenage, and urban areas
-at the centre of less confined districts like Burnley, the digital hub
-of Milton Keynes, and Northampton. These urban locations are on the
-right side of the first level digital divide, and the regression results
+larger LADs in terms of population, including districts within five of
+England's ten largest cities, and a number of other stand-alone urban
+areas and large market towns -- see Appendix
+\protect\hyperlink{appendix1}{1}. These urban locations are on the right
+side of the first level digital divide, and the regression results
 suggest that internet resilience is supporting a wide range of small and
@@
 of the first level digital divide, experiencing higher than average mean
-speeds and ranking high to middle on measures of service reliability.
-However, only LADs in cluster \(2\) and \(10\) are more likely to have
-highly skilled workers to take advantage of working from home
+upload speeds and ranking high to middle on measures of service
+reliability. LADs in clusters \(2\) and \(10\) are also more likely to
+have highly skilled workers to take advantage of working from home
 opportunities. Cluster \(2\) is comprised of just two LADs, with the
@@
 in cluster \(7\), which include the suburbs South of Belfast, a suburban
-district of Greater Manchester, the villages between York and Leeds, and
-a couple of small West Midlands towns have the highest earnings, and
-thus may also be making the most of their resilient internet. In
-contrast, LADs in cluster \(3\) are associated more with lower skills
-and have the highest proportion of individuals in machine operation
-jobs. Arguably, LADs in clusters \(3\) benefit the least from their
-resilient internet infrastructure in an era when working from home
-became a vehicle for economic resilience.
+district of Greater Manchester, and some other Midlands and North
+Yorkshire towns and villages have the highest earnings, and thus may
+also be making the most of their resilient internet. In contrast, LADs
+in cluster \(3\) are associated more with lower skills and have the
+highest proportion of individuals in machine operation jobs. Arguably,
+LADs in clusters \(3\) benefit the least from their resilient internet
+infrastructure in an era when working from home became a vehicle for
+economic resilience.
 
@@
 educated residents, but are not significantly likely to earn more than
-other clusters. Including suburbs near Leicester, Cardiff, Birmingham,
-the green belt area south of London straddling the M25 motorway, and
-rural West Devon, the LADs in cluster \(8\) seem to likely to have the
-skills and occupations that would benefit from quality internet
-services, but suffer from poor internet resilience and reliability.
+other clusters. Including suburbs near Leicester, Cardiff, and
+Birmingham and south of London, the LADs in cluster \(8\) seem to likely
+to have the skills and occupations that would benefit from quality
+internet services, but suffer from poor internet resilience and
+reliability.
 
@@
 home, business premises might well have been abandoned. Cluster \(12\)
-is home to \(1.5\) million people spread across twelve LADs, including
-the City of Westminster, another London Borough, and two others outside
-London, but also a LAD in Northern Ireland, two in Wales, Preston in the
-Northwest, and Fenland in the East of England. This spatial diversity
-demonstrates that the temporal clustering of internet resilience is not
-necessarily spatially dependent, and digital divides do not necessarily
-overlap with economic ones.
+is home to \(1.5\) million people spread across twelve LADs, from London
+to Wales as listed in Appendix \protect\hyperlink{appendix1}{1}. This
+spatial diversity demonstrates that the temporal clustering of internet
+resilience is not necessarily spatially dependent, and digital divides
+do not necessarily overlap with economic ones.
 
@@
 with the necessary skills to effectively use the internet to
-telecommute, this paper raises the issue that without good internet
-connectivity, these places would struggle more to achieve economic
-resilience in a period like the current pandemic in any case. Indeed,
-our analysis demonstrated that the temporal profiles of twelve of our
-thirteen clusters had slower upload speeds in the morning than in the
-evening. The opposite is likely to have been the norm prior to the
-pandemic, as the level of demand and bandwidth management is the most
-common cause of temporal variation in experienced speeds, and why
-evening download speeds, rather than daytime upload speeds, have been
-used to benchmark the performance of internet services. Thus, the new
-patterns can be taken as evidence of widespread telecommuting and other
-daytime internet use which changed the temporal profile of internet
-activity throughout the UK, not just in areas with more digital industry
-or better skills.
+telecommute, this paper raises the issue that places without good
+internet connectivity will still struggle more than other places to
+achieve economic resilience in a period like the current pandemic when
+internet resilience is so vital. Indeed, our analysis demonstrated that
+the temporal profiles of twelve of our thirteen clusters had slower
+upload speeds in the morning than in the evening. The opposite is likely
+to have been the norm prior to the pandemic, as level of demand and
+bandwidth management is the most common cause of temporal variation in
+experienced speeds, and why evening download speeds, rather than daytime
+upload speeds, have been used to benchmark the performance of internet
+services. Thus, the new patterns can be taken as evidence of widespread
+telecommuting and other daytime internet use which changed the temporal
+profile of internet activity throughout the UK, not just in areas with
+more digital industry or better skills.
 
@@
 intersection of digital and economic divides, such as those in cluster
-\(1\), others found their digital infrastructure to be less reliable
-when confronted with the sudden change in the timing and type of demand.
-Still other LADs, such as Milton Keynes in cluster \(9\) were able to
-benefit both from reliable internet connections and populations which
-were familiar with working from home and could capitalise on their
-digital infrastructure. Yet the nuanced picture we gained through our
-analysis of the UK case study suggests that quality internet
-connectivity would also enable other LADs to gain ground during the
-pandemic despite lower skills or earnings currently, especially in a
-future where telecommuting might be a more common means of accessing
-work and broadband services must be fit for purpose.
-
-Indeed, the long-term effects of such drastic changes in telecommuting
+\(1\), other areas, including cluster \(8\) as well as \(6\) and \(13\)
+also found their digital infrastructure to be less than reliable when
+confronted with the sudden change in the timing and type of demand. In
+contrast, LADs such as the digital hub, Milton Keynes, in cluster \(9\)
+were able to benefit both from reliable internet connections and
+populations which were familiar with working from home and could
+capitalise on their digital infrastructure. Yet the nuanced picture we
+gained through our analysis of the UK case study suggests that quality
+internet connectivity may also have enabled other LADs, such as those in
+cluster \(3\) or \(11\) to gain ground during the pandemic despite lower
+skill levels.
+
+Digital infrastructure which considers upload speeds and working day
+reliability as well as availability are likely to be particularly
+important in a future where telecommuting might be a more common means
+of accessing work and broadband services must be fit for purpose,
+although the long-term effects of such drastic changes in telecommuting
 and attitudes towards working from home are difficult to predict.
@@
 
-\pagebreak
-
 \bibliographystyle{sageh}

### 2021-01-15 paper/v1_old_sage/broadband.speed.covid.Rmd
naming differnet versions of the paper

--- 2020-12-30 paper/v1/broadband.speed.covid.Rmd
@@
 author:
-- name: A. Uthor*
-  num: 1,2
-- name: O. Tro
-  num: 2
-- name: O. Vriga
-  num: 3
+- name: Hannah Budnitz
+  num: 1
+- name: Emmanouil Tranos
+  num: 2,3
 address:
 - num: 1
-  org: Department of Incredible Research, University A, City A, Country A
+  org: Transport Studies Unit, University of Oxford, UK
 - num: 2
-  org: Department of Applied Things, University B, City B, Country B
+  org: School of Geographic Sciences, University of Bristol, UK
 - num: 3
-  org: Very Important Stuff Committee, Institute C, City C, Country C
-corrauth: "Corresponding author name, This is sample corresponding address."
-email: correspondingauthor@protonmail.com
+  org: The Alan Turing Institute, UK
+corrauth: "Emmanouil Tranos"
+email: e.tranos@bristol.ac.uk
 abstract: "This paper offers a new perspective on telecommuting from the viewpoint of the complex web of digital divides.
@@
 ```
-
-\pagebreak

### 2021-01-15 paper/v1_old_sage/broadband.speed.covid_anonymised.Rmd
naming differnet versions of the paper

--- 2021-01-15 paper/v1_old_sage/broadband.speed.covid.Rmd
@@
 title: "Working from home and digital divides: resilience during the pandemic"
-author:
-- name: Hannah Budnitz
-  num: 1
-- name: Emmanouil Tranos
-  num: 2,3
-address:
-- num: 1
-  org: Transport Studies Unit, University of Oxford, UK
-- num: 2
-  org: School of Geographic Sciences, University of Bristol, UK
-- num: 3
-  org: The Alan Turing Institute, UK
-corrauth: "Emmanouil Tranos"
-email: e.tranos@bristol.ac.uk
+
 abstract: "This paper offers a new perspective on telecommuting from the viewpoint of the complex web of digital divides.

### 2021-01-15 paper/v1_taylor_francis/broadband.speed.covid.Rmd
naming differnet versions of the paper

--- 2021-01-15 paper/v1_old_sage/broadband.speed.covid_anonymised.Rmd
@@
 ---
-title: "Working from home and digital divides: resilience during the pandemic"
-
-abstract: "This paper offers a new perspective on telecommuting from the viewpoint of the complex web of digital divides.
+title: |
+  Working from home and digital divides: resilience during the pandemic
+author:
+  - name: Hannah Budnitz
+    affil: a
+  - name: Emmanouil Tranos
+    affil: b, c
+    email: e.tranos@bristol.ac.uk
+affiliation:
+  - num: a
+    address: |
+      Transport Studies Unit, University of Oxford, UK
+  - num: b
+    address: |
+      School of Geographic Sciences, University of Bristol, UK
+  - num: c
+    address: |
+      The Alan Turing Institute, UK
+bibliography: bibliography.bib
+abstract: |
+  This paper offers a new perspective on telecommuting from the viewpoint of the complex web of digital divides.
 Using the UK as a case study, this paper studies how the quality and reliability of internet services, as reflected in *experienced* internet upload speeds during the spring 2020 lockdown, may reinforce or redress the spatial and social dimensions of digital divisions.
@@
 We then associate these clusters of local authorities with their socioeconomic and geographic characteristics to explore how they overlap with or diverge from the existing economic and digital geography of the UK.
-Our analysis enables us to better understand how the spatial and social distribution of both occupations and online accessibility intersect to enable or hinder the practice of telecommuting at a time of extreme demand."
-keywords:   covid; internet; working from home; broadband speed; time-series clusters;
-classoption:
-  - Royal
-  - times
-bibliography: bibliography
-bibliographystyle: sageh
+Our analysis enables us to better understand how the spatial and social distribution of both occupations and online accessibility intersect to enable or hinder the practice of telecommuting at a time of extreme demand.
+keywords: |
+  covid; internet; working from home; broadband speed; time-series clusters
 header-includes: |
@@
 
-# Appendix 1 {#appendix1}
+# Appendix 1 {- #appendix1}
 
@@
 
-# Appendix 2 {#appendix2}
+# Appendix 2 {- #appendix2}
 

### 2021-03-04 src/LAs_Clusters_k5.Rmd
clsuters and regressions for k = 5. descriptive_Clusters.Rmd has NOT been updated yet.

--- 2020-10-07 src/LAs_Clusters.Rmd
@@
 ---
-title: "LAs_clusters"
+title: "LAs_clusters k = 5"
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
@@
 library(raster)
+library(httr)
+library(readxl)
 
@@
 # load the data created by Data_Spatial.Rmd and ts_clusters.Rmd
-path.cluster <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
+path.cluster <- paste(path, "/data/temp/clusters_nodiff_5.csv", sep = "")
 clusters <- read_csv(path.cluster)
@@
 
-## Distance to Metro Area
+## Load socio-econ data and merge
 
 ```{r}
-# get LA 
-# (i) directly from the web
-#la <- readOGR("http://geoportal1-ons.opendata.arcgis.com/datasets/b6d2e15801de45328b760a4f55d74318_0.geojson?outSR={%22latestWkid%22:3857,%22wkid%22:102100}")#, layer="OGRGeoJSON")
-# or, (ii) from the json I saved locally
-path.json <- paste(path, "./data/raw/Local_Authority_Districts_(April_2019)_Boundaries_UK_BFE.geojson", sep = "")
-la <- readOGR(path.json)#, layer="OGRGeoJSON")
-# source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe
-
-# spatial transformations
-la <- spTransform(la, CRS("+init=epsg:4326"))
-la@data$LAD19NM <- as.character(la@data$LAD19NM)
-ladDist <- la@data[,c("LAD19CD", "LAD19NM")]
-names(ladDist)[1] <- "lad19cd"
-
-#distance from London and other 
-coords_LAD <- la@data[,c("LONG", "LAT")]
-London <- la[which(la@data$LAD19NM == "City of London"),]
-London <- gCentroid(London)
-distLondon <- distm(coords_LAD, London, fun=distCosine) 
-#result in meters - change to km
-distLondon <- round((distLondon/1000),2)
-Birmingham <- la[which(la@data$LAD19NM == "Birmingham"),]
-Birmingham <- gCentroid(Birmingham)
-distBham <- distm(coords_LAD, Birmingham, fun=distCosine)
-distBham <- round((distBham/1000),2)
-Manchester <- la[which(la@data$LAD19NM == "Manchester"),]
-Manchester <- gCentroid(Manchester)
-distManc <- distm(coords_LAD, Manchester, fun=distCosine)
-distManc <- round((distManc/1000),2)
-Leeds <- la[which(la@data$LAD19NM == "Leeds"),]
-Leeds <- gCentroid(Leeds)
-distLeeds <- distm(coords_LAD, Leeds, fun=distCosine)
-distLeeds <- round((distLeeds/1000),2)
-Liverpool <- la[which(la@data$LAD19NM == "Liverpool"),]
-Liverpool <- gCentroid(Liverpool)
-distLivp <- distm(coords_LAD, Liverpool, fun=distCosine)
-distLivp <- round((distLivp/1000),2)
-Newcastle <- la[which(la@data$LAD19NM == "Newcastle upon Tyne"),]
-Newcastle <- gCentroid(Newcastle)
-distNewc <- distm(coords_LAD, Newcastle, fun=distCosine)
-distNewc <- round((distNewc/1000),2)
-Sheffield <- la[which(la@data$LAD19NM == "Sheffield"),]
-Sheffield <- gCentroid(Sheffield)
-distShef <- distm(coords_LAD, Sheffield, fun=distCosine)
-distShef <- round((distShef/1000),2)
-SouthHants <- la[which(la@data$LAD19NM == "Fareham"),]
-SouthHants <- gCentroid(SouthHants)
-distSHam <- distm(coords_LAD, SouthHants, fun=distCosine)
-distSHam <- round((distSHam/1000),2)
-Nottingham <- la[which(la@data$LAD19NM == "Nottingham"),]
-Nottingham <- gCentroid(Nottingham)
-distNotts <- distm(coords_LAD, Nottingham, fun=distCosine)
-distNotts <- round((distNotts/1000),2)
-Bristol <- la[which(la@data$LAD19NM == "Bristol, City of"),]
-Bristol <- gCentroid(Bristol)
-distBris <- distm(coords_LAD, Bristol, fun=distCosine)
-distBris <- round((distBris/1000),2)
-Cardiff <- la[which(la@data$LAD19NM == "Cardiff"),]
-Cardiff <- gCentroid(Cardiff)
-distCard <- distm(coords_LAD, Cardiff, fun=distCosine)
-distCard <- round((distCard/1000),2)
-Glasgow <- la[which(la@data$LAD19NM == "Glasgow City"),]
-Glasgow <- gCentroid(Glasgow)
-distGlas <- distm(coords_LAD, Glasgow, fun=distCosine) 
-distGlas <- round((distGlas/1000),2)
-
-ladDist <- cbind(ladDist, distLondon, distBham, distManc, distLeeds,
-                 distLivp, distNewc, distShef, distSHam, distNotts, 
-                 distBris, distCard, distGlas)
-
-ladDist$distMet <- apply(ladDist[,3:14], 1, min)
-ladDist <- ladDist %>%
-  dplyr::select(lad19cd, distLondon, distMet)
-
-#test <- left_join(lad, ladDist)
-```
-
-## Job density 2018
-
-```{r}
-
-job.dens <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_57_1.data.csv?geography=1816133633...1816133848,1820327937...1820328318&date=latest&item=3&measures=20100")
-
-job.dens <- job.dens %>%
-  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, OBS_VALUE) %>%
-  rename(job.dens2018 = OBS_VALUE) %>%
-  rename(lad19cd = GEOGRAPHY_CODE) %>%
-  distinct()
-
-# test <- left_join(lad, job.dens, by = "lad19cd")
-```
-
-## Population 2018
-
-```{r}
-# pop <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_31_1.data.csv?geography=1807745025...1807745028,1807745030...1807745032,1807745034...1807745155,1807745157...1807745164,1807745166...1807745170,1807745172...1807745177,1807745179...1807745194,1807745196,1807745197,1807745199,1807745201...1807745218,1807745221,1807745222,1807745224,1807745226...1807745231,1807745233,1807745234,1807745236...1807745244,1807745271...1807745281,1811939329...1811939332,1811939334...1811939336,1811939338...1811939497,1811939499...1811939501,1811939503,1811939505...1811939507,1811939509...1811939517,1811939519,1811939520,1811939524...1811939570,1811939575...1811939599,1811939601...1811939628,1811939630...1811939634,1811939636...1811939647,1811939649,1811939655...1811939664,1811939667...1811939680,1811939682,1811939683,1811939685,1811939687...1811939704,1811939707,1811939708,1811939710,1811939712...1811939717,1811939719,1811939720,1811939722...1811939730,1811939757...1811939767&date=latestMINUS1-latest&sex=7&age=0,22&measures=20100")
-
-pop <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_31_1.data.csv?geography=1820327937...1820328318,1816133633...1816133848&date=latestMINUS1-latest&sex=7&age=0,22&measures=20100")
-
-pop <- pop %>%
-  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, AGE_NAME, OBS_VALUE) %>%
-  distinct(GEOGRAPHY_CODE, AGE_NAME, .keep_all = T) %>% 
-  spread(AGE_NAME, OBS_VALUE) %>% 
-  rename(pop16_64 = `Aged 16 - 64`) %>%
-  rename(pop = `All ages`) %>%
-  rename(lad19cd = GEOGRAPHY_CODE)
-
-# test <- left_join(lad, pop, by = "lad19cd")
-# sapply(test, function(x) sum(is.na(x)))
-```
-
-## Population density 2018
-
-```{r}
-
-# this is just population
-pop.dens <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_31_1.data.csv?geography=1820327937...1820328318,1816133633...1816133848&date=latestMINUS1-latest&sex=7&age=0,22&measures=20100")
-
-pop.dens <- pop.dens %>%
-  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, AGE_NAME, OBS_VALUE) %>%
-  distinct(GEOGRAPHY_CODE, AGE_NAME, .keep_all = T) %>% 
-  spread(AGE_NAME, OBS_VALUE) %>% 
-  rename(pop16_64 = `Aged 16 - 64`) %>%
-  rename(pop = `All ages`) %>%
-  rename(lad19cd = GEOGRAPHY_CODE) %>%
-  dplyr::select(-pop16_64)
-
-crs(la)
-la$area_sqkm <- area(la) / 1000000
-
-pop.dens <- merge(pop.dens, la@data, by.x = "lad19cd", by.y = "LAD19CD", all.x = T)
-
-pop.dens <- pop.dens %>%
-  mutate(pop.dens = pop/area_sqkm) %>%
-  dplyr::select(lad19cd,pop.dens)
-
-# sapply(pop.dens, function(x) sum(is.na(x)))
-# test <- pop.dens[is.na(pop.dens$area_sqkm),]
-```
-
-## N/S divide
-
-```{r}
-
-# It only includes only the south
-
-ns <- read_csv("https://opendata.arcgis.com/datasets/3ba3daf9278f47daba0f561889c3521a_0.csv") %>%
-  mutate(south = ifelse(RGN19NM=="London" | 
-                        RGN19NM=="South West" |
-                        RGN19NM=="East of England" |
-                        RGN19NM=="South East", 1, 0)) %>%
-  dplyr::select(LAD19CD, south) %>% 
-  dplyr::filter(south==1) %>%
-  rename(lad19cd = LAD19CD)
-```
-
-## Labour supply March 2020
-
-```{r}
-
-labour <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_17_5.data.csv?geography=1816133633...1816133837,1820327937...1820328307&date=latest&variable=18,1532...1540,290,720&measures=20599,21001,21002,21003")
-
-labour <- labour %>%
-  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, VARIABLE_NAME, OBS_VALUE) %>%
-  distinct(GEOGRAPHY_CODE, VARIABLE_NAME, .keep_all = T) %>% 
-  spread(VARIABLE_NAME, OBS_VALUE) %>% 
-  rename(lad19cd = GEOGRAPHY_CODE) %>%
-  rename(managers =  `% all in employment who are - 1: managers, directors and senior officials (SOC2010)`) %>%
-  rename(prof = `% all in employment who are - 2: professional occupations (SOC2010)`) %>%
-  rename(tech = `% all in employment who are - 3: associate prof & tech occupations (SOC2010)`) %>%
-  rename(admin = `% all in employment who are - 4: administrative and secretarial occupations (SOC2010)`) %>%
-  rename(skilled = `% all in employment who are - 5: skilled trades occupations (SOC2010)`) %>%
-  rename(caring = `% all in employment who are - 6: caring, leisure and other service occupations (SOC2010)`) %>%
-  rename(sales = `% all in employment who are - 7: sales and customer service occupations (SOC2010)`) %>%
-  rename(plant = `% all in employment who are - 8: process, plant and machine operatives (SOC2010)`) %>%
-  rename(elementary = `% all in employment who are - 9: elementary occupations (SOC2010)`) %>%
-  rename(NVQ3 = `% with NVQ3+ - aged 16-64`) %>%
-  rename(NVQ4 = `% with NVQ4+ - aged 16-64`) %>%
-  rename(econ.act = `Economic activity rate - aged 16-64`)
-
-# test <- left_join(lad, labour, by = "lad19cd")
-# sapply(test, function(x) sum(is.na(x)))
-```
-
-## Earnings 2019
-
-```{r}
-
-earnings <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_30_1.data.csv?geography=1946157348,2092957698,2013265929,1816133633...1816133837,1820327937...1820328307&date=latest&sex=8&item=2&pay=1&measures=20100,20701")
-
-earnings <- earnings %>%
-  filter(MEASURES_NAME == "Value") %>% # dropiing the confedence
-  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, OBS_VALUE) %>%
-  rename(lad19cd = GEOGRAPHY_CODE) %>%
-  rename(earnings =  OBS_VALUE) %>%
-  distinct()
-
-# test <- left_join(lad, earnings, by = "lad19cd")
-# sapply(test, function(x) sum(is.na(x)))
-```
-
-## Business counts 2019
-
-```{r}
-
-# busi <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_141_1.data.csv?geography=1816133633...1816133848,1820327937...1820328318,1870659585...1870659791,1870659801,1870659792...1870659800,1879048193...1879048573,1879048583,1879048574...1879048582&date=latest&industry=163577857...163577874&employment_sizeband=0&legal_status=0&measures=20100")
-
-busi <- read_csv("https://www.nomisweb.co.uk/api/v01/dataset/NM_141_1.data.csv?geography=1816133633...1816133848,1820327937...1820328318,1870659585...1870659791,1870659801,1870659792...1870659800,1879048193...1879048573,1879048583,1879048574...1879048582&date=latest&industry=150994945...150994965&employment_sizeband=0&legal_status=0&measures=20100")
-
-busi <- busi %>%
-  dplyr::select(GEOGRAPHY_CODE, GEOGRAPHY_NAME, INDUSTRY_NAME, OBS_VALUE) %>%
-  distinct(GEOGRAPHY_CODE, INDUSTRY_NAME, .keep_all = T) %>% 
-  spread(INDUSTRY_NAME, OBS_VALUE) %>% 
-  rename(lad19cd = GEOGRAPHY_CODE) %>%
-  rename(A = `A : Agriculture, forestry and fishing`) %>%
-  rename(B = `B : Mining and quarrying`) %>%
-  rename(C = `C : Manufacturing`) %>%
-  rename(D = `D : Electricity, gas, steam and air conditioning supply`) %>%
-  rename(E = `E : Water supply; sewerage, waste management and remediation activities`) %>%
-  rename(F = `F : Construction`) %>%
-  rename(G = `G : Wholesale and retail trade; repair of motor vehicles and motorcycles`) %>%
-  rename(H = `H : Transportation and storage`) %>%
-  rename(I = `I : Accommodation and food service activities`) %>%
-  rename(J = `J : Information and communication`) %>%
-  rename(K = `K : Financial and insurance activities`) %>%
-  rename(L = `L : Real estate activities`) %>%
-  rename(M = `M : Professional, scientific and technical activities`) %>%
-  rename(N = `N : Administrative and support service activities`) %>%
-  rename(O = `O : Public administration and defence; compulsory social security`) %>%
-  rename(P = `P : Education`) %>%
-  rename(Q = `Q : Human health and social work activities`) %>%
-  rename(R = `R : Arts, entertainment and recreation`) %>%
-  rename(S = `S : Other service activities`) %>%
-  rename(T = `T : Activities of households as employers;undifferentiated goods-and services-producing activities of households for own use`) %>%
-  rename(U = `U : Activities of extraterritorial organisations and bodies`) %>%
-  mutate(total.busi = rowSums(.[3:23]))
-
-# test <- left_join(lad, busi, by = "lad19cd")
-# sapply(test, function(x) sum(is.na(x)))
-```
-
-## merge
-
-```{r}
-data <- list(lad, ladDist, job.dens, pop, pop.dens, labour, earnings, busi) # ns 
-sapply(data, function(x) dim(x))
-sapply(data, function(x) names(x))
-
-# merge with reduce
-data <- data %>% 
-  reduce(inner_join, by = "lad19cd") %>%
-  #distinct(.keep_all = T)
-  unique()
-sapply(data, function(x) sum(is.na(x)))
-
-data <- merge(data, ns, all.x = T)
-data$south <- ifelse(is.na(data$south),0, data$south)
+data.old.path <- paste0(path, "/data/temp/data_for_aux.csv")
+data <- read_csv(data.old.path) %>% 
+  dplyr::select(-c(1,4:19)) %>% 
+  left_join(x = lad, y = . )
 
@@
 # export file for future reference / backup
-# data.out.path <- paste0(path, "/data/temp/data_for_aux.csv")
+# data.out.path <- paste0(path, "/data/temp/data_for_aux_k5.csv")
 # write.csv(data, data.out.path)
@@
 data$cluster.up <- as.factor(data$cluster.up)
-data$cluster.up <- relevel(data$cluster.up, "3") # base category: cluster 3
+data$cluster.up <- relevel(data$cluster.up, "4") # base category: cluster 4
 
-aux <- multinom(cluster.up ~ 
+aux.up <- multinom(cluster.up ~ 
                      #pop16_64 + # non sign, while pop is
@@
                      managers + tech + skilled + prof + admin + caring + plant +
-                  earnings + I(total.busi/pop) +
+                     earnings + I(total.busi/pop) + nvq4 +
                      I(LAD.tests.am20/pop) + # non sensitive to addition of other tests pm
@@
 # McFadden's R squared
-aux.null <- multinom(cluster.up ~ 1, data = data)
-r2 <- 1-logLik(aux)/logLik(aux.null)
+aux.up.null <- multinom(cluster.up ~ 1, data = data)
+r2.up <- 1-logLik(aux.up)/logLik(aux.up.null)
 
-n.obs <- dim(aux$fitted.values)[1]
+n.obs <- dim(aux.up$fitted.values)[1]
 
-stargazer(aux, 
+stargazer(aux.up, 
           type = "html",
@@
             "n. business est. per hab., 2019",
-                             "n. of AM broadband tests per hab., 2019",
+            "% of NVQ4+",
+            "n. of AM broadband tests per hab., 2020",
             "% of Virgin Media connections"),
           add.lines = list(c("McFadden's R squared",
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3)),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
+                             , round(r2.up[1],3), round(r2.up[1],3)),
                            (c("N",
@@
                               n.obs, n.obs, n.obs,
-                            n.obs, n.obs, n.obs))))
+                              n.obs, n.obs, n.obs
+                              , n.obs, n.obs))))
 
@@
 
-## Auxiliary regression, dep. var.: test clusters
-
-```{r test regression, results= 'asis', message=F, error=F, warning=F}
-# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
-# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/
-
-# convert dependent variable to factor
-data$cluster.tests <- as.factor(data$cluster.tests)
-data$cluster.tests <- relevel(data$cluster.tests, "5") # base category: cluster 5
-
-aux <- multinom(cluster.tests ~ 
-                  #pop16_64 + # non sign, while pop is
-                  pop +
-                  #pop.dens + 
-                  job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
-                  distMet + distLondon + south +
-                  managers + tech + skilled + prof + admin + caring + plant +
-                  earnings + I(total.busi/pop) +
-                  I(LAD.tests.am20/pop) + # non sensitive to addition of other tests pm
-                  virgin
-                , data = data, 
-                trace = F) # to avoid messages
-
-# McFadden's R squared
-aux.null <- multinom(cluster.tests ~ 1, data = data)
-r2 <- 1-logLik(aux)/logLik(aux.null)
-
-n.obs <- dim(aux$fitted.values)[1]
-
-stargazer(aux, 
-          type = "html",
-          #type = "text",
-          digits = 3,
-          covariate.labels=c(#"pop. 16-64, 2018",
-                             "pop, 2018",
-                             #"pop. densuty, 2018",
-                             "job density, 2018",
-                             "distance to nearest met. area",
-                             "distance to London",
-                             "South of the UK",
-                             "% of managerial jobs, 2020",
-                             "% of tech jobs, 2020",
-                             "% of skilled trade jobs, 2020",
-                             "% of professional jobs, 2020",
-                             "% of administrative jobs, 2020",
-                             "% of leisure jobs, 2020",
-                             "% of machine operation jobs, 2020",
-                             "earnings, 2019",
-                             "n. business est. per hab., 2019",
-                             "n. of AM broadband tests per hab., 2019",
-                             "% of Virgin Media connections"),
-          add.lines = list(c("McFadden's R squared",
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3)),
-                           (c("N",
-                            n.obs, n.obs, n.obs,
-                            n.obs))))
-
+```{r}
+path.out <- paste0(path, "/data/temp/LAs_Clusters_k5.RData")
+save.image(path.out)
 ```
-
-## Auxiliary regression, dep. var.: down clusters
-
-```{r down regression, results= 'asis', message=F, error=F, warning=F}
-# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
-# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/
-
-# convert dependent variable to factor
-data$cluster.down <- as.factor(data$cluster.down)
-data$cluster.down <- relevel(data$cluster.down, "2") # base category: cluster 2
-
-aux <- multinom(cluster.down ~ 
-                  #pop16_64 + # non sign, while pop is
-                  pop +
-                  #pop.dens + 
-                  job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
-                  distMet + distLondon + south +
-                  managers + tech + skilled + prof + admin + caring + plant +
-                  earnings + I(total.busi/pop) +
-                  I(LAD.tests.am20/pop) + # non sensitive to addition of other tests pm
-                  virgin
-                , data = data, 
-                trace = F) # to avoid messages
-
-# McFadden's R squared
-aux.null <- multinom(cluster.down ~ 1, data = data)
-r2 <- 1-logLik(aux)/logLik(aux.null)
-
-n.obs <- dim(aux$fitted.values)[1]
-
-stargazer(aux, 
-          type = "html",
-          #type = "text",
-          digits = 3,
-          covariate.labels=c(#"pop. 16-64, 2018",
-                             "pop, 2018",
-                             #"pop. densuty, 2018",
-                             "job density, 2018",
-                             "distance to nearest met. area",
-                             "distance to London",
-                             "South of the UK",
-                             "% of managerial jobs, 2020",
-                             "% of tech jobs, 2020",
-                             "% of skilled trade jobs, 2020",
-                             "% of professional jobs, 2020",
-                             "% of administrative jobs, 2020",
-                             "% of leisure jobs, 2020",
-                             "% of machine operation jobs, 2020",
-                             "earnings, 2019",
-                             "n. business est. per hab., 2019",
-                             "n. of AM broadband tests per hab., 2019",
-                             "% of Virgin Media connections"),
-          add.lines = list(c("McFadden's R squared",
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3)),
-                           (c("N",
-                            n.obs, n.obs, n.obs,
-                            n.obs, n.obs, n.obs,
-                            n.obs, n.obs, n.obs,
-                            n.obs, n.obs, n.obs,
-                            n.obs, n.obs, n.obs,
-                            n.obs, n.obs, n.obs,
-                            n.obs))))
-
-```

### 2021-03-04 src/ts_clusters_k5.Rmd
clsuters and regressions for k = 5. descriptive_Clusters.Rmd has NOT been updated yet.

--- 2020-12-02 src/ts_clusters.Rmd
@@
 ---
-title: "Time series clusters"
+title: "Time series clusters, k = 5"
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
@@
 library(ggmap)
+library(cowplot)
 
@@
 
-# 5 - 20 cluster sollutions
-pc_k <- tsclust(ts.wide, k = c(5:15), seed = 94L,
-                type="partitional",
-                distance = "dtw_basic", centroid = "pam", trace = T, 
-                args = tsclust_args(dist = list(window.size = 20L)))
+# Define k instead of selecting it
+k <- 5
 
-# decide k, see RJ-2019-023 for max/min
-names(pc_k) <- paste0("k_", c("5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15"))
-cvis <- sapply(pc_k, cvi, type = "internal")
-cvis
-# max Sil
-# max SF
-# max CH
-# min DB
-# min DBstar
-# max D
-# min COP
-
-# loop for k decision
-pref.cvi <- NULL
-
-max.cvis <- c(1:3,6)
-for (i in max.cvis){
-  a <- which.max(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-min.cvis <- c(4:5,7)
-for (i in min.cvis){
-  a <- which.min(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-print(pref.cvi) # cvis[2,] is always 0
-
-pc_k <- tsclust(ts.wide, k = 13, seed = 94L, 
+pc_k <- tsclust(ts.wide, k = k, seed = 94L, 
                 type="partitional", 
@@
 
-# 5 - 20 cluster sollutions
-pc_k <- tsclust(ts.wide, k = c(5:15), seed = 94L,
-                type="partitional",
-                distance = "dtw_basic", centroid = "pam", trace = T, 
-                args = tsclust_args(dist = list(window.size = 20L)))
+# Define k instead of selecting it
 
-# decide k, see RJ-2019-023 for max/min
-names(pc_k) <- paste0("k_", c("5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15"))
-cvis <- sapply(pc_k, cvi, type = "internal")
-cvis
-# max Sil
-# max SF
-# max CH
-# min DB
-# min DBstar
-# max D
-# min COP
-
-# loop for k decision
-pref.cvi <- NULL
-
-max.cvis <- c(1:3,6)
-for (i in max.cvis){
-  a <- which.max(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-min.cvis <- c(4:5,7)
-for (i in min.cvis){
-  a <- which.min(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-print(pref.cvi) # cvis[2,] is always 0
-
-pc_k <- tsclust(ts.wide, k = 10, seed = 94L, 
+pc_k <- tsclust(ts.wide, k = k, seed = 94L, 
                 type="partitional", 
@@
 
-# 5 - 20 cluster sollutions
-pc_k <- tsclust(ts.wide, k = c(5:15), seed = 94L,
-                type="partitional",
-                distance = "dtw_basic", centroid = "pam", trace = T, 
-                args = tsclust_args(dist = list(window.size = 20L)))
+# Define k instead of selecting it
 
-# decide k, see RJ-2019-023 for max/min
-names(pc_k) <- paste0("k_", c("5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15"))
-cvis <- sapply(pc_k, cvi, type = "internal")
-cvis
-# max Sil
-# max SF
-# max CH
-# min DB
-# min DBstar
-# max D
-# min COP
-
-# loop for k decision
-pref.cvi <- NULL
-
-max.cvis <- c(1:3,6)
-for (i in max.cvis){
-  a <- which.max(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-min.cvis <- c(4:5,7)
-for (i in min.cvis){
-  a <- which.min(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-print(pref.cvi) # cvis[2,] is always 0
-
-pc_k <- tsclust(ts.wide, k = 7, seed = 94L, 
+pc_k <- tsclust(ts.wide, k = k, seed = 94L, 
                 type="partitional", 
@@
 
-# 5 - 20 cluster sollutions
-pc_k <- tsclust(ts.wide, k = c(5:15), seed = 94L,
-                type="partitional",
-                distance = "dtw_basic", centroid = "pam", trace = T, 
-                args = tsclust_args(dist = list(window.size = 20L)))
+# Define k instead of selecting it
 
-# decide k, see RJ-2019-023 for max/min
-names(pc_k) <- paste0("k_", c("5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15"))
-cvis <- sapply(pc_k, cvi, type = "internal")
-cvis
-# max Sil
-# max SF
-# max CH
-# min DB
-# min DBstar
-# max D
-# min COP
-
-# loop for k decision
-pref.cvi <- NULL
-
-max.cvis <- c(1:3,6)
-for (i in max.cvis){
-  a <- which.max(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-min.cvis <- c(4:5,7)
-for (i in min.cvis){
-  a <- which.min(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-print(pref.cvi) # cvis[2,] is always 0
-
-pc_k <- tsclust(ts.wide, k = 6, seed = 94L, 
+pc_k <- tsclust(ts.wide, k = k, seed = 94L, 
                 type="partitional", 
@@
 
-# 5 - 20 cluster sollutions
-pc_k <- tsclust(ts.wide, k = c(5:15), seed = 94L, #HERE
-                type="partitional",
-                distance = "dtw_basic", centroid = "pam", trace = T, 
-                args = tsclust_args(dist = list(window.size = 20L)))
+# Define k instead of selecting it
 
-# decide k, see RJ-2019-023 for max/min
-names(pc_k) <- paste0("k_", c("5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15"))
-cvis <- sapply(pc_k, cvi, type = "internal")
-cvis
-# max Sil
-# max SF
-# max CH
-# min DB
-# min DBstar
-# max D
-# min COP
-
-# loop for k decision
-pref.cvi <- NULL
-
-max.cvis <- c(1:3,6)
-for (i in max.cvis){
-  a <- which.max(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-min.cvis <- c(4:5,7)
-for (i in min.cvis){
-  a <- which.min(cvis[i,])
-  pref.cvi <- c(pref.cvi, a)
-}
-
-print(pref.cvi) # cvis[2,] is always 0
-
-pc_k <- tsclust(ts.wide, k = 15, seed = 94L, 
+pc_k <- tsclust(ts.wide, k = k, seed = 94L, 
                 type="partitional", 
@@
 # save clusters file
-out.clusters <- paste(path, "/data/temp/clusters_all.csv", sep = "")
+out.clusters <- paste(path, "/data/temp/clusters_all_5.csv", sep = "")
 write_csv(clusters, out.clusters)
@@
 # save clusters file
-out.clusters <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
+out.clusters <- paste(path, "/data/temp/clusters_nodiff_5.csv", sep = "")
 write_csv(clusters.nodiff, out.clusters)
@@
   geom_polygon(aes( group = group, fill = as.factor(cluster.tests))) +
-  theme_nothing(legend = TRUE) +
+  #theme_nothing(legend = TRUE) +
   labs(title = "Test clusters") +
   scale_fill_viridis_d()+
+  guides(fill=guide_legend(title="Clusters")) +
   theme_void()
@@
   geom_polygon(aes( group = group, fill = as.factor(cluster.up))) +
-  theme_nothing(legend = TRUE) +
+  #theme_nothing(legend = TRUE) +
   labs(title = "Upload clusters") +
   scale_fill_viridis_d()+
+  guides(fill=guide_legend(title="Clusters")) +
   theme_void()
 
-ggplot(la.f, aes(x = long, y = lat)) +
-  geom_polygon(aes( group = group, fill = as.factor(cluster.down))) +
-  theme_nothing(legend = TRUE) +
-  labs(title = "Download clusters") +
-  scale_fill_viridis_d()+
-  theme_void()
-
+#ggsave("all_maps2.png") # this is the colour map for the preprint
+# out.path.map <- paste0(path, "/paper/v1/figures/map.up.clusters5.png")
+# ggsave(out.path.map)#,
+#width = 210, #A4
+#height = 297, #A4
+#units = "mm")
 ```

### 2021-03-04 paper/v2_taylor_francis/response_reviewers.md
v.2

---
output:
  pdf_document: default
---
Dear Prof. Bian,

Thank you for providing the opportunity to revise and resubmit our paper.
We found the reviewer comments very useful and constructive and we believe they enabled us to improve the paper.
We were also very pleased to see that the reviewers found our paper ...

blah blah

Kind Regards,

[removed for anonymity]

## Reviewer: 1

>The paper reports research of detecting clusters of network access speed, especially under the pandemic for resilience. Such a topic is of great importance as we are all in the pandemic and we rely on the network cyberinfrastructure for work as stated, and also for online education for all levels of students. Therefore, it worth supports with a few major revisions needed:

> 1. The VGI approach of obtaining network speed maybe bias given that this probably is the digital guru group that will respond, maybe a discussion on this and picture the steps of selecting/soliciting and how representative this is.

> 2. The results for network speed for the different regions are not very different with 1-2 out of 10Mbps difference and probably can not reflect the social disparities. Not sure if this is the facts in U.K. If so, then it's not a problem?

> 3. The relevance to pandemic is not very straightforward. This may be improved by adding some studies, e.g., how the clusters impact the effectiveness of work from home through survey. Or relates to the employment rate.

> 4. The time series may be enhanced by putting a time range according to different events of the pandemic or other factors that relates to the pandemic.

> 5. Would suggest a review of the spatiotemporal challenges of the covid-19.

> 6. A review of cyberinfrastructure and how network varies in different region would be useful.

> 7. The first part introduction can be shorted to reflect the importance of network bandwidth to resilience of pandemic impacted life.

## Reviewer: 2

>This article presents interesting and timely research in digital geography, which is directly relevant to digital divide associated with the COVID-19 pandemic. The pandemic provides a once-in-a-lifetime opportunity to examine the barriers of broadband usage for telecommuting. The first two sections were very well written. I enjoyed reading them. However, I have a few major concerns on the methodology and a lack of discussion.

> 1. The paper used k-means, a popular unsupervised classification in this study. The optimal number of clusters was determined as 13 based on cluster validation indices (CVIs). While this number could be well justified, I feel it is too high to properly interpret the results. The interpretation and knowledge discovery from clustering results with a high k tend to be tedious. Being affected by the high k, the Results section appears to be very difficult to read and follow. As there are many different CVIs, I feel a lower k will benefit the interpretation of results.

ET

> 2. Another major comment is on the multinomial logit regression (MLR). This is partly related to my first comment on a larger k. While there is no hard limit on the number of categories in MLR, the interpretation of results will be difficult with a high k as well. In addition, if the dependence among independent variables (collinearity) occurs, the estimation of the multinomial logit model parameters becomes inaccurate. According to Table 2, it is very likely some of the variables could present collinearity (e.g., distance to nearest metro and tech jobs).

ET

> 3. Although Fig. 1 and 2 exhibit different speed tests patterns in 2019 and 2020, the data mining outcome lacks a baseline (prior to 2020) to compare with. It is very likely that the clusters in 2019 would be different from that in 2020. It will make the results more robust. Of course, I will not push this given the page limit. But some insights may be helpful in the Discussion (however, a Discussion section is missing).

ET: after we finalise k, we can rerun k-means for 2019 and decide

> 4. No discussion was included in this manuscript. I expect some insights into various important issues, such as problem causes, any broadband policies and socioeconomic solutions, equitable development, and research uncertainties.

> 5.  There are three levels of digital divide. According to the last paragraph in Section 2, only the intersection of level-1 and level-3 was included in this research. I am wondering how level-2 may fit in the general research framework. Could education and demographic variables be used to represent level-2 digital divide?

> 6. A few suggestions on the figures. Fig 1 and 2 have crowded tick labels for x-axis month and weekday. Fig 3 and 4 may be placed side by side with the same y-axis for easy comparison. Actually, I feel “Figure 3 shows that the high standard deviation” may be inaccurate as Fig 4 appears to exhibit higher variability. Fig 5 maps may include the locations of metropolitan or major cities to help interpretation. The warmer-cooler color scheme may match the population size instead of the ordinal cluster numbers. The current color scheme does not help interpretation at all.

> 7. A few minor things. In Sec 2.2, the statement “These returns have been even greater …” appears to be controversial. LAD acronym appears before it was defined. Virgin Media as a major broadband provider should be briefly introduced in the manuscript.

### 2021-03-04 paper/v2_taylor_francis/broadband.speed.covid.tex
v.2

--- 2021-01-02 paper/v1/broadband.speed.covid_anonymised.tex
@@
-\documentclass[Royal,times,sageh]{sagej}
-
-\usepackage{moreverb,url,natbib, multirow, tabularx}
-\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
+% interactcadsample.tex
+% v1.03 - April 2017
+
+\documentclass[]{interact}
+
+\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
+\usepackage{subfigure}% Support for small, `sub' figures and tables
+%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
+
+\usepackage{natbib}% Citation support using natbib.sty
+\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
+\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
+
+\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
+
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
+\newtheorem{notation}{Notation}
+
+% see https://stackoverflow.com/a/47122900
 
@@
 
+\articletype{}
+
 \title{Working from home and digital divides: resilience during the pandemic}
 
-\runninghead{}
-
-\author{}
-
-\affiliation{}
+\author{\name{Hannah Budnitz$^{a}$, Emmanouil Tranos$^{b, c}$}
+\affil{$^{a}$Transport Studies Unit, University of Oxford, UK; $^{b}$School of Geographic Sciences, University of Bristol, UK; $^{c}$The Alan Turing Institute, UK}
+}
+
+\thanks{CONTACT Hannah Budnitz. Email: , Emmanouil Tranos. Email: \href{mailto:e.tranos@bristol.ac.uk}{\nolinkurl{e.tranos@bristol.ac.uk}}}
+
+\maketitle
 
@@
 
-\keywords{covid; internet; working from home; broadband speed; time-series
-clusters;}
-
-\maketitle
+\begin{keywords}
+covid; internet; working from home; broadband speed; time-series
+clusters
+\end{keywords}
 
@@
 \hypertarget{appendix1}{%
-\section{Appendix 1}\label{appendix1}}
+\section*{Appendix 1}\label{appendix1}}
+\addcontentsline{toc}{section}{Appendix 1}
 
@@
 \hypertarget{appendix2}{%
-\section{Appendix 2}\label{appendix2}}
+\section*{Appendix 2}\label{appendix2}}
+\addcontentsline{toc}{section}{Appendix 2}
 
@@
 
-\bibliographystyle{sageh}
-\bibliography{bibliography}
+\bibliographystyle{tfcad}
+\bibliography{bibliography.bib}
 

### 2021-03-10 paper/v2_taylor_francis/broadband.speed.covid.Rmd
Answered most points from reviewer 1 and point 7 from reviewer 2.

--- 2021-01-15 paper/v1_taylor_francis/broadband.speed.covid.Rmd
@@
 Whilst the basis for this pandemic-generated divide has been viewed as mainly
-occupational, here we consider whether it is also technological and, consequently, geographical.
-
-<!-- our approach -->
-Using the UK as a case study, this paper aims to understand how the quality and
-reliability of internet service, as reflected in *experienced* internet speeds
-during the spring 2020 lockdown,
-may reinforce or redress the spatial and social dimensions of digital divisions.
-We employ volunteered geographic data on individual broadband speed tests
-and state-of-the-art time-series clustering methods to create clusters of UK local
-authorities with similar temporal signatures of experienced internet speeds.
-We then associate these clusters of local authorities with their socioeconomic
-and geographic characteristics to explore how they overlap with or diverge from
-the existing economic and digital geography of the UK.
-Our analysis enables us to better understand how the spatial and social distribution
-of education, occupation and online accessibility intersect to enable or hinder the
-practice of telecommuting at a time of extreme demand.
-We also consider what lessons can be learned from this time for a future where
-telecommuting is likely to remain a more common means of accessing work,
-at least in comparison to the pre-Covid era.
+occupational, here we consider whether it is also technological and geographical.
 
@@
 internet connection speeds, enable the completion of work tasks with a minimum of 
-delay or interruption. Prior to the pandemic, the performance of broadband services 
-with respect to telecommuters was never tested at scale, as working from home and 
-connecting to colleagues and workplace resources via the internet was the purview of 
-a small minority of workers.
-Instead, leisure use in the evening, when video streaming services are at their peak,
-has been used to benchmark broadband performance and service delivery by different
-Internet Service Providers (ISPs) [@ofcom2017].
-
-Yet the shift towards telecommuting during various stages of lockdown around the world
-has been drastic and there are speculations that post-Covid, the tendency to work from
-home will be much higher, raising questions around whether internet
+delay or interruption. High levels of demand are one of the main causes of reduced 
+reliability and slower connection speeds, as network bandwidth becomes congested. 
+Yet prior to the pandemic, the performance of broadband services with respect to 
+telecommuters was never tested at scale, as working from home and connecting to 
+colleagues and workplace resources via the internet was the purview of a small 
+minority of workers. Instead, leisure use in the evening, when video streaming 
+services are at their peak, has been used to benchmark broadband performance and 
+service delivery by different Internet Service Providers (ISPs) [@ofcom2017].
+
+The shift towards telecommuting during various stages of lockdown around the world
+has been drastic and there are speculations that post-Covid, the tendency to work 
+from home will be much higher, raising questions around whether internet
 services can accommodate the increased demand.
@@
 @ons2020lm2019].
-A back of the envelope calculation suggests that up to 40% of the labour force
-could work from home on an ongoing basis [@batty2020editorial].
 Similar figures have been reported for other countries [@felstead2020homeworking].
@@
 
-<!-- ICT infrastructure and digital divides-->
-The literature describes first level digital divides in terms of the
-availability and quality of internet connectivity, such as that manifest in
-different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
-Second level digital divides consider the presence or lack of the necessary skills
-to effectively utilise digital technologies and the internet
-[@blank2014dimensions; @van2011internet].
-The third level focuses on the heterogenous returns of internet usage among
-different socioeconomic groups and, consequently, how digital technologies can
-assist in bridging or further enhancing existing socioeconomic divides.
-[@stern2009levels; @van2014digital; @van2015third].
-The capability to telecommute is related to all three levels of digital divides,
-but more importantly leads to differentiated outcomes regarding the economic
-resilience of people and places to overcome a systemic shock such as the current
-pandemic.
-
+<!-- our approach -->
+Using the UK as a case study, this paper aims to understand how the quality and
+reliability of internet service, as reflected in *experienced* internet speeds
+during the spring 2020 lockdown,
+may reinforce or redress the spatial and social dimensions of digital divisions.
+We employ volunteered geographic data on individual broadband speed tests
+and state-of-the-art time-series clustering methods to create clusters of UK local
+authorities with similar temporal signatures of experienced internet speeds.
+We then associate these clusters of local authorities with their socioeconomic
+and geographic characteristics to explore how they overlap with or diverge from
+the existing economic and digital geography of the UK.
 As the quality of internet infrastructure and services, as well as the concentration
@@
 between digital divisions geographically and socioeconomically.
+Our analysis enables us to better understand how the spatial and social distribution
+of education, occupation and online accessibility intersect to enable or hinder the
+practice of telecommuting at a time of extreme demand.
+We also consider what lessons can be learned from this time for a future where
+telecommuting is likely to remain a more common means of accessing work,
+at least in comparison to the pre-Covid era.
+
 By asking how resilient broadband speeds, and particularly upload speeds, are as
@@
 <!-- WFH and digital divides-->
+The literature describes first level digital divides in terms of the
+availability and quality of internet connectivity, such as that manifest in
+different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
+Second level digital divides consider the presence or lack of the necessary skills
+to effectively utilise digital technologies and the internet
+[@blank2014dimensions; @van2011internet].
+The third level focuses on the heterogenous returns of internet usage among
+different socioeconomic groups and, consequently, how digital technologies can
+assist in bridging or further enhancing existing socioeconomic divides.
+[@stern2009levels; @van2014digital; @van2015third].
+The capability to telecommute is related to all three levels of digital divides,
+but more importantly leads to differentiated outcomes regarding the economic
+resilience of people and places to overcome a systemic shock such as the current
+pandemic.
+
 Our understanding of telecommuting as a product of enabled occupations can be
@@
 commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
-These returns have been even greater during the Covid-19 crisis,
+In some aspects, these returns may have increased during the Covid-19 crisis,
 when those with the capability to telecommute also have the ability to maintain
@@
 first level digital divide in infrastructure quality between urban and rural areas
-in various advanced economies.
-Rural areas, predictably, fare worse, yet as @blank2018local highlight, variation
+in various advanced economies. Rural areas, predictably, fare worse.
+This finding broadly applies in the UK, where studies that also analysed broadband 
+speed checks concluded that average speeds are lower in rural areas, something that
+has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
+@riddlesden2014broadband found that levels of deprivation did not correlate with 
+first level digital divides.
+
+However, as @blank2018local highlight, variation
 in individual internet uptake and use is a product of more complex spatial and 
 demographic characteristics than simple rurality or urbanisation.
-Yet whether this variation in infrastructure quality affects the spatial footprint of
-telecommuting has not previously been investigated, in part because telecommuting has
-not previously been a cause of concentrated demand and pressure on internet services.
-
+Dense urban areas were shown to suffer more from slow down during peak hours, 
+although these services were more likely to be improved by increased competition 
+between providers, such as between new entrants and Virgin Media cable connections
+[@riddlesden2014broadband; @nardotto2015unbundling]. The latter were historically 
+available to only $45$% of premises in the UK [@ofcom2016], where the more
+lucrative and competitive market originally attracted the cable TV provider. 
+Whether the variation in infrastructure quality and reliability affects the 
+spatial footprint of telecommuting has not previously been investigated, in part 
+because telecommuting has not previously been a cause of concentrated demand and 
+pressure on internet services. Other intersections between digital divides have been
+subject to study.
 There are indications that those who purchase high speed connections consume
@@
 reduction in household transport spend [@bris2017ict].
-Independently of whether the additional internet use and reduced travel is
-because of increased telecommuting, these studies suggest that better internet
-services enable households to make savings and efficiencies,
-an example of the first level digital divide reinforcing the third level.
+These studies suggest that better internet services enable households to make 
+savings and efficiencies, an example of the first level digital divide reinforcing 
+the third level.
 
@@
 ISPs.
+Indeed, those who seek to test their broadband are most likely to do so because 
+they are experiencing slower speeds than expected, although any skew towards slower
+speeds is balanced by the likelihood that those who test their broadband are also 
+more ‘tech-savvy’ and / or have purchased higher speed packages that are not 
+delivering the promised level of service.
+
 We are particularly interested in upload speeds and the frequency of speed tests
-over the period from March to May $2020$, as government statements indicate this is
-when UK workers were first told to work from home if at all possible [@GovUK2020].
+over the period from March to May $2020$, as government statements indicate this 
+encompasses the period when UK workers were told to work from home if at all possible [@GovUK2020].
+Schools and various retail, leisure and hospitality businesses were closed from late 
+March, and restrictions were gradually eased from late May.
 Average upload speeds are slower than average download speeds, at $9.3$Mb/s
@@
 However, there were insufficient observations -- only $631$ speed
-tests per LAD on average -- for each for each working hour of each working day in
-each Local Authority District (LAD) to profile speeds at that level of detail.
+tests per Local Authority District (LAD) on average -- 
+for each for each working hour of each working day in each LAD to profile speeds at 
+that level of detail.
 Spatial aggregation was also necessary because we could not follow individuals or
@@
 connections to test their internet speeds more often, and the results validate our
-priors. Meanwhile, fast Virgin Media cable connections have historically
-only been available to $45$% of premises in the UK [@ofcom2016], where the more
-lucrative and competitive market originally attracted the cable TV provider.
+priors. 
 Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of

### 2021-03-11 paper/v2_taylor_francis/response_reviewers.md

@@
 The VGI approach of obtaining network speed maybe bias given that this probably is the digital guru group that will respond, maybe a discussion on this and picture the steps of selecting/soliciting and how representative this is.
+We have added the following sentence to the first paragraph in section 3.2, explaining why we believe any bias balances out: 
+'Indeed, those who seek to test their broadband are most likely to do so because 
+they are experiencing slower speeds than expected, although any skew towards slower
+speeds is balanced by the likelihood that those who test their broadband are also 
+more ‘tech-savvy’ and / or have purchased higher speed packages that are not 
+delivering the promised level of service.'
+As we explain, our intentions were to analyse a dataset of experienced internet speeds.
+Those who use the speedcheck service are not solicited in any way, and no personally identifying data is included other than the geolocation of the test.
 
@@
 If so, then it's not a problem?
+Firstly, mean speeds are not likely to be that different between regions of the UK (e.g.
+North or South England).
+The greater differences are between urban, built-up areas, and more rural areas within regions, although there are some local initiatives to address this in some areas.
+However, we assume that your comment refers to Table 1 and the differences in mean network speed between clusters?
+Whilst these differences are not large, the range for upload speeds is smaller than for download speeds, and a difference of 1-2Mbps slower may well be noticeable at the wrong moment.
+This is why we created the clusters by how the time series of speeds changes by day of the week and hour of the day, rather than focusing our analysis on mean speeds.
 
@@
 Or relates to the employment rate.
+We did not have the opportunity to run a survey.
+As we say in the introduction, prior to the pandemic, the demand for telecommuting would not have put a strain on internet connection speeds.
+We have clarified this with the addition of the following sentence:
+'High levels of demand are one of the main causes of reduced reliability and slower connection speeds, as network bandwidth becomes congested.'
+
+Also consider whether we add % furlough to regression? https://www.gov.uk/government/statistics/coronavirus-job-retention-scheme-statistics-june-2020
 
 > 4. The time series may be enhanced by putting a time range according to different events of the pandemic or other factors that relates to the pandemic.
+We chose the time range to reflect the period of first lockdown in the UK, as we say in the first paragraph of 3.2: 
+'We are particularly interested in upload speeds and the frequency of speed tests
+over the period from March to May $2020$, as government statements indicate this 
+encompasses the period when UK workers were first told to work from home if at all 
+possible [@GovUK2020].'
+We have added:
+'Schools and various retail, leisure and hospitality businesses were closed from late 
+March, and restrictions were gradually eased from late May.'
 
@@
 
+??
+
 > 6. A review of cyberinfrastructure and how network varies in different region would be useful.
+In section 2.2 of the literature review, we have added additional detail specific to the UK:
+'This finding broadly applies in the UK, where studies that also analysed broadband 
+speed checks concluded that average speeds are lower in rural areas, something that
+has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
+@riddlesden2014broadband found that levels of deprivation did not correlate with 
+first level digital divides.'
+AND 'Dense urban areas were shown to suffer more from slow down during peak hours, 
+although these services were more likely to be improved by increased competition 
+between providers, such as between new entrants and Virgin Media cable connections
+[@riddlesden2014broadband; @nardotto2015unbundling]. The latter were historically 
+available to only $45$% of premises in the UK [@ofcom2016], where the more
+lucrative and competitive market originally attracted the cable TV provider.'
+The latter bit about Virgin Media is moved from the results section, in response to reviewer 2, q7.
 
 > 7. The first part introduction can be shorted to reflect the importance of network bandwidth to resilience of pandemic impacted life.
+As mentioned in response to comment 3, we have added a sentence of clarification to highlight the link between speed and bandwidth.
+We have also restructured the introduction, so this becomes the second paragraph, the 3rd paragraph is slightly shorter, and we introduce our case study approach in the 4th paragraph.
+We also move the paragraph describing the three levels of digital divides to the literature review.
 
@@
 
-ET: after we finalise k, we can rerun k-means for 2019 and decide
+ET: after we finalise k, we can rerun k-means for 2019 and decide - could also show mean speeds on a map for 2019 - this would be part of the response to reviewer 1 q2 and q6.
 
@@
 Virgin Media as a major broadband provider should be briefly introduced in the manuscript.
+We have toned down the statement on returns to say:
+'In some aspects, these returns may have increased during the Covid-19 crisis...'
+Apologies on LAD acronym, now corrected.
+We also moved and integrated introduction of Virgin Media to the literature review s2.2. See also response to reviewer 1, q6.

### 2021-03-11 src/LAs_Clusters_k11.Rmd
updating files -- moving between computers

--- 2021-03-04 src/LAs_Clusters_k5.Rmd
@@
 ---
-title: "LAs_clusters k = 5"
+title: "LAs_clusters k = 11"
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
@@
 # load the data created by Data_Spatial.Rmd and ts_clusters.Rmd
-path.cluster <- paste(path, "/data/temp/clusters_nodiff_5.csv", sep = "")
+path.cluster <- paste(path, "/data/temp/clusters_nodiff_11.csv", sep = "")
 clusters <- read_csv(path.cluster)
@@
 # export file for future reference / backup
-# data.out.path <- paste0(path, "/data/temp/data_for_aux_k5.csv")
+# data.out.path <- paste0(path, "/data/temp/data_for_aux_k11.csv")
 # write.csv(data, data.out.path)
@@
 ```{r}
-path.out <- paste0(path, "/data/temp/LAs_Clusters_k5.RData")
+path.out <- paste0(path, "/data/temp/LAs_Clusters_k11.RData")
 save.image(path.out)

### 2021-03-11 src/LAs_Clusters_k7.Rmd
updating files -- moving between computers

--- 2021-03-04 src/LAs_Clusters_k5.Rmd
@@
 ---
-title: "LAs_clusters k = 5"
+title: "LAs_clusters k = 7"
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
@@
 # load the data created by Data_Spatial.Rmd and ts_clusters.Rmd
-path.cluster <- paste(path, "/data/temp/clusters_nodiff_5.csv", sep = "")
+path.cluster <- paste(path, "/data/temp/clusters_nodiff_7.csv", sep = "")
 clusters <- read_csv(path.cluster)
@@
 # export file for future reference / backup
-# data.out.path <- paste0(path, "/data/temp/data_for_aux_k5.csv")
+# data.out.path <- paste0(path, "/data/temp/data_for_aux_k7.csv")
 # write.csv(data, data.out.path)
@@
 ```{r}
-path.out <- paste0(path, "/data/temp/LAs_Clusters_k5.RData")
+path.out <- paste0(path, "/data/temp/LAs_Clusters_k7.RData")
 save.image(path.out)

### 2021-03-11 src/LAs_Clusters_k9.Rmd
updating files -- moving between computers

--- 2021-03-11 src/LAs_Clusters_k7.Rmd
@@
 ---
-title: "LAs_clusters k = 7"
+title: "LAs_clusters k = 9"
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
@@
 # load the data created by Data_Spatial.Rmd and ts_clusters.Rmd
-path.cluster <- paste(path, "/data/temp/clusters_nodiff_7.csv", sep = "")
+path.cluster <- paste(path, "/data/temp/clusters_nodiff_9.csv", sep = "")
 clusters <- read_csv(path.cluster)
@@
 # export file for future reference / backup
-# data.out.path <- paste0(path, "/data/temp/data_for_aux_k7.csv")
+# data.out.path <- paste0(path, "/data/temp/data_for_aux_k9.csv")
 # write.csv(data, data.out.path)
@@
 ```{r}
-path.out <- paste0(path, "/data/temp/LAs_Clusters_k7.RData")
+path.out <- paste0(path, "/data/temp/LAs_Clusters_k9.RData")
 save.image(path.out)

### 2021-03-11 src/ts_clusters_k9.Rmd
updating files -- moving between computers

--- 2021-03-04 src/ts_clusters_k5.Rmd
@@
 ---
-title: "Time series clusters, k = 5"
+title: "Time series clusters, k = 9"
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
@@
 # Define k instead of selecting it
-k <- 5
+k <- 9
 
@@
 # save clusters file
-out.clusters <- paste(path, "/data/temp/clusters_all_5.csv", sep = "")
+out.clusters <- paste(path, "/data/temp/clusters_all_9.csv", sep = "")
 write_csv(clusters, out.clusters)
@@
 # save clusters file
-out.clusters <- paste(path, "/data/temp/clusters_nodiff_5.csv", sep = "")
+out.clusters <- paste(path, "/data/temp/clusters_nodiff_9.csv", sep = "")
 write_csv(clusters.nodiff, out.clusters)
@@
 #ggsave("all_maps2.png") # this is the colour map for the preprint
-# out.path.map <- paste0(path, "/paper/v1/figures/map.up.clusters5.png")
+# out.path.map <- paste0(path, "/paper/v1/figures/map.up.clusters9.png")
 # ggsave(out.path.map)#,

### 2021-03-11 paper/v2_taylor_francis/response_reviewers.md
k = 5, 7, 9, 11

--- 2021-03-04 paper/v2_taylor_francis/response_reviewers.md
@@
   pdf_document: default
+bibliography: bibliography.bib
 ---
@@
 
->The paper reports research of detecting clusters of network access speed, especially under the pandemic for resilience.
+> The paper reports research of detecting clusters of network access speed, especially under the pandemic for resilience.
 Such a topic is of great importance as we are all in the pandemic and we rely on the network cyberinfrastructure for work as stated, and also for online education for all levels of students.
@@
 
+Thank you for your very supportive comments.
+
 > 1.
 The VGI approach of obtaining network speed maybe bias given that this probably is the digital guru group that will respond, maybe a discussion on this and picture the steps of selecting/soliciting and how representative this is.
+
+We have added the following sentence to the first paragraph in section 3.2, explaining why we believe any bias balances out: 
+"Indeed, those who seek to test their broadband are most likely to do so because 
+they are experiencing slower speeds than expected, although any skew towards slower
+speeds is balanced by the likelihood that those who test their broadband are also 
+more 'tech-savvy' and / or have purchased higher speed packages that are not 
+delivering the promised level of service."
+
+As we explain, our intentions were to analyse a dataset of experienced internet speeds.
+Those who use the speedcheck service are not solicited in any way, and no personally 
+identifying data is included other than the geolocation of the test.
 
@@
 
+Firstly, mean speeds are not likely to be that different between regions of the UK 
+(e.g. North or South England). The greater differences are between urban, built-up 
+areas, and more rural areas within regions, although there are some local initiatives 
+to address this in some areas. However, we assume that your comment refers to Table 1 
+and the differences in mean network speed between clusters? Whilst these differences 
+are not large, the range for upload speeds is smaller than for download speeds, 
+and a difference of 1-2Mbps slower may well be noticeable at the wrong moment. 
+Importantly, we only consider the test which took place during working hours, so 
+when speed drops during, let's say, a work Zoom call or when uploading large files,
+it will be noticeable. This is why we created the clusters by how the time series 
+of speeds changes by day of the week and hour of the day, rather than focusing our 
+analysis on mean speeds.
+
 > 3.
@@
 
+Regrettably, we did not have the opportunity to run a survey. As we say in the introduction, 
+prior to the pandemic, the demand for telecommuting would not have put a strain 
+on internet connection speeds. We have clarified this with the addition of the following sentence:
+"High levels of demand are one of the main causes of reduced reliability and slower connection speeds, as network bandwidth becomes congested".
+
+**Also consider whether we add % furlough to regression? https://www.gov.uk/government/statistics/coronavirus-job-retention-scheme-statistics-june-2020**
+
 > 4. The time series may be enhanced by putting a time range according to different events of the pandemic or other factors that relates to the pandemic.
+
+We chose the time range to reflect the period of first lockdown in the UK, as we say in the first paragraph of 3.2: 
+"We are particularly interested in upload speeds and the frequency of speed tests
+over the period from March to May $2020$, as government statements indicate this 
+encompasses the period when UK workers were first told to work from home if at all 
+possible [@GovUK2020]".
+
+We have also added the following text:
+"Schools and various retail, leisure and hospitality businesses were closed from late 
+March, and restrictions were gradually eased from late May".
 
@@
 
+??
+
 > 6. A review of cyberinfrastructure and how network varies in different region would be useful.
 
+In section 2.2 of the literature review, we have added additional detail specific to the UK:
+
+"This finding broadly applies in the UK, where studies that also analysed broadband 
+speed checks concluded that average speeds are lower in rural areas, something that
+has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
+@riddlesden2014broadband found that levels of deprivation did not correlate with 
+first level digital divides".
+
+and
+
+"Dense urban areas were shown to suffer more from slow down during peak hours, 
+although these services were more likely to be improved by increased competition 
+between providers, such as between new entrants and Virgin Media cable connections
+[@riddlesden2014broadband; @nardotto2015unbundling]. The latter were historically 
+available to only $45$% of premises in the UK [@ofcom2016], where the more
+lucrative and competitive market originally attracted the cable TV provider".
+
+The latter bit about Virgin Media is moved from the results section, in response to reviewer 2, q7.
+
 > 7. The first part introduction can be shorted to reflect the importance of network bandwidth to resilience of pandemic impacted life.
+
+As mentioned in response to comment 3, we have added a sentence of clarification 
+to highlight the link between speed and bandwidth. We have also restructured the 
+introduction, so this becomes the second paragraph, the 3rd paragraph is slightly 
+shorter, and we introduce our case study approach in the 4th paragraph. We also 
+moved the paragraph describing the three levels of digital divides to the literature review.
 
@@
 However, I have a few major concerns on the methodology and a lack of discussion.
+
+Thank you for your very supportive comments.
 
@@
 Virgin Media as a major broadband provider should be briefly introduced in the manuscript.
+
+We have toned down the statement on returns to say:
+"In some aspects, these returns may have increased during the Covid-19 crisis..."
+
+Apologies on LAD acronym, now corrected.
+
+We also moved and integrated introduction of Virgin Media to the literature review s2.2. See also response to reviewer 1, q6.

### 2021-03-11 src/LAs_Clusters_k9.Rmd
regressions for k=9

@@
 library(readxl)
+library(httr)
 
@@
 
+## furlough
+
+```{r}
+url <- "https://assets.publishing.service.gov.uk/government/uploads/system/uploads/attachment_data/file/891607/Coronavirus_Job_Retention_Scheme_Statistics_Tables_June_2020.xlsx"
+GET(url, write_disk(tf <- tempfile(fileext = ".xlsx")))
+f <- readxl::read_excel(tf, 6L, range = "B7:D486", col_names = F) %>% 
+  rename(lad19cd = 1,
+         lad19nm = 2,
+         furlough = 3) %>% 
+  filter(!is.na(furlough) & !is.na(lad19cd)) %>% 
+  dplyr::select(-lad19nm)
+```
+
 ## Load socio-econ data and merge
@@
   dplyr::select(-c(1,4:19)) %>% 
-  left_join(x = lad, y = . )
+  left_join(x = lad, y = . ) %>% 
+  left_join(x = ., y = f )
 
@@
 
-data.cor <- data %>%
-  dplyr::select(LAD.tests.am20,
-                pop, pop16_64,
-                managers, tech,
-                skilled, earnings,
-                total.busi)
+# data.cor <- data %>%
+#   dplyr::select(LAD.tests.am20,
+#                 pop, pop16_64,
+#                 managers, tech,
+#                 skilled, earnings,
+#                 total.busi)
 
 data.cor <- data %>%
-  dplyr::select(pop16_64, 
-                pop.dens,
+  dplyr::select(#pop16_64,
+                pop,
+                #pop.dens,
                 job.dens2018, 
@@
                 total.busi,
+                nvq4,
+                furlough,
                 LAD.tests.am20,
-                pop,
                 virgin) %>%
@@
   mutate('total.busi/pop' = total.busi/pop) %>%
-  dplyr::select(-LAD.tests.am20,-pop, -total.busi)
+  mutate('furlough/pop' = furlough/pop) %>% 
+  dplyr::select(-LAD.tests.am20,-pop, -total.busi, -furlough)
 
@@
                      managers + tech + skilled + prof + admin + caring + plant +
-                     earnings + I(total.busi/pop) + nvq4 +
+                     earnings + I(total.busi/pop) + nvq4 + I(furlough/pop) +
                      I(LAD.tests.am20/pop) + # non sensitive to addition of other tests pm
@@
             "pop, 2018",
-            #"pop. densuty, 2018",
+            #"pop. density, 2018",
             "job density, 2018",
@@
             "% of NVQ4+",
+            "% furloughed pop., 2020",
+            "n. of AM broadband tests per hab., 2020",
+            "% of Virgin Media connections"),
+          add.lines = list(c("McFadden's R squared",
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
+                             , round(r2.up[1],3), round(r2.up[1],3)),
+                           (c("N",
+                              n.obs, n.obs, n.obs,
+                              n.obs, n.obs, n.obs,
+                              n.obs, n.obs, n.obs
+                              , n.obs, n.obs))))
+
+```
+
+## Auxiliary regression, dep. var.: upload clusters, NO NVQ4
+
+```{r upload regression no NVQ4, results= 'asis', message=F, error=F, warning=F}
+aux.up <- multinom(cluster.up ~ 
+                     #pop16_64 + # non sign, while pop is
+                     pop +
+                     #pop.dens + 
+                     job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
+                     distMet + distLondon + south +
+                     managers + tech + skilled + prof + admin + caring + plant +
+                     earnings + I(total.busi/pop) + I(furlough/pop) +
+                     I(LAD.tests.am20/pop) + # non sensitive to addition of other tests pm
+                     virgin
+                   , data = data, 
+                   trace = F) # to avoid messages
+
+# McFadden's R squared
+aux.up.null <- multinom(cluster.up ~ 1, data = data)
+r2.up <- 1-logLik(aux.up)/logLik(aux.up.null)
+
+n.obs <- dim(aux.up$fitted.values)[1]
+
+stargazer(aux.up, 
+          type = "html",
+          #type = "text",
+          nobs = T,
+          digits = 3,
+          covariate.labels=c(#"pop. 16-64, 2018",
+            "pop, 2018",
+            #"pop. density, 2018",
+            "job density, 2018",
+            "distance to nearest met. area",
+            "distance to London",
+            "South of the UK",
+            "% of managerial jobs, 2020",
+            "% of tech jobs, 2020",
+            "% of skilled trade jobs, 2020",
+            "% of professional jobs, 2020",
+            "% of administrative jobs, 2020",
+            "% of leisure jobs, 2020",
+            "% of machine operation jobs, 2020",
+            "earnings, 2019",
+            "n. business est. per hab., 2019",
+            "% furloughed pop., 2020",
             "n. of AM broadband tests per hab., 2020",

### 2021-03-11 src/LAs_Clusters_k9.Rmd
no prof

@@
 
+## Auxiliary regression, dep. var.: upload clusters, NO PROF
+
+```{r upload regression no prof, results= 'asis', message=F, error=F, warning=F}
+# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
+# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/
+
+# convert dependent variable to factor
+data$cluster.up <- as.factor(data$cluster.up)
+data$cluster.up <- relevel(data$cluster.up, "4") # base category: cluster 4
+
+aux.up <- multinom(cluster.up ~ 
+                     #pop16_64 + # non sign, while pop is
+                     pop +
+                     #pop.dens + 
+                     job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
+                     distMet + distLondon + south +
+                     managers + tech + skilled + admin + caring + plant +
+                     earnings + I(total.busi/pop) + nvq4 + I(furlough/pop) +
+                     I(LAD.tests.am20/pop) + # non sensitive to addition of other tests pm
+                     virgin
+                   , data = data, 
+                   trace = F) # to avoid messages
+
+# McFadden's R squared
+aux.up.null <- multinom(cluster.up ~ 1, data = data)
+r2.up <- 1-logLik(aux.up)/logLik(aux.up.null)
+
+n.obs <- dim(aux.up$fitted.values)[1]
+
+stargazer(aux.up, 
+          type = "html",
+          #type = "text",
+          nobs = T,
+          digits = 3,
+          covariate.labels=c(#"pop. 16-64, 2018",
+            "pop, 2018",
+            #"pop. density, 2018",
+            "job density, 2018",
+            "distance to nearest met. area",
+            "distance to London",
+            "South of the UK",
+            "% of managerial jobs, 2020",
+            "% of tech jobs, 2020",
+            "% of skilled trade jobs, 2020",
+            #"% of professional jobs, 2020",
+            "% of administrative jobs, 2020",
+            "% of leisure jobs, 2020",
+            "% of machine operation jobs, 2020",
+            "earnings, 2019",
+            "n. business est. per hab., 2019",
+            "% of NVQ4+",
+            "% furloughed pop., 2020",
+            "n. of AM broadband tests per hab., 2020",
+            "% of Virgin Media connections"),
+          add.lines = list(c("McFadden's R squared",
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
+                             , round(r2.up[1],3), round(r2.up[1],3)),
+                           (c("N",
+                              n.obs, n.obs, n.obs,
+                              n.obs, n.obs, n.obs,
+                              n.obs, n.obs, n.obs
+                              , n.obs, n.obs))))
+
+```
+
 ```{r}

### 2021-03-13 paper/v2_taylor_francis/broadband.speed.covid.Rmd
Updated descriptive clusters, paper section 4, and response to reviewers.

@@
 and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
+**ET edit this bit for rationale k = 9**
 Following @sardatime the majority vote for the following CVIs was used: Silhouette
@@
 In Section [4.1](#sec:4.1), we review the temporal profile of upload speed
-by hour of the day and day of the composite week for each cluster.
+by hour of the day and day of the composite week for each of the clusters.
 Since the quality and reliability of internet services vary in time and space due
@@
 The temporal profiles of the local authority clusters have been summarised in
-Figures \ref{UpClusterL} and \ref{UpClusterS} and Table \ref{up.cluster.descr}.
-The graphs show a composite profile of mean upload speeds per hour
-per day for each cluster, with the largest five clusters, 
-in terms of the LAD membership and population, in Figure \ref{UpClusterL},
-and the next six in Figure \ref{UpClusterS}.
-These figures and table provide a comprehensive overview of the quality and
-reliability of experienced broadband in different parts of the UK.
+Figure \ref{UpCluster} and Table \ref{up.cluster.descr}, to provide an overview of 
+the quality and reliability of experienced broadband in different parts of the UK.
+Figure \ref{UpCluster} shows a composite profile of mean upload speeds per hour
+per day for each of the largest six clusters, in terms of the LAD membership and 
+population.
 
@@
 ```
-The second largest cluster, comprising $126$ local authorities and over $20$ million
-people, is cluster $13$, which has the slowest aggregate mean upload speed,
-and the second highest ratio of the standard deviation to the mean.
+The largest cluster, comprising $229$ local authorities and over $40$ million
+people, is cluster $6$, which has the slowest aggregate mean upload speed of any
+cluster, and the highest ratio of the standard deviation to the mean.
 This suggests that those living in local authorities in this cluster experienced
-some of the lowest quality broadband services in terms of upload speeds in the UK.
-However, Figure \ref{UpClusterL} shows that the high standard deviation,
-which is one indication of unreliable internet, did not disproportionately affect
-the morning peak from $9:00$-$10:59$, when upload speeds were, on average,
-only $2.6$% slower than in the evening peak period between $19:00$ and $20:59$,
-when entertainment purposes are likely to be using the most bandwidth.
-In comparison, the five LADs in cluster $1$, home to $343$ thousand people, not only
-experience the second slowest mean upload speeds and the highest ratio of standard
-deviation to the mean, but are also much more affected during the morning peak.
-
-Meanwhile, those living in the largest cluster -- $6$, with $155$ LADs and $29.5$
-million people -- experienced aggregate mean upload speeds of about $1$Mb/s faster
-than those in cluster $13$, but still lower than the other three large clusters
-and most of the smaller clusters, suggesting a middling quality of service.
-The temporal profile for cluster $6$ in Figure \ref{UpClusterL} shows that upload
-speeds are highest at $6:00$ on a Monday morning and between $23:00$ and midnight 
-on Wednesday and Thursday, but tend to be lower during the working day.
-Furthermore, experienced mean upload speeds in the morning peak are $4.4$% lower
-than in the evening peak -- a greater, more noticeable change than any of the other
-large clusters experience, suggesting poorer reliability during the working day.
-This difference is less, however than any of the clusters
-included in Figure \ref{UpClusterS}.
-
-Clusters $8$ and $12$ also have mean upload speeds under $10$Mb/s, but higher than
-clusters $1$, $13$, and $6$. However, mean upload speeds are much lower between
-$9:00-10:59$ than between $19:00-20:59$ in cluster $8$, but slightly faster in the
-morning than in the evening in cluster $12$.
-Indeed, cluster $12$ is the only cluster to experience higher speeds in the morning
-peak compared to the evening, and thus the only cluster where the temporal profile
-of internet use is closer to what might have been expected pre-pandemic.
-Among the other clusters, however, the reliability of internet services
-during the working day still varies considerably.
-Interpreting this variation from the large spikes and dips shown on
-Figures \ref{UpClusterL} and \ref{UpClusterS} is difficult,
-but the statistics in Table \ref{up.cluster.descr} show that clusters $9$ and $11$
-have the most reliable internet services.
-The ratio of standard deviation to mean in both these clusters is lowest,
-and speeds are only about 2% slower in the morning than the evening.
-Mean speeds are also higher than in any other cluster, excluding cluster $2$,
-where measures of reliability suggest poorer performance.
-
-Thus, broadband services in clusters $9$ and $11$, 
-home to over twelve million people, performed the best during the study period, 
-in terms of both quality and reliability.
-In Figure \ref{UpClusterL}, cluster $11$ shows more noticeable peaks and troughs,
-with the lowest points occurring, on average, between $6:00-7:00$ on Monday morning, 
-$14:00-15:00$ on Friday, and $16:00-17:00$ on Thursday, not the workday peak. 
-Furthermore, these slow periods offer better speeds than the average hourly
-profile of cluster $13$. Clusters $3$, $7$ and $10$ also have relatively high 
-mean upload speeds. Clusters $3$ and $10$ pair high mean speeds with
-low standard deviations relative to the mean speeds, suggesting reliability and
-resilience, as well as quality broadband services. Cluster $7$ has a higher ratio of
-standard deviation to mean, but there is less difference in average speeds between
-the morning and evening peaks than in clusters $3$ and $10$.
-
-In summary, LADs in clusters $9$ and $11$ experienced resilient
-broadband internet that could support high levels of telecommuting.
-Those in clusters $2$, $3$, $7$, and $10$ also experience higher than average
-mean speeds and rank middle to high on measures of service reliability.
-These LADs are *not* on the wrong side of the first level digital divide,
-but how likely are they to be able to take advantage of their resilient ICT
-infrastructure and services?
-Meanwhile, cluster $6$ is not only the largest in terms of number of LADs
-and population, it has the closest mean upload speed to the pre-clustered average
-for the whole sample.
-As well as average quality internet services, those in cluster $6$ also experience
-average reliability for work purposes, ranking fifth behind the four other clusters
-with populations over one million, but ahead of the smaller clusters.
-Clusters $8$ and $12$ are also close to average mean upload speeds, but show very
-different patterns in terms of reliability, whilst clusters $1$ and $13$ appear to
-suffer most from a lack of quality internet services, with slow speeds and
-high standard deviations.
-With those in cluster $1$ more likely to experience that poor reliability during the
-morning peak, is this first level digital divide occurring in areas where few are
-occupationally able to telecommute anyway, and what are the implications for
-economic resilience?
+some of the lowest quality broadband services in terms of upload speeds and
+reliability in the UK. However, as shown in Figure \ref{map.up.clusters}, some of the
+most rural areas of the UK are included in this cluster. If these areas
+suffer most from first level digital divides as described in the literature review,
+the low speeds in rural areas might be pulling down the averages in other areas in
+this largest cluster. Also, since the areas are clustered by their temporal profile
+across the working week, the graph in Figure \ref{UpCluster} indicates that speeds
+in cluster $6$ are some of the more reliable. Table \ref{up.cluster.descr} confirms
+that upload speeds in cluster $6$ during the morning peak from $9:00$-$10:59$ were, 
+on average, only $4.5$% slower than in the evening peak period between $19:00$ 
+and $20:59$, when entertainment purposes are likely to be using the most bandwidth.
+
+```{r echo=FALSE, out.width = "95%",fig.cap="\\label{map.up.clusters}Upload speed clusters for LADs"}
+img1_path <- "figures/map.up.clusters.png"
+include_graphics(img1_path)
+```
+
+Those living in the second largest cluster -- $9$, with $115$ LADs and $21.5$
+million people -- experienced aggregate mean upload speeds of more than $1.5$Mb/s 
+faster than those in cluster $6$, a lower ratio of standard deviation to mean, and
+almost the same upload speeds in the AM Peak as the PM Peak.
+Indeed, the temporal profile for cluster $9$ in Figure \ref{UpCluster}, like that
+for cluster $6$, is fairly flat. Whilst it is likely that the large numbers of 
+tests being performed by the large populations in these two clusters of LADs result 
+in less varied averages to show in the graph, the method of shape-based clustering
+suggests that these results indicate that the majority of the population of the 
+UK experience less temporal variation, and so belong in these two clusters.
+The results also confirm that those in cluster $9$ experienced a consistently better 
+service during lockdown than those in cluster $6$, in terms of both average speed 
+and reliability.
+
+Of the smaller clusters, cluster $7$ experiences quite similar mean speeds to 
+cluster $9$, and a not too dissimilar standard deviation, but mean speeds in the 
+morning peak are almost $10$% lower than the evening peak.
+Cluster $3$ boasts speeds almost $1$Mb/s higher again, but suffers from even greater
+slowdown in the morning peak. In comparison, clusters $2$ and $5$, home to a 
+little less than half a million people, experience not only above average mean 
+speeds, but also faster speeds in the morning peak compared to the evening.
+Thus, in the five LADs in these two clusters, the temporal profile of internet use 
+may be closer to what might have been expected pre-pandemic.
+In comparison, the $15$ LADs in clusters $1$ and $8$, home to over $1.4$ million 
+people, not only experience fairly average upload speeds and high ratios of 
+standard deviation to the mean, but also experience much lower speeds during 
+the morning peak than the evening peak. From the large spikes and dips shown on
+Figure \ref{UpCluster}, it is likely that this poor reliability or consistency of 
+morning internet speeds was more noticeable to those in cluster $1$, whilst those 
+in cluster $8$ experienced the lack of reliable service as a problem throughout 
+the day.
+
+In summary, the reliability of internet services during the working day appears to
+have altered for the vast majority of locations due to increased use from residents 
+told to stay at home, and, if they were fortunate, work from home.
+Yet this experience has been different in different locations. In particular, 
+LADs in cluster $9$ experienced higher speeds and more resilient
+broadband internet than those in the largest cluster, $6$.
+Those in clusters $3$ and $7$ also experience higher mean speeds and better service 
+reliability, and can be more confident that they are on the right side of the first 
+level digital divide, and that their more resilient ICT infrastructure and services
+can robustly support higher levels of telecommuting.
+On the other hand, whilst high levels of telecommuting may be the cause of morning 
+speeds over 15% lower than those in the evening, those in clusters $1$ and $8$ 
+also experience lower mean speeds than clusters $3$, $7$, and $9$, but not as low as 
+cluster $6$. Whilst the differences may not be large, the lower upload speeds in 
+cluster $6$ or the poor reliability in clusters $1$ and $8$ could still have had 
+implications for economic resilience if combined with other types of digital divide.
+This potential will be explored in the next section.
 
@@
 Using an auxiliary multinomial logit regression, we test whether the clusters that
-have higher mean speeds and more reliable services consist of LADs that are more
-urban, affluent, and / or more likely to benefit from a choice of high quality
-internet services.
-We also estimate which of our clusters are more likely to have a higher proportion of
-occupations where the nature of the work enables telecommuting.
+have higher mean speeds and more reliable services do indeed consist of LADs that 
+are more urban or closer to major urban areas and are more likely to benefit from 
+a choice of high quality internet services. This part of the regression aims to
+confirm any first level digital divides.
+Next, to better understand how the clusters fare in relation to the second level
+digital divide, we consider which LADs in which clusters have a higher proportion of 
+occupations where the nature of the work and the skills that occupation employs 
+enable telecommuting. 
+Finally, we consider the rate of return on internet use, or third level digital 
+divides, by reviewing which clusters have the highest earnings and job density,
+and also which clusters experienced the highest share of population furloughed 
+during this period in the pandemic.
+
 The results of the auxiliary regression are presented in Table \ref{aux}.
 The dependent variable is the LAD cluster membership as described in the
-[methods and data](#sec:3) section
-and equation \ref{eq1}. Each column represents a different cluster.
+[methods and data](#sec:3) section and equation \ref{eq1}. 
+Each column represents a different cluster.
 The reference case is cluster $4$, which includes only the local authority of
 Hambleton in North Yorkshire, a rural area of just over ninety thousand people.
-Mean, experienced upload speeds in cluster $4$ (see Table \ref{up.cluster.descr})
-are between the average speeds for the $13$ clusters ($9.9$Mb/s) and the
-pre-clustered average for the whole sample ($9.3$Mb/s).
+Mean, experienced upload speed in cluster $4$ (see Table \ref{up.cluster.descr})
+is close to the pre-clustered average for the whole sample ($9.3$Mb/s).
 However, the standard deviation for cluster $4$ and the difference between average
@@
 connections. Regarding the former, we expect people in LADs with more unreliable
-connections to test their internet speeds more often, and the results validate our
-priors. 
-Those in clusters $2$, $3$, $9$ and $11$ benefit from a higher proportion of
-Virgin connections, which is an indication that people in these clusters are
-more likely to live in urban areas, with more choice of broadband services.
-In other words, they are more likely to be on the right side of the first level,
-infrastructure-based digital divide, as we expected from the analysis in Section
-[4.1](#sec:4.1).
-
-We employ distance from London and from the nearest metropolitan area
-(including London) as two variables depicting peripherality, urban structure and,
-potentially, first level digital divides.
-The broadband speed tests run in the authorities in cluster $3$ are more likely to
-be taking place close to London than those run in any of the other clusters, and
-two of the four authorities in cluster $3$ are the London commuter towns of Harlow
-and Luton.
-However, even though London was also included in the variable calculating distance
-from the centre of one of either the ten largest metropolitan areas in England,
-or Glasgow or Cardiff, tests run in cluster $3$ are likely to be furthest away.
-Thus, it is important to consider the membership of each cluster as well as the
-regression results. Corby and Eastbourne, the other two authorities in cluster $3$
-are large, accessible towns, although not part of metropolitan areas.
-
-```{r echo=FALSE, out.width = "95%",fig.cap="\\label{map.up.clusters}Upload speed clusters for LADs"}
-img1_path <- "figures/map.up.clusters.png"
-include_graphics(img1_path)
-```
-
-Meanwhile, LADs in cluster $7$ are most likely to be near the centre of a large
-metropolitan area, even though the four local authorities of cluster $7$ include
-no central urban boroughs and only one LAD that is part of a metropolitan area of
-governance - Tameside in Greater Manchester. This may explain why those in cluster
-$7$ are the second least likely to be served by Virgin Media.
-It is also a demonstration of the complexity of both experienced broadband 
-upload speeds as captured by time-series clustering, and the geography of 
-first level digital divides as a product of quality as well as availability.
-In comparison, cluster $1$, which our analysis suggests lacks broadband resilience, 
-contains five, mainly rural authorities, but they are closer to a metropolitan area 
-than authorities in cluster $3$, although furthest from London, perhaps because 
-they are scattered around the country -- see Figure \ref{map.up.clusters}.
-
-Internet resilience is also more nuanced than our -- arguably crude
--- dummy variable depicting the North-South economic divide, which assigns $1$ to
-LADs located in Greater London, Southeast, Southwest and East of England regions,
-and $0$ to the rest.
-The authorities most likely to be in the South are those in cluster $13$,
-which was identified as having the slowest mean upload speeds of any of the clusters,
-and thus a low level of service.
-However, cluster $13$ includes some rural, remote areas of the country, such as
-Northwest Scotland, Cornwall and Powys in Wales as shown on Figure
-\ref{map.up.clusters}.
-It also includes major metropolitan centres in the North -- Liverpool, Newcastle --
-and South -- Bristol, and nine (of $32$) London Boroughs.
-There are also plenty of Southern home county and suburban areas.
-The standard deviation measure for cluster $13$ is high, but speed variation is
-low during the morning peak, suggesting that the estimates for reliability are
-inconsistent. Considering that this is one of the largest clusters, and thus the
-averages incorporate more noise than some of the smaller clusters,
-it may be that the LADs in this cluster do not all suffer equally from
-a first level digital divide.
-
-Yet we need to consider the results for other variables in Table \ref{aux} to better
-determine whether clusters $1$ and $13$, which appear to suffer most from poor
-quality internet services, are also more likely to have a low skilled workforce,
-less able to benefit from telecommuting.
-Cluster $1$ has the lowest proportion of educated people,
-and the lowest earnings, despite recording the highest proportion of managerial and
-professional jobs, and the second highest proportion of tech jobs.
-Rural areas such as those in cluster $1$ are home to many older, retired people 
-[@blank2018local], which might explain these results or perhaps, 
-as these figures are relative, we could note that cluster $1$ also has a greater 
-proportion of skilled trades than other clusters. 
-In any case, it appears that the first level digital divide 
-reinforces other inequities in cluster $1$, where the even slower than average 
-morning upload speeds shown in Table \ref{up.cluster.descr} suggest that internet
-users were more active during the working day. Meanwhile, those in cluster $13$ are 
-more likely to earn more and have a better education despite poor internet services.
-Cluster $13$ also has the most businesses per inhabitant, but is somewhere in the
-middle in terms of job density. If this is an indication of a high number of SMEs,
-it could explain the variable internet quality, considering that small businesses
-have not been seen as the most valuable customers for higher speed broadband services
-if they are not located near residential customers [@ofcom2016].
-
-Cluster $6$ is the largest cluster, and thus, like cluster $13$, there is more
-noise within the averages we use to measure internet quality and reliability.
-Our results indicate average mean speeds, and Table \ref{aux} shows that
-cluster $6$ also falls towards the middle of the clusters on many of the
-socioeconomic variables. It ranks third or fourth out of the eleven in terms of
-the likelihood of having a higher proportion of managerial, professional, and
-tech jobs, as well as higher earnings, but is fourth from bottom for educational
-attainment. The LADs in cluster $6$ are also diverse, with few truly remote areas,
-but urban areas throughout England, including Birmingham, Leeds, Sheffield,
-twelve London Boroughs, and many suburban areas and smaller cities like Oxford
-and Cambridge. The capital cities of the other UK nations, Belfast, Cardiff and
-Edinburgh, are also in this cluster, suggesting perhaps that the lower level of
-reliability discussed in Section [4.1](#sec:4.1) may be due to increased demand,
-e.g. for telecommuting, despite lacking the most resilient internet connections.
-
-In contrast, Clusters $9$ and $11$ enjoy resilient internet connections, 
-but LADs in Cluster $9$ are more likely to host highly educated individuals 
-with higher earnings than cluster $11$, which has a negative coefficient for 
-the NVQ4 variable. Cluster $11$ has fewer businesses per inhabitant and the 
-second highest job density and cluster $9$ the second lowest.
-These coefficients might indicate that resilient broadband infrastructure
-generates higher returns for those in cluster $9$, where slightly more slowdown in
-the AM peak was detected -- see Table \ref{up.cluster.descr}.
-LADs in cluster $9$ are less likely to be in the South and those in cluster $11$
-more likely, although the coefficient for cluster $9$ might be skewed by
-the presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
-different economic profile than England. Still, both clusters consist of larger
-LADs in terms of population, including districts within five of England's ten
-largest cities, and a number of other stand-alone urban areas and large market towns 
--- see Appendix [1](#appendix1).
-These urban locations are on the right side of the first level digital divide,
-and the regression results suggest that internet resilience is supporting a wide
-range of small and large urban economies.
-
-LADs in clusters $2$, $3$, $7$, $10$ are also on the right side of the first level
-digital divide, experiencing higher than average mean upload speeds and ranking 
-high to middle on measures of service reliability.
-LADs in clusters $2$ and $10$ are also more likely to have highly skilled
-workers to take advantage of working from home opportunities.
-Cluster $2$ is comprised of just two LADs, with the lowest job density of any
-cluster. Cluster $10$ is comprised of four LADs, including two
-peripheral suburban areas of Birmingham, the East London Borough of Newham,
-and Dundee. Suburbs are considered the most likely urban form in which telecommuters
-live [@e2018does], and Dundee has a reputation for tech startups [@technation2017].
-LADs in cluster $7$, which include the suburbs South of Belfast, a suburban
-district of Greater Manchester, and some other Midlands and North Yorkshire towns 
-and villages have the highest earnings, and thus may also be making the
-most of their resilient internet. In contrast, LADs in cluster $3$ are associated
-more with lower skills and have the highest proportion of individuals in machine
-operation jobs. Arguably, LADs in clusters $3$ benefit the least from their
-resilient internet infrastructure in an era when working from home became a vehicle
-for economic resilience.
-
-Finally, clusters $8$ and $12$ consist of LADs enjoying close to average
-mean upload speeds, but opposing patterns of internet reliability during the morning
-peak. LADs in cluster $8$ are likely to host more individuals employed in tech and
-administrative occupations than any other cluster, as well as many in managerial
-occupations, whilst the opposite applies to skilled trade and leisure occupations.
-These LADs are characterised by a very small positive likelihood of more educated
-residents, but are not significantly likely to earn more than other clusters.
-Including suburbs near Leicester, Cardiff, and Birmingham and south of London, 
-the LADs in cluster $8$ seem to likely to have the skills and occupations that 
-would benefit from quality internet services, but suffer from poor internet 
-resilience and reliability.
-
-In comparison, LADs in cluster $12$ are less likely to be able to benefit from
-quality internet services if they had them, with fewer individuals achieving
-NVQ4 or better and lower levels of occupations that would benefit from homeworking.
-Yet LADs in this cluster benefit from the second highest level of earnings, and
-have the largest stock of businesses per inhabitant and the highest job density.
-This density of businesses could be why cluster $12$ is the only cluster with higher
-speeds in the morning peak during the study period. If people are at home, business
-premises might well have been abandoned. Cluster $12$ is home to $1.5$ million people
-spread across twelve LADs, from London to Wales as listed in Appendix [1](#appendix1). 
-This spatial diversity demonstrates that the temporal clustering of internet
-resilience is not necessarily spatially dependent, and digital divides do not
-necessarily overlap with economic ones.
-
-# Conclusions {#sec:5}
+connections to test their internet speeds more often, and the results show that 
+those in cluster $6$ are by far the most concerned about reliability at that time 
+of day. Meanwhile, those in clusters $2$, $3$, and $9$ benefit from a higher 
+proportion of Virgin connections, which is an indication that people in these 
+clusters are more likely to live in urban areas, with more profitable broadband 
+markets, and a better choice of broadband services. 
+We also employ distance from London and from the nearest metropolitan area
+(including London) as two variables depicting peripherality and urban structure.
+However, whilst significant, the pattern of coefficients of these variables is 
+inconclusive in confirmed first level digital divides. 
+Even though London was one of the ten largest metropolitan areas in 
+England, which, along with Glasgow and Cardiff, were identified to calculate the 
+variable estimating the impact of distance from the centre of a metropolitan area, 
+the coefficients for the two variables are mostly small and the signs for some are 
+opposite.
+
+When looking at the constituent authorities (see full list of LADs in 
+Appendix [1](#appendix1) as well as Figure \ref{map.up.clusters}), 
+however, Cluster $9$ is clearly more urban, including $13$ of $32$ London Boroughs, 
+five of the seven constitutent LADs of the West Midlands conurbation, 
+nine of the ten boroughs of Greater Manchester, and five of nine other 
+large metropolitan areas coded in for the 'distance to nearest met area' variable. 
+Cluster $9$ also includes the other main cities in the East Midlands, Leicester 
+and Derby, as well as smaller cities known for their knowledge economy, such as 
+Oxford and Milton Keynes. 
+However, as previously mentioned, there is more noise within the variables for the 
+two largest clusters than the smaller ones, and cluster $6$ also contains many urban 
+areas. These include $16$ London boroughs, Birmingham City, Bolton in Greater 
+Manchester, Leeds, Liverpool, Newcastle and Bristol, as well as small cities with 
+knowledge economies such as Cambridge, Edinburgh, and Reading and its neighbours in 
+the high-tech agglomeration of the Thames Valley. 
+And yet cluster $6$ also includes some of the most rural areas in the country, 
+showing how important it is to interpretation to consider the membership of the 
+cluster as well as the regression results.
+
+The coefficients in Table \ref{aux} demonstrate that those in cluster $6$ were more 
+likely to hold managerial, tech and professional jobs than those in cluster $9$, and 
+were the least likely to be furloughed of any cluster. 
+Thus, although speeds were slow and not as reliable during the morning peak 
+as in cluster $9$, the skills that enabled telecommuting also enabled greater 
+returns from doing so, despite lower than average earnings.
+Cluster $6$ also has the second highest job density, with more jobs to keep.
+Likewise, those in LADs in clusters $1$ and $8$ had the highest and second highest 
+proportions of residents working in tech or managerial occupations despite suffering 
+from lower speeds and unreliable services, another mismatch between first and 
+second level digital divides. Yet unlike cluster $6$ and despite occupations with 
+better digital skills, cluster $8$, with its five peripheral suburban areas in the 
+Midlands, Scotland and south of London, as well as rural West Devon saw the third 
+highest proportion of its working population furloughed. It also had the second 
+lowest average earnings in 2019, suggesting that its population was more likely to
+be on the wrong side of the third level digital divide before, as well as during 
+the pandemic. In contrast, the nine LADs of cluster $1$, which include more rural 
+areas, plus Westminster in central London had the lowest earnings of any cluster, 
+but also lowest level of furlough after cluster $6$.
+
+The regression results for cluster $7$ also reveal the complexity of intersecting 
+and diverging digital divides. Along with $3$ and $9$, the analysis in 
+Section [4.1](#sec:4.1) suggested this cluster was on the 'right' side of any first
+level digital divides, yet the regression shows that it has the lowest 
+proportion of Virgin connections. It is furthest from a metropolitan area, 
+but closest to London. The five LADs in cluster $7$ are all suburban, making it 
+sensible that the cluster is in the group with better infrastructure, but none are 
+central urban boroughs and, along with a London suburb, Leeds suburb, Manchester 
+suburb, and West Midlands suburb, the fifth LAD is outside Belfast. 
+Belfast was not included among the metropolitan areas, and whilst there has been 
+substantial investment in high speed broadband in Northern Ireland in recent years, 
+including by Virgin, take-up may still lag behind historic networks in other 
+urban markets.
+
+Cluster $7$ has the lowest proportion of residents currently employed in 
+professional occupations, and among the lowest proportions in managerial or 
+technical occupations. This suggests potentially a low level of skills to take 
+advantage of the quality internet service available, a second level digital divide. 
+Cluster $7$ also has the lowest job density. However, with the largest proportion 
+of skilled tradespeople, and the second highest number of businesses per inhabitant, 
+those in cluster $7$ ranked second in terms of earnings in 2019. On the other hand, 
+the lack of some of the occupations most likely to telecommute during lockdown 
+may have contributed both to a substantial proportion being furloughed in 2020, 
+and even to the relative reliability of broadband speeds compared to many of the 
+other smaller clusters.
+
+Cluster $3$, meanwhile, had the greatest proportion of its workforce furloughed of
+any cluster, despite having the highest job density and the second highest average 
+speeds. Ranking in the middle for most of the occupations estimated, at least some 
+of its population should have the skills to work from home. The geography would 
+suggest economic advantage as well, with eight of its twelve LADs in the South 
+of England, including two London Boroughs, four suburban areas north of London, 
+and the two tightly bounded urban areas of Eastbourne and Ipswich -- a greater 
+proportion than any other cluster. 
+Yet geographic position and job density do not appear to equate to particularly high 
+average earnings in 2019, nor did they protect residents from furlough in 2020. 
+In contrast, cluster $5$, comprised of three suburban districts in the Northwest, 
+Midlands, and the London green belt, had the highest average earnings in 2019, 
+despite low levels of residents working in managerial, tech and professional 
+occupations. Suburbs are considered the most likely urban form in which 
+telecommuters live [@e2018does], yet the contrast between these small, mainly 
+suburban clusters -- $7$, $3$, and $5$, suggest that they are not all equally
+economically resilient in a pandemic.
+
+Cluster $2$, which, like cluster $5$, had faster morning broadband speeds than 
+evening speeds, as shown in Table \ref{up.cluster.descr} contains two remote LADs, 
+the Isles of Scilly and Northeast Lincolnshire.
+Whilst the low demand may be due to the least residents in tech occupations, it may
+also be due to high numbers of retired people. Rural areas such as those in cluster 
+$2$ are often home to many older, retired people [@blank2018local], which may also 
+explain some of the contradictions in cluster $1$ with its low levels of furlough
+and low earnings, whilst the presence of Westminster in that cluster might be why
+there are the highest levels of tech and managerial occupations.
+
+In summary, the regression results indicate the complexity of measuring second and 
+third level digital divides within spatial aggregations where the geography of 
+first level digital divides has been captured using time-series clustering of
+experienced broadband upload speeds as a product of reliability not just
+availability. Whilst the differences in mean speeds between clusters were not large,
+the temporal clustering of internet resilience showed much greater variation, 
+and was not spatially dependent upon distance to large urban areas, or relative
+location. Internet resilience supports a wide range of small and large urban 
+economies, but also fails, or at least frustrates a wide range of other urban 
+economies, as demonstrated by the number of tests per capita run in cluster $6$.
+Yet our analysis of the likelihood of furlough within the population, especially in 
+the smaller suburban clusters $7$, $3$, $5$, and also $8$ shows that digital divides 
+were as likely to diverge as intersect during the pandemic, and did not necessarily 
+overlap with prior economic divides. With a much greater share of the population 
+likely to continue to work from home in the future, and with changing attitudes 
+towards residential locations, our analysis suggests that first level digital divides
+should be seen as a function of reliability as well as availability, of upload 
+speeds as well as download speeds, and of greater spatial variation than might
+have previously been considered.
+
+# Discussion and Conclusions {#sec:5}
 
@@
 telecommuting nor with the necessary skills to effectively use the internet to
-telecommute, this paper raises the issue that places without good internet 
-connectivity will still struggle more than other places to achieve economic 
-resilience in a period like the current pandemic when internet resilience is 
-so vital.
-Indeed, our analysis demonstrated that the temporal profiles of twelve of our
-thirteen clusters had slower upload speeds in the morning than in the evening.
+telecommute, this paper raises the issue that places may depend upon good internet 
+reliability as well as connectivity to achieve economic resilience in a period like 
+the current pandemic when internet resilience is so vital.
+
+Our analysis demonstrated that the temporal profiles of seven of our
+nine clusters had slower upload speeds in the morning than in the evening.
 The opposite is likely to have been the norm prior to the pandemic, as level of
@@
 internet activity throughout the UK, not just in areas with more digital industry
-or better skills.
-
-Upload speeds have not previously been seen as integral to universal service,
-considering there has never before been such extreme demand for telecommuting
-and operations such as video calls.
-This may be why the average upload speeds for the largest two clusters
--- $6$ and $13$ -- contain so many LADs that are centres of the knowledge economy,
-from Oxford and Cambridge in the former, to areas like Reading in the latter, 
-which has the highest concentration of digital businesses in the country 
-[@technation2017].
-Thus, whilst some areas suffer from an intersection of digital and economic divides,
-such as those in cluster $1$, other areas, including cluster $8$ as well as $6$ and
-$13$ also found their digital infrastructure to be less than reliable when 
-confronted with the sudden change in the timing and type of demand.
-In contrast, LADs such as the digital hub, Milton Keynes, in cluster $9$ were able 
-to benefit both from reliable internet connections and populations which were 
-familiar with working from home and could capitalise on their digital infrastructure.
-Yet the nuanced picture we gained through our analysis of the UK case study suggests
-that quality internet connectivity may also have enabled other LADs, such as those in
-cluster $3$ or $11$ to gain ground during the pandemic despite lower skill levels.
+or better skills. Furthermore, upload speeds have not previously been seen as 
+integral to universal service, considering there has never before been such extreme 
+demand for telecommuting and operations such as video calls. Yet those in the 
+largest cluster, $6$, were clearly frustrated by lower upload speeds. 
+Their speeds were also less reliable, perhaps because they had higher levels of
+demand than the second largest cluster, $9$, where many more employees were
+furloughed.
+
+Almost all of the largest urban areas in the UK, as well as many smaller cities, 
+were split between the largest two clusters -- $6$ and $9$ -- and both contain LADs 
+that are centres of the knowledge economy, from Cambridge in the former to Oxford 
+in the latter, or have high concentrations of digital businesses, like Reading in 
+the former and Milton Keynes in the latter [@technation2017].
+Yet there are both intersections and divergences of digital and economic divides.
+LADs in cluster $9$ were able to benefit both from reliable internet connections and
+populations which were familiar with working from home and could capitalise on their 
+digital infrastructure, although overall, they appeared to have a lower rate of
+return in the pandemic, with far greater numbers furloughed. 
+Is this an indication of both historic economic division, as well as the third level 
+digital divide in cities in the North and Midlands unable to capitalise on their 
+digital infrastructure?
+Meanwhile, did the greater presence in London and the South of LADs in cluster $6$,
+as well as the inclusion of many rural areas less impacted by the pandemic help
+those in this cluster be more resilient and gain greater returns, despite using 
+less reliable internet services?
+The nuanced picture we gained through our analysis of the UK case study suggests
+that in cluster $1$ as well as $6$, being on the right side of the second level
+digital divide had a greater impact on economic resilience and therefore the third
+level digital divide, than having quality internet connectivity.
 

### 2021-03-13 src/descriptive_Clusters.Rmd
Updated descriptive clusters, paper section 4, and response to reviewers.

@@
 
-LAD.cluster <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
+LAD.cluster <- paste(path, "/data/temp/clusters_nodiff_9.csv", sep = "")
 
@@
 #summarise upload speed clusters by hour of each weekday and plot
-#plot largest 5 clusters first - all over 1 million pop
+#plot largest 6 clusters - all over 500k pop
 cluster.up <- TS2020 %>%
@@
 cluster.up <- left_join(cluster.up, LAD.cluster.up)
-cluster.up <- cluster.up[which(cluster.up$up.LADs > 9),]
+cluster.up <- cluster.up[which(cluster.up$up.LADs > 4),]
 cluster.up$clusterID <- as.factor(cluster.up$clusterID)
@@
 #plot cluster statistic for mean upload speed
-UpClusterL <- ggplot(cluster.up, aes(help.date.time, mean.up, 
+UpCluster <- ggplot(cluster.up, aes(help.date.time, mean.up, 
                                       colour = clusterID)) +
@@
 #export UpClusterL plot for paper
-out.plot <- paste(path, "/paper/v1/figures/UpClusterL.png", sep = "")
+out.plot <- paste(path, "/paper/v1/figures/UpCluster.png", sep = "")
 png(out.plot)
@@
 #export UpClusterS plot for paper
-out.plot <- paste(path, "/paper/v1/figures/UpClusterS.png", sep = "")
-png(out.plot)
-UpClusterS
-dev.off()
+#out.plot <- paste(path, "/paper/v1/figures/UpClusterS.png", sep = "")
+#png(out.plot)
+#UpClusterS
+#dev.off()
 

### 2021-03-15 paper/v2_taylor_francis/response_reviewers.md

@@
 "High levels of demand are one of the main causes of reduced reliability and slower connection speeds, as network bandwidth becomes congested".
-
-**Also consider whether we add % furlough to regression? https://www.gov.uk/government/statistics/coronavirus-job-retention-scheme-statistics-june-2020**
+We also have added a variable to the auxiliary regression tracking the share of
+the population that was put on furlough. This was the scheme in the UK which, at 
+least initially, prevented much change in the employment rate, but did show who
+could work and who could not at the height of the first wave of the pandemic.
 
@@
 
-??
+**?? - a little in lit review, but also in discussion / conclusion added for reviewer 2?**
 
@@
 
+We have tightened up the results section, and begun this discussion in the last 
+paragraph of the results section and then expanded the conclusion to include more
+discussion.
+
 > 5.
@@
 Could education and demographic variables be used to represent level-2 digital divide?
+
+We clarify that the occupations included in the regression are more about the level-2 
+digital divide, whereas the average earnings and, particularly during the pandemic 
+the % furloughed, are indications of level-3 digital divides. We did have education
+in the previous submission (NVQ4+), but have removed it due to multicollinearity with
+some of the professions, which give more indication of not only education, but digital
+skills.
 

### 2021-03-22 paper/v2_taylor_francis/response_reviewers.md
ET changes / responses to reviewer comments

@@
 ---
+
 Dear Prof. Bian,
@@
 
-We have added the following sentence to the first paragraph in section 3.2, explaining why we believe any bias balances out: 
+We have added the following sentence to the first paragraph in section 3.2, explaining
+why we believe any bias balances out:
 "Indeed, those who seek to test their broadband are most likely to do so because
@@
 
-We chose the time range to reflect the period of first lockdown in the UK, as we say in the first paragraph of 3.2: 
+We chose the time range to reflect the period of first lockdown in the UK, as we
+say in the first paragraph of 3.2:
 "We are particularly interested in upload speeds and the frequency of speed tests
@@
 
-**?? - a little in lit review, but also in discussion / conclusion added for reviewer 2?**
+We have now added the following paragraph in the 'Digital divides and economic resilience'
+section.
+
+"Currently there is a hotly debated discussion in the literature regarding these
+exact changes in transportation and the structure of cities. We now have enough
+hard data to observe the drastic change in the space-time geography of cities around
+the world [e.g. @google2020; @shibayama2021impact].
+Although there is a broad agreement that these changes
+during the pandemic have played a pivotal role in stopping the spread of the virus
+[@jia2020population; @yang2020taking; @mu2020interplay],
+we still have no evidence to what extend these increased levels of working from
+home, and the consequent decrease in commuting flows and the structure of cities,
+will remain post-pandemic".
 
@@
 
-ET
+Please see our response to the next point.
 
@@
 
-ET
+Following your suggestion, we have now opted to present the cluster analysis and
+the subsequent multinomial regression using $k = 13$. We indeed believe that the
+results are more clear now. we have added the following justification in the Methods
+and Data section:
+
+"We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$, calculated
+the cluster validity indices (CVIs) and then run the
+subsequent multinomial regression -- see the end of this section for more details
+for the different *k*.
+Following @sardatime, to identify the optimal *k* we used the majority vote for
+the following CVIs: Silhouette (max), Score function (max), Calinski-Harabasz (max),
+Davies-Bouldin (min), Modified Davies-Bouldin (DB*, min), Dunn (max), COP (min).
+Nevertheless, we opted against using the optimal $k = 13$ solution as it was too
+large to allow for communicable LAD clusters. Instead we opted for a smaller $k = 9$,
+which led to a rather similar spatial pattern and, importantly, to a much higher
+R-squared in the subsequent explanatory regression ($0.44$ instead of $0.34$)."
+
+Regarding multicolinearity, the highest correlations were observed between $NVQ4+$
+and $professional jobs$ ($0.78$) and $NVQ4+$ and earnings ($0.69$). Hence, we decided
+to exclude $NVQ4+$ from the regression.
 
@@
 ET: after we finalise k, we can rerun k-means for 2019 and decide - could also show mean speeds on a map for 2019 - this would be part of the response to reviewer 1 q2 and q6.
+**Let's discuss this**
 
@@
 We have tightened up the results section, and begun this discussion in the last
-paragraph of the results section and then expanded the conclusion to include more
-discussion.
+paragraph of the results section and then expanded the conclusion section to include
+a discussion of the results.
 
@@
 the % furloughed, are indications of level-3 digital divides. We did have education
-in the previous submission (NVQ4+), but have removed it due to multicollinearity with
-some of the professions, which give more indication of not only education, but digital
+in the previous submission (NVQ4+), but have removed it due to multicollinearity
+-- see response for point 2, which give more indication of not only education, but digital
 skills.
@@
 The current color scheme does not help interpretation at all.
+
+x-axis for Figures 1 and 2 have been amended accordingly.
+
+The temporal profiles of the upload clusters are now included in Figure 3.
+
+Figure 5 now labelled Figure 4: the map has now been redesigned using a different
+colour palette to better reflect the discrete nature of the underpinning data --
+the time-series clusters. Points and labels of the main UK cities were also added.
 
@@
 We also moved and integrated introduction of Virgin Media to the literature review s2.2. See also response to reviewer 1, q6.
+
+## References

### 2021-03-22 paper/v2_taylor_francis/broadband.speed.covid.Rmd
ET changes / responses to reviewer comments

@@
 
-<!-- [our google doc](https://docs.google.com/document/d/1PWjkmgzWGYKR9wFogKYw7l-8mZLoORt593x-Tu-f2-M/edit#heading=h.i5om1o8wpcd9) -->
-
 During the pandemic, working from home using digital technologies, whether partially
@@
 The ability to work from home or telecommute meant millions retained their jobs and,
-to a varying extent, maintained productivity during periods of strict
-lockdowns around the world.
+to a varying extent, maintained productivity during periods of strict lockdowns 
+around the world.
 However, this ability has not been evenly distributed socially or spatially,
@@
 For example, $47$% of people in employment in the UK worked solely from home in April
-$2020$, whilst the same figure only reached $5$% the year before [@ons2020;
-@ons2020lm2019].
+$2020$, whilst the same figure only reached $5$% the year before [@ons2020; @ons2020lm2019].
 Similar figures have been reported for other countries [@felstead2020homeworking].
@@
 Our analysis enables us to better understand how the spatial and social distribution
-of education, occupation and online accessibility intersect to enable or hinder the
-practice of telecommuting at a time of extreme demand.
+of education, occupation and online accessibility intersect to enable or hinder 
+the practice of telecommuting at a time of extreme demand.
 We also consider what lessons can be learned from this time for a future where
@@
 <!-- Who works from home-->
-In this analysis, the terms ‘telecommuting’ and ‘working from home^[See also the popular social media hashtag #WFM]’ are used
-interchangeably, as most remote labour during the Covid-19 crisis was carried out
-in the homes of individual employees rather than any other location [@eurofound2020].
+In this analysis, the terms 'telecommuting' and 'working from home^[See also the popular social media hashtag #WFM]' 
+are used interchangeably, as most remote labour during the Covid-19 crisis was carried 
+out in the homes of individual employees rather than any other location [@eurofound2020].
 However, previous research has explored how telecommuting
@@
 ## Digital divides and economic resilience {#sec:2.2}
+
 <!-- WFH and digital divides-->
@@
 has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
-@riddlesden2014broadband found that levels of deprivation did not correlate with 
+Moreover, @riddlesden2014broadband found that levels of deprivation did not correlate with 
 first level digital divides.
@@
 demographic characteristics than simple rurality or urbanisation.
-Dense urban areas were shown to suffer more from slow down during peak hours, 
+Dense urban areas were shown to suffer more from slowdown during peak hours, 
 although these services were more likely to be improved by increased competition 
@@
 the third level.
+
+<!-- spatio-temporal -->
+Currently there is a hotly debated discussion in the literature regarding these 
+exact changes in transportation and the structure of cities. We now have enough 
+hard data to observe the drastic change in the space-time geography of cities around 
+the world [e.g. @google2020; @shibayama2021impact]. 
+Although there is a broad agreement that these changes 
+during the pandemic have played a pivotal role in stopping the spread of the virus 
+[@jia2020population; @yang2020taking; @mu2020interplay],
+we still have no evidence to what extend these increased levels of working from 
+home, and the consequent decrease in commuting flows and the structure of cities, 
+will remain post-pandemic.
 
@@
 ## Time-Series clustering {#sec:3.1}
+
 <!-- clustering in general -->
@@
 ## Experienced Broadband Speeds {#sec:3.2}
+
 <!-- introduction of our broadband speeds data -->
@@
 The time-series were calculated for each of the $382$ LADs in the UK, standardised,
-and then a *k*-means partitioning around medoids clustering algorithm
-was applied using DTW.
-We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$
-and used cluster validity indices (CVIs) to pick the optimal solution of $k = 13$.
-**ET edit this bit for rationale k = 9**
-Following @sardatime the majority vote for the following CVIs was used: Silhouette
-(max), Score function (max), Calinski-Harabasz (max), Davies-Bouldin (min), Modified
-Davies-Bouldin (DB*, min), Dunn (max), COP (min).
+and then a *k*-means partitioning around medoids clustering algorithm was applied using DTW.
+We initially run the algorithm for $k \in \mathbb{N} \bigcap [5,15]$, calculated 
+the cluster validity indices (CVIs) and then run the 
+subsequent multinomial regression -- see the end of this section for more details
+for the different *k*.
+Following @sardatime, to identify the optimal *k* we used the majority vote for 
+the following CVIs: Silhouette (max), Score function (max), Calinski-Harabasz (max), 
+Davies-Bouldin (min), Modified Davies-Bouldin (DB*, min), Dunn (max), COP (min). 
+Nevertheless, we opted against using the optimal $k = 13$ solution as it was too 
+large to allow for communicable LAD clusters. Instead we opted for a smaller $k = 9$, 
+which led to a rather similar spatial pattern and, importantly, to a much higher 
+R-squared in the subsequent explanatory regression ($0.44$ instead of $0.34$).
 
@@
 
-```{r echo=FALSE, out.width = "95%",fig.cap="\\label{UpClusterL}Temporal profilies for upload speed large clusters"}
-img1_path <- "figures/upClusterL.png"
+```{r echo=FALSE, out.width = "95%",fig.cap="\\label{UpCluster}Temporal profilies for upload speed clusters"}
+img1_path <- "figures/UpCluster.png"
 include_graphics(img1_path)
@@
 
-```{r echo=FALSE, out.width = "95%",fig.cap="\\label{UpClusterS}Temporal profilies for upload speed small clusters"}
-img1_path <- "figures/upClusterS.png"
-include_graphics(img1_path)
-```
-
-``````{r echo=FALSE, message=FALSE, results = "asis"}
+```{r echo=FALSE, message=FALSE, results = "asis"}
 path.clustersLADup <- paste(path, "./data/temp/LAD.cluster.up.csv", sep = "")
@@
 ```
+
 The largest cluster, comprising $229$ local authorities and over $40$ million
@@
 ```{r echo=FALSE, message=FALSE, results = "asis"}
-path.LAs_Clusters <- paste0(path, "/data/temp/LAs_Clusters.RData")
+path.LAs_Clusters <- paste0(path, "/data/temp/LAs_Clusters_k9.RData")
 load(path.LAs_Clusters)
@@
                              "administrative jobs, 2020",
-                             "leisure jobs, 2020",
+                             "service jobs, 2020",
                              "machine operation jobs, 2020",
@@
                              "n. business est. per hab., 2019",
-                             "NVQ4+",
+                             "furloughed per hab., 2020",
                              "AM tests per hab., 2020",
@@
 demand for telecommuting and operations such as video calls. Yet those in the 
-largest cluster, $6$, were clearly frustrated by lower upload speeds. 
+largest cluster, $6$, clearly experienced lower upload speeds. 
 Their speeds were also less reliable, perhaps because they had higher levels of
@@
 
+Early research on Covid-19 and cities speculated the changes that potential extensive
+post-Covid working from home patterns might generate for spatial structure: from 
+more walkable cities to more localised production and consumption patterns and from
+more extensive urban sprawl to decline in public transportation and increased private 
+car usage [@batty2020editorial]. Despite the essential role it played during the 
+pandemic, less effort has been spent in understanding the current and future role
+of digital infrastructure. Our results indicate that probably for the first time 
+the future of cities and spatial structure are so much intertwined with digital 
+infrastructure. If the post-covid world is a world with extensive working from home, 
+then we need to build, among other things, resilient digital infrastructure capable 
+of bridging the first layer of the digital divide. Contrary to previous broadband 
+business and deployment plans, emphasis should be placed not only on download,
+but also on upload speeds [@brake2020lessons. In essence, Covid-19 accentuated the 
+old, but more valid than ever argument: digital infrastructure, just like any other 
+network infrastructure, only becomes visible when it stops working [@star1999; @tranos2013geography].
+
 <!-- # Acknowledgement(s) {-} -->
@@
 # Cluster table
-clusters <- paste(path, "/data/temp/clusters_nodiff.csv", sep = "")
+clusters <- paste(path, "/data/temp/clusters_nodiff_9.csv", sep = "")
 clusters <- read_csv(clusters)
@@
 cluster9 <- as.vector(unlist(clusters[9]))
-cluster10 <- as.vector(unlist(clusters[10]))
-cluster11 <- as.vector(unlist(clusters[11]))
-cluster12 <- as.vector(unlist(clusters[12]))
-cluster13 <- as.vector(unlist(clusters[13]))
+# cluster10 <- as.vector(unlist(clusters[10]))
+# cluster11 <- as.vector(unlist(clusters[11]))
+# cluster12 <- as.vector(unlist(clusters[12]))
+# cluster13 <- as.vector(unlist(clusters[13]))
 ```
@@
 
-**Cluster 10: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster10, type="cmd2"), collapse=", "))
-```
-
-**Cluster 11: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster11, type="cmd2"), collapse=", "))
-```
-
-**Cluster 12: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster12, type="cmd2"), collapse=", "))
-```
-
-**Cluster 13: **
-```{r echo=FALSE, message=FALSE, results = "asis"}
-cat(paste(shQuote(cluster13, type="cmd2"), collapse=", "))
-```
+<!-- **Cluster 10: ** -->
+<!-- ```{r echo=FALSE, message=FALSE, results = "asis"} -->
+<!-- cat(paste(shQuote(cluster10, type="cmd2"), collapse=", ")) -->
+<!-- ``` -->
+
+<!-- **Cluster 11: ** -->
+<!-- ```{r echo=FALSE, message=FALSE, results = "asis"} -->
+<!-- cat(paste(shQuote(cluster11, type="cmd2"), collapse=", ")) -->
+<!-- ``` -->
+
+<!-- **Cluster 12: ** -->
+<!-- ```{r echo=FALSE, message=FALSE, results = "asis"} -->
+<!-- cat(paste(shQuote(cluster12, type="cmd2"), collapse=", ")) -->
+<!-- ``` -->
+
+<!-- **Cluster 13: ** -->
+<!-- ```{r echo=FALSE, message=FALSE, results = "asis"} -->
+<!-- cat(paste(shQuote(cluster13, type="cmd2"), collapse=", ")) -->
+<!-- ``` -->
 
@@
 # Descriptive statistics table
-path.LAs_Clusters <- paste0(path, "/data/temp/LAs_Clusters.RData")
+path.LAs_Clusters <- paste0(path, "/data/temp/LAs_Clusters_k9.RData")
 load(path.LAs_Clusters)
@@
                 admin, caring, plant,
-                earnings, nvq4,
+                earnings, furlough,
                 virgin,
@@
   mutate('total.busi/pop' = total.busi/pop,
-         'LAD.tests.am20/pop' = LAD.tests.am20/pop) %>%
-  dplyr::select(-total.busi, -LAD.tests.am20)
+         'LAD.tests.am20/pop' = LAD.tests.am20/pop,
+         'furlough/pop' = furlough/pop) %>%
+  dplyr::select(-total.busi, -LAD.tests.am20, -furlough)
 
@@
                              "administrative jobs, 2020",
-                             "leisure jobs, 2020",
+                             "service jobs, 2020",
                              "machine operation jobs, 2020",
                              "earnings, 2019",
-                             "NVQ4+",
                              "Virgin Media \\%, 2020",
                              "n. business est. per hab., 2019",
-                             "AM tests per hab., 2020"),
+                             "AM tests per hab., 2020",
+                             "furloughed per hab., 2020"),
           rownames = F,
@@
           title = "Descriptive statistics for the auxiliary regression explanatory variables\\label{descr.aux}")
-
-```
+```

### 2021-03-22 paper/v2_taylor_francis/broadband.speed.covid.tex
ET changes / responses to reviewer comments

@@
 for this pandemic-generated divide has been viewed as mainly
-occupational, here we consider whether it is also technological and,
-consequently, geographical.
+occupational, here we consider whether it is also technological and
+geographical.
+
+The capability to work from home has previously been studied from the
+perspective of whether work tasks in a given occupation both can be and
+are allowed to be performed using digital technologies independently of
+location or co-location with colleagues, including supervisors
+\citep{allen2015effective, singh2013modeling}. However, successful
+telecommuting also requires that the quality and reliability of digital
+services, particularly home internet connection speeds, enable the
+completion of work tasks with a minimum of delay or interruption. High
+levels of demand are one of the main causes of reduced reliability and
+slower connection speeds, as network bandwidth becomes congested. Yet
+prior to the pandemic, the performance of broadband services with
+respect to telecommuters was never tested at scale, as working from home
+and connecting to colleagues and workplace resources via the internet
+was the purview of a small minority of workers. Instead, leisure use in
+the evening, when video streaming services are at their peak, has been
+used to benchmark broadband performance and service delivery by
+different Internet Service Providers (ISPs) \citep{ofcom2017}.
+
+The shift towards telecommuting during various stages of lockdown around
+the world has been drastic and there are speculations that post-Covid,
+the tendency to work from home will be much higher, raising questions
+around whether internet services can accommodate the increased demand.
+For example, \(47\)\% of people in employment in the UK worked solely
+from home in April \(2020\), whilst the same figure only reached \(5\)\%
+the year before \citep{ons2020, ons2020lm2019}. Similar figures have
+been reported for other countries \citep{felstead2020homeworking}.
+Approximately \(37\)\% of the European workforce worked from home in
+April \(2020\) with countries like Finland reaching \(60\)\%
+\citep{eurofound2020}. In the US, almost half of the working population
+worked from home during the same period because of the pandemic
+\citep{brynjolfsson2020covid}, and a recent estimate indicated that
+\(37\)\% of all jobs in the US can be permanently performed entirely
+from home \citep{NBERw26948}. None of these changes could have happened
+in the absence of reliable information and communication technology
+(ICT) infrastructure -- both in terms of software and hardware. But
+while software innovations are easily diffused across space and
+society\footnote{See for example the huge success of videoconferencing
+  apps such as Zoom \citep{marks2020zoom}.}, the same does not apply for
+ICT hardware infrastructure such as internet broadband connectivity.
 
@@
 characteristics to explore how they overlap with or diverge from the
-existing economic and digital geography of the UK. Our analysis enables
-us to better understand how the spatial and social distribution of
-education, occupation and online accessibility intersect to enable or
-hinder the practice of telecommuting at a time of extreme demand. We
-also consider what lessons can be learned from this time for a future
-where telecommuting is likely to remain a more common means of accessing
-work, at least in comparison to the pre-Covid era.
-
-The capability to work from home has previously been studied from the
-perspective of whether work tasks in a given occupation both can be and
-are allowed to be performed using digital technologies independently of
-location or co-location with colleagues, including supervisors
-\citep{allen2015effective, singh2013modeling}. However, successful
-telecommuting also requires that the quality and reliability of digital
-services, particularly home internet connection speeds, enable the
-completion of work tasks with a minimum of delay or interruption. Prior
-to the pandemic, the performance of broadband services with respect to
-telecommuters was never tested at scale, as working from home and
-connecting to colleagues and workplace resources via the internet was
-the purview of a small minority of workers. Instead, leisure use in the
-evening, when video streaming services are at their peak, has been used
-to benchmark broadband performance and service delivery by different
-Internet Service Providers (ISPs) \citep{ofcom2017}.
-
-Yet the shift towards telecommuting during various stages of lockdown
-around the world has been drastic and there are speculations that
-post-Covid, the tendency to work from home will be much higher, raising
-questions around whether internet services can accommodate the increased
-demand. For example, \(47\)\% of people in employment in the UK worked
-solely from home in April \(2020\), whilst the same figure only reached
-\(5\)\% the year before \citep{ons2020, ons2020lm2019}. A back of the
-envelope calculation suggests that up to 40\% of the labour force could
-work from home on an ongoing basis \citep{batty2020editorial}. Similar
-figures have been reported for other countries
-\citep{felstead2020homeworking}. Approximately \(37\)\% of the European
-workforce worked from home in April \(2020\) with countries like Finland
-reaching \(60\)\% \citep{eurofound2020}. In the US, almost half of the
-working population worked from home during the same period because of
-the pandemic \citep{brynjolfsson2020covid}, and a recent estimate
-indicated that \(37\)\% of all jobs in the US can be permanently
-performed entirely from home \citep{NBERw26948}. None of these changes
-could have happened in the absence of reliable information and
-communication technology (ICT) infrastructure -- both in terms of
-software and hardware. But while software innovations are easily
-diffused across space and society\footnote{See for example the huge
-  success of videoconferencing apps such as Zoom \citep{marks2020zoom}.},
-the same does not apply for ICT hardware infrastructure such as internet
-broadband connectivity.
-
-The literature describes first level digital divides in terms of the
-availability and quality of internet connectivity, such as that manifest
-in different geographies in the UK
-\citep{riddlesden2014broadband, philip2017digital}. Second level digital
-divides consider the presence or lack of the necessary skills to
-effectively utilise digital technologies and the internet
-\citep{blank2014dimensions, van2011internet}. The third level focuses on
-the heterogenous returns of internet usage among different socioeconomic
-groups and, consequently, how digital technologies can assist in
-bridging or further enhancing existing socioeconomic divides.
-\citep{stern2009levels, van2014digital, van2015third}. The capability to
-telecommute is related to all three levels of digital divides, but more
-importantly leads to differentiated outcomes regarding the economic
-resilience of people and places to overcome a systemic shock such as the
-current pandemic.
-
-As the quality of internet infrastructure and services, as well as the
-concentration of different occupations are spatially dependent and
-clustered in space, our approach offers a framework for understanding
-the impact of and interactions between digital divisions geographically
-and socioeconomically. By asking how resilient broadband speeds, and
-particularly upload speeds, are as experienced in different parts of the
-UK during a time of extreme demand, we interrogate which places benefit
-from the greater economic resilience digital technologies can offer. The
-structure of this paper is as follows. First we review the literature on
+existing economic and digital geography of the UK. As the quality of
+internet infrastructure and services, as well as the concentration of
+different occupations are spatially dependent and clustered in space,
+our approach offers a framework for understanding the impact of and
+interactions between digital divisions geographically and
+socioeconomically. Our analysis enables us to better understand how the
+spatial and social distribution of education, occupation and online
+accessibility intersect to enable or hinder the practice of
+telecommuting at a time of extreme demand. We also consider what lessons
+can be learned from this time for a future where telecommuting is likely
+to remain a more common means of accessing work, at least in comparison
+to the pre-Covid era.
+
+By asking how resilient broadband speeds, and particularly upload
+speeds, are as experienced in different parts of the UK during a time of
+extreme demand, we interrogate which places benefit from the greater
+economic resilience digital technologies can offer. The structure of
+this paper is as follows. First we review the literature on
 telecommuting and digital divides to better understand their structural
@@
 
+The literature describes first level digital divides in terms of the
+availability and quality of internet connectivity, such as that manifest
+in different geographies in the UK
+\citep{riddlesden2014broadband, philip2017digital}. Second level digital
+divides consider the presence or lack of the necessary skills to
+effectively utilise digital technologies and the internet
+\citep{blank2014dimensions, van2011internet}. The third level focuses on
+the heterogenous returns of internet usage among different socioeconomic
+groups and, consequently, how digital technologies can assist in
+bridging or further enhancing existing socioeconomic divides.
+\citep{stern2009levels, van2014digital, van2015third}. The capability to
+telecommute is related to all three levels of digital divides, but more
+importantly leads to differentiated outcomes regarding the economic
+resilience of people and places to overcome a systemic shock such as the
+current pandemic.
+
 Our understanding of telecommuting as a product of enabled occupations
@@
 \citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
-These returns have been even greater during the Covid-19 crisis, when
-those with the capability to telecommute also have the ability to
-maintain their employment whilst protecting their health. However, the
-success of these arrangements has been dependent upon the first level
-digital divide, which is associated with access to and quality of
-internet connectivity. \citet{SALEMINK2017360} provides a systematic
-review of the pre-pandemic, first level digital divide in infrastructure
-quality between urban and rural areas in various advanced economies.
-Rural areas, predictably, fare worse, yet as \citet{blank2018local}
-highlight, variation in individual internet uptake and use is a product
-of more complex spatial and demographic characteristics than simple
-rurality or urbanisation. Yet whether this variation in infrastructure
-quality affects the spatial footprint of telecommuting has not
-previously been investigated, in part because telecommuting has not
-previously been a cause of concentrated demand and pressure on internet
-services.
-
-There are indications that those who purchase high speed connections
-consume more data of all sorts and use their connections for a greater
-variety of purposes \citep{hauge2011consumer}. There is also a
-correlation between access to internet services and a reduction in
-household transport spend \citep{bris2017ict}. Independently of whether
-the additional internet use and reduced travel is because of increased
-telecommuting, these studies suggest that better internet services
+In some aspects, these returns may have increased during the Covid-19
+crisis, when those with the capability to telecommute also have the
+ability to maintain their employment whilst protecting their health.
+However, the success of these arrangements has been dependent upon the
+first level digital divide, which is associated with access to and
+quality of internet connectivity. \citet{SALEMINK2017360} provides a
+systematic review of the pre-pandemic, first level digital divide in
+infrastructure quality between urban and rural areas in various advanced
+economies. Rural areas, predictably, fare worse. This finding broadly
+applies in the UK, where studies that also analysed broadband speed
+checks concluded that average speeds are lower in rural areas, something
+that has not been improved by policy measures to increase competition
+\citep{riddlesden2014broadband, nardotto2015unbundling}. Moreover,
+\citet{riddlesden2014broadband} found that levels of deprivation did not
+correlate with first level digital divides.
+
+However, as \citet{blank2018local} highlight, variation in individual
+internet uptake and use is a product of more complex spatial and
+demographic characteristics than simple rurality or urbanisation. Dense
+urban areas were shown to suffer more from slowdown during peak hours,
+although these services were more likely to be improved by increased
+competition between providers, such as between new entrants and Virgin
+Media cable connections
+\citep{riddlesden2014broadband, nardotto2015unbundling}. The latter were
+historically available to only \(45\)\% of premises in the UK
+\citep{ofcom2016}, where the more lucrative and competitive market
+originally attracted the cable TV provider. Whether the variation in
+infrastructure quality and reliability affects the spatial footprint of
+telecommuting has not previously been investigated, in part because
+telecommuting has not previously been a cause of concentrated demand and
+pressure on internet services. Other intersections between digital
+divides have been subject to study. There are indications that those who
+purchase high speed connections consume more data of all sorts and use
+their connections for a greater variety of purposes
+\citep{hauge2011consumer}. There is also a correlation between access to
+internet services and a reduction in household transport spend
+\citep{bris2017ict}. These studies suggest that better internet services
 enable households to make savings and efficiencies, an example of the
 first level digital divide reinforcing the third level.
+
+Currently there is a hotly debated discussion in the literature
+regarding these exact changes in transportation and the structure of
+cities. We now have enough hard data to observe the drastic change in
+the space-time geography of cities around the world
+\citep[e.g.~][]{google2020, shibayama2021impact}. Although there is a
+broad agreement that these changes during the pandemic have played a
+pivotal role in stopping the spread of the virus
+\citep{jia2020population, yang2020taking, mu2020interplay}, we still
+have no evidence to what extend these increased levels of working from
+home, and the consequent decrease in commuting flows and the structure
+of cities, will remain post-pandemic.
 
@@
 to assess the internet speed \emph{experienced} by users, which may
-differ from the maximum speeds \emph{advertised} by ISPs. We are
-particularly interested in upload speeds and the frequency of speed
-tests over the period from March to May \(2020\), as government
-statements indicate this is when UK workers were first told to work from
-home if at all possible \citep{GovUK2020}. Average upload speeds are
-slower than average download speeds, at \(9.3\)Mb/s mean upload speed
-for the whole sample, compared to \(29.6\)Mb/s for download speeds, but
-they are also less associated with internet-based, high-demand, leisure
-activities such as video streaming. Therefore, upload speeds are more
-relevant to work-related activities such as uploading documents or
-two-way audio, video, and text-based communication systems.
+differ from the maximum speeds \emph{advertised} by ISPs. Indeed, those
+who seek to test their broadband are most likely to do so because they
+are experiencing slower speeds than expected, although any skew towards
+slower speeds is balanced by the likelihood that those who test their
+broadband are also more `tech-savvy' and / or have purchased higher
+speed packages that are not delivering the promised level of service.
+
+We are particularly interested in upload speeds and the frequency of
+speed tests over the period from March to May \(2020\), as government
+statements indicate this encompasses the period when UK workers were
+told to work from home if at all possible \citep{GovUK2020}. Schools and
+various retail, leisure and hospitality businesses were closed from late
+March, and restrictions were gradually eased from late May. Average
+upload speeds are slower than average download speeds, at \(9.3\)Mb/s
+mean upload speed for the whole sample, compared to \(29.6\)Mb/s for
+download speeds, but they are also less associated with internet-based,
+high-demand, leisure activities such as video streaming. Therefore,
+upload speeds are more relevant to work-related activities such as
+uploading documents or two-way audio, video, and text-based
+communication systems.
 
@@
 However, there were insufficient observations -- only \(631\) speed
-tests per LAD on average -- for each for each working hour of each
-working day in each Local Authority District (LAD) to profile speeds at
-that level of detail. Spatial aggregation was also necessary because we
-could not follow individuals or households and connect data points.
-Therefore, we aggregate the \(241,088\) individual, geolocated and
-time-stamped speed-checks during the \(13\) weeks of March to May
-inclusive for weekdays in \(2020\) by each hour of the day and day of
-the week. As our research aims to identify the geography of internet
-service resilience for work purposes, bank holidays and the hours
-between midnight and \(6:00\) were excluded, as well as weekend days.
-The composite week time-series thus comprise \(18\) hours multiplied by
-\(5\) weekdays or \(90\) time points per series. The time-series were
-calculated for each of the \(382\) LADs in the UK, standardised, and
-then a \emph{k}-means partitioning around medoids clustering algorithm
-was applied using DTW. We initially run the algorithm for
-\(k \in \mathbb{N} \bigcap [5,15]\) and used cluster validity indices
-(CVIs) to pick the optimal solution of \(k = 13\). Following
-\citet{sardatime} the majority vote for the following CVIs was used:
-Silhouette (max), Score function (max), Calinski-Harabasz (max),
-Davies-Bouldin (min), Modified Davies-Bouldin (DB*, min), Dunn (max),
-COP (min).
+tests per Local Authority District (LAD) on average -- for each for each
+working hour of each working day in each LAD to profile speeds at that
+level of detail. Spatial aggregation was also necessary because we could
+not follow individuals or households and connect data points. Therefore,
+we aggregate the \(241,088\) individual, geolocated and time-stamped
+speed-checks during the \(13\) weeks of March to May inclusive for
+weekdays in \(2020\) by each hour of the day and day of the week. As our
+research aims to identify the geography of internet service resilience
+for work purposes, bank holidays and the hours between midnight and
+\(6:00\) were excluded, as well as weekend days. The composite week
+time-series thus comprise \(18\) hours multiplied by \(5\) weekdays or
+\(90\) time points per series. The time-series were calculated for each
+of the \(382\) LADs in the UK, standardised, and then a \emph{k}-means
+partitioning around medoids clustering algorithm was applied using DTW.
+We initially run the algorithm for \(k \in \mathbb{N} \bigcap [5,15]\),
+calculated the cluster validity indices (CVIs) and then run the
+subsequent multinomial regression -- see the end of this section for
+more details for the different \emph{k}. Following \citet{sardatime}, to
+identify the optimal \emph{k} we used the majority vote for the
+following CVIs: Silhouette (max), Score function (max),
+Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin
+(DB*, min), Dunn (max), COP (min). Nevertheless, we opted against using
+the optimal \(k = 13\) solution as it was too large to allow for
+communicable LAD clusters. Instead we opted for a smaller \(k = 9\),
+which led to a rather similar spatial pattern and, importantly, to a
+much higher R-squared in the subsequent explanatory regression (\(0.44\)
+instead of \(0.34\)).
 
@@
 profile of upload speed by hour of the day and day of the composite week
-for each cluster. Since the quality and reliability of internet services
-vary in time and space due to both supply and demand-side influences, we
-also use a number of different measures to describe experienced upload
-speeds per cluster. These include: i) mean, experienced connection
-speed, ii) standard deviation or the amount of fluctuation from the
-mean, and iii) the variation in speeds during the new morning peak of
-testing when working from home is more likely to take place. We take
-account of all three measurements in order to determine how resilient
-broadband speeds are as experienced in different parts of the UK during
-a time of extreme demand.
+for each of the clusters. Since the quality and reliability of internet
+services vary in time and space due to both supply and demand-side
+influences, we also use a number of different measures to describe
+experienced upload speeds per cluster. These include: i) mean,
+experienced connection speed, ii) standard deviation or the amount of
+fluctuation from the mean, and iii) the variation in speeds during the
+new morning peak of testing when working from home is more likely to
+take place. We take account of all three measurements in order to
+determine how resilient broadband speeds are as experienced in different
+parts of the UK during a time of extreme demand.
 
@@
 The temporal profiles of the local authority clusters have been
-summarised in Figures \ref{UpClusterL} and \ref{UpClusterS} and Table
-\ref{up.cluster.descr}. The graphs show a composite profile of mean
-upload speeds per hour per day for each cluster, with the largest five
-clusters, in terms of the LAD membership and population, in Figure
-\ref{UpClusterL}, and the next six in Figure \ref{UpClusterS}. These
-figures and table provide a comprehensive overview of the quality and
-reliability of experienced broadband in different parts of the UK.
+summarised in Figure \ref{UpCluster} and Table \ref{up.cluster.descr},
+to provide an overview of the quality and reliability of experienced
+broadband in different parts of the UK. Figure \ref{UpCluster} shows a
+composite profile of mean upload speeds per hour per day for each of the
+largest six clusters, in terms of the LAD membership and population.
 
 \begin{figure}
-\includegraphics[width=0.95\linewidth]{figures/upClusterL} \caption{\label{UpClusterL}Temporal profilies for upload speed large clusters}\label{fig:unnamed-chunk-2}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.95\linewidth]{figures/upClusterS} \caption{\label{UpClusterS}Temporal profilies for upload speed small clusters}\label{fig:unnamed-chunk-3}
+\includegraphics[width=0.95\linewidth]{figures/UpCluster} \caption{\label{UpCluster}Temporal profilies for upload speed clusters}\label{fig:unnamed-chunk-2}
 \end{figure}
@@
 \hline \\[-1.8ex] 
-1 & 5 & 343100 & 8557 & 6139 & 7747 & 9563 \\ 
-2 & 2 & 265600 & 10922 & 6687 & 9674 & 10645 \\ 
-3 & 4 & 474700 & 10201 & 5658 & 9470 & 11236 \\ 
+1 & 9 & 903200 & 9564 & 6314 & 8798 & 10457 \\ 
+2 & 2 & 162000 & 12085 & 6537 & 11882 & 10866 \\ 
+3 & 12 & 1785800 & 11047 & 6079 & 10029 & 11634 \\ 
 4 & 1 & 91100 & 9689 & 6122 & 7816 & 9689 \\ 
-5 & 1 & 79800 & 10127 & 6024 & 9030 & 11101 \\ 
-6 & 155 & 29535700 & 9397 & 5839 & 9161 & 9580 \\ 
-7 & 4 & 559800 & 10119 & 6102 & 9813 & 11070 \\ 
-8 & 5 & 436300 & 9429 & 6254 & 8682 & 10434 \\ 
-9 & 32 & 6355500 & 10878 & 5957 & 10832 & 11071 \\ 
-10 & 4 & 699600 & 10795 & 6005 & 9258 & 10697 \\ 
-11 & 33 & 5771400 & 10845 & 5936 & 10781 & 10988 \\ 
-12 & 10 & 1544900 & 9551 & 6166 & 9254 & 9048 \\ 
-13 & 126 & 20277700 & 8392 & 5849 & 8299 & 8522 \\ 
+5 & 3 & 280000 & 10802 & 6116 & 11010 & 10084 \\ 
+6 & 229 & 40552800 & 8761 & 5847 & 8555 & 8955 \\ 
+7 & 5 & 682500 & 10326 & 6102 & 10045 & 11149 \\ 
+8 & 6 & 510000 & 9769 & 6352 & 8989 & 10836 \\ 
+9 & 115 & 21467800 & 10328 & 5915 & 10283 & 10333 \\ 
 \hline \\[-1.8ex] 
@@
 
-The second largest cluster, comprising \(126\) local authorities and
-over \(20\) million people, is cluster \(13\), which has the slowest
-aggregate mean upload speed, and the second highest ratio of the
-standard deviation to the mean. This suggests that those living in local
+The largest cluster, comprising \(229\) local authorities and over
+\(40\) million people, is cluster \(6\), which has the slowest aggregate
+mean upload speed of any cluster, and the highest ratio of the standard
+deviation to the mean. This suggests that those living in local
 authorities in this cluster experienced some of the lowest quality
-broadband services in terms of upload speeds in the UK. However, Figure
-\ref{UpClusterL} shows that the high standard deviation, which is one
-indication of unreliable internet, did not disproportionately affect the
-morning peak from \(9:00\)-\(10:59\), when upload speeds were, on
-average, only \(2.6\)\% slower than in the evening peak period between
-\(19:00\) and \(20:59\), when entertainment purposes are likely to be
-using the most bandwidth. In comparison, the five LADs in cluster \(1\),
-home to \(343\) thousand people, not only experience the second slowest
-mean upload speeds and the highest ratio of standard deviation to the
-mean, but are also much more affected during the morning peak.
-
-Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
-LADs and \(29.5\) million people -- experienced aggregate mean upload
-speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
-lower than the other three large clusters and most of the smaller
-clusters, suggesting a middling quality of service. The temporal profile
-for cluster \(6\) in Figure \ref{UpClusterL} shows that upload speeds
-are highest at \(6:00\) on a Monday morning and between \(23:00\) and
-midnight on Wednesday and Thursday, but tend to be lower during the
-working day. Furthermore, experienced mean upload speeds in the morning
-peak are \(4.4\)\% lower than in the evening peak -- a greater, more
-noticeable change than any of the other large clusters experience,
-suggesting poorer reliability during the working day. This difference is
-less, however than any of the clusters included in Figure
-\ref{UpClusterS}.
-
-Clusters \(8\) and \(12\) also have mean upload speeds under \(10\)Mb/s,
-but higher than clusters \(1\), \(13\), and \(6\). However, mean upload
-speeds are much lower between \(9:00-10:59\) than between
-\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning
-than in the evening in cluster \(12\). Indeed, cluster \(12\) is the
-only cluster to experience higher speeds in the morning peak compared to
-the evening, and thus the only cluster where the temporal profile of
-internet use is closer to what might have been expected pre-pandemic.
-Among the other clusters, however, the reliability of internet services
-during the working day still varies considerably. Interpreting this
-variation from the large spikes and dips shown on Figures
-\ref{UpClusterL} and \ref{UpClusterS} is difficult, but the statistics
-in Table \ref{up.cluster.descr} show that clusters \(9\) and \(11\) have
-the most reliable internet services. The ratio of standard deviation to
-mean in both these clusters is lowest, and speeds are only about 2\%
-slower in the morning than the evening. Mean speeds are also higher than
-in any other cluster, excluding cluster \(2\), where measures of
-reliability suggest poorer performance.
-
-Thus, broadband services in clusters \(9\) and \(11\), home to over
-twelve million people, performed the best during the study period, in
-terms of both quality and reliability. In Figure \ref{UpClusterL},
-cluster \(11\) shows more noticeable peaks and troughs, with the lowest
-points occurring, on average, between \(6:00-7:00\) on Monday morning,
-\(14:00-15:00\) on Friday, and \(16:00-17:00\) on Thursday, not the
-workday peak. Furthermore, these slow periods offer better speeds than
-the average hourly profile of cluster \(13\). Clusters \(3\), \(7\) and
-\(10\) also have relatively high mean upload speeds. Clusters \(3\) and
-\(10\) pair high mean speeds with low standard deviations relative to
-the mean speeds, suggesting reliability and resilience, as well as
-quality broadband services. Cluster \(7\) has a higher ratio of standard
-deviation to mean, but there is less difference in average speeds
-between the morning and evening peaks than in clusters \(3\) and \(10\).
-
-In summary, LADs in clusters \(9\) and \(11\) experienced resilient
-broadband internet that could support high levels of telecommuting.
-Those in clusters \(2\), \(3\), \(7\), and \(10\) also experience higher
-than average mean speeds and rank middle to high on measures of service
-reliability. These LADs are \emph{not} on the wrong side of the first
-level digital divide, but how likely are they to be able to take
-advantage of their resilient ICT infrastructure and services? Meanwhile,
-cluster \(6\) is not only the largest in terms of number of LADs and
-population, it has the closest mean upload speed to the pre-clustered
-average for the whole sample. As well as average quality internet
-services, those in cluster \(6\) also experience average reliability for
-work purposes, ranking fifth behind the four other clusters with
-populations over one million, but ahead of the smaller clusters.
-Clusters \(8\) and \(12\) are also close to average mean upload speeds,
-but show very different patterns in terms of reliability, whilst
-clusters \(1\) and \(13\) appear to suffer most from a lack of quality
-internet services, with slow speeds and high standard deviations. With
-those in cluster \(1\) more likely to experience that poor reliability
-during the morning peak, is this first level digital divide occurring in
-areas where few are occupationally able to telecommute anyway, and what
-are the implications for economic resilience?
+broadband services in terms of upload speeds and reliability in the UK.
+However, as shown in Figure \ref{map.up.clusters}, some of the most
+rural areas of the UK are included in this cluster. If these areas
+suffer most from first level digital divides as described in the
+literature review, the low speeds in rural areas might be pulling down
+the averages in other areas in this largest cluster. Also, since the
+areas are clustered by their temporal profile across the working week,
+the graph in Figure \ref{UpCluster} indicates that speeds in cluster
+\(6\) are some of the more reliable. Table \ref{up.cluster.descr}
+confirms that upload speeds in cluster \(6\) during the morning peak
+from \(9:00\)-\(10:59\) were, on average, only \(4.5\)\% slower than in
+the evening peak period between \(19:00\) and \(20:59\), when
+entertainment purposes are likely to be using the most bandwidth.
+
+\begin{figure}
+\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-4}
+\end{figure}
+
+Those living in the second largest cluster -- \(9\), with \(115\) LADs
+and \(21.5\) million people -- experienced aggregate mean upload speeds
+of more than \(1.5\)Mb/s faster than those in cluster \(6\), a lower
+ratio of standard deviation to mean, and almost the same upload speeds
+in the AM Peak as the PM Peak. Indeed, the temporal profile for cluster
+\(9\) in Figure \ref{UpCluster}, like that for cluster \(6\), is fairly
+flat. Whilst it is likely that the large numbers of tests being
+performed by the large populations in these two clusters of LADs result
+in less varied averages to show in the graph, the method of shape-based
+clustering suggests that these results indicate that the majority of the
+population of the UK experience less temporal variation, and so belong
+in these two clusters. The results also confirm that those in cluster
+\(9\) experienced a consistently better service during lockdown than
+those in cluster \(6\), in terms of both average speed and reliability.
+
+Of the smaller clusters, cluster \(7\) experiences quite similar mean
+speeds to cluster \(9\), and a not too dissimilar standard deviation,
+but mean speeds in the morning peak are almost \(10\)\% lower than the
+evening peak. Cluster \(3\) boasts speeds almost \(1\)Mb/s higher again,
+but suffers from even greater slowdown in the morning peak. In
+comparison, clusters \(2\) and \(5\), home to a little less than half a
+million people, experience not only above average mean speeds, but also
+faster speeds in the morning peak compared to the evening. Thus, in the
+five LADs in these two clusters, the temporal profile of internet use
+may be closer to what might have been expected pre-pandemic. In
+comparison, the \(15\) LADs in clusters \(1\) and \(8\), home to over
+\(1.4\) million people, not only experience fairly average upload speeds
+and high ratios of standard deviation to the mean, but also experience
+much lower speeds during the morning peak than the evening peak. From
+the large spikes and dips shown on Figure \ref{UpCluster}, it is likely
+that this poor reliability or consistency of morning internet speeds was
+more noticeable to those in cluster \(1\), whilst those in cluster \(8\)
+experienced the lack of reliable service as a problem throughout the
+day.
+
+In summary, the reliability of internet services during the working day
+appears to have altered for the vast majority of locations due to
+increased use from residents told to stay at home, and, if they were
+fortunate, work from home. Yet this experience has been different in
+different locations. In particular, LADs in cluster \(9\) experienced
+higher speeds and more resilient broadband internet than those in the
+largest cluster, \(6\). Those in clusters \(3\) and \(7\) also
+experience higher mean speeds and better service reliability, and can be
+more confident that they are on the right side of the first level
+digital divide, and that their more resilient ICT infrastructure and
+services can robustly support higher levels of telecommuting. On the
+other hand, whilst high levels of telecommuting may be the cause of
+morning speeds over 15\% lower than those in the evening, those in
+clusters \(1\) and \(8\) also experience lower mean speeds than clusters
+\(3\), \(7\), and \(9\), but not as low as cluster \(6\). Whilst the
+differences may not be large, the lower upload speeds in cluster \(6\)
+or the poor reliability in clusters \(1\) and \(8\) could still have had
+implications for economic resilience if combined with other types of
+digital divide. This potential will be explored in the next section.
 
@@
 \tiny 
-\begin{tabular}{@{\extracolsep{5pt}}lccccccccccc} 
+\begin{tabular}{@{\extracolsep{5pt}}lcccccccc} 
 \\[-1.8ex]\hline 
 \hline \\[-1.8ex] 
-\\[-1.8ex] & 1 & 2 & 3 & 6 & 7 & 8 & 9 & 10 & 11 & 12 & 13 \\ 
-\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8) & (9) & (10) & (11)\\ 
+\\[-1.8ex] & 1 & 2 & 3 & 5 & 6 & 7 & 8 & 9 \\ 
+\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8)\\ 
 \hline \\[-1.8ex] 
- pop, 2018 & $-$0.00004$^{***}$ & 0.00002$^{*}$ & 0.00001 & 0.00002$^{***}$ & 0.00002$^{***}$ & 0.00000 & 0.00002$^{***}$ & 0.00002$^{***}$ & 0.00002$^{***}$ & 0.00002$^{**}$ & 0.00002$^{***}$ \\ 
-  & (0.00002) & (0.00001) & (0.00001) & (0.00001) & (0.00001) & (0.00002) & (0.00001) & (0.00001) & (0.00001) & (0.00001) & (0.00001) \\ 
-  & & & & & & & & & & & \\ 
- job density, 2018 & $-$0.536$^{***}$ & $-$1.834$^{***}$ & $-$0.132$^{***}$ & $-$0.925$^{***}$ & $-$1.208$^{***}$ & $-$0.299$^{***}$ & $-$1.746$^{***}$ & $-$1.436$^{***}$ & 3.350$^{***}$ & 3.400$^{***}$ & 0.630$^{***}$ \\
-  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
-  & & & & & & & & & & & \\ 
- distance to nearest met.
-area & $-$0.034$^{***}$ & $-$0.014$^{***}$ & 0.002$^{***}$ & $-$0.020$^{***}$ & $-$0.074$^{***}$ & $-$0.044$^{***}$ & $-$0.013$^{***}$ & $-$0.036$^{***}$ & $-$0.031$^{***}$ & $-$0.036$^{***}$ & $-$0.024$^{***}$ \\
-  & (0.0005) & (0.001) & (0.0002) & (0.002) & (0.0001) & (0.0002) & (0.002) & (0.0005) & (0.0003) & (0.0002) & (0.002) \\ 
-  & & & & & & & & & & & \\ 
- distance to London & 0.007$^{***}$ & 0.002 & $-$0.016$^{***}$ & 0.001 & 0.004$^{***}$ & 0.004$^{***}$ & $-$0.002$^{*}$ & 0.005$^{**}$ & $-$0.002 & 0.003 & 0.006$^{***}$ \\ 
-  & (0.001) & (0.002) & (0.0004) & (0.001) & (0.001) & (0.001) & (0.001) & (0.002) & (0.002) & (0.002) & (0.001) \\ 
-  & & & & & & & & & & & \\ 
- South of the UK & $-$0.410$^{***}$ & $-$1.451$^{***}$ & $-$0.039$^{***}$ & $-$0.111$^{***}$ & $-$0.048$^{***}$ & $-$0.841$^{***}$ & $-$0.798$^{***}$ & 1.492$^{***}$ & 0.610$^{***}$ & 0.798$^{***}$ & 2.403$^{***}$ \\
-  & (0.00000) & (0.00000) & (0.00000) & (0.00001) & (0.00000) & (0.00000) & (0.00001) & (0.00000) & (0.00001) & (0.00001) & (0.00001) \\ 
-  & & & & & & & & & & & \\ 
- managerial jobs, 2020 & 0.939$^{***}$ & 0.704$^{***}$ & 0.435$^{***}$ & 0.704$^{***}$ & 0.316$^{***}$ & 0.786$^{***}$ & 0.576$^{***}$ & 0.311$^{***}$ & 0.476$^{***}$ & 0.594$^{***}$ & 0.615$^{***}$ \\
-  & (0.00004) & (0.0001) & (0.00004) & (0.00004) & (0.00002) & (0.0001) & (0.00003) & (0.00003) & (0.00002) & (0.00004) & (0.00003) \\ 
-  & & & & & & & & & & & \\ 
- tech jobs, 2020 & 0.096$^{***}$ & $-$0.257$^{***}$ & $-$0.071$^{***}$ & $-$0.111$^{***}$ & $-$0.206$^{***}$ & 0.199$^{***}$ & $-$0.126$^{***}$ & $-$0.606$^{***}$ & $-$0.180$^{***}$ & $-$0.398$^{***}$ & $-$0.112$^{***}$ \\
-  & (0.00004) & (0.00004) & (0.00004) & (0.00003) & (0.00003) & (0.0001) & (0.00003) & (0.00003) & (0.00002) & (0.00004) & (0.00003) \\ 
-  & & & & & & & & & & & \\ 
- skilled trade jobs, 2020 & 0.651$^{***}$ & 0.160$^{***}$ & $-$0.191$^{***}$ & 0.236$^{***}$ & 0.604$^{***}$ & $-$0.184$^{***}$ & 0.205$^{***}$ & 0.597$^{***}$ & 0.108$^{***}$ & $-$0.022$^{***}$ & 0.295$^{***}$ \\
-  & (0.00004) & (0.00004) & (0.00003) & (0.00003) & (0.00003) & (0.00004) & (0.00002) & (0.00003) & (0.00003) & (0.00004) & (0.00003) \\ 
-  & & & & & & & & & & & \\ 
- professional jobs, 2020 & $-$0.118$^{***}$ & $-$0.234$^{***}$ & $-$0.121$^{***}$ & $-$0.172$^{***}$ & $-$0.514$^{***}$ & $-$0.172$^{***}$ & $-$0.349$^{***}$ & $-$0.351$^{***}$ & $-$0.245$^{***}$ & $-$0.344$^{***}$ & $-$0.229$^{***}$ \\
-  & (0.00005) & (0.0001) & (0.0001) & (0.00005) & (0.00003) & (0.0001) & (0.00004) & (0.0001) & (0.00003) & (0.00005) & (0.0001) \\ 
-  & & & & & & & & & & & \\ 
- administrative jobs, 2020 & 0.019$^{***}$ & $-$0.836$^{***}$ & $-$0.040$^{***}$ & $-$0.117$^{***}$ & $-$0.139$^{***}$ & 0.206$^{***}$ & $-$0.058$^{***}$ & $-$0.200$^{***}$ & $-$0.055$^{***}$ & $-$0.168$^{***}$ & $-$0.177$^{***}$ \\
-  & (0.00003) & (0.00002) & (0.00004) & (0.00001) & (0.00003) & (0.00003) & (0.00001) & (0.00002) & (0.00002) & (0.00002) & (0.00002) \\ 
-  & & & & & & & & & & & \\ 
- leisure jobs, 2020 & $-$0.198$^{***}$ & $-$0.180$^{***}$ & $-$0.225$^{***}$ & $-$0.476$^{***}$ & $-$0.654$^{***}$ & $-$0.820$^{***}$ & $-$0.537$^{***}$ & $-$0.935$^{***}$ & $-$0.353$^{***}$ & $-$0.625$^{***}$ & $-$0.491$^{***}$ \\
-  & (0.00002) & (0.00004) & (0.00004) & (0.00002) & (0.00002) & (0.00003) & (0.00002) & (0.00001) & (0.00002) & (0.00003) & (0.00002) \\ 
-  & & & & & & & & & & & \\ 
- machine operation jobs, 2020 & $-$0.336$^{***}$ & 0.207$^{***}$ & 0.392$^{***}$ & 0.010$^{***}$ & $-$0.433$^{***}$ & 0.139$^{***}$ & $-$0.099$^{***}$ & $-$0.139$^{***}$ & $-$0.144$^{***}$ & 0.098$^{***}$ & $-$0.179$^{***}$ \\
-  & (0.00002) & (0.00003) & (0.00003) & (0.00002) & (0.00001) & (0.00003) & (0.00002) & (0.00001) & (0.00001) & (0.00002) & (0.00001) \\ 
-  & & & & & & & & & & & \\ 
- earnings, 2019 & $-$0.003$^{*}$ & 0.010$^{***}$ & 0.012$^{***}$ & 0.020$^{***}$ & 0.027$^{***}$ & 0.001 & 0.020$^{***}$ & 0.016$^{***}$ & 0.015$^{***}$ & 0.025$^{***}$ & 0.014$^{***}$ \\ 
-  & (0.002) & (0.002) & (0.002) & (0.001) & (0.001) & (0.003) & (0.001) & (0.002) & (0.001) & (0.001) & (0.001) \\ 
-  & & & & & & & & & & & \\ 
- n.
-business est.
-per hab., 2019 & 0.126$^{***}$ & $-$0.120$^{***}$ & $-$0.094$^{***}$ & $-$0.133$^{***}$ & 0.123$^{***}$ & $-$0.051$^{***}$ & $-$0.334$^{***}$ & $-$0.133$^{***}$ & $-$0.150$^{***}$ & 0.289$^{***}$ & 0.377$^{***}$ \\
-  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
-  & & & & & & & & & & & \\ 
- NVQ4+ & $-$0.141$^{***}$ & 0.064$^{***}$ & $-$0.091$^{***}$ & $-$0.070$^{***}$ & $-$0.010$^{***}$ & 0.004$^{***}$ & 0.016$^{***}$ & 0.170$^{***}$ & $-$0.110$^{***}$ & $-$0.038$^{***}$ & $-$0.035$^{***}$ \\
-  & (0.0001) & (0.0001) & (0.0001) & (0.0001) & (0.0001) & (0.0002) & (0.0001) & (0.0001) & (0.0001) & (0.0001) & (0.0001) \\ 
-  & & & & & & & & & & & \\ 
- AM tests per hab., 2020 & 0.0005$^{***}$ & $-$0.002$^{***}$ & $-$0.005$^{***}$ & 0.010$^{***}$ & 0.0004$^{***}$ & $-$0.001$^{***}$ & $-$0.002$^{***}$ & $-$0.005$^{***}$ & $-$0.013$^{***}$ & $-$0.001$^{***}$ & 0.016$^{***}$ \\
-  & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) \\ 
-  & & & & & & & & & & & \\ 
- Virgin Media \%, 2020 & $-$0.044$^{***}$ & 1.578$^{***}$ & 1.210$^{***}$ & 0.248$^{***}$ & $-$1.724$^{***}$ & $-$0.242$^{***}$ & 3.109$^{***}$ & $-$0.085$^{***}$ & 1.214$^{***}$ & $-$3.889$^{***}$ & $-$0.745$^{***}$ \\
-  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
-  & & & & & & & & & & & \\ 
- Constant & 0.321$^{***}$ & $-$0.436$^{***}$ & 0.199$^{***}$ & $-$2.953$^{***}$ & 0.278$^{***}$ & 0.002$^{***}$ & $-$0.866$^{***}$ & 0.788$^{***}$ & 2.600$^{***}$ & 0.017$^{***}$ & 0.022$^{***}$ \\ 
-  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
-  & & & & & & & & & & & \\ 
+ pop, 2018 & $-$0.00004$^{**}$ & 0.00002 & 0.00001 & 0.00002 & 0.00001 & 0.00002 & $-$0.00000 & 0.00001 \\ 
+  & (0.00002) & (0.00001) & (0.00001) & (0.00002) & (0.00001) & (0.00001) & (0.00002) & (0.00001) \\ 
+  & & & & & & & & \\ 
+ job density, 2018 & $-$0.534$^{***}$ & $-$0.362$^{***}$ & 4.211$^{***}$ & $-$0.890$^{***}$ & 3.516$^{***}$ & $-$8.624$^{***}$ & $-$3.259$^{***}$ & 3.337$^{***}$ \\ 
+  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
+  & & & & & & & & \\ 
+ distance to nearest met. area & $-$0.007$^{***}$ & 0.013$^{***}$ & 0.002$^{***}$ & 0.023$^{***}$ & $-$0.001 & $-$0.158$^{***}$ & 0.005$^{***}$ & $-$0.003$^{**}$ \\ 
+  & (0.0004) & (0.00001) & (0.0003) & (0.0002) & (0.002) & (0.0001) & (0.0003) & (0.002) \\ 
+  & & & & & & & & \\ 
+ distance to London & 0.011$^{***}$ & 0.004$^{***}$ & 0.005$^{**}$ & $-$0.013$^{***}$ & 0.010$^{***}$ & 0.012$^{***}$ & 0.008$^{***}$ & 0.008$^{***}$ \\ 
+  & (0.001) & (0.00004) & (0.002) & (0.001) & (0.001) & (0.001) & (0.002) & (0.001) \\ 
+  & & & & & & & & \\ 
+ South of the UK & $-$3.628$^{***}$ & 0.958$^{***}$ & 4.936$^{***}$ & $-$10.243$^{***}$ & 4.830$^{***}$ & 3.919$^{***}$ & $-$1.140$^{***}$ & 3.188$^{***}$ \\ 
+  & (0.000) & (0.000) & (0.00000) & (0.00000) & (0.00001) & (0.00000) & (0.00000) & (0.00001) \\ 
+  & & & & & & & & \\ 
+ managerial jobs, 2020 & 0.944$^{***}$ & 0.628$^{***}$ & 0.593$^{***}$ & $-$0.544$^{***}$ & 0.634$^{***}$ & $-$0.128$^{***}$ & 0.725$^{***}$ & 0.511$^{***}$ \\ 
+  & (0.00002) & (0.00000) & (0.00003) & (0.00003) & (0.00002) & (0.00003) & (0.00003) & (0.00002) \\ 
+  & & & & & & & & \\ 
+ tech jobs, 2020 & 0.350$^{***}$ & $-$0.222$^{***}$ & $-$0.068$^{***}$ & $-$0.151$^{***}$ & 0.018$^{***}$ & $-$0.139$^{***}$ & 0.143$^{***}$ & $-$0.071$^{***}$ \\ 
+  & (0.00002) & (0.00000) & (0.00002) & (0.0001) & (0.00003) & (0.0001) & (0.00003) & (0.00003) \\ 
+  & & & & & & & & \\ 
+ skilled trade jobs, 2020 & 0.521$^{***}$ & 0.171$^{***}$ & 0.214$^{***}$ & $-$0.578$^{***}$ & 0.231$^{***}$ & 1.067$^{***}$ & 0.284$^{***}$ & 0.049$^{***}$ \\ 
+  & (0.00002) & (0.00000) & (0.00004) & (0.00004) & (0.00004) & (0.0001) & (0.00003) & (0.00003) \\ 
+  & & & & & & & & \\ 
+ professional jobs, 2020 & $-$0.577$^{***}$ & $-$0.962$^{***}$ & $-$0.565$^{***}$ & $-$0.860$^{***}$ & $-$0.566$^{***}$ & $-$1.139$^{***}$ & $-$0.389$^{***}$ & $-$0.702$^{***}$ \\ 
+  & (0.00003) & (0.00000) & (0.00004) & (0.0001) & (0.00004) & (0.0001) & (0.0001) & (0.00004) \\ 
+  & & & & & & & & \\ 
+ administrative jobs, 2020 & 0.200$^{***}$ & $-$1.324$^{***}$ & $-$0.050$^{***}$ & 0.136$^{***}$ & $-$0.137$^{***}$ & $-$0.101$^{***}$ & $-$0.018$^{***}$ & $-$0.155$^{***}$ \\ 
+  & (0.00002) & (0.00000) & (0.00002) & (0.00004) & (0.00002) & (0.00005) & (0.00002) & (0.00002) \\ 
+  & & & & & & & & \\ 
+ service jobs, 2020 & $-$0.082$^{***}$ & $-$0.180$^{***}$ & $-$0.659$^{***}$ & $-$1.066$^{***}$ & $-$0.746$^{***}$ & $-$0.944$^{***}$ & $-$0.773$^{***}$ & $-$0.783$^{***}$ \\ 
+  & (0.00002) & (0.00000) & (0.00003) & (0.00002) & (0.00002) & (0.00003) & (0.00002) & (0.00001) \\ 
+  & & & & & & & & \\ 
+ machine operation jobs, 2020 & 0.210$^{***}$ & 0.574$^{***}$ & 0.439$^{***}$ & $-$0.589$^{***}$ & 0.229$^{***}$ & $-$0.260$^{***}$ & 0.080$^{***}$ & 0.127$^{***}$ \\ 
+  & (0.00001) & (0.00000) & (0.00002) & (0.00002) & (0.00001) & (0.00002) & (0.00001) & (0.00001) \\ 
+  & & & & & & & & \\ 
+ earnings, 2019 & $-$0.003$^{***}$ & 0.029$^{***}$ & 0.019$^{***}$ & 0.069$^{***}$ & 0.018$^{***}$ & 0.050$^{***}$ & 0.016$^{***}$ & 0.026$^{***}$ \\ 
+  & (0.001) & (0.0001) & (0.001) & (0.002) & (0.001) & (0.002) & (0.001) & (0.001) \\ 
+  & & & & & & & & \\ 
+ n. business est. per hab., 2019 & 1.930$^{***}$ & $-$0.377$^{***}$ & 0.826$^{***}$ & 0.251$^{***}$ & 0.315$^{***}$ & 1.515$^{***}$ & $-$0.948$^{***}$ & $-$4.011$^{***}$ \\ 
+  & (0.00000) & (0.000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
+  & & & & & & & & \\ 
+ furloughed per hab., 2020 & $-$1.518$^{***}$ & $-$0.152$^{***}$ & 11.014$^{***}$ & 0.972$^{***}$ & $-$26.969$^{***}$ & 2.059$^{***}$ & 4.236$^{***}$ & 10.500$^{***}$ \\ 
+  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
+  & & & & & & & & \\ 
+ AM tests per hab., 2020 & 0.002$^{***}$ & $-$0.011$^{***}$ & $-$0.140$^{***}$ & 0.015$^{***}$ & 0.231$^{***}$ & 0.005$^{***}$ & $-$0.025$^{***}$ & $-$0.078$^{***}$ \\ 
+  & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) \\ 
+  & & & & & & & & \\ 
+ Virgin Media \%, 2020 & $-$0.349$^{***}$ & 12.236$^{***}$ & 5.619$^{***}$ & 0.489$^{***}$ & 1.721$^{***}$ & $-$16.289$^{***}$ & $-$5.581$^{***}$ & 6.126$^{***}$ \\ 
+  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
+  & & & & & & & & \\ 
+ Constant & $-$4.570$^{***}$ & $-$1.045$^{***}$ & $-$6.539$^{***}$ & 5.629$^{***}$ & 2.957$^{***}$ & 3.386$^{***}$ & $-$2.531$^{***}$ & 2.979$^{***}$ \\ 
+  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
+  & & & & & & & & \\ 
 \hline \\[-1.8ex] 
-McFadden's R squared & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 & 0.338 \\ 
-N & 323 & 323 & 323 & 323 & 323 & 323 & 323 & 323 & 323 & 323 & 323 \\ 
-Akaike Inf. Crit. & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 & 1,148.027 \\ 
+McFadden's R squared & 0.435 & 0.435 & 0.435 & 0.435 & 0.435 & 0.435 & 0.435 & 0.435 \\ 
+N & 322 & 322 & 322 & 322 & 322 & 322 & 322 & 322 \\ 
+Akaike Inf. Crit. & 742.901 & 742.901 & 742.901 & 742.901 & 742.901 & 742.901 & 742.901 & 742.901 \\ 
 \hline 
 \hline \\[-1.8ex] 
-\textit{Note:}  & \multicolumn{11}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
+\textit{Note:}  & \multicolumn{8}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
 \end{tabular} 
@@
 Using an auxiliary multinomial logit regression, we test whether the
-clusters that have higher mean speeds and more reliable services consist
-of LADs that are more urban, affluent, and / or more likely to benefit
-from a choice of high quality internet services. We also estimate which
-of our clusters are more likely to have a higher proportion of
-occupations where the nature of the work enables telecommuting. The
-results of the auxiliary regression are presented in Table \ref{aux}.
-The dependent variable is the LAD cluster membership as described in the
-\protect\hyperlink{sec:3}{methods and data} section and equation
-\ref{eq1}. Each column represents a different cluster. The reference
-case is cluster \(4\), which includes only the local authority of
-Hambleton in North Yorkshire, a rural area of just over ninety thousand
-people. Mean, experienced upload speeds in cluster \(4\) (see Table
-\ref{up.cluster.descr}) are between the average speeds for the \(13\)
-clusters (\(9.9\)Mb/s) and the pre-clustered average for the whole
-sample (\(9.3\)Mb/s). However, the standard deviation for cluster \(4\)
-and the difference between average speeds in the morning compared to the
-evening peak periods are indications of worse reliability than many of
-the other clusters. Hence, the results in Table \ref{aux} should be seen
-as relative rather than absolute probabilities.
+clusters that have higher mean speeds and more reliable services do
+indeed consist of LADs that are more urban or closer to major urban
+areas and are more likely to benefit from a choice of high quality
+internet services. This part of the regression aims to confirm any first
+level digital divides. Next, to better understand how the clusters fare
+in relation to the second level digital divide, we consider which LADs
+in which clusters have a higher proportion of occupations where the
+nature of the work and the skills that occupation employs enable
+telecommuting. Finally, we consider the rate of return on internet use,
+or third level digital divides, by reviewing which clusters have the
+highest earnings and job density, and also which clusters experienced
+the highest share of population furloughed during this period in the
+pandemic.
+
+The results of the auxiliary regression are presented in Table
+\ref{aux}. The dependent variable is the LAD cluster membership as
+described in the \protect\hyperlink{sec:3}{methods and data} section and
+equation \ref{eq1}. Each column represents a different cluster. The
+reference case is cluster \(4\), which includes only the local authority
+of Hambleton in North Yorkshire, a rural area of just over ninety
+thousand people. Mean, experienced upload speed in cluster \(4\) (see
+Table \ref{up.cluster.descr}) is close to the pre-clustered average for
+the whole sample (\(9.3\)Mb/s). However, the standard deviation for
+cluster \(4\) and the difference between average speeds in the morning
+compared to the evening peak periods are indications of worse
+reliability than many of the other clusters. Hence, the results in Table
+\ref{aux} should be seen as relative rather than absolute probabilities.
 
@@
 people in LADs with more unreliable connections to test their internet
-speeds more often, and the results validate our priors. Meanwhile, fast
-Virgin Media cable connections have historically only been available to
-\(45\)\% of premises in the UK \citep{ofcom2016}, where the more
-lucrative and competitive market originally attracted the cable TV
-provider. Those in clusters \(2\), \(3\), \(9\) and \(11\) benefit from
-a higher proportion of Virgin connections, which is an indication that
+speeds more often, and the results show that those in cluster \(6\) are
+by far the most concerned about reliability at that time of day.
+Meanwhile, those in clusters \(2\), \(3\), and \(9\) benefit from a
+higher proportion of Virgin connections, which is an indication that
 people in these clusters are more likely to live in urban areas, with
-more choice of broadband services. In other words, they are more likely
-to be on the right side of the first level, infrastructure-based digital
-divide, as we expected from the analysis in Section
-\protect\hyperlink{sec:4.1}{4.1}.
-
-We employ distance from London and from the nearest metropolitan area
-(including London) as two variables depicting peripherality, urban
-structure and, potentially, first level digital divides. The broadband
-speed tests run in the authorities in cluster \(3\) are more likely to
-be taking place close to London than those run in any of the other
-clusters, and two of the four authorities in cluster \(3\) are the
-London commuter towns of Harlow and Luton. However, even though London
-was also included in the variable calculating distance from the centre
-of one of either the ten largest metropolitan areas in England, or
-Glasgow or Cardiff, tests run in cluster \(3\) are likely to be furthest
-away. Thus, it is important to consider the membership of each cluster
-as well as the regression results. Corby and Eastbourne, the other two
-authorities in cluster \(3\) are large, accessible towns, although not
-part of metropolitan areas.
-
-\begin{figure}
-\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-6}
-\end{figure}
-
-Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
-of a large metropolitan area, even though the four local authorities of
-cluster \(7\) include no central urban boroughs and only one LAD that is
-part of a metropolitan area of governance - Tameside in Greater
-Manchester. This may explain why those in cluster \(7\) are the second
-least likely to be served by Virgin Media. It is also a demonstration of
-the complexity of both experienced broadband upload speeds as captured
-by time-series clustering, and the geography of first level digital
-divides as a product of quality as well as availability. In comparison,
-cluster \(1\), which our analysis suggests lacks broadband resilience,
-contains five, mainly rural authorities, but they are closer to a
-metropolitan area than authorities in cluster \(3\), although furthest
-from London, perhaps because they are scattered around the country --
-see Figure \ref{map.up.clusters}.
-
-Internet resilience is also more nuanced than our -- arguably crude --
-dummy variable depicting the North-South economic divide, which assigns
-\(1\) to LADs located in Greater London, Southeast, Southwest and East
-of England regions, and \(0\) to the rest. The authorities most likely
-to be in the South are those in cluster \(13\), which was identified as
-having the slowest mean upload speeds of any of the clusters, and thus a
-low level of service. However, cluster \(13\) includes some rural,
-remote areas of the country, such as Northwest Scotland, Cornwall and
-Powys in Wales as shown on Figure \ref{map.up.clusters}. It also
-includes major metropolitan centres in the North -- Liverpool, Newcastle
--- and South -- Bristol, and nine (of \(32\)) London Boroughs. There are
-also plenty of Southern home county and suburban areas. The standard
-deviation measure for cluster \(13\) is high, but speed variation is low
-during the morning peak, suggesting that the estimates for reliability
-are inconsistent. Considering that this is one of the largest clusters,
-and thus the averages incorporate more noise than some of the smaller
-clusters, it may be that the LADs in this cluster do not all suffer
-equally from a first level digital divide.
-
-Yet we need to consider the results for other variables in Table
-\ref{aux} to better determine whether clusters \(1\) and \(13\), which
-appear to suffer most from poor quality internet services, are also more
-likely to have a low skilled workforce, less able to benefit from
-telecommuting. Cluster \(1\) has the lowest proportion of educated
-people, and the lowest earnings, despite recording the highest
-proportion of managerial and professional jobs, and the second highest
-proportion of tech jobs. Rural areas such as those in cluster \(1\) are
-home to many older, retired people \citep{blank2018local}, which might
-explain these results or perhaps, as these figures are relative, we
-could note that cluster \(1\) also has a greater proportion of skilled
-trades than other clusters. In any case, it appears that the first level
-digital divide reinforces other inequities in cluster \(1\), where the
-even slower than average morning upload speeds shown in Table
-\ref{up.cluster.descr} suggest that internet users were more active
-during the working day. Meanwhile, those in cluster \(13\) are more
-likely to earn more and have a better education despite poor internet
-services. Cluster \(13\) also has the most businesses per inhabitant,
-but is somewhere in the middle in terms of job density. If this is an
-indication of a high number of SMEs, it could explain the variable
-internet quality, considering that small businesses have not been seen
-as the most valuable customers for higher speed broadband services if
-they are not located near residential customers \citep{ofcom2016}.
-
-Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
-there is more noise within the averages we use to measure internet
-quality and reliability. Our results indicate average mean speeds, and
-Table \ref{aux} shows that cluster \(6\) also falls towards the middle
-of the clusters on many of the socioeconomic variables. It ranks third
-or fourth out of the eleven in terms of the likelihood of having a
-higher proportion of managerial, professional, and tech jobs, as well as
-higher earnings, but is fourth from bottom for educational attainment.
-The LADs in cluster \(6\) are also diverse, with few truly remote areas,
-but urban areas throughout England, including Birmingham, Leeds,
-Sheffield, twelve London Boroughs, and many suburban areas and smaller
-cities like Oxford and Cambridge. The capital cities of the other UK
-nations, Belfast, Cardiff and Edinburgh, are also in this cluster,
-suggesting perhaps that the lower level of reliability discussed in
-Section \protect\hyperlink{sec:4.1}{4.1} may be due to increased demand,
-e.g.~for telecommuting, despite lacking the most resilient internet
-connections.
-
-In contrast, Clusters \(9\) and \(11\) enjoy resilient internet
-connections, but LADs in Cluster \(9\) are more likely to host highly
-educated individuals with higher earnings than cluster \(11\), which has
-a negative coefficient for the NVQ4 variable. Cluster \(11\) has fewer
-businesses per inhabitant and the second highest job density and cluster
-\(9\) the second lowest. These coefficients might indicate that
-resilient broadband infrastructure generates higher returns for those in
-cluster \(9\), where slightly more slowdown in the AM peak was detected
--- see Table \ref{up.cluster.descr}. LADs in cluster \(9\) are less
-likely to be in the South and those in cluster \(11\) more likely,
-although the coefficient for cluster \(9\) might be skewed by the
-presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
-different economic profile than England. Still, both clusters consist of
-larger LADs in terms of population, including districts within five of
-England's ten largest cities, and a number of other stand-alone urban
-areas and large market towns -- see Appendix
-\protect\hyperlink{appendix1}{1}. These urban locations are on the right
-side of the first level digital divide, and the regression results
-suggest that internet resilience is supporting a wide range of small and
-large urban economies.
-
-LADs in clusters \(2\), \(3\), \(7\), \(10\) are also on the right side
-of the first level digital divide, experiencing higher than average mean
-upload speeds and ranking high to middle on measures of service
-reliability. LADs in clusters \(2\) and \(10\) are also more likely to
-have highly skilled workers to take advantage of working from home
-opportunities. Cluster \(2\) is comprised of just two LADs, with the
-lowest job density of any cluster. Cluster \(10\) is comprised of four
-LADs, including two peripheral suburban areas of Birmingham, the East
-London Borough of Newham, and Dundee. Suburbs are considered the most
-likely urban form in which telecommuters live \citep{e2018does}, and
-Dundee has a reputation for tech startups \citep{technation2017}. LADs
-in cluster \(7\), which include the suburbs South of Belfast, a suburban
-district of Greater Manchester, and some other Midlands and North
-Yorkshire towns and villages have the highest earnings, and thus may
-also be making the most of their resilient internet. In contrast, LADs
-in cluster \(3\) are associated more with lower skills and have the
-highest proportion of individuals in machine operation jobs. Arguably,
-LADs in clusters \(3\) benefit the least from their resilient internet
-infrastructure in an era when working from home became a vehicle for
-economic resilience.
-
-Finally, clusters \(8\) and \(12\) consist of LADs enjoying close to
-average mean upload speeds, but opposing patterns of internet
-reliability during the morning peak. LADs in cluster \(8\) are likely to
-host more individuals employed in tech and administrative occupations
-than any other cluster, as well as many in managerial occupations,
-whilst the opposite applies to skilled trade and leisure occupations.
-These LADs are characterised by a very small positive likelihood of more
-educated residents, but are not significantly likely to earn more than
-other clusters. Including suburbs near Leicester, Cardiff, and
-Birmingham and south of London, the LADs in cluster \(8\) seem to likely
-to have the skills and occupations that would benefit from quality
-internet services, but suffer from poor internet resilience and
-reliability.
-
-In comparison, LADs in cluster \(12\) are less likely to be able to
-benefit from quality internet services if they had them, with fewer
-individuals achieving NVQ4 or better and lower levels of occupations
-that would benefit from homeworking. Yet LADs in this cluster benefit
-from the second highest level of earnings, and have the largest stock of
-businesses per inhabitant and the highest job density. This density of
-businesses could be why cluster \(12\) is the only cluster with higher
-speeds in the morning peak during the study period. If people are at
-home, business premises might well have been abandoned. Cluster \(12\)
-is home to \(1.5\) million people spread across twelve LADs, from London
-to Wales as listed in Appendix \protect\hyperlink{appendix1}{1}. This
-spatial diversity demonstrates that the temporal clustering of internet
-resilience is not necessarily spatially dependent, and digital divides
-do not necessarily overlap with economic ones.
+more profitable broadband markets, and a better choice of broadband
+services. We also employ distance from London and from the nearest
+metropolitan area (including London) as two variables depicting
+peripherality and urban structure. However, whilst significant, the
+pattern of coefficients of these variables is inconclusive in confirmed
+first level digital divides. Even though London was one of the ten
+largest metropolitan areas in England, which, along with Glasgow and
+Cardiff, were identified to calculate the variable estimating the impact
+of distance from the centre of a metropolitan area, the coefficients for
+the two variables are mostly small and the signs for some are opposite.
+
+When looking at the constituent authorities (see full list of LADs in
+Appendix \protect\hyperlink{appendix1}{1} as well as Figure
+\ref{map.up.clusters}), however, Cluster \(9\) is clearly more urban,
+including \(13\) of \(32\) London Boroughs, five of the seven
+constitutent LADs of the West Midlands conurbation, nine of the ten
+boroughs of Greater Manchester, and five of nine other large
+metropolitan areas coded in for the `distance to nearest met area'
+variable. Cluster \(9\) also includes the other main cities in the East
+Midlands, Leicester and Derby, as well as smaller cities known for their
+knowledge economy, such as Oxford and Milton Keynes. However, as
+previously mentioned, there is more noise within the variables for the
+two largest clusters than the smaller ones, and cluster \(6\) also
+contains many urban areas. These include \(16\) London boroughs,
+Birmingham City, Bolton in Greater Manchester, Leeds, Liverpool,
+Newcastle and Bristol, as well as small cities with knowledge economies
+such as Cambridge, Edinburgh, and Reading and its neighbours in the
+high-tech agglomeration of the Thames Valley. And yet cluster \(6\) also
+includes some of the most rural areas in the country, showing how
+important it is to interpretation to consider the membership of the
+cluster as well as the regression results.
+
+The coefficients in Table \ref{aux} demonstrate that those in cluster
+\(6\) were more likely to hold managerial, tech and professional jobs
+than those in cluster \(9\), and were the least likely to be furloughed
+of any cluster. Thus, although speeds were slow and not as reliable
+during the morning peak as in cluster \(9\), the skills that enabled
+telecommuting also enabled greater returns from doing so, despite lower
+than average earnings. Cluster \(6\) also has the second highest job
+density, with more jobs to keep. Likewise, those in LADs in clusters
+\(1\) and \(8\) had the highest and second highest proportions of
+residents working in tech or managerial occupations despite suffering
+from lower speeds and unreliable services, another mismatch between
+first and second level digital divides. Yet unlike cluster \(6\) and
+despite occupations with better digital skills, cluster \(8\), with its
+five peripheral suburban areas in the Midlands, Scotland and south of
+London, as well as rural West Devon saw the third highest proportion of
+its working population furloughed. It also had the second lowest average
+earnings in 2019, suggesting that its population was more likely to be
+on the wrong side of the third level digital divide before, as well as
+during the pandemic. In contrast, the nine LADs of cluster \(1\), which
+include more rural areas, plus Westminster in central London had the
+lowest earnings of any cluster, but also lowest level of furlough after
+cluster \(6\).
+
+The regression results for cluster \(7\) also reveal the complexity of
+intersecting and diverging digital divides. Along with \(3\) and \(9\),
+the analysis in Section \protect\hyperlink{sec:4.1}{4.1} suggested this
+cluster was on the `right' side of any first level digital divides, yet
+the regression shows that it has the lowest proportion of Virgin
+connections. It is furthest from a metropolitan area, but closest to
+London. The five LADs in cluster \(7\) are all suburban, making it
+sensible that the cluster is in the group with better infrastructure,
+but none are central urban boroughs and, along with a London suburb,
+Leeds suburb, Manchester suburb, and West Midlands suburb, the fifth LAD
+is outside Belfast. Belfast was not included among the metropolitan
+areas, and whilst there has been substantial investment in high speed
+broadband in Northern Ireland in recent years, including by Virgin,
+take-up may still lag behind historic networks in other urban markets.
+
+Cluster \(7\) has the lowest proportion of residents currently employed
+in professional occupations, and among the lowest proportions in
+managerial or technical occupations. This suggests potentially a low
+level of skills to take advantage of the quality internet service
+available, a second level digital divide. Cluster \(7\) also has the
+lowest job density. However, with the largest proportion of skilled
+tradespeople, and the second highest number of businesses per
+inhabitant, those in cluster \(7\) ranked second in terms of earnings in
+2019. On the other hand, the lack of some of the occupations most likely
+to telecommute during lockdown may have contributed both to a
+substantial proportion being furloughed in 2020, and even to the
+relative reliability of broadband speeds compared to many of the other
+smaller clusters.
+
+Cluster \(3\), meanwhile, had the greatest proportion of its workforce
+furloughed of any cluster, despite having the highest job density and
+the second highest average speeds. Ranking in the middle for most of the
+occupations estimated, at least some of its population should have the
+skills to work from home. The geography would suggest economic advantage
+as well, with eight of its twelve LADs in the South of England,
+including two London Boroughs, four suburban areas north of London, and
+the two tightly bounded urban areas of Eastbourne and Ipswich -- a
+greater proportion than any other cluster. Yet geographic position and
+job density do not appear to equate to particularly high average
+earnings in 2019, nor did they protect residents from furlough in 2020.
+In contrast, cluster \(5\), comprised of three suburban districts in the
+Northwest, Midlands, and the London green belt, had the highest average
+earnings in 2019, despite low levels of residents working in managerial,
+tech and professional occupations. Suburbs are considered the most
+likely urban form in which telecommuters live \citep{e2018does}, yet the
+contrast between these small, mainly suburban clusters -- \(7\), \(3\),
+and \(5\), suggest that they are not all equally economically resilient
+in a pandemic.
+
+Cluster \(2\), which, like cluster \(5\), had faster morning broadband
+speeds than evening speeds, as shown in Table \ref{up.cluster.descr}
+contains two remote LADs, the Isles of Scilly and Northeast
+Lincolnshire. Whilst the low demand may be due to the least residents in
+tech occupations, it may also be due to high numbers of retired people.
+Rural areas such as those in cluster \(2\) are often home to many older,
+retired people \citep{blank2018local}, which may also explain some of
+the contradictions in cluster \(1\) with its low levels of furlough and
+low earnings, whilst the presence of Westminster in that cluster might
+be why there are the highest levels of tech and managerial occupations.
+
+In summary, the regression results indicate the complexity of measuring
+second and third level digital divides within spatial aggregations where
+the geography of first level digital divides has been captured using
+time-series clustering of experienced broadband upload speeds as a
+product of reliability not just availability. Whilst the differences in
+mean speeds between clusters were not large, the temporal clustering of
+internet resilience showed much greater variation, and was not spatially
+dependent upon distance to large urban areas, or relative location.
+Internet resilience supports a wide range of small and large urban
+economies, but also fails, or at least frustrates a wide range of other
+urban economies, as demonstrated by the number of tests per capita run
+in cluster \(6\). Yet our analysis of the likelihood of furlough within
+the population, especially in the smaller suburban clusters \(7\),
+\(3\), \(5\), and also \(8\) shows that digital divides were as likely
+to diverge as intersect during the pandemic, and did not necessarily
+overlap with prior economic divides. With a much greater share of the
+population likely to continue to work from home in the future, and with
+changing attitudes towards residential locations, our analysis suggests
+that first level digital divides should be seen as a function of
+reliability as well as availability, of upload speeds as well as
+download speeds, and of greater spatial variation than might have
+previously been considered.
 
 \hypertarget{sec:5}{%
-\section{Conclusions}\label{sec:5}}
+\section{Discussion and Conclusions}\label{sec:5}}
 
@@
 with the necessary skills to effectively use the internet to
-telecommute, this paper raises the issue that places without good
-internet connectivity will still struggle more than other places to
-achieve economic resilience in a period like the current pandemic when
-internet resilience is so vital. Indeed, our analysis demonstrated that
-the temporal profiles of twelve of our thirteen clusters had slower
-upload speeds in the morning than in the evening. The opposite is likely
-to have been the norm prior to the pandemic, as level of demand and
-bandwidth management is the most common cause of temporal variation in
-experienced speeds, and why evening download speeds, rather than daytime
-upload speeds, have been used to benchmark the performance of internet
-services. Thus, the new patterns can be taken as evidence of widespread
-telecommuting and other daytime internet use which changed the temporal
-profile of internet activity throughout the UK, not just in areas with
-more digital industry or better skills.
-
-Upload speeds have not previously been seen as integral to universal
-service, considering there has never before been such extreme demand for
-telecommuting and operations such as video calls. This may be why the
-average upload speeds for the largest two clusters -- \(6\) and \(13\)
--- contain so many LADs that are centres of the knowledge economy, from
-Oxford and Cambridge in the former, to areas like Reading in the latter,
-which has the highest concentration of digital businesses in the country
-\citep{technation2017}. Thus, whilst some areas suffer from an
-intersection of digital and economic divides, such as those in cluster
-\(1\), other areas, including cluster \(8\) as well as \(6\) and \(13\)
-also found their digital infrastructure to be less than reliable when
-confronted with the sudden change in the timing and type of demand. In
-contrast, LADs such as the digital hub, Milton Keynes, in cluster \(9\)
-were able to benefit both from reliable internet connections and
-populations which were familiar with working from home and could
-capitalise on their digital infrastructure. Yet the nuanced picture we
-gained through our analysis of the UK case study suggests that quality
-internet connectivity may also have enabled other LADs, such as those in
-cluster \(3\) or \(11\) to gain ground during the pandemic despite lower
-skill levels.
+telecommute, this paper raises the issue that places may depend upon
+good internet reliability as well as connectivity to achieve economic
+resilience in a period like the current pandemic when internet
+resilience is so vital.
+
+Our analysis demonstrated that the temporal profiles of seven of our
+nine clusters had slower upload speeds in the morning than in the
+evening. The opposite is likely to have been the norm prior to the
+pandemic, as level of demand and bandwidth management is the most common
+cause of temporal variation in experienced speeds, and why evening
+download speeds, rather than daytime upload speeds, have been used to
+benchmark the performance of internet services. Thus, the new patterns
+can be taken as evidence of widespread telecommuting and other daytime
+internet use which changed the temporal profile of internet activity
+throughout the UK, not just in areas with more digital industry or
+better skills. Furthermore, upload speeds have not previously been seen
+as integral to universal service, considering there has never before
+been such extreme demand for telecommuting and operations such as video
+calls. Yet those in the largest cluster, \(6\), clearly experienced
+lower upload speeds. Their speeds were also less reliable, perhaps
+because they had higher levels of demand than the second largest
+cluster, \(9\), where many more employees were furloughed.
+
+Almost all of the largest urban areas in the UK, as well as many smaller
+cities, were split between the largest two clusters -- \(6\) and \(9\)
+-- and both contain LADs that are centres of the knowledge economy, from
+Cambridge in the former to Oxford in the latter, or have high
+concentrations of digital businesses, like Reading in the former and
+Milton Keynes in the latter \citep{technation2017}. Yet there are both
+intersections and divergences of digital and economic divides. LADs in
+cluster \(9\) were able to benefit both from reliable internet
+connections and populations which were familiar with working from home
+and could capitalise on their digital infrastructure, although overall,
+they appeared to have a lower rate of return in the pandemic, with far
+greater numbers furloughed. Is this an indication of both historic
+economic division, as well as the third level digital divide in cities
+in the North and Midlands unable to capitalise on their digital
+infrastructure? Meanwhile, did the greater presence in London and the
+South of LADs in cluster \(6\), as well as the inclusion of many rural
+areas less impacted by the pandemic help those in this cluster be more
+resilient and gain greater returns, despite using less reliable internet
+services? The nuanced picture we gained through our analysis of the UK
+case study suggests that in cluster \(1\) as well as \(6\), being on the
+right side of the second level digital divide had a greater impact on
+economic resilience and therefore the third level digital divide, than
+having quality internet connectivity.
 
@@
 
+Early research on Covid-19 and cities speculated the changes that
+potential extensive post-Covid working from home patterns might generate
+for spatial structure: from more walkable cities to more localised
+production and consumption patterns and from more extensive urban sprawl
+to decline in public transportation and increased private car usage
+\citep{batty2020editorial}. Despite the essential role it played during
+the pandemic, less effort has been spent in understanding the current
+and future role of digital infrastructure. Our results indicate that
+probably for the first time the future of cities and spatial structure
+are so much intertwined with digital infrastructure. If the post-covid
+world is a world with extensive working from home, then we need to
+build, among other things, resilient digital infrastructure capable of
+bridging the first layer of the digital divide. Contrary to previous
+broadband business and deployment plans, emphasis should be placed not
+only on download, but also on upload speeds {[}\citet{brake2020lessons}.
+In essence, Covid-19 accentuated the old, but more valid than ever
+argument: digital infrastructure, just like any other network
+infrastructure, only becomes visible when it stops working
+\citep{star1999, tranos2013geography}.
+
 \hypertarget{appendix1}{%
@@
 
-\textbf{Cluster 1: } Ceredigion, Darlington, Eden, Rossendale, Rutland
-
-\textbf{Cluster 2: } East Lothian, North East Lincolnshire
-
-\textbf{Cluster 3: } Corby, Eastbourne, Harlow, Luton
+\textbf{Cluster 1: } Ceredigion, Darlington, Eden, Erewash, Kettering,
+Rossendale, Runnymede, Rutland, Westminster
+
+\textbf{Cluster 2: } Isles of Scilly, North East Lincolnshire
+
+\textbf{Cluster 3: } Broxbourne, Carlisle, Corby, Dundee City,
+Eastbourne, Harlow, Hertsmere, Hounslow, Ipswich, Luton, Newham, South
+Northamptonshire
 
@@
 
-\textbf{Cluster 5: } Fylde
-
-\textbf{Cluster 6: } Allerdale, Amber Valley, Angus, Ashfield, Ashford,
-Aylesbury Vale, Barnet, Basingstoke and Deane, Bath and North East
-Somerset, Belfast, Bexley, Birmingham, Blaenau Gwent, Bournemouth,
-Christchurch and Poole, Bradford, Braintree, Brentwood, Bridgend,
-Bromley, Broxtowe, Bury, Calderdale, Cambridge, Canterbury, Cardiff,
-Castle Point, Chelmsford, Cheltenham, Cherwell, Chesterfield, City of
-Edinburgh, Clackmannanshire, Colchester, Copeland, County Durham,
-Coventry, Croydon, Dartford, Daventry, Denbighshire, Derby, Derry City
-and Strabane, Dorset, Ealing, East Ayrshire, East Hampshire, East
-Lindsey, East Northamptonshire, East Renfrewshire, East Riding of
-Yorkshire, East Suffolk, Eastleigh, Elmbridge, Enfield, Falkirk,
-Fareham, Gateshead, Gedling, Gosport, Gravesham, Great Yarmouth,
-Guildford, Harborough, Haringey, Harrogate, Harrow, Hart, Hartlepool,
-Havering, Herefordshire, County of, High Peak, Hinckley and Bosworth,
-Horsham, Islington, Kettering, King's Lynn and West Norfolk, Kingston
-upon Thames, Kirklees, Leeds, Leicester, Lincoln, Maidstone, Maldon,
-Mansfield, Medway, Mendip, Mid Sussex, Middlesbrough, Monmouthshire,
-Moray, New Forest, Newcastle-under-Lyme, Newport, North Ayrshire, North
-East Derbyshire, North Hertfordshire, North Kesteven, North Lanarkshire,
-North Lincolnshire, North Norfolk, North Tyneside, North West
-Leicestershire, Northumberland, Nuneaton and Bedworth, Oxford,
-Pembrokeshire, Pendle, Renfrewshire, Ribble Valley, Rochford, Runnymede,
-Rushcliffe, Ryedale, Salford, Sefton, Sheffield, Shropshire, Solihull,
-South Ayrshire, South Hams, South Holland, South Lanarkshire, South
-Oxfordshire, South Staffordshire, St Albans, Staffordshire Moorlands,
-Stockport, Stoke-on-Trent, Surrey Heath, Sutton, Swale, Tamworth,
-Tendring, Test Valley, Thurrock, Tonbridge and Malling, Torfaen,
-Wakefield, Warrington, Warwick, Wealden, Wellingborough, West Berkshire,
-West Dunbartonshire, West Lancashire, West Lothian, West Oxfordshire,
-West Suffolk, Wigan, Wiltshire, Woking, Worcester, Wrexham, Wycombe,
-York
-
-\textbf{Cluster 7: } Lisburn and Castlereagh, Selby, Tameside, Wyre
-Forest
-
-\textbf{Cluster 8: } Lichfield, Oadby and Wigston, Tandridge, Vale of
-Glamorgan, West Devon
-
-\textbf{Cluster 9: } Aberdeen City, Barnsley, Broxbourne, Charnwood,
-Chorley, Erewash, Glasgow City, Greenwich, Halton, Havant, Knowsley,
-Lewisham, Merton, Mid and East Antrim, Milton Keynes, Newark and
-Sherwood, Northampton, Oldham, Portsmouth, Richmond upon Thames, Rugby,
-Sandwell, South Derbyshire, South Kesteven, South Northamptonshire,
-Southampton, Spelthorne, Stockton-on-Tees, Telford and Wrekin, Trafford,
-Walsall, Welwyn Hatfield
-
-\textbf{Cluster 10: } Bromsgrove, Cannock Chase, Dundee City, Newham
-
-\textbf{Cluster 11: } Barking and Dagenham, Blaby, Blackpool, Bolsover,
-Brent, Burnley, Caerphilly, Carlisle, Causeway Coast and Glens, Crawley,
-Doncaster, Dudley, Hertsmere, Hounslow, Hyndburn, Ipswich, Isles of
-Scilly, Kensington and Chelsea, Lewes, Manchester, North Warwickshire,
-Norwich, Nottingham, Peterborough, Redditch, Rochdale, Scarborough,
-Slough, St.~Helens, Stevenage, Sunderland, Vale of White Horse,
-Wolverhampton
-
-\textbf{Cluster 12: } Ards and North Down, Conwy, East Staffordshire,
-Epping Forest, Fenland, Hammersmith and Fulham, Preston, Rhondda Cynon
-Taf, Three Rivers, Westminster
-
-\textbf{Cluster 13: } Aberdeenshire, Adur, Antrim and Newtownabbey,
-Argyll and Bute, Armagh City, Banbridge and Craigavon, Arun, Babergh,
-Barrow-in-Furness, Basildon, Bassetlaw, Bedford, Blackburn with Darwen,
-Bolton, Boston, Bracknell Forest, Breckland, Brighton and Hove, Bristol,
-City of, Broadland, Camden, Carmarthenshire, Central Bedfordshire,
-Cheshire East, Cheshire West and Chester, Chichester, Chiltern, City of
-London, Cornwall, Cotswold, Craven, Dacorum, Derbyshire Dales, Dover,
-Dumfries and Galloway, East Cambridgeshire, East Devon, East
-Dunbartonshire, East Hertfordshire, Epsom and Ewell, Exeter, Fermanagh
-and Omagh, Fife, Flintshire, Folkestone and Hythe, Forest of Dean,
-Gloucester, Gwynedd, Hackney, Hastings, Highland, Hillingdon,
-Huntingdonshire, Inverclyde, Isle of Anglesey, Isle of Wight, Kingston
-upon Hull, City of, Lambeth, Lancaster, Liverpool, Malvern Hills,
-Melton, Merthyr Tydfil, Mid Devon, Mid Suffolk, Mid Ulster, Midlothian,
-Mole Valley, Na h-Eileanan Siar, Neath Port Talbot, Newcastle upon Tyne,
-Newry, Mourne and Down, North Devon, North Somerset, Orkney Islands,
-Perth and Kinross, Plymouth, Powys, Reading, Redbridge, Redcar and
-Cleveland, Reigate and Banstead, Richmondshire, Rother, Rotherham,
-Rushmoor, Scottish Borders, Sedgemoor, Sevenoaks, Shetland Islands,
-Somerset West and Taunton, South Bucks, South Cambridgeshire, South
-Gloucestershire, South Lakeland, South Norfolk, South Ribble, South
-Somerset, South Tyneside, Southend-on-Sea, Southwark, Stafford,
-Stirling, Stratford-on-Avon, Stroud, Swansea, Swindon, Teignbridge,
-Tewkesbury, Thanet, Torbay, Torridge, Tower Hamlets, Tunbridge Wells,
-Uttlesford, Waltham Forest, Wandsworth, Watford, Waverley, West Lindsey,
-Winchester, Windsor and Maidenhead, Wirral, Wokingham, Worthing,
-Wychavon, Wyre
+\textbf{Cluster 5: } Fylde, Rugby, Three Rivers
+
+\textbf{Cluster 6: } Aberdeenshire, Adur, Allerdale, Amber Valley,
+Angus, Antrim and Newtownabbey, Argyll and Bute, Armagh City, Banbridge
+and Craigavon, Arun, Ashford, Aylesbury Vale, Babergh, Barnet,
+Barrow-in-Furness, Basildon, Bassetlaw, Bath and North East Somerset,
+Bedford, Belfast, Birmingham, Blackburn with Darwen, Blackpool, Blaenau
+Gwent, Bolton, Boston, Bournemouth, Christchurch and Poole, Bracknell
+Forest, Bradford, Braintree, Breckland, Brentwood, Bridgend, Brighton
+and Hove, Bristol, City of, Broadland, Bromley, Calderdale, Cambridge,
+Camden, Canterbury, Carmarthenshire, Causeway Coast and Glens, Central
+Bedfordshire, Chelmsford, Cherwell, Cheshire East, Cheshire West and
+Chester, Chesterfield, Chichester, Chiltern, City of Edinburgh, City of
+London, Clackmannanshire, Conwy, Copeland, Cornwall, Cotswold, County
+Durham, Coventry, Craven, Croydon, Dacorum, Dartford, Denbighshire,
+Derbyshire Dales, Derry City and Strabane, Dorset, Dover, Dumfries and
+Galloway, Ealing, East Cambridgeshire, East Devon, East Dunbartonshire,
+East Hampshire, East Hertfordshire, East Lindsey, East Renfrewshire,
+East Riding of Yorkshire, East Suffolk, Elmbridge, Epping Forest, Epsom
+and Ewell, Exeter, Fareham, Fenland, Fermanagh and Omagh, Fife,
+Flintshire, Folkestone and Hythe, Forest of Dean, Gateshead, Gloucester,
+Gosport, Great Yarmouth, Greenwich, Gwynedd, Hackney, Harborough,
+Haringey, Hartlepool, Hastings, Havering, Herefordshire, County of, High
+Peak, Highland, Hillingdon, Horsham, Huntingdonshire, Inverclyde, Isle
+of Anglesey, King's Lynn and West Norfolk, Kingston upon Hull, City of,
+Kingston upon Thames, Kirklees, Lambeth, Lancaster, Leeds, Lincoln,
+Liverpool, Maidstone, Malvern Hills, Mansfield, Melton, Merthyr Tydfil,
+Mid Devon, Mid Suffolk, Mid Ulster, Midlothian, Mole Valley,
+Monmouthshire, Moray, Na h-Eileanan Siar, Neath Port Talbot, New Forest,
+Newcastle upon Tyne, Newry, Mourne and Down, North Devon, North East
+Derbyshire, North Lanarkshire, North Lincolnshire, North Norfolk, North
+Somerset, North Tyneside, North Warwickshire, North West Leicestershire,
+Northumberland, Orkney Islands, Pembrokeshire, Pendle, Perth and
+Kinross, Peterborough, Plymouth, Powys, Preston, Reading, Redcar and
+Cleveland, Reigate and Banstead, Rhondda Cynon Taf, Ribble Valley,
+Richmondshire, Rother, Rotherham, Rushcliffe, Rushmoor, Ryedale,
+Scottish Borders, Sedgemoor, Sefton, Sevenoaks, Shetland Islands,
+Shropshire, Somerset West and Taunton, South Ayrshire, South Bucks,
+South Cambridgeshire, South Gloucestershire, South Hams, South Holland,
+South Lakeland, South Norfolk, South Oxfordshire, South Ribble, South
+Somerset, South Staffordshire, Southend-on-Sea, Southwark, St Albans,
+Stafford, Stirling, Stoke-on-Trent, Stratford-on-Avon, Stroud, Swale,
+Swansea, Swindon, Teignbridge, Tendring, Test Valley, Tewkesbury,
+Thanet, Tonbridge and Malling, Torbay, Torfaen, Torridge, Tower Hamlets,
+Tunbridge Wells, Uttlesford, Wakefield, Waltham Forest, Wandsworth,
+Warrington, Watford, Waverley, Wellingborough, West Berkshire, West
+Lancashire, West Lindsey, West Oxfordshire, Wiltshire, Winchester,
+Windsor and Maidenhead, Wirral, Wokingham, Worcester, Worthing, Wrexham,
+Wychavon, Wycombe, Wyre
+
+\textbf{Cluster 7: } Lisburn and Castlereagh, Selby, Tameside, Welwyn
+Hatfield, Wyre Forest
+
+\textbf{Cluster 8: } Cannock Chase, East Lothian, Lichfield, Oadby and
+Wigston, Tandridge, West Devon
+
+\textbf{Cluster 9: } Aberdeen City, Ards and North Down, Ashfield,
+Barking and Dagenham, Barnsley, Basingstoke and Deane, Bexley, Blaby,
+Bolsover, Brent, Bromsgrove, Broxtowe, Burnley, Bury, Caerphilly,
+Cardiff, Castle Point, Charnwood, Cheltenham, Chorley, Colchester,
+Crawley, Daventry, Derby, Doncaster, Dudley, East Ayrshire, East
+Northamptonshire, East Staffordshire, Eastleigh, Enfield, Falkirk,
+Gedling, Glasgow City, Gravesham, Guildford, Halton, Hammersmith and
+Fulham, Harrogate, Harrow, Hart, Havant, Hinckley and Bosworth,
+Hyndburn, Isle of Wight, Islington, Kensington and Chelsea, Knowsley,
+Leicester, Lewes, Lewisham, Maldon, Manchester, Medway, Mendip, Merton,
+Mid and East Antrim, Mid Sussex, Middlesbrough, Milton Keynes, Newark
+and Sherwood, Newcastle-under-Lyme, Newport, North Ayrshire, North
+Hertfordshire, North Kesteven, Northampton, Norwich, Nottingham,
+Nuneaton and Bedworth, Oldham, Oxford, Portsmouth, Redbridge, Redditch,
+Renfrewshire, Richmond upon Thames, Rochdale, Rochford, Salford,
+Sandwell, Scarborough, Sheffield, Slough, Solihull, South Derbyshire,
+South Kesteven, South Lanarkshire, South Tyneside, Southampton,
+Spelthorne, St.~Helens, Staffordshire Moorlands, Stevenage, Stockport,
+Stockton-on-Tees, Sunderland, Surrey Heath, Sutton, Tamworth, Telford
+and Wrekin, Thurrock, Trafford, Vale of Glamorgan, Vale of White Horse,
+Walsall, Warwick, Wealden, West Dunbartonshire, West Lothian, West
+Suffolk, Wigan, Woking, Wolverhampton, York
 
@@
 \hline \\[-1.8ex] 
-pop, 2018 & 365 & 174,952.100 & 119,557.100 & 8,700 & 100,400 & 214,900 & 1,141,400 \\ 
+pop, 2018 & 365 & 174,952.100 & 119,557.100 & 8,700.000 & 100,400.000 & 214,900.000 & 1,141,400.000 \\ 
 job density, 2018 & 365 & 1.137 & 5.726 & 0.400 & 0.700 & 0.930 & 110.110 \\ 
@@
 distance to London & 365 & 201.558 & 173.634 & 0.150 & 76.180 & 278.880 & 1,003.950 \\ 
-south of the UK & 365 & 0.463 & 0.499 & 0 & 0 & 1 & 1 \\ 
+south of the UK & 365 & 0.463 & 0.499 & 0.000 & 0.000 & 1.000 & 1.000 \\ 
 managerial jobs, 2020 & 363 & 12.009 & 4.013 & 3.600 & 9.000 & 14.300 & 27.900 \\ 
@@
 administrative jobs, 2020 & 359 & 9.965 & 2.738 & 3.200 & 8.100 & 11.400 & 21.300 \\ 
-leisure jobs, 2020 & 362 & 9.261 & 2.827 & 2.800 & 7.300 & 11.400 & 17.800 \\ 
+service jobs, 2020 & 362 & 9.261 & 2.827 & 2.800 & 7.300 & 11.400 & 17.800 \\ 
 machine operation jobs, 2020 & 337 & 6.339 & 2.847 & 1.200 & 4.400 & 7.900 & 19.800 \\ 
 earnings, 2019 & 360 & 592.184 & 81.129 & 437.600 & 534.625 & 633.875 & 893.200 \\ 
-NVQ4+ & 365 & 39.329 & 11.076 & 15.000 & 31.800 & 45.300 & 100.000 \\ 
-Virgin Media \%, 2020 & 365 & 0.152 & 0.141 & 0.000 & 0.018 & 0.241 & 0.753 \\ 
+Virgin Media \%, 2020 & 381 & 0.150 & 0.140 & 0.000 & 0.018 & 0.238 & 0.753 \\ 
 n. business est. per hab., 2019 & 365 & 0.057 & 0.164 & 0.023 & 0.038 & 0.056 & 3.174 \\ 
 AM tests per hab., 2020 & 365 & 0.0005 & 0.0002 & 0.0001 & 0.0003 & 0.001 & 0.001 \\ 
+furloughed per hab., 2020 & 362 & 0.115 & 0.013 & 0.077 & 0.107 & 0.122 & 0.178 \\ 
 \hline \\[-1.8ex] 

### 2021-03-22 src/LAs_Clusters_k9.Rmd
ET changes / responses to reviewer comments

@@
 
-```{r upload regression , results= 'asis', message=F, error=F, warning=F}
+```{r upload regression, results= 'asis', message=F, error=F, warning=F}
+
 # helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
@@
 
-aux.up <- multinom(cluster.up ~ 
+aux.up.v1 <- multinom(cluster.up ~ 
                      #pop16_64 + # non sign, while pop is
@@
 # McFadden's R squared
-aux.up.null <- multinom(cluster.up ~ 1, data = data)
-r2.up <- 1-logLik(aux.up)/logLik(aux.up.null)
+aux.up.v1.null <- multinom(cluster.up ~ 1, data = data)
+r2.up.v1 <- 1-logLik(aux.up.v1)/logLik(aux.up.v1.null)
 
-n.obs <- dim(aux.up$fitted.values)[1]
+n.obs <- dim(aux.up.v1$fitted.values)[1]
 
-stargazer(aux.up, 
+stargazer(aux.up.v1, 
           type = "html",
@@
           add.lines = list(c("McFadden's R squared",
-                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
-                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
-                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
-                             , round(r2.up[1],3), round(r2.up[1],3)),
+                             round(r2.up.v1[1],3), round(r2.up.v1[1],3), round(r2.up.v1[1],3),
+                             round(r2.up.v1[1],3), round(r2.up.v1[1],3), round(r2.up.v1[1],3),
+                             round(r2.up.v1[1],3), round(r2.up.v1[1],3), round(r2.up.v1[1],3)
+                             , round(r2.up.v1[1],3), round(r2.up.v1[1],3)),
                            (c("N",
@@
 ```{r upload regression no NVQ4, results= 'asis', message=F, error=F, warning=F}
+
+# helpful sources: http://www.princeton.edu/~otorres/LogitR101.pdf
+# https://thestatsgeek.com/2014/02/08/r-squared-in-logistic-regression/
+
+# convert dependent variable to factor
+data$cluster.up <- as.factor(data$cluster.up)
+data$cluster.up <- relevel(data$cluster.up, "4") # base category: cluster 4
+
 aux.up <- multinom(cluster.up ~ 
@@
 
-aux.up <- multinom(cluster.up ~ 
+aux.up.noprof <- multinom(cluster.up ~ 
                      #pop16_64 + # non sign, while pop is
@@
 # McFadden's R squared
-aux.up.null <- multinom(cluster.up ~ 1, data = data)
-r2.up <- 1-logLik(aux.up)/logLik(aux.up.null)
+aux.up.noprof.null <- multinom(cluster.up ~ 1, data = data)
+r2.up.noprof <- 1-logLik(aux.up.noprof)/logLik(aux.up.noprof.null)
 
-n.obs <- dim(aux.up$fitted.values)[1]
+n.obs <- dim(aux.up.noprof$fitted.values)[1]
 
-stargazer(aux.up, 
+stargazer(aux.up.noprof, 
           type = "html",
@@
           add.lines = list(c("McFadden's R squared",
-                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
-                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
-                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
-                             , round(r2.up[1],3), round(r2.up[1],3)),
+                             round(r2.up.noprof[1],3), round(r2.up.noprof[1],3), round(r2.up.noprof[1],3),
+                             round(r2.up.noprof[1],3), round(r2.up.noprof[1],3), round(r2.up.noprof[1],3),
+                             round(r2.up.noprof[1],3), round(r2.up.noprof[1],3), round(r2.up.noprof[1],3)
+                             , round(r2.up.noprof[1],3), round(r2.up.noprof[1],3)),
                            (c("N",

### 2021-03-22 src/descriptive_Clusters.Rmd
ET changes / responses to reviewer comments

@@
 names(TS2019)[13] <- "test frequency 2019"
-time.var.plot2019 <- timeVariation(TS2019, pollutant = "test frequency 2019", statistic = "mean")
+time.var.plot2019 <- timeVariation(TS2019, pollutant = "test frequency 2019", statistic = "mean",
+                                   par.settings=list(fontsize=list(text=8)))
 Tests20 <- count(TS2020, date)
@@
 names(TS2020)[13] <- "test frequency 2020"
-time.var.plot2020 <- timeVariation(TS2020, pollutant = "test frequency 2020", statistic = "mean")
+time.var.plot2020 <- timeVariation(TS2020, pollutant = "test frequency 2020", statistic = "mean",
+                                   par.settings=list(fontsize=list(text=8)))
 
 #export time variation plots for paper
-out.plot <- paste(path, "/paper/v1/figures/time.var.plot2019.png", sep = "")
+out.plot <- paste(path, "/paper/v2_taylor_francis/figures/time.var.plot2019.png", sep = "")
 png(out.plot)
@@
 
-out.plot <- paste(path, "/paper/v1/figures/time.var.plot2020.png", sep = "")
+out.plot <- paste(path, "/paper/v2_taylor_francis/figures/time.var.plot2020.png", sep = "")
 png(out.plot)
@@
 #export UpClusterL plot for paper
-out.plot <- paste(path, "/paper/v1/figures/UpCluster.png", sep = "")
-png(out.plot)
-UpClusterL
-dev.off()
+out.plot <- paste(path, "/paper/v2_taylor_francis/figures/UpCluster.png", sep = "")
+UpCluster
+ggsave(out.plot)
 
@@
 
+# The figure with the small clusters is no included in the paper
 #export UpClusterS plot for paper
-#out.plot <- paste(path, "/paper/v1/figures/UpClusterS.png", sep = "")
+#out.plot <- paste(path, "/paper/v2_taylor_francis/figures/UpClusterS.png", sep = "")
 #png(out.plot)
@@
 
+# Not used in the paper
 #summarise download speed clusters by hour of each weekday and plot

### 2021-03-22 src/ts_clusters_k9.Rmd
ET changes / responses to reviewer comments

@@
 library(cowplot)
+library(ggrepel)
+library(maps)
+library(gridExtra)
 
@@
 # source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe
+# also saved locally \data\raw\Local_Authority_Districts_(December_2019)_Boundaries_UK_BGC.geojson
 
@@
 # ggplot2 maps
-la.f <- fortify(la, region = "lad19nm")
+la.f <- broom::tidy(la, region = "lad19nm")
 la.f <- merge(la.f, clusters.nodiff, by.x = "id", by.y = "LAD")
@@
 
+# ggplot(la.f, aes(x = long, y = lat)) +
+#   geom_polygon(aes( group = group, fill = as.factor(cluster.tests))) +
+#   #theme_nothing(legend = TRUE) +
+#   labs(title = "Test clusters") +
+#   scale_fill_viridis_d()+
+#   guides(fill=guide_legend(title="Clusters")) +
+#   theme_void()
+
+# using world.cities from `maps`
+cities <- world.cities %>% filter(country.etc=="UK")
+
 ggplot(la.f, aes(x = long, y = lat)) +
-  geom_polygon(aes( group = group, fill = as.factor(cluster.tests))) +
-  #theme_nothing(legend = TRUE) +
-  labs(title = "Test clusters") +
-  scale_fill_viridis_d()+
+  geom_polygon(aes( group = group, fill = as.factor(cluster.up))) +
+  #labs(title = "Upload clusters") +
+  scale_fill_brewer(palette="Set1", 
+                    direction = -1) +
+  geom_text_repel( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=4) +
+  geom_point( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat), color="black", size=1) +
   guides(fill=guide_legend(title="Clusters")) +
@@
 
-ggplot(la.f, aes(x = long, y = lat)) +
-  geom_polygon(aes( group = group, fill = as.factor(cluster.up))) +
-  #theme_nothing(legend = TRUE) +
-  labs(title = "Upload clusters") +
-  scale_fill_viridis_d()+
-  guides(fill=guide_legend(title="Clusters")) +
-  theme_void()
-
-#ggsave("all_maps2.png") # this is the colour map for the preprint
-# out.path.map <- paste0(path, "/paper/v1/figures/map.up.clusters9.png")
-# ggsave(out.path.map)#,
+# Run only if needed
+# out.path.map <- paste0(path, "/paper/v2_taylor_francis/figures/map.up.clusters.png")
+# ggsave(out.path.map)
 #width = 210, #A4
@@
 
-```{r, eval=FALSE}
+## Mapping median upload speeds for appendix 2019 and 2020
+
+```{r}
+
+# upload 2020
+ts.up2020median <- TS2020 %>%
+  group_by(LAD19NM) %>%
+  summarise(mean.up = median(speedup))
+
+ts.up2019median <- TS2019 %>%
+  group_by(LAD19NM) %>%
+  summarise(mean.up = median(speedup))
+
+# scatter plots
+sp2020 <- ggplot(ts.up2020median, 
+       aes(x=mean.up)) + 
+  geom_histogram() +
+  geom_vline(aes(xintercept=mean(mean.up)), # This line of code adds a vertical line to represent the mean
+             color="blue", linetype="dashed", size=1) +
+  xlim(0, 20000) +
+  ylab("count 2020")
+
+sp2019 <- ggplot(ts.up2019median, 
+       aes(x=mean.up)) + 
+  geom_histogram() +
+  geom_vline(aes(xintercept=mean(mean.up)), # This line of code adds a vertical line to represent the mean
+             color="blue", linetype="dashed", size=1) +
+  xlim(0, 20000) +
+  ylab("count 2019")
+
+sp20192020 <- grid.arrange(sp2019, sp2020, ncol = 1, nrow = 2) #legend
+
+out.plot <- paste0(path, "/paper/v2_taylor_francis/figures/sc2019_2020.png")
+ggsave(out.plot, sp20192020)
+
+la.f.up <- broom::tidy(la, region = "lad19nm")
+la.f.up <- merge(la.f.up, ts.up2020median, by.x = "id", by.y = "LAD19NM")
+la.f.up <- la.f.up %>% 
+  rename(mean.up2020 = mean.up)
+la.f.up <- merge(la.f.up, ts.up2019median, by.x = "id", by.y = "LAD19NM")
+la.f.up <- la.f.up %>% 
+  rename(mean.up2019 = mean.up)
+la.f.up <- la.f.up[order(la.f.up$order),] # if i don't order merge.nuts.f loses order and the map has gaps
+
+# Already defined
+# using world.cities from `maps`
+# cities <- world.cities %>% filter(country.etc=="UK")
+
+no_classes <- 4
+quantiles2019 <- quantile(ts.up2019median$mean.up, 
+                      probs = seq(0, 1, length.out = no_classes + 1), na.rm = T)
+la.f.up$quantile2019 <- cut(la.f.up$mean.up2019,
+                             breaks = quantiles2019,
+                             labels = c("1st quartile", "2nd quartile", "3rd quartile", "4th quartile"),
+                             right = T)
+quantiles2020 <- quantile(ts.up2020median$mean.up, 
+                      probs = seq(0, 1, length.out = no_classes + 1), na.rm = T)
+la.f.up$quantile2020 <- cut(la.f.up$mean.up2020,
+                             breaks = quantiles2020,
+                             labels = c("1st quartile", "2nd quartile", "3rd quartile", "4th quartile"),
+                             right = T)
+
+map2019 <- ggplot(la.f.up, aes(x = long, y = lat)) +
+  geom_polygon(aes( group = group, fill = quantile2019)) + #mean.up2019
+  #labs(title = "Upload clusters") +
+  scale_fill_brewer(palette = "Greens") +
+  geom_text_repel( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=4) +
+  geom_point( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat), color="black", size=1) +
+  guides(fill=guide_legend(title="Median upload speed")) + #, 2019
+  theme_void() #+ 
+  # theme(legend.text = element_text(size = 11), legend.position="bottom",
+  #       legend.key.size = unit(0.25, "cm"), legend.key.width = unit(0.25,"cm"), 
+  #       legend.box = "horizontal",
+  #       legend.margin=margin(0,0,0,0), legend.box.margin=margin(-40,20,40,20), # This is to bring the legend closer to the map for the grid
+  #       plot.margin = unit(c(0, 1.5, 0, -1.5), "cm"))
+
+map2020 <- ggplot(la.f.up, aes(x = long, y = lat)) +
+  geom_polygon(aes( group = group, fill = quantile2020)) +
+  #labs(title = "Upload clusters") +
+  scale_fill_brewer(palette = "Greens") +
+  geom_text_repel( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=4) +
+  geom_point( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat), color="black", size=1) +
+  guides(fill=guide_legend(title="Median upload speed, 2020")) +
+  theme_void()
+
+# get legend
+legend <- get_legend(map2019)
+
+# remove legends
+map2019 <- map2019 + theme(legend.position = "map2019")
+map2020 <- map2020 + theme(legend.position = "map2020")
+
+up20192020 <- grid.arrange(map2019, map2020, ncol = 2, nrow = 2) #legend
+
+out.plot <- paste0(path, "/paper/v2_taylor_francis/figures/up2019_2020.png")
+ggsave(out.plot, up20192020)
+
+# plot_grid(couk.pop2000, couk.pop2010, noncouk.pop2000, noncouk.pop2010, legend, nrow = 3, ncol = 2,
+#           align = "h", axis = "l", 
+#           #labels = c(".co.uk / pop., 2000", ".co.uk / pop., 2010", "non .co.uk / pop., 2000", "non .co.uk / pop., 2010", ""), 
+#           #label_size = 12,
+#           greedy = T,
+#           rel_heights = c(1, 1, .3)) # this is the relative position of the 3 lines
+
+```
+
+```{r, eval=FALSE, include=FALSE}
+# test for ploty
+
+library(plotly)
+
+p <- 
+  ggplotly(test, layerData = 1, 
+           originalData = T,
+           add_trace(z=~cluster.tests))#, tooltip="text")
+
+```
+
+```{r, eval=FALSE, include=FALSE}
 # leaflet maps

### 2021-03-24 paper/v2_taylor_francis/broadband.speed.covid.Rmd
Most changes discussed complete.

@@
 Our analysis enables us to better understand how the spatial and social distribution
-of education, occupation and online accessibility intersect to enable or hinder 
+of occupation and online accessibility intersect to enable or hinder 
 the practice of telecommuting at a time of extreme demand.
@@
 has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
-Moreover, @riddlesden2014broadband found that levels of deprivation did not correlate with 
+In contrast, @riddlesden2014broadband found that levels of deprivation did not correlate with 
 first level digital divides.
 
-However, as @blank2018local highlight, variation
-in individual internet uptake and use is a product of more complex spatial and 
+However, as @blank2018local highlight, variation in individual internet 
+availability and uptake is a product of more complex spatial and 
 demographic characteristics than simple rurality or urbanisation.
@@
 because telecommuting has not previously been a cause of concentrated demand and 
-pressure on internet services. Other intersections between digital divides have been
-subject to study.
+pressure on internet services.
+
+Other intersections between digital divides have been subject to study.
 There are indications that those who purchase high speed connections consume
@@
 savings and efficiencies, an example of the first level digital divide reinforcing 
-the third level.
-
-<!-- spatio-temporal -->
-Currently there is a hotly debated discussion in the literature regarding these 
-exact changes in transportation and the structure of cities. We now have enough 
-hard data to observe the drastic change in the space-time geography of cities around 
-the world [e.g. @google2020; @shibayama2021impact]. 
-Although there is a broad agreement that these changes 
+the third level. Such digital intersections and efficiencies have also changed 
+transportation, connectivity and the structure of cities during the pandemic.
+We now have enough data to confirm the drastic alteration observed in the space-time 
+geography of cities around the world in the last year [e.g. @google2020; @shibayama2021impact]. 
+Yet, although there is a broad agreement that these changes 
 during the pandemic have played a pivotal role in stopping the spread of the virus 
-[@jia2020population; @yang2020taking; @mu2020interplay],
-we still have no evidence to what extend these increased levels of working from 
-home, and the consequent decrease in commuting flows and the structure of cities, 
-will remain post-pandemic.
+[@jia2020population; @yang2020taking; @mu2020interplay], the extent to which the 
+increased levels of working from home, the consequent decrease in commuting flows, 
+and the altered structure of cities will remain post-pandemic is the subject
+of considerable debate in the literature.
 
 <!-- Covid-19 and working from home, cities, urban structure -->
-Multi-layered digital divides may also intersect with material divides
+Meanwhile, multi-layered digital divides may also intersect with material divides
 and the economic geography of the UK.
@@
 increase the spread of infection.
-We seek to answer how internet service resilience could contributes to or reduce 
+We seek to answer how internet service resilience could contribute to or reduce 
 economic resilience when the latter is dependent upon the capability to 
@@
 economies, but also fails, or at least frustrates a wide range of other urban 
-economies, as demonstrated by the number of tests per capita run in cluster $6$.
+economies, as demonstrated by the number of morning peak tests per capita run in 
+cluster $6$.
 Yet our analysis of the likelihood of furlough within the population, especially in 
-the smaller suburban clusters $7$, $3$, $5$, and also $8$ shows that digital divides 
+the smaller suburban clusters $7$, $3$, $5$, and $8$ shows that digital divides 
 were as likely to diverge as intersect during the pandemic, and did not necessarily 
@@
 # Discussion and Conclusions {#sec:5}
-
-This paper offers a new perspective on telecommuting from the viewpoint of the
-complex web of digital divides. We employ novel data regarding
-experienced upload speeds and time-series clustering methods,
-a family of unsupervised machine learning techniques which are rarely utilised
-in geographical research. Fast, reliable internet connections are
-necessary for the population to be able to work from home.
-Although not every place hosts individuals in occupations which allow for
-telecommuting nor with the necessary skills to effectively use the internet to
-telecommute, this paper raises the issue that places may depend upon good internet 
-reliability as well as connectivity to achieve economic resilience in a period like 
-the current pandemic when internet resilience is so vital.
-
 Our analysis demonstrated that the temporal profiles of seven of our
@@
 demand than the second largest cluster, $9$, where many more employees were
-furloughed.
-
+furloughed. Indeed, those with the skills and jobs to work from home were
+often left with the least robust services and greatest slowdown.
+
+<!-- WFH broader consequences-->
+Therefore, home-based digital infrastructure which considers upload speeds and 
+working day reliability as well as availability are likely to be particularly 
+important in a future where telecommuting might be a more common means of accessing 
+work and broadband services must be fit for purpose. 
+Although the long-term effects of such drastic changes in telecommuting and 
+attitudes towards working from home are difficult to predict, the reliability of 
+home broadband services deserve more consideration than in the past. 
+This would represent a switch from previous broadband policies in the UK which 
+tended to be aimed at supporting small and medium enterprises on industrial estates 
+and business parks []. Such policies also made some assumptions about the quality of 
+service in residential markets that could be assured without government intervention 
+- an assumption which our analysis suggest is not entirely accurate.
+
+Policy and development proposals should also consider how the potential changes 
+wrought by the pandemic span various aspects of economy and society. 
+Changes to transportation planning due to altered commuting patterns also imply 
+changes in land use and urban planning to accommodate 
+people who work from home [@BUDNITZ2020102713; @ELLDER2020102777].
+Productivity and innovation changes will reflect upon changes in agglomeration 
+externalities and the attraction of large cities [@econobs].
+Further research may be able to measure the economic resilience of the different
+clusters of places discussed in this paper once this pandemic is firmly past.
+However, our analysis demonstrates that the economic resilience made possible by
+working from home cannot be understood without considering the underpinning digital
+divides and cannot be achieved without planning for how the levels of digital,
+social and economic divides might intersect. For example, our results suggest that 
+broadband policies cannot improve the economic resilience of places where the 
+industrial structure does not align with occupations that incorporate the digital 
+skills and capabilities to work from home. Such places instead experienced higher 
+proportions of their labour force being placed on furlough.
+
+Early research on Covid-19 and cities also speculated upon the changes that 
+potential extensive post-Covid working from home patterns might generate for 
+spatial structure: from more walkable cities and more localised production and 
+consumption patterns to more extensive urban sprawl, the decline in public 
+transportation and increased private car usage [@batty2020editorial]. 
+Despite the essential role it played during the pandemic, less effort has been spent 
+in understanding the current and future role of digital infrastructure. 
+Our results indicate that probably for the first time the future of cities and 
+spatial structure are so intertwined with digital infrastructure. 
+If the post-covid world is a world with extensive working from home, 
+then we need to build, among other things, resilient digital infrastructure capable 
+of bridging the first layer of the digital divide. Contrary to previous broadband 
+business and deployment plans, emphasis should be placed not only on download,
+but also on upload speeds [@brake2020lessons]. In essence, Covid-19 accentuated the 
+old argument, now more valid than ever: digital infrastructure, just like any other 
+network infrastructure, only becomes visible when it stops working [@star1999; @tranos2013geography].
+
+On the other hand, the nuanced picture we gained through our analysis of the UK 
+case study suggests that being on the right side of the second level digital divide 
+had a greater impact on economic resilience and therefore the third level digital 
+divide, than having quality internet connectivity. 
+Our regressions results for cluster $1$ and $6$ are a demonstration of this. 
+Our analysis shows both the intersections and divergences of digital and economic 
+division.
 Almost all of the largest urban areas in the UK, as well as many smaller cities, 
@@
 the former and Milton Keynes in the latter [@technation2017].
-Yet there are both intersections and divergences of digital and economic divides.
-LADs in cluster $9$ were able to benefit both from reliable internet connections and
-populations which were familiar with working from home and could capitalise on their 
-digital infrastructure, although overall, they appeared to have a lower rate of
-return in the pandemic, with far greater numbers furloughed. 
-Is this an indication of both historic economic division, as well as the third level 
-digital divide in cities in the North and Midlands unable to capitalise on their 
+Yet whilst LADs in cluster $9$ were able to benefit both from reliable internet 
+connections and populations familiar with working from home to capitalise on  their 
+digital infrastructure, they appeared to have a lower rate of return in 
+the pandemic, with far greater numbers furloughed. 
+Is this an indication of historic economic division, as well as the third level 
+digital divide, in cities in the North and Midlands unable to capitalise on their 
 digital infrastructure?
@@
 those in this cluster be more resilient and gain greater returns, despite using 
-less reliable internet services?
-The nuanced picture we gained through our analysis of the UK case study suggests
-that in cluster $1$ as well as $6$, being on the right side of the second level
-digital divide had a greater impact on economic resilience and therefore the third
-level digital divide, than having quality internet connectivity.
-
-<!-- WFH broader consequences-->
-Digital infrastructure which considers upload speeds and working day reliability as 
-well as availability are likely to be particularly important in a future where
-telecommuting might be a more common means of accessing work and broadband services 
-must be fit for purpose, although the long-term effects of such drastic changes in 
-telecommuting and attitudes towards working from home are difficult to predict. 
-Nevertheless, they span various aspects of economy and society: from changes to
-transportation planning due to altered commuting patterns to changes in land
-use and urban planning to accommodate people who work from home
-[@BUDNITZ2020102713; @ELLDER2020102777], and from productivity
-and innovation changes to changes in agglomeration externalities and the
-attraction of large cities [@econobs].
-Further research may be able to measure the economic resilience of the different
-clusters of places discussed in this paper once this pandemic is firmly past.
-However, our analysis demonstrates that the economic resilience made possible by
-working from home cannot be understood without considering the underpinning digital
-divides and cannot be achieved without planning for how the levels of digital,
-social and economic divides might intersect.
-
-Early research on Covid-19 and cities speculated the changes that potential extensive
-post-Covid working from home patterns might generate for spatial structure: from 
-more walkable cities to more localised production and consumption patterns and from
-more extensive urban sprawl to decline in public transportation and increased private 
-car usage [@batty2020editorial]. Despite the essential role it played during the 
-pandemic, less effort has been spent in understanding the current and future role
-of digital infrastructure. Our results indicate that probably for the first time 
-the future of cities and spatial structure are so much intertwined with digital 
-infrastructure. If the post-covid world is a world with extensive working from home, 
-then we need to build, among other things, resilient digital infrastructure capable 
-of bridging the first layer of the digital divide. Contrary to previous broadband 
-business and deployment plans, emphasis should be placed not only on download,
-but also on upload speeds [@brake2020lessons. In essence, Covid-19 accentuated the 
-old, but more valid than ever argument: digital infrastructure, just like any other 
-network infrastructure, only becomes visible when it stops working [@star1999; @tranos2013geography].
+less reliable internet services? Further research would be necessary to prise apart
+the detail, particularly in the largest clusters where averages are less distinct.
+
+In conclusion, this paper offers a new perspective on telecommuting from the 
+viewpoint of the complex web of digital divides. We employ novel data regarding
+experienced upload speeds and time-series clustering methods,
+a family of unsupervised machine learning techniques which are rarely utilised
+in geographical research. Fast, reliable internet connections are
+necessary for the population to be able to work from home.
+Although not every place hosts individuals in occupations which allow for
+telecommuting nor with the necessary skills to effectively use the internet to
+telecommute, this paper raises the issue that places may depend upon good internet 
+reliability as well as connectivity to achieve economic resilience in a period like 
+the current pandemic when internet resilience is so vital.
 

### 2021-03-24 paper/v2_taylor_francis/response_reviewers.md

@@
 
-Regrettably, we did not have the opportunity to run a survey. As we say in the introduction,
-prior to the pandemic, the demand for telecommuting would not have put a strain
-on internet connection speeds. We have clarified this with the addition of the following sentence:
+Regrettably, we did not have the opportunity to run a survey. 
+As we say in the introduction, prior to the pandemic, the demand for telecommuting 
+would not have put a strain on internet connection speeds. We have clarified this 
+with the addition of the following sentence:
 "High levels of demand are one of the main causes of reduced reliability and slower connection speeds, as network bandwidth becomes congested".
@@
 could work and who could not at the height of the first wave of the pandemic.
+Finally, as we note at the end of the methods and data section, an auxiliary 
+regression was the preferred methodology because:
+"Such individual data could be found though surveys, but these offer less
+detailed information about the experience of internet resilience due to enforced
+demand, which is the main contribution of this paper."
 
@@
 
-We have now added the following paragraph in the 'Digital divides and economic resilience'
-section.
+We have now added the following short discussion in the 'Digital divides and 
+economic resilience' section.
 
-"Currently there is a hotly debated discussion in the literature regarding these
-exact changes in transportation and the structure of cities. We now have enough
-hard data to observe the drastic change in the space-time geography of cities around
-the world [e.g. @google2020; @shibayama2021impact].
-Although there is a broad agreement that these changes
+"Such digital intersections and efficiencies have also changed 
+transportation, connectivity and the structure of cities during the pandemic.
+We now have enough data to confirm the drastic alteration observed in the space-time 
+geography of cities around the world in the last year [e.g. @google2020; @shibayama2021impact]. 
+Yet, although there is a broad agreement that these changes 
 during the pandemic have played a pivotal role in stopping the spread of the virus 
-[@jia2020population; @yang2020taking; @mu2020interplay],
-we still have no evidence to what extend these increased levels of working from
-home, and the consequent decrease in commuting flows and the structure of cities,
-will remain post-pandemic".
+[@jia2020population; @yang2020taking; @mu2020interplay], the extent to which the 
+increased levels of working from home, the consequent decrease in commuting flows, 
+and the altered structure of cities will remain post-pandemic is the subject
+of considerable debate in the literature."
 
@@
 Following your suggestion, we have now opted to present the cluster analysis and
-the subsequent multinomial regression using $k = 13$. We indeed believe that the
+the subsequent multinomial regression using $k = 9$. We indeed believe that the
 results are more clear now. we have added the following justification in the Methods
@@
 paragraph of the results section and then expanded the conclusion section to include
-a discussion of the results.
+a more in-depth discussion of the results and how they link to some of the issues you 
+suggest at the start. We mention some uncertainties or avenues for future research,
+but also see our response to Reviewer 1, point 1 for an explanation of biases.
 

### 2021-03-25 paper/v2_taylor_francis/response_reviewers.md
new appendix with 2019 clusters, text for discussion, new images of higher res, ref and footnote for openair added

@@
 
-ET: after we finalise k, we can rerun k-means for 2019 and decide - could also show mean speeds on a map for 2019 - this would be part of the response to reviewer 1 q2 and q6.
-**Let's discuss this**
+This is useful. We have not added a new Appendix 2, which plots the clusters for 
+the same period in 2019. The picture is very different, which is an indication of 
+the changes in internet usage during the pandemic.
+
+"It is worth noting here that clusters of LAD based on the upload speeds are very 
+different to the ones occur when we apply the same methods to upload speed data
+for the same time period in 2019 -- see Appendix 2. This drastic 
+difference between the LAD upload speed profiles before and during the pandemic is
+an indication of the changes in internet usage that took place during the pandemic."
 

### 2021-03-25 paper/v2_taylor_francis/broadband.speed.covid.Rmd
new appendix with 2019 clusters, text for discussion, new images of higher res, ref and footnote for openair added

@@
 this trend masks substantial variation by not only the day of the week,
-but also time of day, as can be seen in Figure \ref{test2020}.
+but also time of day, as can be seen in Figure \ref{test2020}
+^[Figures \ref{test2020} and \ref{test2019} were created with the `R` package `openair` [@openair].].
 Thus, a daily aggregation of upload speeds would mask the variation in experienced
@@
 the day.
+
+It is worth noting here that clusters of LAD based on the upload speeds are very 
+different to the ones occur when we apply the same methods to upload speed data
+for the same time period in 2019 -- see Appendix [2](#appendix2). This drastic 
+difference between the LAD upload speed profiles before and during the pandemic is
+an indication of the changes in internet usage that took place during the pandemic.
 
@@
 between $9:00$-$10:59$ as well as the share of fast Virgin Media internet
-connections. Regarding the former, we expect people in LADs with more unreliable
+connections
+^[See Appendix [3](#appendix3) for the descriptive statistics.].
+Regarding the former, we expect people in LADs with more unreliable
 connections to test their internet speeds more often, and the results show that 
@@
 home broadband services deserve more consideration than in the past. 
-This would represent a switch from previous broadband policies in the UK which 
-tended to be aimed at supporting small and medium enterprises on industrial estates 
-and business parks []. Such policies also made some assumptions about the quality of 
+This would represent a switch from previous demand-side broadband policies in the UK
+which tended to be aimed at supporting small and medium enterprises [@HENDERSON2020102024].
+SUch policies were based on previous research regarding the productivity effects 
+of broadband infrastructure [@DESTEFANO2018110]. However, this stream of research 
+tended not to consider residential locations as places that host economic activity.
+Such policies also made some assumptions about the quality of 
 service in residential markets that could be assured without government intervention 
-- an assumption which our analysis suggest is not entirely accurate.
+-- an assumption which our analysis suggest is not entirely accurate.
 
@@
 
-<!-- \pagebreak -->
+\pagebreak
 
@@
 
+```{r echo=FALSE, out.width = "100%",fig.cap="\\label{map.up.2019.clusters}Upload speed clusters for LADs in before the pandemic (2019)"}
+img_path <- "figures/map.up.2019.clusters.png"
+include_graphics(img_path)
+```
+
+\pagebreak
+
+# Appendix 3 {- #appendix3}
+
 ```{r echo=FALSE, message=FALSE, results = "asis"}
@@
 ```
+
+\pagebreak
+
+# References

### 2021-03-25 paper/v2_taylor_francis/broadband.speed.covid.tex
new appendix with 2019 clusters, text for discussion, new images of higher res, ref and footnote for openair added

@@
 socioeconomically. Our analysis enables us to better understand how the
-spatial and social distribution of education, occupation and online
-accessibility intersect to enable or hinder the practice of
-telecommuting at a time of extreme demand. We also consider what lessons
-can be learned from this time for a future where telecommuting is likely
-to remain a more common means of accessing work, at least in comparison
-to the pre-Covid era.
+spatial and social distribution of occupation and online accessibility
+intersect to enable or hinder the practice of telecommuting at a time of
+extreme demand. We also consider what lessons can be learned from this
+time for a future where telecommuting is likely to remain a more common
+means of accessing work, at least in comparison to the pre-Covid era.
 
@@
 that has not been improved by policy measures to increase competition
-\citep{riddlesden2014broadband, nardotto2015unbundling}. Moreover,
+\citep{riddlesden2014broadband, nardotto2015unbundling}. In contrast,
 \citet{riddlesden2014broadband} found that levels of deprivation did not
@@
 However, as \citet{blank2018local} highlight, variation in individual
-internet uptake and use is a product of more complex spatial and
-demographic characteristics than simple rurality or urbanisation. Dense
-urban areas were shown to suffer more from slowdown during peak hours,
-although these services were more likely to be improved by increased
-competition between providers, such as between new entrants and Virgin
-Media cable connections
+internet availability and uptake is a product of more complex spatial
+and demographic characteristics than simple rurality or urbanisation.
+Dense urban areas were shown to suffer more from slowdown during peak
+hours, although these services were more likely to be improved by
+increased competition between providers, such as between new entrants
+and Virgin Media cable connections
 \citep{riddlesden2014broadband, nardotto2015unbundling}. The latter were
@@
 telecommuting has not previously been a cause of concentrated demand and
-pressure on internet services. Other intersections between digital
-divides have been subject to study. There are indications that those who
-purchase high speed connections consume more data of all sorts and use
-their connections for a greater variety of purposes
-\citep{hauge2011consumer}. There is also a correlation between access to
-internet services and a reduction in household transport spend
-\citep{bris2017ict}. These studies suggest that better internet services
-enable households to make savings and efficiencies, an example of the
-first level digital divide reinforcing the third level.
-
-Currently there is a hotly debated discussion in the literature
-regarding these exact changes in transportation and the structure of
-cities. We now have enough hard data to observe the drastic change in
-the space-time geography of cities around the world
-\citep[e.g.~][]{google2020, shibayama2021impact}. Although there is a
-broad agreement that these changes during the pandemic have played a
-pivotal role in stopping the spread of the virus
-\citep{jia2020population, yang2020taking, mu2020interplay}, we still
-have no evidence to what extend these increased levels of working from
-home, and the consequent decrease in commuting flows and the structure
-of cities, will remain post-pandemic.
-
-Multi-layered digital divides may also intersect with material divides
-and the economic geography of the UK. The regional economic resilience
-literature underlines the differentiated capacity of cities and regions
-to escape or recover from economic crises
+pressure on internet services.
+
+Other intersections between digital divides have been subject to study.
+There are indications that those who purchase high speed connections
+consume more data of all sorts and use their connections for a greater
+variety of purposes \citep{hauge2011consumer}. There is also a
+correlation between access to internet services and a reduction in
+household transport spend \citep{bris2017ict}. These studies suggest
+that better internet services enable households to make savings and
+efficiencies, an example of the first level digital divide reinforcing
+the third level. Such digital intersections and efficiencies have also
+changed transportation, connectivity and the structure of cities during
+the pandemic. We now have enough data to confirm the drastic alteration
+observed in the space-time geography of cities around the world in the
+last year \citep[e.g.~][]{google2020, shibayama2021impact}. Yet,
+although there is a broad agreement that these changes during the
+pandemic have played a pivotal role in stopping the spread of the virus
+\citep{jia2020population, yang2020taking, mu2020interplay}, the extent
+to which the increased levels of working from home, the consequent
+decrease in commuting flows, and the altered structure of cities will
+remain post-pandemic is the subject of considerable debate in the
+literature.
+
+Meanwhile, multi-layered digital divides may also intersect with
+material divides and the economic geography of the UK. The regional
+economic resilience literature underlines the differentiated capacity of
+cities and regions to escape or recover from economic crises
 \citep{martin2012regional, kitsos2018economic}. As different places have
@@
 face-to-face contact that could increase the spread of infection. We
-seek to answer how internet service resilience could contributes to or
+seek to answer how internet service resilience could contribute to or
 reduce economic resilience when the latter is dependent upon the
@@
 not only the day of the week, but also time of day, as can be seen in
-Figure \ref{test2020}. Thus, a daily aggregation of upload speeds would
-mask the variation in experienced service over the course of each
-weekday. Furthermore, the importance of this variation is highlighted by
-a comparison with the same period in \(2019\), as in Figure
-\ref{test2019}, when the volume of testing and thus of experienced
-internet service quality peaked in the evening, presumably in response
-to demand for leisure activities and download speeds. In contrast, the
-majority of the increase in testing in \(2020\) is during the working
-day, creating a new morning peak in Figure \ref{test2020}. Therefore, we
-include a measure of hourly variation in our temporal profiles to
-reflect the change in users' perception of the workday reliability of
-internet services.
+Figure \ref{test2020} \footnote{Figures \ref{test2020} and
+  \ref{test2019} were created with the \texttt{R} package
+  \texttt{openair} \citep{openair}.}. Thus, a daily aggregation of
+upload speeds would mask the variation in experienced service over the
+course of each weekday. Furthermore, the importance of this variation is
+highlighted by a comparison with the same period in \(2019\), as in
+Figure \ref{test2019}, when the volume of testing and thus of
+experienced internet service quality peaked in the evening, presumably
+in response to demand for leisure activities and download speeds. In
+contrast, the majority of the increase in testing in \(2020\) is during
+the working day, creating a new morning peak in Figure \ref{test2020}.
+Therefore, we include a measure of hourly variation in our temporal
+profiles to reflect the change in users' perception of the workday
+reliability of internet services.
 
@@
 
+It is worth noting here that clusters of LAD based on the upload speeds
+are very different to the ones occur when we apply the same methods to
+upload speed data for the same time period in 2019 -- see Appendix
+\protect\hyperlink{appendix2}{2}. This drastic difference between the
+LAD upload speed profiles before and during the pandemic is an
+indication of the changes in internet usage that took place during the
+pandemic.
+
 In summary, the reliability of internet services during the working day
@@
 inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
-Virgin Media internet connections. Regarding the former, we expect
-people in LADs with more unreliable connections to test their internet
-speeds more often, and the results show that those in cluster \(6\) are
-by far the most concerned about reliability at that time of day.
-Meanwhile, those in clusters \(2\), \(3\), and \(9\) benefit from a
-higher proportion of Virgin connections, which is an indication that
-people in these clusters are more likely to live in urban areas, with
-more profitable broadband markets, and a better choice of broadband
-services. We also employ distance from London and from the nearest
-metropolitan area (including London) as two variables depicting
-peripherality and urban structure. However, whilst significant, the
-pattern of coefficients of these variables is inconclusive in confirmed
-first level digital divides. Even though London was one of the ten
-largest metropolitan areas in England, which, along with Glasgow and
-Cardiff, were identified to calculate the variable estimating the impact
-of distance from the centre of a metropolitan area, the coefficients for
-the two variables are mostly small and the signs for some are opposite.
+Virgin Media internet connections \footnote{See Appendix
+  \protect\hyperlink{appendix3}{3} for the descriptive statistics.}.
+Regarding the former, we expect people in LADs with more unreliable
+connections to test their internet speeds more often, and the results
+show that those in cluster \(6\) are by far the most concerned about
+reliability at that time of day. Meanwhile, those in clusters \(2\),
+\(3\), and \(9\) benefit from a higher proportion of Virgin connections,
+which is an indication that people in these clusters are more likely to
+live in urban areas, with more profitable broadband markets, and a
+better choice of broadband services. We also employ distance from London
+and from the nearest metropolitan area (including London) as two
+variables depicting peripherality and urban structure. However, whilst
+significant, the pattern of coefficients of these variables is
+inconclusive in confirmed first level digital divides. Even though
+London was one of the ten largest metropolitan areas in England, which,
+along with Glasgow and Cardiff, were identified to calculate the
+variable estimating the impact of distance from the centre of a
+metropolitan area, the coefficients for the two variables are mostly
+small and the signs for some are opposite.
 
@@
 economies, but also fails, or at least frustrates a wide range of other
-urban economies, as demonstrated by the number of tests per capita run
-in cluster \(6\). Yet our analysis of the likelihood of furlough within
-the population, especially in the smaller suburban clusters \(7\),
-\(3\), \(5\), and also \(8\) shows that digital divides were as likely
-to diverge as intersect during the pandemic, and did not necessarily
-overlap with prior economic divides. With a much greater share of the
-population likely to continue to work from home in the future, and with
-changing attitudes towards residential locations, our analysis suggests
-that first level digital divides should be seen as a function of
-reliability as well as availability, of upload speeds as well as
-download speeds, and of greater spatial variation than might have
-previously been considered.
+urban economies, as demonstrated by the number of morning peak tests per
+capita run in cluster \(6\). Yet our analysis of the likelihood of
+furlough within the population, especially in the smaller suburban
+clusters \(7\), \(3\), \(5\), and \(8\) shows that digital divides were
+as likely to diverge as intersect during the pandemic, and did not
+necessarily overlap with prior economic divides. With a much greater
+share of the population likely to continue to work from home in the
+future, and with changing attitudes towards residential locations, our
+analysis suggests that first level digital divides should be seen as a
+function of reliability as well as availability, of upload speeds as
+well as download speeds, and of greater spatial variation than might
+have previously been considered.
 
@@
 \section{Discussion and Conclusions}\label{sec:5}}
-
-This paper offers a new perspective on telecommuting from the viewpoint
-of the complex web of digital divides. We employ novel data regarding
-experienced upload speeds and time-series clustering methods, a family
-of unsupervised machine learning techniques which are rarely utilised in
-geographical research. Fast, reliable internet connections are necessary
-for the population to be able to work from home. Although not every
-place hosts individuals in occupations which allow for telecommuting nor
-with the necessary skills to effectively use the internet to
-telecommute, this paper raises the issue that places may depend upon
-good internet reliability as well as connectivity to achieve economic
-resilience in a period like the current pandemic when internet
-resilience is so vital.
 
@@
 because they had higher levels of demand than the second largest
-cluster, \(9\), where many more employees were furloughed.
-
-Almost all of the largest urban areas in the UK, as well as many smaller
-cities, were split between the largest two clusters -- \(6\) and \(9\)
--- and both contain LADs that are centres of the knowledge economy, from
-Cambridge in the former to Oxford in the latter, or have high
-concentrations of digital businesses, like Reading in the former and
-Milton Keynes in the latter \citep{technation2017}. Yet there are both
-intersections and divergences of digital and economic divides. LADs in
-cluster \(9\) were able to benefit both from reliable internet
-connections and populations which were familiar with working from home
-and could capitalise on their digital infrastructure, although overall,
+cluster, \(9\), where many more employees were furloughed. Indeed, those
+with the skills and jobs to work from home were often left with the
+least robust services and greatest slowdown.
+
+Therefore, home-based digital infrastructure which considers upload
+speeds and working day reliability as well as availability are likely to
+be particularly important in a future where telecommuting might be a
+more common means of accessing work and broadband services must be fit
+for purpose. Although the long-term effects of such drastic changes in
+telecommuting and attitudes towards working from home are difficult to
+predict, the reliability of home broadband services deserve more
+consideration than in the past. This would represent a switch from
+previous demand-side broadband policies in the UK which tended to be
+aimed at supporting small and medium enterprises
+\citep{HENDERSON2020102024}. SUch policies were based on previous
+research regarding the productivity effects of broadband infrastructure
+\citep{DESTEFANO2018110}. However, this stream of research tended not to
+consider residential locations as places that host economic activity.
+Such policies also made some assumptions about the quality of service in
+residential markets that could be assured without government
+intervention -- an assumption which our analysis suggest is not entirely
+accurate.
+
+Policy and development proposals should also consider how the potential
+changes wrought by the pandemic span various aspects of economy and
+society. Changes to transportation planning due to altered commuting
+patterns also imply changes in land use and urban planning to
+accommodate people who work from home
+\citep{BUDNITZ2020102713, ELLDER2020102777}. Productivity and innovation
+changes will reflect upon changes in agglomeration externalities and the
+attraction of large cities \citep{econobs}. Further research may be able
+to measure the economic resilience of the different clusters of places
+discussed in this paper once this pandemic is firmly past. However, our
+analysis demonstrates that the economic resilience made possible by
+working from home cannot be understood without considering the
+underpinning digital divides and cannot be achieved without planning for
+how the levels of digital, social and economic divides might intersect.
+For example, our results suggest that broadband policies cannot improve
+the economic resilience of places where the industrial structure does
+not align with occupations that incorporate the digital skills and
+capabilities to work from home. Such places instead experienced higher
+proportions of their labour force being placed on furlough.
+
+Early research on Covid-19 and cities also speculated upon the changes
+that potential extensive post-Covid working from home patterns might
+generate for spatial structure: from more walkable cities and more
+localised production and consumption patterns to more extensive urban
+sprawl, the decline in public transportation and increased private car
+usage \citep{batty2020editorial}. Despite the essential role it played
+during the pandemic, less effort has been spent in understanding the
+current and future role of digital infrastructure. Our results indicate
+that probably for the first time the future of cities and spatial
+structure are so intertwined with digital infrastructure. If the
+post-covid world is a world with extensive working from home, then we
+need to build, among other things, resilient digital infrastructure
+capable of bridging the first layer of the digital divide. Contrary to
+previous broadband business and deployment plans, emphasis should be
+placed not only on download, but also on upload speeds
+\citep{brake2020lessons}. In essence, Covid-19 accentuated the old
+argument, now more valid than ever: digital infrastructure, just like
+any other network infrastructure, only becomes visible when it stops
+working \citep{star1999, tranos2013geography}.
+
+On the other hand, the nuanced picture we gained through our analysis of
+the UK case study suggests that being on the right side of the second
+level digital divide had a greater impact on economic resilience and
+therefore the third level digital divide, than having quality internet
+connectivity. Our regressions results for cluster \(1\) and \(6\) are a
+demonstration of this. Our analysis shows both the intersections and
+divergences of digital and economic division. Almost all of the largest
+urban areas in the UK, as well as many smaller cities, were split
+between the largest two clusters -- \(6\) and \(9\) -- and both contain
+LADs that are centres of the knowledge economy, from Cambridge in the
+former to Oxford in the latter, or have high concentrations of digital
+businesses, like Reading in the former and Milton Keynes in the latter
+\citep{technation2017}. Yet whilst LADs in cluster \(9\) were able to
+benefit both from reliable internet connections and populations familiar
+with working from home to capitalise on their digital infrastructure,
 they appeared to have a lower rate of return in the pandemic, with far
-greater numbers furloughed. Is this an indication of both historic
-economic division, as well as the third level digital divide in cities
-in the North and Midlands unable to capitalise on their digital
-infrastructure? Meanwhile, did the greater presence in London and the
-South of LADs in cluster \(6\), as well as the inclusion of many rural
-areas less impacted by the pandemic help those in this cluster be more
-resilient and gain greater returns, despite using less reliable internet
-services? The nuanced picture we gained through our analysis of the UK
-case study suggests that in cluster \(1\) as well as \(6\), being on the
-right side of the second level digital divide had a greater impact on
-economic resilience and therefore the third level digital divide, than
-having quality internet connectivity.
-
-Digital infrastructure which considers upload speeds and working day
-reliability as well as availability are likely to be particularly
-important in a future where telecommuting might be a more common means
-of accessing work and broadband services must be fit for purpose,
-although the long-term effects of such drastic changes in telecommuting
-and attitudes towards working from home are difficult to predict.
-Nevertheless, they span various aspects of economy and society: from
-changes to transportation planning due to altered commuting patterns to
-changes in land use and urban planning to accommodate people who work
-from home \citep{BUDNITZ2020102713, ELLDER2020102777}, and from
-productivity and innovation changes to changes in agglomeration
-externalities and the attraction of large cities \citep{econobs}.
-Further research may be able to measure the economic resilience of the
-different clusters of places discussed in this paper once this pandemic
-is firmly past. However, our analysis demonstrates that the economic
-resilience made possible by working from home cannot be understood
-without considering the underpinning digital divides and cannot be
-achieved without planning for how the levels of digital, social and
-economic divides might intersect.
-
-Early research on Covid-19 and cities speculated the changes that
-potential extensive post-Covid working from home patterns might generate
-for spatial structure: from more walkable cities to more localised
-production and consumption patterns and from more extensive urban sprawl
-to decline in public transportation and increased private car usage
-\citep{batty2020editorial}. Despite the essential role it played during
-the pandemic, less effort has been spent in understanding the current
-and future role of digital infrastructure. Our results indicate that
-probably for the first time the future of cities and spatial structure
-are so much intertwined with digital infrastructure. If the post-covid
-world is a world with extensive working from home, then we need to
-build, among other things, resilient digital infrastructure capable of
-bridging the first layer of the digital divide. Contrary to previous
-broadband business and deployment plans, emphasis should be placed not
-only on download, but also on upload speeds {[}\citet{brake2020lessons}.
-In essence, Covid-19 accentuated the old, but more valid than ever
-argument: digital infrastructure, just like any other network
-infrastructure, only becomes visible when it stops working
-\citep{star1999, tranos2013geography}.
+greater numbers furloughed. Is this an indication of historic economic
+division, as well as the third level digital divide, in cities in the
+North and Midlands unable to capitalise on their digital infrastructure?
+Meanwhile, did the greater presence in London and the South of LADs in
+cluster \(6\), as well as the inclusion of many rural areas less
+impacted by the pandemic help those in this cluster be more resilient
+and gain greater returns, despite using less reliable internet services?
+Further research would be necessary to prise apart the detail,
+particularly in the largest clusters where averages are less distinct.
+
+In conclusion, this paper offers a new perspective on telecommuting from
+the viewpoint of the complex web of digital divides. We employ novel
+data regarding experienced upload speeds and time-series clustering
+methods, a family of unsupervised machine learning techniques which are
+rarely utilised in geographical research. Fast, reliable internet
+connections are necessary for the population to be able to work from
+home. Although not every place hosts individuals in occupations which
+allow for telecommuting nor with the necessary skills to effectively use
+the internet to telecommute, this paper raises the issue that places may
+depend upon good internet reliability as well as connectivity to achieve
+economic resilience in a period like the current pandemic when internet
+resilience is so vital.
+
+\pagebreak
 
@@
 
+\begin{figure}
+\includegraphics[width=1\linewidth]{figures/map.up.2019.clusters} \caption{\label{map.up.2019.clusters}Upload speed clusters for LADs in before the pandemic (2019)}\label{fig:unnamed-chunk-17}
+\end{figure}
+
+\pagebreak
+
+\hypertarget{appendix3}{%
+\section*{Appendix 3}\label{appendix3}}
+\addcontentsline{toc}{section}{Appendix 3}
+
 \begin{table}[!htbp] \centering 
@@
 
+\pagebreak
+
 \bibliographystyle{tfcad}

### 2021-03-25 src/descriptive_Clusters.Rmd
leftovers -- sorry

--- 2021-03-13 src/descriptive_Clusters.Rmd
@@
 
+## Time variation plots (Figures 1 and 2)
+
 ```{r include = FALSE}
@@
 names(TS2020)[13] <- "test frequency 2020"
-time.var.plot2020 <- timeVariation(TS2020, pollutant = "test frequency 2020", statistic = "mean")
+time.var.plot2020 <- timeVariation(TS2020, pollutant = "test frequency 2020", statistic = "mean",
+                                   par.settings=list(fontsize=list(text=8)))
 
 #export time variation plots for paper
-out.plot <- paste(path, "/paper/v1/figures/time.var.plot2019.png", sep = "")
-png(out.plot)
+out.plot <- paste(path, "/paper/v2_taylor_francis/figures/time.var.plot2019.png", sep = "")
+png(out.plot, res = 100, width = 615, height = 615) # 72dpi default res. 480x480. 
+                                                    # ((100-72)/100)*480+480 = 615 
 time.var.plot2019
-dev.off()
+dev.off() # if dev.off() doesn't work try dev.set(dev.next()) 2-3 times
 
-out.plot <- paste(path, "/paper/v1/figures/time.var.plot2020.png", sep = "")
-png(out.plot)
+out.plot <- paste(path, "/paper/v2_taylor_francis/figures/time.var.plot2020.png", sep = "")
+png(out.plot, res = 100, width = 615, height = 615)
 time.var.plot2020
 dev.off()
+```
 
+```{r}
 #join clusters to TS2020
@@
 
+## Summarise 2020 test frequency clusters by hour of each weekday and plot
+
 ```{r }
-#summarise test frequency clusters by hour of each weekday and plot
 cluster.tests <- TS2020 %>%
@@
                      c("Mon", "Tues", "Weds", "Thurs", "Fri")) + geom_point()
-#summarise upload speed clusters by hour of each weekday and plot
-#plot largest 6 clusters - all over 500k pop
+```
+
+## Summarise 2020 upload speed clusters by hour of each weekday and plot
+
+```{r}
+
+# plot largest 6 clusters - all over 500k pop
 cluster.up <- TS2020 %>%
@@
 cluster.up$clusterID <- as.factor(cluster.up$clusterID)
-#add fake dates for plotting
+
+# add fake dates for plotting
 help.dates <- tibble(help.date.time = c(
@@
 cluster.up$help.date.time <- with(cluster.up, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
-#plot cluster statistic for mean upload speed
+
+# plot cluster statistic for mean upload speed
 UpCluster <- ggplot(cluster.up, aes(help.date.time, mean.up, 
@@
 
-#export UpClusterL plot for paper
-out.plot <- paste(path, "/paper/v1/figures/UpCluster.png", sep = "")
-png(out.plot)
-UpClusterL
-dev.off()
+#export UpCluster plot for paper
+out.plot <- paste(path, "/paper/v2_taylor_francis/figures/UpCluster.png", sep = "")
+UpCluster
+ggsave(out.plot)
 
-#plot next 6 smaller clusters, but exclude 2 clusters with 1 LAD and under 100k pop
+# EXCLUDED from paper
+# plot next 6 smaller clusters, but exclude 2 clusters with 1 LAD and under 100k pop
 cluster.upS <- TS2020 %>%
@@
 cluster.upS$clusterID <- as.factor(cluster.upS$clusterID)
-#add fake dates for plotting
+
+# add fake dates for plotting
 help.dates <- tibble(help.date.time = c(
@@
 cluster.upS$help.date.time <- with(cluster.upS, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
-#plot cluster statistic for mean upload speed
+
+# plot cluster statistic for mean upload speed
 UpClusterS <- ggplot(cluster.upS, aes(help.date.time, mean.up, 
@@
 
+# The figure with the small clusters is not included in the paper
 #export UpClusterS plot for paper
-#out.plot <- paste(path, "/paper/v1/figures/UpClusterS.png", sep = "")
+#out.plot <- paste(path, "/paper/v2_taylor_francis/figures/UpClusterS.png", sep = "")
 #png(out.plot)
@@
 #dev.off() 
+```
+
+## Summarise 2019 upload speed clusters by hour of each weekday and plot
+
+This is a test to compare 2020 vs. 2019 line graphs.
+
+```{r}
+
+# plot largest 6 clusters - all over 500k pop
+cluster.up <- TS2019 %>%
+  group_by(cluster.up, weekday, hour) %>%
+  summarise(mean.up = mean(speedup))
+names(cluster.up)[1] <- "clusterID"
+names(LAD.cluster.up)[1] <- "clusterID"
+cluster.up <- left_join(cluster.up, LAD.cluster.up)
+cluster.up <- cluster.up[which(cluster.up$up.LADs > 4),]
+cluster.up$clusterID <- as.factor(cluster.up$clusterID)
+
+# add fake dates for plotting
+help.dates <- tibble(help.date.time = c(
+  make_datetime(year = 2019, month = 3, day = 4),
+  make_datetime(year = 2019, month = 3, day = 5),
+  make_datetime(year = 2019, month = 3, day = 6),
+  make_datetime(year = 2019, month = 3, day = 7),
+  make_datetime(year = 2019, month = 3, day = 8)),
+                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+cluster.up <- merge(cluster.up, help.dates, by = "weekday")
+cluster.up$hour <- paste(cluster.up$hour, ":00", sep="")
+cluster.up$help.date.time <- with(cluster.up, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+
+# plot cluster statistic for mean upload speed
+UpCluster2019 <- ggplot(cluster.up, aes(help.date.time, mean.up, 
+                                      colour = clusterID)) +
+  theme_minimal() +
+  geom_line() + 
+  labs(x = "Weekdays March-May", y = "mean upload speed", col = "Cluster ID") + 
+  scale_x_datetime(breaks = as.POSIXct(c("2019-03-04 15:00:00", 
+                    "2019-03-05 15:00:00", "2019-03-06 15:00:00",
+                    "2019-03-07 15:00:00", "2019-03-08 15:00:00"), 
+                    labels = "%a"), labels = 
+                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + 
+  geom_point() +
+  theme(legend.position = c(0.7, 0.08),
+        legend.direction = "horizontal")
+
+#export UpCluster plot for paper
+out.plot <- paste(path, "/paper/v2_taylor_francis/figures/UpCluster2019.png", sep = "")
+UpCluster2019
+ggsave(out.plot)
+```
+
+```{r eval=F}
+# Not used in the paper
 

### 2021-03-25 src/ts_clusters_k9.Rmd
leftovers -- sorry

@@
 
-### Upload speed
+## Line graphs for upload speed (not clusters) 2019 and 2020
+
+```{r}
+
+#test <- 
+line2020 <- ts.up2020 %>% 
+  mutate(day = recode(weekday, # these are dates from 2020
+                            Mon=2,
+                            Tue = 3,
+                            Wed = 4,
+                            Thu = 5,
+                            Fri = 6)) %>% 
+  mutate(help.date.time = make_datetime(year = 2020, month = 3, day = day),
+         hour = paste0(hour, ":00"),
+         help.date.time = (as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))) %>% 
+  group_by(help.date.time) %>%
+  summarise(mean.up = mean(mean.up)) %>% 
+  ggplot(aes(x=help.date.time, y=mean.up), #, group = LAD19NM, colour = LAD19NM),
+         show.legend = F) + 
+  geom_line() + guides(colour=FALSE) +
+  xlab(2020)
+  # + xlab("Year") +
+  # ylab("Pareto exponent")
+
+# library(plotly)
+# ggplotly(test,
+#          show.legend=F)
+
+line2019 <- ts.up2019 %>% 
+  mutate(day = recode(weekday, # these are dates from 2020
+                            Mon=2,
+                            Tue = 3,
+                            Wed = 4,
+                            Thu = 5,
+                            Fri = 6)) %>% 
+  mutate(help.date.time = make_datetime(year = 2020, month = 3, day = day),
+         hour = paste0(hour, ":00"),
+         help.date.time = (as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))) %>% 
+  group_by(help.date.time) %>%
+  summarise(mean.up = mean(mean.up)) %>% 
+  ggplot(aes(x=help.date.time, y=mean.up), #, group = LAD19NM, colour = LAD19NM),
+         show.legend = F) + 
+  geom_line() + guides(colour=FALSE)+
+  xlab(2019)
+
+grid.arrange(line2019, line2020, ncol = 1, nrow = 2) #legend
+```
+
+## Upload speed 2020
 
@@
 
-### Download speed
+## Download speed 2020
 
@@
 
-### N. of speed tests
+## N. of speed tests 2020
 
@@
 
-### Upload speed diff 2019-2020
+## Upload speed diff 2019-2020
 
@@
 
-### download speed diff 2019-2020
+## download speed diff 2019-2020
 
@@
 
-## Mapping the clusters
+## Upload speed 2019
+
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# 2019
+
+# create a 'fake' week for a ts id
+help.dates <- tibble(help.date.time = c(make_datetime(year = 2019, month = 3, day = 4),
+                make_datetime(year = 2019, month = 3, day = 5),
+                make_datetime(year = 2019, month = 3, day = 6),
+                make_datetime(year = 2019, month = 3, day = 7),
+                make_datetime(year = 2019, month = 3, day = 8)),
+                weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
+ts.wide2019 <- merge(ts.up2019, help.dates, by = "weekday")
+ts.wide2019$hour <- paste(ts.wide2019$hour, ":00", sep="")
+ts.wide2019$help.date.time <- with(ts.wide2019, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
+
+ts.wide2019 <- ts.wide2019 %>%
+  dplyr::select(help.date.time, LAD19NM, mean.up) #%>%
+  #filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide2019 <- ts_wide(ts.wide2019)
+
+# impute with means of each TS
+# TODO check sensitivity
+ts.wide2019 <- na_mean(ts.wide2019)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# extract names
+ts.wide2019.names <- names(ts.wide2019[,-1])
+ts.wide2019 <- ts.wide2019[,-1]
+
+# standardise
+ts.wide2019 <- zscore(ts.wide2019)
+
+# convert to list for tsclust
+ts.wide2019 <- split(ts.wide2019, rep(1:ncol(ts.wide2019), each = nrow(ts.wide2019)))
+
+# k = 9
+pc_k <- tsclust(ts.wide, k = 9, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+plot(pc_k)
+
+cluster_up.2019 <- data.frame(LAD=ts.wide2019.names,
+                           cluster=pc_k@cluster)
+```
+
+## Export clusters
 
@@
 
+# 2019 upload cluster
+out.2019.clusters <- paste(path, "/data/temp/clusters_up2019.csv", sep = "")
+write_csv(cluster_up.2019, out.2019.clusters)
+```
+
+## Mapping the clusters: upload 2020
+
+```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
 # get LA
+
 # this is the heaviest version:
@@
 # out.path.map <- paste0(path, "/paper/v2_taylor_francis/figures/map.up.clusters.png")
+# ggsave(out.path.map)
+#width = 210, #A4
+#height = 297, #A4
+#units = "mm")
+```
+
+## Mapping the clusters: upload 2019
+
+```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+# get LA
+
+# ggplot2 maps
+la.f <- broom::tidy(la, region = "lad19nm")
+la.f <- merge(la.f, cluster_up.2019, by.x = "id", by.y = "LAD")
+la.f <- la.f[order(la.f$order),] # if i don't order merge.nuts.f loses order and the map has gaps
+
+# using world.cities from `maps`
+cities <- world.cities %>% filter(country.etc=="UK")
+
+ggplot(la.f, aes(x = long, y = lat)) +
+  geom_polygon(aes( group = group, fill = as.factor(cluster))) +
+  #labs(title = "Upload clusters") +
+  scale_fill_brewer(palette="Set1", 
+                    direction = -1) +
+  geom_text_repel( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=4) +
+  geom_point( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat), color="black", size=1) +
+  guides(fill=guide_legend(title="Clusters")) +
+  theme_void()
+
+# Run only if needed
+# out.path.map <- paste0(path, "/paper/v2_taylor_francis/figures/map.up.2019.clusters.png")
 # ggsave(out.path.map)

### 2021-03-26 paper/v2_taylor_francis/response_reviewers.md

--- 2021-03-24 paper/v2_taylor_francis/response_reviewers.md
@@
 Thank you for providing the opportunity to revise and resubmit our paper.
-We found the reviewer comments very useful and constructive and we believe they enabled us to improve the paper.
-We were also very pleased to see that the reviewers found our paper ...
-
-blah blah
+We found the reviewer comments very useful and constructive and we believe they 
+enabled us to improve the paper and clarify its findings.
+We were also very pleased to see that the reviewers found our paper a timely and
+important topic of research, and that we have addressed all their questions and 
+concerns below and in the paper itself.
 
@@
 and a difference of 1-2Mbps slower may well be noticeable at the wrong moment.
-Importantly, we only consider the test which took place during working hours, so
-when speed drops during, let's say, a work Zoom call or when uploading large files,
+Importantly, we only consider the tests which took place during working hours, so
+when speed drops, let's say, during a work Zoom call or when uploading large files,
 it will be noticeable. This is why we created the clusters by how the time series
 of speeds changes by day of the week and hour of the day, rather than focusing our
-analysis on mean speeds.
+analysis on mean speeds for the period as a whole.
 
@@
 "High levels of demand are one of the main causes of reduced reliability and slower connection speeds, as network bandwidth becomes congested".
+
 We also have added a variable to the auxiliary regression tracking the share of
@@
 could work and who could not at the height of the first wave of the pandemic.
+
 Finally, as we note at the end of the methods and data section, an auxiliary 
@@
 speed checks concluded that average speeds are lower in rural areas, something that
-has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
+has not been improved by policy measures to increase competition 
+[@riddlesden2014broadband; @nardotto2015unbundling]. In contrast,
 @riddlesden2014broadband found that levels of deprivation did not correlate with
@@
 
-Regarding multicolinearity, the highest correlations were observed between $NVQ4+$
+Regarding multicolinearity, the highest correlations were observed between $NVQ4+$, 
+which is a measure of being educated to degree level or higher,
 and $professional jobs$ ($0.78$) and $NVQ4+$ and earnings ($0.69$). Hence, we decided
@@
 
-ET: after we finalise k, we can rerun k-means for 2019 and decide - could also show mean speeds on a map for 2019 - this would be part of the response to reviewer 1 q2 and q6.
-**Let's discuss this**
+This is useful. We have added a new Appendix 2, which plots the clusters for 
+the same period in 2019. The picture is very different, which is an indication of 
+the changes in the temporal profiles of internet usage during the pandemic, as 
+we say towards the end of section 4.1:
+
+"It is worth noting here that when we apply the same methods to upload speed data for 
+the same time period in 2019, the spatial distribution and composition of the nine 
+clusters of LADs are very different -- see Appendix [2](#appendix2). This 
+difference is an indication of the changes in the temporal profile of internet usage 
+that took place during the pandemic, and which would be independent of any wider
+trend in improving internet services."
 
@@
 
-We have tightened up the results section, and begun this discussion in the last
-paragraph of the results section and then expanded the conclusion section to include
-a more in-depth discussion of the results and how they link to some of the issues you 
-suggest at the start. We mention some uncertainties or avenues for future research,
+We have tightened up the results section, which had to be re-written for the nine
+clusters instead of 13, and begin our discussion of insights in the last paragraph 
+of the results section. We then expand the conclusion section to include
+a more in-depth discussion of how the results link to some of the issues you 
+suggest.
+
+We mention some uncertainties or avenues for future research in this final section,
 but also see our response to Reviewer 1, point 1 for an explanation of biases.
@@
 We clarify that the occupations included in the regression are more about the level-2
-digital divide, whereas the average earnings and, particularly during the pandemic
-the % furloughed, are indications of level-3 digital divides. We did have education
+digital divide, providing an indication of not only education, but digital
+skills. The average earnings and, particularly during the pandemic, the new variable 
+-- % furloughed, are indications of level-3 digital divides. We did have education
 in the previous submission (NVQ4+), but have removed it due to multicollinearity
--- see response for point 2, which give more indication of not only education, but digital
-skills.
+-- see response for point 2.
 

### 2021-03-26 paper/v2_taylor_francis/broadband.speed.covid.Rmd

@@
 has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
-In contrast, @riddlesden2014broadband found that levels of deprivation did not correlate with 
-first level digital divides.
+In contrast, @riddlesden2014broadband found that levels of deprivation did not 
+correlate with first level digital divides.
 
@@
 
-Of the smaller clusters, cluster $7$ experiences quite similar mean speeds to 
+Of the smaller clusters, the $15$ LADs in clusters $1$ and $8$, home to over $1.4$ 
+million people, not only experience fairly average upload speeds and high ratios of 
+standard deviation to the mean, but also experience much lower speeds during 
+the morning peak than the evening peak. From the large spikes and dips shown on
+Figure \ref{UpCluster}, it is likely that this poor reliability or consistency of 
+morning internet speeds was more noticeable to those in cluster $1$, whilst those 
+in cluster $8$ experienced the lack of reliable service as a problem throughout 
+the day. Meanwhile, cluster $7$ experiences quite similar mean speeds to 
 cluster $9$, and a not too dissimilar standard deviation, but mean speeds in the 
@@
 may be closer to what might have been expected pre-pandemic.
-In comparison, the $15$ LADs in clusters $1$ and $8$, home to over $1.4$ million 
-people, not only experience fairly average upload speeds and high ratios of 
-standard deviation to the mean, but also experience much lower speeds during 
-the morning peak than the evening peak. From the large spikes and dips shown on
-Figure \ref{UpCluster}, it is likely that this poor reliability or consistency of 
-morning internet speeds was more noticeable to those in cluster $1$, whilst those 
-in cluster $8$ experienced the lack of reliable service as a problem throughout 
-the day.
-
-It is worth noting here that clusters of LAD based on the upload speeds are very 
-different to the ones occur when we apply the same methods to upload speed data
-for the same time period in 2019 -- see Appendix [2](#appendix2). This drastic 
-difference between the LAD upload speed profiles before and during the pandemic is
-an indication of the changes in internet usage that took place during the pandemic.
+
+It is worth noting here that when we apply the same methods to upload speed data for 
+the same time period in 2019, the spatial distribution and composition of the nine 
+clusters of LADs are very different -- see Appendix [2](#appendix2). This 
+difference is an indication of the changes in the temporal profile of internet usage 
+that took place during the pandemic, and which would be independent of any wider
+trend in improving internet services.
 

### 2021-03-26 paper/v2_taylor_francis/response_reviewers.md
second submission

@@
 bibliography: bibliography.bib
+suppress-bibliography: true
 ---
-
-Dear Prof. Bian,
-
-Thank you for providing the opportunity to revise and resubmit our paper.
-We found the reviewer comments very useful and constructive and we believe they 
-enabled us to improve the paper and clarify its findings.
-We were also very pleased to see that the reviewers found our paper a timely and
-important topic of research, and that we have addressed all their questions and 
-concerns below and in the paper itself.
-
-Kind Regards,
-
-[removed for anonymity]
 
@@
 
-Firstly, mean speeds are not likely to be that different between regions of the UK
+We have no reasons to expect that mean speeds are that different between regions of the UK
 (e.g. North or South England). The greater differences are between urban, built-up
@@
 with the addition of the following sentence:
-"High levels of demand are one of the main causes of reduced reliability and slower connection speeds, as network bandwidth becomes congested".
+"High levels of demand are one of the main causes of reduced reliability and slower 
+connection speeds, as network bandwidth becomes congested".
 
@@
 the population that was put on furlough. This was the scheme in the UK which, at
-least initially, prevented much change in the employment rate, but did show who
+least until now, prevented much change in the employment rate, but did show who
 could work and who could not at the height of the first wave of the pandemic.
@@
 
-The latter bit about Virgin Media is moved from the results section, in response to reviewer 2, q7.
+The latter bit about Virgin Media is moved from the results section, in response to Reviewer 2, q7.
 
@@
 the subsequent multinomial regression using $k = 9$. We indeed believe that the
-results are more clear now. we have added the following justification in the Methods
+results are clearer now. We have added the following justification in the Methods
 and Data section:
@@
 
-x-axis for Figures 1 and 2 have been amended accordingly.
+- x-axis for Figures 1 and 2 have been amended accordingly.
 
-The temporal profiles of the upload clusters are now included in Figure 3.
+- The temporal profiles of the upload clusters are now included in Figure 3.
 
-Figure 5 now labelled Figure 4: the map has now been redesigned using a different
+- Figure 5, which is now labeled Figure 4: the map has now been redesigned using a different
 colour palette to better reflect the discrete nature of the underpinning data --
@@
 
-We have toned down the statement on returns to say:
+- We have toned down the statement on returns to say:
 "In some aspects, these returns may have increased during the Covid-19 crisis..."
 
-Apologies on LAD acronym, now corrected.
+- Apologies on LAD acronym, now corrected.
 
-We also moved and integrated introduction of Virgin Media to the literature review s2.2. See also response to reviewer 1, q6.
-
-## References
+- We also moved and integrated introduction of Virgin Media to the literature review s2.2. See also response to Reviewer 1, q6.

### 2021-03-26 paper/v2_taylor_francis/broadband.speed.covid.tex
second submission

@@
 
-\author{\name{Hannah Budnitz$^{a}$, Emmanouil Tranos$^{b, c}$}
-\affil{$^{a}$Transport Studies Unit, University of Oxford, UK; $^{b}$School of Geographic Sciences, University of Bristol, UK; $^{c}$The Alan Turing Institute, UK}
+\author{\name{Hannah Budnitz$^{a}$, Emmanouil Tranos$^{b, c, \dagger}$}
+\affil{$^{a}$Transport Studies Unit, University of Oxford, UK; $^{b}$School of Geographic Sciences, University of Bristol, UK; $^{c}$The Alan Turing Institute, UK; $^{\dagger}$Corresponding author}
 }
 
-\thanks{CONTACT Hannah Budnitz. Email: , Emmanouil Tranos. Email: \href{mailto:e.tranos@bristol.ac.uk}{\nolinkurl{e.tranos@bristol.ac.uk}}}
+\thanks{CONTACT Hannah Budnitz.
+Email: \href{mailto:hannah.budnitz@ouce.ox.ac.uk}{\nolinkurl{hannah.budnitz@ouce.ox.ac.uk}}, Emmanouil Tranos.
+Email: \href{mailto:e.tranos@bristol.ac.uk}{\nolinkurl{e.tranos@bristol.ac.uk}}}
 
@@
 
-Of the smaller clusters, cluster \(7\) experiences quite similar mean
-speeds to cluster \(9\), and a not too dissimilar standard deviation,
-but mean speeds in the morning peak are almost \(10\)\% lower than the
-evening peak. Cluster \(3\) boasts speeds almost \(1\)Mb/s higher again,
-but suffers from even greater slowdown in the morning peak. In
-comparison, clusters \(2\) and \(5\), home to a little less than half a
-million people, experience not only above average mean speeds, but also
-faster speeds in the morning peak compared to the evening. Thus, in the
-five LADs in these two clusters, the temporal profile of internet use
-may be closer to what might have been expected pre-pandemic. In
-comparison, the \(15\) LADs in clusters \(1\) and \(8\), home to over
-\(1.4\) million people, not only experience fairly average upload speeds
-and high ratios of standard deviation to the mean, but also experience
-much lower speeds during the morning peak than the evening peak. From
-the large spikes and dips shown on Figure \ref{UpCluster}, it is likely
-that this poor reliability or consistency of morning internet speeds was
-more noticeable to those in cluster \(1\), whilst those in cluster \(8\)
-experienced the lack of reliable service as a problem throughout the
-day.
-
-It is worth noting here that clusters of LAD based on the upload speeds
-are very different to the ones occur when we apply the same methods to
-upload speed data for the same time period in 2019 -- see Appendix
-\protect\hyperlink{appendix2}{2}. This drastic difference between the
-LAD upload speed profiles before and during the pandemic is an
-indication of the changes in internet usage that took place during the
-pandemic.
+Of the smaller clusters, the \(15\) LADs in clusters \(1\) and \(8\),
+home to over \(1.4\) million people, not only experience fairly average
+upload speeds and high ratios of standard deviation to the mean, but
+also experience much lower speeds during the morning peak than the
+evening peak. From the large spikes and dips shown on Figure
+\ref{UpCluster}, it is likely that this poor reliability or consistency
+of morning internet speeds was more noticeable to those in cluster
+\(1\), whilst those in cluster \(8\) experienced the lack of reliable
+service as a problem throughout the day. Meanwhile, cluster \(7\)
+experiences quite similar mean speeds to cluster \(9\), and a not too
+dissimilar standard deviation, but mean speeds in the morning peak are
+almost \(10\)\% lower than the evening peak. Cluster \(3\) boasts speeds
+almost \(1\)Mb/s higher again, but suffers from even greater slowdown in
+the morning peak. In comparison, clusters \(2\) and \(5\), home to a
+little less than half a million people, experience not only above
+average mean speeds, but also faster speeds in the morning peak compared
+to the evening. Thus, in the five LADs in these two clusters, the
+temporal profile of internet use may be closer to what might have been
+expected pre-pandemic.
+
+It is worth noting here that when we apply the same methods to upload
+speed data for the same time period in 2019, the spatial distribution
+and composition of the nine clusters of LADs are very different -- see
+Appendix \protect\hyperlink{appendix2}{2}. This difference is an
+indication of the changes in the temporal profile of internet usage that
+took place during the pandemic, and which would be independent of any
+wider trend in improving internet services.
 

### 2021-03-26 paper/v2_taylor_francis/broadband.speed.covid_anonymised.Rmd
second submission

--- 2021-03-26 paper/v2_taylor_francis/broadband.speed.covid.Rmd
@@
 author:
-  - name: Hannah Budnitz
+  - name: 
     affil: a
-  - name: Emmanouil Tranos
-    affil: b, c
-    email: e.tranos@bristol.ac.uk
 affiliation:
   - num: a
-    address: |
-      Transport Studies Unit, University of Oxford, UK
-  - num: b
-    address: |
-      School of Geographic Sciences, University of Bristol, UK
-  - num: c
-    address: |
-      The Alan Turing Institute, UK
+    address: 
 bibliography: bibliography.bib
@@
 
-\pagebreak
+\newpage
 

### 2021-03-26 paper/v2_taylor_francis/broadband.speed.covid_anonymised.tex
v.2
second submission

--- 2021-03-04 paper/v2_taylor_francis/broadband.speed.covid.tex
@@
 
-\author{\name{Hannah Budnitz$^{a}$, Emmanouil Tranos$^{b, c}$}
-\affil{$^{a}$Transport Studies Unit, University of Oxford, UK; $^{b}$School of Geographic Sciences, University of Bristol, UK; $^{c}$The Alan Turing Institute, UK}
+\author{\name{$^{a}$}
+\affil{$^{a}$}
 }
 
-\thanks{CONTACT Hannah Budnitz. Email: , Emmanouil Tranos. Email: \href{mailto:e.tranos@bristol.ac.uk}{\nolinkurl{e.tranos@bristol.ac.uk}}}
+\thanks{CONTACT . Email: }
 
@@
 characteristics to explore how they overlap with or diverge from the
-existing economic and digital geography of the UK. Our analysis enables
-us to better understand how the spatial and social distribution of
-education, occupation and online accessibility intersect to enable or
-hinder the practice of telecommuting at a time of extreme demand. We
-also consider what lessons can be learned from this time for a future
-where telecommuting is likely to remain a more common means of accessing
-work, at least in comparison to the pre-Covid era.
-
-The capability to work from home has previously been studied from the
-perspective of whether work tasks in a given occupation both can be and
-are allowed to be performed using digital technologies independently of
-location or co-location with colleagues, including supervisors
-\citep{allen2015effective, singh2013modeling}. However, successful
-telecommuting also requires that the quality and reliability of digital
-services, particularly home internet connection speeds, enable the
-completion of work tasks with a minimum of delay or interruption. Prior
-to the pandemic, the performance of broadband services with respect to
-telecommuters was never tested at scale, as working from home and
-connecting to colleagues and workplace resources via the internet was
-the purview of a small minority of workers. Instead, leisure use in the
-evening, when video streaming services are at their peak, has been used
-to benchmark broadband performance and service delivery by different
-Internet Service Providers (ISPs) \citep{ofcom2017}.
-
-Yet the shift towards telecommuting during various stages of lockdown
-around the world has been drastic and there are speculations that
-post-Covid, the tendency to work from home will be much higher, raising
-questions around whether internet services can accommodate the increased
-demand. For example, \(47\)\% of people in employment in the UK worked
-solely from home in April \(2020\), whilst the same figure only reached
-\(5\)\% the year before \citep{ons2020, ons2020lm2019}. A back of the
-envelope calculation suggests that up to 40\% of the labour force could
-work from home on an ongoing basis \citep{batty2020editorial}. Similar
-figures have been reported for other countries
-\citep{felstead2020homeworking}. Approximately \(37\)\% of the European
-workforce worked from home in April \(2020\) with countries like Finland
-reaching \(60\)\% \citep{eurofound2020}. In the US, almost half of the
-working population worked from home during the same period because of
-the pandemic \citep{brynjolfsson2020covid}, and a recent estimate
-indicated that \(37\)\% of all jobs in the US can be permanently
-performed entirely from home \citep{NBERw26948}. None of these changes
-could have happened in the absence of reliable information and
-communication technology (ICT) infrastructure -- both in terms of
-software and hardware. But while software innovations are easily
-diffused across space and society\footnote{See for example the huge
-  success of videoconferencing apps such as Zoom \citep{marks2020zoom}.},
-the same does not apply for ICT hardware infrastructure such as internet
-broadband connectivity.
-
-The literature describes first level digital divides in terms of the
-availability and quality of internet connectivity, such as that manifest
-in different geographies in the UK
-\citep{riddlesden2014broadband, philip2017digital}. Second level digital
-divides consider the presence or lack of the necessary skills to
-effectively utilise digital technologies and the internet
-\citep{blank2014dimensions, van2011internet}. The third level focuses on
-the heterogenous returns of internet usage among different socioeconomic
-groups and, consequently, how digital technologies can assist in
-bridging or further enhancing existing socioeconomic divides.
-\citep{stern2009levels, van2014digital, van2015third}. The capability to
-telecommute is related to all three levels of digital divides, but more
-importantly leads to differentiated outcomes regarding the economic
-resilience of people and places to overcome a systemic shock such as the
-current pandemic.
-
-As the quality of internet infrastructure and services, as well as the
-concentration of different occupations are spatially dependent and
-clustered in space, our approach offers a framework for understanding
-the impact of and interactions between digital divisions geographically
-and socioeconomically. By asking how resilient broadband speeds, and
-particularly upload speeds, are as experienced in different parts of the
-UK during a time of extreme demand, we interrogate which places benefit
-from the greater economic resilience digital technologies can offer. The
-structure of this paper is as follows. First we review the literature on
+existing economic and digital geography of the UK. As the quality of
+internet infrastructure and services, as well as the concentration of
+different occupations are spatially dependent and clustered in space,
+our approach offers a framework for understanding the impact of and
+interactions between digital divisions geographically and
+socioeconomically. Our analysis enables us to better understand how the
+spatial and social distribution of occupation and online accessibility
+intersect to enable or hinder the practice of telecommuting at a time of
+extreme demand. We also consider what lessons can be learned from this
+time for a future where telecommuting is likely to remain a more common
+means of accessing work, at least in comparison to the pre-Covid era.
+
+By asking how resilient broadband speeds, and particularly upload
+speeds, are as experienced in different parts of the UK during a time of
+extreme demand, we interrogate which places benefit from the greater
+economic resilience digital technologies can offer. The structure of
+this paper is as follows. First we review the literature on
 telecommuting and digital divides to better understand their structural
@@
 \citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
-These returns have been even greater during the Covid-19 crisis, when
-those with the capability to telecommute also have the ability to
-maintain their employment whilst protecting their health. However, the
-success of these arrangements has been dependent upon the first level
-digital divide, which is associated with access to and quality of
-internet connectivity. \citet{SALEMINK2017360} provides a systematic
-review of the pre-pandemic, first level digital divide in infrastructure
-quality between urban and rural areas in various advanced economies.
-Rural areas, predictably, fare worse, yet as \citet{blank2018local}
-highlight, variation in individual internet uptake and use is a product
-of more complex spatial and demographic characteristics than simple
-rurality or urbanisation. Yet whether this variation in infrastructure
-quality affects the spatial footprint of telecommuting has not
-previously been investigated, in part because telecommuting has not
-previously been a cause of concentrated demand and pressure on internet
-services.
-
+In some aspects, these returns may have increased during the Covid-19
+crisis, when those with the capability to telecommute also have the
+ability to maintain their employment whilst protecting their health.
+However, the success of these arrangements has been dependent upon the
+first level digital divide, which is associated with access to and
+quality of internet connectivity. \citet{SALEMINK2017360} provides a
+systematic review of the pre-pandemic, first level digital divide in
+infrastructure quality between urban and rural areas in various advanced
+economies. Rural areas, predictably, fare worse. This finding broadly
+applies in the UK, where studies that also analysed broadband speed
+checks concluded that average speeds are lower in rural areas, something
+that has not been improved by policy measures to increase competition
+\citep{riddlesden2014broadband, nardotto2015unbundling}. In contrast,
+\citet{riddlesden2014broadband} found that levels of deprivation did not
+correlate with first level digital divides.
+
+However, as \citet{blank2018local} highlight, variation in individual
+internet availability and uptake is a product of more complex spatial
+and demographic characteristics than simple rurality or urbanisation.
+Dense urban areas were shown to suffer more from slowdown during peak
+hours, although these services were more likely to be improved by
+increased competition between providers, such as between new entrants
+and Virgin Media cable connections
+\citep{riddlesden2014broadband, nardotto2015unbundling}. The latter were
+historically available to only \(45\)\% of premises in the UK
+\citep{ofcom2016}, where the more lucrative and competitive market
+originally attracted the cable TV provider. Whether the variation in
+infrastructure quality and reliability affects the spatial footprint of
+telecommuting has not previously been investigated, in part because
+telecommuting has not previously been a cause of concentrated demand and
+pressure on internet services.
+
+Other intersections between digital divides have been subject to study.
 There are indications that those who purchase high speed connections
@@
 correlation between access to internet services and a reduction in
-household transport spend \citep{bris2017ict}. Independently of whether
-the additional internet use and reduced travel is because of increased
-telecommuting, these studies suggest that better internet services
-enable households to make savings and efficiencies, an example of the
-first level digital divide reinforcing the third level.
-
-Multi-layered digital divides may also intersect with material divides
-and the economic geography of the UK. The regional economic resilience
-literature underlines the differentiated capacity of cities and regions
-to escape or recover from economic crises
+household transport spend \citep{bris2017ict}. These studies suggest
+that better internet services enable households to make savings and
+efficiencies, an example of the first level digital divide reinforcing
+the third level. Such digital intersections and efficiencies have also
+changed transportation, connectivity and the structure of cities during
+the pandemic. We now have enough data to confirm the drastic alteration
+observed in the space-time geography of cities around the world in the
+last year \citep[e.g.~][]{google2020, shibayama2021impact}. Yet,
+although there is a broad agreement that these changes during the
+pandemic have played a pivotal role in stopping the spread of the virus
+\citep{jia2020population, yang2020taking, mu2020interplay}, the extent
+to which the increased levels of working from home, the consequent
+decrease in commuting flows, and the altered structure of cities will
+remain post-pandemic is the subject of considerable debate in the
+literature.
+
+Meanwhile, multi-layered digital divides may also intersect with
+material divides and the economic geography of the UK. The regional
+economic resilience literature underlines the differentiated capacity of
+cities and regions to escape or recover from economic crises
 \citep{martin2012regional, kitsos2018economic}. As different places have
@@
 not only the day of the week, but also time of day, as can be seen in
-Figure \ref{test2020}. Thus, a daily aggregation of upload speeds would
-mask the variation in experienced service over the course of each
-weekday. Furthermore, the importance of this variation is highlighted by
-a comparison with the same period in \(2019\), as in Figure
-\ref{test2019}, when the volume of testing and thus of experienced
-internet service quality peaked in the evening, presumably in response
-to demand for leisure activities and download speeds. In contrast, the
-majority of the increase in testing in \(2020\) is during the working
-day, creating a new morning peak in Figure \ref{test2020}. Therefore, we
-include a measure of hourly variation in our temporal profiles to
-reflect the change in users' perception of the workday reliability of
-internet services.
+Figure \ref{test2020} \footnote{Figures \ref{test2020} and
+  \ref{test2019} were created with the \texttt{R} package
+  \texttt{openair} \citep{openair}.}. Thus, a daily aggregation of
+upload speeds would mask the variation in experienced service over the
+course of each weekday. Furthermore, the importance of this variation is
+highlighted by a comparison with the same period in \(2019\), as in
+Figure \ref{test2019}, when the volume of testing and thus of
+experienced internet service quality peaked in the evening, presumably
+in response to demand for leisure activities and download speeds. In
+contrast, the majority of the increase in testing in \(2020\) is during
+the working day, creating a new morning peak in Figure \ref{test2020}.
+Therefore, we include a measure of hourly variation in our temporal
+profiles to reflect the change in users' perception of the workday
+reliability of internet services.
 
 However, there were insufficient observations -- only \(631\) speed
-tests per LAD on average -- for each for each working hour of each
-working day in each Local Authority District (LAD) to profile speeds at
-that level of detail. Spatial aggregation was also necessary because we
-could not follow individuals or households and connect data points.
-Therefore, we aggregate the \(241,088\) individual, geolocated and
-time-stamped speed-checks during the \(13\) weeks of March to May
-inclusive for weekdays in \(2020\) by each hour of the day and day of
-the week. As our research aims to identify the geography of internet
-service resilience for work purposes, bank holidays and the hours
-between midnight and \(6:00\) were excluded, as well as weekend days.
-The composite week time-series thus comprise \(18\) hours multiplied by
-\(5\) weekdays or \(90\) time points per series. The time-series were
-calculated for each of the \(382\) LADs in the UK, standardised, and
-then a \emph{k}-means partitioning around medoids clustering algorithm
-was applied using DTW. We initially run the algorithm for
-\(k \in \mathbb{N} \bigcap [5,15]\) and used cluster validity indices
-(CVIs) to pick the optimal solution of \(k = 13\). Following
-\citet{sardatime} the majority vote for the following CVIs was used:
-Silhouette (max), Score function (max), Calinski-Harabasz (max),
-Davies-Bouldin (min), Modified Davies-Bouldin (DB*, min), Dunn (max),
-COP (min).
+tests per Local Authority District (LAD) on average -- for each for each
+working hour of each working day in each LAD to profile speeds at that
+level of detail. Spatial aggregation was also necessary because we could
+not follow individuals or households and connect data points. Therefore,
+we aggregate the \(241,088\) individual, geolocated and time-stamped
+speed-checks during the \(13\) weeks of March to May inclusive for
+weekdays in \(2020\) by each hour of the day and day of the week. As our
+research aims to identify the geography of internet service resilience
+for work purposes, bank holidays and the hours between midnight and
+\(6:00\) were excluded, as well as weekend days. The composite week
+time-series thus comprise \(18\) hours multiplied by \(5\) weekdays or
+\(90\) time points per series. The time-series were calculated for each
+of the \(382\) LADs in the UK, standardised, and then a \emph{k}-means
+partitioning around medoids clustering algorithm was applied using DTW.
+We initially run the algorithm for \(k \in \mathbb{N} \bigcap [5,15]\),
+calculated the cluster validity indices (CVIs) and then run the
+subsequent multinomial regression -- see the end of this section for
+more details for the different \emph{k}. Following \citet{sardatime}, to
+identify the optimal \emph{k} we used the majority vote for the
+following CVIs: Silhouette (max), Score function (max),
+Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin
+(DB*, min), Dunn (max), COP (min). Nevertheless, we opted against using
+the optimal \(k = 13\) solution as it was too large to allow for
+communicable LAD clusters. Instead we opted for a smaller \(k = 9\),
+which led to a rather similar spatial pattern and, importantly, to a
+much higher R-squared in the subsequent explanatory regression (\(0.44\)
+instead of \(0.34\)).
 
@@
 
-The second largest cluster, comprising \(126\) local authorities and
-over \(20\) million people, is cluster \(13\), which has the slowest
-aggregate mean upload speed, and the second highest ratio of the
-standard deviation to the mean. This suggests that those living in local
+The largest cluster, comprising \(229\) local authorities and over
+\(40\) million people, is cluster \(6\), which has the slowest aggregate
+mean upload speed of any cluster, and the highest ratio of the standard
+deviation to the mean. This suggests that those living in local
 authorities in this cluster experienced some of the lowest quality
-broadband services in terms of upload speeds in the UK. However, Figure
-\ref{UpClusterL} shows that the high standard deviation, which is one
-indication of unreliable internet, did not disproportionately affect the
-morning peak from \(9:00\)-\(10:59\), when upload speeds were, on
-average, only \(2.6\)\% slower than in the evening peak period between
-\(19:00\) and \(20:59\), when entertainment purposes are likely to be
-using the most bandwidth. In comparison, the five LADs in cluster \(1\),
-home to \(343\) thousand people, not only experience the second slowest
-mean upload speeds and the highest ratio of standard deviation to the
-mean, but are also much more affected during the morning peak.
-
-Meanwhile, those living in the largest cluster -- \(6\), with \(155\)
-LADs and \(29.5\) million people -- experienced aggregate mean upload
-speeds of about \(1\)Mb/s faster than those in cluster \(13\), but still
-lower than the other three large clusters and most of the smaller
-clusters, suggesting a middling quality of service. The temporal profile
-for cluster \(6\) in Figure \ref{UpClusterL} shows that upload speeds
-are highest at \(6:00\) on a Monday morning and between \(23:00\) and
-midnight on Wednesday and Thursday, but tend to be lower during the
-working day. Furthermore, experienced mean upload speeds in the morning
-peak are \(4.4\)\% lower than in the evening peak -- a greater, more
-noticeable change than any of the other large clusters experience,
-suggesting poorer reliability during the working day. This difference is
-less, however than any of the clusters included in Figure
-\ref{UpClusterS}.
-
-Clusters \(8\) and \(12\) also have mean upload speeds under \(10\)Mb/s,
-but higher than clusters \(1\), \(13\), and \(6\). However, mean upload
-speeds are much lower between \(9:00-10:59\) than between
-\(19:00-20:59\) in cluster \(8\), but slightly faster in the morning
-than in the evening in cluster \(12\). Indeed, cluster \(12\) is the
-only cluster to experience higher speeds in the morning peak compared to
-the evening, and thus the only cluster where the temporal profile of
-internet use is closer to what might have been expected pre-pandemic.
-Among the other clusters, however, the reliability of internet services
-during the working day still varies considerably. Interpreting this
-variation from the large spikes and dips shown on Figures
-\ref{UpClusterL} and \ref{UpClusterS} is difficult, but the statistics
-in Table \ref{up.cluster.descr} show that clusters \(9\) and \(11\) have
-the most reliable internet services. The ratio of standard deviation to
-mean in both these clusters is lowest, and speeds are only about 2\%
-slower in the morning than the evening. Mean speeds are also higher than
-in any other cluster, excluding cluster \(2\), where measures of
-reliability suggest poorer performance.
-
-Thus, broadband services in clusters \(9\) and \(11\), home to over
-twelve million people, performed the best during the study period, in
-terms of both quality and reliability. In Figure \ref{UpClusterL},
-cluster \(11\) shows more noticeable peaks and troughs, with the lowest
-points occurring, on average, between \(6:00-7:00\) on Monday morning,
-\(14:00-15:00\) on Friday, and \(16:00-17:00\) on Thursday, not the
-workday peak. Furthermore, these slow periods offer better speeds than
-the average hourly profile of cluster \(13\). Clusters \(3\), \(7\) and
-\(10\) also have relatively high mean upload speeds. Clusters \(3\) and
-\(10\) pair high mean speeds with low standard deviations relative to
-the mean speeds, suggesting reliability and resilience, as well as
-quality broadband services. Cluster \(7\) has a higher ratio of standard
-deviation to mean, but there is less difference in average speeds
-between the morning and evening peaks than in clusters \(3\) and \(10\).
-
-In summary, LADs in clusters \(9\) and \(11\) experienced resilient
-broadband internet that could support high levels of telecommuting.
-Those in clusters \(2\), \(3\), \(7\), and \(10\) also experience higher
-than average mean speeds and rank middle to high on measures of service
-reliability. These LADs are \emph{not} on the wrong side of the first
-level digital divide, but how likely are they to be able to take
-advantage of their resilient ICT infrastructure and services? Meanwhile,
-cluster \(6\) is not only the largest in terms of number of LADs and
-population, it has the closest mean upload speed to the pre-clustered
-average for the whole sample. As well as average quality internet
-services, those in cluster \(6\) also experience average reliability for
-work purposes, ranking fifth behind the four other clusters with
-populations over one million, but ahead of the smaller clusters.
-Clusters \(8\) and \(12\) are also close to average mean upload speeds,
-but show very different patterns in terms of reliability, whilst
-clusters \(1\) and \(13\) appear to suffer most from a lack of quality
-internet services, with slow speeds and high standard deviations. With
-those in cluster \(1\) more likely to experience that poor reliability
-during the morning peak, is this first level digital divide occurring in
-areas where few are occupationally able to telecommute anyway, and what
-are the implications for economic resilience?
+broadband services in terms of upload speeds and reliability in the UK.
+However, as shown in Figure \ref{map.up.clusters}, some of the most
+rural areas of the UK are included in this cluster. If these areas
+suffer most from first level digital divides as described in the
+literature review, the low speeds in rural areas might be pulling down
+the averages in other areas in this largest cluster. Also, since the
+areas are clustered by their temporal profile across the working week,
+the graph in Figure \ref{UpCluster} indicates that speeds in cluster
+\(6\) are some of the more reliable. Table \ref{up.cluster.descr}
+confirms that upload speeds in cluster \(6\) during the morning peak
+from \(9:00\)-\(10:59\) were, on average, only \(4.5\)\% slower than in
+the evening peak period between \(19:00\) and \(20:59\), when
+entertainment purposes are likely to be using the most bandwidth.
+
+\begin{figure}
+\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-4}
+\end{figure}
+
+Those living in the second largest cluster -- \(9\), with \(115\) LADs
+and \(21.5\) million people -- experienced aggregate mean upload speeds
+of more than \(1.5\)Mb/s faster than those in cluster \(6\), a lower
+ratio of standard deviation to mean, and almost the same upload speeds
+in the AM Peak as the PM Peak. Indeed, the temporal profile for cluster
+\(9\) in Figure \ref{UpCluster}, like that for cluster \(6\), is fairly
+flat. Whilst it is likely that the large numbers of tests being
+performed by the large populations in these two clusters of LADs result
+in less varied averages to show in the graph, the method of shape-based
+clustering suggests that these results indicate that the majority of the
+population of the UK experience less temporal variation, and so belong
+in these two clusters. The results also confirm that those in cluster
+\(9\) experienced a consistently better service during lockdown than
+those in cluster \(6\), in terms of both average speed and reliability.
+
+Of the smaller clusters, the \(15\) LADs in clusters \(1\) and \(8\),
+home to over \(1.4\) million people, not only experience fairly average
+upload speeds and high ratios of standard deviation to the mean, but
+also experience much lower speeds during the morning peak than the
+evening peak. From the large spikes and dips shown on Figure
+\ref{UpCluster}, it is likely that this poor reliability or consistency
+of morning internet speeds was more noticeable to those in cluster
+\(1\), whilst those in cluster \(8\) experienced the lack of reliable
+service as a problem throughout the day. Meanwhile, cluster \(7\)
+experiences quite similar mean speeds to cluster \(9\), and a not too
+dissimilar standard deviation, but mean speeds in the morning peak are
+almost \(10\)\% lower than the evening peak. Cluster \(3\) boasts speeds
+almost \(1\)Mb/s higher again, but suffers from even greater slowdown in
+the morning peak. In comparison, clusters \(2\) and \(5\), home to a
+little less than half a million people, experience not only above
+average mean speeds, but also faster speeds in the morning peak compared
+to the evening. Thus, in the five LADs in these two clusters, the
+temporal profile of internet use may be closer to what might have been
+expected pre-pandemic.
+
+It is worth noting here that when we apply the same methods to upload
+speed data for the same time period in 2019, the spatial distribution
+and composition of the nine clusters of LADs are very different -- see
+Appendix \protect\hyperlink{appendix2}{2}. This difference is an
+indication of the changes in the temporal profile of internet usage that
+took place during the pandemic, and which would be independent of any
+wider trend in improving internet services.
+
+In summary, the reliability of internet services during the working day
+appears to have altered for the vast majority of locations due to
+increased use from residents told to stay at home, and, if they were
+fortunate, work from home. Yet this experience has been different in
+different locations. In particular, LADs in cluster \(9\) experienced
+higher speeds and more resilient broadband internet than those in the
+largest cluster, \(6\). Those in clusters \(3\) and \(7\) also
+experience higher mean speeds and better service reliability, and can be
+more confident that they are on the right side of the first level
+digital divide, and that their more resilient ICT infrastructure and
+services can robustly support higher levels of telecommuting. On the
+other hand, whilst high levels of telecommuting may be the cause of
+morning speeds over 15\% lower than those in the evening, those in
+clusters \(1\) and \(8\) also experience lower mean speeds than clusters
+\(3\), \(7\), and \(9\), but not as low as cluster \(6\). Whilst the
+differences may not be large, the lower upload speeds in cluster \(6\)
+or the poor reliability in clusters \(1\) and \(8\) could still have had
+implications for economic resilience if combined with other types of
+digital divide. This potential will be explored in the next section.
 
@@
 inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
-Virgin Media internet connections. Regarding the former, we expect
-people in LADs with more unreliable connections to test their internet
-speeds more often, and the results validate our priors. Meanwhile, fast
-Virgin Media cable connections have historically only been available to
-\(45\)\% of premises in the UK \citep{ofcom2016}, where the more
-lucrative and competitive market originally attracted the cable TV
-provider. Those in clusters \(2\), \(3\), \(9\) and \(11\) benefit from
-a higher proportion of Virgin connections, which is an indication that
-people in these clusters are more likely to live in urban areas, with
-more choice of broadband services. In other words, they are more likely
-to be on the right side of the first level, infrastructure-based digital
-divide, as we expected from the analysis in Section
-\protect\hyperlink{sec:4.1}{4.1}.
-
-We employ distance from London and from the nearest metropolitan area
-(including London) as two variables depicting peripherality, urban
-structure and, potentially, first level digital divides. The broadband
-speed tests run in the authorities in cluster \(3\) are more likely to
-be taking place close to London than those run in any of the other
-clusters, and two of the four authorities in cluster \(3\) are the
-London commuter towns of Harlow and Luton. However, even though London
-was also included in the variable calculating distance from the centre
-of one of either the ten largest metropolitan areas in England, or
-Glasgow or Cardiff, tests run in cluster \(3\) are likely to be furthest
-away. Thus, it is important to consider the membership of each cluster
-as well as the regression results. Corby and Eastbourne, the other two
-authorities in cluster \(3\) are large, accessible towns, although not
-part of metropolitan areas.
-
-\begin{figure}
-\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-6}
-\end{figure}
-
-Meanwhile, LADs in cluster \(7\) are most likely to be near the centre
-of a large metropolitan area, even though the four local authorities of
-cluster \(7\) include no central urban boroughs and only one LAD that is
-part of a metropolitan area of governance - Tameside in Greater
-Manchester. This may explain why those in cluster \(7\) are the second
-least likely to be served by Virgin Media. It is also a demonstration of
-the complexity of both experienced broadband upload speeds as captured
-by time-series clustering, and the geography of first level digital
-divides as a product of quality as well as availability. In comparison,
-cluster \(1\), which our analysis suggests lacks broadband resilience,
-contains five, mainly rural authorities, but they are closer to a
-metropolitan area than authorities in cluster \(3\), although furthest
-from London, perhaps because they are scattered around the country --
-see Figure \ref{map.up.clusters}.
-
-Internet resilience is also more nuanced than our -- arguably crude --
-dummy variable depicting the North-South economic divide, which assigns
-\(1\) to LADs located in Greater London, Southeast, Southwest and East
-of England regions, and \(0\) to the rest. The authorities most likely
-to be in the South are those in cluster \(13\), which was identified as
-having the slowest mean upload speeds of any of the clusters, and thus a
-low level of service. However, cluster \(13\) includes some rural,
-remote areas of the country, such as Northwest Scotland, Cornwall and
-Powys in Wales as shown on Figure \ref{map.up.clusters}. It also
-includes major metropolitan centres in the North -- Liverpool, Newcastle
--- and South -- Bristol, and nine (of \(32\)) London Boroughs. There are
-also plenty of Southern home county and suburban areas. The standard
-deviation measure for cluster \(13\) is high, but speed variation is low
-during the morning peak, suggesting that the estimates for reliability
-are inconsistent. Considering that this is one of the largest clusters,
-and thus the averages incorporate more noise than some of the smaller
-clusters, it may be that the LADs in this cluster do not all suffer
-equally from a first level digital divide.
-
-Yet we need to consider the results for other variables in Table
-\ref{aux} to better determine whether clusters \(1\) and \(13\), which
-appear to suffer most from poor quality internet services, are also more
-likely to have a low skilled workforce, less able to benefit from
-telecommuting. Cluster \(1\) has the lowest proportion of educated
-people, and the lowest earnings, despite recording the highest
-proportion of managerial and professional jobs, and the second highest
-proportion of tech jobs. Rural areas such as those in cluster \(1\) are
-home to many older, retired people \citep{blank2018local}, which might
-explain these results or perhaps, as these figures are relative, we
-could note that cluster \(1\) also has a greater proportion of skilled
-trades than other clusters. In any case, it appears that the first level
-digital divide reinforces other inequities in cluster \(1\), where the
-even slower than average morning upload speeds shown in Table
-\ref{up.cluster.descr} suggest that internet users were more active
-during the working day. Meanwhile, those in cluster \(13\) are more
-likely to earn more and have a better education despite poor internet
-services. Cluster \(13\) also has the most businesses per inhabitant,
-but is somewhere in the middle in terms of job density. If this is an
-indication of a high number of SMEs, it could explain the variable
-internet quality, considering that small businesses have not been seen
-as the most valuable customers for higher speed broadband services if
-they are not located near residential customers \citep{ofcom2016}.
-
-Cluster \(6\) is the largest cluster, and thus, like cluster \(13\),
-there is more noise within the averages we use to measure internet
-quality and reliability. Our results indicate average mean speeds, and
-Table \ref{aux} shows that cluster \(6\) also falls towards the middle
-of the clusters on many of the socioeconomic variables. It ranks third
-or fourth out of the eleven in terms of the likelihood of having a
-higher proportion of managerial, professional, and tech jobs, as well as
-higher earnings, but is fourth from bottom for educational attainment.
-The LADs in cluster \(6\) are also diverse, with few truly remote areas,
-but urban areas throughout England, including Birmingham, Leeds,
-Sheffield, twelve London Boroughs, and many suburban areas and smaller
-cities like Oxford and Cambridge. The capital cities of the other UK
-nations, Belfast, Cardiff and Edinburgh, are also in this cluster,
-suggesting perhaps that the lower level of reliability discussed in
-Section \protect\hyperlink{sec:4.1}{4.1} may be due to increased demand,
-e.g.~for telecommuting, despite lacking the most resilient internet
-connections.
-
-In contrast, Clusters \(9\) and \(11\) enjoy resilient internet
-connections, but LADs in Cluster \(9\) are more likely to host highly
-educated individuals with higher earnings than cluster \(11\), which has
-a negative coefficient for the NVQ4 variable. Cluster \(11\) has fewer
-businesses per inhabitant and the second highest job density and cluster
-\(9\) the second lowest. These coefficients might indicate that
-resilient broadband infrastructure generates higher returns for those in
-cluster \(9\), where slightly more slowdown in the AM peak was detected
--- see Table \ref{up.cluster.descr}. LADs in cluster \(9\) are less
-likely to be in the South and those in cluster \(11\) more likely,
-although the coefficient for cluster \(9\) might be skewed by the
-presence of the Scottish cities of Glasgow and Aberdeen. Scotland has a
-different economic profile than England. Still, both clusters consist of
-larger LADs in terms of population, including districts within five of
-England's ten largest cities, and a number of other stand-alone urban
-areas and large market towns -- see Appendix
-\protect\hyperlink{appendix1}{1}. These urban locations are on the right
-side of the first level digital divide, and the regression results
-suggest that internet resilience is supporting a wide range of small and
-large urban economies.
-
-LADs in clusters \(2\), \(3\), \(7\), \(10\) are also on the right side
-of the first level digital divide, experiencing higher than average mean
-upload speeds and ranking high to middle on measures of service
-reliability. LADs in clusters \(2\) and \(10\) are also more likely to
-have highly skilled workers to take advantage of working from home
-opportunities. Cluster \(2\) is comprised of just two LADs, with the
-lowest job density of any cluster. Cluster \(10\) is comprised of four
-LADs, including two peripheral suburban areas of Birmingham, the East
-London Borough of Newham, and Dundee. Suburbs are considered the most
-likely urban form in which telecommuters live \citep{e2018does}, and
-Dundee has a reputation for tech startups \citep{technation2017}. LADs
-in cluster \(7\), which include the suburbs South of Belfast, a suburban
-district of Greater Manchester, and some other Midlands and North
-Yorkshire towns and villages have the highest earnings, and thus may
-also be making the most of their resilient internet. In contrast, LADs
-in cluster \(3\) are associated more with lower skills and have the
-highest proportion of individuals in machine operation jobs. Arguably,
-LADs in clusters \(3\) benefit the least from their resilient internet
-infrastructure in an era when working from home became a vehicle for
-economic resilience.
-
-Finally, clusters \(8\) and \(12\) consist of LADs enjoying close to
-average mean upload speeds, but opposing patterns of internet
-reliability during the morning peak. LADs in cluster \(8\) are likely to
-host more individuals employed in tech and administrative occupations
-than any other cluster, as well as many in managerial occupations,
-whilst the opposite applies to skilled trade and leisure occupations.
-These LADs are characterised by a very small positive likelihood of more
-educated residents, but are not significantly likely to earn more than
-other clusters. Including suburbs near Leicester, Cardiff, and
-Birmingham and south of London, the LADs in cluster \(8\) seem to likely
-to have the skills and occupations that would benefit from quality
-internet services, but suffer from poor internet resilience and
-reliability.
-
-In comparison, LADs in cluster \(12\) are less likely to be able to
-benefit from quality internet services if they had them, with fewer
-individuals achieving NVQ4 or better and lower levels of occupations
-that would benefit from homeworking. Yet LADs in this cluster benefit
-from the second highest level of earnings, and have the largest stock of
-businesses per inhabitant and the highest job density. This density of
-businesses could be why cluster \(12\) is the only cluster with higher
-speeds in the morning peak during the study period. If people are at
-home, business premises might well have been abandoned. Cluster \(12\)
-is home to \(1.5\) million people spread across twelve LADs, from London
-to Wales as listed in Appendix \protect\hyperlink{appendix1}{1}. This
-spatial diversity demonstrates that the temporal clustering of internet
-resilience is not necessarily spatially dependent, and digital divides
-do not necessarily overlap with economic ones.
+Virgin Media internet connections \footnote{See Appendix
+  \protect\hyperlink{appendix3}{3} for the descriptive statistics.}.
+Regarding the former, we expect people in LADs with more unreliable
+connections to test their internet speeds more often, and the results
+show that those in cluster \(6\) are by far the most concerned about
+reliability at that time of day. Meanwhile, those in clusters \(2\),
+\(3\), and \(9\) benefit from a higher proportion of Virgin connections,
+which is an indication that people in these clusters are more likely to
+live in urban areas, with more profitable broadband markets, and a
+better choice of broadband services. We also employ distance from London
+and from the nearest metropolitan area (including London) as two
+variables depicting peripherality and urban structure. However, whilst
+significant, the pattern of coefficients of these variables is
+inconclusive in confirmed first level digital divides. Even though
+London was one of the ten largest metropolitan areas in England, which,
+along with Glasgow and Cardiff, were identified to calculate the
+variable estimating the impact of distance from the centre of a
+metropolitan area, the coefficients for the two variables are mostly
+small and the signs for some are opposite.
+
+When looking at the constituent authorities (see full list of LADs in
+Appendix \protect\hyperlink{appendix1}{1} as well as Figure
+\ref{map.up.clusters}), however, Cluster \(9\) is clearly more urban,
+including \(13\) of \(32\) London Boroughs, five of the seven
+constitutent LADs of the West Midlands conurbation, nine of the ten
+boroughs of Greater Manchester, and five of nine other large
+metropolitan areas coded in for the `distance to nearest met area'
+variable. Cluster \(9\) also includes the other main cities in the East
+Midlands, Leicester and Derby, as well as smaller cities known for their
+knowledge economy, such as Oxford and Milton Keynes. However, as
+previously mentioned, there is more noise within the variables for the
+two largest clusters than the smaller ones, and cluster \(6\) also
+contains many urban areas. These include \(16\) London boroughs,
+Birmingham City, Bolton in Greater Manchester, Leeds, Liverpool,
+Newcastle and Bristol, as well as small cities with knowledge economies
+such as Cambridge, Edinburgh, and Reading and its neighbours in the
+high-tech agglomeration of the Thames Valley. And yet cluster \(6\) also
+includes some of the most rural areas in the country, showing how
+important it is to interpretation to consider the membership of the
+cluster as well as the regression results.
+
+The coefficients in Table \ref{aux} demonstrate that those in cluster
+\(6\) were more likely to hold managerial, tech and professional jobs
+than those in cluster \(9\), and were the least likely to be furloughed
+of any cluster. Thus, although speeds were slow and not as reliable
+during the morning peak as in cluster \(9\), the skills that enabled
+telecommuting also enabled greater returns from doing so, despite lower
+than average earnings. Cluster \(6\) also has the second highest job
+density, with more jobs to keep. Likewise, those in LADs in clusters
+\(1\) and \(8\) had the highest and second highest proportions of
+residents working in tech or managerial occupations despite suffering
+from lower speeds and unreliable services, another mismatch between
+first and second level digital divides. Yet unlike cluster \(6\) and
+despite occupations with better digital skills, cluster \(8\), with its
+five peripheral suburban areas in the Midlands, Scotland and south of
+London, as well as rural West Devon saw the third highest proportion of
+its working population furloughed. It also had the second lowest average
+earnings in 2019, suggesting that its population was more likely to be
+on the wrong side of the third level digital divide before, as well as
+during the pandemic. In contrast, the nine LADs of cluster \(1\), which
+include more rural areas, plus Westminster in central London had the
+lowest earnings of any cluster, but also lowest level of furlough after
+cluster \(6\).
+
+The regression results for cluster \(7\) also reveal the complexity of
+intersecting and diverging digital divides. Along with \(3\) and \(9\),
+the analysis in Section \protect\hyperlink{sec:4.1}{4.1} suggested this
+cluster was on the `right' side of any first level digital divides, yet
+the regression shows that it has the lowest proportion of Virgin
+connections. It is furthest from a metropolitan area, but closest to
+London. The five LADs in cluster \(7\) are all suburban, making it
+sensible that the cluster is in the group with better infrastructure,
+but none are central urban boroughs and, along with a London suburb,
+Leeds suburb, Manchester suburb, and West Midlands suburb, the fifth LAD
+is outside Belfast. Belfast was not included among the metropolitan
+areas, and whilst there has been substantial investment in high speed
+broadband in Northern Ireland in recent years, including by Virgin,
+take-up may still lag behind historic networks in other urban markets.
+
+Cluster \(7\) has the lowest proportion of residents currently employed
+in professional occupations, and among the lowest proportions in
+managerial or technical occupations. This suggests potentially a low
+level of skills to take advantage of the quality internet service
+available, a second level digital divide. Cluster \(7\) also has the
+lowest job density. However, with the largest proportion of skilled
+tradespeople, and the second highest number of businesses per
+inhabitant, those in cluster \(7\) ranked second in terms of earnings in
+2019. On the other hand, the lack of some of the occupations most likely
+to telecommute during lockdown may have contributed both to a
+substantial proportion being furloughed in 2020, and even to the
+relative reliability of broadband speeds compared to many of the other
+smaller clusters.
+
+Cluster \(3\), meanwhile, had the greatest proportion of its workforce
+furloughed of any cluster, despite having the highest job density and
+the second highest average speeds. Ranking in the middle for most of the
+occupations estimated, at least some of its population should have the
+skills to work from home. The geography would suggest economic advantage
+as well, with eight of its twelve LADs in the South of England,
+including two London Boroughs, four suburban areas north of London, and
+the two tightly bounded urban areas of Eastbourne and Ipswich -- a
+greater proportion than any other cluster. Yet geographic position and
+job density do not appear to equate to particularly high average
+earnings in 2019, nor did they protect residents from furlough in 2020.
+In contrast, cluster \(5\), comprised of three suburban districts in the
+Northwest, Midlands, and the London green belt, had the highest average
+earnings in 2019, despite low levels of residents working in managerial,
+tech and professional occupations. Suburbs are considered the most
+likely urban form in which telecommuters live \citep{e2018does}, yet the
+contrast between these small, mainly suburban clusters -- \(7\), \(3\),
+and \(5\), suggest that they are not all equally economically resilient
+in a pandemic.
+
+Cluster \(2\), which, like cluster \(5\), had faster morning broadband
+speeds than evening speeds, as shown in Table \ref{up.cluster.descr}
+contains two remote LADs, the Isles of Scilly and Northeast
+Lincolnshire. Whilst the low demand may be due to the least residents in
+tech occupations, it may also be due to high numbers of retired people.
+Rural areas such as those in cluster \(2\) are often home to many older,
+retired people \citep{blank2018local}, which may also explain some of
+the contradictions in cluster \(1\) with its low levels of furlough and
+low earnings, whilst the presence of Westminster in that cluster might
+be why there are the highest levels of tech and managerial occupations.
+
+In summary, the regression results indicate the complexity of measuring
+second and third level digital divides within spatial aggregations where
+the geography of first level digital divides has been captured using
+time-series clustering of experienced broadband upload speeds as a
+product of reliability not just availability. Whilst the differences in
+mean speeds between clusters were not large, the temporal clustering of
+internet resilience showed much greater variation, and was not spatially
+dependent upon distance to large urban areas, or relative location.
+Internet resilience supports a wide range of small and large urban
+economies, but also fails, or at least frustrates a wide range of other
+urban economies, as demonstrated by the number of morning peak tests per
+capita run in cluster \(6\). Yet our analysis of the likelihood of
+furlough within the population, especially in the smaller suburban
+clusters \(7\), \(3\), \(5\), and \(8\) shows that digital divides were
+as likely to diverge as intersect during the pandemic, and did not
+necessarily overlap with prior economic divides. With a much greater
+share of the population likely to continue to work from home in the
+future, and with changing attitudes towards residential locations, our
+analysis suggests that first level digital divides should be seen as a
+function of reliability as well as availability, of upload speeds as
+well as download speeds, and of greater spatial variation than might
+have previously been considered.
 
 \hypertarget{sec:5}{%
-\section{Conclusions}\label{sec:5}}
-
-This paper offers a new perspective on telecommuting from the viewpoint
-of the complex web of digital divides. We employ novel data regarding
-experienced upload speeds and time-series clustering methods, a family
-of unsupervised machine learning techniques which are rarely utilised in
-geographical research. Fast, reliable internet connections are necessary
-for the population to be able to work from home. Although not every
-place hosts individuals in occupations which allow for telecommuting nor
-with the necessary skills to effectively use the internet to
-telecommute, this paper raises the issue that places without good
-internet connectivity will still struggle more than other places to
-achieve economic resilience in a period like the current pandemic when
-internet resilience is so vital. Indeed, our analysis demonstrated that
-the temporal profiles of twelve of our thirteen clusters had slower
-upload speeds in the morning than in the evening. The opposite is likely
-to have been the norm prior to the pandemic, as level of demand and
-bandwidth management is the most common cause of temporal variation in
-experienced speeds, and why evening download speeds, rather than daytime
-upload speeds, have been used to benchmark the performance of internet
-services. Thus, the new patterns can be taken as evidence of widespread
-telecommuting and other daytime internet use which changed the temporal
-profile of internet activity throughout the UK, not just in areas with
-more digital industry or better skills.
-
-Upload speeds have not previously been seen as integral to universal
-service, considering there has never before been such extreme demand for
-telecommuting and operations such as video calls. This may be why the
-average upload speeds for the largest two clusters -- \(6\) and \(13\)
--- contain so many LADs that are centres of the knowledge economy, from
-Oxford and Cambridge in the former, to areas like Reading in the latter,
-which has the highest concentration of digital businesses in the country
-\citep{technation2017}. Thus, whilst some areas suffer from an
-intersection of digital and economic divides, such as those in cluster
-\(1\), other areas, including cluster \(8\) as well as \(6\) and \(13\)
-also found their digital infrastructure to be less than reliable when
-confronted with the sudden change in the timing and type of demand. In
-contrast, LADs such as the digital hub, Milton Keynes, in cluster \(9\)
-were able to benefit both from reliable internet connections and
-populations which were familiar with working from home and could
-capitalise on their digital infrastructure. Yet the nuanced picture we
-gained through our analysis of the UK case study suggests that quality
-internet connectivity may also have enabled other LADs, such as those in
-cluster \(3\) or \(11\) to gain ground during the pandemic despite lower
-skill levels.
-
-Digital infrastructure which considers upload speeds and working day
-reliability as well as availability are likely to be particularly
-important in a future where telecommuting might be a more common means
-of accessing work and broadband services must be fit for purpose,
-although the long-term effects of such drastic changes in telecommuting
-and attitudes towards working from home are difficult to predict.
-Nevertheless, they span various aspects of economy and society: from
-changes to transportation planning due to altered commuting patterns to
-changes in land use and urban planning to accommodate people who work
-from home \citep{BUDNITZ2020102713, ELLDER2020102777}, and from
-productivity and innovation changes to changes in agglomeration
-externalities and the attraction of large cities \citep{econobs}.
-Further research may be able to measure the economic resilience of the
-different clusters of places discussed in this paper once this pandemic
-is firmly past. However, our analysis demonstrates that the economic
-resilience made possible by working from home cannot be understood
-without considering the underpinning digital divides and cannot be
-achieved without planning for how the levels of digital, social and
-economic divides might intersect.
+\section{Discussion and Conclusions}\label{sec:5}}
+
+Our analysis demonstrated that the temporal profiles of seven of our
+nine clusters had slower upload speeds in the morning than in the
+evening. The opposite is likely to have been the norm prior to the
+pandemic, as level of demand and bandwidth management is the most common
+cause of temporal variation in experienced speeds, and why evening
+download speeds, rather than daytime upload speeds, have been used to
+benchmark the performance of internet services. Thus, the new patterns
+can be taken as evidence of widespread telecommuting and other daytime
+internet use which changed the temporal profile of internet activity
+throughout the UK, not just in areas with more digital industry or
+better skills. Furthermore, upload speeds have not previously been seen
+as integral to universal service, considering there has never before
+been such extreme demand for telecommuting and operations such as video
+calls. Yet those in the largest cluster, \(6\), clearly experienced
+lower upload speeds. Their speeds were also less reliable, perhaps
+because they had higher levels of demand than the second largest
+cluster, \(9\), where many more employees were furloughed. Indeed, those
+with the skills and jobs to work from home were often left with the
+least robust services and greatest slowdown.
+
+Therefore, home-based digital infrastructure which considers upload
+speeds and working day reliability as well as availability are likely to
+be particularly important in a future where telecommuting might be a
+more common means of accessing work and broadband services must be fit
+for purpose. Although the long-term effects of such drastic changes in
+telecommuting and attitudes towards working from home are difficult to
+predict, the reliability of home broadband services deserve more
+consideration than in the past. This would represent a switch from
+previous demand-side broadband policies in the UK which tended to be
+aimed at supporting small and medium enterprises
+\citep{HENDERSON2020102024}. SUch policies were based on previous
+research regarding the productivity effects of broadband infrastructure
+\citep{DESTEFANO2018110}. However, this stream of research tended not to
+consider residential locations as places that host economic activity.
+Such policies also made some assumptions about the quality of service in
+residential markets that could be assured without government
+intervention -- an assumption which our analysis suggest is not entirely
+accurate.
+
+Policy and development proposals should also consider how the potential
+changes wrought by the pandemic span various aspects of economy and
+society. Changes to transportation planning due to altered commuting
+patterns also imply changes in land use and urban planning to
+accommodate people who work from home
+\citep{BUDNITZ2020102713, ELLDER2020102777}. Productivity and innovation
+changes will reflect upon changes in agglomeration externalities and the
+attraction of large cities \citep{econobs}. Further research may be able
+to measure the economic resilience of the different clusters of places
+discussed in this paper once this pandemic is firmly past. However, our
+analysis demonstrates that the economic resilience made possible by
+working from home cannot be understood without considering the
+underpinning digital divides and cannot be achieved without planning for
+how the levels of digital, social and economic divides might intersect.
+For example, our results suggest that broadband policies cannot improve
+the economic resilience of places where the industrial structure does
+not align with occupations that incorporate the digital skills and
+capabilities to work from home. Such places instead experienced higher
+proportions of their labour force being placed on furlough.
+
+Early research on Covid-19 and cities also speculated upon the changes
+that potential extensive post-Covid working from home patterns might
+generate for spatial structure: from more walkable cities and more
+localised production and consumption patterns to more extensive urban
+sprawl, the decline in public transportation and increased private car
+usage \citep{batty2020editorial}. Despite the essential role it played
+during the pandemic, less effort has been spent in understanding the
+current and future role of digital infrastructure. Our results indicate
+that probably for the first time the future of cities and spatial
+structure are so intertwined with digital infrastructure. If the
+post-covid world is a world with extensive working from home, then we
+need to build, among other things, resilient digital infrastructure
+capable of bridging the first layer of the digital divide. Contrary to
+previous broadband business and deployment plans, emphasis should be
+placed not only on download, but also on upload speeds
+\citep{brake2020lessons}. In essence, Covid-19 accentuated the old
+argument, now more valid than ever: digital infrastructure, just like
+any other network infrastructure, only becomes visible when it stops
+working \citep{star1999, tranos2013geography}.
+
+On the other hand, the nuanced picture we gained through our analysis of
+the UK case study suggests that being on the right side of the second
+level digital divide had a greater impact on economic resilience and
+therefore the third level digital divide, than having quality internet
+connectivity. Our regressions results for cluster \(1\) and \(6\) are a
+demonstration of this. Our analysis shows both the intersections and
+divergences of digital and economic division. Almost all of the largest
+urban areas in the UK, as well as many smaller cities, were split
+between the largest two clusters -- \(6\) and \(9\) -- and both contain
+LADs that are centres of the knowledge economy, from Cambridge in the
+former to Oxford in the latter, or have high concentrations of digital
+businesses, like Reading in the former and Milton Keynes in the latter
+\citep{technation2017}. Yet whilst LADs in cluster \(9\) were able to
+benefit both from reliable internet connections and populations familiar
+with working from home to capitalise on their digital infrastructure,
+they appeared to have a lower rate of return in the pandemic, with far
+greater numbers furloughed. Is this an indication of historic economic
+division, as well as the third level digital divide, in cities in the
+North and Midlands unable to capitalise on their digital infrastructure?
+Meanwhile, did the greater presence in London and the South of LADs in
+cluster \(6\), as well as the inclusion of many rural areas less
+impacted by the pandemic help those in this cluster be more resilient
+and gain greater returns, despite using less reliable internet services?
+Further research would be necessary to prise apart the detail,
+particularly in the largest clusters where averages are less distinct.
+
+In conclusion, this paper offers a new perspective on telecommuting from
+the viewpoint of the complex web of digital divides. We employ novel
+data regarding experienced upload speeds and time-series clustering
+methods, a family of unsupervised machine learning techniques which are
+rarely utilised in geographical research. Fast, reliable internet
+connections are necessary for the population to be able to work from
+home. Although not every place hosts individuals in occupations which
+allow for telecommuting nor with the necessary skills to effectively use
+the internet to telecommute, this paper raises the issue that places may
+depend upon good internet reliability as well as connectivity to achieve
+economic resilience in a period like the current pandemic when internet
+resilience is so vital.
+
+\pagebreak
 
@@
 \addcontentsline{toc}{section}{Appendix 2}
+
+\begin{figure}
+\includegraphics[width=1\linewidth]{figures/map.up.2019.clusters} \caption{\label{map.up.2019.clusters}Upload speed clusters for LADs in before the pandemic (2019)}\label{fig:unnamed-chunk-17}
+\end{figure}
+
+\pagebreak
+
+\hypertarget{appendix3}{%
+\section*{Appendix 3}\label{appendix3}}
+\addcontentsline{toc}{section}{Appendix 3}
 
@@
 
+\newpage
+
 \bibliographystyle{tfcad}

### 2021-04-13 src/descriptive_Clusters.Rmd
Tweaked lit review with new reference added. Cleaned up descriptive_Clusters.Rmd.

@@
 
-#summarise by LADs per cluster and Total tests / Mean speed per cluster
-#Test Frequency cluster
-LAD.cluster.tests <- clusters %>%
-  group_by(cluster.tests) %>%
-  summarise(n.LADs = n())
-LAD.cluster.pop <- clusters %>%
-  group_by(cluster.tests) %>%
-  summarise(pop.LADs = sum(pop))
-LAD.cluster.testsTotal <- TS2020 %>%
-  group_by(cluster.tests) %>%
-  summarise(n.cluster = n()) 
-LAD.cluster.tests19 <- TS2019 %>%
-  group_by(cluster.tests) %>%
-  summarise(n.cluster19 = n()) 
-LAD.cluster.tests <- cbind(LAD.cluster.tests, LAD.cluster.pop [2], LAD.cluster.testsTotal[,2], LAD.cluster.tests19[,2])
-LAD.cluster.tests$TestsPop20 <- LAD.cluster.tests$n.cluster / LAD.cluster.tests$pop.LADs
-LAD.cluster.tests$TestsPop19 <- LAD.cluster.tests$n.cluster19 / LAD.cluster.tests$pop.LADs
-LAD.cluster.tests$n.20_19.diff <- LAD.cluster.tests$n.cluster - LAD.cluster.tests$n.cluster19
-LAD.cluster.tests$perCent.change <- LAD.cluster.tests$n.20_19.diff/
-  LAD.cluster.tests$n.cluster19
-LAD.cluster.tests$perCent.change <- paste(round(100*LAD.cluster.tests$perCent.change, 2), "%", sep="")
+#summarise by LADs per upload cluster
 #Upload Speed Clusters
@@
 LAD.cluster.up$SDtoMean <- paste(round(100*LAD.cluster.up$SDtoMean, 2), "%", sep="")
-#Download speed clusters  
-LAD.cluster.down <- clusters %>%
-  group_by(cluster.down) %>%
-  summarise(down.LADs = n())
-LAD.cluster.downPop <- clusters %>%
-  group_by(cluster.down) %>%
-  summarise(pop.LADs = sum(pop))
-LAD.cluster.downMean <- summarise(group_by(TS2020, cluster.down),
-  mean(speeddown))
-LAD.cluster.downSD <- summarise(group_by(TS2020, cluster.down),
-  sd(speeddown))
-LAD.cluster.downAM <- summarise(group_by(TS2020[which(TS2020$hour == 9 | TS2020$hour == 10),], cluster.down), mean(speeddown))
-LAD.cluster.downPM <- summarise(group_by(TS2020[which(TS2020$hour == 19 | TS2020$hour == 20),], cluster.down), mean(speeddown))
-LAD.cluster.down <- cbind(LAD.cluster.down, LAD.cluster.downPop[,2], LAD.cluster.downMean[,2], LAD.cluster.downSD[,2], LAD.cluster.downAM[,2], LAD.cluster.downPM[,2])
-names(LAD.cluster.down)[4:7] <- c("downMean", "downSD", "downAMmean",
-                                  "downPMmean")
-LAD.cluster.down$AM_slower <- (LAD.cluster.down$downPMmean - LAD.cluster.down$downAMmean) / LAD.cluster.down$downPMmean
-LAD.cluster.down$SDtoMean <- LAD.cluster.down$downSD / LAD.cluster.down$downMean
-LAD.cluster.down$AM_slower <- paste(round(100 * LAD.cluster.down$AM_slower, 2), "%", sep="")
-LAD.cluster.down$SDtoMean <- paste(round(100*LAD.cluster.down$SDtoMean, 2), "%", sep="")
-# commented to run the script
-# LAD.cluster.diffUp <- clusters %>%
-#   group_by(cluster.up.diff) %>%
-#   summarise(diffUp.LADs = n())
-# LAD.cluster.diffDown <- clusters %>%
-#   group_by(cluster.down.diff) %>%
-#   summarise(diffDown.LADs = n())
-
-#write csvs to outputs
-out.clustersLAD <- paste(path, "./data/temp/LAD.cluster.tests.csv", sep = "")
-write_csv(LAD.cluster.tests, out.clustersLAD)
+#write csv to outputs
 out.clustersLADup <- paste(path, "./data/temp/LAD.cluster.up.csv", sep = "")
 write_csv(LAD.cluster.up, out.clustersLADup)
-out.clustersLADdown <- paste(path, "./data/temp/LAD.cluster.down.csv", sep = "")
-write_csv(LAD.cluster.down, out.clustersLADdown)
 
-```
-
-As people came to rely more on their broadband services, it is unsurprising that both the amount and rate of testing increased in 2020 during the lockdown compared to the same period in 2019.
-This increase was greatest in cluster 2, where 56% more tests were run compared to the same period in 2019.
-Cluster 2 also has the most local authorities, 148, which are home to more than 28.5 million people, so the sum of tests run in all these authorities is much higher than in all the other clusters, both in 2019 and 2020.
-Cluster 1 in comparison, not only has has the lowest  frequency of testing per capita in both 2019 and 2020, it also shows the lowest rate of increase in testing between the two time periods at just 33%.
-
-Meanwhile, cluster 5 has the fewest local authorities, at 28, but the second highest total number of tests in both periods and the second largest increase in testing between 2019 and 2020, at 52%.
-It also has the highest number of tests per capita in both periods.
-A possible explanation is that cluster 5, with a population of just under 12 million contains some of the most populous local authorities in the country, including the cities of Birmingham, Manchester, Leeds, Sheffield, Glasgow and Edinburgh.
-However, this cluster also includes large rural counties, such as Cornwall and Northumberland.
-Are urban dwellers or rural households more likely to test their broadband speeds?
-Auxiliary regressions suggest that cluster 5 also had the highest earnings of the clusters, but was least likely to have a Virgin Media connection.
-
-Yet the temporal profile in 2020 is very different to 2019, as can be seen in graphs [.] and [.].
-The number of broadband speed tests run in local authorities in cluster 2 peaked each Monday to Thursday in the hour 19.00-19.59 in March through May 2019.
-On Fridays, the peak was slightly earlier.
-A similar, although less stark pattern could be seen in the other clusters, along with much smaller peaks around the beginning of the working day, at either 9.00-9.59 or 10.00-10.59.
-In comparison, the daily profile of broadband speed testing in 2020 is flatter, with less variation hour to hour.
-Although there is still an identifiable evening peak for almost all days of the week and all clusters, it is much less pronounced throughout, there is less of a gap between the number of tests in the morning peak and the evening peak, and there is less of a dip in testing in the middle of the day.
-All this suggests that more of the additional testing generated in 2020 compared to 2019 is occurring outside the traditional peak in the evening at streaming primetime.
-
-The increase in daytime compared to evening testing is particularly noticeable for cluster 2, but the change is less pronounced for cluster 5.
-Indeed, in an auxiliary regression, the probability of additional tests per person run between 9.00 and 11.00 was significantly higher for those living in cluster 2 than in cluster 5, but both were higher than the other three clusters.
-??
-perhaps because of the diversity of urban and rural populations represented in cluster 5 ??
-Cluster 5 mostly same as cluster 6 upload, few cluster 9, but same for cluster 2?
-#re-base auxiliary regression so reference case is either 1 or 3 - we wanted to talk most about 5?
-The graphs also show that Cluster 1, with the lowest increase in testing, falls from the middle position in terms of total number of tests to the second from bottom.
-
-For upload speeds, 345 of 382 local authorities, or over 62 million people, fall into cluster 6 or cluster 9.
-Graph [.] shows that both of these clusters have relatively similar temporal profiles, with relatively stable speeds.
-However, the upload speeds at all times for cluster 9 are substantially higher than for cluster 6.
-This difference is reflected not only in the mean speeds for these clusters for the whole sample, but also the mean upload speeds in the morning peak from 9:00-10:59, as well as the evening peak period from 19:00-20:59.
-The quality of broadband service is thus much better in the 115 Local Authority areas in cluster 9, which are mostly in urban or suburban areas.
-These include 13 London Boroughs (of 32), 8 of the 10 local authorities of Greater Manchester, 5 of the 7 constituent authorities of the West Midlands Combined Authority, as well as cities like Glasgow, Leicester, Nottingham, Sheffield, and the Portsmouth and Southampton conurbation.
-There are also some notable medium-sized cities, including Aberdeen, Cardiff, Oxford, Milton Keynes, and York, and many suburban districts from the South East of England to South Tyneside.
-Auxiliary regressions indicate that the speed tests in these authorities are also more likely to have been run on services provided by Virgin Media, suggesting they are in the half of the country with the most lucrative ICT market, which originally attracted the cable TV provider (OfCom...).
-
-Meanwhile, the 230 local authorities in cluster 6 still include major urban areas, such as Bristol, Liverpool and Leeds, and many suburban areas, but also include some of the most rural areas in the country.
-For example, although auxiliary regressions show that Cluster 6 local authorities are more likely to be in the South of the UK than Cluster 9, the cluster notably includes Southern rural districts from Cornwall to North Norfolk.
-Slower speeds could reflect the lower quality of service in rural areas compared to urban and suburban areas.
-The reliability of service in these areas is also in question.
-For although upload speeds in cluster 6 are only 4% slower in the morning than in the evening, compared to 12-19% slower for the almost 3 million people living in five of the smaller clusters, auxiliary regressions show that those living in cluster 6 had the highest probability of testing their broadband between 9:00 and 11:00 of any of the 10 upload speed clusters.
-Cluster 6 also experienced the joint highest ratio of standard deviation to mean across the time period under assessment.
-This suggests that although the time profile is relatively flat in graph [.], the experience is one of speeds that fluctuate from a lower mean, and therefore might more often impact on the activities users are doing.
-
-In comparison, there is negligible difference between morning and evening upload speeds for cluster 9, at 1% slower in the morning.
-This suggests a high level of reliability of service, a conclusion supported by the finding in the auxiliary regressions that those in cluster 9 ran the fewest speed tests per person during the morning period of any cluster.
-Now this may be an indication of fewer people working from home, less contention, and less resultant frustration.
-Cluster 9 comprises many central urban areas and has the lowest number of established businesses per inhabitant, which could be interpreted as a dominance of large employers.
-However, the job density is lower than cluster 6, meaning there are not as many jobs per resident in these areas.
-As the cluster includes many suburban areas too, which may be largely residential, could the quality and reliability of internet service be reinforcing patterns of telecommuting by those in wealthier suburbs who can work from home?
-Earnings in cluster 9 are second highest of all the clusters, with only Cluster 2 (comprising just North East Lincolnshire and East Lothian, population 265k) earning more per person.
-
-Meanwhile, Cluster 1, with 10 local authorities that are home to over 1 million people has the second slowest speeds in the morning compared to the evening 'peaks' and the second highest ratio of standard deviation to mean.
-Indeed, the time profile in graph [.] shows upload speeds in cluster 1 are on average lower at certain times of day during the study period than any other cluster.
-Yet the auxiliary regression suggests that there are not many tests being run during the am peak in cluster 1.
-This may be because there are fewer people working at home checking their broadband than in most other clusters.
-Indeed, the auxiliary regressions indicate that cluster 1 has the highest job density or proportion of jobs to working-age population, which is likely to due to the presence of Westminster, central London, cluster 1's most populous local authority.
-Westminster not only has more workplaces than residents, but it is reasonable to presume that many who would normally work in Westminster, but be able to work from home during lockdown are likely to live outside central London and not be subject to the fluctuating speeds there.
-Workplaces, meanwhile, some of which would still have been open, could be running programmes that cause the slowdown and variation, but would be more likely to have their own in-house diagnostics, rather than using a service like Speedchecker Ltd.
-
-Indeed, the only cluster where upload speeds were slower in the evening than in the morning was cluster 5, made up of 2 local authorities with less than 200,000 people: Three Rivers, a suburban district north of London, and Fylde, a seaside suburb of Blackpool.
-However, these are likely to be outliers and may not have many tests from which the clusters are calcualted.
-**Take out smaller clusters!?** The auxiliary regressions show that compared to these two places, all the other clusters had a lower percentage of working people in managerial, professional and administrative jobs.???
-
-This is an indication of the reliability **or maybe not - 5% slowdown from low base can be noticeable, and sd of cluster 6 bigger** of broadband services in places in both clusters, and suggests that upload speeds have generally been robust throughout the study period.
-The exceptions can be found in seven of the eight much smaller clusters including 35 local authority districts, where AM peak upload speeds are between 6% and 18% slower than PM peak upload speeds, although the mean speeds for each cluster are higher than cluster 6.
-The temporal profiles on Graph [.] for the three of those clusters with more than 5 local authorities also shows that speeds are far from dependable in these clusters.
-Indeed, in 25 local authorities with a combined population of almost 3 million, speeds are 13% or more slower in the morning than in the evening.
-Included in this latter group are central London borough of Westminster and the London Borough of Newham, rural authorities like Eden and West Devon, and small cities like Dundee and Carlisle.
-
-Download speeds are the most commonly used measure of the quality and reliability of broadband services.
-They are most sensitive, which does result in substantial noise.
-This is clear from the way that download speeds vary the most, with the largest range and standard deviations in speed tending to be in a similar range to the mean speeds for each cluster, compared to upload speeds, where standard deviations are half to two thirds the level of the mean speeds.
-This temporal variation can be seen in Graph [.], which shows the 10 of 20 clusters with over five local authorities in them.
-The second largest of these 10 clusters, number 19 with 123 Local Authorities and a population of almost 18 million, along with cluster 1 with 16 local authorities and an additional 1.5million people have lower mean speeds than the other clusters.
-This is consistent throughout the temporal profile on the graph.
-However, cluster 19 also has the second lowest standard deviation at 88% of the mean speed, whilst the standard deviation of cluster 1 is 110% and thus greater than the mean speed.
-
-What is of greater interest, however, is where download speeds are slower in the morning peak when broadband is more likely to be used for work purposes than in the evening peak.
-This difference is most stark in some of the smaller clusters, although since there are only 11 Local Authorities and a little less than 1.2 million people in these clusters altogether, the differences in speeds may be a reflection of quite localised issues.
-The largest two clusters of Local Authorities, which are home to over 39 million people, have speeds only 1-3% slower in the AM peak, and a few clusters have faster download speeds in the AM peak.
-However, there are 4 clusters with over 9 million people in 47 Local Authorities with download speeds 5-8% lower in the morning peak than in the evening peak, a noticeable reduction and one equivalent to the contention monitored in the evening peak by OfCom to measure service performance.
-Is the 5% or greater reduction in download speeds in the morning peak compared to the evening peak suffered in more places by more people than suffer a similar reduction in upload speeds a result of ISP management to support reliable services for telecommuters?
-There are only eight Local Authorities where both upload and download speeds are more than 5% slower in the AM peak than the PM peak.
-
-## Summarise 2020 test frequency clusters by hour of each weekday and plot
-
-```{r }
-cluster.tests <- TS2020 %>%
-  group_by(cluster.tests, weekday, hour) %>%
-  summarise(n.tests = n())
-names(cluster.tests)[1] <- "clusterID"
-cluster.tests$clusterID <- as.factor(cluster.tests$clusterID)
-#add fake dates for plotting
-help.dates <- tibble(help.date.time = c(
-  make_datetime(year = 2020, month = 3, day = 2),
-  make_datetime(year = 2020, month = 3, day = 3),
-  make_datetime(year = 2020, month = 3, day = 4),
-  make_datetime(year = 2020, month = 3, day = 5),
-  make_datetime(year = 2020, month = 3, day = 6)),
-                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
-cluster.tests <- merge(cluster.tests, help.dates, by = "weekday")
-cluster.tests$hour <- paste(cluster.tests$hour, ":00", sep="")
-cluster.tests$help.date.time <- with(cluster.tests, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
-#plot cluster statistic for number tests
-TestCluster <- ggplot(cluster.tests, aes(help.date.time, n.tests, 
-                                         colour = clusterID))
-TestCluster + geom_line() + labs(x = "Weekdays March-May 2020") + 
-  scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
-                    "2020-03-03 15:00:00", "2020-03-04 15:00:00",
-                    "2020-03-05 15:00:00", "2020-03-06 15:00:00"), 
-                    labels = "%a"), labels = 
-                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + geom_point()
-#summarise test frequency clusters by hour of each weekday 2019 and plot
-cluster.tests19 <- TS2019 %>%
-  group_by(cluster.tests, weekday, hour) %>%
-  summarise(n.tests = n())
-names(cluster.tests19)[1] <- "clusterID"
-cluster.tests19$clusterID <- as.factor(cluster.tests19$clusterID)
-#add fake dates for plotting
-help.dates <- tibble(help.date.time = c(
-  make_datetime(year = 2019, month = 3, day = 4),
-  make_datetime(year = 2019, month = 3, day = 5),
-  make_datetime(year = 2019, month = 3, day = 6),
-  make_datetime(year = 2019, month = 3, day = 7),
-  make_datetime(year = 2019, month = 3, day = 8)),
-                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
-cluster.tests19 <- merge(cluster.tests19, help.dates, by = "weekday")
-cluster.tests19$hour <- paste(cluster.tests19$hour, ":00", sep="")
-cluster.tests19$help.date.time <- with(cluster.tests19, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
-#plot cluster statistic for number tests
-TestCluster19 <- ggplot(cluster.tests19, aes(help.date.time, n.tests, 
-                                         colour = clusterID))
-TestCluster19 + geom_line() + labs(x = "Weekdays March-May 2019") + 
-  scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
-                    "2020-03-03 15:00:00", "2020-03-04 15:00:00",
-                    "2020-03-05 15:00:00", "2020-03-06 15:00:00"), 
-                    labels = "%a"), labels = 
-                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + geom_point()
 ```
@@
 ```
-
-```{r eval=F}
-# Not used in the paper
-
-#summarise download speed clusters by hour of each weekday and plot
-cluster.down <- TS2020 %>%
-  group_by(cluster.down, weekday, hour) %>%
-  summarise(mean.down = mean(speeddown))
-names(cluster.down)[1] <- "clusterID"
-names(LAD.cluster.down)[1] <- "clusterID"
-cluster.down <- left_join(cluster.down, LAD.cluster.down)
-cluster.down <- cluster.down[which(cluster.down$down.LADs > 5),]
-cluster.down$clusterID <- as.factor(cluster.down$clusterID)
-#add fake dates for plotting
-help.dates <- tibble(help.date.time = c(
-  make_datetime(year = 2020, month = 3, day = 2),
-  make_datetime(year = 2020, month = 3, day = 3),
-  make_datetime(year = 2020, month = 3, day = 4),
-  make_datetime(year = 2020, month = 3, day = 5),
-  make_datetime(year = 2020, month = 3, day = 6)),
-                     weekday = c("Mon", "Tue", "Wed", "Thu", "Fri"))
-cluster.down <- merge(cluster.down, help.dates, by = "weekday")
-cluster.down$hour <- paste(cluster.down$hour, ":00", sep="")
-cluster.down$help.date.time <- with(cluster.down, as.POSIXct(paste(help.date.time, hour), format="%Y-%m-%d %H:%M"))
-#plot cluster statistic for mean download speeds
-TestCluster <- ggplot(cluster.down, aes(help.date.time, mean.down, 
-                                         colour = clusterID))
-TestCluster + geom_line() + labs(x = "Weekdays March-May") + 
-  scale_x_datetime(breaks = as.POSIXct(c("2020-03-02 15:00:00", 
-                    "2020-03-03 15:00:00", "2020-03-04 15:00:00",
-                    "2020-03-05 15:00:00", "2020-03-06 15:00:00"), 
-                    labels = "%a"), labels = 
-                     c("Mon", "Tues", "Weds", "Thurs", "Fri")) + geom_point()
-
-```

### 2021-04-13 src/ts_clusters_k9.Rmd
online appendix. we need to add some text to make it almost a standalone document.

@@
   xlab(2020)
-  # + xlab("Year") +
-  # ylab("Pareto exponent")
-
-# library(plotly)
-# ggplotly(test,
-#          show.legend=F)
 
@@
 # la <- readOGR(path.json)#, layer="OGRGeoJSON")
+
 # UK BGC: 
 la <- readOGR("https://opendata.arcgis.com/datasets/0e07a8196454415eab18c40a54dfbbef_0.geojson")
-# UK BFC: la <- readOGR("https://opendata.arcgis.com/datasets/1d78d47c87df4212b79fe2323aae8e08_0.geojson")
+# also saved locally \data\raw\Local_Authority_Districts_(December_2019)_Boundaries_UK_BGC.geojson
+# source: https://geoportal.statistics.gov.uk/
 
-# source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe
-# also saved locally \data\raw\Local_Authority_Districts_(December_2019)_Boundaries_UK_BGC.geojson
+# fully generalised
+#la <- readOGR("https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.geojson")
 
@@
   theme_void()
-
-# Run only if needed
-# out.path.map <- paste0(path, "/paper/v2_taylor_francis/figures/map.up.clusters.png")
-# ggsave(out.path.map)
-#width = 210, #A4
-#height = 297, #A4
-#units = "mm")
 ```

### 2021-04-13 paper/v2_taylor_francis/broadband.speed.covid.Rmd

@@
     affil: a
+    email: hannah.budnitz@ouce.ox.ac.uk
   - name: Emmanouil Tranos
-    affil: b, c
+    affil: b, c, \dagger
     email: e.tranos@bristol.ac.uk
@@
       The Alan Turing Institute, UK
+  - num: \dagger
+    address: |
+      Corresponding author
 bibliography: bibliography.bib
@@
 [@stern2009levels; @van2014digital; @van2015third].
-The capability to telecommute is related to all three levels of digital divides,
-but more importantly leads to differentiated outcomes regarding the economic
-resilience of people and places to overcome a systemic shock such as the current
-pandemic.
+The potential for these different levels of digital divide to interact is growing in 
+parallel with the digitisation of various sectors of the economy and society.
+For example, @li2020challenges highlights how the intersection of new demands upon
+ICT infrastructure and a paucity of end users means that some areas are being left
+behind in the roll-out of smart applications and services. They argue that this 
+could create a geospatial 'smart divide', with areas unable to capitalise on the 
+opportunities offered by such applications and services.
+Likewise, the capability to telecommute is related to all three levels of 
+digital divides, but more importantly leads to differentiated outcomes regarding 
+the economic resilience of people and places to overcome a systemic shock such as 
+the current pandemic.
 
@@
 However, the success of these arrangements has been dependent upon the first
-level digital divide, which is associated with access to and quality of internet
-connectivity.
+level digital divide, which is associated with geospatial access to and quality 
+of internet connectivity.
 @SALEMINK2017360 provides a systematic review of the pre-pandemic,
@@
 In contrast, @riddlesden2014broadband found that levels of deprivation did not 
-correlate with first level digital divides.
+correlate with first level digital divides, perhaps because many deprived areas 
+benefit from spillover effects of quality installed infrastructure nearby.
 
 However, as @blank2018local highlight, variation in individual internet 
-availability and uptake is a product of more complex spatial and 
-demographic characteristics than simple rurality or urbanisation.
+availability and uptake is a product of more complex geospatial and 
+demographic characteristics than simple rurality, urbanisation or levels of deprivation.
 Dense urban areas were shown to suffer more from slowdown during peak hours, 
@@
 lucrative and competitive market originally attracted the cable TV provider. 
-Whether the variation in infrastructure quality and reliability affects the 
-spatial footprint of telecommuting has not previously been investigated, in part 
-because telecommuting has not previously been a cause of concentrated demand and 
-pressure on internet services.
+Whether the geospatial variation in infrastructure quality and reliability affects 
+the spatial footprint of telecommuting has not previously been investigated, 
+in part because telecommuting has not previously been a cause of concentrated 
+demand and pressure on internet services.
 
@@
 We also aim to improve our understanding of the impact of first level digital
-divisions on telecommuting, and whether this results in more fundamental third
-level digital divisions than has previously been perceived.
+divisions on telecommuting, in terms of both geospatial and temporal quality of 
+service, and whether this results in more fundamental third level digital divisions 
+between local geographies than has previously been perceived.
 
@@
 which tended to be aimed at supporting small and medium enterprises [@HENDERSON2020102024].
-SUch policies were based on previous research regarding the productivity effects 
+Such policies were based on previous research regarding the productivity effects 
 of broadband infrastructure [@DESTEFANO2018110]. However, this stream of research 

### 2021-04-14 src/online_appendix_v2.Rmd
online appendix v.2

---
title: "Online Appendix"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    social: menu
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../paper/v2_taylor_francis/")
  })
---

```{r, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
library(rmarkdown)
library(tidyverse)
library(plotly)
library(lubridate)
library(gridExtra)
library(openair)
library(rgdal)
library(maps)
library(rprojroot)
library(kableExtra)
#library(cowplot)
library(ggthemes)

path <- find_rstudio_root_file()
```

```{r, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
# load the data created by Data_Spatial.Rmd

path.data <- paste(path, "/data/temp/TSbb19_20sp.csv", sep = "")

TSbb19_20sp <- read_csv(path.data)

#split by year before create stats by each hour of each day of the week
TS2019 <- selectByDate(TSbb19_20sp, year = 2019) #272248 obs
TS2020 <- selectByDate(TSbb19_20sp, year = 2020) #343268 obs

# save clusters file
path.clusters <- paste(path, "/data/temp/clusters_nodiff_9.csv", sep = "")
clusters.nodiff <- read_csv(path.clusters)
```

Column 
-----------------------------------------------------------------------

### LAD, Upload speed clusters, 2020

```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE, fig.height=10}
#out.width='100%'

# LAD fully generalised
la <- readOGR("https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.geojson",
              verbose = FALSE)
# saved locally in /data/raw/Local_Authority_Districts_(December_2019)_Boundaries_UK_BUC.geojson

# spatial transformations
la <- spTransform(la, CRS("+init=epsg:4326"))
#la@data$LAD19NM <- as.character(la@data$LAD19NM)
la@data$lad19nm <- as.character(la@data$lad19nm)

# upload metrics
for.tip.mean.up.am <- TS2020 %>% # raw data, individual tests
  filter(hour(date)==9 | hour(date)==10) %>% 
  group_by(LAD19NM) %>%
  summarise(mean.up.am = mean(speedup)) %>% 
  mutate(rank = dense_rank(desc(mean.up.am)))

# not used
# for.tip.mean.up.pm <- TS2020 %>% # raw data, individual tests
#   filter(hour(date)==19 | hour(date)==20) %>% 
#   group_by(LAD19NM) %>%
#   summarise(mean.up.pm = mean(speedup)) 
# 
# for.tip.mean.up <- TS2020 %>% # raw data, individual tests
#   group_by(LAD19NM) %>%
#   summarise(mean.up = mean(speedup)) 
# 
# # download metrics
# for.tip.mean.down.am <- TS2020 %>% # raw data, individual tests
#   filter(hour(date)==9 | hour(date)==10) %>% 
#   group_by(LAD19NM) %>%
#   summarise(mean.down.am = mean(speeddown)) 
# 
# for.tip.mean.up.pm <- TS2020 %>% # raw data, individual tests
#   filter(hour(date)==19 | hour(date)==20) %>% 
#   group_by(LAD19NM) %>%
#   summarise(mean.down.pm = mean(speeddown)) 
# 
# for.tip.mean.up <- TS2020 %>% # raw data, individual tests
#   group_by(LAD19NM) %>%
#   summarise(mean.down = mean(speeddown))

# fortify
la.f <- broom::tidy(la, region = "lad19nm")
la.f <- merge(la.f, clusters.nodiff, by.x = "id", by.y = "LAD")

# create the text to tooltip
la.f <- merge(la.f, for.tip.mean.up.am, by.x = "id", by.y = "LAD19NM", all.x=T)
#sapply(la.f, function(x) sum(is.na(x)))
#No am data points for Isles of Scilly
la.f$tip <- paste0(la.f$id, "<br>",
                   "Upload AM speed: ", round(la.f$mean.up.am), " Kbps", "<br>",
                   "Rank: ", la.f$rank)
la.f <- la.f[order(la.f$order),] # if i don't order merge.nuts.f loses order and the map has gaps

# using world.cities from `maps`
cities <- world.cities %>% filter(country.etc=="UK")

p <- ggplot(la.f, aes(x = long, y = lat)) +
  geom_polygon(aes(group = group, fill = as.factor(cluster.up), 
                   text = tip)) +# ,text =paste("test1", long, "test2", lat)
  #labs(title = "Upload clusters") +
  scale_fill_brewer(palette="Set1", 
                    direction = -1) +
  #geom_text_repel( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=4) +
  geom_point( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat), color="black", size=0.5) +
  geom_text( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=3) +

  guides(fill=guide_legend(title="Cluster")) +
  theme_map()
#   theme(panel.border = element_blank(), panel.grid.major = element_blank(),
# panel.grid.minor = element_blank())

p %>%
  ggplotly(tooltip = "text") %>%
  style(hoverlabel = list(bgcolor = "white"), hoveron = "fill")
```

Column
-----------------------------------------------------------------------

### LAD mean AM upload speeds and ranking

```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE,fig.height=10}
for.tip.mean.up.am <- merge(for.tip.mean.up.am, clusters.nodiff, 
                            by.x = "LAD19NM", by.y = "LAD", all.x = T)

for.tip.mean.up.am %>% 
  select(-cluster.down, -cluster.tests) %>% 
  arrange(rank) %>% 
  kable(digits = 0,
        format.args = list(big.mark = ","),
        col.names = c("Local Authority",
                      "Mean AM upload sped",
                      "Rank",
                      "Cluster")) %>% 
  kable_styling("striped", full_width = F) %>% 
  scroll_box() #width = "500px", height = "400px"
```

### Description

The below map illustrates the different clusters of LAD based on the upload 
experienced internet speeds. Hovering over the LAD polygons, the map also presents
the mean AM upload internet speed and the relevant ranking in the country for each 
LAD.

**Is there a chance to briefly describe the clusters with 1-2 words?**

- Cluster 1:

- Cluster 2:

...

### 2021-04-16 src/online_appendix.Rmd
online appendix to add text

--- 2021-04-14 src/online_appendix_v2.Rmd
@@
 
-```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE, fig.height=10}
+```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE, fig.height=10, fig.asp=T}
 #out.width='100%'
@@
 
+for.tip.mean.up.pm <- TS2020 %>% # raw data, individual tests
+  filter(hour(date)==19 | hour(date)==20) %>%
+  group_by(LAD19NM) %>%
+  summarise(mean.up.pm = mean(speedup))
+
 # not used
-# for.tip.mean.up.pm <- TS2020 %>% # raw data, individual tests
-#   filter(hour(date)==19 | hour(date)==20) %>% 
-#   group_by(LAD19NM) %>%
-#   summarise(mean.up.pm = mean(speedup)) 
-# 
 # for.tip.mean.up <- TS2020 %>% # raw data, individual tests
@@
 la.f <- merge(la.f, for.tip.mean.up.am, by.x = "id", by.y = "LAD19NM", all.x=T)
+la.f <- merge(la.f, for.tip.mean.up.pm, by.x = "id", by.y = "LAD19NM", all.x=T)
+
 #sapply(la.f, function(x) sum(is.na(x)))
@@
 la.f$tip <- paste0(la.f$id, "<br>",
-                   "Upload AM speed: ", round(la.f$mean.up.am), " Kbps", "<br>",
-                   "Rank: ", la.f$rank)
+                   "Mean upload AM speed: ", round(la.f$mean.up.am), " Kbps", "<br>",
+                   "Mean upload PM speed: ", round(la.f$mean.up.pm), " Kbps")
 la.f <- la.f[order(la.f$order),] # if i don't order merge.nuts.f loses order and the map has gaps
@@
 
-### LAD mean AM upload speeds and ranking
+### LAD mean AM and PM upload speeds
 
@@
                             by.x = "LAD19NM", by.y = "LAD", all.x = T)
+for.tip.mean.up.am <- merge(for.tip.mean.up.am, for.tip.mean.up.pm, 
+                            by.x = "LAD19NM", all.x = T)
 
 for.tip.mean.up.am %>% 
-  select(-cluster.down, -cluster.tests) %>% 
-  arrange(rank) %>% 
+  select(-cluster.down, -cluster.tests, -rank) %>% 
+  arrange(-mean.up.am) %>% 
+  relocate(cluster.up, .after=mean.up.pm) %>% 
   kable(digits = 0,
         format.args = list(big.mark = ","),
-        col.names = c("Local Authority",
-                      "Mean AM upload sped",
-                      "Rank",
+        col.names = c("LAD",
+                      "Mean AM upload speed",
+                      "Mean PM upload speed",
                       "Cluster")) %>% 
@@
 
+- Cluster 3:
+
+- Cluster 4:
+
+- Cluster 5:
+
+- Cluster 6:
+
 ...

### 2021-04-19 src/online_appendix.Rmd

@@
 
-- Cluster 1:
+- Cluster 1: Medium sized mostly rural cluster; slow mean upload speeds; high workday temporal variation.
 
-- Cluster 2:
+- Cluster 2: Small rural cluster; highest mean upload speeds; slowdown more in evening.
 
-- Cluster 3:
+- Cluster 3: Medium sized mixed cluster; fast mean upload speeds; medium temporal variation.
 
-- Cluster 4:
+- Cluster 4: Rural reference LAD; medium mean upload speeds; high temporal variation.
 
-- Cluster 5:
+- Cluster 5: Small suburban cluster; fast mean upload speeds; slowdown more in evening.
 
-- Cluster 6:
+- Cluster 6: Large mixed cluster; slowest aggregate mean upload speeds; low workday temporal variation.
+
+- Cluster 7: Small suburban cluster; fast mean upload speeds; medium temporal variation.
+
+- Cluster 8: Small mainly surburban cluster; slow mean upload speeds; high workday temporal variation.
+
+- Cluster 9: Large, more urban cluster; fast mean upload speeds; lowest workday temporal variation.
 

### 2021-04-19 paper/v2_taylor_francis/broadband.speed.covid.Rmd
submission

@@
 
-<!-- # Acknowledgement(s) {-} -->
-
-<!-- An unnumbered section, e.g.\ `\section*{Acknowledgements}`, may be used for thanks, etc.\ if required and included _in the non-anonymous version_ before any Notes or References. -->
+# Supplemental Material {-}
+
+The LAD upload clusters as well as the mean AM and PM upload speeds are presented 
+in the Supplemental Material. Download the .html file and open it using your internet
+browser. Hovering over the LAD polygons provides the mean internet speeds.
+
+# Acknowledgement {-}
+
+The authors would like to thank [https://www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)
+for providing the main data used for this paper.
 

### 2021-04-19 paper/v2_taylor_francis/broadband.speed.covid.tex
submission

@@
 
-\title{Working from home and digital divides: resilience during the pandemic}
+\title{Working from home and digital divides: resilience during the
+pandemic}
 
 \author{\name{Hannah Budnitz$^{a}$, Emmanouil Tranos$^{b, c, \dagger}$}
-\affil{$^{a}$Transport Studies Unit, University of Oxford, UK; $^{b}$School of Geographic Sciences, University of Bristol, UK; $^{c}$The Alan Turing Institute, UK; $^{\dagger}$Corresponding author}
+\affil{$^{a}$Transport Studies Unit, University of Oxford,
+UK; $^{b}$School of Geographic Sciences, University of Bristol,
+UK; $^{c}$The Alan Turing Institute, UK; $^{\dagger}$Corresponding
+author}
 }
 
-\thanks{CONTACT Hannah Budnitz.
-Email: \href{mailto:hannah.budnitz@ouce.ox.ac.uk}{\nolinkurl{hannah.budnitz@ouce.ox.ac.uk}}, Emmanouil Tranos.
-Email: \href{mailto:e.tranos@bristol.ac.uk}{\nolinkurl{e.tranos@bristol.ac.uk}}}
+\thanks{CONTACT Hannah
+Budnitz. Email: \href{mailto:hannah.budnitz@ouce.ox.ac.uk}{\nolinkurl{hannah.budnitz@ouce.ox.ac.uk}}, Emmanouil
+Tranos. Email: \href{mailto:e.tranos@bristol.ac.uk}{\nolinkurl{e.tranos@bristol.ac.uk}}}
 
@@
 bridging or further enhancing existing socioeconomic divides.
-\citep{stern2009levels, van2014digital, van2015third}. The capability to
-telecommute is related to all three levels of digital divides, but more
-importantly leads to differentiated outcomes regarding the economic
-resilience of people and places to overcome a systemic shock such as the
-current pandemic.
+\citep{stern2009levels, van2014digital, van2015third}. The potential for
+these different levels of digital divide to interact is growing in
+parallel with the digitisation of various sectors of the economy and
+society. For example, \citet{li2020challenges} highlights how the
+intersection of new demands upon ICT infrastructure and a paucity of end
+users means that some areas are being left behind in the roll-out of
+smart applications and services. They argue that this could create a
+geospatial `smart divide', with areas unable to capitalise on the
+opportunities offered by such applications and services. Likewise, the
+capability to telecommute is related to all three levels of digital
+divides, but more importantly leads to differentiated outcomes regarding
+the economic resilience of people and places to overcome a systemic
+shock such as the current pandemic.
 
@@
 However, the success of these arrangements has been dependent upon the
-first level digital divide, which is associated with access to and
-quality of internet connectivity. \citet{SALEMINK2017360} provides a
-systematic review of the pre-pandemic, first level digital divide in
-infrastructure quality between urban and rural areas in various advanced
-economies. Rural areas, predictably, fare worse. This finding broadly
-applies in the UK, where studies that also analysed broadband speed
-checks concluded that average speeds are lower in rural areas, something
-that has not been improved by policy measures to increase competition
+first level digital divide, which is associated with geospatial access
+to and quality of internet connectivity. \citet{SALEMINK2017360}
+provides a systematic review of the pre-pandemic, first level digital
+divide in infrastructure quality between urban and rural areas in
+various advanced economies. Rural areas, predictably, fare worse. This
+finding broadly applies in the UK, where studies that also analysed
+broadband speed checks concluded that average speeds are lower in rural
+areas, something that has not been improved by policy measures to
+increase competition
 \citep{riddlesden2014broadband, nardotto2015unbundling}. In contrast,
 \citet{riddlesden2014broadband} found that levels of deprivation did not
-correlate with first level digital divides.
+correlate with first level digital divides, perhaps because many
+deprived areas benefit from spillover effects of quality installed
+infrastructure nearby.
 
 However, as \citet{blank2018local} highlight, variation in individual
-internet availability and uptake is a product of more complex spatial
-and demographic characteristics than simple rurality or urbanisation.
-Dense urban areas were shown to suffer more from slowdown during peak
-hours, although these services were more likely to be improved by
-increased competition between providers, such as between new entrants
-and Virgin Media cable connections
+internet availability and uptake is a product of more complex geospatial
+and demographic characteristics than simple rurality, urbanisation or
+levels of deprivation. Dense urban areas were shown to suffer more from
+slowdown during peak hours, although these services were more likely to
+be improved by increased competition between providers, such as between
+new entrants and Virgin Media cable connections
 \citep{riddlesden2014broadband, nardotto2015unbundling}. The latter were
@@
 \citep{ofcom2016}, where the more lucrative and competitive market
-originally attracted the cable TV provider. Whether the variation in
-infrastructure quality and reliability affects the spatial footprint of
-telecommuting has not previously been investigated, in part because
-telecommuting has not previously been a cause of concentrated demand and
-pressure on internet services.
+originally attracted the cable TV provider. Whether the geospatial
+variation in infrastructure quality and reliability affects the spatial
+footprint of telecommuting has not previously been investigated, in part
+because telecommuting has not previously been a cause of concentrated
+demand and pressure on internet services.
 
@@
 observed in the space-time geography of cities around the world in the
-last year \citep[e.g.~][]{google2020, shibayama2021impact}. Yet,
-although there is a broad agreement that these changes during the
-pandemic have played a pivotal role in stopping the spread of the virus
+last year \citep[e.g.][]{google2020, shibayama2021impact}. Yet, although
+there is a broad agreement that these changes during the pandemic have
+played a pivotal role in stopping the spread of the virus
 \citep{jia2020population, yang2020taking, mu2020interplay}, the extent
@@
 capability to work from home. We also aim to improve our understanding
-of the impact of first level digital divisions on telecommuting, and
-whether this results in more fundamental third level digital divisions
-than has previously been perceived.
+of the impact of first level digital divisions on telecommuting, in
+terms of both geospatial and temporal quality of service, and whether
+this results in more fundamental third level digital divisions between
+local geographies than has previously been perceived.
 
@@
 aimed at supporting small and medium enterprises
-\citep{HENDERSON2020102024}. SUch policies were based on previous
+\citep{HENDERSON2020102024}. Such policies were based on previous
 research regarding the productivity effects of broadband infrastructure
@@
 resilience is so vital.
+
+\hypertarget{supplemental-material}{%
+\section*{Supplemental Material}\label{supplemental-material}}
+\addcontentsline{toc}{section}{Supplemental Material}
+
+The LAD upload clusters as well as the mean AM and PM upload speeds are
+presented in the Supplemental Material. Download the .html file and open
+it using your internet browser. Hovering over the LAD polygons provides
+the mean internet speeds.
+
+\hypertarget{acknowledgement}{%
+\section*{Acknowledgement}\label{acknowledgement}}
+\addcontentsline{toc}{section}{Acknowledgement}
+
+The authors would like to thank
+\url{https://www.broadbandspeedchecker.co.uk/} for providing the main
+data used for this paper.
 

### 2021-04-19 src/online_appendix.Rmd
submission

@@
 
-### Description
-
-The below map illustrates the different clusters of LAD based on the upload 
-experienced internet speeds. Hovering over the LAD polygons, the map also presents
-the mean AM upload internet speed and the relevant ranking in the country for each 
-LAD.
-
-**Is there a chance to briefly describe the clusters with 1-2 words?**
+### Cluster description
 
@@
 - Cluster 9: Large, more urban cluster; fast mean upload speeds; lowest workday temporal variation.
-
-...

### 2021-05-11 paper/v2_taylor_francis_for_proof/broadband.speed.covid_for_tables.Rmd
word version for final submission

--- 2021-04-19 paper/v2_taylor_francis/broadband.speed.covid.Rmd
@@
   \def\tightlist{}
+  \usepackage{setspace}
+  \doublespacing
 output: rticles::tf_article
@@
 
+path.out = paste0(path, "/paper/v2_taylor_francis_for_proof/tables/table_1.txt")
+
 stargazer(clustersLADup[1:7],
           summary = F,
+          type = "text",
+          out = path.out,
           font.size = "footnotesize",
-          type = "latex",
           header = F,
@@
 
+path.out = paste0(path, "/paper/v2_taylor_francis_for_proof/tables/table_2.txt")
+
 stargazer(aux.up,
-          type = "latex",
+          type = "text",
+          out = path.out,
           nobs = T,
@@
 
+path.out = paste0(path, "/paper/v2_taylor_francis_for_proof/tables/table_A1.txt")
+
 stargazer(for.descr,
           summary = T,
+          out = path.out,
           font.size = "footnotesize",
-          type = "latex",
+          type = "text",
           header = F,

### 2021-05-11 src/online_appendix.Rmd
word version for final submission

@@
 ---
-title: "Online Appendix"
+title: "Supplemental Material"
 output: 

### 2021-05-11 paper/v2_taylor_francis_for_proof/render_word_bw_images.R

@@ -1,0 +1,24 @@
+library(rmarkdown)
+library(magick)
+
+# render to .Rmd to word
+path <- "C:/Users/nw19521/OneDrive - University of Bristol/projects/broadband.speed.covid/paper/v2_taylor_francis_for_proof/broadband.speed.covid.Rmd"
+render(path, word_document())
+
+# convert Fig 1 to bw
+path <- "C:/Users/nw19521/OneDrive - University of Bristol/projects/broadband.speed.covid/paper/v2_taylor_francis_for_proof/figures_for_word_submission"
+
+fig1.path <- paste0(path, "/Fig1.png")
+
+fig1 <- image_read(fig1.path)
+fig1 <- fig1 %>% image_quantize(colorspace = 'gray')
+
+image_write(fig1, path = fig1.path, format = "png")
+
+# convert Fig 2 and 2 to bw
+fig2.path <- paste0(path, "/Fig2.png")
+
+fig2 <- image_read(fig1.path)
+fig2 <- fig2 %>% image_quantize(colorspace = 'gray')
+
+image_write(fig2, path = fig2.path, format = "png")

### 2021-05-11 paper/v2_taylor_francis_for_proof/tables/table_1.tex
typos and .tex files for tables

--- 2021-03-25 paper/v2_taylor_francis/broadband.speed.covid.tex
@@
-% interactcadsample.tex
-% v1.03 - April 2017
-
-\documentclass[]{interact}
-
-\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
-\usepackage{subfigure}% Support for small, `sub' figures and tables
-%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
-
-\usepackage{natbib}% Citation support using natbib.sty
-\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
-\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
-
-\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
-\newtheorem{theorem}{Theorem}[section]
-\newtheorem{lemma}[theorem]{Lemma}
-\newtheorem{corollary}[theorem]{Corollary}
-\newtheorem{proposition}[theorem]{Proposition}
-
-\theoremstyle{definition}
-\newtheorem{definition}[theorem]{Definition}
-\newtheorem{example}[theorem]{Example}
-
-\theoremstyle{remark}
-\newtheorem{remark}{Remark}
-\newtheorem{notation}{Notation}
-
-% see https://stackoverflow.com/a/47122900
-
-\usepackage{hyperref}
-\usepackage[utf8]{inputenc}
-\usepackage{dcolumn}
-\def\tightlist{}
-
-\begin{document}
-
-\articletype{}
-
-\title{Working from home and digital divides: resilience during the pandemic}
-
-\author{\name{Hannah Budnitz$^{a}$, Emmanouil Tranos$^{b, c}$}
-\affil{$^{a}$Transport Studies Unit, University of Oxford, UK; $^{b}$School of Geographic Sciences, University of Bristol, UK; $^{c}$The Alan Turing Institute, UK}
-}
-
-\thanks{CONTACT Hannah Budnitz. Email: , Emmanouil Tranos. Email: \href{mailto:e.tranos@bristol.ac.uk}{\nolinkurl{e.tranos@bristol.ac.uk}}}
-
-\maketitle
-
-\begin{abstract}
-This paper offers a new perspective on telecommuting from the viewpoint
-of the complex web of digital divides. Using the UK as a case study,
-this paper studies how the quality and reliability of internet services,
-as reflected in \emph{experienced} internet upload speeds during the
-spring 2020 lockdown, may reinforce or redress the spatial and social
-dimensions of digital divisions. Fast, reliable internet connections are
-necessary for the population to be able to work from home. Although not
-every place hosts individuals in occupations which allow for
-telecommuting nor with the necessary skills to effectively use the
-internet to telecommute, good internet connectivity is also essential to
-local economic resilience in a period like the current pandemic.
-Employing data on individual broadband speed tests and state-of-the-art
-time-series clustering methods, we create clusters of UK local
-authorities with similar temporal signatures of experienced upload
-speeds. We then associate these clusters of local authorities with their
-socioeconomic and geographic characteristics to explore how they overlap
-with or diverge from the existing economic and digital geography of the
-UK. Our analysis enables us to better understand how the spatial and
-social distribution of both occupations and online accessibility
-intersect to enable or hinder the practice of telecommuting at a time of
-extreme demand.
-\end{abstract}
-
-\begin{keywords}
-covid; internet; working from home; broadband speed; time-series
-clusters
-\end{keywords}
-
-\hypertarget{sec:1}{%
-\section{Introduction}\label{sec:1}}
-
-During the pandemic, working from home using digital technologies,
-whether partially or exclusively, was transformed from a niche means of
-accessing work, albeit one that had been on a slow, upward trend, to a
-widespread way of life in many countries. The ability to work from home
-or telecommute meant millions retained their jobs and, to a varying
-extent, maintained productivity during periods of strict lockdowns
-around the world. However, this ability has not been evenly distributed
-socially or spatially, creating new intersections of economic and
-digital divisions. On one side are those who can work from home,
-supported by digital technologies, and have thus been able to enjoy both
-economic resilience and greater personal safety. On the other side,
-previously employed individuals have been forced to accept furlough or
-redundancy packages unless they are part of the cadre of essential
-workers, who are potentially at high risk of infection. Whilst the basis
-for this pandemic-generated divide has been viewed as mainly
-occupational, here we consider whether it is also technological and
-geographical.
-
-The capability to work from home has previously been studied from the
-perspective of whether work tasks in a given occupation both can be and
-are allowed to be performed using digital technologies independently of
-location or co-location with colleagues, including supervisors
-\citep{allen2015effective, singh2013modeling}. However, successful
-telecommuting also requires that the quality and reliability of digital
-services, particularly home internet connection speeds, enable the
-completion of work tasks with a minimum of delay or interruption. High
-levels of demand are one of the main causes of reduced reliability and
-slower connection speeds, as network bandwidth becomes congested. Yet
-prior to the pandemic, the performance of broadband services with
-respect to telecommuters was never tested at scale, as working from home
-and connecting to colleagues and workplace resources via the internet
-was the purview of a small minority of workers. Instead, leisure use in
-the evening, when video streaming services are at their peak, has been
-used to benchmark broadband performance and service delivery by
-different Internet Service Providers (ISPs) \citep{ofcom2017}.
-
-The shift towards telecommuting during various stages of lockdown around
-the world has been drastic and there are speculations that post-Covid,
-the tendency to work from home will be much higher, raising questions
-around whether internet services can accommodate the increased demand.
-For example, \(47\)\% of people in employment in the UK worked solely
-from home in April \(2020\), whilst the same figure only reached \(5\)\%
-the year before \citep{ons2020, ons2020lm2019}. Similar figures have
-been reported for other countries \citep{felstead2020homeworking}.
-Approximately \(37\)\% of the European workforce worked from home in
-April \(2020\) with countries like Finland reaching \(60\)\%
-\citep{eurofound2020}. In the US, almost half of the working population
-worked from home during the same period because of the pandemic
-\citep{brynjolfsson2020covid}, and a recent estimate indicated that
-\(37\)\% of all jobs in the US can be permanently performed entirely
-from home \citep{NBERw26948}. None of these changes could have happened
-in the absence of reliable information and communication technology
-(ICT) infrastructure -- both in terms of software and hardware. But
-while software innovations are easily diffused across space and
-society\footnote{See for example the huge success of videoconferencing
-  apps such as Zoom \citep{marks2020zoom}.}, the same does not apply for
-ICT hardware infrastructure such as internet broadband connectivity.
-
-Using the UK as a case study, this paper aims to understand how the
-quality and reliability of internet service, as reflected in
-\emph{experienced} internet speeds during the spring 2020 lockdown, may
-reinforce or redress the spatial and social dimensions of digital
-divisions. We employ volunteered geographic data on individual broadband
-speed tests and state-of-the-art time-series clustering methods to
-create clusters of UK local authorities with similar temporal signatures
-of experienced internet speeds. We then associate these clusters of
-local authorities with their socioeconomic and geographic
-characteristics to explore how they overlap with or diverge from the
-existing economic and digital geography of the UK. As the quality of
-internet infrastructure and services, as well as the concentration of
-different occupations are spatially dependent and clustered in space,
-our approach offers a framework for understanding the impact of and
-interactions between digital divisions geographically and
-socioeconomically. Our analysis enables us to better understand how the
-spatial and social distribution of occupation and online accessibility
-intersect to enable or hinder the practice of telecommuting at a time of
-extreme demand. We also consider what lessons can be learned from this
-time for a future where telecommuting is likely to remain a more common
-means of accessing work, at least in comparison to the pre-Covid era.
-
-By asking how resilient broadband speeds, and particularly upload
-speeds, are as experienced in different parts of the UK during a time of
-extreme demand, we interrogate which places benefit from the greater
-economic resilience digital technologies can offer. The structure of
-this paper is as follows. First we review the literature on
-telecommuting and digital divides to better understand their structural
-and spatial development pre-pandemic, and thus their importance to the
-economic resilience of different places. We then describe our data and
-methodology. Our results section first offers a classification of how
-internet services vary across clusters of UK local authorities and then
-assesses whether these clusters replicate or repudiate other
-socio-economic and geographic patterns of economic resilience. We
-conclude with a discussion of the insights we have gained from our new
-perspective on digital divisions.
-
-\hypertarget{sec:2}{%
-\section{Literature review}\label{sec:2}}
-
-\hypertarget{sec:2.1}{%
-\subsection{From telecommuting to \#WFH}\label{sec:2.1}}
-
-In this analysis, the terms `telecommuting' and `working from
-home\footnote{See also the popular social media hashtag \#WFM}' are used
-interchangeably, as most remote labour during the Covid-19 crisis was
-carried out in the homes of individual employees rather than any other
-location \citep{eurofound2020}. However, previous research has explored
-how telecommuting can occur in other places, including satellite offices
-or on public transport \citep{felstead2012rapid, siha2006telecommuting}.
-Previous research has also used a variety of definitions to measure the
-level of telecommuting within different workforces, distinguishing
-between those directly employed, indirectly employed, self-employed,
-full-time or part-time, and those who use digital technologies to work
-remotely full-days or part-days
-\citep{allen2015effective, bailey2002review, haddad2009examination}. No
-matter the definition, the option and capability to telecommute or work
-from home has never been equally distributed spatially or
-socio-economically any more than different industries and employment
-opportunities have. Studies from the United States, the Netherlands, and
-the UK found that telecommuters are most likely to hold professional,
-managerial, and technical occupations where the workforce is better
-educated and wealthier, and that there is suppressed demand among women
-and part-time workers
-\citep{headicar2016move, peters2004employees, singh2013modeling}.
-
-Opportunities for working from home during the current pandemic have
-likewise not been equally spread across the workforce.
-\citet{NBERw26948} indicated that in the US, managers, educators, those
-working in computer-related occupations, finance, and law can easily
-work from home, and that occupations with opportunities to telecommute
-are associated with higher earnings. This is not the case for the
-workforce occupied in more spatially fixed occupations, from farming,
-construction and manufacturing to hospitality and care services. In the
-US, these occupations tend to be lower-income, non-white, without a
-university degree, live in rental accommodation and lack health
-insurance \citep{NBERw27085}. Similar trends can be observed for other
-countries. For example, \(75\)\% of workers with tertiary education
-worked from home in Europe during spring \(2020\), whilst only \(34\)\%
-of workers with secondary education and \(14\)\% of those primary
-education did so \citep{eurofound2020}.
-
-\hypertarget{sec:2.2}{%
-\subsection{Digital divides and economic resilience}\label{sec:2.2}}
-
-The literature describes first level digital divides in terms of the
-availability and quality of internet connectivity, such as that manifest
-in different geographies in the UK
-\citep{riddlesden2014broadband, philip2017digital}. Second level digital
-divides consider the presence or lack of the necessary skills to
-effectively utilise digital technologies and the internet
-\citep{blank2014dimensions, van2011internet}. The third level focuses on
-the heterogenous returns of internet usage among different socioeconomic
-groups and, consequently, how digital technologies can assist in
-bridging or further enhancing existing socioeconomic divides.
-\citep{stern2009levels, van2014digital, van2015third}. The capability to
-telecommute is related to all three levels of digital divides, but more
-importantly leads to differentiated outcomes regarding the economic
-resilience of people and places to overcome a systemic shock such as the
-current pandemic.
-
-Our understanding of telecommuting as a product of enabled occupations
-can be described as a manifestation of the third level digital divide,
-as those who are able to use digital technologies to work from home
-benefit from a high rate of return on their use of the internet in terms
-of autonomy, flexibility, and time saved from commuting
-\citep{peters2004employees, siha2006telecommuting, singh2013modeling}.
-In some aspects, these returns may have increased during the Covid-19
-crisis, when those with the capability to telecommute also have the
-ability to maintain their employment whilst protecting their health.
-However, the success of these arrangements has been dependent upon the
-first level digital divide, which is associated with access to and
-quality of internet connectivity. \citet{SALEMINK2017360} provides a
-systematic review of the pre-pandemic, first level digital divide in
-infrastructure quality between urban and rural areas in various advanced
-economies. Rural areas, predictably, fare worse. This finding broadly
-applies in the UK, where studies that also analysed broadband speed
-checks concluded that average speeds are lower in rural areas, something
-that has not been improved by policy measures to increase competition
-\citep{riddlesden2014broadband, nardotto2015unbundling}. In contrast,
-\citet{riddlesden2014broadband} found that levels of deprivation did not
-correlate with first level digital divides.
-
-However, as \citet{blank2018local} highlight, variation in individual
-internet availability and uptake is a product of more complex spatial
-and demographic characteristics than simple rurality or urbanisation.
-Dense urban areas were shown to suffer more from slowdown during peak
-hours, although these services were more likely to be improved by
-increased competition between providers, such as between new entrants
-and Virgin Media cable connections
-\citep{riddlesden2014broadband, nardotto2015unbundling}. The latter were
-historically available to only \(45\)\% of premises in the UK
-\citep{ofcom2016}, where the more lucrative and competitive market
-originally attracted the cable TV provider. Whether the variation in
-infrastructure quality and reliability affects the spatial footprint of
-telecommuting has not previously been investigated, in part because
-telecommuting has not previously been a cause of concentrated demand and
-pressure on internet services.
-
-Other intersections between digital divides have been subject to study.
-There are indications that those who purchase high speed connections
-consume more data of all sorts and use their connections for a greater
-variety of purposes \citep{hauge2011consumer}. There is also a
-correlation between access to internet services and a reduction in
-household transport spend \citep{bris2017ict}. These studies suggest
-that better internet services enable households to make savings and
-efficiencies, an example of the first level digital divide reinforcing
-the third level. Such digital intersections and efficiencies have also
-changed transportation, connectivity and the structure of cities during
-the pandemic. We now have enough data to confirm the drastic alteration
-observed in the space-time geography of cities around the world in the
-last year \citep[e.g.~][]{google2020, shibayama2021impact}. Yet,
-although there is a broad agreement that these changes during the
-pandemic have played a pivotal role in stopping the spread of the virus
-\citep{jia2020population, yang2020taking, mu2020interplay}, the extent
-to which the increased levels of working from home, the consequent
-decrease in commuting flows, and the altered structure of cities will
-remain post-pandemic is the subject of considerable debate in the
-literature.
-
-Meanwhile, multi-layered digital divides may also intersect with
-material divides and the economic geography of the UK. The regional
-economic resilience literature underlines the differentiated capacity of
-cities and regions to escape or recover from economic crises
-\citep{martin2012regional, kitsos2018economic}. As different places have
-different industrial and occupational profiles, these affect their
-aggregated potential capacity for telecommuting. Such profiles are
-associated with longstanding inequalities in the UK and their spatial
-representation as a North-South divide \citep{martin_north_south}.
-Various studies have illustrated severe inequalities between the north
-and the south regions of England in terms of skills and human capital,
-unemployment, productivity and prosperity
-\citep{lee2014grim, mccann2020perceptions, dorling2018peak}. Some
-scholars have even argued that the UK suffers from some of the highest
-levels of interregional inequalities in the global north
-\citep{gal2018reducing, mccann2016uk}. All three levels of digital
-divides are associated with or shaped by the geography of the UK. Yet
-this is the first time that the intersection of digital and material
-divides is relevant to understanding the economic resilience of places
-and large swathes of the population, as digital technologies became an
-essential tool of productivity during the Covid-19 pandemic.
-
-The extreme demand during the pandemic thus provides a new opportunity
-to understand how internet infrastructure quality, and reliability
-affects telecommuting, particularly in light of the high volumes of
-bandwidth-intensive video conferencing required in order to avoid the
-face-to-face contact that could increase the spread of infection. We
-seek to answer how internet service resilience could contribute to or
-reduce economic resilience when the latter is dependent upon the
-capability to work from home. We also aim to improve our understanding
-of the impact of first level digital divisions on telecommuting, and
-whether this results in more fundamental third level digital divisions
-than has previously been perceived.
-
-\hypertarget{sec:3}{%
-\section{Methods and data}\label{sec:3}}
-
-\hypertarget{sec:3.1}{%
-\subsection{Time-Series clustering}\label{sec:3.1}}
-
-Our chosen methodological framework is cluster analysis, which can be
-defined within machine learning approaches as an unsupervised learning
-task, partitioning unlabelled observations into homogeneous groups known
-as clusters \citep{montero2014tsclust}. The key idea is that
-observations within clusters tend to be more similar than observations
-between clusters. Clustering is particularly useful for exploratory
-studies as it identifies structures within the data
-\citep{aghabozorgi2015time}. Cluster analysis is a widely used in
-geography \citep{gordon1977classification, everitt1974cluster}, for
-example to solve \emph{regionalisation} problems
-\citep{niesterowicz2016}. Clustering methods are also the basis of
-\emph{geodemographics}, a research domain which aims to create small
-area indicators or typologies of neighbourhoods based on diverse
-variables \citep{SINGLETON2009289, harris2005geodemographics}. These
-studies usually employ cross-sectional data, and most clustering
-problems in geography deal with observations that are fixed in time.
-However, for this paper we are interested in internet speeds, which vary
-over time. Therefore, we create clusters of local authorities in the UK
-with similar temporal signatures of experienced internet speeds.
-
-To do so, we employ time-series clustering methods, which have been
-developed to deal with clustering problems linked to, for instance,
-stock or other financial data, economic, governmental or medical data as
-well as machine monitoring
-\citep{aggarwal2013time, aggarwal2001surprising, hyndman2015large, WARRENLIAO20051857}.
-The main challenge, which does not apply to cross-sectional clustering
-problems, is data dimensionality, with a multiplicity of data points for
-every individual object included in the data set, and how their value
-changes dynamically as a function of time \citep{aghabozorgi2015time}.
-This high dimensionality leads to (i) computational and algorithmic
-challenges regarding handling these data and building algorithms to
-perform clustering over long time-series, and (ii) open questions
-regarding the choice of similarity measures in order to cluster similar
-times-series objects together considering the whole length of the
-time-series and overcoming issues around noise, outliers and shifts
-\citep{lin2004iterative, aghabozorgi2015time}.
-
-For this paper we utilise a category of time-series clustering methods
-known as shape-based approaches. These methods match two separate
-time-series objects based on the similarity of their shapes through the
-calculation of distances between the shapes, and are better equipped to
-capture similarities between short length time-series
-\citep{aghabozorgi2015time}, such as our data. We thus identify clusters
-of UK local authorities with similar temporal signatures -- i.e.~shapes
--- of experienced internet speeds. The clusters are identified using the
-common partitioning algorithm, where no clusters overlap, known as
-\emph{k}-means. This iterative algorithm is popular because of the
-simplicity of the implementation and the interpretability of the
-results. It begins with defining the desired number of clusters:
-\emph{k}. Then each observation is randomly assigned to a cluster from
-the \([1,k]\) space. This initial cluster assignment is followed by
-iterations in order to minimise the distance between the centroids of
-the clusters and the observations assigned to these clusters
-\citep{james2013introduction}.
-
-There are a number of differences between the application of
-\emph{k}-means for cross-sectional and times-series data. Instead of
-creating clusters around centroids, a common approach is to create
-clusters around \emph{medoids}, which are representative time-series
-objects with a minimal distance to all other cluster objects
-\citep{sardatime}. Also, instead of calculating the Euclidean distance
-between centroids and data points, more complex distance measures need
-to be employed to capture the similarity between a time-series object
-and a medoid. A common distance measure for shape-based time-series
-clustering is Dynamic Time Warping (DTW), an algorithm comparing two
-time-series objects to find the optimum warping path between them. DTW
-is widely used in order to overcome limitations linked to the use of
-Euclidean distance
-\citep{sardatime, berndt1994using, ratanamahatana2004everything}. The
-\texttt{R} package \texttt{dtwclust} has been used for the time-series
-clustering \citep{dtwclust}.
-
-\hypertarget{sec:3.2}{%
-\subsection{Experienced Broadband Speeds}\label{sec:3.2}}
-
-To assess the internet quality and reliability across local authorities
-in the UK, we utilise unique data comprising individual internet speed
-tests from Speedchecker Ltd\footnote{\url{https://www.broadbandspeedchecker.co.uk/}}.
-This is a private company that allows internet users to check their own
-broadband upload and download speeds, and stores every speed-check with
-timestamp and geolocation information. These data have been used before
-to assess digital divides \citep{riddlesden2014broadband} and the impact
-of local loop unbundling regulatory processes
-\citep{nardotto2015unbundling}, and we followed the former's approach to
-remove outliers. By using this volunteered geographic data, we are able
-to assess the internet speed \emph{experienced} by users, which may
-differ from the maximum speeds \emph{advertised} by ISPs. Indeed, those
-who seek to test their broadband are most likely to do so because they
-are experiencing slower speeds than expected, although any skew towards
-slower speeds is balanced by the likelihood that those who test their
-broadband are also more `tech-savvy' and / or have purchased higher
-speed packages that are not delivering the promised level of service.
-
-We are particularly interested in upload speeds and the frequency of
-speed tests over the period from March to May \(2020\), as government
-statements indicate this encompasses the period when UK workers were
-told to work from home if at all possible \citep{GovUK2020}. Schools and
-various retail, leisure and hospitality businesses were closed from late
-March, and restrictions were gradually eased from late May. Average
-upload speeds are slower than average download speeds, at \(9.3\)Mb/s
-mean upload speed for the whole sample, compared to \(29.6\)Mb/s for
-download speeds, but they are also less associated with internet-based,
-high-demand, leisure activities such as video streaming. Therefore,
-upload speeds are more relevant to work-related activities such as
-uploading documents or two-way audio, video, and text-based
-communication systems.
-
-The frequency of speed tests was important in identifying the temporal
-profile which would give us most insight into experienced internet
-service and resilience over units of time. Whilst there is an overall
-trend of increased testing from March to April and then a slight
-reduction from April to May, this trend masks substantial variation by
-not only the day of the week, but also time of day, as can be seen in
-Figure \ref{test2020} \footnote{Figures \ref{test2020} and
-  \ref{test2019} were created with the \texttt{R} package
-  \texttt{openair} \citep{openair}.}. Thus, a daily aggregation of
-upload speeds would mask the variation in experienced service over the
-course of each weekday. Furthermore, the importance of this variation is
-highlighted by a comparison with the same period in \(2019\), as in
-Figure \ref{test2019}, when the volume of testing and thus of
-experienced internet service quality peaked in the evening, presumably
-in response to demand for leisure activities and download speeds. In
-contrast, the majority of the increase in testing in \(2020\) is during
-the working day, creating a new morning peak in Figure \ref{test2020}.
-Therefore, we include a measure of hourly variation in our temporal
-profiles to reflect the change in users' perception of the workday
-reliability of internet services.
-
-However, there were insufficient observations -- only \(631\) speed
-tests per Local Authority District (LAD) on average -- for each for each
-working hour of each working day in each LAD to profile speeds at that
-level of detail. Spatial aggregation was also necessary because we could
-not follow individuals or households and connect data points. Therefore,
-we aggregate the \(241,088\) individual, geolocated and time-stamped
-speed-checks during the \(13\) weeks of March to May inclusive for
-weekdays in \(2020\) by each hour of the day and day of the week. As our
-research aims to identify the geography of internet service resilience
-for work purposes, bank holidays and the hours between midnight and
-\(6:00\) were excluded, as well as weekend days. The composite week
-time-series thus comprise \(18\) hours multiplied by \(5\) weekdays or
-\(90\) time points per series. The time-series were calculated for each
-of the \(382\) LADs in the UK, standardised, and then a \emph{k}-means
-partitioning around medoids clustering algorithm was applied using DTW.
-We initially run the algorithm for \(k \in \mathbb{N} \bigcap [5,15]\),
-calculated the cluster validity indices (CVIs) and then run the
-subsequent multinomial regression -- see the end of this section for
-more details for the different \emph{k}. Following \citet{sardatime}, to
-identify the optimal \emph{k} we used the majority vote for the
-following CVIs: Silhouette (max), Score function (max),
-Calinski-Harabasz (max), Davies-Bouldin (min), Modified Davies-Bouldin
-(DB*, min), Dunn (max), COP (min). Nevertheless, we opted against using
-the optimal \(k = 13\) solution as it was too large to allow for
-communicable LAD clusters. Instead we opted for a smaller \(k = 9\),
-which led to a rather similar spatial pattern and, importantly, to a
-much higher R-squared in the subsequent explanatory regression (\(0.44\)
-instead of \(0.34\)).
-
-\begin{figure}
-\centering
-\includegraphics[width=0.85\textwidth,height=0.5\textheight]{figures/time.var.plot2020.png}
-\caption{Speed tests over time, 2020 \label{test2020}}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[width=0.85\textwidth,height=0.5\textheight]{figures/time.var.plot2019.png}
-\caption{Speed tests over time, 2019 \label{test2019}}
-\end{figure}
-
-In Section \protect\hyperlink{sec:4.1}{4.1}, we review the temporal
-profile of upload speed by hour of the day and day of the composite week
-for each of the clusters. Since the quality and reliability of internet
-services vary in time and space due to both supply and demand-side
-influences, we also use a number of different measures to describe
-experienced upload speeds per cluster. These include: i) mean,
-experienced connection speed, ii) standard deviation or the amount of
-fluctuation from the mean, and iii) the variation in speeds during the
-new morning peak of testing when working from home is more likely to
-take place. We take account of all three measurements in order to
-determine how resilient broadband speeds are as experienced in different
-parts of the UK during a time of extreme demand.
-
-The cause of these different experiences of broadband resilience may
-vary between and within clusters, as they may reflect either patterns of
-demand or quality of infrastructure. Our approach is also limited by
-potential endogeneity, as for example, better quality connections with
-high mean speeds may enable more working from home, but greater demand
-can cause slower speeds, less reliability, or greater variability of
-speed at different times of day or week. Therefore, we avoid attributing
-any cause to our analysis of the experienced level of quality and
-reliability of upload speeds. Instead, we run an auxiliary regression to
-understand how the spatial and temporal patterns of internet service
-relate to the economic geography of the UK. More specifically, we
-estimate the following multinomial logit model:
-
-\begin{align}
-Pr(Y_{i}=j) = \frac{exp^{X_{i}\beta_{j}}}{\sum_{i=1}^j exp^{X_{i}\beta_{j}}}
-\begin{cases}
-    i = 1, 2, ... , N \\  
-    j = 1, 2, ... , J
-\end{cases}\label{eq1}
-\end{align}
-
-Based on the outcomes of the time-series clustering, we identify \(J\)
-distinct and crisp clusters. We then regress this cluster membership
-against a vector \(X_{i}\) of socio-economic and geographic variables,
-which are discussed in detail in the relevant Section
-\protect\hyperlink{sec:4.2}{4.2}. Because we cannot identify individuals
-or households and consequently aggregated our data at the LAD level, our
-results offer correlations between the socioeconomic characteristics of
-certain geographic locations and internet service quality, not a record
-of who was telecommuting. Such individual data could be found though
-surveys, but these offer less detailed information about the experience
-of internet resilience due to enforced demand, which is the main
-contribution of this paper. Our auxiliary regression, therefore,
-provides an indication of how internet connectivity can reinforce or
-redress existing spatial and social inequalities in different places.
-However, it opens a path to future research by highlighting the
-importance of understanding of how telecommuting capabilities and
-digital infrastructure divisions intersect.
-
-\hypertarget{sec:4}{%
-\section{Results}\label{sec:4}}
-
-\hypertarget{sec:4.1}{%
-\subsection{Upload Clusters / cluster description}\label{sec:4.1}}
-
-The temporal profiles of the local authority clusters have been
-summarised in Figure \ref{UpCluster} and Table \ref{up.cluster.descr},
-to provide an overview of the quality and reliability of experienced
-broadband in different parts of the UK. Figure \ref{UpCluster} shows a
-composite profile of mean upload speeds per hour per day for each of the
-largest six clusters, in terms of the LAD membership and population.
-
-\begin{figure}
-\includegraphics[width=0.95\linewidth]{figures/UpCluster} \caption{\label{UpCluster}Temporal profilies for upload speed clusters}\label{fig:unnamed-chunk-2}
-\end{figure}
-
 \begin{table}[!htbp] \centering 
@@
 \end{table}
-
-The largest cluster, comprising \(229\) local authorities and over
-\(40\) million people, is cluster \(6\), which has the slowest aggregate
-mean upload speed of any cluster, and the highest ratio of the standard
-deviation to the mean. This suggests that those living in local
-authorities in this cluster experienced some of the lowest quality
-broadband services in terms of upload speeds and reliability in the UK.
-However, as shown in Figure \ref{map.up.clusters}, some of the most
-rural areas of the UK are included in this cluster. If these areas
-suffer most from first level digital divides as described in the
-literature review, the low speeds in rural areas might be pulling down
-the averages in other areas in this largest cluster. Also, since the
-areas are clustered by their temporal profile across the working week,
-the graph in Figure \ref{UpCluster} indicates that speeds in cluster
-\(6\) are some of the more reliable. Table \ref{up.cluster.descr}
-confirms that upload speeds in cluster \(6\) during the morning peak
-from \(9:00\)-\(10:59\) were, on average, only \(4.5\)\% slower than in
-the evening peak period between \(19:00\) and \(20:59\), when
-entertainment purposes are likely to be using the most bandwidth.
-
-\begin{figure}
-\includegraphics[width=0.95\linewidth]{figures/map.up.clusters} \caption{\label{map.up.clusters}Upload speed clusters for LADs}\label{fig:unnamed-chunk-4}
-\end{figure}
-
-Those living in the second largest cluster -- \(9\), with \(115\) LADs
-and \(21.5\) million people -- experienced aggregate mean upload speeds
-of more than \(1.5\)Mb/s faster than those in cluster \(6\), a lower
-ratio of standard deviation to mean, and almost the same upload speeds
-in the AM Peak as the PM Peak. Indeed, the temporal profile for cluster
-\(9\) in Figure \ref{UpCluster}, like that for cluster \(6\), is fairly
-flat. Whilst it is likely that the large numbers of tests being
-performed by the large populations in these two clusters of LADs result
-in less varied averages to show in the graph, the method of shape-based
-clustering suggests that these results indicate that the majority of the
-population of the UK experience less temporal variation, and so belong
-in these two clusters. The results also confirm that those in cluster
-\(9\) experienced a consistently better service during lockdown than
-those in cluster \(6\), in terms of both average speed and reliability.
-
-Of the smaller clusters, cluster \(7\) experiences quite similar mean
-speeds to cluster \(9\), and a not too dissimilar standard deviation,
-but mean speeds in the morning peak are almost \(10\)\% lower than the
-evening peak. Cluster \(3\) boasts speeds almost \(1\)Mb/s higher again,
-but suffers from even greater slowdown in the morning peak. In
-comparison, clusters \(2\) and \(5\), home to a little less than half a
-million people, experience not only above average mean speeds, but also
-faster speeds in the morning peak compared to the evening. Thus, in the
-five LADs in these two clusters, the temporal profile of internet use
-may be closer to what might have been expected pre-pandemic. In
-comparison, the \(15\) LADs in clusters \(1\) and \(8\), home to over
-\(1.4\) million people, not only experience fairly average upload speeds
-and high ratios of standard deviation to the mean, but also experience
-much lower speeds during the morning peak than the evening peak. From
-the large spikes and dips shown on Figure \ref{UpCluster}, it is likely
-that this poor reliability or consistency of morning internet speeds was
-more noticeable to those in cluster \(1\), whilst those in cluster \(8\)
-experienced the lack of reliable service as a problem throughout the
-day.
-
-It is worth noting here that clusters of LAD based on the upload speeds
-are very different to the ones occur when we apply the same methods to
-upload speed data for the same time period in 2019 -- see Appendix
-\protect\hyperlink{appendix2}{2}. This drastic difference between the
-LAD upload speed profiles before and during the pandemic is an
-indication of the changes in internet usage that took place during the
-pandemic.
-
-In summary, the reliability of internet services during the working day
-appears to have altered for the vast majority of locations due to
-increased use from residents told to stay at home, and, if they were
-fortunate, work from home. Yet this experience has been different in
-different locations. In particular, LADs in cluster \(9\) experienced
-higher speeds and more resilient broadband internet than those in the
-largest cluster, \(6\). Those in clusters \(3\) and \(7\) also
-experience higher mean speeds and better service reliability, and can be
-more confident that they are on the right side of the first level
-digital divide, and that their more resilient ICT infrastructure and
-services can robustly support higher levels of telecommuting. On the
-other hand, whilst high levels of telecommuting may be the cause of
-morning speeds over 15\% lower than those in the evening, those in
-clusters \(1\) and \(8\) also experience lower mean speeds than clusters
-\(3\), \(7\), and \(9\), but not as low as cluster \(6\). Whilst the
-differences may not be large, the lower upload speeds in cluster \(6\)
-or the poor reliability in clusters \(1\) and \(8\) could still have had
-implications for economic resilience if combined with other types of
-digital divide. This potential will be explored in the next section.
-
-\hypertarget{sec:4.2}{%
-\subsection{Post-clustering regression analysis}\label{sec:4.2}}
-
-\begin{sidewaystable}[!htbp] \centering 
-  \caption{Auxiliary multinomial regression of upload speed clusters on socio-economic and geographic LAD variables\label{aux}} 
-  \label{} 
-\tiny 
-\begin{tabular}{@{\extracolsep{5pt}}lcccccccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-\\[-1.8ex] & 1 & 2 & 3 & 5 & 6 & 7 & 8 & 9 \\ 
-\\[-1.8ex] & (1) & (2) & (3) & (4) & (5) & (6) & (7) & (8)\\ 
-\hline \\[-1.8ex] 
- pop, 2018 & $-$0.00004$^{**}$ & 0.00002 & 0.00001 & 0.00002 & 0.00001 & 0.00002 & $-$0.00000 & 0.00001 \\ 
-  & (0.00002) & (0.00001) & (0.00001) & (0.00002) & (0.00001) & (0.00001) & (0.00002) & (0.00001) \\ 
-  & & & & & & & & \\ 
- job density, 2018 & $-$0.534$^{***}$ & $-$0.362$^{***}$ & 4.211$^{***}$ & $-$0.890$^{***}$ & 3.516$^{***}$ & $-$8.624$^{***}$ & $-$3.259$^{***}$ & 3.337$^{***}$ \\ 
-  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
-  & & & & & & & & \\ 
- distance to nearest met. area & $-$0.007$^{***}$ & 0.013$^{***}$ & 0.002$^{***}$ & 0.023$^{***}$ & $-$0.001 & $-$0.158$^{***}$ & 0.005$^{***}$ & $-$0.003$^{**}$ \\ 
-  & (0.0004) & (0.00001) & (0.0003) & (0.0002) & (0.002) & (0.0001) & (0.0003) & (0.002) \\ 
-  & & & & & & & & \\ 
- distance to London & 0.011$^{***}$ & 0.004$^{***}$ & 0.005$^{**}$ & $-$0.013$^{***}$ & 0.010$^{***}$ & 0.012$^{***}$ & 0.008$^{***}$ & 0.008$^{***}$ \\ 
-  & (0.001) & (0.00004) & (0.002) & (0.001) & (0.001) & (0.001) & (0.002) & (0.001) \\ 
-  & & & & & & & & \\ 
- South of the UK & $-$3.628$^{***}$ & 0.958$^{***}$ & 4.936$^{***}$ & $-$10.243$^{***}$ & 4.830$^{***}$ & 3.919$^{***}$ & $-$1.140$^{***}$ & 3.188$^{***}$ \\ 
-  & (0.000) & (0.000) & (0.00000) & (0.00000) & (0.00001) & (0.00000) & (0.00000) & (0.00001) \\ 
-  & & & & & & & & \\ 
- managerial jobs, 2020 & 0.944$^{***}$ & 0.628$^{***}$ & 0.593$^{***}$ & $-$0.544$^{***}$ & 0.634$^{***}$ & $-$0.128$^{***}$ & 0.725$^{***}$ & 0.511$^{***}$ \\ 
-  & (0.00002) & (0.00000) & (0.00003) & (0.00003) & (0.00002) & (0.00003) & (0.00003) & (0.00002) \\ 
-  & & & & & & & & \\ 
- tech jobs, 2020 & 0.350$^{***}$ & $-$0.222$^{***}$ & $-$0.068$^{***}$ & $-$0.151$^{***}$ & 0.018$^{***}$ & $-$0.139$^{***}$ & 0.143$^{***}$ & $-$0.071$^{***}$ \\ 
-  & (0.00002) & (0.00000) & (0.00002) & (0.0001) & (0.00003) & (0.0001) & (0.00003) & (0.00003) \\ 
-  & & & & & & & & \\ 
- skilled trade jobs, 2020 & 0.521$^{***}$ & 0.171$^{***}$ & 0.214$^{***}$ & $-$0.578$^{***}$ & 0.231$^{***}$ & 1.067$^{***}$ & 0.284$^{***}$ & 0.049$^{***}$ \\ 
-  & (0.00002) & (0.00000) & (0.00004) & (0.00004) & (0.00004) & (0.0001) & (0.00003) & (0.00003) \\ 
-  & & & & & & & & \\ 
- professional jobs, 2020 & $-$0.577$^{***}$ & $-$0.962$^{***}$ & $-$0.565$^{***}$ & $-$0.860$^{***}$ & $-$0.566$^{***}$ & $-$1.139$^{***}$ & $-$0.389$^{***}$ & $-$0.702$^{***}$ \\ 
-  & (0.00003) & (0.00000) & (0.00004) & (0.0001) & (0.00004) & (0.0001) & (0.0001) & (0.00004) \\ 
-  & & & & & & & & \\ 
- administrative jobs, 2020 & 0.200$^{***}$ & $-$1.324$^{***}$ & $-$0.050$^{***}$ & 0.136$^{***}$ & $-$0.137$^{***}$ & $-$0.101$^{***}$ & $-$0.018$^{***}$ & $-$0.155$^{***}$ \\ 
-  & (0.00002) & (0.00000) & (0.00002) & (0.00004) & (0.00002) & (0.00005) & (0.00002) & (0.00002) \\ 
-  & & & & & & & & \\ 
- service jobs, 2020 & $-$0.082$^{***}$ & $-$0.180$^{***}$ & $-$0.659$^{***}$ & $-$1.066$^{***}$ & $-$0.746$^{***}$ & $-$0.944$^{***}$ & $-$0.773$^{***}$ & $-$0.783$^{***}$ \\ 
-  & (0.00002) & (0.00000) & (0.00003) & (0.00002) & (0.00002) & (0.00003) & (0.00002) & (0.00001) \\ 
-  & & & & & & & & \\ 
- machine operation jobs, 2020 & 0.210$^{***}$ & 0.574$^{***}$ & 0.439$^{***}$ & $-$0.589$^{***}$ & 0.229$^{***}$ & $-$0.260$^{***}$ & 0.080$^{***}$ & 0.127$^{***}$ \\ 
-  & (0.00001) & (0.00000) & (0.00002) & (0.00002) & (0.00001) & (0.00002) & (0.00001) & (0.00001) \\ 
-  & & & & & & & & \\ 
- earnings, 2019 & $-$0.003$^{***}$ & 0.029$^{***}$ & 0.019$^{***}$ & 0.069$^{***}$ & 0.018$^{***}$ & 0.050$^{***}$ & 0.016$^{***}$ & 0.026$^{***}$ \\ 
-  & (0.001) & (0.0001) & (0.001) & (0.002) & (0.001) & (0.002) & (0.001) & (0.001) \\ 
-  & & & & & & & & \\ 
- n. business est. per hab., 2019 & 1.930$^{***}$ & $-$0.377$^{***}$ & 0.826$^{***}$ & 0.251$^{***}$ & 0.315$^{***}$ & 1.515$^{***}$ & $-$0.948$^{***}$ & $-$4.011$^{***}$ \\ 
-  & (0.00000) & (0.000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
-  & & & & & & & & \\ 
- furloughed per hab., 2020 & $-$1.518$^{***}$ & $-$0.152$^{***}$ & 11.014$^{***}$ & 0.972$^{***}$ & $-$26.969$^{***}$ & 2.059$^{***}$ & 4.236$^{***}$ & 10.500$^{***}$ \\ 
-  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
-  & & & & & & & & \\ 
- AM tests per hab., 2020 & 0.002$^{***}$ & $-$0.011$^{***}$ & $-$0.140$^{***}$ & 0.015$^{***}$ & 0.231$^{***}$ & 0.005$^{***}$ & $-$0.025$^{***}$ & $-$0.078$^{***}$ \\ 
-  & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) & (0.000) \\ 
-  & & & & & & & & \\ 
- Virgin Media \%, 2020 & $-$0.349$^{***}$ & 12.236$^{***}$ & 5.619$^{***}$ & 0.489$^{***}$ & 1.721$^{***}$ & $-$16.289$^{***}$ & $-$5.581$^{***}$ & 6.126$^{***}$ \\ 
-  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
-  & & & & & & & & \\ 
- Constant & $-$4.570$^{***}$ & $-$1.045$^{***}$ & $-$6.539$^{***}$ & 5.629$^{***}$ & 2.957$^{***}$ & 3.386$^{***}$ & $-$2.531$^{***}$ & 2.979$^{***}$ \\ 
-  & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) & (0.00000) \\ 
-  & & & & & & & & \\ 
-\hline \\[-1.8ex] 
-McFadden's R squared & 0.435 & 0.435 & 0.435 & 0.435 & 0.435 & 0.435 & 0.435 & 0.435 \\ 
-N & 322 & 322 & 322 & 322 & 322 & 322 & 322 & 322 \\ 
-Akaike Inf. Crit. & 742.901 & 742.901 & 742.901 & 742.901 & 742.901 & 742.901 & 742.901 & 742.901 \\ 
-\hline 
-\hline \\[-1.8ex] 
-\textit{Note:}  & \multicolumn{8}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
-\end{tabular} 
-\end{sidewaystable}
-
-Using an auxiliary multinomial logit regression, we test whether the
-clusters that have higher mean speeds and more reliable services do
-indeed consist of LADs that are more urban or closer to major urban
-areas and are more likely to benefit from a choice of high quality
-internet services. This part of the regression aims to confirm any first
-level digital divides. Next, to better understand how the clusters fare
-in relation to the second level digital divide, we consider which LADs
-in which clusters have a higher proportion of occupations where the
-nature of the work and the skills that occupation employs enable
-telecommuting. Finally, we consider the rate of return on internet use,
-or third level digital divides, by reviewing which clusters have the
-highest earnings and job density, and also which clusters experienced
-the highest share of population furloughed during this period in the
-pandemic.
-
-The results of the auxiliary regression are presented in Table
-\ref{aux}. The dependent variable is the LAD cluster membership as
-described in the \protect\hyperlink{sec:3}{methods and data} section and
-equation \ref{eq1}. Each column represents a different cluster. The
-reference case is cluster \(4\), which includes only the local authority
-of Hambleton in North Yorkshire, a rural area of just over ninety
-thousand people. Mean, experienced upload speed in cluster \(4\) (see
-Table \ref{up.cluster.descr}) is close to the pre-clustered average for
-the whole sample (\(9.3\)Mb/s). However, the standard deviation for
-cluster \(4\) and the difference between average speeds in the morning
-compared to the evening peak periods are indications of worse
-reliability than many of the other clusters. Hence, the results in Table
-\ref{aux} should be seen as relative rather than absolute probabilities.
-
-First, we control for the number of speed tests run per cluster
-inhabitant between \(9:00\)-\(10:59\) as well as the share of fast
-Virgin Media internet connections \footnote{See Appendix
-  \protect\hyperlink{appendix3}{3} for the descriptive statistics.}.
-Regarding the former, we expect people in LADs with more unreliable
-connections to test their internet speeds more often, and the results
-show that those in cluster \(6\) are by far the most concerned about
-reliability at that time of day. Meanwhile, those in clusters \(2\),
-\(3\), and \(9\) benefit from a higher proportion of Virgin connections,
-which is an indication that people in these clusters are more likely to
-live in urban areas, with more profitable broadband markets, and a
-better choice of broadband services. We also employ distance from London
-and from the nearest metropolitan area (including London) as two
-variables depicting peripherality and urban structure. However, whilst
-significant, the pattern of coefficients of these variables is
-inconclusive in confirmed first level digital divides. Even though
-London was one of the ten largest metropolitan areas in England, which,
-along with Glasgow and Cardiff, were identified to calculate the
-variable estimating the impact of distance from the centre of a
-metropolitan area, the coefficients for the two variables are mostly
-small and the signs for some are opposite.
-
-When looking at the constituent authorities (see full list of LADs in
-Appendix \protect\hyperlink{appendix1}{1} as well as Figure
-\ref{map.up.clusters}), however, Cluster \(9\) is clearly more urban,
-including \(13\) of \(32\) London Boroughs, five of the seven
-constitutent LADs of the West Midlands conurbation, nine of the ten
-boroughs of Greater Manchester, and five of nine other large
-metropolitan areas coded in for the `distance to nearest met area'
-variable. Cluster \(9\) also includes the other main cities in the East
-Midlands, Leicester and Derby, as well as smaller cities known for their
-knowledge economy, such as Oxford and Milton Keynes. However, as
-previously mentioned, there is more noise within the variables for the
-two largest clusters than the smaller ones, and cluster \(6\) also
-contains many urban areas. These include \(16\) London boroughs,
-Birmingham City, Bolton in Greater Manchester, Leeds, Liverpool,
-Newcastle and Bristol, as well as small cities with knowledge economies
-such as Cambridge, Edinburgh, and Reading and its neighbours in the
-high-tech agglomeration of the Thames Valley. And yet cluster \(6\) also
-includes some of the most rural areas in the country, showing how
-important it is to interpretation to consider the membership of the
-cluster as well as the regression results.
-
-The coefficients in Table \ref{aux} demonstrate that those in cluster
-\(6\) were more likely to hold managerial, tech and professional jobs
-than those in cluster \(9\), and were the least likely to be furloughed
-of any cluster. Thus, although speeds were slow and not as reliable
-during the morning peak as in cluster \(9\), the skills that enabled
-telecommuting also enabled greater returns from doing so, despite lower
-than average earnings. Cluster \(6\) also has the second highest job
-density, with more jobs to keep. Likewise, those in LADs in clusters
-\(1\) and \(8\) had the highest and second highest proportions of
-residents working in tech or managerial occupations despite suffering
-from lower speeds and unreliable services, another mismatch between
-first and second level digital divides. Yet unlike cluster \(6\) and
-despite occupations with better digital skills, cluster \(8\), with its
-five peripheral suburban areas in the Midlands, Scotland and south of
-London, as well as rural West Devon saw the third highest proportion of
-its working population furloughed. It also had the second lowest average
-earnings in 2019, suggesting that its population was more likely to be
-on the wrong side of the third level digital divide before, as well as
-during the pandemic. In contrast, the nine LADs of cluster \(1\), which
-include more rural areas, plus Westminster in central London had the
-lowest earnings of any cluster, but also lowest level of furlough after
-cluster \(6\).
-
-The regression results for cluster \(7\) also reveal the complexity of
-intersecting and diverging digital divides. Along with \(3\) and \(9\),
-the analysis in Section \protect\hyperlink{sec:4.1}{4.1} suggested this
-cluster was on the `right' side of any first level digital divides, yet
-the regression shows that it has the lowest proportion of Virgin
-connections. It is furthest from a metropolitan area, but closest to
-London. The five LADs in cluster \(7\) are all suburban, making it
-sensible that the cluster is in the group with better infrastructure,
-but none are central urban boroughs and, along with a London suburb,
-Leeds suburb, Manchester suburb, and West Midlands suburb, the fifth LAD
-is outside Belfast. Belfast was not included among the metropolitan
-areas, and whilst there has been substantial investment in high speed
-broadband in Northern Ireland in recent years, including by Virgin,
-take-up may still lag behind historic networks in other urban markets.
-
-Cluster \(7\) has the lowest proportion of residents currently employed
-in professional occupations, and among the lowest proportions in
-managerial or technical occupations. This suggests potentially a low
-level of skills to take advantage of the quality internet service
-available, a second level digital divide. Cluster \(7\) also has the
-lowest job density. However, with the largest proportion of skilled
-tradespeople, and the second highest number of businesses per
-inhabitant, those in cluster \(7\) ranked second in terms of earnings in
-2019. On the other hand, the lack of some of the occupations most likely
-to telecommute during lockdown may have contributed both to a
-substantial proportion being furloughed in 2020, and even to the
-relative reliability of broadband speeds compared to many of the other
-smaller clusters.
-
-Cluster \(3\), meanwhile, had the greatest proportion of its workforce
-furloughed of any cluster, despite having the highest job density and
-the second highest average speeds. Ranking in the middle for most of the
-occupations estimated, at least some of its population should have the
-skills to work from home. The geography would suggest economic advantage
-as well, with eight of its twelve LADs in the South of England,
-including two London Boroughs, four suburban areas north of London, and
-the two tightly bounded urban areas of Eastbourne and Ipswich -- a
-greater proportion than any other cluster. Yet geographic position and
-job density do not appear to equate to particularly high average
-earnings in 2019, nor did they protect residents from furlough in 2020.
-In contrast, cluster \(5\), comprised of three suburban districts in the
-Northwest, Midlands, and the London green belt, had the highest average
-earnings in 2019, despite low levels of residents working in managerial,
-tech and professional occupations. Suburbs are considered the most
-likely urban form in which telecommuters live \citep{e2018does}, yet the
-contrast between these small, mainly suburban clusters -- \(7\), \(3\),
-and \(5\), suggest that they are not all equally economically resilient
-in a pandemic.
-
-Cluster \(2\), which, like cluster \(5\), had faster morning broadband
-speeds than evening speeds, as shown in Table \ref{up.cluster.descr}
-contains two remote LADs, the Isles of Scilly and Northeast
-Lincolnshire. Whilst the low demand may be due to the least residents in
-tech occupations, it may also be due to high numbers of retired people.
-Rural areas such as those in cluster \(2\) are often home to many older,
-retired people \citep{blank2018local}, which may also explain some of
-the contradictions in cluster \(1\) with its low levels of furlough and
-low earnings, whilst the presence of Westminster in that cluster might
-be why there are the highest levels of tech and managerial occupations.
-
-In summary, the regression results indicate the complexity of measuring
-second and third level digital divides within spatial aggregations where
-the geography of first level digital divides has been captured using
-time-series clustering of experienced broadband upload speeds as a
-product of reliability not just availability. Whilst the differences in
-mean speeds between clusters were not large, the temporal clustering of
-internet resilience showed much greater variation, and was not spatially
-dependent upon distance to large urban areas, or relative location.
-Internet resilience supports a wide range of small and large urban
-economies, but also fails, or at least frustrates a wide range of other
-urban economies, as demonstrated by the number of morning peak tests per
-capita run in cluster \(6\). Yet our analysis of the likelihood of
-furlough within the population, especially in the smaller suburban
-clusters \(7\), \(3\), \(5\), and \(8\) shows that digital divides were
-as likely to diverge as intersect during the pandemic, and did not
-necessarily overlap with prior economic divides. With a much greater
-share of the population likely to continue to work from home in the
-future, and with changing attitudes towards residential locations, our
-analysis suggests that first level digital divides should be seen as a
-function of reliability as well as availability, of upload speeds as
-well as download speeds, and of greater spatial variation than might
-have previously been considered.
-
-\hypertarget{sec:5}{%
-\section{Discussion and Conclusions}\label{sec:5}}
-
-Our analysis demonstrated that the temporal profiles of seven of our
-nine clusters had slower upload speeds in the morning than in the
-evening. The opposite is likely to have been the norm prior to the
-pandemic, as level of demand and bandwidth management is the most common
-cause of temporal variation in experienced speeds, and why evening
-download speeds, rather than daytime upload speeds, have been used to
-benchmark the performance of internet services. Thus, the new patterns
-can be taken as evidence of widespread telecommuting and other daytime
-internet use which changed the temporal profile of internet activity
-throughout the UK, not just in areas with more digital industry or
-better skills. Furthermore, upload speeds have not previously been seen
-as integral to universal service, considering there has never before
-been such extreme demand for telecommuting and operations such as video
-calls. Yet those in the largest cluster, \(6\), clearly experienced
-lower upload speeds. Their speeds were also less reliable, perhaps
-because they had higher levels of demand than the second largest
-cluster, \(9\), where many more employees were furloughed. Indeed, those
-with the skills and jobs to work from home were often left with the
-least robust services and greatest slowdown.
-
-Therefore, home-based digital infrastructure which considers upload
-speeds and working day reliability as well as availability are likely to
-be particularly important in a future where telecommuting might be a
-more common means of accessing work and broadband services must be fit
-for purpose. Although the long-term effects of such drastic changes in
-telecommuting and attitudes towards working from home are difficult to
-predict, the reliability of home broadband services deserve more
-consideration than in the past. This would represent a switch from
-previous demand-side broadband policies in the UK which tended to be
-aimed at supporting small and medium enterprises
-\citep{HENDERSON2020102024}. SUch policies were based on previous
-research regarding the productivity effects of broadband infrastructure
-\citep{DESTEFANO2018110}. However, this stream of research tended not to
-consider residential locations as places that host economic activity.
-Such policies also made some assumptions about the quality of service in
-residential markets that could be assured without government
-intervention -- an assumption which our analysis suggest is not entirely
-accurate.
-
-Policy and development proposals should also consider how the potential
-changes wrought by the pandemic span various aspects of economy and
-society. Changes to transportation planning due to altered commuting
-patterns also imply changes in land use and urban planning to
-accommodate people who work from home
-\citep{BUDNITZ2020102713, ELLDER2020102777}. Productivity and innovation
-changes will reflect upon changes in agglomeration externalities and the
-attraction of large cities \citep{econobs}. Further research may be able
-to measure the economic resilience of the different clusters of places
-discussed in this paper once this pandemic is firmly past. However, our
-analysis demonstrates that the economic resilience made possible by
-working from home cannot be understood without considering the
-underpinning digital divides and cannot be achieved without planning for
-how the levels of digital, social and economic divides might intersect.
-For example, our results suggest that broadband policies cannot improve
-the economic resilience of places where the industrial structure does
-not align with occupations that incorporate the digital skills and
-capabilities to work from home. Such places instead experienced higher
-proportions of their labour force being placed on furlough.
-
-Early research on Covid-19 and cities also speculated upon the changes
-that potential extensive post-Covid working from home patterns might
-generate for spatial structure: from more walkable cities and more
-localised production and consumption patterns to more extensive urban
-sprawl, the decline in public transportation and increased private car
-usage \citep{batty2020editorial}. Despite the essential role it played
-during the pandemic, less effort has been spent in understanding the
-current and future role of digital infrastructure. Our results indicate
-that probably for the first time the future of cities and spatial
-structure are so intertwined with digital infrastructure. If the
-post-covid world is a world with extensive working from home, then we
-need to build, among other things, resilient digital infrastructure
-capable of bridging the first layer of the digital divide. Contrary to
-previous broadband business and deployment plans, emphasis should be
-placed not only on download, but also on upload speeds
-\citep{brake2020lessons}. In essence, Covid-19 accentuated the old
-argument, now more valid than ever: digital infrastructure, just like
-any other network infrastructure, only becomes visible when it stops
-working \citep{star1999, tranos2013geography}.
-
-On the other hand, the nuanced picture we gained through our analysis of
-the UK case study suggests that being on the right side of the second
-level digital divide had a greater impact on economic resilience and
-therefore the third level digital divide, than having quality internet
-connectivity. Our regressions results for cluster \(1\) and \(6\) are a
-demonstration of this. Our analysis shows both the intersections and
-divergences of digital and economic division. Almost all of the largest
-urban areas in the UK, as well as many smaller cities, were split
-between the largest two clusters -- \(6\) and \(9\) -- and both contain
-LADs that are centres of the knowledge economy, from Cambridge in the
-former to Oxford in the latter, or have high concentrations of digital
-businesses, like Reading in the former and Milton Keynes in the latter
-\citep{technation2017}. Yet whilst LADs in cluster \(9\) were able to
-benefit both from reliable internet connections and populations familiar
-with working from home to capitalise on their digital infrastructure,
-they appeared to have a lower rate of return in the pandemic, with far
-greater numbers furloughed. Is this an indication of historic economic
-division, as well as the third level digital divide, in cities in the
-North and Midlands unable to capitalise on their digital infrastructure?
-Meanwhile, did the greater presence in London and the South of LADs in
-cluster \(6\), as well as the inclusion of many rural areas less
-impacted by the pandemic help those in this cluster be more resilient
-and gain greater returns, despite using less reliable internet services?
-Further research would be necessary to prise apart the detail,
-particularly in the largest clusters where averages are less distinct.
-
-In conclusion, this paper offers a new perspective on telecommuting from
-the viewpoint of the complex web of digital divides. We employ novel
-data regarding experienced upload speeds and time-series clustering
-methods, a family of unsupervised machine learning techniques which are
-rarely utilised in geographical research. Fast, reliable internet
-connections are necessary for the population to be able to work from
-home. Although not every place hosts individuals in occupations which
-allow for telecommuting nor with the necessary skills to effectively use
-the internet to telecommute, this paper raises the issue that places may
-depend upon good internet reliability as well as connectivity to achieve
-economic resilience in a period like the current pandemic when internet
-resilience is so vital.
-
-\pagebreak
-
-\hypertarget{appendix1}{%
-\section*{Appendix 1}\label{appendix1}}
-\addcontentsline{toc}{section}{Appendix 1}
-
-This is the LAD cluster membership for the upload speed timeseries.
-
-\textbf{Cluster 1: } Ceredigion, Darlington, Eden, Erewash, Kettering,
-Rossendale, Runnymede, Rutland, Westminster
-
-\textbf{Cluster 2: } Isles of Scilly, North East Lincolnshire
-
-\textbf{Cluster 3: } Broxbourne, Carlisle, Corby, Dundee City,
-Eastbourne, Harlow, Hertsmere, Hounslow, Ipswich, Luton, Newham, South
-Northamptonshire
-
-\textbf{Cluster 4: } Hambleton
-
-\textbf{Cluster 5: } Fylde, Rugby, Three Rivers
-
-\textbf{Cluster 6: } Aberdeenshire, Adur, Allerdale, Amber Valley,
-Angus, Antrim and Newtownabbey, Argyll and Bute, Armagh City, Banbridge
-and Craigavon, Arun, Ashford, Aylesbury Vale, Babergh, Barnet,
-Barrow-in-Furness, Basildon, Bassetlaw, Bath and North East Somerset,
-Bedford, Belfast, Birmingham, Blackburn with Darwen, Blackpool, Blaenau
-Gwent, Bolton, Boston, Bournemouth, Christchurch and Poole, Bracknell
-Forest, Bradford, Braintree, Breckland, Brentwood, Bridgend, Brighton
-and Hove, Bristol, City of, Broadland, Bromley, Calderdale, Cambridge,
-Camden, Canterbury, Carmarthenshire, Causeway Coast and Glens, Central
-Bedfordshire, Chelmsford, Cherwell, Cheshire East, Cheshire West and
-Chester, Chesterfield, Chichester, Chiltern, City of Edinburgh, City of
-London, Clackmannanshire, Conwy, Copeland, Cornwall, Cotswold, County
-Durham, Coventry, Craven, Croydon, Dacorum, Dartford, Denbighshire,
-Derbyshire Dales, Derry City and Strabane, Dorset, Dover, Dumfries and
-Galloway, Ealing, East Cambridgeshire, East Devon, East Dunbartonshire,
-East Hampshire, East Hertfordshire, East Lindsey, East Renfrewshire,
-East Riding of Yorkshire, East Suffolk, Elmbridge, Epping Forest, Epsom
-and Ewell, Exeter, Fareham, Fenland, Fermanagh and Omagh, Fife,
-Flintshire, Folkestone and Hythe, Forest of Dean, Gateshead, Gloucester,
-Gosport, Great Yarmouth, Greenwich, Gwynedd, Hackney, Harborough,
-Haringey, Hartlepool, Hastings, Havering, Herefordshire, County of, High
-Peak, Highland, Hillingdon, Horsham, Huntingdonshire, Inverclyde, Isle
-of Anglesey, King's Lynn and West Norfolk, Kingston upon Hull, City of,
-Kingston upon Thames, Kirklees, Lambeth, Lancaster, Leeds, Lincoln,
-Liverpool, Maidstone, Malvern Hills, Mansfield, Melton, Merthyr Tydfil,
-Mid Devon, Mid Suffolk, Mid Ulster, Midlothian, Mole Valley,
-Monmouthshire, Moray, Na h-Eileanan Siar, Neath Port Talbot, New Forest,
-Newcastle upon Tyne, Newry, Mourne and Down, North Devon, North East
-Derbyshire, North Lanarkshire, North Lincolnshire, North Norfolk, North
-Somerset, North Tyneside, North Warwickshire, North West Leicestershire,
-Northumberland, Orkney Islands, Pembrokeshire, Pendle, Perth and
-Kinross, Peterborough, Plymouth, Powys, Preston, Reading, Redcar and
-Cleveland, Reigate and Banstead, Rhondda Cynon Taf, Ribble Valley,
-Richmondshire, Rother, Rotherham, Rushcliffe, Rushmoor, Ryedale,
-Scottish Borders, Sedgemoor, Sefton, Sevenoaks, Shetland Islands,
-Shropshire, Somerset West and Taunton, South Ayrshire, South Bucks,
-South Cambridgeshire, South Gloucestershire, South Hams, South Holland,
-South Lakeland, South Norfolk, South Oxfordshire, South Ribble, South
-Somerset, South Staffordshire, Southend-on-Sea, Southwark, St Albans,
-Stafford, Stirling, Stoke-on-Trent, Stratford-on-Avon, Stroud, Swale,
-Swansea, Swindon, Teignbridge, Tendring, Test Valley, Tewkesbury,
-Thanet, Tonbridge and Malling, Torbay, Torfaen, Torridge, Tower Hamlets,
-Tunbridge Wells, Uttlesford, Wakefield, Waltham Forest, Wandsworth,
-Warrington, Watford, Waverley, Wellingborough, West Berkshire, West
-Lancashire, West Lindsey, West Oxfordshire, Wiltshire, Winchester,
-Windsor and Maidenhead, Wirral, Wokingham, Worcester, Worthing, Wrexham,
-Wychavon, Wycombe, Wyre
-
-\textbf{Cluster 7: } Lisburn and Castlereagh, Selby, Tameside, Welwyn
-Hatfield, Wyre Forest
-
-\textbf{Cluster 8: } Cannock Chase, East Lothian, Lichfield, Oadby and
-Wigston, Tandridge, West Devon
-
-\textbf{Cluster 9: } Aberdeen City, Ards and North Down, Ashfield,
-Barking and Dagenham, Barnsley, Basingstoke and Deane, Bexley, Blaby,
-Bolsover, Brent, Bromsgrove, Broxtowe, Burnley, Bury, Caerphilly,
-Cardiff, Castle Point, Charnwood, Cheltenham, Chorley, Colchester,
-Crawley, Daventry, Derby, Doncaster, Dudley, East Ayrshire, East
-Northamptonshire, East Staffordshire, Eastleigh, Enfield, Falkirk,
-Gedling, Glasgow City, Gravesham, Guildford, Halton, Hammersmith and
-Fulham, Harrogate, Harrow, Hart, Havant, Hinckley and Bosworth,
-Hyndburn, Isle of Wight, Islington, Kensington and Chelsea, Knowsley,
-Leicester, Lewes, Lewisham, Maldon, Manchester, Medway, Mendip, Merton,
-Mid and East Antrim, Mid Sussex, Middlesbrough, Milton Keynes, Newark
-and Sherwood, Newcastle-under-Lyme, Newport, North Ayrshire, North
-Hertfordshire, North Kesteven, Northampton, Norwich, Nottingham,
-Nuneaton and Bedworth, Oldham, Oxford, Portsmouth, Redbridge, Redditch,
-Renfrewshire, Richmond upon Thames, Rochdale, Rochford, Salford,
-Sandwell, Scarborough, Sheffield, Slough, Solihull, South Derbyshire,
-South Kesteven, South Lanarkshire, South Tyneside, Southampton,
-Spelthorne, St.~Helens, Staffordshire Moorlands, Stevenage, Stockport,
-Stockton-on-Tees, Sunderland, Surrey Heath, Sutton, Tamworth, Telford
-and Wrekin, Thurrock, Trafford, Vale of Glamorgan, Vale of White Horse,
-Walsall, Warwick, Wealden, West Dunbartonshire, West Lothian, West
-Suffolk, Wigan, Woking, Wolverhampton, York
-
-\hypertarget{appendix2}{%
-\section*{Appendix 2}\label{appendix2}}
-\addcontentsline{toc}{section}{Appendix 2}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{figures/map.up.2019.clusters} \caption{\label{map.up.2019.clusters}Upload speed clusters for LADs in before the pandemic (2019)}\label{fig:unnamed-chunk-17}
-\end{figure}
-
-\pagebreak
-
-\hypertarget{appendix3}{%
-\section*{Appendix 3}\label{appendix3}}
-\addcontentsline{toc}{section}{Appendix 3}
-
-\begin{table}[!htbp] \centering 
-  \caption{Descriptive statistics for the auxiliary regression explanatory variables\label{descr.aux}} 
-  \label{} 
-\footnotesize 
-\begin{tabular}{@{\extracolsep{0pt}}lccccccc} 
-\\[-1.8ex]\hline 
-\hline \\[-1.8ex] 
-Statistic & \multicolumn{1}{c}{N} & \multicolumn{1}{c}{Mean} & \multicolumn{1}{c}{St.
-Dev.} & \multicolumn{1}{c}{Min} & \multicolumn{1}{c}{Pctl(25)} & \multicolumn{1}{c}{Pctl(75)} & \multicolumn{1}{c}{Max} \\
-\hline \\[-1.8ex] 
-pop, 2018 & 365 & 174,952.100 & 119,557.100 & 8,700.000 & 100,400.000 & 214,900.000 & 1,141,400.000 \\ 
-job density, 2018 & 365 & 1.137 & 5.726 & 0.400 & 0.700 & 0.930 & 110.110 \\ 
-distance to nearest met. area & 365 & 53.269 & 57.700 & 0.150 & 22.050 & 69.290 & 544.090 \\ 
-distance to London & 365 & 201.558 & 173.634 & 0.150 & 76.180 & 278.880 & 1,003.950 \\ 
-south of the UK & 365 & 0.463 & 0.499 & 0.000 & 0.000 & 1.000 & 1.000 \\ 
-managerial jobs, 2020 & 363 & 12.009 & 4.013 & 3.600 & 9.000 & 14.300 & 27.900 \\ 
-tech jobs, 2020 & 364 & 14.505 & 4.057 & 3.500 & 11.800 & 16.900 & 29.600 \\ 
-skilled trade jobs, 2020 & 358 & 10.513 & 3.764 & 1.000 & 8.025 & 12.500 & 21.600 \\ 
-professional jobs, 2020 & 364 & 21.223 & 6.902 & 4.400 & 16.775 & 24.850 & 71.600 \\ 
-administrative jobs, 2020 & 359 & 9.965 & 2.738 & 3.200 & 8.100 & 11.400 & 21.300 \\ 
-service jobs, 2020 & 362 & 9.261 & 2.827 & 2.800 & 7.300 & 11.400 & 17.800 \\ 
-machine operation jobs, 2020 & 337 & 6.339 & 2.847 & 1.200 & 4.400 & 7.900 & 19.800 \\ 
-earnings, 2019 & 360 & 592.184 & 81.129 & 437.600 & 534.625 & 633.875 & 893.200 \\ 
-Virgin Media \%, 2020 & 381 & 0.150 & 0.140 & 0.000 & 0.018 & 0.238 & 0.753 \\ 
-n. business est. per hab., 2019 & 365 & 0.057 & 0.164 & 0.023 & 0.038 & 0.056 & 3.174 \\ 
-AM tests per hab., 2020 & 365 & 0.0005 & 0.0002 & 0.0001 & 0.0003 & 0.001 & 0.001 \\ 
-furloughed per hab., 2020 & 362 & 0.115 & 0.013 & 0.077 & 0.107 & 0.122 & 0.178 \\ 
-\hline \\[-1.8ex] 
-\end{tabular} 
-\end{table}
-
-\pagebreak
-
-\bibliographystyle{tfcad}
-\bibliography{bibliography.bib}
-
-\end{document}

### 2021-05-11 paper/v2_taylor_francis_for_proof/broadband.speed.covid_for_tables.Rmd
not my day... the revised .Rmd for the .tex tables

--- 2021-04-19 paper/v2_taylor_francis/broadband.speed.covid.Rmd
@@
 
+path.out = paste0(path, "/paper/v2_taylor_francis_for_proof/tables/table_1.tex")
+
 stargazer(clustersLADup[1:7],
           summary = F,
+          type = "latex",
+          out = path.out,
           font.size = "footnotesize",
-          type = "latex",
           header = F,
@@
 
+path.out = paste0(path, "/paper/v2_taylor_francis_for_proof/tables/table_2.tex")
+
 stargazer(aux.up,
           type = "latex",
+          out = path.out,
           nobs = T,
@@
 
+path.out = paste0(path, "/paper/v2_taylor_francis_for_proof/tables/table_A1.tex")
+
 stargazer(for.descr,
           summary = T,
+          out = path.out,
           font.size = "footnotesize",

### 2021-06-09 presentations/RinR2021.Rmd
presentation for Monday + renaming a script for rendering the Word file

---
title: "Working from home and digital divides: resilience during the pandemic"
author: |
  | 
  | Hannah Budnitz & Emmanouil Tranos
  |
  | <small>University of Bristol, Alan Turing Institute</small> 
  | <small>e.tranos@bristol.ac.uk [\@EmmanouilTranos](https://twitter.com/EmmanouilTranos) [etranos.info](https://etranos.info/)</small>
output: 
  revealjs::revealjs_presentation:
    theme: simple
    self_contained: true
    reveal_options:
      width: 1150
      height: 720
bibliography: ./bibliography.bib
biblio-style: "apalike"
css: style.css
---

```{r, echo=FALSE, results=FALSE, include=FALSE}
library(tidyverse)
library(geojsonio)
library(rprojroot)
library(knitr)
library(stargazer)
library(nnet)

knitr::opts_chunk$set(echo = FALSE, message = F, warning = F) # By default, hide code; set to TRUE to see code
#knitr::opts_chunk$set(out.width = '100%', dpi=300) # Figure resolution and size

# This is the project path
path <- find_rstudio_root_file()

```

## Cite:

<br>
Budnitz, H., and E. Tranos (2021) Working from home and digital divides: resilience during the pandemic. *Annals of the AAG*. In press. [link](https://doi.org/10.1080/24694452.2021.1939647)
<br>
<br>
[etranos.info/post/rir2021](https://etranos.info/post/rir2021)

## Contents

- Introduction
- Experienced broadband speed data
- Analysis
- Conclusions

# Introduction

## WFH / Telecommuting
- From a niche means of accessing work to a widespread way of life
- From $5$% to $47$% in the UK [@ons2020; @ons2020lm2019]
- $37$% WFM in European April $2020$ 
- $60$% in Finland  [@eurofound2020]
- Almost $50$% in the US, $37$% could be permanently moved online [@brynjolfsson2020covid; @NBERw26948]

## WFH / Telecommuting
- WFM = economic resilience during Covid-19
- capability to WFM:
  - occupations
  - quality digital infrastructure
- Spatial clustering and dependence

## Digital divides
- 1st level: availability and quality of internet connectivity
- 2nd level: skills to effectively utilise digital technologies
- 3rd level: returns of internet usage
- Cross-level interactions

## Our approach
- Intersection of digital and material divides
- Economic resilience of places as digital technologies became an essential
tool of productivity during the pandemic
<br>
<br>
**internet services + occupations $\Rightarrow$ WFM $\Rightarrow$ economic resilience**

## Our approach
- UK Local Authorities Districts (LAD) as a case study
- Quality of internet service: *experienced* internet speeds during the spring $2020$ lockdown
- Time-series clustering to create clusters of local authorities with similar temporal signatures of internet speeds
  - $\Rightarrow$ 1st level of digital divide
- Regress cluster membership against their socioeconomic and geographic characteristics 
  - $\Rightarrow$ how the 1st level of digital divide intersects with the 2nd and 3rd

# Data

## Experienced Broadband Speeds 
- *Experienced* vs. advertised internet speeds
- $\Uparrow$ internet demand \ usage $\Downarrow$ experienced internet speed
- [www.broadbandspeedchecker.co.uk/](https://www.broadbandspeedchecker.co.uk/)
- @riddlesden2014broadband & @nardotto2015unbundling
- Stream data, anonymised individual speed tests
- Upload speeds and the frequency of speed tests March to May $2020$
<br>
<br>
<center>
![](speedtest.png){width=40%}
</center>

---

```{r echo=FALSE, out.width='60%'}
path.image = paste0(path,"/paper/v2_taylor_francis/figures/time.var.plot2020.png")
knitr::include_graphics(path.image)
```

---

```{r echo=FALSE, out.width='60%'}
path.image = paste0(path,"/paper/v2_taylor_francis/figures/time.var.plot2019.png")
knitr::include_graphics(path.image)
```

## Experienced Broadband Speeds 
- An overall trend of increased testing from March to April and then a slight reduction from April to May
- $2019$ evening peak of testing, leisure activities, download speeds
- $2020$ **new morning peak**

# Analysis

## Aggregation 
- Relatively scarce data to analyse as stream
- Composite week time series
- Mon-Friday, hourly $6:00$ -- $24:00$, excl. bank holidays
- $18$ hourly data points $x$ $5$ weekdays ($90$) for each local authority ($382$)
- Focus on upload speeds

## Time series clustering
- *Aim*: create clusters of local authorities with similar temporal signatures of internet speed
- Shape-based approach
- Match two separate time-series objects based on the similarity of their shapes 
- Distances between the shapes
- Simplicity: 
  - *k*-means
  - medoids, time-series objects with a minimal distance to all other cluster objects

---
```{r echo=FALSE, message=FALSE, results = "asis"}
path.clustersLADup <- paste(path, "./data/temp/LAD.cluster.up.csv", sep = "")

clustersLADup <- read_csv(path.clustersLADup)

# clustersLADup <- clustersLADup %>%
#   dplyr::rename('N. of LADs' = up.LADs,
#          'mean upload speed' = upMean,
#          'SD upload speed' = upSD,
#          'mean AM upload speed' = upAMmean)

clustersLADup <- clustersLADup %>%
  mutate(across(4:7, round, 0))

stargazer(clustersLADup[1:7],
          summary = F,
          font.size = "footnotesize",
          type = "html",
          header = F,
          digit.separator = ",",
          digit.separate = c(3,3),
          covariate.labels=c("Cluster",
                             "N. of LADs",
                             "LAD population",
                             "mean speed",
                             "SD speed",
                             "mean AM speed",
                             "mean PM speed"),
                             #"AM_slower",
                             #"SDtoMean"),
          rownames = F,
          no.space=TRUE,
          column.sep.width = "0pt",
          notes = "Note: All speed measures are upload speeds",
          title = "Upload speed cluster characteristics")
```

## Time series clustering

<small>

- Cluster 1: Medium sized mostly rural cluster; slow mean upload speeds; high workday temporal variation.

- Cluster 2: Small rural cluster; highest mean upload speeds; slowdown more in evening.

- Cluster 3: Medium sized mixed cluster; fast mean upload speeds; medium temporal variation.

- Cluster 4: Rural reference LAD; medium mean upload speeds; high temporal variation.

- Cluster 5: Small suburban cluster; fast mean upload speeds; slowdown more in evening.

- **Cluster 6: Large mixed cluster; slowest aggregate mean upload speeds; low workday temporal variation.**

- Cluster 7: Small suburban cluster; fast mean upload speeds; medium temporal variation.

- Cluster 8: Small mainly suburban cluster; slow mean upload speeds; high workday temporal variation.

- **Cluster 9: Large, more urban cluster; fast mean upload speeds; lowest workday temporal variation.**

</small>

---

```{r echo=FALSE, out.width='60%'}
path.image = paste0(path,"/paper/v2_taylor_francis/figures/map.up.clusters.png")
knitr::include_graphics(path.image)
```

## Post-clustering regression analysis
- Explain the characteristics of clusters
- How the 1st level of the digital divide intersects with the 2nd and 3rd?
- Multinomial logit regression: cluster membership against:
<small>

  - population
  - pop. and job density
  - distance to nearest met. area and to London, South of the UK dummy
  - % of managerial, tech, skilled trade, professional, administrative, leisure, machine operation jobs
  - earnings, n. business est. per hab.
  - % furloughed pop., 2020
  - AM broadband tests per hab.
  - % of Virgin Media connections

</small>

---

<center>
![](regression.png){width=120%}
</center>

---

```{r eval=FALSE, echo=FALSE, message=FALSE, results = "asis"}
path.LAs_Clusters <- paste0(path, "/data/temp/LAs_Clusters_k9.RData")
load(path.LAs_Clusters)

aux.up <- multinom(cluster.up ~ 
                     #pop16_64 + # non sign, while pop is
                     pop +
                     #pop.dens + 
                     job.dens2018 + # pop.dens is still insignificant when job.dens is excluded
                     distMet + distLondon + south +
                     managers + tech + skilled + prof + admin + caring + plant +
                     earnings + I(total.busi/pop) + I(furlough/pop) +
                     I(LAD.tests.am20/pop) + # non sensitive to addition of other tests pm
                     virgin
                   , data = data, 
                   trace = F) # to avoid messages

# McFadden's R squared
aux.up.null <- multinom(cluster.up ~ 1, data = data)
r2.up <- 1-logLik(aux.up)/logLik(aux.up.null)

n.obs <- dim(aux.up$fitted.values)[1]

stargazer(aux.up, 
          type = "html",
          #type = "text",
          nobs = T,
          digits = 3,
          font.size = "footnotesize",
          covariate.labels=c(#"pop. 16-64, 2018",
            "pop, 2018",
            #"pop. density, 2018",
            "job density, 2018",
            "distance to nearest met. area",
            "distance to London",
            "South of the UK",
            "% of managerial jobs, 2020",
            "% of tech jobs, 2020",
            "% of skilled trade jobs, 2020",
            "% of professional jobs, 2020",
            "% of administrative jobs, 2020",
            "% of leisure jobs, 2020",
            "% of machine operation jobs, 2020",
            "earnings, 2019",
            "n. business est. per hab., 2019",
            "% furloughed pop., 2020",
            "n. of AM broadband tests per hab., 2020",
            "% of Virgin Media connections"),
          add.lines = list(c("McFadden's R squared",
                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
                             , round(r2.up[1],3), round(r2.up[1],3)),
                           (c("N",
                              n.obs, n.obs, n.obs,
                              n.obs, n.obs, n.obs,
                              n.obs, n.obs, n.obs
                              , n.obs, n.obs))))

# ls()
# 
# str(aux.up)
# 
# summary(aux.up)

# tidy_model <- broom::tidy(aux.up)
# stargazer(tidy_model)

# stargazer(aux.up,
#           type = "text",
#           nobs = T,
#           digits = 3,
#           font.size = "footnotesize",
#           #float.env = "sidewaystable",
#           dep.var.caption = "",
#           title = "Auxiliary multinomial regression of upload speed clusters on socio-economic and geographic LAD variables",
#           header = F,
#           covariate.labels=c(#"pop. 16-64, 2018",
#                              "pop, 2018",
#                              #"pop. densuty, 2018",
#                              "job density, 2018",
#                              "distance to nearest met. area",
#                              "distance to London",
#                              "South of the UK",
#                              "managerial jobs, 2020",
#                              "tech jobs, 2020",
#                              "skilled trade jobs, 2020",
#                              "professional jobs, 2020",
#                              "administrative jobs, 2020",
#                              "service jobs, 2020",
#                              "machine operation jobs, 2020",
#                              "earnings, 2019",
#                              "n. business est. per hab., 2019",
#                              "furloughed per hab., 2020",
#                              "AM tests per hab., 2020",
#                              "Virgin Media \\%, 2020"),
#           add.lines = list(c("McFadden's R squared",
#                              round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
#                              round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
#                              round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
#                              , round(r2.up[1],3), round(r2.up[1],3)),
#                            (c("N",
#                             n.obs, n.obs, n.obs,
#                             n.obs, n.obs, n.obs,
#                             n.obs, n.obs, n.obs
#                             , n.obs, n.obs))))
```
## Clusters 6 and 9

- LADs in Cluster 6 are more likely to hold managerial, tech and professional jobs
- Least likely to be furloughed
- Slow and less reliable speeds
- High job density

## Clusters 6 and 9

- Although speeds were slow and not as reliable during the morning peak as in cluster 9, the skills that enabled telecommuting
also enabled **greater returns** from doing so ($\Downarrow$ furloughed)
- LADs in cluster 9 were able to benefit both from reliable internet connections and populations
able to work from home to capitalise on their digital infrastructure
- But **lower returns** in the pandemic $\Rightarrow$ greater numbers furloughed

## Being on the right side of digital divides

- Clusters 3, 7 and 9 are on the right side of the 1st layer of digital divide
- Internet resilience supports, but also frustrates a wide range of other urban economies: 
  - $\Uparrow$ morning peak tests per capita run in cluster 6
- Being on the right side of the 2nd level digital divide had a **greater impact on economic resilience** (3rd level digital divide), than having quality internet connectivity 
  - $\Rightarrow$ cluster 1 and 6

# Conclusions

---

- Seven of our nine clusters had slower upload speeds in the morning than in the evening

- Evidence of widespread telecommuting and other daytime internet use (e.g. education)
which changed the temporal profile of internet activity throughout the UK

- Upload speeds have not previously been seen as integral to universal service, considering there has never before been such extreme demand for telecommuting and operations such as video calls

---

- Τelecommuting from the viewpoint of the complex web of digital divides

- Quality of internet infrastructure

- Νot every place hosts individuals with the necessary *skills* and in *occupations* to effectively use the internet to telecommute

- Places may depend upon good internet reliability and connectivity to achieve economic resilience in a period like the current pandemic

```{css eval=FALSE, echo=FALSE, message=FALSE}
.slide {
    height: 750px;
    overflow-y: auto !important;
}

# it aligns all the slides
# .reveal section {
#         text-align: left;
#     }

```

## References {.allowframebreaks}

### 2021-08-20 presentations/RinR2021.Rmd
online appendix

@@
 <br>
-[etranos.info/post/rir2021](https://etranos.info/post/rir2021)
+[etranos.info/post/rinr2021](https://etranos.info/post/rinr2021)
 
@@
 ## WFH / Telecommuting
-- WFM = economic resilience during Covid-19
-- capability to WFM:
+- WFH = economic resilience during Covid-19
+
+- Capability to WFH:
   - occupations
   - quality digital infrastructure
+
 - Spatial clustering and dependence
@@
 - 1st level: availability and quality of internet connectivity
+
 - 2nd level: skills to effectively utilise digital technologies
+
 - 3rd level: returns of internet usage
+
 - Cross-level interactions
@@
 - Intersection of digital and material divides
+
 - Economic resilience of places as digital technologies became an essential
@@
 ## Experienced Broadband Speeds 
-- An overall trend of increased testing from March to April and then a slight reduction from April to May
+- An overall trend of increased testing from March to April and then a slight reduction from April to May in 2020, compared to less testing as Spring progresses in 2019
+
 - $2019$ evening peak of testing, leisure activities, download speeds
-- $2020$ **new morning peak**
+- $2020$ **new morning peak** $\Rightarrow$ flatter hourly temporal profile
 
@@
 - Relatively scarce data to analyse as stream
+
 - Composite week time series
+
 - Mon-Friday, hourly $6:00$ -- $24:00$, excl. bank holidays
+
 - $18$ hourly data points $x$ $5$ weekdays ($90$) for each local authority ($382$)
+
 - Focus on upload speeds
@@
 - *Aim*: create clusters of local authorities with similar temporal signatures of internet speed
+
 - Shape-based approach
@@
 ```
-## Clusters 6 and 9
+## Cluster 6
 
-- LADs in Cluster 6 are more likely to hold managerial, tech and professional jobs
+- Workers are more likely to hold managerial, tech and professional jobs
 - Least likely to be furloughed
@@
 also enabled **greater returns** from doing so ($\Downarrow$ furloughed)
+
 - LADs in cluster 9 were able to benefit both from reliable internet connections and populations
 able to work from home to capitalise on their digital infrastructure
+
 - But **lower returns** in the pandemic $\Rightarrow$ greater numbers furloughed
@@
 - Clusters 3, 7 and 9 are on the right side of the 1st layer of digital divide
-- Internet resilience supports, but also frustrates a wide range of other urban economies: 
+
+- Internet resilience supports, but also frustrates urban economies in different geographies: 
   - $\Uparrow$ morning peak tests per capita run in cluster 6
+
 - Being on the right side of the 2nd level digital divide had a **greater impact on economic resilience** (3rd level digital divide), than having quality internet connectivity 
@@
 
+- Places may depend upon good internet reliability and connectivity to achieve economic resilience in a period like the current pandemic
+
 - Νot every place hosts individuals with the necessary *skills* and in *occupations* to effectively use the internet to telecommute
-
-- Places may depend upon good internet reliability and connectivity to achieve economic resilience in a period like the current pandemic
 
@@
 
-## References {.allowframebreaks}
+# References {.allowframebreaks}

### 2021-08-20 presentations/RinR2021_HB.Rmd
online appendix

--- 2021-08-20 presentations/RinR2021.Rmd
@@
 <br>
-[etranos.info/post/rinr2021](https://etranos.info/post/rinr2021)
+[etranos.info/post/rir2021](https://etranos.info/post/rir2021)
 
@@
 - From $5$% to $47$% in the UK [@ons2020; @ons2020lm2019]
-- $37$% WFM in European April $2020$ 
-- $60$% in Finland  [@eurofound2020]
+- $37$% WFH in Europe April $2020$ 
 - Almost $50$% in the US, $37$% could be permanently moved online [@brynjolfsson2020covid; @NBERw26948]
@@
 <br>
-**internet services + occupations $\Rightarrow$ WFM $\Rightarrow$ economic resilience**
+**internet services + occupations $\Rightarrow$ WFH $\Rightarrow$ economic resilience**
 
@@
 
-- Composite week time series
+- Composite week time series for March-May 2020
 
@@
           notes = "Note: All speed measures are upload speeds",
-          title = "Upload speed cluster characteristics")
+          title = "Upload speed cluster characteristics Spring 2020")
 ```
@@
 
-- Workers are more likely to hold managerial, tech and professional jobs
+- Workers in LADs more likely to hold managerial, tech and professional jobs
 - Least likely to be furloughed
@@
 
-- Although speeds were slow and not as reliable during the morning peak as in cluster 9, the skills that enabled telecommuting
+- Although speeds in cluster 6 were slow and not as reliable during the morning peak as in cluster 9, the skills that enabled telecommuting
 also enabled **greater returns** from doing so ($\Downarrow$ furloughed)
 
-- LADs in cluster 9 were able to benefit both from reliable internet connections and populations
+- LADs in cluster 9 were able to benefit from reliable internet connections and still included substantial populations
 able to work from home to capitalise on their digital infrastructure
@@
 
-- Clusters 3, 7 and 9 are on the right side of the 1st layer of digital divide
+- Some clusters, such as 9, are on the right side of the 1st layer of digital divide
 
@@
 - Being on the right side of the 2nd level digital divide had a **greater impact on economic resilience** (3rd level digital divide), than having quality internet connectivity
-  - $\Rightarrow$ cluster 1 and 6
 
@@
 
-# References {.allowframebreaks}
+## References {.allowframebreaks}

### 2021-08-20 src/online_appendix.Rmd
online appendix

--- 2021-04-19 src/online_appendix.Rmd
@@
 ---
-title: "Online Appendix"
+title: 'Working from home and digital divides: resilience during the pandemic. Supplemental Material.'
 output: 
   flexdashboard::flex_dashboard:
+    vertical_layout: fill
     orientation: columns
-    vertical_layout: fill
     social: menu
@@
 library(tidyverse)
-library(plotly)
 library(lubridate)
@@
 library(ggthemes)
+library(leaflet)
 
@@
 
-# save clusters file
+# read clusters file
 path.clusters <- paste(path, "/data/temp/clusters_nodiff_9.csv", sep = "")
@@
 
-Column 
+Column {data-width=800}
 -----------------------------------------------------------------------
 
-### LAD, Upload speed clusters, 2020
+### Upload speed clusters for Local Authority Districts, 2020. See [Budnitz and Tranos (2021)](https://doi.org/10.1080/24694452.2021.1939647) for more details.
 
-```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE, fig.height=10, fig.asp=T}
-#out.width='100%'
+```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE, fig.asp=T}
 
 # LAD fully generalised
-la <- readOGR("https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.geojson",
-              verbose = FALSE)
+# la <- readOGR("https://opendata.arcgis.com/datasets/3a4fa2ce68f642e399b4de07643eeed3_0.geojson",
+#               verbose = FALSE)
 # saved locally in /data/raw/Local_Authority_Districts_(December_2019)_Boundaries_UK_BUC.geojson
+
+path.geojson <- paste0(path, "/data/raw/Local_Authority_Districts_(December_2019)_Boundaries_UK_BGC.geojson")
+la <- readOGR(path.geojson, verbose = FALSE)
 
@@
 la <- spTransform(la, CRS("+init=epsg:4326"))
-#la@data$LAD19NM <- as.character(la@data$LAD19NM)
+
 la@data$lad19nm <- as.character(la@data$lad19nm)
@@
 
-# fortify
-la.f <- broom::tidy(la, region = "lad19nm")
-la.f <- merge(la.f, clusters.nodiff, by.x = "id", by.y = "LAD")
+# merge with the data
+la <- sp::merge(la, clusters.nodiff, by.x = "lad19nm", by.y = "LAD")
 
 # create the text to tooltip
-la.f <- merge(la.f, for.tip.mean.up.am, by.x = "id", by.y = "LAD19NM", all.x=T)
-la.f <- merge(la.f, for.tip.mean.up.pm, by.x = "id", by.y = "LAD19NM", all.x=T)
+la <- sp::merge(la, for.tip.mean.up.am, by.x = "lad19nm", by.y = "LAD19NM", all.x=T)
+la <- sp::merge(la, for.tip.mean.up.pm, by.x = "lad19nm", by.y = "LAD19NM", all.x=T)
 
-#sapply(la.f, function(x) sum(is.na(x)))
+#sapply(la@data, function(x) sum(is.na(x)))
 #No am data points for Isles of Scilly
-la.f$tip <- paste0(la.f$id, "<br>",
-                   "Mean upload AM speed: ", round(la.f$mean.up.am), " Kbps", "<br>",
-                   "Mean upload PM speed: ", round(la.f$mean.up.pm), " Kbps")
-la.f <- la.f[order(la.f$order),] # if i don't order merge.nuts.f loses order and the map has gaps
+
+# text for labels
+la@data$tip <- paste0(la@data$lad19nm, "<br>",
+                   "Mean upload AM speed: ", round(la@data$mean.up.am), " Kbps", "<br>",
+                   "Mean upload PM speed: ", round(la@data$mean.up.pm), " Kbps") %>% 
+  lapply(htmltools::HTML)
 
 # using world.cities from `maps`
-cities <- world.cities %>% filter(country.etc=="UK")
+cities <- world.cities %>% 
+  filter(country.etc=="UK") %>% 
+  arrange(pop) %>% tail(10)
 
-p <- ggplot(la.f, aes(x = long, y = lat)) +
-  geom_polygon(aes(group = group, fill = as.factor(cluster.up), 
-                   text = tip)) +# ,text =paste("test1", long, "test2", lat)
-  #labs(title = "Upload clusters") +
-  scale_fill_brewer(palette="Set1", 
-                    direction = -1) +
-  #geom_text_repel( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=4) +
-  geom_point( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat), color="black", size=0.5) +
-  geom_text( data=cities %>% arrange(pop) %>% tail(10), aes(x=long, y=lat, label=name), size=3) +
+# Create a color palette for the map:
+mypalette <- colorFactor(palette="Set1", 
+                         domain=la$cluster.up, 
+                         na.color="transparent",
+                         reverse = T)
 
-  guides(fill=guide_legend(title="Cluster")) +
-  theme_map()
-#   theme(panel.border = element_blank(), panel.grid.major = element_blank(),
-# panel.grid.minor = element_blank())
+leaflet(la) %>% 
+  addTiles()  %>% 
+  fitBounds(-5.7, 50, -1.2, 58.7) %>% 
+  addPolygons(fillColor = ~mypalette(cluster.up), 
+              stroke=T,
+              color = "white",
+              weight = .8,
+              opacity = .3,
+              fillOpacity = .5,
+              label = ~tip) %>% 
+  addCircles(lng = cities$long, lat = cities$lat,
+             weight = 2, radius = 5, color = "black",
+             label = cities$name,
+             labelOptions = labelOptions(noHide = T, direction = 'right', 
+                                         textOnly = TRUE, permanent = F),
+             options = markerOptions(riseOnHover = TRUE)) %>% 
+  # addLabelOnlyMarkers(data = cities,
+  #                   lng = ~long, lat = ~lat, label = ~name,
+  #                   labelOptions = labelOptions(noHide = T, direction = 'auto', textOnly = TRUE)) %>%
+  addProviderTiles(providers$Esri.WorldTerrain) %>% 
+  addLegend(pal = mypalette,
+            values = ~cluster.up,
+            position = "topleft",
+            opacity = .5,
+            title = "Upload speed<br>clusters")
 
-p %>%
-  ggplotly(tooltip = "text") %>%
-  style(hoverlabel = list(bgcolor = "white"), hoveron = "fill")
 ```
@@
 
-### LAD mean AM and PM upload speeds
+### Local Authority District mean AM and PM upload speeds
 

### 2021-09-22 src/old/LAs_Clusters.Rmd
tidy up

--- 2020-12-02 src/LAs_Clusters.Rmd
@@
 
-aux <- multinom(cluster.up ~ 
+aux.up <- multinom(cluster.up ~ 
                   #pop16_64 + # non sign, while pop is
@@
 # McFadden's R squared
-aux.null <- multinom(cluster.up ~ 1, data = data)
-r2 <- 1-logLik(aux)/logLik(aux.null)
+aux.up.null <- multinom(cluster.up ~ 1, data = data)
+r2.up <- 1-logLik(aux.up)/logLik(aux.up.null)
 
-n.obs <- dim(aux$fitted.values)[1]
+n.obs <- dim(aux.up$fitted.values)[1]
 
-stargazer(aux, 
-          type = "html",
-          #type = "text",
+stargazer(aux.up, 
+          #type = "html",
+          type = "text",
           nobs = T,
@@
                              "% of NVQ4+",
-                             "n. of AM broadband tests per hab., 2019",
+                             "n. of AM broadband tests per hab., 2020",
                              "% of Virgin Media connections"),
           add.lines = list(c("McFadden's R squared",
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3)
-                             , round(r2[1],3), round(r2[1],3)),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3),
+                             round(r2.up[1],3), round(r2.up[1],3), round(r2.up[1],3)
+                             , round(r2.up[1],3), round(r2.up[1],3)),
                            (c("N",
@@
 
-aux <- multinom(cluster.tests ~ 
+aux.tests <- multinom(cluster.tests ~ 
                   #pop16_64 + # non sign, while pop is
@@
 # McFadden's R squared
-aux.null <- multinom(cluster.tests ~ 1, data = data)
-r2 <- 1-logLik(aux)/logLik(aux.null)
+aux.tests.null <- multinom(cluster.tests ~ 1, data = data)
+r2.tests <- 1-logLik(aux.tests)/logLik(aux.tests.null)
 
-n.obs <- dim(aux$fitted.values)[1]
+n.obs <- dim(aux.tests$fitted.values)[1]
 
-stargazer(aux, 
+stargazer(aux.tests, 
           type = "html",
@@
                              "% of NVQ4+",
-                             "n. of AM broadband tests per hab., 2019",
+                             "n. of AM broadband tests per hab., 2020",
                              "% of Virgin Media connections"),
           add.lines = list(c("McFadden's R squared",
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3)),
+                             round(r2.tests[1],3), round(r2.tests[1],3), round(r2.tests[1],3),
+                             round(r2.tests[1],3), round(r2.tests[1],3), round(r2.tests[1],3)),
                            (c("N",
@@
 
-aux <- multinom(cluster.down ~ 
+aux.down <- multinom(cluster.down ~ 
                   #pop16_64 + # non sign, while pop is
@@
 # McFadden's R squared
-aux.null <- multinom(cluster.down ~ 1, data = data)
-r2 <- 1-logLik(aux)/logLik(aux.null)
+aux.down.null <- multinom(cluster.down ~ 1, data = data)
+r2.down <- 1-logLik(aux.down)/logLik(aux.down.null)
 
-n.obs <- dim(aux$fitted.values)[1]
+n.obs <- dim(aux.down$fitted.values)[1]
 
-stargazer(aux, 
+stargazer(aux.down, 
           type = "html",
@@
                              "% of NVQ4+",
-                             "n. of AM broadband tests per hab., 2019",
+                             "n. of AM broadband tests per hab., 2020",
                              "% of Virgin Media connections"),
           add.lines = list(c("McFadden's R squared",
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3), round(r2[1],3), round(r2[1],3),
-                             round(r2[1],3)),
+                             round(r2.down[1],3), round(r2.down[1],3), round(r2.down[1],3),
+                             round(r2.down[1],3), round(r2.down[1],3), round(r2.down[1],3),
+                             round(r2.down[1],3), round(r2.down[1],3), round(r2.down[1],3),
+                             round(r2.down[1],3), round(r2.down[1],3), round(r2.down[1],3),
+                             round(r2.down[1],3), round(r2.down[1],3), round(r2.down[1],3),
+                             round(r2.down[1],3), round(r2.down[1],3), round(r2.down[1],3),
+                             round(r2.down[1],3)),
                            (c("N",
@@
 ```
+
+```{r}
+path.out <- paste0(path, "/data/temp/LAs_Clusters.RData")
+save.image(path.out)
+```

### 2021-09-22 src/old/online_appendix_old.Rmd
tidy up

--- 2021-04-14 src/online_appendix_v2.Rmd
@@
 title: "Online Appendix"
-output: 
-  flexdashboard::flex_dashboard:
-    orientation: columns
-    vertical_layout: fill
-    social: menu
+author: ""
+output: html_document
+header-includes:
+- \usepackage{multirow}
 knit: (function(inputFile, encoding) {
@@
 ```
+## Local Authority Districts
 
-Column 
------------------------------------------------------------------------
+The below map illustrates the different clusters of LAD based on the upload 
+experienced internet speeds. Hovering over the LAD polygons, the map also presents
+the mean AM upload internet speed and the relevant ranking in the country for each 
+LAD.
 
-### LAD, Upload speed clusters, 2020
+**Is there a chance to briefly describe the clusters with 1-2 words?**
 
-```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE, fig.height=10}
-#out.width='100%'
+- Cluster 1:
+
+- Cluster 2: 
+...
+
+<br>
+```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE,fig.height=10}
 
@@
 la.f$tip <- paste0(la.f$id, "<br>",
-                   "Upload AM speed: ", round(la.f$mean.up.am), " Kbps", "<br>",
+                   "Upload am: ", round(la.f$mean.up.am), " Kbps", "<br>",
                    "Rank: ", la.f$rank)
@@
 
-  guides(fill=guide_legend(title="Cluster")) +
+  guides(fill=guide_legend(title="Clusters")) +
   theme_map()
@@
 
-Column
------------------------------------------------------------------------
-
-### LAD mean AM upload speeds and ranking
-
 ```{r, echo=FALSE, warning=FALSE, message = FALSE, error=FALSE,fig.height=10}
-for.tip.mean.up.am <- merge(for.tip.mean.up.am, clusters.nodiff, 
-                            by.x = "LAD19NM", by.y = "LAD", all.x = T)
-
 for.tip.mean.up.am %>% 
-  select(-cluster.down, -cluster.tests) %>% 
     arrange(rank) %>% 
   kable(digits = 0,
-        format.args = list(big.mark = ","),
-        col.names = c("Local Authority",
-                      "Mean AM upload sped",
-                      "Rank",
-                      "Cluster")) %>% 
+        format.args = list(big.mark = ",")) %>% 
   kable_styling("striped", full_width = F) %>% 
-  scroll_box() #width = "500px", height = "400px"
+  scroll_box(width = "500px", height = "400px")
+
 ```
-
-### Description
-
-The below map illustrates the different clusters of LAD based on the upload 
-experienced internet speeds. Hovering over the LAD polygons, the map also presents
-the mean AM upload internet speed and the relevant ranking in the country for each 
-LAD.
-
-**Is there a chance to briefly describe the clusters with 1-2 words?**
-
-- Cluster 1:
-
-- Cluster 2:
-
-...

### 2021-09-22 src/old/timeseries_plots.Rmd
tidy up

---
title: "Time series plots"
date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
knit: (function(inputFile, encoding) {
    rmarkdown::render(inputFile, encoding = encoding, output_dir = "../output")
  })
---

```{r settings, include = FALSE}
library(tsbox)
library(ggplot2)
library(tidyverse)
library(geojsonio)
library(tsfeatures)
library(rprojroot)
library(openair)
library(reshape2)
library(lubridate)
library(rgdal)
library(maptools)
library(sp)
library(rgeos)
library(classInt)
library(RColorBrewer)
library(knitr)
library(rprojroot)
library(viridis)
library(hrbrthemes)

# This is the project path
path <- find_rstudio_root_file()
```

```{r include = FALSE}

# load and clean the data.
# copied from bbData2020.Rmd

#upload data file
path.data <- paste(path, "./data/raw/speedtest.csv", sep = "")

#bb2020 <- read.csv("./data/raw/speedtest.csv")
bb2020 <- read.csv(path.data)

#name columns
names(bb2020)[1:6] <- c("datetext","speeddown","speedup","provider","lat", "lon")

str(bb2020)
summary(bb2020) #2030205 obs
#conversions needed to numeric? df$col <- as.numeric(df$col) 
#remove empty rows or coerced to NA? df[complete.cases(df),] or df <- df[!is.na(df$col),]

#remove outliers based on Riddlesden and Singleton 2014 for slower end
#check fastest upload and download speeds commercially available from Virgin Media (Gigaclear offers even faster!)
bb2020 <- bb2020[bb2020$speeddown>512,] 
bb2020 <- bb2020[bb2020$speeddown<362000,]
bb2020 <- bb2020[bb2020$speedup<21000,] #1839333

#remove below seconds, create columns for date and hour
bb2020$datetext <- strtrim(x = bb2020$datetext, width = 19) 
bb2020$datetext2 <- bb2020$datetext
bb2020$datetext2 <- strtrim(x = bb2020$datetext2, width = 13)
bb2020$datetext2 <- colsplit(bb2020$datetext2, pattern = " ", c("date", "hour"))
bb2020$hour <- bb2020$datetext2$hour
bb2020$date <- bb2020$datetext2$date
#delete datetext2 column(s)
bb2020 <- bb2020[ ,c(1:6,8:9)]
#convert to date form datetext and date columns
bb2020$datetext <- as.POSIXct(strptime(bb2020$datetext, format = "%Y-%m-%d %H:%M:%S", "GMT")) 
bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
#create dataframe with number of tests, mean and standard deviation by month for adding to dataframe later
MonthFreq <- timeAverage(bb2020, avg.time = "month", statistic = "frequency")
MonthMean <- timeAverage(bb2020, avg.time = "month", statistic = "mean")
MonthSD <- timeAverage(bb2020, avg.time = "month", statistic = "sd")
MonthStats <- cbind(MonthFreq[,1:2], MonthMean[,c(3:4)], 
                    MonthSD[,c(3:4)])
names(MonthStats) <- c("month", "monthlyTests", "monthlyMeanSpDown", 
                       "monthlyMeanSpUp", "monthlySDSpDown", 
                       "monthlySDSpUp")
MonthStats$month <- month(MonthStats$month, label = TRUE, abbr = TRUE)
#add column for month and day of the week to create stats below
bb2020$month <- month(bb2020$date, label = TRUE, abbr = TRUE)
bb2020$weekday <- wday(bb2020$date, label = TRUE)
#divide MonthStats for future join with separate 2019 and 2020 dataframes
MonthStats19 <- MonthStats[1:6,]
MonthStats20 <- MonthStats[13:17,]
```

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}

# Structured TS

# convert to spatial object
coords_bb <- cbind(bb2020$lon, bb2020$lat)
bb.la <- SpatialPointsDataFrame(coords_bb, data = data.frame(bb2020))
proj4string(bb.la) <- CRS("+init=epsg:4326") #define projection

# get LA 
# (i) directly from the web
#la <- readOGR("http://geoportal1-ons.opendata.arcgis.com/datasets/b6d2e15801de45328b760a4f55d74318_0.geojson?outSR={%22latestWkid%22:3857,%22wkid%22:102100}")#, layer="OGRGeoJSON")
# or, (ii) from the json I saved locally
path.json <- paste(path, "./data/raw/Local_Authority_Districts_(April_2019)_Boundaries_UK_BFE.geojson", sep = "")
la <- readOGR(path.json)#, layer="OGRGeoJSON")
# UK BGC la <- readOGR("https://opendata.arcgis.com/datasets/0e07a8196454415eab18c40a54dfbbef_0.geojson")
# UK BFC la <- readOGR("https://opendata.arcgis.com/datasets/1d78d47c87df4212b79fe2323aae8e08_0.geojson")

# source: https://data.gov.uk/dataset/7c387c64-d25f-474a-b07e-b933578caea2/local-authority-districts-april-2019-boundaries-uk-bfe

# spatial transformations
la <- spTransform(la, CRS("+init=epsg:4326"))
la@data$LAD19NM <- as.character(la@data$LAD19NM)

# spatial join to LAD
bb.la.sp <- over(bb.la, la[, "LAD19NM"]) #not a spatial object
bb.la$LAD19NM <- bb.la.sp$LAD19NM

# create dataframe object to analyse (remove lat and lon?)
bb2020sp <- bb.la@data

# drop NAs
bb2020sp <- bb2020sp[!is.na(bb2020sp$LAD19NM),]
#13431/1839332 = .0073 NA LAD19NM, coord outside UK

# summarise by date and geography
la.date.count <- count(bb2020sp, LAD19NM, date)
la.dateAvg <- summarise(group_by(la.date.count, LAD19NM), mean(n))
range(la.dateAvg$`mean(n)`)

# map LA mean test frequency per date
la@data <- left_join(la@data, la.dateAvg, by = "LAD19NM")
var <- la@data[ ,'mean(n)']
breaks <- classIntervals(var, n = 5, style = "quantile")
my_colours <- brewer.pal(5, "Blues")
plot(la, col = my_colours[findInterval(var, breaks$brks, all.inside = TRUE)], axes = FALSE,
                 border = rgb(0.8, 0.8, 0.8, 0))
# drop LA
rm(la)
```

## Sub-daily timeseries

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}

bb2020sp <- bb2020sp %>% arrange(datetext)

# create sub daily blocks
bb2020sp$block[bb2020sp$hour>=0 & bb2020sp$hour<8] <- 1
bb2020sp$block[bb2020sp$hour>=8 & bb2020sp$hour<16] <- 2
bb2020sp$block[bb2020sp$hour>=16 & bb2020sp$hour<24] <- 3

#bb2020sp$date.block <- paste(bb2020sp$date, bb2020sp$date.block, sep = "_")
# I use block as hour for the ts_wide()
bb2020sp$date.block = ymd_h(paste(bb2020sp$date, bb2020sp$block))

head(bb2020sp)
```

## Download speeds

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
ts.down <- bb2020sp %>%
  group_by(date.block,LAD19NM) %>%
  summarise(mean.down = mean(speeddown))#, mean.down = mean(speeddown), n.tests = n())
sapply(ts.down, function(x) sum(is.na(x)))

# https://www.data-to-viz.com/caveat/spaghetti.html

mean.down.uk <- ts.down %>%
  group_by(date.block) %>%
  summarise(mean.down.uk = mean(mean.down))#, mean.down = mean(speeddown), n.tests = n())

start.date =  "2019-12-31"

ggplot(subset(ts.down,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = "red")) +
  geom_line() + guides(colour=FALSE) + ggtitle("Download speeds") + #xlab("") +
  ylab("download speed") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~LAD19NM, ncol = 2) +
  geom_line(data = subset(mean.down.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.down.uk, group = 1), colour="green") #

```

## Upload speeds

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
ts.up <- bb2020sp %>%
  group_by(date.block,LAD19NM) %>%
  summarise(mean.up = mean(speedup))
sapply(ts.up, function(x) sum(is.na(x)))

# https://www.data-to-viz.com/caveat/spaghetti.html

mean.up.uk <- ts.up %>%
  group_by(date.block) %>%
  summarise(mean.up.uk = mean(mean.up))

start.date =  "2019-12-31"

ggplot(subset(ts.up,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.up, group = LAD19NM, colour = "red")) +
  geom_line() + guides(colour=FALSE) + ggtitle("Upload speeds") + #xlab("") +
  ylab("upload speed") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~LAD19NM, ncol = 2) +
  geom_line(data = subset(mean.up.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.up.uk, group = 1), colour="green") #

```

## Speed tests

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
ts.tests <- bb2020sp %>%
  group_by(date.block,LAD19NM) %>%
  summarise(n.tests = n())
sapply(ts.tests, function(x) sum(is.na(x)))

# https://www.data-to-viz.com/caveat/spaghetti.html

# n.tests.uk <- ts.tests %>%
#   group_by(date.block) %>%
#   summarise(n.tests = n()/382)

mean.tests.uk <- ts.tests %>%
  group_by(date.block) %>%
  summarise(mean.tests.uk = mean(n.tests))

start.date =  "2019-12-31"

ggplot(subset(ts.tests,date.block > start.date), aes(x=as.POSIXct(date.block), y=n.tests, group = LAD19NM, colour = "red")) +
  geom_line() + guides(colour=FALSE) + ggtitle("N. of tests") + #xlab("") +
  ylab("tests") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~LAD19NM, ncol = 2, scales = "free_y") +
  geom_line(data = subset(mean.tests.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.tests.uk, group = 1), colour="green") #.uk

```

## Download diff 2020 - 2019

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}

ts.down.diff <- ts.down
ts.down.diff$date.block.lag1y <- ts.down.diff$date.block - years(1)
ts.down.diff <- merge(ts.down.diff,ts.down.diff, by.x = c("date.block.lag1y", "LAD19NM"), by.y = c("date.block","LAD19NM"), all = F)

ts.down.diff <- ts.down.diff %>%
  rename(mean.down2019 = mean.down.y, mean.down2020 = mean.down.x) %>%
  select(-date.block.lag1y.y)

ts.down.diff$diff <- ts.down.diff$mean.down2020 - ts.down.diff$mean.down2019

head(ts.down.diff)
sapply(ts.down.diff, function(x) sum(is.na(x)))

# https://www.data-to-viz.com/caveat/spaghetti.html

mean.down.diff.uk <- ts.down.diff %>%
  group_by(date.block) %>%
  summarise(mean.down.diff.uk = mean(diff))#, mean.down = mean(speeddown), n.tests = n())

start.date =  "2019-12-31"

ggplot(subset(ts.down.diff,date.block > start.date), aes(x=as.POSIXct(date.block), y=diff, group = LAD19NM, colour = "red")) +
  geom_line() + guides(colour=FALSE) + ggtitle("Diff. in download speeds") + #xlab("") +
  ylab("diff. in download speed") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~LAD19NM, ncol = 2) +
  geom_line(data = subset(mean.down.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.down.uk, group = 1), colour="green") #
```

## Upload diff 2020 - 2019

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}

ts.up.diff <- ts.up
ts.up.diff$date.block.lag1y <- ts.up.diff$date.block - years(1)
ts.up.diff <- merge(ts.up.diff,ts.up.diff, by.x = c("date.block.lag1y", "LAD19NM"), by.y = c("date.block","LAD19NM"), all = F)

ts.up.diff <- ts.up.diff %>%
  rename(mean.up2019 = mean.up.y, mean.up2020 = mean.up.x) %>%
  select(-date.block.lag1y.y)

ts.up.diff$diff <- ts.up.diff$mean.up2020 - ts.up.diff$mean.up2019

head(ts.up.diff)
sapply(ts.up.diff, function(x) sum(is.na(x)))

# https://www.data-to-viz.com/caveat/spaghetti.html

mean.up.diff.uk <- ts.up.diff %>%
  group_by(date.block) %>%
  summarise(mean.up.diff.uk = mean(diff))#, mean.up = mean(speedup), n.tests = n())

start.date =  "2019-12-31"

ggplot(subset(ts.up.diff,date.block > start.date), aes(x=as.POSIXct(date.block), y=diff, group = LAD19NM, colour = "red")) +
  geom_line() + guides(colour=FALSE) + ggtitle("Diff. in upload speeds") + #xlab("") +
  ylab("diff. in upload speed") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~LAD19NM, ncol = 2) +
  geom_line(data = subset(mean.up.diff.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.up.diff.uk, group = 1), colour="green") #
```

## Speed tests diff. 2020 - 2019

```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
ts.tests.diff <- ts.tests
ts.tests.diff$date.block.lag1y <- ts.tests.diff$date.block - years(1)
ts.tests.diff <- merge(ts.tests.diff,ts.tests.diff, by.x = c("date.block.lag1y", "LAD19NM"), by.y = c("date.block","LAD19NM"), all = F)

ts.tests.diff <- ts.tests.diff %>%
  rename(n.tests2019 = n.tests.y, n.tests2020 = n.tests.x) %>%
  select(-date.block.lag1y.y)

ts.tests.diff$diff <- ts.tests.diff$n.tests2020 - ts.tests.diff$n.tests2019

head(ts.tests.diff)
sapply(ts.tests.diff, function(x) sum(is.na(x)))

mean.tests.diff.uk <- ts.tests.diff %>%
  group_by(date.block) %>%
  summarise(mean.tests.diff.uk = mean(diff))

start.date =  "2019-12-31"

ggplot(subset(ts.tests.diff,date.block > start.date), aes(x=as.POSIXct(date.block), y=diff, group = LAD19NM, colour = "red")) +
  geom_line() + guides(colour=FALSE) + ggtitle("Diff. in speed tests") + #xlab("") +
  ylab("diff. in speed tests") +
  scale_y_continuous(labels = scales::comma) +
  scale_x_datetime(date_breaks = "1 month") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~LAD19NM, ncol = 2) +
  geom_line(data = subset(mean.tests.diff.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.tests.diff.uk, group = 1), colour="green") #
```

### 2021-09-22 src/old/timeseries_plots0916.Rmd
tidy up

--- 2021-09-22 src/old/timeseries_plots.Rmd
@@
 
-## Sub-daily timeseries
+## Keep working hours
 
@@
 
-# create sub daily blocks
-bb2020sp$block[bb2020sp$hour>=0 & bb2020sp$hour<8] <- 1
-bb2020sp$block[bb2020sp$hour>=8 & bb2020sp$hour<16] <- 2
-bb2020sp$block[bb2020sp$hour>=16 & bb2020sp$hour<24] <- 3
+# keep working hours
+bb2020sp <- bb2020sp[bb2020sp$hour>=9 & bb2020sp$hour<17,]
 
-#bb2020sp$date.block <- paste(bb2020sp$date, bb2020sp$date.block, sep = "_")
-# I use block as hour for the ts_wide()
-bb2020sp$date.block = ymd_h(paste(bb2020sp$date, bb2020sp$block))
+# drop w/e
+bb2020sp <- bb2020sp %>%
+  filter(bb2020sp$weekday!="Sat" & bb2020sp$weekday!="Sun")
 
@@
 ts.down <- bb2020sp %>%
-  group_by(date.block,LAD19NM) %>%
+  group_by(date,LAD19NM) %>%
   summarise(mean.down = mean(speeddown))#, mean.down = mean(speeddown), n.tests = n())
@@
 mean.down.uk <- ts.down %>%
-  group_by(date.block) %>%
+  group_by(date) %>%
   summarise(mean.down.uk = mean(mean.down))#, mean.down = mean(speeddown), n.tests = n())
@@
 
-ggplot(subset(ts.down,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.down, group = LAD19NM, colour = "red")) +
+ggplot(subset(ts.down,date > start.date), aes(x=as.POSIXct(date), y=mean.down, group = LAD19NM, colour = "red")) +
   geom_line() + guides(colour=FALSE) + ggtitle("Download speeds") + #xlab("") +
@@
   facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.down.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.down.uk, group = 1), colour="green") #
+  geom_line(data = subset(mean.down.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.down.uk, group = 1), colour="green") #
 
@@
 ts.up <- bb2020sp %>%
-  group_by(date.block,LAD19NM) %>%
+  group_by(date,LAD19NM) %>%
   summarise(mean.up = mean(speedup))
@@
 mean.up.uk <- ts.up %>%
-  group_by(date.block) %>%
+  group_by(date) %>%
   summarise(mean.up.uk = mean(mean.up))
@@
 
-ggplot(subset(ts.up,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.up, group = LAD19NM, colour = "red")) +
+ggplot(subset(ts.up,date > start.date), aes(x=as.POSIXct(date), y=mean.up, group = LAD19NM, colour = "red")) +
   geom_line() + guides(colour=FALSE) + ggtitle("Upload speeds") + #xlab("") +
@@
   facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.up.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.up.uk, group = 1), colour="green") #
+  geom_line(data = subset(mean.up.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.up.uk, group = 1), colour="green") #
 
@@
 ts.tests <- bb2020sp %>%
-  group_by(date.block,LAD19NM) %>%
+  group_by(date,LAD19NM) %>%
   summarise(n.tests = n())
@@
 mean.tests.uk <- ts.tests %>%
-  group_by(date.block) %>%
+  group_by(date) %>%
   summarise(mean.tests.uk = mean(n.tests))
@@
 
-ggplot(subset(ts.tests,date.block > start.date), aes(x=as.POSIXct(date.block), y=n.tests, group = LAD19NM, colour = "red")) +
+ggplot(subset(ts.tests,date > start.date), aes(x=as.POSIXct(date), y=n.tests, group = LAD19NM, colour = "red")) +
   geom_line() + guides(colour=FALSE) + ggtitle("N. of tests") + #xlab("") +
@@
   facet_wrap(~LAD19NM, ncol = 2, scales = "free_y") +
-  geom_line(data = subset(mean.tests.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.tests.uk, group = 1), colour="green") #.uk
+  geom_line(data = subset(mean.tests.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.tests.uk, group = 1), colour="green") #.uk
 
@@
 ts.down.diff <- ts.down
-ts.down.diff$date.block.lag1y <- ts.down.diff$date.block - years(1)
-ts.down.diff <- merge(ts.down.diff,ts.down.diff, by.x = c("date.block.lag1y", "LAD19NM"), by.y = c("date.block","LAD19NM"), all = F)
+ts.down.diff$date.lag1y <- ts.down.diff$date - years(1)
+ts.down.diff <- merge(ts.down.diff,ts.down.diff, by.x = c("date.lag1y", "LAD19NM"), by.y = c("date","LAD19NM"), all = F)
 
@@
   rename(mean.down2019 = mean.down.y, mean.down2020 = mean.down.x) %>%
-  select(-date.block.lag1y.y)
+  select(-date.lag1y.y)
 
@@
 mean.down.diff.uk <- ts.down.diff %>%
-  group_by(date.block) %>%
+  group_by(date) %>%
   summarise(mean.down.diff.uk = mean(diff))#, mean.down = mean(speeddown), n.tests = n())
@@
 
-ggplot(subset(ts.down.diff,date.block > start.date), aes(x=as.POSIXct(date.block), y=diff, group = LAD19NM, colour = "red")) +
+ggplot(subset(ts.down.diff,date > start.date), aes(x=as.POSIXct(date), y=diff, group = LAD19NM, colour = "red")) +
   geom_line() + guides(colour=FALSE) + ggtitle("Diff. in download speeds") + #xlab("") +
@@
   facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.down.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.down.uk, group = 1), colour="green") #
+  geom_line(data = subset(mean.down.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.down.uk, group = 1), colour="green") #
 ```
@@
 ts.up.diff <- ts.up
-ts.up.diff$date.block.lag1y <- ts.up.diff$date.block - years(1)
-ts.up.diff <- merge(ts.up.diff,ts.up.diff, by.x = c("date.block.lag1y", "LAD19NM"), by.y = c("date.block","LAD19NM"), all = F)
+ts.up.diff$date.lag1y <- ts.up.diff$date - years(1)
+ts.up.diff <- merge(ts.up.diff,ts.up.diff, by.x = c("date.lag1y", "LAD19NM"), by.y = c("date","LAD19NM"), all = F)
 
@@
   rename(mean.up2019 = mean.up.y, mean.up2020 = mean.up.x) %>%
-  select(-date.block.lag1y.y)
+  select(-date.lag1y.y)
 
@@
 mean.up.diff.uk <- ts.up.diff %>%
-  group_by(date.block) %>%
+  group_by(date) %>%
   summarise(mean.up.diff.uk = mean(diff))#, mean.up = mean(speedup), n.tests = n())
@@
 
-ggplot(subset(ts.up.diff,date.block > start.date), aes(x=as.POSIXct(date.block), y=diff, group = LAD19NM, colour = "red")) +
+ggplot(subset(ts.up.diff,date > start.date), aes(x=as.POSIXct(date), y=diff, group = LAD19NM, colour = "red")) +
   geom_line() + guides(colour=FALSE) + ggtitle("Diff. in upload speeds") + #xlab("") +
@@
   facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.up.diff.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.up.diff.uk, group = 1), colour="green") #
+  geom_line(data = subset(mean.up.diff.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.up.diff.uk, group = 1), colour="green") #
 ```
@@
 ts.tests.diff <- ts.tests
-ts.tests.diff$date.block.lag1y <- ts.tests.diff$date.block - years(1)
-ts.tests.diff <- merge(ts.tests.diff,ts.tests.diff, by.x = c("date.block.lag1y", "LAD19NM"), by.y = c("date.block","LAD19NM"), all = F)
+ts.tests.diff$date.lag1y <- ts.tests.diff$date - years(1)
+ts.tests.diff <- merge(ts.tests.diff,ts.tests.diff, by.x = c("date.lag1y", "LAD19NM"), by.y = c("date","LAD19NM"), all = F)
 
@@
   rename(n.tests2019 = n.tests.y, n.tests2020 = n.tests.x) %>%
-  select(-date.block.lag1y.y)
+  select(-date.lag1y.y)
 
@@
 mean.tests.diff.uk <- ts.tests.diff %>%
-  group_by(date.block) %>%
+  group_by(date) %>%
   summarise(mean.tests.diff.uk = mean(diff))
@@
 
-ggplot(subset(ts.tests.diff,date.block > start.date), aes(x=as.POSIXct(date.block), y=diff, group = LAD19NM, colour = "red")) +
+ggplot(subset(ts.tests.diff,date > start.date), aes(x=as.POSIXct(date), y=diff, group = LAD19NM, colour = "red")) +
   geom_line() + guides(colour=FALSE) + ggtitle("Diff. in speed tests") + #xlab("") +
@@
   facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.tests.diff.uk,date.block > start.date), aes(x=as.POSIXct(date.block), y=mean.tests.diff.uk, group = 1), colour="green") #
+  geom_line(data = subset(mean.tests.diff.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.tests.diff.uk, group = 1), colour="green") #
 ```

### 2021-09-22 src/old/ts_clusters.Rmd
tidy up

--- 2020-12-02 src/ts_clusters.Rmd
@@
 ggplot(la.f, aes(x = long, y = lat)) +
-  geom_polygon(aes( group = group, fill = as.factor(cluster.up))) +
-  theme_nothing(legend = TRUE) +
-  labs(title = "Upload clusters") +
-  scale_fill_viridis_d()+
-  theme_void()
-
-ggplot(la.f, aes(x = long, y = lat)) +
   geom_polygon(aes( group = group, fill = as.factor(cluster.down))) +
@@
 
+cluster.upL <- c(6, 9, 11, 12, 13)
+cluster.upS <- c(1, 2, 3, 7, 8, 10)
+
+la.f_ <- la.f #plot the background
+
+cluster.upL.plot <- la.f %>% 
+  filter(cluster.up %in% cluster.upL) %>% 
+  ggplot(aes(x = long, y = lat)) +
+  geom_polygon(data = la.f_, color='grey', fill='white', aes(x=long, y=lat, group=group, colour = 'grey30')) +
+  geom_polygon(aes( group = group, fill = as.factor(cluster.up))) +
+  labs(title = "Larger clusters, upload speeds") + 
+  scale_fill_viridis_d() +
+  guides(fill=guide_legend(title="Clusters")) +
+  theme_void()
+
+cluster.upS.plot <- la.f %>% 
+  filter(cluster.up %in% cluster.upS) %>% 
+  ggplot(aes(x = long, y = lat)) +
+  geom_polygon(data = la.f_, color='grey', fill='white', aes(x=long, y=lat, group=group, colour = 'grey30')) +
+  geom_polygon(aes( group = group, fill = as.factor(cluster.up))) +
+  labs(title = "Smaller clusters, upload speeds") + 
+  scale_fill_viridis_d() +
+  guides(fill=guide_legend(title="Clusters")) +
+  theme_void()
+
+plot_grid(cluster.upL.plot, cluster.upS.plot, nrow = 1,
+          align = "h", axis = "l", 
+          #labels = c(".co.uk / pop., 2000", ".co.uk / pop., 2010", "non .co.uk / pop., 2000", "non .co.uk / pop., 2010", ""), 
+          #label_size = 12,
+          greedy = T)
+          #rel_heights = c(1, 1, .3)) # this is the relative position of the 3 lines
+
+#ggsave("all_maps2.png") # this is the colour map for the preprint
+out.path.map <- paste0(path, "/paper/v1/figures/map.up.clusters.png")
+ggsave(out.path.map)#,
+       #width = 210, #A4
+       #height = 297, #A4
+       #units = "mm")
 ```

### 2021-09-22 src/old/ts_clusters_k11.Rmd
tidy up

--- 2021-03-04 src/ts_clusters_k5.Rmd
@@
 ---
-title: "Time series clusters, k = 5"
+title: "Time series clusters, k = 11"
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
@@
 # Define k instead of selecting it
-k <- 5
+k <- 11
 
@@
 # save clusters file
-out.clusters <- paste(path, "/data/temp/clusters_all_5.csv", sep = "")
+out.clusters <- paste(path, "/data/temp/clusters_all_11.csv", sep = "")
 write_csv(clusters, out.clusters)
@@
 # save clusters file
-out.clusters <- paste(path, "/data/temp/clusters_nodiff_5.csv", sep = "")
+out.clusters <- paste(path, "/data/temp/clusters_nodiff_11.csv", sep = "")
 write_csv(clusters.nodiff, out.clusters)
@@
 #ggsave("all_maps2.png") # this is the colour map for the preprint
-# out.path.map <- paste0(path, "/paper/v1/figures/map.up.clusters5.png")
+# out.path.map <- paste0(path, "/paper/v1/figures/map.up.clusters11.png")
 # ggsave(out.path.map)#,

### 2021-09-22 src/old/ts_clusters_k7.Rmd
tidy up

--- 2021-03-04 src/ts_clusters_k5.Rmd
@@
 ---
-title: "Time series clusters, k = 5"
+title: "Time series clusters, k = 7"
 date: "`r format(Sys.time(), '%d %B, %Y, %H:%M')`"
@@
 # Define k instead of selecting it
-k <- 5
+k <- 7
 
@@
 # save clusters file
-out.clusters <- paste(path, "/data/temp/clusters_all_5.csv", sep = "")
+out.clusters <- paste(path, "/data/temp/clusters_all_7.csv", sep = "")
 write_csv(clusters, out.clusters)
@@
 # save clusters file
-out.clusters <- paste(path, "/data/temp/clusters_nodiff_5.csv", sep = "")
+out.clusters <- paste(path, "/data/temp/clusters_nodiff_7.csv", sep = "")
 write_csv(clusters.nodiff, out.clusters)
@@
 #ggsave("all_maps2.png") # this is the colour map for the preprint
-# out.path.map <- paste0(path, "/paper/v1/figures/map.up.clusters5.png")
+# out.path.map <- paste0(path, "/paper/v1/figures/map.up.clusters7.png")
 # ggsave(out.path.map)#,

### 2021-09-22 src/old/ts_clusters_old.Rmd
tidy up

--- 2021-09-22 src/old/timeseries_plots0916.Rmd
@@
                  border = rgb(0.8, 0.8, 0.8, 0))
-# drop LA
-rm(la)
+
 ```
 
-## Keep working hours
+## Keep working hours and days
+
+The time series are based on mean upload and download speeds as well as the total number of tests for working days and during working hours
 
@@
 
-## Download speeds
+## Time series clusters
 
-```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
-ts.down <- bb2020sp %>%
-  group_by(date,LAD19NM) %>%
+I create clusters based on upload, download and number of tests time series for tests working hours and days from 16/2/2020 onwards.
+
+More work needs to be done to select the optimal *k* as well as the clustering methods.
+The `dtwclust` package comes with quite some readings, saved in the `.\literature\`.
+
+Also, we need to think about upload / download / differnces / period + **imputation**.
+
+```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# download
+ts.down <- TS2020 %>%
+  group_by(weekday, hour, LAD19NM) %>%
   summarise(mean.down = mean(speeddown))#, mean.down = mean(speeddown), n.tests = n())
-sapply(ts.down, function(x) sum(is.na(x)))
 
-# https://www.data-to-viz.com/caveat/spaghetti.html
+# upload
+ts.up <- TS2020 %>%
+  group_by(weekday, hour, LAD19NM) %>%
+  summarise(mean.up = mean(speedup))
 
-mean.down.uk <- ts.down %>%
-  group_by(date) %>%
-  summarise(mean.down.uk = mean(mean.down))#, mean.down = mean(speeddown), n.tests = n())
+# tests
+ts.tests <- TS2020 %>%
+  group_by(weekday, hour, LAD19NM) %>%
+  summarise(n.tests = n())
+```
 
-start.date =  "2019-12-31"
+### Upload speed
 
-ggplot(subset(ts.down,date > start.date), aes(x=as.POSIXct(date), y=mean.down, group = LAD19NM, colour = "red")) +
-  geom_line() + guides(colour=FALSE) + ggtitle("Download speeds") + #xlab("") +
-  ylab("download speed") +
-  scale_y_continuous(labels = scales::comma) +
-  scale_x_datetime(date_breaks = "1 month") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
-        plot.title = element_text(hjust = 0.5)) +
-  facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.down.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.down.uk, group = 1), colour="green") #
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# upload working hours and days
+
+ts.wide <- ts.up %>%
+  select(date, LAD19NM, mean.up) %>%
+  filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide <- ts_wide(ts.wide)
+
+# impute with means
+library(imputeTS)
+ts.wide <- na_mean(ts.wide)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# extract names
+ts.wide.names <- names(ts.wide[,-1])
+ts.wide <- ts.wide[,-1]
+
+# standardise
+ts.wide <- zscore(ts.wide)
+
+# convert to list for tsclust
+ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
+
+# 5 - 20 cluster sollutions
+pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
+                type="partitional",
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+# decide k, see RJ-2019-023 for max/min
+names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+sapply(pc_k, cvi, type = "internal")
+
+# I need to work on the selection
+
+pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+plot(pc_k)
+
+cluster_up <- data.frame(LAD=ts.wide.names,
+                           cluster=pc_k@cluster)
+```
+
+### Download speed
+
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# download working hours and days
+
+ts.wide <- ts.down %>%
+  select(date, LAD19NM, mean.down) %>%
+  filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide <- ts_wide(ts.wide)
+
+# impute with means
+library(imputeTS)
+ts.wide <- na_mean(ts.wide)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# extract names
+ts.wide.names <- names(ts.wide[,-1])
+ts.wide <- ts.wide[,-1]
+
+# standardise
+ts.wide <- zscore(ts.wide)
+
+# convert to list for tsclust
+ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
+
+# 5 - 20 cluster sollutions
+pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
+                type="partitional",
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+# decide k, see RJ-2019-023 for max/min
+names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+sapply(pc_k, cvi, type = "internal")
+
+# I need to work on the selection
+
+pc_k <- tsclust(ts.wide, k = 20, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+plot(pc_k)
+
+cluster_down <- data.frame(LAD=ts.wide.names,
+                         cluster=pc_k@cluster)
+```
+
+### N. of speeds
+
+```{r eval=TRUE, echo=TRUE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+# tests working hours and days
+
+ts.wide <- ts.tests %>%
+  select(date, LAD19NM, n.tests) %>%
+  filter(date > "2020-2-15")
+
+# turn to a wide format
+ts.wide <- ts_wide(ts.wide)
+
+# impute with means
+library(imputeTS)
+ts.wide <- na_mean(ts.wide)
+#sapply(ts.wide, function(x) sum(is.na(x)))
+
+# extract names
+ts.wide.names <- names(ts.wide[,-1])
+ts.wide <- ts.wide[,-1]
+
+# standardise
+ts.wide <- zscore(ts.wide)
+
+# convert to list for tsclust
+ts.wide <- split(ts.wide, rep(1:ncol(ts.wide), each = nrow(ts.wide)))
+
+# 5 - 20 cluster sollutions
+pc_k <- tsclust(ts.wide, k = c(5,10,15,20), seed = 94L,
+                type="partitional",
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+# decide k, see RJ-2019-023 for max/min
+names(pc_k) <- paste0("k_", c("5","10","15", "20"))
+sapply(pc_k, cvi, type = "internal")
+
+# I need to work on the selection
+
+pc_k <- tsclust(ts.wide, k = 15, seed = 94L, 
+                type="partitional", 
+                distance = "dtw_basic", centroid = "pam", trace = T, 
+                args = tsclust_args(dist = list(window.size = 20L)))
+
+plot(pc_k)
+
+cluster_tests <- data.frame(LAD=ts.wide.names,
+                           cluster=pc_k@cluster)
+
+clusters <- merge(cluster_up, cluster_down, by = "LAD")
+names(clusters)[2] <- "cluster.up"
+names(clusters)[3] <- "cluster.down"
+clusters <- merge(clusters, cluster_tests, by = "LAD")
+names(clusters)[4] <- "cluster.tests"
+#sapply(clusters, function(x) sum(is.na(x)))
+
+clusters  %>% arrange(cluster.up) %>% kable() 
+```
+
+## Mapping the clusters
+
+```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=10, fig.width=10}
+
+la@data <- merge(la@data, clusters, by.x = "LAD19NM", by.y = "LAD")
+
+var <- la@data[ ,'cluster.up']
+plot(la, col = var, main = "Upload clusters",
+     border = rgb(0.8, 0.8, 0.8, 0))
+
+var <- la@data[ ,'cluster.down']
+plot(la, col = var, main = "Download clusters",
+     border = rgb(0.8, 0.8, 0.8, 0))
+
+var <- la@data[ ,'cluster.tests']
+plot(la, col = var, main = "Tests clusters",
+     border = rgb(0.8, 0.8, 0.8, 0))
 
 ```
-
-## Upload speeds
-
-```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
-ts.up <- bb2020sp %>%
-  group_by(date,LAD19NM) %>%
-  summarise(mean.up = mean(speedup))
-sapply(ts.up, function(x) sum(is.na(x)))
-
-# https://www.data-to-viz.com/caveat/spaghetti.html
-
-mean.up.uk <- ts.up %>%
-  group_by(date) %>%
-  summarise(mean.up.uk = mean(mean.up))
-
-start.date =  "2019-12-31"
-
-ggplot(subset(ts.up,date > start.date), aes(x=as.POSIXct(date), y=mean.up, group = LAD19NM, colour = "red")) +
-  geom_line() + guides(colour=FALSE) + ggtitle("Upload speeds") + #xlab("") +
-  ylab("upload speed") +
-  scale_y_continuous(labels = scales::comma) +
-  scale_x_datetime(date_breaks = "1 month") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
-        plot.title = element_text(hjust = 0.5)) +
-  facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.up.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.up.uk, group = 1), colour="green") #
-
-```
-
-## Speed tests
-
-```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
-ts.tests <- bb2020sp %>%
-  group_by(date,LAD19NM) %>%
-  summarise(n.tests = n())
-sapply(ts.tests, function(x) sum(is.na(x)))
-
-# https://www.data-to-viz.com/caveat/spaghetti.html
-
-# n.tests.uk <- ts.tests %>%
-#   group_by(date.block) %>%
-#   summarise(n.tests = n()/382)
-
-mean.tests.uk <- ts.tests %>%
-  group_by(date) %>%
-  summarise(mean.tests.uk = mean(n.tests))
-
-start.date =  "2019-12-31"
-
-ggplot(subset(ts.tests,date > start.date), aes(x=as.POSIXct(date), y=n.tests, group = LAD19NM, colour = "red")) +
-  geom_line() + guides(colour=FALSE) + ggtitle("N. of tests") + #xlab("") +
-  ylab("tests") +
-  scale_y_continuous(labels = scales::comma) +
-  scale_x_datetime(date_breaks = "1 month") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
-        plot.title = element_text(hjust = 0.5)) +
-  facet_wrap(~LAD19NM, ncol = 2, scales = "free_y") +
-  geom_line(data = subset(mean.tests.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.tests.uk, group = 1), colour="green") #.uk
-
-```
-
-## Download diff 2020 - 2019
-
-```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
-
-ts.down.diff <- ts.down
-ts.down.diff$date.lag1y <- ts.down.diff$date - years(1)
-ts.down.diff <- merge(ts.down.diff,ts.down.diff, by.x = c("date.lag1y", "LAD19NM"), by.y = c("date","LAD19NM"), all = F)
-
-ts.down.diff <- ts.down.diff %>%
-  rename(mean.down2019 = mean.down.y, mean.down2020 = mean.down.x) %>%
-  select(-date.lag1y.y)
-
-ts.down.diff$diff <- ts.down.diff$mean.down2020 - ts.down.diff$mean.down2019
-
-head(ts.down.diff)
-sapply(ts.down.diff, function(x) sum(is.na(x)))
-
-# https://www.data-to-viz.com/caveat/spaghetti.html
-
-mean.down.diff.uk <- ts.down.diff %>%
-  group_by(date) %>%
-  summarise(mean.down.diff.uk = mean(diff))#, mean.down = mean(speeddown), n.tests = n())
-
-start.date =  "2019-12-31"
-
-ggplot(subset(ts.down.diff,date > start.date), aes(x=as.POSIXct(date), y=diff, group = LAD19NM, colour = "red")) +
-  geom_line() + guides(colour=FALSE) + ggtitle("Diff. in download speeds") + #xlab("") +
-  ylab("diff. in download speed") +
-  scale_y_continuous(labels = scales::comma) +
-  scale_x_datetime(date_breaks = "1 month") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
-        plot.title = element_text(hjust = 0.5)) +
-  facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.down.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.down.uk, group = 1), colour="green") #
-```
-
-## Upload diff 2020 - 2019
-
-```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
-
-ts.up.diff <- ts.up
-ts.up.diff$date.lag1y <- ts.up.diff$date - years(1)
-ts.up.diff <- merge(ts.up.diff,ts.up.diff, by.x = c("date.lag1y", "LAD19NM"), by.y = c("date","LAD19NM"), all = F)
-
-ts.up.diff <- ts.up.diff %>%
-  rename(mean.up2019 = mean.up.y, mean.up2020 = mean.up.x) %>%
-  select(-date.lag1y.y)
-
-ts.up.diff$diff <- ts.up.diff$mean.up2020 - ts.up.diff$mean.up2019
-
-head(ts.up.diff)
-sapply(ts.up.diff, function(x) sum(is.na(x)))
-
-# https://www.data-to-viz.com/caveat/spaghetti.html
-
-mean.up.diff.uk <- ts.up.diff %>%
-  group_by(date) %>%
-  summarise(mean.up.diff.uk = mean(diff))#, mean.up = mean(speedup), n.tests = n())
-
-start.date =  "2019-12-31"
-
-ggplot(subset(ts.up.diff,date > start.date), aes(x=as.POSIXct(date), y=diff, group = LAD19NM, colour = "red")) +
-  geom_line() + guides(colour=FALSE) + ggtitle("Diff. in upload speeds") + #xlab("") +
-  ylab("diff. in upload speed") +
-  scale_y_continuous(labels = scales::comma) +
-  scale_x_datetime(date_breaks = "1 month") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
-        plot.title = element_text(hjust = 0.5)) +
-  facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.up.diff.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.up.diff.uk, group = 1), colour="green") #
-```
-
-## Speed tests diff. 2020 - 2019
-
-```{r include=TRUE, echo=FALSE, results= 'markup', message=FALSE, warning = FALSE, fig.height=600, fig.width=14}
-ts.tests.diff <- ts.tests
-ts.tests.diff$date.lag1y <- ts.tests.diff$date - years(1)
-ts.tests.diff <- merge(ts.tests.diff,ts.tests.diff, by.x = c("date.lag1y", "LAD19NM"), by.y = c("date","LAD19NM"), all = F)
-
-ts.tests.diff <- ts.tests.diff %>%
-  rename(n.tests2019 = n.tests.y, n.tests2020 = n.tests.x) %>%
-  select(-date.lag1y.y)
-
-ts.tests.diff$diff <- ts.tests.diff$n.tests2020 - ts.tests.diff$n.tests2019
-
-head(ts.tests.diff)
-sapply(ts.tests.diff, function(x) sum(is.na(x)))
-
-mean.tests.diff.uk <- ts.tests.diff %>%
-  group_by(date) %>%
-  summarise(mean.tests.diff.uk = mean(diff))
-
-start.date =  "2019-12-31"
-
-ggplot(subset(ts.tests.diff,date > start.date), aes(x=as.POSIXct(date), y=diff, group = LAD19NM, colour = "red")) +
-  geom_line() + guides(colour=FALSE) + ggtitle("Diff. in speed tests") + #xlab("") +
-  ylab("diff. in speed tests") +
-  scale_y_continuous(labels = scales::comma) +
-  scale_x_datetime(date_breaks = "1 month") +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.title.x = element_blank(),
-        plot.title = element_text(hjust = 0.5)) +
-  facet_wrap(~LAD19NM, ncol = 2) +
-  geom_line(data = subset(mean.tests.diff.uk,date > start.date), aes(x=as.POSIXct(date), y=mean.tests.diff.uk, group = 1), colour="green") #
-```

### 2021-09-22 src/bbData2020.R

@@ -1,270 +1,0 @@
-#key libraries
-library(dplyr)
-library(reshape2)
-library(lubridate)
-library(ggplot2)
-
-#upload data file
-bb2020 <- read.csv("./data/raw/speedtest.csv")
-#name columns
-names(bb2020)[1:6] <- c("datetext","speeddown","speedup","provider","lat", "lon") 
-
-str(bb2020)
-summary(bb2020) #2030205 obs
-#conversions needed to numeric? df$col <- as.numeric(df$col) 
-#remove empty rows or coerced to NA? df[complete.cases(df),] or df <- df[!is.na(df$col),]
-
-#remove outliers based on Riddlesden and Singleton 2014 for slower end
-#check fastest upload and download speeds commercially available from Virgin Media (Gigaclear offers even faster!)
-bb2020 <- bb2020[bb2020$speeddown>512,] 
-bb2020 <- bb2020[bb2020$speeddown<362000,]
-bb2020 <- bb2020[bb2020$speedup<21000,] #1839332 obs
-
-#remove below seconds, create columns for date and hour
-bb2020$datetext <- strtrim(x = bb2020$datetext, width = 19) 
-bb2020$datetext2 <- bb2020$datetext
-bb2020$datetext2 <- strtrim(x = bb2020$datetext2, width = 13)
-bb2020$datetext2 <- colsplit(bb2020$datetext2, pattern = " ", c("date", "hour"))
-bb2020$hour <- bb2020$datetext2$hour
-bb2020$date <- bb2020$datetext2$date
-#delete datetext2 column(s)
-bb2020 <- bb2020[ ,c(1:6,8:9)]
-#convert to date form datetext and date columns
-bb2020$datetext <- as.POSIXct(strptime(bb2020$datetext, format = "%Y-%m-%d %H:%M:%S", "GMT")) 
-bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
-
-#add library to explore data temporally
-library(openair)
-#create dataframe with number of tests, mean and standard deviation by month for adding to dataframe later
-MonthFreq <- timeAverage(bb2020, avg.time = "month", statistic = "frequency")
-MonthMean <- timeAverage(bb2020, avg.time = "month", statistic = "mean")
-MonthSD <- timeAverage(bb2020, avg.time = "month", statistic = "sd")
-MonthStats <- cbind(MonthFreq[,1:2], MonthMean[,c(3:4)], 
[…]

### 2021-09-22 src/old/bbData2020.R

@@ -1,0 +1,270 @@
+#key libraries
+library(dplyr)
+library(reshape2)
+library(lubridate)
+library(ggplot2)
+
+#upload data file
+bb2020 <- read.csv("./data/raw/speedtest.csv")
+#name columns
+names(bb2020)[1:6] <- c("datetext","speeddown","speedup","provider","lat", "lon") 
+
+str(bb2020)
+summary(bb2020) #2030205 obs
+#conversions needed to numeric? df$col <- as.numeric(df$col) 
+#remove empty rows or coerced to NA? df[complete.cases(df),] or df <- df[!is.na(df$col),]
+
+#remove outliers based on Riddlesden and Singleton 2014 for slower end
+#check fastest upload and download speeds commercially available from Virgin Media (Gigaclear offers even faster!)
+bb2020 <- bb2020[bb2020$speeddown>512,] 
+bb2020 <- bb2020[bb2020$speeddown<362000,]
+bb2020 <- bb2020[bb2020$speedup<21000,] #1839332 obs
+
+#remove below seconds, create columns for date and hour
+bb2020$datetext <- strtrim(x = bb2020$datetext, width = 19) 
+bb2020$datetext2 <- bb2020$datetext
+bb2020$datetext2 <- strtrim(x = bb2020$datetext2, width = 13)
+bb2020$datetext2 <- colsplit(bb2020$datetext2, pattern = " ", c("date", "hour"))
+bb2020$hour <- bb2020$datetext2$hour
+bb2020$date <- bb2020$datetext2$date
+#delete datetext2 column(s)
+bb2020 <- bb2020[ ,c(1:6,8:9)]
+#convert to date form datetext and date columns
+bb2020$datetext <- as.POSIXct(strptime(bb2020$datetext, format = "%Y-%m-%d %H:%M:%S", "GMT")) 
+bb2020$date <- as.POSIXct(strptime(bb2020$date, format = "%Y-%m-%d", "GMT"))
+
+#add library to explore data temporally
+library(openair)
+#create dataframe with number of tests, mean and standard deviation by month for adding to dataframe later
+MonthFreq <- timeAverage(bb2020, avg.time = "month", statistic = "frequency")
+MonthMean <- timeAverage(bb2020, avg.time = "month", statistic = "mean")
+MonthSD <- timeAverage(bb2020, avg.time = "month", statistic = "sd")
+MonthStats <- cbind(MonthFreq[,1:2], MonthMean[,c(3:4)], 
[…]
